### 2023-05-17 doc/SAP/sap.tex
SAP version 0.4

\documentclass{article}

% spelling
\usepackage[english]{babel}

% page layout
\usepackage[letterpaper,top=2cm,bottom=2cm,left=2cm,right=2cm,marginparwidth=1.75cm]{geometry}

% math
\usepackage{amsmath}
\usepackage{amssymb}

% layout
\setlength\parindent{0pt}
\usepackage{setspace}
\usepackage{parskip}
\onehalfspacing
\usepackage{graphicx}

% referencing
\usepackage[colorlinks=true, allcolors=black]{hyperref}
\usepackage[capitalize, nameinlink]{cleveref}
\crefdefaultlabelformat{#2\textbf{#1}#3} % 
\crefname{figure}{\textbf{Figure}}{\textbf{Figures}}
\Crefname{table}{\textbf{Table}}{\textbf{Tables}}

% captions
\usepackage[labelfont=bf]{caption}
\usepackage[table]{xcolor}

% tables
\usepackage{booktabs}
\renewcommand{\arraystretch}{1.25} % vertical spacing

\begin{document}

\begin{titlepage}
        \vspace*{1cm}

        \LARGE
        \textbf{Transmission of airborne respiratory infections with and without air cleaners in a school in Switzerland: A modeling study of epidemiological, environmental, and molecular data}

        \vspace{0.5cm}
        \Large
        Statistical analysis plan (SAP)

        \vspace{1.5cm}

        \textbf{SAP authors:} Nicolas Banholzer$^{1*}$, Lukas Fenner$^1$ \\
        \textbf{Study authors:} Nicolas Banholzer$^{1*}$, Philipp Jent$^{2}$, Pascal Bittel$^{2}$, Lavinia Furrer$^{2}$, Tina Hascher$^{3}$, Lukas Fenner$^1$

        \vspace{1cm}

        $^1$ Institute of Social and Preventive Medicine, University of Bern, Switzerland \\
        $^2$ Department of Infectious Diseases, University of Bern, Switzerland \\
        $^3$ Institute of Educational Science, University of Bern, Switzerland \\
        $^*$ Corresponding author: nicolas.banholzer@unibe.ch

        \vfill

        \Large
        Version number: 0.4 \\
        Version date: \today

        \vspace*{1cm}
\end{titlepage}

\tableofcontents

\clearpage

\section{Introduction}

This document describes the proposed presentation and analysis for the main paper(s) reporting results from a study estimating and comparing the transmission of airborne respiratory infections with and without air cleaners in schools. To study the effectiveness of air cleaners as an infection control measure, we installed air cleaners in the classrooms of two classes in a secondary school in the Canton of Solothurn, Switzerland (\Cref{fig:study_setting}). Over the study period between January 16 and March 11, 2023, we collected the following data (\Cref{tab:data}):

\begin{enumerate}
    \item \textbf{Molecular data}: Bioaerosol and human saliva samples.
    \item \textbf{Environmental data}: CO$_2$, aerosol and particle concentrations.
    \item \textbf{Epidemiological data}: Daily absences from school.
\end{enumerate}

Laboratory and molecular analysis will detect which respiratory infections (e.g. SARS-CoV-2 or Influenza virus) were spreading over the study. Molecular data (positive samples), environmental (particle concentrations), and epidemiological data (absences related to respiratory infections) will then be compared between study conditions with and without air cleaners. The aim of the comparison is to assess whether air cleaners reduced the risk of airborne infection. \smallskip

Any deviations from the statistical analysis plan will be described and justified in the final report. The analysis will be carried out by identified, appropriately qualified, and experienced statisticians, who will ensure the integrity of the data during their processing.

\begin{figure}[!htpb]
    \centering
    \caption{\textbf{Study setting.} Schematic study setup of classrooms where molecular and environmental data were collected. One air cleaner was placed in the front and the other in the back of the classrooms. All devices were placed at the head level of students when they were seated. Both classrooms were not equipped with an active HVAC (Heating, Ventilation, Air conditioning) system, but were ventilated using passive window ventilation.}
    \includegraphics{study_setting.pdf}
    \label{fig:study_setting}
\end{figure}

\begin{table}[!htpb]
    \footnotesize
    \centering
    \caption{\textbf{Data}. Type of data collected, method/device, and frequency in each class.}
    \begin{tabular}{p{3.5cm}p{6cm} p{1cm} p{1cm} p{3cm}}
    \midrule
    Type of data & Method / device & Class A & Class B & Frequency \\
    \midrule
    \textbf{Molecular data} \\
    \midrule
    Bioaerosol sampling & Bioaerosol sampling devices: BioSpot-VIVAS (Aerosol Devices Inc., Ft. Collins, Colorado, United States) and Coriolis Micro Air (Bertin Instruments Montigny-le/Bretonneux, France) & x & (x) \newline only Coriolis & daily \\
    Saliva samples & Saliva samples & x & x & twice per week \\
    Swabs from air cleaners & Swab samples from Xiaomi HEPA filters (Xiaomi Mi Air Pro 70m2, Shenzhen, China) & x & x & after each intervention phase (both classes) and before the vacation (only class B) \\ 
    \midrule
    \textbf{Environmental data} \\
    \midrule
    CO$_2$, aerosol/particle concentrations, humidity, temperature & Air quality device: AQ Guard (Palas GmbH, Karlsruhe, Germany) & x & x & daily by minute \\
    Coughs & Sounds collected via microphones and coughs detected with a Machine Learning model\cite{Bertschinger2023} & x & x & daily by seconds \\
    \midrule
    \textbf{Epidemiological data} \\
    \midrule
    Absences & Survey & x & x & daily \\
    Student characteristics & Survey & x & x & at study start \\
    \bottomrule
    \end{tabular}
    \label{tab:data}
\end{table}

\section{Background}

In a previous study, we estimated and compared the transmission of airborne infections with and without air cleaners and mask mandates in five classes of two secondary schools in the Canton of Solothurn, Switzerland, between January and March 2022\cite{Banholzer2023}. Air cleaners (commercially available portable HEPA-filtration devices) were found to reduce particle concentrations, but no association was found with transmission. In the previous study, air cleaners were introduced towards the end of the study when most students were already infected with SARS-CoV-2, thus limiting their potential to reduce transmission. By contrast, in this study, we only assess the effectiveness of air cleaners and used a more efficient cross-over design (\cref{tab:study_design}) to assess whether air cleaners are associated with reduced risk of airborne infection in schools.

\begin{table}[!htpb]
    \footnotesize
    \centering
    \caption{\textbf{Study design.} Seven-week study period from January 16 to March 11, 2023, excluding a week of vacation between February 6 and 11. Cross-over design where two portable air cleaners were installed during different time periods in each class (Class A: weeks 1, 2, 6, and 7; Class B: weeks 3, 4, and 5) but not at the same time in both classes.}
    \begin{tabular}{l l l l l l l l l}
    \toprule
      & \textbf{Week 1} & \textbf{Week 2} & \textbf{Week 3} & Vacation & \textbf{Week 4} & \textbf{Week 5} & \textbf{Week 6} & \textbf{Week 7} \\
      & Jan 16-21 & Jan 23-28 & Jan 30-Feb 3 & Feb 6-11 & Feb 13-18 & Feb 20-25 & Feb 27-March 3 & March 6-11 \\
      \midrule
      \textbf{Class A} & \cellcolor{gray!50} Air cleaner & \cellcolor{gray!50} Air cleaner & \cellcolor{gray!10} None & & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} Air cleaner & \cellcolor{gray!50} Air cleaner \\
      \textbf{Class B} & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} Air cleaner & & \cellcolor{gray!50} Air cleaner & \cellcolor{gray!50} Air cleaner & \cellcolor{gray!10} None & \cellcolor{gray!10} None \\
      \bottomrule
    \end{tabular}
    \label{tab:study_design}
\end{table}

\subsection{Objectives of the study}

The primary objective of the study is to estimate and compare the risk of airborne infection in schools. To this end, the following comparisons between study conditions (with and without air cleaners) are of interest:

\begin{enumerate}
    \item \textbf{Molecular data}: Comparison of positive bioaerosol/saliva samples and viral load concentration.
    \item \textbf{Environmental data}: Comparison of aerosol/particle concentrations.
    \item \textbf{Epidemiological data}: Comparison of the number of infections.  
\end{enumerate}

\subsection{Definitions}

\subsubsection{Case of respiratory infection}

Based on epidemiological data, we will define a case of respiratory infection as an absence where the student reports (a)~a positive lab test result confirming the respiratory infection, or (b)~a sickness with at least one of the following symptoms: fever, coughing, tiredness, loss of test or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, stomach.

Students were asked to report the first date when they experienced symptoms. For example, a student absent on Monday may report Saturday or Sunday as the symptom onset date. We will denote cases by their symptom onset date. Note that, during the week, the symptom onset date will usually correspond to the absence date, unless the student went to school despite experiencing symptoms.

\subsection{Primary outcome}

The number of respiratory cases (absences related to respiratory infections by date of symptom onset).

\subsection{Secondary outcomes}

\begin{enumerate}
    \item The number of positive bioaerosol and saliva samples.
    \item Viral load concentration in positive bioaerosol and saliva samples.
    \item The aerosol/particle number, and matter mass concentrations.
    \item The number of coughs.
\end{enumerate}

\subsection{Variables}

\begin{enumerate}
    \item Molecular data
    \begin{enumerate}
        \item Number and proportion (\%) of positive saliva and bioaerosol samples
        \item Viral load concentration (viral genome load in copies per liter)
        \item Spectrum of viral pathogens detected in bioaerosol and saliva samples, and on the HEPA filters of the air cleaners
    \end{enumerate}
    \item Environmental data 
    \begin{enumerate}
        \item CO$_2$ (parts per million [ppm])
        \item Relative humidity (\%) and temperature (°C)
        \item Particle number (in 1/cm$^3$, particle diameter between 175nm to 20$\mu$m) and matter mass concentrations (PM in $\mu$gm$^{-3}$; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$, i.e. particles of sizes $<$1 to $<$10$\mu$m)
        \item Number of detected coughs
    \end{enumerate}
    \item Epidemiological data
    \begin{enumerate}
        \item Aggregated (baseline) data per class
        \begin{enumerate}
            \item Number of students per class
            \item Vaccination and recovery status regarding COVID-19 of both students and teachers
            \item Age and gender
        \end{enumerate}
        \item Daily reporting per class
        \begin{enumerate}
            \item Number of absent students and teachers by sex
            \item Reason for absence: sickness, career days, or other
            \item Symptoms: fever, coughing, tiredness, loss of test or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, stomach, or other
            \item Date of symptom onset
            \item Laboratory test performed for respiratory viruses: yes, no, or unknown
            \item Date when the laboratory test was taken
            \item Viruses tested: COVID-19, Influenza A and B, RSV, or other
            \item Laboratory test result: positive or negative
            \item Date when the absent student came back to school
        \end{enumerate}
    \end{enumerate}
    \item Additional data (collected daily at 10min intervals)
    \begin{enumerate}
        \item Working atmosphere: silent working, soft speaking, or loud speaking 
        \item Ventilation: times of window or door opening
        \item Occupancy: breaks, lessons outside the classroom, lessons partly outside the classroom, or school-free hours
    \end{enumerate}
    \item Secondary data
    \begin{enumerate}
        \item Weekly number cases and deaths, laboratory tests and share of positive tests for COVID-19 in the Canton of Solothurn \cite{FOPH2023-cases}
        \item Weekly number of consultations per 100,000 people due to influenza-like illnesses (ILI) \cite{FOPH2023-consultations}
    \end{enumerate}
\end{enumerate}

\subsection{Hypothesis framework}

The null hypothesis will be that there is no true difference in the outcomes between study conditions. We will use a Bayesian framework to assess the null hypothesis and use the posterior distribution of the parameter modeling the difference in the outcome by study condition to assess the direction and magnitude of the effects of air cleaners.

\section{Study populations}

The target population includes all students from two classes in a secondary school in the Canton of Solothurn, Switzerland. In total, there were 38 students in both classes (Class A: 20, Class B: 18).

\section{Statistical analyses}

\subsection{Comparative analysis}

We will employ descriptive statistics to summarize epidemiological, environmental, and molecular data. Categorical variables will be presented as the number and percentage of students, while continuous variables will be reported with the median and interquartile range (IQR). We will create summary tables to describe student characteristics (age and gender), vaccination status, recovery status, absences, number and share of positive samples, and viral load concentration. We will create figures (box-, bar- and line plots) to compare the share of positive samples, viral load concentration, and particle concentration.

\subsection{Primary analysis}

\subsubsection*{Linking observed cases to unobserved infections}

The number of respiratory cases (absences related to respiratory infections by date of symptom onset) will be modeled with a Bayesian latent hierarchical regression model, which is similar to the model used in our previous study \Cref{Banholzer2023}. The model links the unobserved number of new infections to the observed number of new cases. The number of new cases in class $j$ at time $t$ is modeled with a Negative Binomial distribution   
\begin{align*}
    C_{jt} \sim \text{Negative-Binomial}(\mu_{jt},\phi),
\end{align*}
where $\mu_{jt}$ is the expected number of new cases and $\phi$ is the parameter modeling over-dispersion. The expected number of new cases is the weighted sum of the number of new infections $I_{jt}$ in the  previous days
\begin{align*}
    \mu_{jt} = \sum_{s<t}I_{js} \cdot p_{\text{IN}}(t-s),
\end{align*}
where $p_{\text{IN}}(t-s)$ denotes the probability distribution of the incubation period. The  number of new infections is related to the presence of air cleaners using a log-link
\begin{align*}
    \log I_{jt} &= \log N_{jt} + \beta_0 + \beta_1 \cdot \text{AirCleaner}_{jt} + \beta_2 \cdot \text{Class}_t + \beta_3 \cdot \text{Students}_{jt} + \beta_4 \cdot \text{Ventilation}_{jt} + \beta_5 \cdot \text{CoV}_t + \beta_6 \cdot \text{ILI}_t,
\end{align*}
where $N_{jt} = \sum_{s<t} I_{js}$ is the cumulative number of infections (model offset), $\beta_0$ is the rate of new infections without air cleaners (model intercept), and $\beta_1$ is the effect of air cleaners. The effect of air cleaners will be adjusted for class-specific effects, the number of students in school, the ventilation rate (air changes per hour computed from indoor CO$_2$ levels using the transient mass balance method\cite{Battermann2017}), the proportion of positive tests for SARS-CoV-2 in the community (from secondary data), and the number of consultations for influenza-like illnesses in the community (from secondary data).

\subsubsection*{Specifying the distribution of the incubation period}

The pathogen of each respiratory infection cannot be identified from the epidemiological data unless students obtained a laboratory test result. As a consequence, different incubation periods will be considered in $p_{\text{IN}}$, reflecting a combination of the pathogen-specific incubation periods. The combination will be determined based on the proportion of positive bioaerosol/saliva samples for each pathogen found in the molecular analysis. The prior distributions for pathogen-specific incubation periods will be based on published estimates.

\subsubsection*{Adjusting for under-reporting of cases on weekends}

Our previous study recorded a higher proportion of cases on Mondays than on weekends, suggesting recall bias and under-reporting of cases on weekends\cite{Banholzer2023}. To consider weekday effects in the reporting of cases, we will re-weight the expected number of cases each week as follows. Let $k \in $(1: Saturday, 2: Sunday, 3: Monday, $\dots$, 7: Friday) denote the weekday with the week starting on Saturday. The re-weighted expected number of cases $\Tilde{\mu}$ (class and day indexes omitted) are computed as
\begin{align*}
    \Tilde{\mu}_{k} &= \mu_k \cdot \nu_k \cdot \left(\frac{\sum_k \mu_k}{\sum_k \nu_k \cdot \mu_k}\right)\\
    \sum_k \nu_k &= 1,
\end{align*}
where $\nu_k$ is the weight for weekday $k$. These weights will be modeled with a Dirichlet prior
\begin{align*}
    \nu &\sim \text{Dirichlet}(c) \\
    c_k &= \sum_j \sum_t C_{jt} \mathbb{I}_{\text{Weekday}(t)=k}
\end{align*}
where $c_k$ is the total number of cases reported for weekday $k$ and $\mathbb{I}$ is a binary indicator.

\subsubsection*{Modeling school-free days}

Infections may have occurred during the week of vacation that falls into the study period. The expected number of infections and cases will be computed during vacation, but vacation days are not modeled (i.e. not incorporated into the model likelihood). In addition, we will assume lower transmission of respiratory infections on days without school (weekends and vacations). We incorporate our prior belief into the model intercept $\beta_0$  
\begin{align*}
    \beta_0 &= \alpha + \omega \cdot \text{NoSchool},
\end{align*}
where $\alpha$ is the rate of new infections on school days and $\alpha + \omega$ is the rate on days without school. We will model $\omega$ with an informative prior for a 10\% decrease in new infections on school-free days
\begin{align*}
    \omega &\sim \text{Normal}(\log 1.1, 0.05)~.
\end{align*}

\subsubsection*{Seeding infections before study start}

Cases in the first week of the study could indicate infections before the study commenced. We will therefore seed our model $2 \cdot \text{m}\,$days before the study start, where $\text{m}$ is the average incubation period. The number of infections before the study start will be modeled with an exponential prior
\begin{align*}
    I_{jt} \sim \text{Exponential}(2 \cdot \text{m}) \quad t = -2\cdot\text{m}+1, \dots, 0~. 
\end{align*}

\subsubsection*{Priors for modeling parameters}

Unless stated otherwise, we will use weakly informative priors for all modeling parameters following recommendations on the choice of priors \cite{Gelman2008-Scaling,Gelman2008-Logistic,Gelman2020,Stan2020,Gabry2023}. The continuous adjustment variables will be standardized to have zero mean and a standard deviation of 0.5. 
\begin{align*}
    \alpha &\sim \text{Student-t}_5(0, 10) \\
    \beta_2, \dots, \beta_6 &\sim \text{Student-t}_5(0, 2.5) \\
    \frac{1}{\sqrt{\phi}} &\sim \text{Half-Normal}(0,1),
\end{align*}
All priors represent our default but not definitive choices, and changes in the context of the data will be applied if necessary.

\subsection{Secondary analyses}

\subsubsection{Number of positive bioaerosol and saliva samples}

Molecular analysis will determine whether a bioaerosol or saliva sample is positive for a respiratory pathogen. If a sample is positive for multiple pathogens, we will denote this sample as positive for the pathogen with the highest viral load concentration. Let $p = 0, 1, \dots, P$ denote the pathogen, where $0$ refers to negative samples and $P$ is the number of different pathogens detected over the study. The number of positive bioaerosol and saliva samples $y_p$ in samples of sizes $n$ will be analyzed with a Multinomial logistic regression model
\begin{align*}
    y_0, y_1, \cdots, y_P | n, \beta_0, \beta_1, \beta_2, \beta_3 &\sim \text{Multinomial-Logit}(\theta_0, \theta_1, \dots \theta_P) \\
    \theta &= \text{softmax}(\mu) \\
    \mu &= \begin{cases}
                0, & \text{if } p=0, \\
                \beta_{0[p]} + \beta_{1[p]} \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Susceptibles}_p, & \text{if } p>0.
            \end{cases} \\
    \beta_0 &\sim \text{Student-t}_5(0, 2.5) \\
    \beta_1 &\sim \text{Normal}(\kappa, \tau) \\
    \kappa, \beta_2, \beta_3 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) \\
    \tau &\sim \text{Half-Student-t}_5(0,1),
\end{align*}
where $\text{softmax}(\mu) = \exp(\mu)/\sum\exp(\mu)$, Student-t$_5$ is a Student-t distribution with 5 degrees of freedom, and $s_{x}$ is the empirical standard deviation of each input variable.  Negative tests will be set as the reference category and overall variation in positive samples by pathogen will be modeled with $\beta_0$. The effect of air cleaners will be modeled with a hierarchical prior, where $\kappa$ is the average effect of air cleaners across pathogens and $\tau$ is the variation of the effect between pathogens. The effect of air cleaners will be adjusted for class-specific effects and the decreasing number of susceptibles over time (for each pathogen computed as the number of students minus the total number of positive samples).

\subsubsection{Viral load concentration in positive samples}\label{subsubsec:viral_load}

Molecular analysis will determine the viral load concentration in positive samples. The viral load concentration in positive samples will be analyzed with a hierarchical linear regression model 
\begin{align*}
    y | \beta_0, \beta_1 &\sim \text{Normal}(\mu,s_{y}) \\
    \mu &= \beta_{0[p]} + \beta_{1[p]} \cdot \text{AirCleaner} \\
    \beta_0 &\sim \text{Normal}(\kappa_0, \tau_0) \\
    \beta_1 &\sim \text{Normal}(\kappa_1, \tau_1) \\
    \kappa_0 &\sim \text{Student-t}_5(\mu_{y}, 2.5s_{y}) \\
    \kappa_1 &\sim \text{Student-t}_5\left(0, 2.5\frac{s_{y}}{s_{x}}\right) \\
    \tau_0, \tau_1 &\sim \text{Half-Student-t}_5(0,1) \\
    \sigma &\sim \text{Exponential}\left(\frac{1}{s_{y}}\right),
\end{align*}
where $\mu_{y}$ and $s_{y}$ are the empirical mean and standard deviation of the outcome.

\subsubsection{Aerosol/particle concentration}\label{subsubsec:aerosol}

We will only analyze particle concentration data of the time periods during which students were in the classroom. Daily particle concentrations will be summarized with the mean and a log transform will be applied. Each log-transformed particle concentration will be analyzed separately with a linear regression model 
\begin{align*}
    \log y | \beta_0, \beta_1, \dots, \beta_6 &\sim \text{Normal}(\mu,\sigma) \\
    \mu &= \beta_0 + \beta_1 \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Weekday} + \beta_4 \cdot \text{Students} + \beta_5 \cdot \text{Ventilation} + \beta_6 \cdot \text{Cases}\\
    \beta_0 &\sim \text{Student-t}_5(\mu_y, 2.5s_{y}) \\
    \beta_1, \dots, \beta_6 &\sim \text{Student-t}_5\left(0, 2.5\frac{s_{y}}{s_{x}}\right) \\
    \sigma &\sim \text{Exponential}\left(\frac{1}{s_{y}}\right)~.
\end{align*}
The effect of air cleaners will be adjusted for class- and weekday-specific effects, the number of students in school, the ventilation rate (air changes per hour as computed from indoor CO$_2$ levels), and the cumulative number of cases related to respiratory infections. A log transform will be applied to all continuous input variables.

\subsubsection{Number of coughs}

The daily number of coughs will be analyzed with a Negative Binomial regression model
\begin{align*}
    y | \beta_0, \beta_1, \dots, \beta_6 &\sim \text{Negative-Binomial}(\mu,\phi) \\
    \log \mu &= \log T + \beta_0 + \beta_1 \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Weekday} + \beta_4 \cdot \text{Students} + \beta_5 \cdot \text{Ventilation} + \beta_6 \cdot \text{Cases} \\
    \beta_0 &\sim \text{Student-t}_5(0, 2.5) \\
    \beta_1, \dots, \beta_6 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) \\
    \frac{1}{\sqrt{\phi}} &\sim \text{Half-Normal}(0,1),
\end{align*}
where $T$ is the daily duration that students were in the classroom and $\phi$ is the parameter modeling over-dispersion in count data.

\section{Supplementary analyses}

\subsection{Modeling transmission risks}

We will model transmission risks using the Wells-Riley equation\cite{Wells1963} as modified by Rudnick and Milton\cite{Rudnick2003} 
\begin{align*}
    P = \frac{D}{S} = 1 - \exp\left(-\frac{fIqt}{n}\right),
\end{align*}
where $P$ is the risk of infection, $D$ is the number of disease cases, $S$ is the number of susceptibles, $f$ is the fraction of rebreathed air, $I$, is the number of infectious individuals in space, $n$ is the total number of individuals in space, $q$ is the generation rate of infectious quanta (doses of infectious particles), and $t$ is the duration of exposure. Parameters $D$, $S$, and $I$ can be computed based on epidemiological (or molecular) data. The fraction of rebreathed air can be computed from CO$_2$ levels. The number of individuals in space can be determined based on the number of absences. Exposure times can be determined based on the timetable. Thus, we will first use the Wells-Riley equation to estimate the unknown parameter $q$. The risk of infection will then be modeled using the estimated $q$ for exposure times representing a typical school day or week. These risks will be estimated separately for different pathogens detected over the study.

\subsection{Targeted and deep sequencing}

For positive samples (bioaerosol and saliva samples, swabs from HEPA filters), we will perform targeted sequencing and deep sequencing of viral RNA to compare genetic relatedness between viral strains detected in the air and human samples\cite{Goncalves2021}.

\section{Missing data}

We obtained informed consent from all but one student. As the reason for any absence(s) from this student and the symptoms are unclear, we will consider any absence(s) from this student as unrelated to a respiratory infection. Particle concentrations will be summarized with the daily mean of the minutely values in the time periods when students were in the classroom. Missing values during these time periods will be linearly imputed. Missing values in adjustment variables will be imputed with a fully Bayesian approach where missing values are treated as parameters and a prior is assigned.

\section{Statistical software employed}

The statistical software R (version 4.2.1 or later) with RStudio will be used for all statistical analyses. Data preprocessing and comparative analyses will be performed in the \texttt{tidyverse} package (version 1.3.2). Bayesian analyses will be performed with the probabilistic programming language Stan (version 2.21.0), using the R interface provided by the \texttt{rstan} package (version 2.21.7), the \texttt{rstanarm} package (version 2.21.3), or the \texttt{brms} package (version 2.18.0).

\end{document}

### 2023-05-23 doc/SAP/sap.tex
update SAP with references version 0.5

@@
         \Large
-        Version number: 0.4 \\
+        Version number: 0.5 \\
         Version date: \today
@@
 SARS-CoV-2 or Influenza virus) were spreading over the study.
-Molecular data (positive samples), environmental (particle concentrations), and epidemiological data (absences related to respiratory infections) will then be compared between study conditions with and without air cleaners.
+Molecular (positive samples), environmental (particle concentrations), and epidemiological data (absences related to respiratory infections) will then be compared between study conditions with and without air cleaners.
 The aim of the comparison is to assess whether air cleaners reduced the risk of airborne infection.
@@
 Both classrooms were not equipped with an active HVAC (Heating, Ventilation, Air conditioning) system, but were ventilated using passive window ventilation.}
-    \includegraphics{study_setting.pdf}
+    \includegraphics{../study_setting.pdf}
     \label{fig:study_setting}
@@
     CO$_2$, aerosol/particle concentrations, humidity, temperature & Air quality device: AQ Guard (Palas GmbH, Karlsruhe, Germany) & x & x & daily by minute \\
-    Coughs & Sounds collected via microphones and coughs detected with a Machine Learning model\cite{Bertschinger2023} & x & x & daily by seconds \\
+    Coughs & Sounds collected via microphones and coughs detected with a Machine Learning model \cite{Bertschinger2023CBMS} & x & x & daily by seconds \\
     \midrule
@@
 
-In a previous study, we estimated and compared the transmission of airborne infections with and without air cleaners and mask mandates in five classes of two secondary schools in the Canton of Solothurn, Switzerland, between January and March 2022\cite{Banholzer2023}.
+In a previous study, we estimated and compared the transmission of airborne infections with and without air cleaners and mask mandates in five classes of two secondary schools in the Canton of Solothurn, Switzerland, between January and March 2022 \cite{Banholzer2023PLoSMed}.
 Air cleaners (commercially available portable HEPA-filtration devices) were found to reduce particle concentrations, but no association was found with transmission.
@@
     \caption{\textbf{Study design.} Seven-week study period from January 16 to March 11, 2023, excluding a week of vacation between February 6 and 11.
-Cross-over design where two portable air cleaners were installed during different time periods in each class (Class A: weeks 1, 2, 6, and 7; Class B: weeks 3, 4, and 5) but not at the same time in both classes.}
+Cross-over design where two portable air cleaners were installed during different time periods in each class (Class A: weeks 1, 2, 6, and 7; Class B: weeks 3, 4, and 5).}
     \begin{tabular}{l l l l l l l l l}
@@
     \item \textbf{Environmental data}: Comparison of aerosol/particle concentrations.
-    \item \textbf{Epidemiological data}: Comparison of the number of infections.  
+    \item \textbf{Epidemiological data}: Comparison of the number of absences related to respiratory infections.  
 \end{enumerate}
@@
     \begin{enumerate}
-        \item Weekly number cases and deaths, laboratory tests and share of positive tests for COVID-19 in the Canton of Solothurn \cite{FOPH2023-cases}
-        \item Weekly number of consultations per 100,000 people due to influenza-like illnesses (ILI) \cite{FOPH2023-consultations}
+        \item Weekly number of cases and deaths, laboratory tests and share of positive tests for COVID-19 in the Canton of Solothurn \cite{FOPH2023Cases}
+        \item Weekly number of consultations per 100,000 people due to influenza-like illnesses (ILI) \cite{FOPH2023Consultations}
     \end{enumerate}
@@
 
-The number of respiratory cases (absences related to respiratory infections by date of symptom onset) will be modeled with a Bayesian latent hierarchical regression model, which is similar to the model used in our previous study \Cref{Banholzer2023}.
+The number of respiratory cases (absences related to respiratory infections by date of symptom onset) will be modeled with a Bayesian latent hierarchical regression model, which is similar to the model used in our previous study \cite{Banholzer2023PLoSMed}.
 The model links the unobserved number of new infections to the observed number of new cases.
@@
 where $N_{jt} = \sum_{s<t} I_{js}$ is the cumulative number of infections (model offset), $\beta_0$ is the rate of new infections without air cleaners (model intercept), and $\beta_1$ is the effect of air cleaners.
-The effect of air cleaners will be adjusted for class-specific effects, the number of students in school, the ventilation rate (air changes per hour computed from indoor CO$_2$ levels using the transient mass balance method\cite{Battermann2017}), the proportion of positive tests for SARS-CoV-2 in the community (from secondary data), and the number of consultations for influenza-like illnesses in the community (from secondary data).
+The effect of air cleaners will be adjusted for class-specific effects, the number of students in school, the ventilation rate (air changes per hour computed from indoor CO$_2$ levels using the transient mass balance method \cite{Batterman2017IJERPH}), the proportion of positive tests for SARS-CoV-2 in the community (from secondary data), and the number of consultations for influenza-like illnesses in the community (from secondary data).
 
@@
 The combination will be determined based on the proportion of positive bioaerosol/saliva samples for each pathogen found in the molecular analysis.
-The prior distributions for pathogen-specific incubation periods will be based on published estimates.
+The prior distributions for pathogen-specific incubation periods will be based on estimates published in the literature.
 
@@
 
-Our previous study recorded a higher proportion of cases on Mondays than on weekends, suggesting recall bias and under-reporting of cases on weekends\cite{Banholzer2023}.
+Despite recording cases by date of symptom onset, our previous study recorded a higher proportion of cases on Mondays than on weekends, suggesting recall bias and under-reporting of cases on weekends \cite{Banholzer2023PLoSMed}.
 To consider weekday effects in the reporting of cases, we will re-weight the expected number of cases each week as follows.
@@
 
-Unless stated otherwise, we will use weakly informative priors for all modeling parameters following recommendations on the choice of priors \cite{Gelman2008-Scaling,Gelman2008-Logistic,Gelman2020,Stan2020,Gabry2023}.
+Unless stated otherwise, we will use weakly informative priors for all modeling parameters following recommendations on the choice of priors \cite{Gelman2008StatMed,Gelman2008StatAnnals,Gelman2020RegOther,Stan2020Priors,Gabry2023Priors}.
 The continuous adjustment variables will be standardized to have zero mean and a standard deviation of 0.5.
@@
 \end{align*}
+where Student-t$_5$ is a Student-t distribution with 5 degrees of freedom.
 All priors represent our default but not definitive choices, and changes in the context of the data will be applied if necessary.
@@
             \end{cases} \\
-    \beta_0 &\sim \text{Student-t}_5(0, 2.5) \\
-    \beta_1 &\sim \text{Normal}(\kappa, \tau) \\
-    \kappa, \beta_2, \beta_3 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) \\
-    \tau &\sim \text{Half-Student-t}_5(0,1),
+    \beta_0 &\sim \text{Normal}(\kappa_0, \tau_0) \\
+    \beta_1 &\sim \text{Normal}(\kappa_1, \tau_1) \\
+    \kappa_0 &\sim \text{Student-t}_5(0, 2.5) \\
+    \kappa_1, \beta_2, \beta_3 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) \\
+    \tau_0, \tau_1 &\sim \text{Half-Student-t}_5(0,1),
 \end{align*}
-where $\text{softmax}(\mu) = \exp(\mu)/\sum\exp(\mu)$, Student-t$_5$ is a Student-t distribution with 5 degrees of freedom, and $s_{x}$ is the empirical standard deviation of each input variable.
-Negative tests will be set as the reference category and overall variation in positive samples by pathogen will be modeled with $\beta_0$.
-The effect of air cleaners will be modeled with a hierarchical prior, where $\kappa$ is the average effect of air cleaners across pathogens and $\tau$ is the variation of the effect between pathogens.
+where $\text{softmax}(\mu) = \exp(\mu)/\sum\exp(\mu)$ and $s_{x}$ is the empirical standard deviation of each input variable.
+Negative tests will be set as the reference category.
+Overall variation in positive samples by pathogen will be modeled with $\beta_0$ and the effect of air cleaners will be modeled with $\beta_1$.
+Both $\beta_0$ and $\beta_1$ will be modeled with a hierarchical prior, where $\kappa$ is the average estimate across pathogens and $\tau$ is the variation of the estimate between pathogens.
 The effect of air cleaners will be adjusted for class-specific effects and the decreasing number of susceptibles over time (for each pathogen computed as the number of students minus the total number of positive samples).
@@
 We will only analyze particle concentration data of the time periods during which students were in the classroom.
-Daily particle concentrations will be summarized with the mean and a log transform will be applied.
-Each log-transformed particle concentration will be analyzed separately with a linear regression model
+Particle concentrations will be summarized daily with the mean across these time periods and a log transform will be applied.
+Each log-transformed average particle concentration will be analyzed separately with a linear regression model
 \begin{align*}
@@
 \end{align*}
-The effect of air cleaners will be adjusted for class- and weekday-specific effects, the number of students in school, the ventilation rate (air changes per hour as computed from indoor CO$_2$ levels), and the cumulative number of cases related to respiratory infections.
+The effect of air cleaners will be adjusted for class- and weekday-specific effects, the number of students in school, the ventilation rate (air changes per hour), and the cumulative number of cases related to respiratory infections.
 A log transform will be applied to all continuous input variables.
@@
 
-We will model transmission risks using the Wells-Riley equation\cite{Wells1963} as modified by Rudnick and Milton\cite{Rudnick2003} 
+We will model transmission risks using the Wells-Riley equation \cite{Riley1978AJE} as modified by Rudnick and Milton \cite{Rudnick2003IndoorAir} 
 \begin{align*}
@@
 where $P$ is the risk of infection, $D$ is the number of disease cases, $S$ is the number of susceptibles, $f$ is the fraction of rebreathed air, $I$, is the number of infectious individuals in space, $n$ is the total number of individuals in space, $q$ is the generation rate of infectious quanta (doses of infectious particles), and $t$ is the duration of exposure.
-Parameters $D$, $S$, and $I$ can be computed based on epidemiological (or molecular) data.
-The fraction of rebreathed air can be computed from CO$_2$ levels.
-The number of individuals in space can be determined based on the number of absences.
-Exposure times can be determined based on the timetable.
-Thus, we will first use the Wells-Riley equation to estimate the unknown parameter $q$.
+Parameters $D$, $S$, and $I$ will be computed based on epidemiological (or molecular) data.
+The fraction of rebreathed air will be computed from CO$_2$ levels.
+The number of individuals in space will be determined based on the number of absences.
+Exposure times will be determined based on the students' timetables.
+
+We will first use the Wells-Riley equation to estimate the unknown parameter $q$.
 The risk of infection will then be modeled using the estimated $q$ for exposure times representing a typical school day or week.
@@
 
-For positive samples (bioaerosol and saliva samples, swabs from HEPA filters), we will perform targeted sequencing and deep sequencing of viral RNA to compare genetic relatedness between viral strains detected in the air and human samples\cite{Goncalves2021}.
+For positive samples (bioaerosol and saliva samples, swabs from HEPA filters), we will perform targeted sequencing and deep sequencing of viral RNA to compare genetic relatedness between viral strains detected in the air and human samples \cite{Goncalves2021Microorganisms}.
 
@@
 We obtained informed consent from all but one student.
-As the reason for any absence(s) from this student and the symptoms are unclear, we will consider any absence(s) from this student as unrelated to a respiratory infection.
+As the reason for any absence(s) from this student and the symptoms will be unclear, we will consider any absence(s) from this student as unrelated to a respiratory infection.
 Particle concentrations will be summarized with the daily mean of the minutely values in the time periods when students were in the classroom.
@@
 Data preprocessing and comparative analyses will be performed in the \texttt{tidyverse} package (version 1.3.2).
-Bayesian analyses will be performed with the probabilistic programming language Stan (version 2.21.0), using the R interface provided by the \texttt{rstan} package (version 2.21.7), the \texttt{rstanarm} package (version 2.21.3), or the \texttt{brms} package (version 2.18.0).
+Bayesian analyses will be performed with the probabilistic programming language Stan (version 2.21.0), using the R interfaces provided by the \texttt{rstan} package (version 2.21.7), the \texttt{rstanarm} package (version 2.21.3), or the \texttt{brms} package (version 2.18.0).
+
+\bibliographystyle{vancouver}
+\bibliography{references}
 

### 2023-05-23 preprocessing/prep_palas.r
manual fixes in epidemiological reporting

@@ -7,8 +7,8 @@ library(lubridate)
 
 #### Data ####
 
-files_A <- list.files("data-raw/palas/Palas Wochen 1-3/E3a_3 Palas_AQGuard_SN15988/aqGuard_15988", full.names = T)
-files_B <- list.files("data-raw/palas/Palas Wochen 1-3/E3b_3 Palas_AQGuard_SN15973/aqGuard_15973", full.names = T)
+files_A <- list.files("data-raw/palas/E3a_komplett_aqGuard_15988", full.names = T)
+files_B <- list.files("data-raw/palas/E3b_komplett_aqGuard_15973", full.names = T)
 files_A <- files_A[grepl("txt", files_A)]
 files_B <- files_B[grepl("txt", files_B)]
 
@@ -37,6 +37,17 @@ df <- rbind(data.frame(class = "A", file = files_A),
 
 #### Preprocessing ####
 
+# remove weekends and vacation
+df <- df %>%
+  filter(!(weekdays(date) %in% c("Saturday", "Sunday")),
+         !(date %in% seq.Date(as.Date("2023-02-06"), as.Date("2023-02-11"), by = "1 day")),
+         date >= as.Date("2023-01-16"),
+         date <= as.Date("2023-03-11")) 
+
+# move time in class B (E3b) on 2023-01-16 (first weekend, Monday) by one hour backwards (see data-collection-issues.docx)
+df$datetime[df$class=="B"&df$date==as.Date("2023-01-16")] <- df$datetime[df$class=="B"&df$date==as.Date("2023-01-16")] %m-% hours(1)
+df %>% filter(date == as.Date("2023-01-16")) %>% ggplot(aes(x = datetime, y = CO2, color = class)) + geom_line()
+
 to_hhmm <- function(t) {
   h <- hour(t)
   if (nchar(h) == 1) {
@@ -50,40 +61,207 @@ to_hhmm <- function(t) {
 }
 
 # Filter time when people are in class
-
 rtime <- readRDS("data-clean/time-in-room.rds") 
 rtime$hhmm <- sapply(rtime$datetime, to_hhmm)
-
-df_filt <- df 
-df_filt$hhmm <- sapply(df_filt$datetime, to_hhmm) 
-df_filt <- df_filt %>%
+ 
+df$hhmm <- sapply(df$datetime, to_hhmm) 
+df <- df %>%
   left_join(rtime %>% select(class, date, hhmm, no_palas)) %>%
[…]

### 2023-05-23 preprocessing/prep_redcap.r

@@ -2,13 +2,14 @@
 
 library(tidyverse)
 library(reshape2)
+library(lubridate)
 
 
 #### Data ####
 
 df <- read.csv("data-raw/redcap/MCIDSchulenFollowup_DATA_2023-02-09_1356.csv") %>%
   mutate(class = ifelse(grepl("Kontroll", record_id), "A", "B"),  # A is Kontroll- and B is Studienklasse
-         date = as.Date(date, format = "%d.%m.%y")) 
+         date = as.Date(date, format = "%Y-%m-%d")) 
 
 
 #### In-Room time ####
@@ -36,3 +37,141 @@ rtime <- df %>%
 
 
 saveRDS(rtime, "data-clean/time-in-room.rds")
+write.csv(rtime, "data-clean/time-in-room.csv", row.names = F)
+
+
+#### Absences ####
+
+cases <- df %>%
+  dplyr::select(class, initialen_sus, matches("sus_")) %>%
+  filter(sus_date != "") %>%
+  rename(date_symptom = sus_symp_date,
+         date_absent = sus_date,
+         date_back = sus_date_back,
+         sex = sus_sex,
+         reason = sus_absenz,
+         student_id = initialen_sus) %>%
+  mutate(across(c(date_symptom, date_absent, date_back), as.Date),
+         sex = factor(ifelse(sex == 1, "male", "female"), levels = c("male", "female")),
+         reason = factor(ifelse(is.na(reason), "unknown", ifelse(reason == 1, "sickness", "other")), levels = c("sickness", "other", "unknown")),
+         respiratory_infection = `sus_symp___1` + `sus_symp___2` + `sus_symp___3` + `sus_symp___4` +
+                                       `sus_symp___5` + `sus_symp___6` + `sus_symp___7` + `sus_symp___8` +
+                                       `sus_symp___9` + `sus_symp___10` + `sus_symp___11` + `sus_symp___12` +
+                                       `sus_symp___13`,
+         respiratory_infection = ifelse(respiratory_infection >= 1, "yes", "no"),
+         respiratory_infection = ifelse(sus_testing == 1 & sus_tesing_res == 1, "no", respiratory_infection), # re-label one case that tested negative
+         respiratory_infection = factor(respiratory_infection, levels = c("no", "yes"))) %>% 
+  dplyr::select(class, student_id, date_symptom, date_absent, date_back, sex, reason, respiratory_infection) 
+
+# check delay between symptom onset and absence
+cases <- cases %>%
[…]

### 2023-05-24 tests/multilevel-linear.r
test analyses with simulated data

@@ -1,0 +1,16 @@
+library(tidyverse)
+library(rstanarm)
+library(LaplacesDemon)
+
+aircleaner <- rep(c(0,1), each = 6)
+pathogen <- rep(c("SARS-CoV-2", "Influenza", "RSV"), each = 5)
+dat <- expand_grid(aircleaner = aircleaner, pathogen = factor(pathogen)) 
+binary_pathogen <- model.matrix(~ pathogen - 1, data = dat)
+dat <- cbind(dat, binary_pathogen) %>%
+  mutate(mu_y = 2 - 0.5 * aircleaner - .25 * pathogenInfluenza + 0.5 * `pathogenSARS-CoV-2`,
+         y = sapply(mu_y, function(m) rtrunc(n = 1, spec = "norm", a = 0, mean = m, sd = .5))) 
+
+
+mlm <- stan_glmer(y ~ 1 + aircleaner + (1 + aircleaner | pathogen), data = dat)
+
+summary(mlm)

### 2023-05-24 tests/multinomial.r

@@ -1,0 +1,71 @@
+# libraries
+library(tidyverse)
+library(rstan)
+library(LaplacesDemon)
+library(bayesplot)
+
+# functions
+softmax <- function(x) {
+  exp(x) / sum(exp(x))
+}
+
+# seed
+seed <- 12345
+set.seed(seed)
+
+# population and time
+no_days <- 6 * 2 * 2
+days <- 1:no_days
+no_students <- 38
+students <- 1:no_students
+
+# input variables
+class <- rep(c(0, 1), each = no_days/2)
+aircleaner <- rep(c(0,1), times = 3, each = no_days/6)
+
+# parameters
+prob_test <- runif(no_days, 0.5, 1)
+prob_virus <- c(0.5, 0.3, 0.2) # (SARS-CoV-2, Influenza, RSV)
+no_virus <- length(prob_virus)
+prob_pos <- 3 / no_students
+beta0 <- logit(prob_pos * prob_virus)
+beta1 <- c(1, 2, 1)
+beta2 <- 0.0
+beta3 <- 0.5
+
+# initialize
+susceptibles <- matrix(38, nrow = no_days + 1, ncol = no_virus)
+y <- matrix(0, nrow = no_days, ncol = no_virus + 1)
+
+# simulate data
+mu <- numeric(no_virus + 1)
+for (d in days) {
+  tested_students <- sum(rbernoulli(no_students, prob_test[d]))
+  mu[-1] <- beta0 - beta1 * aircleaner[d] + beta2 * class[d] + beta3 * (susceptibles[d] / 38)
+  y[d,] <- rmultinom(1, tested_students, softmax(mu))
+  susceptibles[d+1,] <- susceptibles[d,] - y[d,-1]
+}
+
+colSums(y)
+
+# create stan data
+sdl <- list(
+  K = ncol(y),
+  N = no_days,
+  y = y,
+  x1 = aircleaner,
+  x2 = class,
+  x3 = (susceptibles[-nrow(susceptibles),] / 38) / (2 * sd(susceptibles[-nrow(susceptibles),] / 38))
+)
+
+
+# model
+mm <- stan(file = "tests/multi-logit.stan",
+           data = sdl,
+           seed = seed)
+
+mm_sum <- summary(mm, pars = c("beta0", "beta1", "beta2", "beta3", "tau", "kappa"), probs = c(0.05, 0.95))
[…]

### 2023-05-24 tests/transmission.r

@@ -1,0 +1,127 @@
+# libraries
+library(tidyverse)
+library(rstanarm)
+library(LaplacesDemon)
+
+# functions
+source("utils/epi.r")
+
+re_weight <- function(x, p) {
+  x * p * (sum(x) / (sum(x * p)))
+}
+
+# seed
+seed <- 12345
+set.seed(seed)
+
+# population and time
+weeks <- 7
+S <- 40
+D <- weeks * 7
+weekday <- c(c("Sat", "Sun"),
+             rep(c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"), times = 6),
+             c("Mon", "Tue", "Wed", "Thu", "Fri"))
+days <- 1:D
+week <- c(c(0,0),
+          rep(1:6, each = 7),
+          rep(7, 5))
+
+# input variables
+aircleaner <- ifelse(!(weekday %in% c("Sat", "Sun")) & week %in% c(3, 5, 6), 1, 0)
+weekend <- ifelse(weekday %in% c("Sat", "Sun"), 1, 0)
+vacation <- ifelse(week == 4, 1, 0)
+noschool <- pmax(vacation, weekend)
+
+# parameters
+t0 <- 7
+I0 <- rexp(t0, rate = t0)
+alpha <- -2
+omega <- -0.15
+beta1 <- -1
+pIN <- vp(D + t0, 0, p_in)
+w <- c(0.025, 0.025, 0.35, 0.15, 0.15, 0.15, 0.15)
+
+
+# simulate data 
+muC <- numeric(D)
+I <- numeric(D)
+for (d in 1:D) {
+  if (d == 1) {
+    N <- sum(I0)
+  } else {
+    N <- sum(c(I0, I[1:(d-1)]))
+  }
+  I[d] <- N * exp(alpha + omega * noschool[d] + beta1 * aircleaner[d])
+  muC[d] <- sum(rev(pIN[1:(d+t0)]) * c(I0, I[1:d]))
+}
+
+# cases without shift
+C <- sapply(muC, rpois, n = 1)
+
+# cases with shift
+muC_weekly <- split(muC, ceiling(seq_along(muC)/7))
+muC_tilde <- unlist(lapply(muC_weekly, re_weight, p = w))
+C_tilde <- sapply(muC_tilde, rpois, n = 1)
+
+# inspect
+df <- data.frame(
+  C = C,
+  C_tilde = C_tilde,
+  aircleaner = aircleaner,
+  day = 1:D,
[…]

### 2023-05-24 utils/bayes.R
add utils

@@ -1,0 +1,89 @@
+require(posterior)
+require(cmdstanr)
+require(tidybayes)
+require(bayesplot)
+require(LaplacesDemon)
+require(loo)
+library(rstan)
+
+
+# Gather draws for posterior draws from read_cmdstan_csv
+gather_draws_csv <- function(model, par, mi, ma) {
+  model$post_warmup_draws[,,paste0(par, "[", mi:ma, "]")]%>% 
+    as_draws_df() %>%
+    data.frame() %>%
+    reshape2::melt(id.vars = c(".chain", ".iteration", ".draw")) %>%
+    mutate(variable = stringi::stri_extract(variable, regex = "\\d+")) 
+}
+
+gather_draws_csv2 <- function(model, par, mi1, ma1, mi2, ma2) {
+  grid <- expand.grid(mi1:ma1, mi2:ma2)
+  model$post_warmup_draws[,,paste0(par, "[", grid$Var1, ",", grid$Var2, "]")]%>% 
+    as_draws_df() %>%
+    data.frame() %>%
+    reshape2::melt(id.vars = c(".chain", ".iteration", ".draw")) %>%
+    mutate(row = stringi::stri_extract(variable, regex = "\\d+"),
+           col = sapply(stringi::stri_extract_all(variable, regex = "\\d+"), function(x) x[2])) 
+}
+
+# To support tidybayes gather_draws for cmdstanr
+tidy_draws.CmdStanMCMC <- function(model, ...) { return(as_draws_df(model$draws(...))) }
+
+tidy_draws.CmdStanMCMC_mat <- function(model, ...) {
+  tidy_draws.CmdStanMCMC(model, ...) %>%
+    select(-`.chain`,-`.iteration`) %>%
+    rename(draw = `.draw`) %>%
+    melt("draw") %>%
+    mutate(variable = stringi::stri_extract(variable, regex = "\\d+,\\d+"),
+           variable_split = str_split(variable, ","),
+           row = map_int(variable_split, function(x) as.integer(x[1])),
+           col = map_int(variable_split, function(x) as.integer(x[2]))) %>%
+    select(row, col, draw, value) 
+}
+
+# loo
+loo.CmdStanMCMC.array <- function(model, ...) {
+  loo_pars <- expand.grid(...) %>%
+    unite("par", sep = ",") 
+  loo <- model$post_warmup_draws[,,paste0("log_lik[",loo_pars$par,"]")] %>%
+    loo()
+  return(loo)
+}
+
[…]

### 2023-05-24 utils/epi.r

@@ -1,0 +1,34 @@
+#' Incubation period
+#' - assume based on meta-analysis from McAloon et al
+#' - https://bmjopen.bmj.com/content/bmjopen/10/8/e039652.full.pdf
+#' - lognormal distribution
+#' - log mu: 1.63 (1.51, 1.75)
+#' - log sigma: 0.50 (0.45, 0.55)
+#' 
+
+
+p_in_mu_m <- 1.63
+p_in_mu_s <- (1.75 - 1.51) / (2 * qnorm(0.975))
+p_in_sigma_m <- 0.50
+p_in_sigma_s <- (0.55 - 0.45) / qnorm(0.975)
+
+p_in <- function(x, log_mu = 1.63, log_sigma = 0.50) { 
+  plnorm(x, log_mu, log_sigma)
+}
+
+# Vectorize delay function (e.g. p_g)
+vp <- function(xT, from0, FUN, ...) {
+  x <- seq(0,xT)
+  y <- length(x)
+  if (from0) {
+    y[1] <- FUN(0.5, ...)
+    y[2] <- FUN(1.5, ...) - FUN(0.5, ...)
+  } else {
+    y[1] <- 0
+    y[2] <- FUN(1.5, ...)
+  }
+  for (i in 3:length(x)) {
+    y[i] = FUN(x[i] + 0.5, ...) - FUN(x[i] - 0.5, ...)
+  }
+  return(y)
+}

### 2023-05-24 utils/plotting.R

@@ -1,0 +1,93 @@
+require(ggpattern)
+require(gridExtra)
+require(GGally)
+require(Cairo)
+
+# ggplot theme for Nature style
+text_size = 8
+update_geom_defaults("text", list(size = text_size))
+theme_nature <- function () { 
+  theme_classic(base_size = text_size, base_family = 'sans') %+replace% 
+    theme(
+      axis.text = element_text(size = text_size),
+      axis.title = element_text(size = text_size),
+      plot.title = element_text(size = text_size + 2, face = "bold", hjust = 0.5, margin = ggplot2::margin(0, 0, 5, 0))
+    )
+}
+
+theme_bw2 <- function () { 
+  theme_bw(base_size = text_size, base_family = 'sans') %+replace% 
+    theme(
+      axis.text = element_text(size = text_size),
+      axis.title = element_text(size = text_size),
+      plot.title = element_text(size = text_size + 2, face = "bold", hjust = 0, margin = ggplot2::margin(0, 0, 5, 0))
+    )
+}
+
+# Correlation plot 
+# Helpful code:
+# https://stackoverflow.com/questions/44961437/how-to-include-density-coloring-in-pairwise-correlation-scatter-plot
+# https://stackoverflow.com/questions/37889222/change-colors-in-ggpairs-now-that-params-is-deprecated?answertab=votes#tab-top
+contours <- function(data, mapping, ...){
+  p <- ggplot(data = data, mapping = mapping) + 
+    stat_density2d() +
+    scale_x_continuous(breaks = seq(-.6,0,.2)) +
+    scale_y_continuous(breaks = seq(-.6,0,.2))
+  p
+}
+
+# Plot for correlation
+contours <- function(data, mapping, ...){
+  p <- ggplot(data = data, mapping = mapping) + 
+    stat_density2d() #+
+  #scale_x_continuous(breaks = seq(-.8,.4,.4)) +
+  #scale_y_continuous(breaks = seq(-.8,.4,.4))
+  p
+}
+hex <- function(data, mapping, ...){
+  p <- ggplot(data = data, mapping = mapping) + 
+    geom_hex() #+
+  #scale_x_continuous(breaks = seq(-.8,.4,.4)) +
+  #scale_y_continuous(breaks = seq(-.8,.4,.4))
+  p
+}
[…]

### 2023-05-24 utils/tex.R

@@ -1,0 +1,56 @@
+# require(tikzDevice)
+# 
+# # Settings for tikzdevice
+# if (Sys.info()['sysname'] == 'Darwin') {
+#   options("tikzLatex" = Sys.which("pdflatex"))
+# } else {
+#   options("tikzLatex" = 'C:/texlive/2018/bin/win32/pdflatex')
+# }
+# 
+
+
+
+# Round function
+round_k <- function(x, k = 2) {
+  trimws(format(round(x, k), nsmall = k))
+}
+
+# Check if value is double
+is.numeric.double <- function(x) {
+  if (is.numeric(x)) {
+    if (all(x %% 1 == 0)) {
+      return(F)
+    }
+    else {
+      return(T)
+    }
+  } 
+  else {
+    return(F)
+  }
+}
+
+
+# Remove leading zero
+rzero <- function(x) {
+  x <- as.character(x)
+  fc <- substr(x, 1, 1)
+  if (fc == "0")
+    return(substr(x, 2, nchar(x)))
+  else
+    return(paste0("$-$", substr(x, 3, nchar(x))))
+}
+
+# Math tex sign
+tex_sign <- function(x) {
+  x_new <- character(length(x))
+  for (i in 1:length(x)) {
+    is_negative <- ifelse(substr(x[i], 1, 1) == "-", T, F)
+    if (is_negative) {
+      x_new[i] <- paste0("$-$", substr(x[i], 2, nchar(x[i])))
+    } else {
+      x_new[i] <- paste0("\\hphantom{$-$}", x[i])
+    }
+  }
+  return(x_new)
+}

### 2023-05-24 doc/SAP/sap.tex

@@
     \item The number of positive bioaerosol and saliva samples.
-    \item Viral load concentration in positive bioaerosol and saliva samples.
+    \item Viral load concentration in bioaerosol samples.
     \item The aerosol/particle number, and matter mass concentrations.
@@
 
-\subsubsection{Viral load concentration in positive samples}\label{subsubsec:viral_load}
+\subsubsection{Viral load concentration in bioaerosol samples}\label{subsubsec:viral_load}
 
-Molecular analysis will determine the viral load concentration in positive samples. The viral load concentration in positive samples will be analyzed with a hierarchical linear regression model 
+Molecular analysis will determine the viral load concentration in bioaerosol samples, which will be analyzed with a hierarchical linear regression model 
 \begin{align*}

### 2023-05-24 preprocessing/prep_palas.r
relabling classes

@@ -7,10 +7,10 @@ library(lubridate)
 
 #### Data ####
 
-files_A <- list.files("data-raw/palas/E3a_komplett_aqGuard_15988", full.names = T)
-files_B <- list.files("data-raw/palas/E3b_komplett_aqGuard_15973", full.names = T)
-files_A <- files_A[grepl("txt", files_A)]
+files_B <- list.files("data-raw/palas/E3a_komplett_aqGuard_15988", full.names = T)
+files_A <- list.files("data-raw/palas/E3b_komplett_aqGuard_15973", full.names = T)
 files_B <- files_B[grepl("txt", files_B)]
+files_A <- files_A[grepl("txt", files_A)]
 
 read_palas <- function(file) {
   df <- read.table(file, sep ="\t")
@@ -29,8 +29,8 @@ read_palas <- function(file) {
 }
 
 
-df <- rbind(data.frame(class = "A", file = files_A), 
-            data.frame(class = "B", file = files_B)) %>%
+df <- rbind(data.frame(class = "B", file = files_B), 
+            data.frame(class = "A", file = files_A)) %>%
   mutate(data = lapply(file, read_palas)) %>%
   unnest(data)
 
@@ -44,8 +44,8 @@ df <- df %>%
          date >= as.Date("2023-01-16"),
          date <= as.Date("2023-03-11")) 
 
-# move time in class B (E3b) on 2023-01-16 (first weekend, Monday) by one hour backwards (see data-collection-issues.docx)
-df$datetime[df$class=="B"&df$date==as.Date("2023-01-16")] <- df$datetime[df$class=="B"&df$date==as.Date("2023-01-16")] %m-% hours(1)
+# move time in class A (E3b) on 2023-01-16 (first weekend, Monday) by one hour backwards (see data-collection-issues.docx)
+df$datetime[df$class=="A"&df$date==as.Date("2023-01-16")] <- df$datetime[df$class=="A"&df$date==as.Date("2023-01-16")] %m-% hours(1)
 df %>% filter(date == as.Date("2023-01-16")) %>% ggplot(aes(x = datetime, y = CO2, color = class)) + geom_line()
 
 to_hhmm <- function(t) {
@@ -195,10 +195,10 @@ env_dat <- left_join(
 ) %>%
   mutate(AER = ifelse(is.na(CO2), NA, AER),
          aircleaner = ifelse(class == "A", 
-                             ifelse(date <= as.Date("2023-01-28"), "yes",
[…]

### 2023-05-24 preprocessing/prep_redcap.r

@@ -8,7 +8,7 @@ library(lubridate)
 #### Data ####
 
 df <- read.csv("data-raw/redcap/MCIDSchulenFollowup_DATA_2023-02-09_1356.csv") %>%
-  mutate(class = ifelse(grepl("Kontroll", record_id), "A", "B"),  # A is Kontroll- and B is Studienklasse
+  mutate(class = ifelse(grepl("Kontroll", record_id), "B", "A"),  # A is Studien- and B is Kontrollklasse
          date = as.Date(date, format = "%Y-%m-%d")) 
 
 
@@ -83,19 +83,19 @@ View(cases)
 #' the following students were absent only in the morning (and return in the afternoon)
 #' nevertheless, we consider them as being absent for the whole day, 
 #' thus shifting their date back by one day
-cases$date_back[cases$class == "A" & cases$student_id == 7 & cases$date_back == as.Date("2023-01-19")] <- as.Date("2023-01-19") + days(1)
-cases$date_back[cases$class == "B" & cases$student_id == 6 & cases$date_back == as.Date("2023-01-26")] <- as.Date("2023-01-26") + days(1)
-cases$date_back[cases$class == "B" & cases$student_id == 15 & cases$date_back == as.Date("2023-02-15")] <- as.Date("2023-02-15") + days(1)
-cases$date_back[cases$class == "B" & cases$student_id == 19 & cases$date_back == as.Date("2023-02-23")] <- as.Date("2023-02-23") + days(1)
-cases$date_back[cases$class == "B" & cases$student_id == 12 & cases$date_back == as.Date("2023-03-02")] <- as.Date("2023-03-02") + days(1)
+cases$date_back[cases$class == "B" & cases$student_id == 7 & cases$date_back == as.Date("2023-01-19")] <- as.Date("2023-01-19") + days(1)
+cases$date_back[cases$class == "A" & cases$student_id == 6 & cases$date_back == as.Date("2023-01-26")] <- as.Date("2023-01-26") + days(1)
+cases$date_back[cases$class == "A" & cases$student_id == 15 & cases$date_back == as.Date("2023-02-15")] <- as.Date("2023-02-15") + days(1)
+cases$date_back[cases$class == "A" & cases$student_id == 19 & cases$date_back == as.Date("2023-02-23")] <- as.Date("2023-02-23") + days(1)
+cases$date_back[cases$class == "A" & cases$student_id == 12 & cases$date_back == as.Date("2023-03-02")] <- as.Date("2023-03-02") + days(1)
 
[…]

### 2023-05-24 analysis/env_analysis.Rmd
environmental data analysis

---
title: "Environmental Data Analysis"
author: "Nicolas Banholzer"
date: "2023-05-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(reshape2)
library(rstanarm)
library(tidybayes)
source("../utils/plotting.r")
source("../utils/tex.r")
```

## Data

```{r}
env_data <- readRDS("../data-clean/environmental.rds")
epi_data <- readRDS("../data-clean/epidemiological.rds")

env_data <- na.omit(env_data)
env_epi_data <- left_join(
  env_data,
  epi_data %>% 
    group_by(class) %>%
    arrange(date) %>%
    mutate(cases = cumsum(new_cases)) %>%
    dplyr::select(class, date, weekday, n_present, cases),
  by = c("class", "date")
)

vars <- c("CO2", "rH", "Temp", "AER", "Cn", "PM1", "PM25", "PM4", "PM10")
vars_labels <- c("CO[2]~(ppm)", "Relative~humidity~('%')", "Temperature~(degree*C)", 
                 "Air~exchange~rate~(changes/h)", "CN~(1/cm^3)", "PM[1]~(mu*gm^{-3})", "PM[2.5]~(mu*gm^{-3})", "PM[4]~(mu*gm^{-3})", "PM[10]~(mu*gm^{-3})")
names(vars_labels) <- vars
vars_labels_no_units <- c("CO[2]", "Relative~humidity", "Temperature", "Air~exchange~rate",
                          "CN", "PM[1]", "PM[2.5]", "PM[4]", "PM[10]")
names(vars_labels_no_units) <- vars
```

## Boxplots

```{r}
# particle concentrations
boxpl_pc <- env_epi_data %>%
  dplyr::select(class, date, weekday, aircleaner, Cn, PM1, PM25, PM4, PM10) %>%
  melt(c("class", "date", "weekday", "aircleaner")) %>%
  rename(`Air Cleaner` = aircleaner,
         Class = class) %>%
  ggplot(aes(x = `Air Cleaner`, color = `Air Cleaner`, y = value)) +
  facet_wrap(~ variable, scales = "free_y", nrow = 1, dir = "h", labeller = as_labeller(vars_labels, label_parsed)) +
  geom_boxplot2() +
  geom_jitter(aes(shape = Class), alpha = .3, width = .2) +
  scale_color_manual(values = wes_palette("Royal1")) +
  scale_shape_manual(values = c(2,20)) +
  labs(title = "a") +
  theme_custom() +
  theme(axis.title.y = element_blank(), legend.position = "top",
        plot.title.position = "plot") +
  guides(color = "none")
save_plot(boxpl_pc, pdf_file = "../results/env-data/particles-boxplot.pdf", eps_file = "../results/env-data/particles-boxplot.eps", w = 16, h = 8)

# other variables
boxpl_oth <- env_epi_data %>%
  dplyr::select(class, date, weekday, aircleaner, CO2, AER, rH, Temp) %>%
  melt(c("class", "date", "weekday", "aircleaner")) %>%
  rename(`Air Cleaner` = aircleaner,
         Class = class) %>%
  ggplot(aes(x = `Air Cleaner`, color = `Air Cleaner`, y = value)) +
  facet_wrap(~ variable, scales = "free_y", nrow = 1, dir = "h", labeller = as_labeller(vars_labels, label_parsed)) +
  geom_boxplot2() +
  geom_jitter(aes(shape = Class), alpha = .3, width = .2) +
  scale_color_manual(values = wes_palette("Royal1")) +
  scale_shape_manual(values = c(2,20)) +
  theme_custom() +
  theme(axis.title.y = element_blank(), legend.position = "top") +
  guides(color = "none")
save_plot(boxpl_oth, pdf_file = "../results/env-data/otherVars-boxplot.pdf", w = 16, h = 8)
```

## Estimated change

```{r}
# log transform
env_epi_data_log <- env_epi_data %>%
  dplyr::select(-CO2, -PMtot, -Temp, -rH) %>%
  mutate(across(c(Cn, PM1, PM25, PM4, PM25, AER, n_present, cases), log))

# Bayesian linear regression model
seed0 <- 12345
draws <- list()
particle_vars <- c("Cn", "PM1", "PM25", "PM4", "PM10")
for (i in 1:length(particle_vars)) {
  f <- as.formula(paste0(particle_vars[[i]], "~ class + weekday + aircleaner + n_present + AER + cases"))
  m <- stan_glm(f, data = env_epi_data_log, seed = seed0)
  draws[[i]] <- as.data.frame(m)
}

# collect results
draws_filt <- lapply(draws, function(D) dplyr::select(D, `aircleanerYes`))
draws_filt <- do.call(rbind, draws_filt) %>%
  mutate(outcome = factor(rep(particle_vars, each = nrow(draws[[i]])), levels = rev(particle_vars))) %>%
  set_names(c("aircleaner", "outcome")) %>%
  mutate(aircleaner = (-1) * 100 * (exp(aircleaner) - 1))

# create estimation results table
effects_tbl <- draws_filt %>%
  mutate(outcome = recode(as.character(outcome), !!! vars_labels_no_units),
         outcome = factor(outcome, levels = rev(vars_labels_no_units))) %>%
  group_by(outcome) %>%
  mean_qi() %>%
  ungroup() %>%
  filter(.width == .95) %>%
  select(outcome, aircleaner, `.lower`, `.upper`) %>%
  mutate_if(is.numeric, round_k) %>%
  mutate(across(-outcome, tex_sign)) %>%
  arrange(-row_number()) %>%
  mutate(outcome = c("CN", "PM$_{1}$", "PM$_{2.5}$", "PM$_{4}$", "PM$_{10}$"),
         space = "") %>%
  as.matrix() %>%
  xtable() %>%
  print.xtable(
    file = "../results/env-data/estimation-results-table.tex",
    type = "latex",
    sanitize.text.function = identity,
    include.colnames = F,
    include.rownames = F,
    only.contents = T,
    hline.after = NULL)

# create estimation results figure
effects_env_main <- draws_filt %>%
  mutate(outcome = factor(outcome, levels = (c("Cn", "PM1", "PM25", "PM4", "PM10")))) %>%
  group_by(outcome) %>%
  mean_qi(.width = c(0.5, 0.8, 0.95)) %>%
  ungroup() %>%
  arrange(desc(.width)) %>%
  mutate(`.width` = factor(`.width`, levels = rev(c(.5, .8, .95)))) %>%
  ggplot() +
  geom_rect(aes(xmin = .lower, xmax = .upper, ymin = 0.75, ymax = 1.25, fill = `.width`)) +
  geom_segment(aes(x = aircleaner, xend = aircleaner, y = 0.75, yend = 1.25), size = 1.5 / cm(1)) +
  geom_point(aes(x = aircleaner, y = 1), size = 3, fill = "white", shape = 21) +
  facet_wrap(~ outcome, nrow = 1, labeller = as_labeller(vars_labels_no_units, label_parsed)) +
  scale_fill_manual(values = wes_palette("Royal1")[c(3,4,2)]) +
  scale_y_continuous(expand = expansion(mult = c(0.25, 0.25))) +
  scale_x_continuous(limits = c(50, 100), expand = expansion(mult = c(0.1, 0.1)), breaks = c(50, 75, 100)) +
  labs(x = "Estimated reduction (%)", fill = "CrI", title = "b") +
  theme_custom() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "top", legend.direction = "horizontal",
        legend.key.height = unit(.25, "cm"), legend.key.width = unit(.35, "cm"), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        plot.title.position = "plot",
        panel.spacing = unit(20, units = "pt")) 
save_plot(effects_env_main, 
          pdf_file = "../results/env-data/estimation-results-figure.pdf",
          eps_file = "../results/env-data/estimation-results-figure.eps",
          w = 16, h = 4)
```

### 2023-05-24 utils/plotting.R

@@ -2,10 +2,13 @@ require(ggpattern)
 require(gridExtra)
 require(GGally)
 require(Cairo)
+require(gg.layers)
+require(wesanderson) # https://github.com/karthik/wesanderson
 
 # ggplot theme for Nature style
 text_size = 8
 update_geom_defaults("text", list(size = text_size))
+
 theme_nature <- function () { 
   theme_classic(base_size = text_size, base_family = 'sans') %+replace% 
     theme(
@@ -13,6 +16,20 @@ theme_nature <- function () {
       axis.title = element_text(size = text_size),
       plot.title = element_text(size = text_size + 2, face = "bold", hjust = 0.5, margin = ggplot2::margin(0, 0, 5, 0))
     )
+}
+
+theme_custom <- function() {
+  theme_minimal() %+replace% 
+    theme(text = element_text(size = text_size),
+          axis.text = element_text(size = text_size),
+          axis.title = element_text(size = text_size),
+          plot.title = element_text(size = text_size + 2, face = "bold", hjust = 0, margin = ggplot2::margin(0, 0, 5, 0)),
+          strip.text = element_text(size = text_size),
+          panel.grid.major = element_blank(),
+          panel.border = element_blank(),
+          axis.line.x = element_line(),
+          axis.line.y = element_line(),
+          legend.text = element_text(size = 8))
 }
 
 theme_bw2 <- function () { 
@@ -80,7 +97,7 @@ hist_fun <- function(data, mapping) {
 cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
 
 
-save_plot <- function(pl, pdf_file = NULL, tikz_file = NULL, w = 8, h = 4) {
+save_plot <- function(pl, pdf_file = NULL, eps_file = NULL, tikz_file = NULL, w = 8, h = 4) {
   print(pl)
   if (!is.null(pdf_file)) {
     ggsave(pdf_file, width = w / cm(1), height = h / cm(1)) 
@@ -89,5 +106,8 @@ save_plot <- function(pl, pdf_file = NULL, tikz_file = NULL, w = 8, h = 4) {
     tikz(tikz_file, width = w / cm(1), height = h / cm(1))
     print(pl)
     dev.off()
-  } 
+  }  
+  if (!is.null(eps_file)) {
+    cairo_ps(filename = eps_file, width = w / cm(1), height = h / cm(1))
+  }
 }

### 2023-05-24 utils/tex.R

@@ -1,14 +1,4 @@
-# require(tikzDevice)
-# 
-# # Settings for tikzdevice
-# if (Sys.info()['sysname'] == 'Darwin') {
-#   options("tikzLatex" = Sys.which("pdflatex"))
-# } else {
-#   options("tikzLatex" = 'C:/texlive/2018/bin/win32/pdflatex')
-# }
-# 
-
-
+require(xtable)
 
 # Round function
 round_k <- function(x, k = 2) {

### 2023-05-30 analysis/env_analysis.Rmd
epidemiological data analysis

@@
 vars_labels <- c("CO[2]~(ppm)", "Relative~humidity~('%')", "Temperature~(degree*C)", 
-                 "Air~exchange~rate~(changes/h)", "CN~(1/cm^3)", "PM[1]~(mu*gm^{-3})", "PM[2.5]~(mu*gm^{-3})", "PM[4]~(mu*gm^{-3})", "PM[10]~(mu*gm^{-3})")
+                 "AER~(air~changes/h)", "CN~(1/cm^3)", "PM[1]~(mu*gm^{-3})", "PM[2.5]~(mu*gm^{-3})", "PM[4]~(mu*gm^{-3})", "PM[10]~(mu*gm^{-3})")
 names(vars_labels) <- vars

### 2023-05-30 analysis/epi_analysis.Rmd
epidemiological data analysis

---
title: "Epidemiological Data Analysis"
author: "Nicolas Banholzer"
date: "2023-05-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(reshape2)
library(tidybayes)
library(lubridate)
library(rstan)
source("../utils/epi.r")
source("../utils/bayes.r")
source("../utils/plotting.r")
source("../utils/tex.r")
```

## Data

```{r}
epi_data <- readRDS("../data-clean/epidemiological.rds")
env_data <- readRDS("../data-clean/environmental.rds")
sec_data <- readRDS("../data-clean/secondary.rds")
```

## Descriptives

```{r}
# number of cases by class
epi_data %>%
  filter(vacation == 0) %>%
  group_by(class) %>%
  summarize(cases = sum(new_cases))

# number of cases by intervention
epi_data %>%
  filter(vacation == 0) %>%
  group_by(aircleaner) %>%
  summarize(cases = sum(new_cases))

# number of cases by week
cases_by_week_pl <- epi_data %>%
  mutate(week = isoweek(date)) %>%
  filter(week != 6) %>%
  group_by(class, week) %>%
  summarize(cases = sum(new_cases),
            aircleaner = first(aircleaner)) %>%
  ungroup() %>%
  ggplot(aes(x = factor(week), y = cases, fill = aircleaner)) +
  geom_bar_pattern(stat = "identity", aes(pattern_angle = class), pattern = "stripe",
                   pattern_fill = "white", pattern_color = NA, color = "black", pattern_density = 0.25, pattern_spacing = 0.05) +
  labs(x = "Study week", y = "Number of cases", fill = "Air cleaner", pattern_angle = "Class", title = "a") +
  scale_fill_manual(values = wes_palette("Royal1")) +
  scale_x_discrete(breaks = c(1:5, 7:10)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_pattern_angle_discrete(range = c(-45, 45)) +
  theme_custom() +
  theme(legend.position = c(0.13, 0.8),
        legend.key.width = unit(0.5, "cm"),
        legend.key.height = unit(0.3, "cm"),
        plot.title.position = "plot") +
  guides(fill = guide_legend(nrow = 1),
         pattern_angle = guide_legend(nrow = 1))

save_plot(cases_by_week_pl,
          pdf_file = "../results/epi-data/cases_by_week.pdf",
          eps_file = "../results/epi-data/cases_by_week.eps",
          w = 12, h = 8) 
```

## Prepare data

```{r}
# stan data
stan_data <- list()

# number of seeding days
stan_data$S <- ceiling(2 * exp(p_in_mu_m + p_in_sigma_m^2/2))

# number of days from Saturday to study start
stan_data$SS <- seq(min(epi_data$date) - days(7), min(epi_data$date), by = "1 day")
stan_data$SS <- abs(as.numeric(stan_data$SS[weekdays(stan_data$SS) == "Saturday"] - min(min(epi_data$date))))

# number of new cases
stan_data$cases <- epi_data %>%
  dplyr::select(class, date, new_cases) %>%
  dcast(date ~ class) %>%
  arrange(date) %>%
  dplyr::select(A, B) %>%
  as.matrix()

# number of classes
stan_data$C <- ncol(stan_data$cases)

# number of days and weeks
stan_data$D <- nrow(stan_data$cases)
stan_data$W <- ceiling(stan_data$D / 7)

# total cases by weekday
stan_data$totCasesWeekday <- epi_data %>%
  group_by(weekday) %>%
  summarize(n = sum(new_cases)) %>%
  ungroup() %>%
  mutate(weekday = factor(weekday, levels = c("Saturday", "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))) %>%
  arrange(weekday) %>%
  dplyr::select(n) %>%
  unlist()

# weekend
stan_data$weekend <- epi_data %>%
  filter(class == "A") %>%
  dplyr::select(weekend) %>%
  unlist()
seeding_weekend <- epi_data$date[epi_data$class=="A"][1] - days(stan_data$S:1)
seeding_weekend <- ifelse(weekdays(seeding_weekend) %in% c("Saturday", "Sunday"), 1, 0)
stan_data$weekend <- c(seeding_weekend, stan_data$weekend)

# vacation
stan_data$vacation <- epi_data %>%
  filter(class == "A") %>%
  dplyr::select(vacation) %>%
  unlist()
stan_data$vacation <- c(rep(0, stan_data$S), stan_data$vacation)

# aircleaner intervention
stan_data$airCleaner <- epi_data %>%
  mutate(aircleaner = ifelse(aircleaner == "Yes", 1, 0)) %>%
  dplyr::select(class, date, aircleaner) %>%
  dcast(date ~ class) %>%
  arrange(date) %>%
  dplyr::select(A, B) %>%
  as.matrix()
seeding_airCleaner <- matrix(0, nrow = stan_data$S, ncol = stan_data$C) 
stan_data$airCleaner <- rbind(seeding_airCleaner, stan_data$airCleaner)

# proportion of students in class
stan_data$stud <- epi_data %>%
  mutate(prop_in_class = n_present / n_class,
         prop_in_class = (prop_in_class - mean(prop_in_class)) / sd(prop_in_class)) %>%
  dplyr::select(class, date, prop_in_class) %>%
  dcast(date ~ class) %>%
  arrange(date) %>%
  dplyr::select(A, B) %>%
  as.matrix()

# air change rate
stan_data$vent <- epi_data %>%
  dplyr::select(class, date, vacation, weekend) %>%
  left_join(env_data %>% dplyr::select(class, date, AER), by = c("class", "date")) %>%
  mutate(AER = (AER - mean(AER, na.rm = T)) / sd(AER, na.rm = T)) %>%
  mutate(AER = ifelse(vacation + weekend > 0, -1001, AER),
         AER = ifelse(is.na(AER), +1001, AER)) %>%
  dplyr::select(class, date, AER) %>%
  dcast(date ~ class) %>%
  arrange(date) %>%
  dplyr::select(A, B) %>%
  as.matrix()
stan_data$missingVent <- apply(stan_data$vent, c(1,2), function(x) ifelse(x == 1001, 1, 0))
stan_data$Mvent <- sum(stan_data$missingVent)
stan_data$missingVent <- apply(stan_data$missingVent, 2, function(x) {
  y <- cumsum(x)
  dy <- c(0, diff(y))
  y <- ifelse(dy == 0, 0, y)
  return(y)
})

# share of positive covid-19 tests in the community
stan_data$cov <- sec_data %>%
  filter(between(date, min(epi_data$date - days(stan_data$S)), max(epi_data$date))) %>%
  mutate(covid_prop_pos = (covid_prop_pos - mean(covid_prop_pos)) / sd(covid_prop_pos)) %>%
  dplyr::select(covid_prop_pos) %>%
  unlist()

# consultations for influenza-like in the community
stan_data$ili <- sec_data %>%
  filter(between(date, min(epi_data$date - days(stan_data$S)), max(epi_data$date))) %>%
  mutate(ili_consult = (ili_consult - mean(ili_consult)) / sd(ili_consult)) %>%
  dplyr::select(ili_consult) %>%
  unlist()

# hyperparameters for prior for incubation period distribution
stan_data$p_in_mu_m <- p_in_mu_m
stan_data$p_in_mu_s <- p_in_mu_s
stan_data$p_in_sigma_m <- p_in_sigma_m
stan_data$p_in_sigma_s <- p_in_sigma_s
```

## Modeling

```{r}
# stan model code file
stan_file <- "../models/latent-infection-model.stan"

# run model
latInfMod <- stan(stan_file, data = stan_data, seed = 12345, cores = 4)

# quick estimation results
summary(latInfMod, probs = c(0.025, 0.975), pars = c("phi", "alpha", "omega", "beta", "vega", "ventNa", "mu_p_in", "sigma_p_in", "I0"))$summary %>%
  data.frame() %>%
  mutate_all(round, 2)
mcmc_intervals(latInfMod, pars = "beta[1]")
```

## Results

### Adjusted Rate Ratio

```{r}
arr <- spread_draws(latInfMod, `beta[1]`) %>%
  rename(effect = `beta[1]`) %>%
  dplyr::select(effect) %>%
  mean_qi(.width = c(0.5, 0.8, .95)) %>%
  mutate(across(c(effect, .lower, .upper), exp))
arr
```

### Avoided infections

```{r}
types <- c("Partial instalation of air cleaners (study design)",
           "No installation of air cleaners (hypothetical)",
           "Full installation of air cleaners (hypothetical)")

avoid_infections <- spread_draws(latInfMod, mu_cases_ac[day,class,type]) %>%
  group_by(.draw, type) %>%
  summarize(infections = sum(mu_cases_ac)) %>%
  ungroup() %>%
  group_by(type) %>%
  mean_qi(.width = c(0.5, 0.8, 0.95)) %>%
  ungroup() %>%
  mutate(type = ifelse(type == 1, types[1], ifelse(type == 2, types[2], types[3])),
         type = factor(type, levels = types[c(2,1,3)])) %>%
  arrange(desc(.width)) %>%
  mutate(`.width` = factor(`.width`, levels = rev(c(.5, .8, .95))))

avoid_infections_pl <- avoid_infections  %>%
  ggplot() +
  geom_rect(aes(xmin = .lower, xmax = .upper, ymin = 0.75, ymax = 1.25, fill = `.width`)) +
  geom_segment(aes(x = infections, xend = infections, y = 0.75, yend = 1.25), size = 1.5 / cm(1)) +
  geom_point(aes(x = infections, y = 1), size = 3, fill = "white", shape = 21) +
  facet_wrap(~ type, ncol = 1) +
  scale_fill_manual(values = wes_palette("Royal1")[c(3,4,2)]) +
  scale_y_continuous(expand = expansion(mult = c(0.25, 0.25))) +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0))) +
  labs(x = "Estimated average number of infections", fill = "CrI", title = "b") +
  theme_custom() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "top", legend.direction = "horizontal",
        legend.key.height = unit(.25, "cm"), legend.key.width = unit(.35, "cm"), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        plot.title.position = "plot",
        panel.spacing = unit(20, units = "pt"),
        strip.text.x = element_text(margin = margin(b = 5)))

save_plot(avoid_infections_pl, 
          pdf_file = "../results/epi-data/avoided-infections.pdf",
          eps_file = "../results/epi-data/avoided_infections.eps",
          w = 16, h = 8)
```

### Estimation results

```{r}
est_res <- summary(latInfMod, probs = c(0.025, 0.975), pars = c("phi", "alpha", "omega", "beta", "vega", "mu_p_in", "sigma_p_in"))$summary %>%
  data.frame() %>%
  add_rownames() %>%
  rename(variable = rowname) %>%
  mutate(variable = c("$\\phi$",
                      "$\\alpha_{\\text{A}}$",
                      "$\\alpha_{\\text{A}}$",
                      "$\\omega$",
                      "$\\beta_1$",
                      "$\\beta_2$",
                      "$\\beta_3$",
                      "$\\beta_4$",
                      "$\\beta_5$",
                      "$\\beta_6$",
                      "$\\vega_{\\text{Saturday}}$",
                      "$\\vega_{\\text{Sunday}}$",
                      "$\\vega_{\\text{Monday}}$",
                      "$\\vega_{\\text{Tuesday}}$",
                      "$\\vega_{\\text{Wednesday}}$",
                      "$\\vega_{\\text{Thursday}}$",
                      "$\\vega_{\\text{Friday}}$",
                      "$\\mu_{p_{\\text{IN}}}",
                      "$\\sigma_{p_{\\text{IN}}}")) %>%
  dplyr::select(variable, mean, `X2.5.`, `X97.5.`, n_eff, Rhat) %>%
  mutate(across(c(mean, `X2.5.`, `X97.5.`), function(x) tex_sign(round_k(x)))) %>%
  as.matrix()

print.xtable(xtable(est_res),
             type = "latex",
             file = "../results/epi-data/estimation-results.tex",
             sanitize.text.function = identity,
             include.colnames = F,
             include.rownames = F,
             only.contents = T,
             hline.after = NULL)
```

### Model fit

```{r}
fitted_cases <- spread_draws(latInfMod, muCases[day,class]) %>%
  mutate(class = ifelse(class == 1, "A", "B")) %>%
  filter(day > stan_data$SS) %>%
  mutate(day = day - stan_data$SS) %>%
  left_join(epi_data %>%
              dplyr::select(class, date, new_cases) %>%
              group_by(class) %>%
              arrange(date) %>%
              mutate(day = 1:n()) %>%
              ungroup(),
            by = c("class", "day")) %>%
  mutate(week = isoweek(date)) %>%
  filter(week != 6) %>%
  group_by(week, .draw) %>%
  summarise(estimated_cases = sum(muCases),
            true_cases = sum(new_cases)) %>%
  ungroup()

fitted_cases_pl <- ggplot(mapping = aes(x = week)) +
  stat_lineribbon(data = fitted_cases,
                  mapping = aes(y = estimated_cases),
                  .width = c(0.5, 0.8, 0.95), color = "brown") +
  geom_line(data = fitted_cases %>% group_by(week) %>% slice(1),
            mapping = aes(y = true_cases, color = "Observed"), size = 1.5) +
  scale_fill_manual(values = wes_palette("Royal1")[c(3,4,2)]) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_color_manual(values = "black") +
  labs(y = "Number of new infections", x = "Study week", fill = "Estimated (CrI)") +
  theme_custom() +
  theme(legend.position = "top") +
  guides(color = guide_legend(title = element_blank()))

save_plot(fitted_cases_pl, pdf_file = "../results/epi-data/model-fit.pdf",
          w = 12, h = 10)
```

### 2023-05-30 preprocessing/prep_palas.r

@@ -42,7 +42,7 @@ df <- df %>%
   filter(!(weekdays(date) %in% c("Saturday", "Sunday")),
          !(date %in% seq.Date(as.Date("2023-02-06"), as.Date("2023-02-11"), by = "1 day")),
          date >= as.Date("2023-01-16"),
-         date <= as.Date("2023-03-11")) 
+         date <= as.Date("2023-03-10")) 
 
 # move time in class A (E3b) on 2023-01-16 (first weekend, Monday) by one hour backwards (see data-collection-issues.docx)
 df$datetime[df$class=="A"&df$date==as.Date("2023-01-16")] <- df$datetime[df$class=="A"&df$date==as.Date("2023-01-16")] %m-% hours(1)

### 2023-05-30 preprocessing/prep_redcap.r

@@ -129,7 +129,7 @@ resp_cases <- cases %>%
   rename(date = date_symptom)
 
 # full join
-epi_data <- data.frame(date = rep(seq.Date(as.Date("2023-01-16"), as.Date("2023-03-11"), by = "1 day"), 2)) %>%
+epi_data <- data.frame(date = rep(seq.Date(min(resp_cases$date), as.Date("2023-03-10"), by = "1 day"), 2)) %>%
   mutate(class = rep(c("B", "A"), each = nrow(.) / 2),
          n_class = rep(c(17, 20), each = nrow(.) / 2),
          weekday = weekdays(date),

### 2023-05-30 preprocessing/prep_secondary.R

@@ -1,0 +1,39 @@
+# libraries
+library(tidyverse)
+library(lubridate)
+
+# raw data
+covid <- read_csv("data-raw/secondary/COVID19Test_geoRegion_w.csv") 
+ili <- readxl::read_excel("data-raw/secondary/inzidenz-grippeverdacht-de.xlsx")
+
+# preprocessing covid 
+covid <- covid %>%
+  filter(geoRegion == "SO") %>%
+  dplyr::select(datum, anteil_pos) %>%
+  mutate(year = as.numeric(substr(datum, 1, 4)),
+         week = as.numeric(substr(datum, 5, 6))) %>%
+  filter(year == 2023) %>%
+  dplyr::select(-datum)
+
+start_week <- as.numeric(paste0("2023", isoweek(as.Date("2023-01-01"))))
+end_week <- as.numeric(paste0("2023", isoweek(as.Date("2023-03-11"))))
+date <- seq.Date(as.Date("2023-01-01"), as.Date("2023-03-11"), by = "1 day")
+
+covid_share_pos <- data.frame(date = date) %>%
+  mutate(year = year(date),
+         week = isoweek(date)) %>%
+  left_join(covid, by = c("year", "week")) %>%
+  rename(covid_prop_pos = anteil_pos)
+  
+
+# preprocessing
+ili <- ili[-c(1,2),12:14] %>%
+  mutate_all(as.numeric) %>%
+  set_names(c("year", "week", "ili_consult")) 
+
+
+# combine
+commun_trans <- left_join(covid_share_pos, ili, by = c('year', 'week')) 
+
+# save
+saveRDS(commun_trans, "data-clean/secondary.rds")

### 2023-06-15 analysis/env_analysis.Rmd
add molecular data, analysis, and results, update epi model and results

@@
 save_plot(boxpl_pc, pdf_file = "../results/env-data/particles-boxplot.pdf", eps_file = "../results/env-data/particles-boxplot.eps", w = 16, h = 8)
+saveRDS(boxpl_pc, file = "../results/env-data/particles-boxplot.rds")
 

### 2023-06-15 analysis/epi_analysis.Rmd
add molecular data, analysis, and results, update epi model and results

@@
 env_data <- readRDS("../data-clean/environmental.rds")
+mol_data <- readRDS("../data-clean/molecular.rds") 
 sec_data <- readRDS("../data-clean/secondary.rds")
@@
 
+### Assumptions
+
+* Incubation periods are informed by https://www.sciencedirect.com/science/article/pii/S1473309909700696?casa_token=MSYyhmCLcLAAAAAA:10ooZqTNmMMfk6sIB1idSETWMusOs3Y-tKWPqGUXOFiD0PPFB0edT3W5OZyDO13S6kWPTv8NDOU and for SARS-CoV-2 from https://bmjopen.bmj.com/content/bmjopen/10/8/e039652.full.pdf.
+* We assume a lognormal distribution and use the median and dispersion estimate
+* We weight the incubation period distributions each week according to the proportion of positive samples for each virus --> effective incubation period.
+* For Metapneumovirus, there was insufficient data. Here we take the mean of the parameters of the other viruses.
+
+```{r}
+# mean estimate
+p_in_mean_m <- c(
+  5.6, # AdV
+  0.6, # Flu B
+  2.6, # PIV
+  1.9 # HRV
+)
+
+# 95% limits of estimate
+p_in_mean_ul <- matrix(c(
+  6.3, 4.8,
+  0.6, 0.5,
+  3.1, 2.1,
+  2.4, 1.4
+), ncol = 2, byrow = T)
+
+# mean dispersion
+p_in_sd_m <- c(
+  1.26,
+  1.51,
+  1.35,
+  1.68
+)
+
+# upper and lower limit of SD
+p_in_sd_ul <- matrix(c(
+  1.38, 1.13,
+  1.64, 1.37,
+  1.55, 1.16,
+  2.01, 1.16
+), ncol = 2, byrow = T)
+
+# Convert to lognormal distr parameters
+p_in_mu_m <- c(log(p_in_mean_m), 1.63) # the last value is for SARS-CoV-2
+p_in_mu_s <- c(log(p_in_mean_ul[,1] / p_in_mean_ul[,2]), (1.75 - 1.51)) / (2 * qnorm(.975))
+p_in_sigma_m <- c(log(p_in_sd_m), 0.5)
+p_in_sigma_s <- c(log(p_in_sd_ul[,1] / p_in_sd_ul[,2]), (0.55 - 0.46)) / (2 * qnorm(.975))
+
+# visualize
+set.seed(12345)
+
+prior_pin_par <- data.frame(
+  virus = c("AdV", "Flu B", "PIV", "HRV", "CoV", "MPV"),
+  mum = c(p_in_mu_m, mean(p_in_mu_m)), 
+  mus = c(p_in_mu_s, mean(p_in_mu_s)),
+  sim = c(p_in_sigma_m, mean(p_in_sigma_m)),
+  sis = c(p_in_sigma_s, mean(p_in_sigma_s))
+) %>%
+  mutate(virus = factor(virus, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV"))) %>%
+  arrange(virus)
+
+prior_pin <- prior_pin_par %>%
+  mutate(mu = map2(mum, mus, function(m, s) rnorm(n = 4e3, mean = m, sd = s)),
+         sigma = map2(sim, sis, function(m, s) rtrunc(n = 4e3, spec = "norm", a = 0, b = Inf, mean = m, sd = s))) %>%
+  unnest() %>%
+  mutate(pin = map2(mu, sigma, function(log_mu, log_sigma) data.frame(x = 0:10, y = vp(10, T, p_in, log_mu, log_sigma)))) %>%
+  unnest()
+
+prior_pin_pl <- prior_pin %>%
+  ggplot(aes(x = x, y = y)) +
+  stat_lineribbon(point_interval = "mean_qi") +
+  scale_fill_brewer() + 
+  facet_wrap(~ virus) +
+  labs(x = "s (delay in days)", y = expression(p[IN]*" (%)")) +
+  scale_x_continuous(expand = c(0,0), breaks = seq(0, 10, 2)) +
+  scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
+  theme_bw2() +
+  theme(plot.title = element_text(hjust = 0), legend.position = "none")
+prior_pin_pl
+
+save_plot(prior_pin_pl, pdf_file = "../results/epi-data/incubation-periods.pdf", w = 16, h = 10)
+```
+
+### Stan data
+
 ```{r}
@@
 # number of seeding days
-stan_data$S <- ceiling(2 * exp(p_in_mu_m + p_in_sigma_m^2/2))
+stan_data$S <- ceiling(2 * exp(max(p_in_mu_m) + max(p_in_sigma_m)^2/2))
 
@@
 
-# hyperparameters for prior for incubation period distribution
-stan_data$p_in_mu_m <- p_in_mu_m
-stan_data$p_in_mu_s <- p_in_mu_s
-stan_data$p_in_sigma_m <- p_in_sigma_m
-stan_data$p_in_sigma_s <- p_in_sigma_s
+# number of viruses
+stan_data$K <- 6
+
+# positive molecular samples by week and class 
+mol_class_week <- mol_data %>%
+  mutate(week = week(date),
+         week = week - min(week) + 1,
+         week = ifelse(week > 4, week - 1, week))  %>%
+  dplyr::select(week, class, AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`) %>%
+  melt(c("week", "class")) %>%
+  group_by(class, week, variable) %>%
+  summarize(n = sum(value)) %>%
+  ungroup() %>%
+  mutate(variable = as.character(variable), 
+         variable = ifelse(variable == "SARS-CoV-2", "CoV", variable),
+         variable = factor(variable, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")))
+mol_class_week_extend <- epi_data %>%
+  dplyr::select(class, date) %>%
+  mutate(week = week(date),
+         week = week - min(week) + 1,
+         week = ifelse(week > 4, week - 1, week))  %>%
+  left_join(mol_class_week) %>%
+  arrange(class, date) %>%
+  dplyr::select(class, date, variable, n) %>%
+  dcast(class + date ~ variable)
+stan_data$pos_samples <- simplify2array(by(mol_class_week_extend, mol_class_week_extend$class, as.matrix))
+stan_data$pos_samples <- apply(stan_data$pos_samples[,-c(1,2),], c(1,2,3), as.integer)
+
+# hyperparameters for prior for incubation period distributions
+stan_data$p_in_mu_m <- prior_pin_par$mum
+stan_data$p_in_mu_s <- prior_pin_par$mus
+stan_data$p_in_sigma_m <- prior_pin_par$sim
+stan_data$p_in_sigma_s <- prior_pin_par$sis
 ```
@@
 # quick estimation results
-summary(latInfMod, probs = c(0.025, 0.975), pars = c("phi", "alpha", "omega", "beta", "vega", "ventNa", "mu_p_in", "sigma_p_in", "I0"))$summary %>%
+summary(latInfMod, probs = c(0.025, 0.5, 0.975), pars = c("phi", "alpha", "omega", "beta", "vega", "mu_p_in", "sigma_p_in", "I0"))$summary %>%
   data.frame() %>%
   mutate_all(round, 2)
-mcmc_intervals(latInfMod, pars = "beta[1]")
 ```
@@
 ```{r}
-types <- c("Partial instalation of air cleaners (study design)",
-           "No installation of air cleaners (hypothetical)",
-           "Full installation of air cleaners (hypothetical)")
+types <- c("Air cleaners partially installed",
+           "Air cleaners never installed",
+           "Air cleaners always installed")
 
@@
   mutate(type = ifelse(type == 1, types[1], ifelse(type == 2, types[2], types[3])),
-         type = factor(type, levels = types[c(2,1,3)])) %>%
+         type = factor(type, levels = types[c(3,1,2)])) %>%
   arrange(desc(.width)) %>%
-  mutate(`.width` = factor(`.width`, levels = rev(c(.5, .8, .95))))
+  mutate(`.width` = factor(`.width`, levels = rev(c(.5, .8, .95)))) %>%
+  filter(type != types[1])
 
@@
   scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0))) +
-  labs(x = "Estimated average number of infections", fill = "CrI", title = "b") +
+  labs(x = "Estimated number of infections", fill = "CrI", title = "b") +
   theme_custom() +
@@
 
+```{r}
+barplot_mol <- readRDS("../results/mol-data/descriptives-cond.rds") + 
+  coord_flip() +
+  guides(fill = guide_legend(reverse = TRUE, nrow = 1)) +
+  theme(legend.position = "bottom", legend.key.height = unit(.1, "cm"), legend.key.width = unit(.5, "cm"), plot.title.position = "plot") +
+  ggtitle("a | Number of positive saliva samples by study condition.",
+          subtitle = "Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza (PIV).")
+boxplot_env <- readRDS("../results/env-data/particles-boxplot.rds") +
+  ggtitle("b | Aerosol number (CN) and particle matter (PM) concentrations by study condition.",
+          subtitle = expression("Particle diameter between 175nm and 20"*mu*"m. PM for sizes <1 to <10"*mu*"m."))
+avoid_infections_pl <- avoid_infections_pl +
+  ggtitle("c | Estimated number of infections with and without air cleaners. ",
+          subtitle = "Counterfactual scenarios with air cleaners always (never) installed, based on the model-estimated effect of air cleaners.")
+env_epi_comb_pl <- grid.arrange(barplot_mol, 
+                                grid.arrange(boxplot_env, avoid_infections_pl, nrow = 2, heights = c(8, 8), widths = 16),
+                                nrow = 2, heights = c(6, 16), widths = 16)
+ggsave(plot = env_epi_comb_pl, "../doc/Epidemics2023/figure.png", width = 16 / cm(1), height = 22 / cm(1))
+```
+
 ### Estimation results
@@
   mutate(variable = c("$\\phi$",
-                      "$\\alpha_{\\text{A}}$",
-                      "$\\alpha_{\\text{A}}$",
+                      "$\\alpha$",
                       "$\\omega$",
@@
                       "$\\vega_{\\text{Friday}}$",
-                      "$\\mu_{p_{\\text{IN}}}",
-                      "$\\sigma_{p_{\\text{IN}}}")) %>%
+                      "$\\mu_{p_{\\text{IN}}^{\\text{Flu B}}}",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{HRV}}}",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{AdV}}}",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{CoV}}}",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{MPV}}}",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{PIV}}}",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{Flu B}}}",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{HRV}}}",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{AdV}}}",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{CoV}}}",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{MPV}}}",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{PIV}}}")) %>%
   dplyr::select(variable, mean, `X2.5.`, `X97.5.`, n_eff, Rhat) %>%
@@
   guides(color = guide_legend(title = element_blank()))
+fitted_cases_pl
 

### 2023-06-15 analysis/mol_analysis.Rmd
add molecular data, analysis, and results, update epi model and results

---
title: "Molecular Analysis"
author: "Nicolas Banholzer"
date: "2023-06-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(reshape2)
library(tidybayes)
library(lubridate)
library(rstan)
source("../utils/epi.r")
source("../utils/bayes.r")
source("../utils/plotting.r")
source("../utils/tex.r")
```

## Data

```{r}
df <- readRDS("../data-clean/molecular.rds")
```

## Descriptives

```{r}
# positive samples by study condition and week
df_pos <- df %>%
  mutate(week = week(date),
         week = week - min(week) + 1,
         week = ifelse(week > 4, week - 1, week)) %>%
  dplyr::select(aircleaner, week, class, AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`) %>%
  melt(c("aircleaner", "week", "class")) %>%
  group_by(aircleaner, class, week, variable) %>%
  summarize(n = sum(value)) %>%
  ungroup() %>%
  mutate(variable = as.character(variable), 
         variable = ifelse(variable == "SARS-CoV-2", "CoV", variable),
         variable = factor(variable, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")))

pos_week_pl <- ggplot(mapping = aes(x = week, y = n, fill = variable)) +
  geom_bar(data = df_pos, stat = "identity") +
  geom_text(data = subset(df_pos, n > 0), mapping = aes(label = n), 
            size = 8 / cm(1), position = position_stack(vjust = 0.5), color = "white") +
  facet_wrap(~ class) +
  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
  scale_y_continuous(expand = expansion(add = c(0,1))) +
  scale_x_continuous(breaks = seq(1, 7, 1)) +
  labs(y = "Number of positive samples", x = "Study week", fill = "Virus", title = "c") +
  theme_custom() +
  theme(panel.grid = element_blank(),
        legend.position = "none") +
  guides(fill = guide_legend(nrow = 1))
pos_week_pl
```

```{r}
# positive samples by study condition
df_pos_stud <- df_pos %>%
  group_by(aircleaner, variable) %>%
  summarize(n = sum(n)) %>%
  ungroup()
pos_cond_pl <- ggplot(mapping = aes(x = aircleaner, y = n, fill = variable)) +
  geom_bar(data = df_pos_stud, stat = "identity", position = "stack") +
  geom_text(data = subset(df_pos_stud, n != 0), mapping = aes(label = n), 
            size = 8 / cm(1), position = position_stack(vjust = 0.5), color = "white") +
  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
  scale_y_continuous(expand = c(0,0)) +
  labs(y = "Number of positive samples", x = "Air cleaner", fill = "Virus", title = "b") +
  theme_custom() +
  theme(panel.grid = element_blank(),
        legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 1))
pos_cond_pl
```

```{r}
# Total counts by virus
df_pos_tot <- df_pos %>%
  group_by(variable) %>%
  summarize(n = sum(n)) %>%
  ungroup()

pos_tot_pl <- df_pos_tot %>%
  ggplot(aes(x = variable, y = n, fill = variable)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = n, y = n + 1), size = 8 / cm(1)) +
  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
  scale_y_continuous(expand = expansion(add = c(0,1))) +
  labs(y = "Number of positive samples", x = "Air cleaner", fill = "Virus", title = "a") +
  theme_custom() +
  theme(panel.grid = element_blank(),
        legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 1))
pos_tot_pl
```

```{r}
# combine
descr <- pos_tot_pl + pos_cond_pl & theme(legend.position = "bottom", legend.key.height = unit(.1, "cm"),
                                          legend.key.width = unit(.3, "cm"))
descr <- descr + plot_layout(guides = "collect", widths = c(10, 6), ncol = 2) 
descr <- descr + pos_week_pl 
descr <- descr + plot_layout(nrow = 2, heights = c(8, 8), design = 'AB
                             CC')
save_plot(descr, pdf_file = "../results/mol-data/descriptives.pdf", eps_file = "../results/mol-data/descriptives.eps", h = 16, w = 16)
saveRDS(pos_tot_pl + ggtitle(""), file = "../results/mol-data/descriptives-total.rds")
saveRDS(pos_cond_pl + ggtitle(""), file = "../results/mol-data/descriptives-cond.rds")
saveRDS(pos_week_pl + ggtitle(""), file = "../results/mol-data/descriptives-week.rds")
```

## Prepare data

```{r}
create.stan_data <- function(dat) {

  # stan data list
  stan_data <- list()

  # dimensions
  stan_data$N <- nrow(dat) / 2
  stan_data$K <- 6 + 1
  stan_data$C <- 2

  # outcome
  stan_data$samples <- dat %>%
    arrange(class, date) %>%
    dplyr::select(Neg, AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`) %>%
    as.matrix()

  # air cleaner indicator
  stan_data$aircleaner <- dat %>%
    arrange(class, date) %>%
    mutate(aircleaner = ifelse(aircleaner == "Yes", 1, 0)) %>%
    dplyr::select(aircleaner) %>%
    unlist()

  # class
  stan_data$schoolclass <- dat %>%
    arrange(class, date) %>%
    mutate(class = ifelse(class == "A", 0, 1)) %>%
    dplyr::select(class) %>%
    unlist()

  # susceptibles 
  stan_data$susceptibles <- dat %>%
    group_by(class) %>%
    arrange(date) %>%
    mutate(across(c(AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`), ~ n_class - cumsum(.x)),
           across(c(AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`), ~ (.x - mean(.x)) / sd(.x)),
           PIV = 0) %>% # no variation in PIV
    ungroup() %>%
    arrange(class, date) %>%
    dplyr::select(AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`)

  return(stan_data)
}

stan_data <- create.stan_data(df)
```

## Modeling

### Main model

```{r}
# stan model code file
stan_file <- "../models/multinomial-logit-model.stan"

# run model
multLogitMod <- stan(stan_file, data = stan_data, seed = 12345, cores = 4)

# quick estimation results
summary(multLogitMod, probs = c(0.025, 0.975), pars = c("beta0", "beta1", "beta2", "beta3"))$summary %>%
  data.frame() %>%
  mutate_all(round, 2)
```

### Sensitivity

```{r}
# lag data by one test date and one week
df_l1d <- df %>%
  group_by(class) %>%
  mutate(aircleaner = lag(aircleaner),
         aircleaner = ifelse(is.na(aircleaner), "No", aircleaner)) %>%
  ungroup()

df_l1w <- df %>%
  group_by(class) %>%
  mutate(aircleaner = lag(aircleaner, 2),
         aircleaner = ifelse(is.na(aircleaner), "No", aircleaner)) %>%
  ungroup()

# stan data list
stan_data_l1d <- create.stan_data(df_l1d)
stan_data_l1w <- create.stan_data(df_l1w)

# model
multLogitMod_l1d <- stan(stan_file, data = stan_data_l1d, seed = 12345, cores = 4)
multLogitMod_l1w <- stan(stan_file, data = stan_data_l1w, seed = 12345, cores = 4)
```

## Results

```{r}
# lag types
lag_types <- c("None",
               "1 test date",
               "1 week")

# collect draws
main_res <- spread_draws(multLogitMod, beta1)
sens_res_l1d <- spread_draws(multLogitMod_l1d, beta1)
sens_res_l1w <- spread_draws(multLogitMod_l1w, beta1)
res <- rbind(
  main_res %>% mutate(lag = lag_types[1]),
  sens_res_l1d %>% mutate(lag = lag_types[2]),
  sens_res_l1w %>% mutate(lag = lag_types[3])
) %>%
  mutate(beta1 = exp(beta1))

# plot
res_pl <- res  %>%
  ggplot() +
  stat_pointinterval(aes(x = beta1, y = lag), point_fill = "white", shape = 21, point_size = 3, .width = c(0.5, 0.8, 0.95)) +
  geom_vline(aes(xintercept = 1), linetype = "dashed") +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0)), breaks = seq(0, 2, .25)) +
  scale_size_continuous(range = c(1, 9)) +
  labs(x = "Relative risk ratio", y = "Lag of sampling results", fill = "CrI") +
  theme_custom() +
  theme(legend.position = "top", legend.direction = "horizontal",
        legend.key.height = unit(.25, "cm"), legend.key.width = unit(.35, "cm"))
res_pl
save_plot(res_pl, pdf_file = "../results/mol-data/model-results.pdf", w = 10, h = 6)
```

### 2023-06-15 preprocessing/prep_molecular.R

@@ -1,0 +1,58 @@
+#### Libraries ####
+
+library(tidyverse)
+library(readxl)
+library(lubridate)
+
+
+#### Data ####
+
+# read data
+df <- read_xlsx("data-raw/molecular/ZF Resultate Airchecker 2023.xlsx", sheet = "Spuckproben") %>%
+  rename(class = Klasse,
+         date = Datum,
+         virus = `Auto   Interpretation`) %>%
+  dplyr::select(class, date, virus) %>% # We set A as the "Studienklasse" and B as the "Kontrollklasse" --> 3a is B and 3b is A
+  mutate(class = ifelse(class == "3a", "B", "A"),
+         date = as.Date(gsub("2022", "2023", as.character(date))))
+
+
+#### Preprocessing ####
+
+# counts by date 
+df <- df %>%
+  group_by(class, date, virus) %>%
+  summarize(n = n()) %>%
+  ungroup()
+
+# add 0s
+expanded_df <- expand.grid(
+  class = unique(df$class), 
+  date = unique(df$date),
+  virus = unique(df$virus))
+df <- expanded_df %>%
+  left_join(df, by = c("class", "date", "virus")) %>%
+  mutate(n = ifelse(is.na(n), 0, n))
+
+# total count
+df_wide <- reshape2::dcast(df, class + date ~ virus) %>%
+  rename(Neg = `-`) %>%
+  mutate(N = Neg + AdV + `Flu B` + HRV + MPV + PIV + `SARS-CoV-2`)
+
+# add additional info
+df_wide <- df_wide %>%
+  mutate(weekday = weekdays(date),
+         aircleaner = ifelse(class == "A", 
+                             ifelse(date <= as.Date("2023-01-28"), "Yes",
+                                    ifelse(date >= as.Date("2023-02-27"), "Yes", "No")),
+                             ifelse(date <= as.Date("2023-01-28"), "No",
+                                    ifelse(date >= as.Date("2023-02-27"), "No", "Yes"))),
+         n_class = ifelse(class == "A", 20, 17),
+         n_class = ifelse(class == "B" & date > as.Date("2023-02-11"), 18, n_class)) %>%
+  dplyr::select(class, date, weekday, aircleaner, everything())
+
+
+#### Save ####
+
+saveRDS(df_wide, file = "data-clean/molecular.rds")
+

### 2023-06-15 preprocessing/prep_redcap.r

@@ -8,7 +8,7 @@ library(lubridate)
 #### Data ####
 
 df <- read.csv("data-raw/redcap/MCIDSchulenFollowup_DATA_2023-02-09_1356.csv") %>%
-  mutate(class = ifelse(grepl("Kontroll", record_id), "B", "A"),  # A is Studien- and B is Kontrollklasse
+  mutate(class = ifelse(grepl("Kontroll", record_id), "B", "A"),  # We set A as the "Studienklasse" and B as the "Kontrollklasse"
          date = as.Date(date, format = "%Y-%m-%d")) 
 
 

### 2023-06-15 utils/epi.r

@@ -10,7 +10,7 @@
 p_in_mu_m <- 1.63
 p_in_mu_s <- (1.75 - 1.51) / (2 * qnorm(0.975))
 p_in_sigma_m <- 0.50
-p_in_sigma_s <- (0.55 - 0.45) / qnorm(0.975)
+p_in_sigma_s <- (0.55 - 0.46) / (2 * qnorm(0.975))
 
 p_in <- function(x, log_mu = 1.63, log_sigma = 0.50) { 
   plnorm(x, log_mu, log_sigma)

### 2023-06-15 utils/plotting.R

@@ -1,3 +1,4 @@
+require(patchwork)
 require(ggpattern)
 require(gridExtra)
 require(GGally)

### 2023-06-20 analysis/epi_analysis.Rmd
add line list data and minor results update

@@
 epi_data <- readRDS("../data-clean/epidemiological.rds")
+epi_data_ll <- readRDS("../data-clean/redcap-long.rds")
 env_data <- readRDS("../data-clean/environmental.rds")
@@
 ## Descriptives
+
+### Line List
+
+```{r}
+line_list_dat <- epi_data_ll %>%
+  filter(respiratory_infection == "yes") %>%
+  dplyr::select(class, date_symptom, date_absent, date_back) %>%
+  as.matrix() %>%
+  xtable::xtable()
+
+print.xtable(line_list_dat,
+             type = "latex",
+             file = "../results/epi-data/line-list-data.tex",
+             sanitize.text.function = identity,
+             include.colnames = F,
+             include.rownames = F,
+             only.contents = T,
+             hline.after = NULL)
+```
+
+### Number of cases
 

### 2023-06-20 analysis/mol_analysis.Rmd
add line list data and minor results update

@@
 df <- readRDS("../data-clean/molecular.rds")
+df_ll <- readRDS("../data-clean/molecular-long.rds")
 ```
@@
 ## Descriptives
+
+### Line list
+
+```{r}
+line_list_dat <- df_ll %>%
+  as.matrix() %>%
+  xtable::xtable()
+
+print.xtable(line_list_dat,
+             type = "latex",
+             file = "../results/mol-data/line-list-data.tex",
+             sanitize.text.function = identity,
+             include.colnames = F,
+             include.rownames = F,
+             only.contents = T,
+             hline.after = NULL)
+```
+
+### Number of positive samples
 
@@
   mutate_all(round, 2)
+
+# estimated effect of air cleaners
+spread_draws(multLogitMod, beta1) %>%
+  mutate(beta1 = exp(beta1)) %>%
+  mean_qi()
 ```

### 2023-06-20 preprocessing/prep_molecular.R

@@ -16,6 +16,17 @@ df <- read_xlsx("data-raw/molecular/ZF Resultate Airchecker 2023.xlsx", sheet = 
   mutate(class = ifelse(class == "3a", "B", "A"),
          date = as.Date(gsub("2022", "2023", as.character(date))))
 
+# line list
+
+long_list <- df %>%
+  mutate(result = ifelse(virus == "-", "Negative", "Positive"),
+         virus = ifelse(virus == "SARS-CoV-2", "CoV", virus)) %>%
+  dplyr::select(class, date, result, virus) %>%
+  arrange(class, virus) %>%
+  ungroup()
+
+saveRDS(long_list, file = "data-clean/molecular-long.rds")
+
 
 #### Preprocessing ####
 

### 2023-06-20 analysis/epi_analysis.Rmd
minor update line list data

@@
   dplyr::select(class, date_symptom, date_absent, date_back) %>%
+  arrange(class, date_absent) %>%
   as.matrix() %>%

### 2023-06-20 doc/paper/manuscript.tex

\PassOptionsToPackage{top=3cm,left=3cm,right=3cm,bottom=3cm}{geometry}
\documentclass[fleqn,11pt]{wlscirep}

\usepackage{import}
\usepackage{main}

\renewcommand{\paragraph}[1]{\vspace{0.3cm}\noindent\underline{\emph{#1}}\hfill\noindent}

% word count
\newcommand{\maincount}[1]{%
  \immediate\write18{texcount -1 -sum=1 -merge -q -nobib #1.tex > #1-words.sum}%
  \input{#1-words.sum}%
}

\newcommand{\abstractcount}[1]{%
  \immediate\write18{texcount -template="{abst}" #1.tex > #1-words.sum}%
  \input{#1-words.sum}%
}

\begin{document}

\doublespacing

\title{\bfseries\LARGE\singlespacing{Transmission of airborne respiratory infections with and without air cleaners in a Swiss school during non-pandemic conditions: A modeling study of epidemiological, environmental, and molecular data}}

% Alternative title: SARS-CoV-2 transmission in schools based on molecular and epidemiological data: potential effects of mask wearing and air cleaners

% author list
\author[1$\ddag$]{Nicolas Banholzer}
\author[2]{Philipp Jent}
\author[3]{Pascal Bittel}
\author[3]{Lavinia Furrer}
\author[1]{Matthias Egger}
\author[4]{Tina Hascher}
\author[1*]{Lukas Fenner}

\affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
\affil[2]{Department of Infectious Diseases, Inselspital, Bern University Hospital, University of Bern, Bern, Switzerland}
\affil[3]{Institute of Infectious Diseases, University of Bern, Bern, Switzerland}
\affil[4]{Institute of Educational Science, University of Bern, Bern, Switzerland}

\affil[*]{Corresponding author: lukas.fenner@ispm.unibe.ch }

\vspace{1em}

\begin{information}\normalfont
\noindent\textbf{Running head}: SARS-COV-2 transmission in schools and effect of air cleaners

%\noindent\textbf{Subject categorization}: 6.20  Indoor Air; 10.11 Pediatrics: Respiratory Infections

\noindent\textbf{Word count}: \maincount{manuscript}words, abstract \abstractcount{manuscript}words (max. 500), title 157 chars (max. 200)

%\noindent\textbf{Inserts:} 2 tables, 6 figures, 44 references

\noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.

\vspace{1em}

\noindent\textbf{Funding}

\noindent This study is funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland. NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05. ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498). \medskip

\noindent\textbf{Contributions}

\noindent Conception and design: NB, LF. Epidemiological and environmental data collection: NB, PJ, TS, LF. Laboratory data collection: PB, LFu. Additional data collection: TH. Statistical analysis: NB, KZ. Genomic analysis: LB, LFu. Paper draft: NB, LF, ME. All authors reviewed and approved the final version of the manuscript.

\par
\end{information}

%TC:newcounter abst Words in abstract
%TC:envir abstract [] abst
\begin{abstract}\normalfont
\noindent\textbf{Background:} Airborne respiratory infections can be transmitted in crowded indoor settings, including schools. Infection control measures such as air ventilation or filtration can reduce the risk of airborne transmission. The aim of this study was to evaluate the effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory infections in schools using a multiple-measurement approach. \medskip

\noindent\textbf{Methods and Findings:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), and molecular data (bioaerosol and saliva samples) over seven weeks from 16 January to 11 March 2023 in a secondary school (n=38 students across two classes) in Switzerland. We performed laboratory and molecular analyses to detect a panel of 19 respiratory viruses. We installed air cleaners in a cross-over design and compared positive samples and particle concentrations between weeks with and without air filters. We modeled the number of new infections using a Bayesian negative binomial regression model, adjusting for the incubation period, the number of recent and cumulative infections, the ventilation rate, the number of students in class, and community transmission. \\
Molecular analysis detected multiple viruses in saliva samples throughout the study (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus), but rarely in bioaerosol samples (2/105 positive: 1~rhinovirus, 1~adenovirus). The detection rate of positive samples was comparable between study conditions (adjusted risk ratio 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61). Daily average CO$_2$ levels were 1,769$\pm$391\,ppm ($\pm$ standard deviation) with air cleaners and 1,636$\pm$341\,ppm without air cleaners, corresponding to air change rates of 0.74$\pm$0.41\,h$^{-1}$ and 0.68$\pm$0.32\,h$^{-1}$, respectively. Daily average aerosol number concentrations were 27$\pm$34\,1/cm$^3$ with and 95$\pm$81\,1/cm$^3$ without air cleaners, corresponding to an estimated adjusted decrease of 77\% (95\%-CrI 63\% to 86\%). Absences due to respiratory infections were less frequent with air cleaners (13~absences) than without them (21~absences). The estimated risk of infection only tended to be lower with air cleaners (adjusted risk ratio 0.72, 95\%-Credible Interval 0.44 to 1.15). \\
Study limitations include unknown specific causes of absences related to respiratory infections (i.e which pathogen caused the sickness). Furthermore, students occasionally had lessons outside the classroom, thus limiting the potential effectiveness of air cleaners.
\medskip %\vspace{-1.3em}

\noindent\textbf{Conclusions:} Molecular detection showed sustained transmission of multiple respiratory viruses over the study. Our results are mixed and, at most, suggest a small benefit of air cleaners in reducing the transmission of airborne respiratory infections in school rooms. Our multiple-measurement approach could be used to continuously monitor the risk of transmission of respiratory infections and to identify the most appropriate infection control measures.

\par
\end{abstract}

%TC:ignore

\flushbottom
\maketitle
\setcounter{page}{1}
\thispagestyle{fancy}

\vspace{2em}

\noindent\textbf{Word count:} \abstractcount{manuscript}words (max. 500)

\vspace{0.5em}

\noindent\textbf{Keywords:} schools, air cleaner, respiratory viruses, molecular detection
% maximum of 3-5 keywords
\newpage

\sloppy
\raggedbottom
%TC:endignore

\newpage

%TC:break main
\section{Introduction}

The spread of respiratory infections such as SARS-CoV-2 and influenza is difficult to control\cite{Leung2021}. Person-to-person transmission mainly occurs through exhaled respiratory particles containing the respective pathogen, particularly via aerosols (more recently defined as respiratory particles <100$\mu$m\cite{Fennelly2020,Wang2021}), rather than larger respiratory droplets. Multiple reports have provided evidence that a considerable part of SARS-CoV-2 transmission is likely to happen through small respiratory particles (<5$\mu$m, also called fine aerosols), allowing for longer suspension times and airborne transmission at short (1-2 m) and at long ranges (>2m)\cite{Coleman2022,Wang2020,Heneghan2021}. Growing evidence suggests they contribute importantly to the overall spread of SARS-CoV-2 in indoor congregate settings such as clinics, workplaces, and schools\cite{Greenhalgh2021,Tellier2022,Heneghan2021,Wang2021}.

Public authorities worldwide closed businesses and schools during the COVID-19 pandemic\cite{Banholzer2022EJE,Banholzer2021PLOS}, but the closure of schools was particularly contentious. While the negative impact of school closures on student well-being and mental health is well documented\cite{Loades2020}, the role of children and adolescents in transmitting SARS-CoV-2 is less clear\cite{Goldstein2021}. A study in Germany estimated that school contacts contributed between 2\% and 20\% to the overall transmission of SARS-CoV-2 in the population\cite{Heinsohn2022}. Studies on the effectiveness of government interventions at the population level are inconclusive regarding the effects of school closures in different epidemic waves\cite{Banholzer2021PLOS,Sharma2021}. The introduction of compulsory face mask wearing\cite{Gettings2021,Heinsohn2022} and improved ventilation\cite{Gettings2021} in schools was associated with a lower incidence of COVID-19. In addition, the installation of portable HEPA-air filtration devices (air cleaners) was shown to remove SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}. Finally, a simulation study of exhaled SARS-CoV-2 bioaerosols in an indoor space demonstrated the efficacy of mask wearing and portable air cleaners in reducing aerosols\cite{Lindsley2021}.

We used a multiple-measurement approach to study the transmission of SARS-CoV-2 and other respiratory viruses in school rooms. We collected epidemiological (cases of respiratory diseases), environmental (CO$_2$, aerosol and particle concentrations), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) over a seven-week study period from January to March 2022 in two secondary schools in Switzerland.  We analyzed changes in environmental and molecular characteristics and estimated the probability of infection with SARS-CoV-2 during three different study conditions with and without infection control measures (general mask wearing and air cleaners).

\newpage

\section{Methods}

\subsection{Study setting and design}

\noindent We collected data in two classrooms of a secondary school (age of students 13-15~years, 700 students in total) in the Canton of Solothurn, Switzerland, over a seven-week study period from 26~January to 11~March 2023. Fig.~\ref{fig:study-setup} shows the schematic study setup and Table~\ref{tab:data} provides an overview of the types of data that were collected in each classroom.

\begin{figure}[!htpb]
    \centering
    \includegraphics{../study_setting.pdf}
    \caption{\textbf{Study setting}. Schematic study setup of the classrooms. One air cleaner was placed in the front and the other in the back of classrooms. All devices were placed at the head level of students when they were seated. Both classrooms were not equipped with an active HVAC (Heating, Ventilation, Air conditioning) system, but were ventilated using passive window ventilation. }
    \label{fig:study-setup}
\end{figure}

\begin{table}[!htpb]
    \footnotesize
    \centering
    \caption{\textbf{Data}. Type of data collected, method/device, and frequency in the rooms of classes A and B.}
    \begin{tabular}{p{3.5cm}p{6cm} p{1cm} p{1cm} p{3cm}}
    \midrule
    Type of data & Method / device & Class A & Class B & Frequency \\
    \midrule
    \textbf{Molecular data} \\
    \midrule
    Bioaerosol sampling & Bioaerosol sampling devices: BioSpot-VIVAS (Aerosol Devices Inc., Ft. Collins, Colorado, United States) and Coriolis Micro Air (Bertin Instruments Montigny-le/Bretonneux, France) & x & (x) \newline only Coriolis & daily \\
    Saliva samples & Saliva samples & x & x & twice per week \\
    Swabs from air cleaners & Swab samples from Xiaomi HEPA filters (Xiaomi Mi Air Pro 70m2, Shenzhen, China) & x & x & after each intervention phase (both classes) and before the vacation (only class B) \\ 
    \midrule
    \textbf{Environmental data} \\
    \midrule
    CO$_2$, aerosol/particle concentrations, humidity, temperature & Air quality device: AQ Guard (Palas GmbH, Karlsruhe, Germany) & x & x & daily by minute \\
    Coughs & Sounds collected via microphones and coughs detected with a Machine Learning model \cite{Bertschinger2023CBMS} & x & x & daily by seconds \\
    \midrule
    \textbf{Epidemiological data} \\
    \midrule
    Absences & Survey & x & x & daily \\
    Student characteristics & Survey & x & x & at study start \\
    \bottomrule
    \end{tabular}
    \label{tab:data}
\end{table}

\subsection{Study intervention}

\noindent We used a cross-over design to study the effectiveness of air cleaners, thereby distinguishing two study conditions in each class: weeks with and weeks without installation of air cleaners in the classroom (Fig.~\ref{tab:study-design}). Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China; approx. USD\,250 per device), running at 2$\,\times\,$600 m$^{3}$/h clean air delivery rate. Passive window ventilation occurred per recommendations by the national public health authorities during both study conditions.

\begin{table}[!htpb]
    \footnotesize
    \centering
    \caption{\textbf{Study design.} Cross-over design where two portable air cleaners (AC) were installed during different time periods in classes A and B over a seven-week study period from 16~January to 11~March 2023, excluding a week of vacation between February 6 and 11.}
    \begin{tabular}{l l l l l l l l l}
    \toprule
      & \textbf{Week 1} & \textbf{Week 2} & \textbf{Week 3} & Vacation & \textbf{Week 4} & \textbf{Week 5} & \textbf{Week 6} & \textbf{Week 7} \\
      & Jan 16-21 & Jan 23-28 & Jan 30-Feb 3 & Feb 6-11 & Feb 13-18 & Feb 20-25 & Feb 27-March 3 & March 6-11 \\
      \midrule
      \textbf{A} & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC & \cellcolor{gray!10} None & & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC \\
      \textbf{B} & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC & & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC & \cellcolor{gray!10} None & \cellcolor{gray!10} None \\
      \bottomrule
    \end{tabular}
    \label{tab:study_design}
\end{table}

\subsection{Data collection}

\noindent\textbf{Epidemiological data} \smallskip

\noindent At baseline, we collected aggregated data on age, sex and COVID-19 vaccination and recovery status in the participating classes. Daily we collected data on each absent student, \ie the absence, symptom onset and return date as well as the reason for the absence (sickness, career days,  or other) and symptoms. We also collected data on laboratory tests but none of the students reported a test result over the study. Based on the epidemiological data, we defined a case of respiratory infection as an absence where the student reported a sickness with at least one of the following symptoms: fever, coughing, tiredness, loss of test or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, stomach. Students were asked to report the first date when they experienced symptoms, \eg a student absent on Monday may report Saturday or Sunday as the symptom onset date. We will denote cases by their symptom onset date. Note that, during the week, the symptom onset date will usually correspond to the absence date, unless the student went to school despite experiencing symptoms. The line list data is shown in \supp~\zref{sec:case-data}. Informed consent of one student could not be obtained and his/her one absence was considered as unrelated to a respiratory infection. Reports about absences were entered electronically into a REDcap database\cite{Harris2009,Harris2019}. Furthermore, both classes participated in repetitive bi-weekly (Mondays and Wednesdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below). The saliva samples were transported to the laboratory and stored at $-$80°C until further processed\cite{Galar2021,To2019,Huber2021}. \medskip

\noindent\textbf{Environmental data} \smallskip

\noindent \emph{CO$_2$ and particle measurements}

\noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute. A transient mass balance model was used to estimate the air change rate from indoor CO$_2$ levels\cite{Batterman2017IJERPH}, assuming that the classrooms were fully occupied during lessons and that room occupancy increased (decreased) linearly 10~minutes before (after) the start (end) of a class. Particle concentrations were filtered according to the times students were in the classroom, which were monitored together with the time windows were opened. \medskip

\noindent \emph{Bioaerosol sampling}

\noindent We collected airborne respiratory viruses in the classroom with a bioaerosol sampling device (BioSpot-VIVAS, Aerosol Devices Inc., Ft. Collins, CO, USA). This device samples airborne virus particles into a viral transport medium (VTM) using a laminar flow water-based condensation method. In parallel, we also used the Coriolis Micro Air (Bertin Instruments Montigny-le-Bretonneux, France) sampler, running at 200\,l/min and collecting into 15\,mL PBS  as previously reported in clinic settings\cite{Moore2021}. BioSpot-VIVAS operated throughout lessons while Coriolis Micro Air only operated shortly before and during break times to reduce noise exposure (approximately 60\,min/day). The removable parts of both sampling devices were regularly autoclaved to avoid contamination. At the end of the day, samples were transported to the Institute of Infectious Diseases (IFIK) and stored at $-$80°C. At the end of the study period, the Xiaomi HEPA filters were carefully removed and 20\,swabs were taken from each filter and stored at $-$80°C until further processed. \medskip

\subsection{Laboratory and molecular analyses}\label{sec:mol_analyses}

\noindent Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL. The human saliva samples were directly analyzed without prior filtration. The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19~respiratory viruses: SARS-CoV-2, Influenza A virus, Influenza B virus, Human respiratory syncytial virus A/B (RSV), Human metapneumovirus (MPV), Human adenovirus A/B/C/D/E/F (AdV), Human rhinovirus A/B/C (HRV), and Human parainfluenza virus 1/2/3/4 (PIV) in one single reaction. Viral genome load (VGL) of specimens was quantified (copies/L) by curve extrapolation derived from standardized ATCC quantitative genomic standard dilutions (ATCC-VR-3347D, LGC Standards, LGC Group UK). For the BioSpot-VIVAS air samples, the air concentration of the DNA was exactly calculated according to the airflow influx into the sampling device. The Limit of detection (LOD) of the assay SARS-CoV-2, IFA/IFB is 7.5x10$^0$ TCID/ml (SARS-CoV-2), 3.7x10$^{-1}$ TCID/ml (IFA), and 4.8x100 TCID/ml (100cp/ml) (IFB) respectively. For positive samples, we performed targeted sequencing to compare genetic relatedness between SARS-CoV-2 strains detected in the air and human samples\cite{Goncalves2021}, but we were unable to amplify and sequence any of the gene targets in the bioaerosol samples due to low RNA concentrations.

\subsection{Statistical analyses and modeling}

All statistical analyses were described in the statistical analysis plan (SAP)\cite{Banholzer2023SAP}, which was published before analyses began. In contrast to the SAP, the number of positive bioaerosol samples and the viral load concentration of positive samples could not be analyzed because there were too few positive samples over the study. Apart from that there were no major deviations from the SAP regarding the conducted analyses. There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in \supp~\zref{sec:transmission-model} to \supp~\zref{sec:env-regression-model}, which provide detailed descriptions of the models used in each analysis.

\noindent\textbf{Risk of infection} \smallskip

\noindent We estimated the relative risk of infection with air cleaners using a Bayesian latent hierarchical regression model. The observed outcome of this model is the number of new respiratory cases $C$ (absences related to respiratory infections by date of symptom onset) on day $t$ in class $j$, which are modeled with a Negative Binomial distribution. The expected number of new cases is the weighted sum of the number of new infections $I_{js}$ (unobserved/latent outcome) in the previous days $s<t$, with the weights corresponding to the probability distribution of the incubation period. The number of new infections is related to the presence of air cleaners using a regression-like log-link
\begin{align}
    \log I_{js} = \log F_{js} - \log N_{js} + \beta_0 + \beta_1 \cdot \text{AirCleaner}_{js},
\end{align}
where $F_{js}$ is the number of infections in the previous week (a proxy for the number of infectious students), $N_{js}$ is the cumulative number of infections, $\beta_0$ is the infection rate without air cleaners, and $\beta_1$ is the effect of air cleaners. Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the community (\ie the Canton of Solothurn).

A detailed description of the model and choice of priors for all model parameters are provided in \supp~\zref{sec:transmission-model}. Importantly, since multiple pathogens were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the pathogen-specific distributions. The weights were inferred based on the proportion of positive saliva samples for each pathogen per week. As a result, the distribution of the incubation period used for modeling could change every week depending on the proportion of pathogens that were detected in the students' saliva. Moreover, respiratory cases by date of symptom onset were under-reported for weekends and thus a re-weighting scheme was used to account for weekday effects. \medskip

\noindent\textbf{Number of positive saliva samples} \smallskip

\noindent Molecular analysis determined which saliva samples were positive for a respiratory pathogen. We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~\zref{sec:multinomial-model}) where the expected number of positive samples for a pathogen was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each pathogen. The latter was computed as the difference between the number of students per class and the cumulative number of positive samples. As a sensitivity analysis, we lagged the test results by one test date (\ie from Wednesday to Monday and from Monday to last Wednesday) and one test week (\ie from Monday to last Monday and Wednesday to last Wednesday), to consider the unknown delay when students were infected and tested positive.

\noindent\textbf{Aerosol and particle concentrations} \smallskip

\noindent We computed the average concentrations per day and compared these between study conditions. We estimated the reduction in concentrations using Bayesian log-linear regression models (\supp~\zref{sec:env-regression-model}), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases. \medskip

\noindent\textbf{Software and model estimation} \smallskip

\noindent All analyses were performed in R software (version 4.2.0)\cite{RCoreTeam2022} and modeling in Stan (version 2.21.0)\cite{Carpenter2017}. Model parameters were estimated with a Bayesian approach. Specifically, Markov chain Monte Carlo (MCMC) sampling was used as implemented by the Hamiltonian Monte Carlo algorithm with the No-U-Turn Sampler (NUTS)\cite{Hoffman2014}. Each model is estimated with 4 Markov chains and 2,000 iterations of which the first 1,000 iterations are discarded as part of the warm-up. Estimation power is evaluated via the effective sample size ESS and convergence of the Markov chains is evaluated with the Gelman-Rubin convergence diagnostic ($\hat{R}$). If not stated otherwise, we report posterior means and credible intervals (CrIs) based on the 50\%, 80\%, and 95\% quantiles of the posterior samples, respectively. The code is available from \url{https://osf.io/38j9g}.

% \subsection{Additional data collection}

% \noindent We also collected data about the students' subjective experiences of well-being and emotions using standardized daily online questionnaires and a pre-post questionnaire\cite{Brandenberger2018}. These data are not part of the current analysis.

\subsection{Ethics statement}

\noindent The Ethics Committee of the Canton of Bern, Switzerland, approved the study (reference no. 2021-02377). For the saliva samples, we included all students who were willing to participate and obtained written informed consent from their caregivers.

\newpage

\section{Results}

The study population consisted of 90~students (39~female, 51~male; Table~\ref{tab:cases-overview-school}). Of these, 27~students were fully vaccinated and 34~students had recovered from an infection within the last year. Over the seven-week study period (3,150 student-days in total), students were absent from school for 644~days (20\% of total) of which 147~days (23\% of absences) were due to isolation related to COVID-19, and 247~(38\% of absences) were due to sickness. Overall, there were 35~confirmed and 73~suspected cases of COVID-19, exceeding the number of students in some classes (Table~\zref{tab:cases-overview}). This suggests that only a proportion of suspected cases were actual cases of COVID-19 as it is unlikely that students were infected twice. We estimated 55 (95\%-CrI 35 to 76) cases of COVID-19 across schools. \deleted{(\supp~Fig.~\zref{fig:redcap-supp-total-cases}).}

\begin{table}[!htpb]
    \centering
    \caption{Overview of the study population, number of COVID-19 cases, and person-days of absences.}
    \label{tab:cases-overview-school}
    \footnotesize
    \begin{tabular}{l r r r}
    \toprule
         &  School 1 & School 2 & Total \\ \midrule 
        \textbf{Students} & \textbf{52 (58\%)} & \textbf{38 (42\%)} & \textbf{90 (100\%)} \\
        \emph{Gender} \\
        $\drsh$ Female & 20 (38\%) & 19 (50\%) & 39 (\hphantom{0}43\%) \\
        $\drsh$ Male & 32 (62\%) & 19 (50\%) & 51 (\hphantom{0}57\%) \\
        \emph{Vaccination status} \\
        $\drsh$ Vaccinated & 16 (31\%) & 11 (29\%) & 27 (\hphantom{0}30\%) \\
        $\drsh$ Not vaccinated & 36 (69\%) & 27 (71\%) & 63 (\hphantom{0}70\%) \\
        \emph{Recovery status} \\
        $\drsh$ Recovered last year & 25 (48\%) & 9 (24\%) & 34 (\hphantom{0}38\%) \\
        $\drsh$ Not recovered & 27 (52\%) & 29 (76\%) & 56 (\hphantom{0}62\%) \\
        & \\
        \textbf{Absent person-days} & \textbf{334 (52\%)} & \textbf{310 (48\%)} & \textbf{644 (100\%)} \\
        $\drsh$ Isolation & 109 (33\%) & 38 (12\%) & 147 (\hphantom{0}23\%) \\
        $\drsh$ Sickness & 95 (28\%) & 152 (49\%) & 247 (\hphantom{0}38\%) \\
        $\drsh$ Quarantine & 55 (17\%) & 5 (\hphantom{0}2\%) & 60 (\hphantom{00}9\%) \\
        $\drsh$ Other & 75 (22\%) & 115 (37\%) & 190 (\hphantom{0}30\%) \\
        & \\
        \textbf{COVID-19 cases} & \textbf{78 (19\%)} & \textbf{30 (13\%)} &  \textbf{108 (100\%)} \\
        $\drsh$ Confirmed & 25 (32\%) & 10 (\hphantom{0}33\%) & 35 (\hphantom{0}32\%) \\
        $\drsh$ Suspected & 53 (68\%) & 20 (\hphantom{0}67\%) & 73 (\hphantom{0}68\%) \\
        \bottomrule
    \end{tabular} 
\end{table}

\subsection{Molecular analyses}

We analyzed 262 saliva, 130 bioaerosol samples and swabs from the filters of air cleaners (20~swabs per filter) in two classrooms. We detected SARS-CoV-2, adenovirus, and influenza virus (Table~\ref{tab:molecular-data} and \supp~Table~\zref{tab:mol-data-1}-\zref{tab:mol-data-2}). SARS-CoV-2 made up the vast majority of positive saliva (19 out of 21) and bioaerosol samples (9 out of 10). We found four positive air-saliva samples in the same respective week (three SARS-CoV-2 and one adenovirus), suggesting they were paired samples. We also detected SARS-CoV-2 and influenza viruses from the HEPA filters of the air cleaners. The number of positive saliva and airborne SARS-CoV-2 samples per week varied by study condition (Fig.~\ref{fig:molecular-descriptives}a). Without interventions, the proportion of positive samples per week was 11.5\% for saliva and 8.1\% for airborne samples. These proportions were lower with mask mandates (Saliva: 5.7\%, Air: 7.1\%) and air cleaners (Saliva: 7.7\%, Air: 5.0\%). There were also differences in viral concentration in positive airborne samples (Fig.~\ref{fig:molecular-descriptives}b). Without interventions, it was 1.1\,copies per liter per week, which was higher than with mask mandates (0.7\,copies/L per week) and air cleaners (0.1\,copies/L per week).

\begin{table}[!htpb]
    \centering
    \caption{\textbf{Analysis of molecular data}. Number of positive and negative saliva and airborne samples in each school.}
    \label{tab:molecular-data}
    \footnotesize
    \begin{tabular}{l rlrl p{.2cm} rlrl p{.2cm} rlrl}
    \toprule
    & \multicolumn{4}{c}{Saliva} & & \multicolumn{4}{c}{Air} & & \multicolumn{4}{c}{Air cleaner (HEPA filter)$^\dag$} \\ \cline{2-5} \cline{7-10} \cline{12-15} 
    & \multicolumn{2}{c}{School 1} & \multicolumn{2}{c}{School 2} & &  \multicolumn{2}{c}{School 1} & \multicolumn{2}{c}{School 2} & & \multicolumn{2}{c}{School 1} & \multicolumn{2}{c}{School 2} \\
    \midrule
    \textbf{Total} & \textbf{173} & & \textbf{89} & & & \textbf{68} & & \textbf{62} & & &  & &  & \\
    Negative & 160 & (\hphantom{0}92\%) & 81 & (\hphantom{0}91\%) & & 64 & (\hphantom{0}94\%) & 56 & (\hphantom{0}90\%) & &  & &  & \\  
    Positive & 13 & (\hphantom{00}8\%) & 8 & (\hphantom{00}9\%) & & 4 & (\hphantom{00}6\%) & 6 & (\hphantom{0}10\%) & & & &  & \\
    \midrule
    \textbf{Positive} & \textbf{13} & & \textbf{8} &  & & \textbf{4} & & \textbf{6} & & & \multicolumn{2}{l}{\textbf{2}}  & \multicolumn{2}{l}{\textbf{6}} \\
    SARS-CoV-2 & 12 & (\hphantom{0}92\%) & 7 & (\hphantom{0}88\%) & & 4 & (100\%) & 5 & (\hphantom{0}83\%) & & 2 & (100\%) & 4 & (\hphantom{0}66\%) \\
    Influenza & 1 & (\hphantom{00}8\%) & 0 & (\hphantom{00}0\%) & & 0 & (\hphantom{00}0\%) & 0 & (\hphantom{00}0\%) & & 0 & (\hphantom{00}0\%) & 1 & (\hphantom{0}17\%) \\
    Adeno & 0 & (\hphantom{00}0\%) & 1 & (\hphantom{0}12\%) & & 0 & (\hphantom{00}0\%) & 1 & (\hphantom{0}17\%) & & 0 & (\hphantom{00}0\%) & 1 & (\hphantom{0}17\%) \\
    \bottomrule
    \multicolumn{15}{l}{\scriptsize \emph{$\dag$: 20 swabs were taken from each filter at the end of the study period.}}
    \end{tabular}
\end{table}

\begin{figure}[!htpb]
    \centering
    \includegraphics{figures/molecular-descriptives.pdf}
    \caption{\textbf{Analysis of molecular data and comparison by study condition}. \textbf{(a)}~Proportion of positive airborne (solid line) and saliva (dashed line) samples (average per week). \textbf{(b)}~Viral concentration in positive airborne samples from BioSpot-VIVAS (average per week).}
    \label{fig:molecular-descriptives}
\end{figure}

\subsection{Analysis of aerosol and particle concentrations}

Particle concentrations differed by study condition (Fig.~\ref{fig:palas-results}a). Aerosol number concentrations were lower with mask mandates (49$\pm$52\,1/cm$^3$) and air cleaners (84$\pm$56\,1/cm$^3$) than without interventions (177$\pm$109\,1/cm$^3$). Similarly, particle mass concentrations (\eg PM$_1$) were lower with mask mandates (2.0$\pm$1.6\,$\mu$gm$^{-3}$) and air cleaners (3.8$\pm$2.9\,$\mu$gm$^{-3}$) than without interventions (6.9$\pm$4.1\,$\mu$gm$^{-3}$). CO$_2$ levels without interventions (953$\pm$198\,ppm) were slightly lower than with mask mandates (1,155$\pm$237\,ppm) and air cleaners (1,088$\pm$224\,ppm), indicating increased ventilation through outdoor air exchange (\supp~Fig.~\zref{fig:palas-supp-descriptives}), although the proportion of times windows were opened did not change with mask mandates (0.03, 95\%-CrI $-$0.07 to 0.12) or air cleaners (0.00, 95\%-CrI $-$0.11 to 0.11). When adjusting for different ventilation rates, the number of students in class, school and weekday effects, the aerosol number concentration decreased by 69\% (95\%-CrI 42\% to 86\%) with mask mandates and by 39\% (95\%-CrI 4\% to 69\%) with air cleaners (Fig.~\ref{fig:palas-results}b and Table~\zref{tab:palas-est-results}). The concentration of smaller particles (PM$_1$ and PM$_{2.5}$) was more reduced by mask mandates and the concentration of larger particles (PM$_4$ and PM$_{10}$) was more reduced by air cleaners.

\begin{figure}[!htpb]
\centering
    \includegraphics[width=\linewidth]{figures/palas-descriptives.pdf}
    \includegraphics[width=\linewidth]{figures/palas-results.pdf}
    \caption{\textbf{Analysis of particle concentrations and comparison by study condition}. \textbf{(a)}~Boxplot of the daily average values for aerosol number concentration (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$). Results for CO$_2$ and other environmental variables are shown in \supp~\zref{sec:detailed-palas}. \textbf{(b)}~Estimated reduction in aerosol number and particle mass concentrations with interventions (posterior mean as dot and 50\%-, 80\%- and 95\%-CrI as lines, respectively). }
    \label{fig:palas-results}
\end{figure}

\subsection{Estimating transmission risk of SARS-CoV-2}

The cumulative number of cases of COVID-19 increased considerably in all classes without intervention and the large majority of students has been infected in School~1 by the time air cleaners were installed, except for class~D in School~2 (Fig.~\ref{fig:redcap-results}a).  Based on our Bayesian transmission model, the probability of getting infected was lower with mask mandates than without interventions (adjusted odds ratio 0.19, 95\%-CrI 0.09 to 0.38) and comparable with air cleaners (1.00, 95\%-CrI 0.15 to 6.51). Accordingly, mask mandates were associated with a considerable number of avoided infections (9.98, 95\%-CrI 2.16 to 19.00), but not air cleaners (Fig.~\ref{fig:redcap-results}b). As an additional analysis, we used a modified Wells-Riley equation\cite{Rudnick2003} and assumed that the change in the rate of emitted infectious quanta were proportional to the \replaced{estimated}{measured} reduction \added{mean and 95\%-CrI) in the aerosol number concentration, while other parameters were kept constant under all study conditions (\supp~\zref{sec:transmission-risk}). Accordingly, the daily risk of infection for a six~hour school day was 1.0\% (95\%-CrI 0.4\% to 1.9\%) with mask mandates and 1.9\% (95\%-CrI 1.0\% to 3.0\%) with air cleaners, compared with a 3.1\% risk \added{for the baseline} without interventions (\supp~Fig.~\zref{fig:results-trans-risk}).

\begin{figure}[!htpb]
    \includegraphics[width=\linewidth]{figures/redcap-descriptives.pdf} 
    \includegraphics[width=\linewidth]{figures/redcap-results.pdf}
    \caption{\textbf{Analysis of epidemiological data and estimated transmission by study condition using Bayesian hierarchical transmission model}. \textbf{(a)}~Estimated mean cumulative number of cases for each school class after probabilistic simulation accounting for uncertainty in suspected cases and the date of symptom onset (\supp~\zref{sec:prob-sim}). Dotted red lines indicate the number of students per class. A comparison of the estimated number of new cases after probabilistic simulation with the observed number of new confirmed and suspected cases is shown in \supp~Fig.\zref{fig:redcap-supp-descriptives}. \textbf{(b)}~Estimated number of avoided infections associated with interventions across schools (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) based on Bayesian hierarchical transmission model. Detailed estimation results are shown in \supp~\zref{sec:detailed-redcap}.}
    \label{fig:redcap-results}
\end{figure}

\FloatBarrier

\newpage

\section{Discussion}

We collected epidemiological, environmental, and molecular data to estimate transmission of respiratory infections in schools and assessed the effectiveness of infection control measures. Airborne detection of SARS-CoV-2 documented sustained SARS-CoV-2 transmission. Bioaerosol SARS-CoV-2 concentrations with general mask mandates were lower, and SARS-CoV-2 was detected on the filters of air cleaners. Aerosol number and particle mass concentrations dropped significantly with both interventions. The Bayesian transmission model estimated that mask mandated avoided SARS-CoV-2 infections, but not air cleaners.

Our study demonstrated airborne detection of SARS-CoV-2 in schools. Although sampling and molecular detection of infectious bioaerosols are challenging and there is no agreed standard\cite{Santarpia2020,Morris2022,Stockwell2019}, it provides important evidence on the airborne transmission of respiratory pathogens. So far, viral RNA in airborne samples of SARS-CoV-2 was mainly found in hospitals and healthcare facilities\cite{Shen2021}. A related study in two South African schools detected airborne \emph{Mycobacterium tuberculosis}\cite{Bunyasi2022}. Tuberculosis was the leading cause of death worldwide due to an infectious disease prior to the COVID-19 pandemic. An improved understanding of airborne transmission and the effectiveness of interventions can benefit the control of both infectious diseases\cite{Rieder2021}. Our study provides evidence on airborne SARS-CoV-2 viral transmission and the potential effects of interventions in schools based on airborne and saliva samples from students. In our study, positive samples mostly pertained to SARS-CoV-2, but we occasionally detected other respiratory viruses such as adenovirus and influenza. General public health measures during the study likely suppressed the spread of other respiratory viruses. It also must be noted that the detected viral concentrations were low, as shown by high cycle thresholds (CTs) in the RT-qPCR results. The molecular presence of other viral pathogens cannot be excluded. The low sensitivity of molecular assays to detect airborne pathogens is a common challenge\cite{Morris2022,Stockwell2019}.

An experimental study demonstrated the effectiveness of infection control measures (universal face mask wearing and air cleaners) using simulated exhaled SARS-CoV-2 bioaerosols in a closed indoor space\cite{Lindsley2021}. In contrast, we studied infection control measures in a real-life setting and demonstrated their effectiveness using a multiple-measurement approach, and obtain similar results. Our findings also align with existing evidence from population-level studies showing that the incidence of COVID-19 was lower schools with mask use\cite{Heinsohn2022,Gettings2021} and improved ventilation\cite{Gettings2021}. Similarly, a field study showed that adequate ventilation was associated with reduced incidence of tuberculosis, a strictly airborne disease, in a university building\cite{Du2020}. Altogether, these findings support arguments in favor of multiple intervention strategies to address airborne transmission of respiratory infections in crowded indoor settings\cite{Morawska2021}.

Indoor ventilation is one of the key determinants of airborne transmission\cite{Fennelly2020,Wang2021}, but schools are often poorly ventilated\cite{Santamouris2008,DiGilio2021}. We showed that air cleaners effectively reduce the concentration of both larger and smaller particles, in line with findings about their effectiveness in hospitals\cite{Morris2022} and simulated indoor environments\cite{Lindsley2021}. The detection of viruses on the filters of air cleaners further supports the evidence on the effective removal of bioaerosols. However, it was difficult to estimate changes in transmission following air cleaners because they were installed at the end of the study period when presumably a large proportion of students were already infected with SARS-CoV-2. In addition, physicochemical properties of aerosols, environmental factors and the distance to infectious people determine the survival of airborne particles and the loss of infectivity over time\cite{Wang2021}. Thus, a predominance of close range high particle density aerosol transmission of SARS-CoV-2 in school settings could further explain why air cleaners appeared not effective in preventing transmission. Our study used portable, affordable air cleaners that could be implemented at scale, rather than large cleaners\cite{Duill2021}. Noise exposure, and a lack of acceptance by teachers\cite{Sanguinetti2022} may still prevent their widespread use. Although not specifically assessed, we perceived good acceptance of our air cleaners during the short study period. Nevertheless, investments in professional building ventilation systems should be preferred to air cleaners in the long term\cite{Nardell2016}.

School closures during the COVID-19 pandemic have been intensely debated as children and adolescents are particularly vulnerable to the negative impact of such interventions on their well-being and mental health\cite{Loades2020}. Furthermore, numerous studies have examined the role of children in transmitting SARS-CoV-2\cite{Goldstein2021} and it remains unclear to what extent transmission of SARS-CoV-2 occurs in schools\cite{Ulyte2021Clustering}. These findings contrast with studies of influenza viruses showing that school children may drive the seasonal influenza epidemic. Community studies in the US demonstrated that influenza transmission rates in children and adolescents were high in schools and that they easily transmit influenza viruses to household members and into their communities\cite{Glezen1978,Shang2018,Fiore2012}. Our study suggests that also transmission of SARS-CoV-2 occurs to a considerable extent in schools.

Our study has several limitations. First, aerosol measurements and molecular detection of pathogens in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (human to air, air to human) cannot be determined. Second, a comparison of viral concentration between study conditions should be interpreted with care due to the technical limitations of molecular detection in bioaerosol samples, and because the number of possibly infectious (and thus susceptible) students decreased towards the end of the study. Third, our epidemiological analysis is based on observational data, thus subject to potential confounding, \eg the incidence of COVID-19 (cases per week) in the community varied over the study period. However, levels were high throughout and included two Omicron subvariant peaks. Community transmission was also considered in our Bayesian transmission model. CO$_2$ levels were not considered in the model but the levels were slightly lower without interventions, suggesting that lower ventilation during intervention phases may have actually reduced the estimated effectiveness of mask mandates and air cleaners. Fourth, epidemiological data may not always be complete due to the under-reporting of COVID-19 among absent students. We thus used a probabilistic approach to estimate the proportion of suspected cases being actual cases of COVID-19 and the date of symptom onset where it was not reported. While this allows us to consider uncertainty in the observed data, the estimated effects of interventions will be less precise as reflected in the larger uncertainty intervals.

In conclusion, using epidemiological, environmental, and molecular data, our study shows that considerable transmission of SARS-CoV-2 occurred in the participating schools. General face mask wearing were associated with reduced SARS-CoV-2 transmission and prevented infections. The effectiveness of interventions was supported by significant decreases in the concentration of aerosols. Taken together, our results suggests that infection control measures can reduce the transmission of respiratory infections in school rooms. Future studies may use our multiple-measurement approach to assess the effectiveness of infection control measures in reducing the transmission of respiratory infections. Ideally, these data should be collected routinely in sentinel schools, thus continuously monitoring transmission risks and alerting health authorities when infection control measures should be taken.

\newpage

%TC:ignore
\section*{Acknowledgements}
We would like to thank the schools, teachers, and students participating in the study. We are also grateful to the Educational Department of the Canton of Solothurn for their support throughout the study. Finally, we are indebted to the student assistants who helped with the data collection at the schools.

\bibliography{references.bib}
%TC:endignore

\end{document}

### 2023-06-20 doc/paper/supplements.tex

\PassOptionsToPackage{top=3cm,left=3cm,right=3cm,bottom=3cm}{geometry}
\documentclass[fleqn,11pt]{wlscirep_supp}

\usepackage[]{minitoc}
\mtcsetdepth{secttoc}{3}
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\mtcsettitle{secttoc}{}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
%\usepackage[top=3cm,left=3cm,right=3cm,bottom=3cm]{geometry}% by courtesy of Mico
\usepackage{lmodern}
\usepackage{bbm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{colortbl}
\usepackage{siunitx}
\sisetup{
  detect-all,
  detect-weight=true,
  detect-family=true,
  mode=text,
%   detect-inline-family=math,
  group-separator={,},
%   group-minimum-digits={3}			
}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{tabu}
\usepackage{authblk}
\usepackage{mathtools}
\usepackage{overpic}
\usepackage{url}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{fit}
\usepackage{multirow}
\usepackage{float}
\usepackage[normalem]{ulem}
\usepackage{bm}
\usepackage{enumerate}
\usepackage[absolute,overlay%,showboxes
                        ]{textpos}
% \usepackage{caption}
\usepackage[font=small,labelfont=bf,justification=justified]{caption}
\usepackage{subcaption}
\usepackage{xspace}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{placeins}
\usepackage{makecell, booktabs}
\usepackage{eqparbox}
\usepackage{rotating}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{setspace}
%\usepackage{comment}
\usepackage[resetlabels,labeled]{multibib}
\newcites{Supp}{References}
\usepackage[sort&compress]{cleveref}
\Crefname{appendix}{Supplement}{Supplements}

%\usepackage{mathabx}

% Table formatting packages
\usepackage{dcolumn} % align decimal points in tables
\newcolumntype{d}[1]{D{.}{.}{#1}}
\usepackage{booktabs} 
\usepackage[flushleft]{threeparttable}
\usepackage{siunitx} % align on decimal point in tables
\usepackage{lineno}
\usepackage{etoolbox}

\usepackage{lscape}
\usepackage{longtable}
\usepackage{arydshln}

% math
\usepackage{amsmath,amsfonts,amssymb}
% additional math symbols
\DeclareFontFamily{U}{mathb}{}
\DeclareFontShape{U}{mathb}{m}{n}{
  <-5.5> mathb5
  <5.5-6.5> mathb6
  <6.5-7.5> mathb7
  <7.5-8.5> mathb8
  <8.5-9.5> mathb9
  <9.5-11.5> mathb10
  <11.5-> mathb12
}{}
\DeclareSymbolFont{mathb}{U}{mathb}{m}{n}
\DeclareMathSymbol{\ulsh}{3}{mathb}{"E8}
\DeclareMathSymbol{\ursh}{3}{mathb}{"E9}
\DeclareMathSymbol{\dlsh}{3}{mathb}{"EA}
\DeclareMathSymbol{\drsh}{3}{mathb}{"EB}

%% Patch 'normal' math environments:
\newcommand*\linenomathpatch[1]{%
  \cspreto{#1}{\linenomath}%
  \cspreto{#1*}{\linenomath}%
  \csappto{end#1}{\endlinenomath}%
  \csappto{end#1*}{\endlinenomath}%
}

\linenomathpatch{equation}
\linenomathpatch{gather}
\linenomathpatch{multline}
\linenomathpatch{align}
\linenomathpatch{alignat}
\linenomathpatch{flalign}

\linenumbers

% PLOS formatting
\makeatletter %only needed in preamble
\renewcommand\Large{\@setfontsize\Large{18pt}{18}}
\renewcommand\large{\@setfontsize\large{16pt}{18}}
\makeatother

\addto\captionsenglish{\renewcommand{\figurename}{Fig }}

% \usepackage{xstring}
% \usepackage{etoolbox}
% \usepackage{caption}

% \captionsetup{labelfont=bf,tableposition=top}

% \makeatletter
% \newcommand\formatlabel[1]{%
%     \noexpandarg
%     \IfSubStr{#1}{.}{%
%       \StrBefore{#1}{.}[\firstcaption]%
%       \StrBehind{#1}{.}[\secondcaption]%
%       \textbf{\firstcaption.} \secondcaption}{%
%       #1}%
%       }

% \patchcmd{\@caption}{#3}{\formatlabel{#3}}
% \makeatother

\renewcommand*{\Affilfont}{\normalsize\normalfont}
\renewcommand*{\Authfont}{\normalfont}

% referencing of unnumbered materials and methods
\newcounter{methods}
\renewcommand{\themethods}{Materials and methods}

% Track changes
%\usepackage[markup=underlined]{changes}
\makeatletter
\@namedef{Changes@AuthorColor}{magenta}
\colorlet{Changes@Color}{magenta}
\makeatother

%=====================================================================% Declare

\DeclareSIUnit\eur{\officialeuro}
\DeclareSIUnit\M{M}
\DeclareSIUnit\k{k}

% Widebar symbol
% \DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
% \DeclareFontShape{U}{mathx}{m}{n}{<-> mathx10}{}
% \DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
% \DeclareMathAccent{\widebar}{0}{mathx}{"73}

%=====================================================================% New commands (Macros)

% def
\def\sym#1{\ifmmode^{#1}\else\(^{#1}\)\fi}
\definecolor{darkgreen}{rgb}{0.0, 0.5, 0.0}

% new command
\newcommand{\smallsim}{\smallsym{\mathrel}{\sim}}
\newcommand{\specialcell}[2][c]{%
  \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
\newcommand{\specialcellc}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
\newcommand\ie{i.\,e.\xspace}
\newcommand\eg{e.\,g.\xspace}
\newcommand{\dd}[1][]{\mathrm{d}#1}
\newcommand{\BK}[1]{{\color{orange}{BK: #1}}}
\newcommand{\figletter}[1]{{{\fontfamily{\sfdefault}\selectfont \textbf{#1}}}}
\newcommand\TODO[1]{{\color{red}#1}}  
\newcommand{\FIX}[1]{{\color{darkgreen}#1}}

% renewcommand
\renewcommand\theadfont{\bfseries}
\renewcommand\theadalign{lc}
\renewcommand\cellalign{tl}

\makeatletter

\newbox\@abstract%
\def\abstitle{\textbf{Abstract}}%
\renewenvironment{abstract}{
  \global\setbox\@abstract\vbox\bgroup%
   \noindent
}{%
   \egroup%
}%

\renewcommand*{\Affilfont}{\normalsize\normalfont}
\renewcommand*{\Authfont}{\normalfont}

\addto\captionsenglish{% Replace "english" with the language you use
  \renewcommand{\contentsname}{List of Texts}
}

\def\@maketitle{%
  \newpage
    {\raggedright\fontsize{18pt}{20pt}\selectfont \@title \par}%
    \vskip 0.5em%
    {\large
      \lineskip .5em%
      \begin{tabular}[t]{l}%
        \raggedright \normalsize\mdseries{\@author} %
      \end{tabular}\par}%
      \vskip 1em
%      \raggedright\Large\abstitle\par
%      \vskip 1em
%    {\unvbox\@abstract\par}%
    \par
  \vskip 0.5em
}

\makeatother

%\renewcommand{\thesection}{Text \arabic{section}}
\usepackage{titlesec}
\titleformat{\section}{\normalfont\Large\bfseries}{Text \thesection.~#1}{1em}{}
\renewcommand{\thefigure}{\Alph{figure}}
\renewcommand{\thetable}{\Alph{table}}

\begin{document}
\doublespacing
\nolinenumbers

\newcommand{\supp}{SI Appendix}

\title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
Transmission of airborne respiratory infections with and without air cleaners in a Swiss school during non-pandemic conditions: A modeling study of epidemiological, environmental, and molecular data}}

% author list
\author[1$\ddag$]{Nicolas Banholzer}
\author[1$\ddag$]{Kathrin Z\"urcher}
\author[2]{Philipp Jent}
\author[3]{Pascal Bittel}
\author[3]{Lavinia Furrer}
\author[1]{Matthias Egger}
\author[4]{Tina Hascher}
\author[1*]{Lukas Fenner}

\affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
\affil[2]{Department of Infectious Diseases, Inselspital, Bern University Hospital, University of Bern, Bern, Switzerland}
\affil[3]{Institute of Infectious Diseases, University of Bern, Bern, Switzerland}
\affil[4]{Institute of Educational Science, University of Bern, Bern, Switzerland}

\affil[*]{Corresponding author: lukas.fenner@ispm.unibe.ch }

\affil[$\ddag$]{These authors contributed equally to this work.}

%\begin{abstract}\normalfont
%The supplementary material contains (1)~the detailed method, (2)~the simulation-based study, (3)~further descriptives, and (4)~the results from the sensitivity analysis.
%\end{abstract}

\flushbottom
\maketitle
\thispagestyle{empty}

%\newpage

\sloppy
\raggedbottom

\newpage

\appendix

\tableofcontents

\listoffigures

\listoftables

%\listoffigures
%\listoftables

\newpage

\section{Epidemiological line list data}\label{sec:case-data}

Table~\ref{tab:epi-data-line-list} shows the line list data for respiratory cases based on epidemiological data.

{\footnotesize\begin{longtable}{l l l l}
    \caption[Line list of respiratory cases over the study period]{Line list of respiratory cases over the study period.}\label{tab:epi-data-line-list} \\
    \toprule
    Class & Date of absence & Date of symptom onset & Date of return \\
    \midrule
    \input ../../results/epi-data/line-list-data
    \bottomrule
\end{longtable}}

{\footnotesize\begin{longtable}{l l l l l}
    \caption[Line list of molecular test results over the study period]{Line list of respiratory cases over the study period.}\label{tab:mol-data-line-list} \\
    \toprule
    Class & Date of test & Weekday & Test result & Pathogen \\
    \midrule
    \input ../../results/mol-data/line-list-data
    \bottomrule
\end{longtable}}

\clearpage

\section{Molecular line list data}\label{sec:mol-data}

Table~\ref{tab:mol-data-line-list} shows the line list data for the laboratory test results from human saliva samples.

\clearpage

\section{Modeling relative risk of respiratory infections}\label{sec:transmission-model}

\subsection{Overall approach}

The aim is to estimate transmission of SARS-CoV-2 and the effects of infection control measures (mask mandates and air cleaners) on the number of new infections with COVID-19. To estimate transmission, our model links two unobserved quantities (the number of new infections and susceptible students) with one observed quantity (the number of new cases among students). Specifically, we link the number of new cases to the number of new infections in the previous days. We then formulate the daily proportion of susceptible students getting infected as a function of the interventions in place. The effects of interventions are estimated with a Bayesian approach, which requires the specification of prior distributions for all model parameters.

\subsection{Notation}

\begin{tabular}{ll} 
$j$  & class \\
$t$ & days since start of study period \\
$Y_{j}$  & total number of students in class $j$ (unobserved)  \\
$N_{jt}$  & number of  new cases among students in class $j$ at day $t$ (observed) \\
$I_{jt}$  & number of new infections in class $j$ at day $t$ (unobserved)  \\
$S_{jt}$  & number of susceptibles in class $j$ at day $t$ (unobserved)  \\
\end{tabular}

\subsection{Relating the number of new infections to the number of susceptibles and the presence of interventions}

A proportion of the susceptible students $P_{jt}$ gets infected in class $j = 1,\dots,J$ at each day $t = 1,\dots,T$
\begin{align}
    I_{jt} = S_{jt} P_{jt} ~.
\end{align}
This proportion (or probability) can be estimated using a logit link function
\begin{align}
    \Phi_{jt} &= \log\left(\frac{P_{jt}}{1-P_{jt}}\right) \\
    \textrm{logit}^{-1}(\Phi_{jt}) &= \frac{\exp{(\Phi_{jt})}}{1 + \exp{(\Phi_{jt})}} = P_{jt} ~.
\end{align}

In the absence of interventions, the probability of getting infected is constant, \ie $\Phi_{jt} = \beta_j$. It can only change at times when interventions are in place. Let $\theta_{j}^M$ be the effect of mask mandates and let $\theta_{j}^A$ be the effect of air cleaners in class $j$, then
\begin{align}
    \Phi_{jt} = \beta_j + \theta_{j}^M\,{M}_{jt}
    + \theta_{j}^A\,{A}_{jt},
\end{align}
where ${M}_{jt}$ and ${A}_{jt}$ are binary variables indicating whether mask mandates and air cleaners are in place in class $j$ at day $t$.

\subsection{Relating the number susceptibles to the number of infections}

The daily number of susceptibles $S_{jt}$ are computed as the difference between the total number of students $Y_j$ and the cumulative number of infections $C_{jt}$ in the previous days
\begin{align}
    S_{jt} = Y_j - C_{jt} = Y_j - \sum_{s < t} I_{js}~.
\end{align}
We assume that all students are susceptible at the beginning of the study period, despite the fact that some students were vaccinated or already infected with COVID-19 prior to the beginning of our study. Note that varying susceptibility before study onset can be subsumed in the class-specific transmission rate without interventions $\beta_j$.

\subsection{Relating the number new observed cases to the number of new infections}

The expected number $\mu$ of new cases $N_{jt}$ in country $j$ at day $t$ can be derived from the number of new infections in the previous days as 
\begin{equation}
\mu^{N_{jt}} = \sum_{s < t} I_{js} \cdot p_\text{IN}(t-s) ~, 
\end{equation}
where $p_\text{IN}(t)$ denotes the probability distribution of the incubation period, \ie the probability that a new infected subject reports symptoms at day $t$ after the infection. This distribution is estimated from our data as part of fitting the overall model using an informative prior (Section~\ref{subsec:prior-incubation-period}).

The observed number of new cases are modeled to follow a negative binomial distribution, \ie 
\begin{equation}
N_{jt}  \sim  \textrm{NegBinom}\left(\mu^{N_{jt}}, \sigma^{N_{jt}}\right) \label{eq:N} 
\end{equation}
with mean $\mu^{N_{jt}}$, standard deviation $\sigma^{N_{jt}} = \sqrt{\mu^{N_{jt}} \, \left(1 + \frac{\mu^{N_{jt}}}{\phi^N}\right)}$, and an overdispersion parameter $\phi^N$.

\subsection{Taking into account transmission outside school days}\label{subsec:consider-weekends}

Absences were reported on school days but symptom onset could have occurred on the weekend, \eg a student absent on Monday due to COVID-19 may report Saturday as the day of symptom onset. Some cases may still have gone unreported because the date of absence was the same as the date of symptom onset for the majority of cases (see Fig.~\ref{fig:delay-weights}). Therefore, we distinguish between the transmission rate during and outside schools days, \ie 
\begin{align}
    \beta_j = \alpha_j + \omega, 
\end{align}
where $\alpha_j$ and $\alpha_j + \omega$ are the logit of the probability of getting infected without interventions during and outside school days, respectively. We estimate $\alpha_j$ and $\omega$ from data using informative priors (Section~\ref{subsec:prior-alpha}). We further assume that transmission outside school days is the same regardless of whether they are weekend or vacation days, but we do not incorporate vacation days into the model likelihood. That is, during vacation, the modeling of $N_{jt}$ is ignored but $I_{jt}$ and $S_{jt}$ are still computed.

\subsection{Seeding phase}

A few cases were reported in the first days of the study period, indicating that the corresponding infections could have occurred before the start of the study. Therefore, we seed the number of new infections and cases before the study period by initiating our model 12~days (double the average incubation period) prior to the start of the study. Before this day, we assume there have been no infections and all students are susceptible.

\subsection{Adjusting for absences and community transmission}

The probability of getting infected may be influenced by factors other than interventions. We account for two such factors. First, we adjust our estimates for the daily proportion of students absent from class. We expect this proportion to decrease the probability of getting infected because absent students cannot infect classmates and, if staying home, they are also less likely to get infected themselves (if not already infected). Second, we adjust our estimates for the risk of community transmission, which we proxy with the average of the median effective reproduction number for the canton of Solothurn, Switzerland\cite{Scire2022}. We expect higher community transmission to increase the probability of getting infected. The modeling parameters of these factors are denoted by $\bm{\gamma}$.

\subsection{Choice of priors for the probability of getting infected without interventions}\label{subsec:prior-alpha}

We choose informative priors for the logit of the daily probability of getting infected in the absence of interventions $\beta_j = \alpha_j + \omega$. First, our prior for $\alpha_j$ is derived from the share of susceptibles by the end of the study period $S_{jT} / Y_j$, \ie 
\begin{align}
    S_{jT} / Y_j = (1 - \textrm{logit}^{-1}(\alpha_j))^{T_{\mathrm{school}}},
\end{align}
where $T_{\mathrm{school}}$ is the number of school days (excluding weekends and vacation), and solving for $\alpha_j$
\begin{align}
    \textrm{logit}^{-1}(\alpha_j) = 1 - \exp\left(\frac{\log(S_{jT} / Y_j)}{T_{\mathrm{school}}}\right) ~.
\end{align} 
Note that $S_{jT} / Y_j$ and correspondingly $\textrm{logit}^{-1}(\alpha_j)$ will be different for each of our probabilistically generated datasets. Fig.~\ref{fig:prior-alpha}a shows the daily proportion of susceptibles getting infected based on the distributions for the proportion of suspected cases being actual cases of COVID-19 (see Fig.~\ref{fig:prior-proportion}). The corresponding distribution for $\alpha_j$ is shown in Fig.~\ref{fig:prior-alpha}b. We choose a prior for $\alpha_j \sim \textrm{Student-t}(\nu = 3, \mu = -4.1, \sigma = 1.6)$ with wider tails than the empirical distribution to accommodate the fact that $\alpha_j$ may be smaller or larger depending on the effects of interventions and adjustment variables.

\begin{figure}[!htpb]
    \centering
    \includegraphics{figures/redcap-supp-prior-alpha.pdf}
    \caption[Prior for the probability of getting infected without interventions]{Prior for the probability of getting infected without interventions. \textbf{(a)}~Empirical distribution for the daily proportion of susceptibles getting infected across classes. The dotted blue line indicates the mean. \textbf{(b)}~Empirical distribution for the logit of the daily proportion of susceptibles getting infected. The dotted blue line indicates the empirical mean. The corresponding choice of prior for the probability of getting infected without interventions $\alpha$ is hown in red. }
    \label{fig:prior-alpha}
\end{figure}

Second, we choose a prior for $\omega \sim \textrm{Normal}(\mu = \log\,0.7, \sigma = 0.2)$ that encodes our prior belief that students will have fewer indoor contacts outside school days and thus lower odds of getting infected. In addition, this informative prior also considers potential underreporting of cases on weekends (see Section~\ref{subsec:consider-weekends}).

\clearpage

\subsection{Choice of priors for the effects of interventions}

We consider that the mandates may not have been equally effective in all classes. Therefore, we model the effects of mask mandates as $\theta_{j}^M = \textrm{Normal}(\theta^M, \tau)$, where $\theta^M \sim \textrm{Student-t}(\nu = 7, \mu = 0, \sigma = 2.5)$ is the average effect across classes and $\tau \sim \textrm{Student-t}^{+}(\nu = 5, \mu = 0, \sigma = 1)$ is the variation in the effect between classes.

Air cleaners only applied to three classes and were installed towards the end of the study period when most students may have already been infected. Therefore, we do not estimate variation in the effect between classes but only the average effect. In addition, we choose a prior leaning towards smaller effects, \ie $\theta^A \sim \textrm{Student-t}(\nu = 7, \mu = 0, \sigma = 1)$.

\subsection{Choice of priors for the incubation period}\label{subsec:prior-incubation-period}

We estimate the distribution of the incubation period from our data as part of fitting the overall model. Our prior choices for the parameters of this distribution are informed by a meta-analysis\cite{McAloon2020}, which reported a Lognormal distribution with $\log \mu \sim \textrm{Normal}(1.63, 0.06)$ and $\log \sigma \sim \textrm{Normal}^{\textrm{+}}(0.50, 0.05)$. The prior for $p_\text{IN}$ is shown in Fig.~\ref{fig:prior-pin}. Note that $p_\text{IN}$ is discretized via $p_\text{IN}(s) = \int_0^{0.5} p_\text{IN}(\tau) \;\text{d}\tau$ for $s = 0$ and $p_\text{IN}(s) = \int_{s-0.5}^{s+0.5} p_\text{IN}(\tau) \;\text{d}\tau$ for $s > 0$, where $p_\text{IN}(\tau) \sim \textrm{Lognormal}(\mu, \sigma)$ is the density of the lognormal distribution.

\begin{figure}[!htpb]
    \centering
    \includegraphics{../../results/epi-data/incubation-periods.pdf}
    \caption[Choices of prior for incubation period]{Choices of prior for incubation period. \textbf{(a)}~Prior for parameter $\log \mu$. \textbf{(a)}~Prior for parameter $\log \sigma$. \textbf{(c)}~Resulting prior $p_\text{IN}$ (posterior mean as line and 50\%, 80\%, and 95\%-quantile as shaded area).}
    \label{fig:prior-pin}
\end{figure}

\subsection{Choice of prior distributions (summary)}

Table~\ref{tbl:prior_choices} provides an overview of the model parameters together with the prior choices. If not tailored to the specifics of the model, our choices are  informed by recommendations on the choice of priors from the Stan Development Team \cite{Stan2020Priors}.

\begin{table}
\centering
\footnotesize
\begin{tabular}{p{7cm}p{1.5cm}p{6.5cm}}
\toprule
Parameter & Notation & (Hyper-)Prior  \\ 
\midrule
Effects of mask mandates & $\theta_{j}^M$ &  $\textrm{Normal}(\theta^M, \tau)$\\
& $\theta^M$ & $\textrm{Student-t}(\nu = 7, \mu = 0, \sigma = 2.5)$\\
& $\tau$ & $\textrm{Student-t}^{+}(\nu = 5, \mu = 0, \sigma = 1)$ \\
Effects of air cleaners & $\theta^A$ & $\textrm{Student-t}(\nu = 7, \mu = 0, \sigma = 1)$\\
Prob. of getting infected w/o interventions (logit) & $\beta_j$ & $\beta_j = \alpha_j + \omega$ \\
Prob. of getting infected during school (logit) & $\alpha_j$ & $\textrm{Student-t}(\nu = 3, \mu = -4.1, \sigma = 1.6)$ \\
Weekend effect & $\omega$ & $\textrm{Normal}(\mu = \log\,0.7, \sigma = 0.2)$ \\
Overdispersion & $\phi^N$ & $\phi^N = \left(\frac{1}{\xi^N}\right)^2$ \\
 & $\xi^N$ & $\textrm{Normal}^{+}(\mu = 0, \sigma = 1)$ \\
Time from infection to new case & $p_\text{IN}$ & $\textrm{Lognormal}(\log \mu, \log \sigma)$ \\
& $\mu$ & $\textrm{Normal}(\mu = 1.63, \sigma = 0.06)$ \\
& $\sigma$ & $\textrm{Normal}^{\textrm{+}}(\mu = 0.50, \sigma = 0.05)$ \\
Adjustments & $\bm{\gamma}$ & $\textrm{Student-t}(\nu = 7, \mu = 0, \sigma = 2.5)$ \\ 
\bottomrule
\end{tabular}
\caption{Prior choices for model parameters.}
\label{tbl:prior_choices}
\end{table}

\clearpage

\section{Modeling positive rate of human saliva samples}\label{sec:multinomial-model}

\section{Modeling changes in particle concentrations}\label{sec:env-regression-model}

The change in aerosol number concentration $CN_{jt}$ in class $j$ at day $t$ (analogously for particle mass concentrations $PM_{jt}$) is estimated using Bayesian log-linear regression models, \ie
\begin{align}
    \log\,CN_{jt} = \alpha + \beta\,\textrm{School}_j + \bm{\gamma}\,\textrm{Weekday}_t + \theta_M\,M_{jt} + \theta_A\,A_{jt} + \zeta\,\log\,\textrm{N}_{jt} + \omega\,\log\,\textrm{AER}_{jt}~, 
\end{align}
where $\alpha$ is the log of the aerosol concentration without interventions, $\beta$ and $\bm{\gamma}$ are school and weekday effects, respectively, $\theta_M$ and $\theta_A$ are the effects of mask mandates ($M_{jt}$) and air cleaners ($A_{jt}$), respectively, $\zeta$ adjusts for the number of students in class ($\textrm{N}_{jt}$), and $\omega$ adjusts for the outdoor air exchange rate ($\textrm{AER}_{jt}$). The latter is computed from measured indoor CO$_2$ levels (see Section~\ref{sec:rav-computation}). The percent reduction in particle concentrations with interventions are quantified as $100 \times (\mathrm{e}^{-\theta} - 1)\,\%$.

\clearpage

\section{Detailed model results for risk of infection model}\label{sec:detailed-redcap}

Table~\ref{tab:estimation-results} presents the posterior mean, credible intervals and model diagnostics for all model parameters. See the main paper for a discussion of the effects of interventions. Here we briefly discuss some of the additional model parameters.

The effective sample size (ESS) and the Gelman-Rubin convergence diagnostic ($\hat{R}$) indicate good estimation power. It further suggests that the Markov chains converged.

The estimate for $\tau$ indicates that there is variation in the effects of mask mandates between classes. However, the credible intervals of $\theta_j^M$ all include zero, indicating that there is only mild deviation of the class-specific estimates from their cross-class average estimate $\theta^M$.

The mean estimate for the proportion of students being absent is negative, indicating that higher proportion of absences decrease the probability of getting infected. In contrast to that, the estimate for the reproduction number in the community is positive, indicating that higher community transmission increases the probability of getting infected. Both estimates are in line with our hypothesized effect, although it should be noted that both estimates have large credible intervals including zero.

The overdispersion parameter ($\phi$) cannot be precisely estimated, but the credible intervals indicate rather small overdispersion ($\phi \rightarrow \infty$).

The estimates for the logit of the probability of getting infected during school days without interventions ($\alpha$) are larger than the prior mean. This is due to the reduction in transmission from mask mandates. The weekend effect ($\omega$) is negative, but smaller than the mean of our informative prior, indicating that the likelihood suggests more comparable transmission on weekends.

The posterior distribution of the parameters of the incubation period ($\mu^{p_{IN}}$ and $\sigma^{p_{IN}}$) are close to their prior, indicating that the data is not informative about the incubation period.

\begin{table}[!htpb]
    \caption[Estimation results from transmission model]{Estimation results from transmission models across the 100 generated datasets for the number of new cases of COVID-19.}
    \label{tab:estimation-results}
    \footnotesize
    \centering
    \begin{tabular}{lrrrrr}
    \toprule
    Parameter & Mean & Lower 95\%-CrI & Upper 95\%-CrI & $\hat{R}$ & ESS \\
    \midrule
    \input ../../results/epi-data/estimation-results
    \bottomrule
    \multicolumn{6}{p{13cm}}{\scriptsize
        ESS is the effective sample size, \ie the number of independent MCMC samples with estimation power equivalent to the total number of autocorrelated samples\cite{Stan2022}, and $\hat{R}$ is the Gelman-Rubin convergence diagnostic \cite{Gelman1992}. Low ESS or $\hat{R}$ or $\hat{R}>1.10$ indicate bad convergence of the model \cite{Gelman2013}.}
    \end{tabular}
\end{table}

Our model accounts for the delay from infection to case confirmation and allows transmission to change only at the dates of interventions. It thus reflects overall trends in transmission during study conditions rather than day-to-day variation. To evaluate how well our model fits these trends, Fig.~\ref{fig:coverage} compares the estimated number of cases from the probabilistic simulation with the estimated number of new cases from the transmission model. Overall the estimates are in good agreement. The 95\%-CrI of the model-based estimates includes the 95\%-quantile of the simulation-based estimates.

\begin{figure}[!htpb]
    \centering
    \includegraphics{../../results/epi-data/model-fit.pdf}
    \caption[Model- and simulation-based estimates of the number of COVID-19 cases]{Estimated (posterior mean as dot and 95\%-CrI as line) number of cases from our probabilistic simulation (blue) and transmission model (black).}
    \label{fig:coverage}
\end{figure}

\clearpage

\section{Detailed model results for positivity rate in human saliva samples}\label{sec:detailed-molecular}

\begin{figure}[!htb]
\centering
    \includegraphics[width=\linewidth]{../../results/mol-data/model-results.pdf}
    \caption[Boxplot of environmental variables by school and study condition]{Boxplot for the daily average values of each environmental variable by school and study condition.}
    \label{fig:mol-estimation-results-sensitivity}
\end{figure}

\clearpage

\section{Detailed model results for changes in particle concentrations}\label{sec:detailed-palas}

In the main paper, we showed results for changes in particle concentrations by study condition and summarized them across schools. In Fig.~\ref{fig:palas-supp-descriptives}, we present additional results disaggregated by schools and including other environmental variables. Furthermore, numerical estimation results for the reduction in particle concentrations with interventions are shown in Table~\ref{tab:palas-est-results}

\begin{figure}[!htb]
\centering
    \includegraphics[width=\linewidth]{../../results/env-data/othVars-boxplot.pdf}
    \caption[Boxplot of environmental variables by school and study condition]{Boxplot for the daily average values of each environmental variable by school and study condition.}
    \label{fig:env-descriptives-other-vars}
\end{figure}

\begin{table}[!htpb]
    \caption[Estimated reduction in aerosol and particle concentrations with interventions]{Estimated reduction in aerosol number (CN) and particle mass (PM) concentrations with interventions (posterior mean and upper and lower estimate from the 95\%-CrI).}
    \label{tab:palas-est-results}
    \centering
    \footnotesize
    \begin{tabular}{l r r r c r r r }
    \toprule
    & \multicolumn{3}{c}{Mask mandate} & & \multicolumn{3}{c}{Air cleaner} \\ \cline{2-4} \cline{6-8}
    Variable & Mean & Lower & Upper & & Mean & Lower & Upper \\
    \midrule
    \input ../../results/env-data/estimation-results-table.tex
    \bottomrule
    \end{tabular}

\end{table}

\clearpage

\bibliography{references.bib}

\end{document}

### 2023-06-20 analysis/epi_analysis.Rmd
minor update of results

@@
                       "$\\sigma_{p_{\\text{IN}}^{\\text{PIV}}}")) %>%
-  dplyr::select(variable, mean, `X2.5.`, `X97.5.`, n_eff, Rhat) %>%
-  mutate(across(c(mean, `X2.5.`, `X97.5.`), function(x) tex_sign(round_k(x)))) %>%
+  dplyr::select(variable, mean, `X2.5.`, `X97.5.`, Rhat, `n_eff`) %>%
+  mutate(across(c(mean, `X2.5.`, `X97.5.`, Rhat), function(x) tex_sign(round_k(x))),
+         `n_eff` = round(`n_eff`)) %>%
   as.matrix()

### 2023-06-20 analysis/mol_analysis.Rmd
minor update line list data
fix line list mol
minor update of results

@@
 line_list_dat <- df_ll %>%
+  mutate(weekday = weekdays(date)) %>%
+  dplyr::select(class, date, weekday, result, virus) %>%
+  arrange(class, date) %>%
   as.matrix() %>%
@@
                "1 test date",
-               "1 week")
+               "1 test week")
 

### 2023-06-21 analysis/epi_analysis.Rmd
minor fixes in estimation results tex tables

--- 2023-06-20 analysis/epi_analysis.Rmd
@@
                       "$\\vega_{\\text{Friday}}$",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{Flu B}}}",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{HRV}}}",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{AdV}}}",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{CoV}}}",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{MPV}}}",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{PIV}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{Flu B}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{HRV}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{AdV}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{CoV}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{MPV}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{PIV}}}")) %>%
-  dplyr::select(variable, mean, `X2.5.`, `X97.5.`, n_eff, Rhat) %>%
-  mutate(across(c(mean, `X2.5.`, `X97.5.`), function(x) tex_sign(round_k(x)))) %>%
+                      "$\\mu_{p_{\\text{IN}}^{\\text{Flu B}}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{HRV}}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{AdV}}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{CoV}}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{MPV}}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{PIV}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{Flu B}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{HRV}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{AdV}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{CoV}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{MPV}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{PIV}}}$")) %>%
+  dplyr::select(variable, mean, `X2.5.`, `X97.5.`, Rhat, `n_eff`) %>%
+  mutate(across(c(mean, `X2.5.`, `X97.5.`, Rhat), function(x) tex_sign(round_k(x))),
+         `n_eff` = round(`n_eff`)) %>%
   as.matrix()

### 2023-06-21 doc/paper/supplements.tex

@@
 
+\clearpage
+
 {\footnotesize\begin{longtable}{l l l l l}
-    \caption[Line list of molecular test results over the study period]{Line list of respiratory cases over the study period.}\label{tab:mol-data-line-list} \\
+    \caption[Line list of molecular test results over the study period]{Line list of respiratory cases over the study period.
+Pathogens: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}\label{tab:mol-data-line-list} \\
     \toprule
@@
 
-\section{Modeling relative risk of respiratory infections}\label{sec:transmission-model}
+\section{Modeling relative risk of infection}\label{sec:transmission-model}
 
@@
 \centering
-    \includegraphics[width=\linewidth]{../../results/mol-data/model-results.pdf}
+    \includegraphics[width=10cm]{../../results/mol-data/model-results.pdf}
     \caption[Boxplot of environmental variables by school and study condition]{Boxplot for the daily average values of each environmental variable by school and study condition.}
@@
 \centering
-    \includegraphics[width=\linewidth]{../../results/env-data/othVars-boxplot.pdf}
+    \includegraphics[width=\linewidth]{../../results/env-data/otherVars-boxplot.pdf}
     \caption[Boxplot of environmental variables by school and study condition]{Boxplot for the daily average values of each environmental variable by school and study condition.}
@@
     \footnotesize
-    \begin{tabular}{l r r r c r r r }
+    \begin{tabular}{l r r r}
     \toprule
-    & \multicolumn{3}{c}{Mask mandate} & & \multicolumn{3}{c}{Air cleaner} \\ \cline{2-4} \cline{6-8}
-    Variable & Mean & Lower & Upper & & Mean & Lower & Upper \\
+    Variable & Mean & Lower & Upper \\
     \midrule

### 2023-06-21 doc/paper/supplements.tex
Updates from Overleaf

@@
 Table~\ref{tab:epi-data-line-list} shows the line list data for respiratory cases based on epidemiological data.
+For each class and case, it shows the date when the student was absent, s/he reported symptoms, and returned to school.
+Recall that a case of respiratory infection was defined as an absence where the student reported a sickness with at least one of the following symptoms: fever, coughing, tiredness, loss of test or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, stomach.
 
@@
 \clearpage
+
+\section{Molecular line list data}\label{sec:mol-data}
+
+Table~\ref{tab:mol-data-line-list} shows the line list data for the laboratory test results from human saliva samples.
+For each class and test, it shows the date when the test was taken, the test result and (if positive) the pathogen that was detected.
+Recall that all students in class were sampled twice per week every Tuesday and Thursday.
 
@@
 
-\section{Molecular line list data}\label{sec:mol-data}
-
-Table~\ref{tab:mol-data-line-list} shows the line list data for the laboratory test results from human saliva samples.
-
-\clearpage
-
 \section{Modeling relative risk of infection}\label{sec:transmission-model}
@@
 
-The aim is to estimate transmission of SARS-CoV-2 and the effects of infection control measures (mask mandates and air cleaners) on the number of new infections with COVID-19.
-To estimate transmission, our model links two unobserved quantities (the number of new infections and susceptible students) with one observed quantity (the number of new cases among students).
-Specifically, we link the number of new cases to the number of new infections in the previous days.
-We then formulate the daily proportion of susceptible students getting infected as a function of the interventions in place.
-The effects of interventions are estimated with a Bayesian approach, which requires the specification of prior distributions for all model parameters.
+The overall aim is to estimate the effects of air cleaners on the daily number of respiratory infections.
+The latter is unobserved/latent and inferred from the daily number of respiratory cases (absences related to respiratory infections by date of symptom onset), considering the delay from infection to symptom onset (\ie the incubation period).
+The effect of air cleaners is estimated with a Bayesian approach, which requires the specification of prior distributions for all model parameters.
 
@@
 $t$ & days since start of study period \\
-$Y_{j}$  & total number of students in class $j$ (unobserved)  \\
-$N_{jt}$  & number of  new cases among students in class $j$ at day $t$ (observed) \\
+$C_{jt}$  & number of  new cases among students in class $j$ at day $t$ (observed) \\
 $I_{jt}$  & number of new infections in class $j$ at day $t$ (unobserved)  \\
-$S_{jt}$  & number of susceptibles in class $j$ at day $t$ (unobserved)  \\
+$F_{jt}$  & number of new infections in class $j$ before day $t$ (unobserved)  \\
+$N_{jt}$  & cumulative number of infections in class $j$ until day $t$ (unobserved)  \\
 \end{tabular}
 
-\subsection{Relating the number of new infections to the number of susceptibles and the presence of interventions}
+\subsection{Relating the number of respiratory cases to the number of infections}
 
-A proportion of the susceptible students $P_{jt}$ gets infected in class $j = 1,\dots,J$ at each day $t = 1,\dots,T$
+The number of respiratory cases $C$ in class $j$ at time $t$ is modeled with a Negative Binomial distribution   
+\begin{align*}
+    C_{jt} \sim \text{Negative-Binomial}(\mu_{jt},\phi),
+\end{align*}
+where $\mu_{jt}$ is the expected number of new cases and $\phi$ is the parameter modeling over-dispersion.
+The expected number of new cases is the weighted sum of the number of new infections $I_{jt}$ in the  previous days
+\begin{align*}
+    \mu_{jt} = \sum_{s<t}I_{js} \cdot p_{\text{IN}}(t-s),
+\end{align*}
+where $p_{\text{IN}}(t-s)$ denotes the probability distribution of the incubation period.
+
+\subsection{Relating the number of new infections to the presence of interventions}
+
+The  number of new infections is related to the presence of air cleaners using a log-link
+\begin{align*}
+    \log I_{jt} &= \log F_{jt} - \log N_{jt} + \beta_0 + \beta_1 \cdot \text{AirCleaner}_{jt} + \beta_2 \cdot \text{Class}_t + \beta_3 \cdot \text{Students}_{jt} \\
+    &\hphantom{= }+ \beta_4 \cdot \text{Ventilation}_{jt} + \beta_5 \cdot \text{CoV}_t + \beta_6 \cdot \text{ILI}_t,
+\end{align*}
+where $F_{jt} = \sum_{t-7}^{t-1} I_{js}$ is the number of infections in the previous seven days (first model offset; a proxy for the number of contagious students), $N_{jt} = \sum_{s<t} I_{js}$ is the cumulative number of infections (second model offset; the inverse is a proxy for the number of susceptible students), $\beta_0$ is the rate of new infections without air cleaners (model intercept), and $\beta_1$ is the effect of air cleaners.
+The effect of air cleaners is adjusted for class-specific effects, the number of students in school, the air change rate, the proportion of positive tests for SARS-CoV-2 in the community, and the number of consultations for influenza-like illnesses in the community.
+
+Note that we deviated from the statistical analysis plan by using as model offset $\log F_{jt} - \log N_{jt}$ instead of just $\log N_{jt}$.
+This change improved the fit of our model.
+The reason is that the ratio $\log (F_{jt} / N_{jt})$ incorporates both ways by which new infections can naturally decrease over time, \ie a decrease in contagious students ($F_{jt}$) or a decrease in susceptible students ($1 / N_{jt}$).
+
+\subsection{Specifying the distribution of the incubation period}
+
+The pathogen of each respiratory infection could not be identified from the epidemiological data because the students never obtained a laboratory test result.
+As a consequence, different incubation periods need to be considered in $p_{\text{IN}}$, reflecting a combination of the pathogen-specific incubation periods.
+The combination is determined based on the weekly proportion of positive saliva samples for each pathogen found in the molecular analysis.
+Formally, let $p_{\text{IN}}^{v}$ be the distribution for the incubation period of respiratory virus $v$ and let $pp_{vw}$ be the proportion of positive saliva samples in study week $w$, then each week the combined incubation period is computed as the weighted sum of the pathogen-specific incubation periods
 \begin{align}
-    I_{jt} = S_{jt} P_{jt} ~.
-\end{align}
-This proportion (or probability) can be estimated using a logit link function
-\begin{align}
-    \Phi_{jt} &= \log\left(\frac{P_{jt}}{1-P_{jt}}\right) \\
-    \textrm{logit}^{-1}(\Phi_{jt}) &= \frac{\exp{(\Phi_{jt})}}{1 + \exp{(\Phi_{jt})}} = P_{jt} ~.
+    p_{\text{IN}}^w(s) = \sum_{v} p_{\text{IN}}^{v}(s) \cdot pp_{vw} \quad \forall s.
 \end{align}
 
-In the absence of interventions, the probability of getting infected is constant, \ie $\Phi_{jt} = \beta_j$.
-It can only change at times when interventions are in place.
-Let $\theta_{j}^M$ be the effect of mask mandates and let $\theta_{j}^A$ be the effect of air cleaners in class $j$, then
-\begin{align}
-    \Phi_{jt} = \beta_j + \theta_{j}^M\,{M}_{jt}
-    + \theta_{j}^A\,{A}_{jt},
-\end{align}
-where ${M}_{jt}$ and ${A}_{jt}$ are binary variables indicating whether mask mandates and air cleaners are in place in class $j$ at day $t$.
+The prior distributions for the pathogen-specific incubation periods are based on estimates published in the literature\cite{McAloon2020,Lessler2009LancetID}.
+The distributions are shown in \Cref{fig:prior-pin}.
+Since we could not obtain prior estimates for the incubation period of metapneumovirus (MPV) from the literature, we instead formed a distribution by using the equally weighted average of the parameters from the other distributions.
 
-\subsection{Relating the number susceptibles to the number of infections}
-
-The daily number of susceptibles $S_{jt}$ are computed as the difference between the total number of students $Y_j$ and the cumulative number of infections $C_{jt}$ in the previous days
-\begin{align}
-    S_{jt} = Y_j - C_{jt} = Y_j - \sum_{s < t} I_{js}~.
-\end{align}
-We assume that all students are susceptible at the beginning of the study period, despite the fact that some students were vaccinated or already infected with COVID-19 prior to the beginning of our study.
-Note that varying susceptibility before study onset can be subsumed in the class-specific transmission rate without interventions $\beta_j$.
-
-\subsection{Relating the number new observed cases to the number of new infections}
-
-The expected number $\mu$ of new cases $N_{jt}$ in country $j$ at day $t$ can be derived from the number of new infections in the previous days as 
-\begin{equation}
-\mu^{N_{jt}} = \sum_{s < t} I_{js} \cdot p_\text{IN}(t-s) ~, 
-\end{equation}
-where $p_\text{IN}(t)$ denotes the probability distribution of the incubation period, \ie the probability that a new infected subject reports symptoms at day $t$ after the infection.
-This distribution is estimated from our data as part of fitting the overall model using an informative prior (Section~\ref{subsec:prior-incubation-period}).
-
-The observed number of new cases are modeled to follow a negative binomial distribution, \ie 
-\begin{equation}
-N_{jt}  \sim  \textrm{NegBinom}\left(\mu^{N_{jt}}, \sigma^{N_{jt}}\right) \label{eq:N} 
-\end{equation}
-with mean $\mu^{N_{jt}}$, standard deviation $\sigma^{N_{jt}} = \sqrt{\mu^{N_{jt}} \, \left(1 + \frac{\mu^{N_{jt}}}{\phi^N}\right)}$, and an overdispersion parameter $\phi^N$.
-
-\subsection{Taking into account transmission outside school days}\label{subsec:consider-weekends}
-
-Absences were reported on school days but symptom onset could have occurred on the weekend, \eg a student absent on Monday due to COVID-19 may report Saturday as the day of symptom onset.
-Some cases may still have gone unreported because the date of absence was the same as the date of symptom onset for the majority of cases (see Fig.~\ref{fig:delay-weights}).
-Therefore, we distinguish between the transmission rate during and outside schools days, \ie
-\begin{align}
-    \beta_j = \alpha_j + \omega, 
-\end{align}
-where $\alpha_j$ and $\alpha_j + \omega$ are the logit of the probability of getting infected without interventions during and outside school days, respectively.
-We estimate $\alpha_j$ and $\omega$ from data using informative priors (Section~\ref{subsec:prior-alpha}).
-We further assume that transmission outside school days is the same regardless of whether they are weekend or vacation days, but we do not incorporate vacation days into the model likelihood.
-That is, during vacation, the modeling of $N_{jt}$ is ignored but $I_{jt}$ and $S_{jt}$ are still computed.
-
-\subsection{Seeding phase}
-
-A few cases were reported in the first days of the study period, indicating that the corresponding infections could have occurred before the start of the study.
-Therefore, we seed the number of new infections and cases before the study period by initiating our model 12~days (double the average incubation period) prior to the start of the study.
-Before this day, we assume there have been no infections and all students are susceptible.
-
-\subsection{Adjusting for absences and community transmission}
-
-The probability of getting infected may be influenced by factors other than interventions.
-We account for two such factors.
-First, we adjust our estimates for the daily proportion of students absent from class.
-We expect this proportion to decrease the probability of getting infected because absent students cannot infect classmates and, if staying home, they are also less likely to get infected themselves (if not already infected).
-Second, we adjust our estimates for the risk of community transmission, which we proxy with the average of the median effective reproduction number for the canton of Solothurn, Switzerland\cite{Scire2022}.
-We expect higher community transmission to increase the probability of getting infected.
-The modeling parameters of these factors are denoted by $\bm{\gamma}$.
-
-\subsection{Choice of priors for the probability of getting infected without interventions}\label{subsec:prior-alpha}
-
-We choose informative priors for the logit of the daily probability of getting infected in the absence of interventions $\beta_j = \alpha_j + \omega$.
-First, our prior for $\alpha_j$ is derived from the share of susceptibles by the end of the study period $S_{jT} / Y_j$, \ie
-\begin{align}
-    S_{jT} / Y_j = (1 - \textrm{logit}^{-1}(\alpha_j))^{T_{\mathrm{school}}},
-\end{align}
-where $T_{\mathrm{school}}$ is the number of school days (excluding weekends and vacation), and solving for $\alpha_j$
-\begin{align}
-    \textrm{logit}^{-1}(\alpha_j) = 1 - \exp\left(\frac{\log(S_{jT} / Y_j)}{T_{\mathrm{school}}}\right) ~.
-\end{align} 
-Note that $S_{jT} / Y_j$ and correspondingly $\textrm{logit}^{-1}(\alpha_j)$ will be different for each of our probabilistically generated datasets.
-Fig.~\ref{fig:prior-alpha}a shows the daily proportion of susceptibles getting infected based on the distributions for the proportion of suspected cases being actual cases of COVID-19 (see Fig.~\ref{fig:prior-proportion}).
-The corresponding distribution for $\alpha_j$ is shown in Fig.~\ref{fig:prior-alpha}b.
-We choose a prior for $\alpha_j \sim \textrm{Student-t}(\nu = 3, \mu = -4.1, \sigma = 1.6)$ with wider tails than the empirical distribution to accommodate the fact that $\alpha_j$ may be smaller or larger depending on the effects of interventions and adjustment variables.
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{figures/redcap-supp-prior-alpha.pdf}
-    \caption[Prior for the probability of getting infected without interventions]{Prior for the probability of getting infected without interventions.
-\textbf{(a)}~Empirical distribution for the daily proportion of susceptibles getting infected across classes.
-The dotted blue line indicates the mean.
-\textbf{(b)}~Empirical distribution for the logit of the daily proportion of susceptibles getting infected.
-The dotted blue line indicates the empirical mean.
-The corresponding choice of prior for the probability of getting infected without interventions $\alpha$ is hown in red.
-}
-    \label{fig:prior-alpha}
-\end{figure}
-
-Second, we choose a prior for $\omega \sim \textrm{Normal}(\mu = \log\,0.7, \sigma = 0.2)$ that encodes our prior belief that students will have fewer indoor contacts outside school days and thus lower odds of getting infected.
-In addition, this informative prior also considers potential underreporting of cases on weekends (see Section~\ref{subsec:consider-weekends}).
-
-\clearpage
-
-\subsection{Choice of priors for the effects of interventions}
-
-We consider that the mandates may not have been equally effective in all classes.
-Therefore, we model the effects of mask mandates as $\theta_{j}^M = \textrm{Normal}(\theta^M, \tau)$, where $\theta^M \sim \textrm{Student-t}(\nu = 7, \mu = 0, \sigma = 2.5)$ is the average effect across classes and $\tau \sim \textrm{Student-t}^{+}(\nu = 5, \mu = 0, \sigma = 1)$ is the variation in the effect between classes.
-
-Air cleaners only applied to three classes and were installed towards the end of the study period when most students may have already been infected.
-Therefore, we do not estimate variation in the effect between classes but only the average effect.
-In addition, we choose a prior leaning towards smaller effects, \ie $\theta^A \sim \textrm{Student-t}(\nu = 7, \mu = 0, \sigma = 1)$.
-
-\subsection{Choice of priors for the incubation period}\label{subsec:prior-incubation-period}
-
-We estimate the distribution of the incubation period from our data as part of fitting the overall model.
-Our prior choices for the parameters of this distribution are informed by a meta-analysis\cite{McAloon2020}, which reported a Lognormal distribution with $\log \mu \sim \textrm{Normal}(1.63, 0.06)$ and $\log \sigma \sim \textrm{Normal}^{\textrm{+}}(0.50, 0.05)$.
-The prior for $p_\text{IN}$ is shown in Fig.~\ref{fig:prior-pin}.
-Note that $p_\text{IN}$ is discretized via $p_\text{IN}(s) = \int_0^{0.5} p_\text{IN}(\tau) \;\text{d}\tau$ for $s = 0$ and $p_\text{IN}(s) = \int_{s-0.5}^{s+0.5} p_\text{IN}(\tau) \;\text{d}\tau$ for $s > 0$, where $p_\text{IN}(\tau) \sim \textrm{Lognormal}(\mu, \sigma)$ is the density of the lognormal distribution.
+We estimate the pathogen-specific distributions from our data as part of fitting the overall model.
+Furthermore, note that $p_\text{IN}$ is discretized via $p_\text{IN}(s) = \int_0^{0.5} p_\text{IN}(\tau) \;\text{d}\tau$ for $s = 0$ and $p_\text{IN}(s) = \int_{s-0.5}^{s+0.5} p_\text{IN}(\tau) \;\text{d}\tau$ for $s > 0$, where $p_\text{IN}(\tau) \sim \textrm{Lognormal}(\mu, \sigma)$ is the density of the lognormal distribution.
 
@@
     \includegraphics{../../results/epi-data/incubation-periods.pdf}
-    \caption[Choices of prior for incubation period]{Choices of prior for incubation period.
-\textbf{(a)}~Prior for parameter $\log \mu$.
-\textbf{(a)}~Prior for parameter $\log \sigma$.
-\textbf{(c)}~Resulting prior $p_\text{IN}$ (posterior mean as line and 50\%, 80\%, and 95\%-quantile as shaded area).}
+    \caption[Choices of priors for the incubation periods]{Choices of priors for the distribution of the pathogen-specific incubation periods.
+Pathogens: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}
     \label{fig:prior-pin}
@@
 
-\subsection{Choice of prior distributions (summary)}
+\subsection{Adjusting for under-reporting of cases on weekends}
 
-Table~\ref{tbl:prior_choices} provides an overview of the model parameters together with the prior choices.
-If not tailored to the specifics of the model, our choices are  informed by recommendations on the choice of priors from the Stan Development Team \cite{Stan2020Priors}.
+Despite recording cases by date of symptom onset, we recorded a higher proportion of cases on Mondays than on weekends, suggesting recall bias and under-reporting of cases on weekends.
+To consider weekday effects in the reporting of cases, we re-weight the expected number of cases each week as follows.
+Let $k \in $(1: Saturday, 2: Sunday, 3: Monday, $\dots$, 7: Friday) denote the weekday with the week starting on Saturday.
+The re-weighted expected number of cases $\Tilde{\mu}$ (class and day indexes omitted) are computed as
+\begin{align*}
+    \Tilde{\mu}_{k} &= \mu_k \cdot \nu_k \cdot \left(\frac{\sum_k \mu_k}{\sum_k \nu_k \cdot \mu_k}\right)\\
+    \sum_k \nu_k &= 1,
+\end{align*}
+where $\nu_k$ is the weight for weekday $k$. These weights are modeled with a Dirichlet prior
+\begin{align*}
+    \nu &\sim \text{Dirichlet}(c) \\
+    c_k &= \sum_j \sum_t C_{jt} \mathbb{I}_{\text{Weekday}(t)=k}
+\end{align*}
+where $c_k$ is the total number of cases reported for weekday $k$ and $\mathbb{I}$ is a binary indicator.
 
-\begin{table}
-\centering
-\footnotesize
-\begin{tabular}{p{7cm}p{1.5cm}p{6.5cm}}
-\toprule
-Parameter & Notation & (Hyper-)Prior  \\ 
-\midrule
-Effects of mask mandates & $\theta_{j}^M$ &  $\textrm{Normal}(\theta^M, \tau)$\\
-& $\theta^M$ & $\textrm{Student-t}(\nu = 7, \mu = 0, \sigma = 2.5)$\\
-& $\tau$ & $\textrm{Student-t}^{+}(\nu = 5, \mu = 0, \sigma = 1)$ \\
-Effects of air cleaners & $\theta^A$ & $\textrm{Student-t}(\nu = 7, \mu = 0, \sigma = 1)$\\
-Prob. of getting infected w/o interventions (logit) & $\beta_j$ & $\beta_j = \alpha_j + \omega$ \\
-Prob. of getting infected during school (logit) & $\alpha_j$ & $\textrm{Student-t}(\nu = 3, \mu = -4.1, \sigma = 1.6)$ \\
-Weekend effect & $\omega$ & $\textrm{Normal}(\mu = \log\,0.7, \sigma = 0.2)$ \\
-Overdispersion & $\phi^N$ & $\phi^N = \left(\frac{1}{\xi^N}\right)^2$ \\
- & $\xi^N$ & $\textrm{Normal}^{+}(\mu = 0, \sigma = 1)$ \\
-Time from infection to new case & $p_\text{IN}$ & $\textrm{Lognormal}(\log \mu, \log \sigma)$ \\
-& $\mu$ & $\textrm{Normal}(\mu = 1.63, \sigma = 0.06)$ \\
-& $\sigma$ & $\textrm{Normal}^{\textrm{+}}(\mu = 0.50, \sigma = 0.05)$ \\
-Adjustments & $\bm{\gamma}$ & $\textrm{Student-t}(\nu = 7, \mu = 0, \sigma = 2.5)$ \\ 
-\bottomrule
-\end{tabular}
-\caption{Prior choices for model parameters.}
-\label{tbl:prior_choices}
-\end{table}
+\subsection{Modeling school-free days}
+
+Infections may have occurred during the week of vacation that falls into the study period.
+The expected number of infections and cases are computed during vacation, but vacation days are not modeled (\ie not incorporated into the model likelihood).
+In addition, we assume lower transmission of respiratory infections on days without school (weekends and vacations).
+We incorporate our prior belief into the model intercept $\beta_0$
+\begin{align*}
+    \beta_0 &= \alpha + \omega \cdot \text{NoSchool},
+\end{align*}
+where $\alpha$ is the rate of new infections on school days and $\alpha + \omega$ is the rate on days without school.
+We model $\omega$ with an informative prior for a 10\% decrease in new infections on school-free days
+\begin{align*}
+    \omega &\sim \text{Normal}(\log 1.1, 0.05)~.
+\end{align*}
+
+\subsection{Seeding infections before study start}
+
+Cases in the first week of the study could indicate infections before the study commenced.
+We will therefore seed our model $2 \cdot \text{m}\,$days before the study start, where $\text{m}$ is the average incubation period of the pathogen with the largest incubation period (\ie adenovirus).
+The number of infections before the study start will be modeled with an exponential prior
+\begin{align*}
+    I_{jt} \sim \text{Exponential}(1) \quad t = -2\cdot\text{m}+1, \dots, 0~. 
+\end{align*}
+
+Note that we deviated from the statistical analysis plan by changing the parameter of the Exponential distribution from $\lambda = 2 \cdot m$ to $\lambda = 1$.
+After inspecting the model fit, we realized that the model could not adequately fit the number of new cases at the start of study, because the expected number of new infections in the seeding period $\mu = 1 / \lambda = 1 / (2 \cdot m)$ were too low.
+After increasing the expected number of new infections in the seeding period to $\mu = 1$ the model could adequately capture new respiratory cases observed at the study start.
+
+\subsection{Priors for modeling parameters}
+
+We use weakly informative priors for all modeling parameters following recommendations on the choice of priors \cite{Gelman2008StatMed,Gelman2008StatAnnals,Gelman2020RegOther,Stan2020Priors,Gabry2023Priors}.
+The continuous adjustment variables are standardized to have zero mean and a standard deviation of 0.5.
+\begin{align*}
+    \alpha &\sim \text{Student-t}_5(0, 10) \\
+    \beta_2, \dots, \beta_6 &\sim \text{Student-t}_5(0, 2.5) \\
+    \frac{1}{\sqrt{\phi}} &\sim \text{Half-Normal}(0,1),
+\end{align*}
+where Student-t$_5$ is a Student-t distribution with 5 degrees of freedom.
 
@@
 
-\section{Modeling positive rate of human saliva samples}\label{sec:multinomial-model}
+\section{Modeling positivity rate of human saliva samples}\label{sec:multinomial-model}
 
@@
 
-\section{Detailed model results for risk of infection model}\label{sec:detailed-redcap}
+\section{Detailed results for risk of infection model}\label{sec:detailed-redcap}
 
@@
 
-\section{Detailed model results for positivity rate in human saliva samples}\label{sec:detailed-molecular}
+\section{Detailed results for positivity rate in saliva samples}
+\label{sec:detailed-molecular}
 
@@
 
-\section{Detailed model results for changes in particle concentrations}\label{sec:detailed-palas}
+\section{Detailed results for changes in particle concentrations}\label{sec:detailed-palas}
 

### 2023-06-21 analysis/env_analysis.Rmd
add line list data and minor results update
minor fix latex estimation results

@@
 names(vars_labels_no_units) <- vars
+```
+
+## Summaries
+
+```{r}
+env_epi_data %>%
+  group_by(aircleaner) %>%
+  summarize(across(c(CO2, AER, Cn), .fns = list(mean = mean, sd = sd), .names = "{.col}_{.fn}")) %>%
+  ungroup() %>%
+  mutate(across(c(CO2_mean, CO2_sd, Cn_mean, Cn_sd), round),
+         across(c(AER_mean, AER_sd), round, 2))
 ```
@@
   arrange(-row_number()) %>%
-  mutate(outcome = c("CN", "PM$_{1}$", "PM$_{2.5}$", "PM$_{4}$", "PM$_{10}$"),
-         space = "") %>%
+  mutate(outcome = c("CN", "PM$_{1}$", "PM$_{2.5}$", "PM$_{4}$", "PM$_{10}$")) %>%
   as.matrix() %>%

### 2023-06-21 analysis/epi_analysis.Rmd
minor fix latex estimation results

--- 2023-06-20 analysis/epi_analysis.Rmd
@@
                       "$\\beta_6$",
-                      "$\\vega_{\\text{Saturday}}$",
-                      "$\\vega_{\\text{Sunday}}$",
-                      "$\\vega_{\\text{Monday}}$",
-                      "$\\vega_{\\text{Tuesday}}$",
-                      "$\\vega_{\\text{Wednesday}}$",
-                      "$\\vega_{\\text{Thursday}}$",
-                      "$\\vega_{\\text{Friday}}$",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{Flu B}}}",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{HRV}}}",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{AdV}}}",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{CoV}}}",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{MPV}}}",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{PIV}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{Flu B}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{HRV}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{AdV}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{CoV}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{MPV}}}",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{PIV}}}")) %>%
-  dplyr::select(variable, mean, `X2.5.`, `X97.5.`, n_eff, Rhat) %>%
-  mutate(across(c(mean, `X2.5.`, `X97.5.`), function(x) tex_sign(round_k(x)))) %>%
+                      "$\\vartheta_{\\text{Saturday}}$",
+                      "$\\vartheta_{\\text{Sunday}}$",
+                      "$\\vartheta_{\\text{Monday}}$",
+                      "$\\vartheta_{\\text{Tuesday}}$",
+                      "$\\vartheta_{\\text{Wednesday}}$",
+                      "$\\vartheta_{\\text{Thursday}}$",
+                      "$\\vartheta_{\\text{Friday}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{Flu B}}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{HRV}}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{AdV}}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{CoV}}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{MPV}}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{PIV}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{Flu B}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{HRV}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{AdV}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{CoV}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{MPV}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{PIV}}}$")) %>%
+  dplyr::select(variable, mean, `X2.5.`, `X97.5.`, Rhat, `n_eff`) %>%
+  mutate(across(c(mean, `X2.5.`, `X97.5.`, Rhat), function(x) tex_sign(round_k(x))),
+         `n_eff` = round(`n_eff`)) %>%
   as.matrix()

### 2023-06-21 doc/paper/manuscript.tex

@@
 Reports about absences were entered electronically into a REDcap database\cite{Harris2009,Harris2019}.
-Furthermore, both classes participated in repetitive bi-weekly (Mondays and Wednesdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below).
+Furthermore, both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below).
 The saliva samples were transported to the laboratory and stored at $-$80°C until further processed\cite{Galar2021,To2019,Huber2021}.
@@
 The latter was computed as the difference between the number of students per class and the cumulative number of positive samples.
-As a sensitivity analysis, we lagged the test results by one test date (\ie from Wednesday to Monday and from Monday to last Wednesday) and one test week (\ie from Monday to last Monday and Wednesday to last Wednesday), to consider the unknown delay when students were infected and tested positive.
+As a sensitivity analysis, we lagged the test results by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday), to consider the unknown delay when students were infected and tested positive.
 

### 2023-06-22 analysis/epi_analysis.Rmd
minor update to result plots

@@
   ungroup() %>%
-  ggplot(aes(x = factor(week), y = cases, fill = aircleaner)) +
-  geom_bar_pattern(stat = "identity", aes(pattern_angle = class), pattern = "stripe",
-                   pattern_fill = "white", pattern_color = NA, color = "black", pattern_density = 0.25, pattern_spacing = 0.05) +
-  labs(x = "Study week", y = "Number of cases", fill = "Air cleaner", pattern_angle = "Class", title = "a") +
-  scale_fill_manual(values = wes_palette("Royal1")) +
+  mutate(aircleaner = factor(aircleaner, levels = c("Yes", "No")),
+         class = factor(class, levels = c("A", "B"))) %>%
+  arrange(class, aircleaner) %>%
+  ggplot(aes(x = factor(week), y = cases, fill = aircleaner, pattern = class)) +
+  geom_bar_pattern(stat = "identity", color = "black", pattern_fill = "white", pattern_angle = 45,
+                   pattern_density = 0.25, pattern_spacing = 0.05) +
+  labs(x = "Study week", y = "Number of cases", fill = "Air cleaner", pattern = "Class", title = "a") +
+  scale_fill_manual(values = rev(wes_palette("Royal1")[1:2])) +
   scale_x_discrete(breaks = c(1:5, 7:10)) +
   scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
-  scale_pattern_angle_discrete(range = c(-45, 45)) +
+  scale_pattern_manual(values = c(A = "stripe", B = "none")) +
   theme_custom() +
@@
         plot.title.position = "plot") +
-  guides(fill = guide_legend(nrow = 1),
-         pattern_angle = guide_legend(nrow = 1))
+  guides(fill = guide_legend(nrow = 1, override.aes = list(pattern = "none", order = 2)),
+         pattern = guide_legend(nrow = 1, override.aes = list(fill = "white"), order = 1))
 
@@
           eps_file = "../results/epi-data/cases_by_week.eps",
-          w = 12, h = 8) 
+          w = 16, h = 8) 
 ```

### 2023-06-22 analysis/mol_analysis.Rmd
minor update to result plots

@@
   scale_x_continuous(breaks = seq(1, 7, 1)) +
-  labs(y = "Number of positive samples", x = "Study week", fill = "Virus", title = "c") +
+  labs(y = "Number of positive samples", x = "Study week", fill = "Pathogen", title = "c") +
   theme_custom() +
@@
   scale_y_continuous(expand = c(0,0)) +
-  labs(y = "Number of positive samples", x = "Air cleaner", fill = "Virus", title = "b") +
+  labs(y = "Number of positive samples", x = "Air cleaner", fill = "Pathogen", title = "b") +
   theme_custom() +
@@
   scale_y_continuous(expand = expansion(add = c(0,1))) +
-  labs(y = "Number of positive samples", x = "Air cleaner", fill = "Virus", title = "a") +
+  labs(y = "Number of positive samples", x = "Pathogen", fill = "Pathogen", title = "a") +
   theme_custom() +

### 2023-06-22 doc/paper/supplements.tex
Updates from Overleaf

@@
 
-Table~\ref{tab:epi-data-line-list} shows the line list data for respiratory cases based on epidemiological data.
+\Cref{tab:epi-data-line-list} shows the line list data for respiratory cases based on epidemiological data.
 For each class and case, it shows the date when the student was absent, s/he reported symptoms, and returned to school.
@@
 
-Table~\ref{tab:mol-data-line-list} shows the line list data for the laboratory test results from human saliva samples.
+\Cref{tab:mol-data-line-list} shows the line list data for the laboratory test results from human saliva samples.
 For each class and test, it shows the date when the test was taken, the test result and (if positive) the pathogen that was detected.
@@
     \log I_{jt} &= \log F_{jt} - \log N_{jt} + \beta_0 + \beta_1 \cdot \text{AirCleaner}_{jt} + \beta_2 \cdot \text{Class}_t + \beta_3 \cdot \text{Students}_{jt} \\
-    &\hphantom{= }+ \beta_4 \cdot \text{Ventilation}_{jt} + \beta_5 \cdot \text{CoV}_t + \beta_6 \cdot \text{ILI}_t,
+    &\hphantom{= }+ \beta_4 \cdot \text{AirChangeRate}_{jt} + \beta_5 \cdot \text{CoV}_t + \beta_6 \cdot \text{ILI}_t,
 \end{align*}
@@
 \begin{align*}
-    \Tilde{\mu}_{k} &= \mu_k \cdot \nu_k \cdot \left(\frac{\sum_k \mu_k}{\sum_k \nu_k \cdot \mu_k}\right)\\
-    \sum_k \nu_k &= 1,
+    \Tilde{\mu}_{k} &= \mu_k \cdot \vartheta_k \cdot \left(\frac{\sum_k \mu_k}{\sum_k \nu_k \cdot \mu_k}\right)\\
+    \sum_k \vartheta_k &= 1,
 \end{align*}
@@
 \begin{align*}
-    \nu &\sim \text{Dirichlet}(c) \\
-    c_k &= \sum_j \sum_t C_{jt} \mathbb{I}_{\text{Weekday}(t)=k}
+    \vartheta &\sim \text{Dirichlet}(c) \\
+    c_k &= \sum_j \sum_t C_{jt} \mathbb{I}_{\text{Weekday}(t)=k} \quad \forall k
 \end{align*}
@@
 
+Molecular analysis determined which saliva samples were positive for a respiratory pathogen.
+Let $p = 0, 1, \dots, P$ denote the pathogen, where $0$ refers to negative samples and $P$ is the number of different pathogens detected over the study.
+The number of positive bioaerosol and saliva samples $y_p$ in samples of sizes $n$ is analyzed with a Multinomial logistic regression model
+\begin{align*}
+    y_0, y_1, \cdots, y_P | n, \beta_0, \beta_1, \beta_2, \beta_3 &\sim \text{Multinomial-Logit}(\theta_0, \theta_1, \dots \theta_P) \\
+    \theta &= \text{softmax}(\mu) \\
+    \mu &= \begin{cases}
+                0, & \text{if } p=0, \\
+                \beta_{0p} + \beta_{1} \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Susceptibles}_p, & \text{if } p>0.
+            \end{cases} \\
+    \beta_{0p} &\sim \text{Normal}(0, 2.5) \quad \forall p \\
+    \beta_1, \beta_2, \beta_3 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) 
+\end{align*}
+where $\text{softmax}(\mu) = \exp(\mu)/\sum\exp(\mu)$ and $s_{x}$ is the empirical standard deviation of each input variable.
+The negative test is set as the reference category.
+Overall variation in positive samples by pathogen will be modeled with $\beta_{0p}$ and the effect of air cleaners will be modeled with $\beta_1$.
+The effect of air cleaners is adjusted for class-specific effects and the decreasing number of susceptibles over time (for each pathogen computed as the number of students minus the total number of positive samples).
+
+Note that we deviated from the statistical analysis plan by estimating $\beta_{0p}$ and $\beta_1$ without a hierarchical prior.
+First, there was considerable variation in the overall positivity rate of each pathogen, so that we decided to use unpooled intercepts $\beta_{0p}$ instead of partially pooled intercepts $\beta_{0[p]}$ for each pathogen.
+Second, there was insufficient variation in the data to inform partially pooled estimates $\beta_{1[p]}$ for the effects of air cleaners.
+Therefore, we decided to only estimate the average effect across pathogens, \ie the completely pooled estimate $\beta_1$.
+
 \section{Modeling changes in particle concentrations}\label{sec:env-regression-model}
 
-The change in aerosol number concentration $CN_{jt}$ in class $j$ at day $t$ (analogously for particle mass concentrations $PM_{jt}$) is estimated using Bayesian log-linear regression models, \ie
-\begin{align}
-    \log\,CN_{jt} = \alpha + \beta\,\textrm{School}_j + \bm{\gamma}\,\textrm{Weekday}_t + \theta_M\,M_{jt} + \theta_A\,A_{jt} + \zeta\,\log\,\textrm{N}_{jt} + \omega\,\log\,\textrm{AER}_{jt}~, 
-\end{align}
-where $\alpha$ is the log of the aerosol concentration without interventions, $\beta$ and $\bm{\gamma}$ are school and weekday effects, respectively, $\theta_M$ and $\theta_A$ are the effects of mask mandates ($M_{jt}$) and air cleaners ($A_{jt}$), respectively, $\zeta$ adjusts for the number of students in class ($\textrm{N}_{jt}$), and $\omega$ adjusts for the outdoor air exchange rate ($\textrm{AER}_{jt}$).
-The latter is computed from measured indoor CO$_2$ levels (see Section~\ref{sec:rav-computation}).
-The percent reduction in particle concentrations with interventions are quantified as $100 \times (\mathrm{e}^{-\theta} - 1)\,\%$.
+We only analyze particle concentration data of the time periods during which students were in the classroom.
+Particle concentrations $Y$ on day $t$ in class $j$ are summarized daily with the mean across these time periods.
+The change in aerosol number and particle matter mass concentrations is then estimated using Bayesian log-linear regression models
+\begin{align*}
+    \log Y | \beta_0, \beta_1, \dots, \beta_6 &\sim \text{Normal}(\mu,\sigma) \\
+    \mu &= \beta_0 + \beta_1 \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Weekday} + \beta_4 \cdot \text{Students} \\
+    &\hphantom{= } + \beta_5 \cdot \text{AirChangeRate} + \beta_6 \cdot \text{Cases} \\
+    \beta_0 &\sim \text{Student-t}_5(\mu_y, 2.5s_{y}) \\
+    \beta_1, \dots, \beta_6 &\sim \text{Student-t}_5\left(0, 2.5\frac{s_{y}}{s_{x}}\right) \\
+    \sigma &\sim \text{Exponential}\left(\frac{1}{s_{y}}\right)~.
+\end{align*}
+The effect of air cleaners is adjusted for class- and weekday-specific effects, the number of students in school, the air change rate, and the cumulative number of cases related to respiratory infections.
+A log transform is applied to all continuous input variables.
 
@@
 
-The estimate for $\tau$ indicates that there is variation in the effects of mask mandates between classes.
-However, the credible intervals of $\theta_j^M$ all include zero, indicating that there is only mild deviation of the class-specific estimates from their cross-class average estimate $\theta^M$.
+The estimate for $\beta_2$ is positive, suggesting a higher risk of infection in class $B$, which is in line with the higher number of cases in this class.
+The estimate for $\beta_3$ is positive, suggesting a higher risk of infection as more students were in class.
+The estimate for $\beta_4$ is negative, suggesting a lower risk of infection with higher air change rates, in line with the intuition that the risk of infection is lower when the classroom is better ventilated.
+The estimate for $\beta_5$ and $\beta_6$ are both positive although only the former is distinguishable from zero, suggesting that higher transmission in the community was also associated with a higher risk of infection in the school.
 
-The mean estimate for the proportion of students being absent is negative, indicating that higher proportion of absences decrease the probability of getting infected.
-In contrast to that, the estimate for the reproduction number in the community is positive, indicating that higher community transmission increases the probability of getting infected.
-Both estimates are in line with our hypothesized effect, although it should be noted that both estimates have large credible intervals including zero.
-
-The overdispersion parameter ($\phi$) cannot be precisely estimated, but the credible intervals indicate rather small overdispersion ($\phi \rightarrow \infty$).
-
-The estimates for the logit of the probability of getting infected during school days without interventions ($\alpha$) are larger than the prior mean.
-This is due to the reduction in transmission from mask mandates.
-The weekend effect ($\omega$) is negative, but smaller than the mean of our informative prior, indicating that the likelihood suggests more comparable transmission on weekends.
-
-The posterior distribution of the parameters of the incubation period ($\mu^{p_{IN}}$ and $\sigma^{p_{IN}}$) are close to their prior, indicating that the data is not informative about the incubation period.
+The overdispersion parameter ($\phi$) cannot be precisely estimated, but the credible interval suggests rather small overdispersion ($\phi \rightarrow \infty$).
+The school-free effect $\omega$, the weekday weights $\vartheta$, and the parameters for the distribution of the incubation period $\mu_{p_{\text{IN}}}$ and $\sigma_{p_{\text{IN}}}$ are all close to their priors, suggest that these parameters are barely informed by date but more by our prior assumptions.
 
 \begin{table}[!htpb]
-    \caption[Estimation results from transmission model]{Estimation results from transmission models across the 100 generated datasets for the number of new cases of COVID-19.}
+    \caption[Estimation results from infection risk model]{Estimation results from infection risk model for the number of new respiratory cases.}
     \label{tab:estimation-results}
@@
 It thus reflects overall trends in transmission during study conditions rather than day-to-day variation.
-To evaluate how well our model fits these trends, Fig.~\ref{fig:coverage} compares the estimated number of cases from the probabilistic simulation with the estimated number of new cases from the transmission model.
-Overall the estimates are in good agreement.
-The 95\%-CrI of the model-based estimates includes the 95\%-quantile of the simulation-based estimates.
+To evaluate how well our model fits these trends, \Cref{fig:coverage} compares the model-estimated expected number of cases  with the observed number cases only on a weekly basis and across classes.
+Overall the estimates are in relatively good agreement.
+The 95\%-CrIs of the model-estimates always include the observed cases.
 
@@
     \includegraphics{../../results/epi-data/model-fit.pdf}
-    \caption[Model- and simulation-based estimates of the number of COVID-19 cases]{Estimated (posterior mean as dot and 95\%-CrI as line) number of cases from our probabilistic simulation (blue) and transmission model (black).}
+    \caption[Comparison of model-estimates with observed case data]{Model-estimated number of new respiratory cases (posterior mean as red line and 50\%-, 80\%, and 95\%-CrIs as shaded areas) and the observed number of new respiratory cases (black line) across classes by study week.}
     \label{fig:coverage}
@@
 
+There can be unknown delay between when students were infected and when they tested positive.
+Therefore, we performed a sensitivity analysis where we lagged the molecular test results (number of positive saliva samples) by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday).
+The results are shown in \Cref{fig:mol-estimation-results-sensitivity}.
+The estimated effect of air cleaners remains indistinguishable from zero regardless of the considered lag.
+
 \begin{figure}[!htb]
@@
     \includegraphics[width=10cm]{../../results/mol-data/model-results.pdf}
-    \caption[Boxplot of environmental variables by school and study condition]{Boxplot for the daily average values of each environmental variable by school and study condition.}
+    \caption[Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples]{Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples.
+Shown is the adjusted relative risk ratio (posterior mean as dot and 50\%-, 80\%, and 95\%-CrIs as lines) for air cleaners by lag of the molecular test results.}
     \label{fig:mol-estimation-results-sensitivity}
@@
 
-In the main paper, we showed results for changes in particle concentrations by study condition and summarized them across schools.
-In Fig.~\ref{fig:palas-supp-descriptives}, we present additional results disaggregated by schools and including other environmental variables.
-Furthermore, numerical estimation results for the reduction in particle concentrations with interventions are shown in Table~\ref{tab:palas-est-results}
+In the main paper, we compared particle concentrations between study conditions.
+\Cref{fig:env-descriptives-other-vars} compares additional environmental variables, \ie CO$_2$, air change rate, relative humidity, and temperature.
+There are no considerable differences between in any of the variables between study conditions.
+Furthermore, numerical estimation results for the reduction in particle concentrations with air cleaners are shown in \Cref{tab:palas-est-results}.
 

### 2023-06-22 analysis/env_analysis.Rmd
minor update result plots

@@
           eps_file = "../results/env-data/estimation-results-figure.eps",
-          w = 16, h = 4)
+          w = 16, h = 5)
 ```

### 2023-06-22 analysis/epi_analysis.Rmd
minor update result plots

@@
   scale_x_discrete(breaks = c(1:5, 7:10)) +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = seq(0, 10, 2)) +
   scale_pattern_manual(values = c(A = "stripe", B = "none")) +

### 2023-06-22 doc/paper/manuscript.tex

@@
     \toprule
-         &  School 1 & School 2 & Total \\ \midrule 
+         &  Class A & Class B & Total \\ \midrule 
         \textbf{Students} & \textbf{52 (58\%)} & \textbf{38 (42\%)} & \textbf{90 (100\%)} \\
@@
         \textbf{Absent person-days} & \textbf{334 (52\%)} & \textbf{310 (48\%)} & \textbf{644 (100\%)} \\
-        $\drsh$ Isolation & 109 (33\%) & 38 (12\%) & 147 (\hphantom{0}23\%) \\
         $\drsh$ Sickness & 95 (28\%) & 152 (49\%) & 247 (\hphantom{0}38\%) \\
-        $\drsh$ Quarantine & 55 (17\%) & 5 (\hphantom{0}2\%) & 60 (\hphantom{00}9\%) \\
         $\drsh$ Other & 75 (22\%) & 115 (37\%) & 190 (\hphantom{0}30\%) \\
         & \\
-        \textbf{COVID-19 cases} & \textbf{78 (19\%)} & \textbf{30 (13\%)} &  \textbf{108 (100\%)} \\
-        $\drsh$ Confirmed & 25 (32\%) & 10 (\hphantom{0}33\%) & 35 (\hphantom{0}32\%) \\
-        $\drsh$ Suspected & 53 (68\%) & 20 (\hphantom{0}67\%) & 73 (\hphantom{0}68\%) \\
+        \textbf{Respiratory cases} & \textbf{78 (19\%)} & \textbf{30 (13\%)} &  \textbf{108 (100\%)} \\
+        $\drsh$ Epidemiological data & 25 (32\%) & 10 (\hphantom{0}33\%) & 35 (\hphantom{0}32\%) \\
+        $\drsh$ Molecular data & 53 (68\%) & 20 (\hphantom{0}67\%) & 73 (\hphantom{0}68\%) \\
         \bottomrule
@@
 
-\begin{table}[!htpb]
-    \centering
-    \caption{\textbf{Analysis of molecular data}. Number of positive and negative saliva and airborne samples in each school.}
-    \label{tab:molecular-data}
-    \footnotesize
-    \begin{tabular}{l rlrl p{.2cm} rlrl p{.2cm} rlrl}
-    \toprule
-    & \multicolumn{4}{c}{Saliva} & & \multicolumn{4}{c}{Air} & & \multicolumn{4}{c}{Air cleaner (HEPA filter)$^\dag$} \\ \cline{2-5} \cline{7-10} \cline{12-15} 
-    & \multicolumn{2}{c}{School 1} & \multicolumn{2}{c}{School 2} & &  \multicolumn{2}{c}{School 1} & \multicolumn{2}{c}{School 2} & & \multicolumn{2}{c}{School 1} & \multicolumn{2}{c}{School 2} \\
-    \midrule
-    \textbf{Total} & \textbf{173} & & \textbf{89} & & & \textbf{68} & & \textbf{62} & & &  & &  & \\
-    Negative & 160 & (\hphantom{0}92\%) & 81 & (\hphantom{0}91\%) & & 64 & (\hphantom{0}94\%) & 56 & (\hphantom{0}90\%) & &  & &  & \\  
-    Positive & 13 & (\hphantom{00}8\%) & 8 & (\hphantom{00}9\%) & & 4 & (\hphantom{00}6\%) & 6 & (\hphantom{0}10\%) & & & &  & \\
-    \midrule
-    \textbf{Positive} & \textbf{13} & & \textbf{8} &  & & \textbf{4} & & \textbf{6} & & & \multicolumn{2}{l}{\textbf{2}}  & \multicolumn{2}{l}{\textbf{6}} \\
-    SARS-CoV-2 & 12 & (\hphantom{0}92\%) & 7 & (\hphantom{0}88\%) & & 4 & (100\%) & 5 & (\hphantom{0}83\%) & & 2 & (100\%) & 4 & (\hphantom{0}66\%) \\
-    Influenza & 1 & (\hphantom{00}8\%) & 0 & (\hphantom{00}0\%) & & 0 & (\hphantom{00}0\%) & 0 & (\hphantom{00}0\%) & & 0 & (\hphantom{00}0\%) & 1 & (\hphantom{0}17\%) \\
-    Adeno & 0 & (\hphantom{00}0\%) & 1 & (\hphantom{0}12\%) & & 0 & (\hphantom{00}0\%) & 1 & (\hphantom{0}17\%) & & 0 & (\hphantom{00}0\%) & 1 & (\hphantom{0}17\%) \\
-    \bottomrule
-    \multicolumn{15}{l}{\scriptsize \emph{$\dag$: 20 swabs were taken from each filter at the end of the study period.}}
-    \end{tabular}
-\end{table}
-
 \begin{figure}[!htpb]
     \centering
-    \includegraphics{figures/molecular-descriptives.pdf}
+    \includegraphics{../../results/mol-data/descriptives.pdf}
     \caption{\textbf{Analysis of molecular data and comparison by study condition}.
@@
 \centering
-    \includegraphics[width=\linewidth]{figures/palas-descriptives.pdf}
-    \includegraphics[width=\linewidth]{figures/palas-results.pdf}
+    \includegraphics[width=\linewidth]{../../results/env-data/particles-boxplot.pdf}
+    \includegraphics[width=\linewidth]{../../results/env-data/estimation-results-figure.pdf}
     \caption{\textbf{Analysis of particle concentrations and comparison by study condition}.
@@
 \begin{figure}[!htpb]
-    \includegraphics[width=\linewidth]{figures/redcap-descriptives.pdf} 
-    \includegraphics[width=\linewidth]{figures/redcap-results.pdf}
+    \includegraphics[width=\linewidth]{../../results/epi-data/cases_by_week.pdf} 
+    \includegraphics[width=\linewidth]{../../results/epi-data/avoided-infections.pdf}
     \caption{\textbf{Analysis of epidemiological data and estimated transmission by study condition using Bayesian hierarchical transmission model}.

### 2023-06-28 analysis/env_analysis.Rmd
update results

@@
 ```{r}
+# overall
+env_epi_data %>%
+  summarize(across(c(CO2, AER, Cn, PM1), .fns = list(mean = mean, sd = sd), .names = "{.col}_{.fn}")) %>%
+  mutate(across(c(CO2_mean, CO2_sd, Cn_mean, Cn_sd), round),
+         across(c(AER_mean, AER_sd, PM1_mean, PM1_sd), round, 2))
+
+# by air cleaner
 env_epi_data %>%
   group_by(aircleaner) %>%
-  summarize(across(c(CO2, AER, Cn), .fns = list(mean = mean, sd = sd), .names = "{.col}_{.fn}")) %>%
+  summarize(across(c(CO2, AER, Cn, PM1), .fns = list(mean = mean, sd = sd), .names = "{.col}_{.fn}")) %>%
   ungroup() %>%
   mutate(across(c(CO2_mean, CO2_sd, Cn_mean, Cn_sd), round),
-         across(c(AER_mean, AER_sd), round, 2))
+         across(c(AER_mean, AER_sd, PM1_mean, PM1_sd), round, 2))
 ```
@@
   theme(axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
+        axis.text.y = element_text(color = "white"),
         axis.ticks.y = element_blank(),

### 2023-06-28 analysis/epi_analysis.Rmd
update results

--- 2023-06-22 analysis/epi_analysis.Rmd
@@
 ## Descriptives
+
+### Baseline data
+
+```{r}
+# absent person days
+absent_pers_days <- epi_data_ll %>%
+  mutate(absent_days = as.numeric(date_back - date_absent),
+         reason = ifelse(as.character(reason) == "unknown", "other", as.character(reason))) %>%
+  group_by(class, reason) %>%
+  summarize(n = sum(absent_days)) %>%
+  ungroup() %>%
+  group_by(class) %>%
+  mutate(p = n / sum(n)) %>%
+  ungroup()
+absent_pers_days <- absent_pers_days %>%
+  add_row(class = "A", reason = "total", n = sum(absent_pers_days$n[absent_pers_days$class=="A"]),
+          p = sum(absent_pers_days$n[absent_pers_days$class=="A"]) / sum(absent_pers_days$n)) %>%
+  add_row(class = "B", reason = "total", n = sum(absent_pers_days$n[absent_pers_days$class=="B"]),
+          p = sum(absent_pers_days$n[absent_pers_days$class=="B"]) / sum(absent_pers_days$n)) 
+absent_pers_days <- absent_pers_days %>%
+  add_row(class = "Total", reason = "total", n = sum(absent_pers_days$n[absent_pers_days$reason=="total"]),
+          p = 1) %>%
+  add_row(class = "Total", reason = "sickness", n = sum(absent_pers_days$n[absent_pers_days$reason=="sickness"]),
+          p = sum(absent_pers_days$n[absent_pers_days$reason=="sickness"])  / sum(absent_pers_days$n[absent_pers_days$reason=="total"])) %>%
+  add_row(class = "Total", reason = "other", n = sum(absent_pers_days$n[absent_pers_days$reason=="other"]),
+          p = sum(absent_pers_days$n[absent_pers_days$reason=="other"])  / sum(absent_pers_days$n[absent_pers_days$reason=="total"])) %>%
+  mutate(p = round(100 * p)) %>%
+  mutate(class = factor(class, levels = c("A", "B", "Total")),
+         reason = factor(reason, levels = c("total", "sickness", "other"))) %>%
+  arrange(class, reason) 
+absent_pers_days
+
+# total absent
+round(absent_pers_days$n[absent_pers_days$class=="Total"] / (7 * 5 * 20 + 7 * 5 * 18) * 100)
+
+# epidemiological cases
+epi_data %>%
+  group_by(class) %>%
+  summarize(n = sum(new_cases)) %>%
+  ungroup() %>%
+  mutate(p = n / sum(n) * 100 %>% round) %>%
+  add_row(class = "Total", n = sum(.$n), p = 100)
+
+# molecular cases
+mol_data %>%
+  mutate(Pos = N - Neg) %>%
+  group_by(class) %>%
+  summarize(n = sum(Pos)) %>%
+  ungroup() %>%
+  mutate(p = n / sum(n) * 100 %>% round) %>%
+  add_row(class = "Total", n = sum(.$n), p = 100)
+```
 
@@
   filter(week != 6) %>%
+  mutate(week = ifelse(week < 6, week - min(week) + 1, week - min(week))) %>%
   group_by(class, week) %>%
@@
   geom_bar_pattern(stat = "identity", color = "black", pattern_fill = "white", pattern_angle = 45,
-                   pattern_density = 0.25, pattern_spacing = 0.05) +
+                   pattern_density = 0.25, pattern_spacing = 0.05, width = .5) +
   labs(x = "Study week", y = "Number of cases", fill = "Air cleaner", pattern = "Class", title = "a") +
   scale_fill_manual(values = rev(wes_palette("Royal1")[1:2])) +
-  scale_x_discrete(breaks = c(1:5, 7:10)) +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
+  scale_x_discrete(breaks = 1:7) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = seq(0, 10, 2)) +
   scale_pattern_manual(values = c(A = "stripe", B = "none")) +
@@
 types <- c("Air cleaners partially installed",
-           "Air cleaners never installed",
-           "Air cleaners always installed")
+           "Scenario 2: Air cleaners not installed",
+           "Scenario 1: Air cleaners installed throughout the study")
 
@@
   scale_y_continuous(expand = expansion(mult = c(0.25, 0.25))) +
-  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0))) +
+  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0)), breaks = seq(0, 200, 25)) +
   labs(x = "Estimated number of infections", fill = "CrI", title = "b") +
@@
   filter(week != 6) %>%
+  ungroup() %>%
+  mutate(week = ifelse(week < 6, week - min(week) + 1, week - min(week))) %>%
   group_by(week, .draw) %>%
@@
   scale_x_continuous(expand = c(0,0)) +
-  scale_y_continuous(expand = c(0,0)) +
+  scale_y_continuous(expand = c(0,0), limits = c(0, NA), breaks = scales::breaks_width(2)) +
   scale_color_manual(values = "black") +
@@
 save_plot(fitted_cases_pl, pdf_file = "../results/epi-data/model-fit.pdf",
-          w = 12, h = 10)
+          w = 16, h = 10)
 ```

### 2023-06-28 analysis/mol_analysis.Rmd
update results

@@
 
-pos_week_pl <- ggplot(mapping = aes(x = week, y = n, fill = variable)) +
-  geom_bar(data = df_pos, stat = "identity") +
-  geom_text(data = subset(df_pos, n > 0), mapping = aes(label = n), 
+pos_week_pl <- ggplot(mapping = aes(x = week, y = n)) +
+  geom_bar(data = df_pos %>% mutate(class = paste("Class", class)), mapping = aes(fill = variable),
+           stat = "identity") +
+  geom_text(data = subset(df_pos, n > 0) %>% mutate(class = paste("Class", class)), mapping = aes(label = n), 
             size = 8 / cm(1), position = position_stack(vjust = 0.5), color = "white") +
   facet_wrap(~ class) +
+  geom_text(
+    data = data.frame(class = c("Class A", "Class B"), axis_title = c("Study week", "Study week")),
+    mapping = aes(label = axis_title), hjust = 0.5, vjust = 3.5,
+    x = min(df_pos$week) + diff(range(df_pos$week)) / 2,
+    y = -Inf, color = 'black', size = 8 / cm(1)
+  ) +
+
   scale_fill_manual(values = wes_palette("IsleofDogs1")) +
-  scale_y_continuous(expand = expansion(add = c(0,1))) +
+  scale_y_continuous(expand = expansion(add = c(0,1)), breaks = scales::breaks_width(2)) +
   scale_x_continuous(breaks = seq(1, 7, 1)) +
@@
   theme_custom() +
+  coord_cartesian(clip = "off") +
   theme(panel.grid = element_blank(),
+        panel.spacing.x = unit(2, "cm"),
+        axis.title.x = element_text(color = "white"),
         legend.position = "none") +

### 2023-06-28 utils/plotting.R

@@ -30,6 +30,7 @@ theme_custom <- function() {
           panel.border = element_blank(),
           axis.line.x = element_line(),
           axis.line.y = element_line(),
+          axis.ticks =  element_line(),
           legend.text = element_text(size = 8))
 }
 

### 2023-06-28 doc/paper/manuscript.tex
Updates from Overleaf

@@
 % word count
-\newcommand{\maincount}[1]{%
-  \immediate\write18{texcount -1 -sum=1 -merge -q -nobib #1.tex > #1-words.sum}%
-  \input{#1-words.sum}%
-}
+% \newcommand{\maincount}[1]{%
+%   \immediate\write18{texcount -1 -sum=1 -merge -q -nobib #1.tex > #1-words.sum}%
+%   \input{#1-words.sum}%
+% }
 
-\newcommand{\abstractcount}[1]{%
-  \immediate\write18{texcount -template="{abst}" #1.tex > #1-words.sum}%
-  \input{#1-words.sum}%
-}
+% \newcommand{\abstractcount}[1]{%
+%   \immediate\write18{texcount -template="{abst}" #1.tex > #1-words.sum}%
+%   \input{#1-words.sum}%
+% }
 
@@
 
-\title{\bfseries\LARGE\singlespacing{Transmission of airborne respiratory infections with and without air cleaners in a Swiss school during non-pandemic conditions: A modeling study of epidemiological, environmental, and molecular data}}
+\title{\bfseries\LARGE\singlespacing{Transmission of airborne respiratory infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}}
 
@@
 \author[3]{Lavinia Furrer}
+\author[1]{Kathrin Zürcher}
 \author[1]{Matthias Egger}
@@
 
-\affil[*]{Corresponding author: lukas.fenner@ispm.unibe.ch }
+\affil[*]{Corresponding author: lukas.fenner@unibe.ch }
 
@@
 
-\begin{information}\normalfont
-\noindent\textbf{Running head}: SARS-COV-2 transmission in schools and effect of air cleaners
+% \begin{information}\normalfont
+% \noindent\textbf{Running head}: SARS-COV-2 transmission in schools and effect of air cleaners
 
-%\noindent\textbf{Subject categorization}: 6.20  Indoor Air; 10.11 Pediatrics: Respiratory Infections
+% %\noindent\textbf{Subject categorization}: 6.20  Indoor Air; 10.11 Pediatrics: Respiratory Infections
 
-\noindent\textbf{Word count}: \maincount{manuscript}words, abstract \abstractcount{manuscript}words (max. 500), title 157 chars (max. 200)
+% \noindent\textbf{Word count}: \maincount{manuscript}words, abstract \abstractcount{manuscript}words (max. 500), title 157 chars (max. 200)
 
-%\noindent\textbf{Inserts:} 2 tables, 6 figures, 44 references
+% %\noindent\textbf{Inserts:} 2 tables, 6 figures, 44 references
 
-\noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.
+% \noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.
 
-\vspace{1em}
+% \vspace{1em}
 
-\noindent\textbf{Funding}
+% \noindent\textbf{Funding}
 
-\noindent This study is funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
+% \noindent This study is funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
 NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
@@
 
-\noindent\textbf{Contributions}
+% \noindent\textbf{Contributions}
 
-\noindent Conception and design: NB, LF.
+% \noindent Conception and design: NB, LF.
 Epidemiological and environmental data collection: NB, PJ, TS, LF.
@@
 
-\par
-\end{information}
+% \par
+% \end{information}
 
@@
 
-\noindent\textbf{Methods and Findings:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), and molecular data (bioaerosol and saliva samples) over seven weeks from 16 January to 11 March 2023 in a secondary school (n=38 students across two classes) in Switzerland.
+\noindent\textbf{Methods and Findings:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), and molecular data (bioaerosol and saliva samples) over seven weeks from 16 January to 11 March 2023 in a secondary school (n=38 students in two classes) in Switzerland.
 We performed laboratory and molecular analyses to detect a panel of 19 respiratory viruses.
@@
 \\
-Molecular analysis detected multiple viruses in saliva samples throughout the study (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus), but rarely in bioaerosol samples (2/105 positive: 1~rhinovirus, 1~adenovirus).
-The detection rate of positive samples was comparable between study conditions (adjusted risk ratio 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61).
-Daily average CO$_2$ levels were 1,769$\pm$391\,ppm ($\pm$ standard deviation) with air cleaners and 1,636$\pm$341\,ppm without air cleaners, corresponding to air change rates of 0.74$\pm$0.41\,h$^{-1}$ and 0.68$\pm$0.32\,h$^{-1}$, respectively.
+Molecular analysis detected multiple viruses in saliva samples throughout the study (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus), but rarely in bioaerosol samples (2/105 positive: 1~rhinovirus and 1~adenovirus) and on the filters of air cleaners (4/160~positive: 1~influenza~B, 1~HRV, 1~adenovirus, and 1~SARS-CoV-2).
+The detection rate of positive saliva samples was comparable between study conditions (adjusted risk ratio 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61).
+Daily average CO$_2$ levels were 1,769$\pm$391\,ppm ($\pm$ standard deviation) with air cleaners and 1,636$\pm$341\,ppm without air cleaners, corresponding to similar air change rates (0.74$\pm$0.41\,h$^{-1}$ vs.
+0.68$\pm$0.32\,h$^{-1}$).
 Daily average aerosol number concentrations were 27$\pm$34\,1/cm$^3$ with and 95$\pm$81\,1/cm$^3$ without air cleaners, corresponding to an estimated adjusted decrease of 77\% (95\%-CrI 63\% to 86\%).
-Absences due to respiratory infections were less frequent with air cleaners (13~absences) than without them (21~absences).
-The estimated risk of infection only tended to be lower with air cleaners (adjusted risk ratio 0.72, 95\%-Credible Interval 0.44 to 1.15).
+Absences related to respiratory infections were less frequent with air cleaners in place (13~absences with vs.
+21~absences without air cleaners).
+The estimated risk of infection tended to be lower with air cleaners (adjusted risk ratio 0.72, 95\%-Credible Interval 0.44 to 1.15).
 \\
-Study limitations include unknown specific causes of absences related to respiratory infections (i.e which pathogen caused the sickness).
+% Study limitations include unknown specific causes of absences related to respiratory infections (i.e which pathogen caused the sickness).
 Furthermore, students occasionally had lessons outside the classroom, thus limiting the potential effectiveness of air cleaners.
@@
 
-\noindent\textbf{Conclusions:} Molecular detection showed sustained transmission of multiple respiratory viruses over the study.
-Our results are mixed and, at most, suggest a small benefit of air cleaners in reducing the transmission of airborne respiratory infections in school rooms.
-Our multiple-measurement approach could be used to continuously monitor the risk of transmission of respiratory infections and to identify the most appropriate infection control measures.
+\noindent\textbf{Conclusions:} A wide range of respiratory viruses were detected  in students, but rarely SARS-CoV-2.
+Airborne molecular detection was also rare.
+Our results show a reduction in aerosol concentrations with air cleaners and suggest a small benefit of air cleaners in reducing the transmission of airborne respiratory infections in school rooms.
 
@@
 
-\noindent\textbf{Word count:} \abstractcount{manuscript}words (max. 500)
+%\noindent\textbf{Word count:} \abstractcount{manuscript}words (max. 500)
 
@@
 
-\noindent\textbf{Keywords:} schools, air cleaner, respiratory viruses, molecular detection
+\noindent\textbf{Keywords:} schools, air cleaner, respiratory viruses, airborne transmission, molecular detection
 % maximum of 3-5 keywords
@@
 
-The spread of respiratory infections such as SARS-CoV-2 and influenza is difficult to control\cite{Leung2021}.
-Person-to-person transmission mainly occurs through exhaled respiratory particles containing the respective pathogen, particularly via aerosols (more recently defined as respiratory particles <100$\mu$m\cite{Fennelly2020,Wang2021}), rather than larger respiratory droplets.
-Multiple reports have provided evidence that a considerable part of SARS-CoV-2 transmission is likely to happen through small respiratory particles (<5$\mu$m, also called fine aerosols), allowing for longer suspension times and airborne transmission at short (1-2 m) and at long ranges (>2m)\cite{Coleman2022,Wang2020,Heneghan2021}.
-Growing evidence suggests they contribute importantly to the overall spread of SARS-CoV-2 in indoor congregate settings such as clinics, workplaces, and schools\cite{Greenhalgh2021,Tellier2022,Heneghan2021,Wang2021}.
+The transmission of respiratory infections such as SARS-CoV-2 and influenza presents significant challenges in terms of mitigation and control.
+Person-to-person transmission primarily occurs through the release of respiratory particles containing the respective pathogens.
+The recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
+Unlike larger respiratory droplets that tend to settle quickly, aerosols can remain suspended in the air for several hours and their respiratory plumes can reach more than 2m\cite{Coleman2022,Wang2020,Heneghan2021}, thus exceeding the recommended rules for social distancing by public health organizations\cite{Bazant2021PNAS,Trivedi2021PhyFluids,Lindsley2010CID}.
 
-Public authorities worldwide closed businesses and schools during the COVID-19 pandemic\cite{Banholzer2022EJE,Banholzer2021PLOS}, but the closure of schools was particularly contentious.
-While the negative impact of school closures on student well-being and mental health is well documented\cite{Loades2020}, the role of children and adolescents in transmitting SARS-CoV-2 is less clear\cite{Goldstein2021}.
-A study in Germany estimated that school contacts contributed between 2\% and 20\% to the overall transmission of SARS-CoV-2 in the population\cite{Heinsohn2022}.
-Studies on the effectiveness of government interventions at the population level are inconclusive regarding the effects of school closures in different epidemic waves\cite{Banholzer2021PLOS,Sharma2021}.
-The introduction of compulsory face mask wearing\cite{Gettings2021,Heinsohn2022} and improved ventilation\cite{Gettings2021} in schools was associated with a lower incidence of COVID-19.
-In addition, the installation of portable HEPA-air filtration devices (air cleaners) was shown to remove SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}.
-Finally, a simulation study of exhaled SARS-CoV-2 bioaerosols in an indoor space demonstrated the efficacy of mask wearing and portable air cleaners in reducing aerosols\cite{Lindsley2021}.
+The generation, spread, and concentration of bioareosols can be influenced with infection control measures.
+For example, compulsory face mask wearing reduced the transmission of SARS-CoV-2 during the COVID-19 pandemic\cite{Gettings2021,Heinsohn2022,Leech2022PNAS,Mitze2020PNAS}.
+More generally, improved ventilation systems are considered crucial for healthy indoor environments both during pandemic and non-pandemic times\cite{Wang2021,Morawska2021}.
+Portable air-filtration devices (air cleaners) may present another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear.
+Various studies showed that air cleaners reduce the concentration of aerosols and particle matter in indoor air\cite{Park2020Build,Buising2022InfContr,Ren2021Dent,Banholzer2023PLoSMed}.
+These findings suggest potential protection from respiratory infection as long-term exposure to higher aerosol and particle matter mass concentration have been associated with increased risks of respiratory infections\cite{Gordon2014Resp} and, more recently, with the incidence and mortality of COVID-19\cite{Kelly2023Atmos,Veronesi2022OccEnvMed,Travaglio2021EnvPoll,DeAngelis2021EnvRes}.
+Furthermore, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}.
+As a result, studies have simulated the potential of air cleaners in reducing indoor transmission risks of SARS-CoV-2\cite{Lindsley2021} and other respiratory pathogens\cite{Cortellessa2023Build}.
+However, empirical evidence linking the use of air cleaners with a reduced risk of infection is lacking.
+Thus, it remains unclear whether the reduction in particle concentrations and the removal of bioaerosols from the air translates into lower transmission of respiratory infections.
 
-We used a multiple-measurement approach to study the transmission of SARS-CoV-2 and other respiratory viruses in school rooms.
-We collected epidemiological (cases of respiratory diseases), environmental (CO$_2$, aerosol and particle concentrations), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) over a seven-week study period from January to March 2022 in two secondary schools in Switzerland.
-We analyzed changes in environmental and molecular characteristics and estimated the probability of infection with SARS-CoV-2 during three different study conditions with and without infection control measures (general mask wearing and air cleaners).
+We used a multiple-measurement approach to study the transmission of respiratory viruses in school rooms under non-pandemic conditions in the winter of 2023.
+We collected epidemiological (absences probably related to respiratory infections), environmental (CO$_2$, aerosol and particle concentrations), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) over a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
+We used a cross-over study design to assess the effectiveness of air cleaners as an infection control measure.
+For this, we estimated changes in particle concentrations, the positivity rate for respiratory viruses, and the risk of infection.
 
@@
 
-\noindent We collected data in two classrooms of a secondary school (age of students 13-15~years, 700 students in total) in the Canton of Solothurn, Switzerland, over a seven-week study period from 26~January to 11~March 2023.
-Fig.~\ref{fig:study-setup} shows the schematic study setup and Table~\ref{tab:data} provides an overview of the types of data that were collected in each classroom.
+\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the Canton of Solothurn, Switzerland, over a seven-week study period from 26~January to 11~March 2023.
+\Cref{fig:study-setup} shows the schematic study setup and \Cref{tab:data} provides an overview of the types of data that were collected in each classroom.
 
@@
     \centering
-    \caption{\textbf{Data}. Type of data collected, method/device, and frequency in the rooms of classes A and B.}
+    \caption{\textbf{Data collection}. Type of data collected, method/device, and frequency in the rooms of classes A and B.}
     \begin{tabular}{p{3.5cm}p{6cm} p{1cm} p{1cm} p{3cm}}
@@
 
-\noindent We used a cross-over design to study the effectiveness of air cleaners, thereby distinguishing two study conditions in each class: weeks with and weeks without installation of air cleaners in the classroom (Fig.~\ref{tab:study-design}).
+\noindent We used a cross-over design to study the effectiveness of air cleaners, thereby distinguishing two study conditions in each class: weeks with and weeks without air cleaners being installed in each classroom (\Cref{tab:study_design}).
 Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China; approx.
-USD\,250 per device), running at 2$\,\times\,$600 m$^{3}$/h clean air delivery rate.
-Passive window ventilation occurred per recommendations by the national public health authorities during both study conditions.
+USD\,250 per device).
+According to the manufacturer, these air cleaners run at a clean air delivery rate of 2$\,\times\,$600 m$^{3}$/h.
+Based on our own experiments conducted in the classrooms, we measured a lower clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h.
+Window ventilation was done at the discretion of the teachers, as there were no specific recommendations from the national health authorities at that time.
 
@@
     \centering
-    \caption{\textbf{Study design.} Cross-over design where two portable air cleaners (AC) were installed during different time periods in classes A and B over a seven-week study period from 16~January to 11~March 2023, excluding a week of vacation between February 6 and 11.}
+    \caption{\textbf{Study design.} Cross-over design where two portable air cleaners (AC) were installed during different time periods in classes A and B over a seven-week study period from 16~January to 11~March 2023, excluding a week of vacation between February 6 and 11.}\label{tab:study_design}
     \begin{tabular}{l l l l l l l l l}
@@
     \end{tabular}
-    \label{tab:study_design}
 \end{table}
@@
 
-\noindent At baseline, we collected aggregated data on age, sex and COVID-19 vaccination and recovery status in the participating classes.
-Daily we collected data on each absent student, \ie the absence, symptom onset and return date as well as the reason for the absence (sickness, career days,  or other) and symptoms.
+\noindent At the beginning of the study, we collected aggregated data on age, sex and COVID-19 vaccination and recovery status in the participating classes.
+Daily we collected data on each absent student, \ie the absence and return date as well as the reason for the absence (sickness, career days,  or other).
+For sickness-related absences, we collected the symptoms and the symptom onset date.
 We also collected data on laboratory tests but none of the students reported a test result over the study.
 Based on the epidemiological data, we defined a case of respiratory infection as an absence where the student reported a sickness with at least one of the following symptoms: fever, coughing, tiredness, loss of test or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, stomach.
-Students were asked to report the first date when they experienced symptoms, \eg a student absent on Monday may report Saturday or Sunday as the symptom onset date.
-We will denote cases by their symptom onset date.
-Note that, during the week, the symptom onset date will usually correspond to the absence date, unless the student went to school despite experiencing symptoms.
-The line list data is shown in \supp~\zref{sec:case-data}.
-Informed consent of one student could not be obtained and his/her one absence was considered as unrelated to a respiratory infection.
+Students were asked to report the first date when they experienced symptoms, \eg a student absent on Monday can also report Saturday or Sunday as the symptom onset date.
+We will denote respiratory cases by their symptom onset date, which usually corresponded to the absence date, unless the student went to school while having symptoms.
+Informed consent of one student could not be obtained and the sole absence of this student was considered as unrelated to a respiratory infection.
 Reports about absences were entered electronically into a REDcap database\cite{Harris2009,Harris2019}.
-Furthermore, both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below).
+The line list data is shown in Table~\zref{tab:epi-data-line-list} in \supp.
+
+In addition, both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below).
 The saliva samples were transported to the laboratory and stored at $-$80°C until further processed\cite{Galar2021,To2019,Huber2021}.
+The line list data is shown in Table~\zref{tab:mol-data-line-list} in \supp.
 \medskip
@@
 \noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute.
-A transient mass balance model was used to estimate the air change rate from indoor CO$_2$ levels\cite{Batterman2017IJERPH}, assuming that the classrooms were fully occupied during lessons and that room occupancy increased (decreased) linearly 10~minutes before (after) the start (end) of a class.
-Particle concentrations were filtered according to the times students were in the classroom, which were monitored together with the time windows were opened.
+This device has been used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}.
+A transient mass balance model was used to estimate the air change rate from indoor CO$_2$ levels\cite{Batterman2017IJERPH}, assuming that the classrooms were fully occupied during lessons and that room occupancy increased (decreased) linearly 10~minutes before (after) the start (end) of a lesson.
+Particle concentrations were filtered according to the times students were in the classroom, which were monitored together with the times during which windows were opened.
 \medskip
 
-\noindent \emph{Bioaerosol sampling}
+\noindent \emph{Bioaerosol sampling and HEPA filters of the air cleaners}
 
-\noindent We collected airborne respiratory viruses in the classroom with a bioaerosol sampling device (BioSpot-VIVAS, Aerosol Devices Inc., Ft.
+\noindent We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France) , running at 200\,l/min and collecting into 15\,mL PBS, as previously reported in clinic settings\cite{Moore2021}.
+The Coriolis Micro Air only operated shortly before and during break times to reduce noise exposure (approximately 60\,min/day).
+In one class, we also used the BioSpot-VIVAS device sampler (Aerosol Devices Inc., Ft.
 Collins, CO, USA).
-This device samples airborne virus particles into a viral transport medium (VTM) using a laminar flow water-based condensation method.
-In parallel, we also used the Coriolis Micro Air (Bertin Instruments Montigny-le-Bretonneux, France) sampler, running at 200\,l/min and collecting into 15\,mL PBS  as previously reported in clinic settings\cite{Moore2021}.
-BioSpot-VIVAS operated throughout lessons while Coriolis Micro Air only operated shortly before and during break times to reduce noise exposure (approximately 60\,min/day).
+This device samples airborne virus particles into a viral transport medium (VTM) using a laminar flow water-based condensation method\cite{Pan2016JAM,Lednicky2016AST}.
+BioSpot-VIVAS operated throughout lessons.
 The removable parts of both sampling devices were regularly autoclaved to avoid contamination.
 At the end of the day, samples were transported to the Institute of Infectious Diseases (IFIK) and stored at $-$80°C.
-At the end of the study period, the Xiaomi HEPA filters were carefully removed and 20\,swabs were taken from each filter and stored at $-$80°C until further processed.
-\medskip
+After each intervention phase (see table xxx), the HEPA filters were carefully removed from the air cleaners, divided into 20~fields, and one swab per field was taken for a total of 20~swabs per filter.\medskip
 
@@
 The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19~respiratory viruses: SARS-CoV-2, Influenza A virus, Influenza B virus, Human respiratory syncytial virus A/B (RSV), Human metapneumovirus (MPV), Human adenovirus A/B/C/D/E/F (AdV), Human rhinovirus A/B/C (HRV), and Human parainfluenza virus 1/2/3/4 (PIV) in one single reaction.
-Viral genome load (VGL) of specimens was quantified (copies/L) by curve extrapolation derived from standardized ATCC quantitative genomic standard dilutions (ATCC-VR-3347D, LGC Standards, LGC Group UK).
+%/Viral genome load (VGL) of specimens was quantified (copies/L) by curve extrapolation derived from standardized ATCC quantitative genomic standard dilutions (ATCC-VR-3347D, LGC Standards, LGC Group UK).
 For the BioSpot-VIVAS air samples, the air concentration of the DNA was exactly calculated according to the airflow influx into the sampling device.
+%/
 The Limit of detection (LOD) of the assay SARS-CoV-2, IFA/IFB is 7.5x10$^0$ TCID/ml (SARS-CoV-2), 3.7x10$^{-1}$ TCID/ml (IFA), and 4.8x100 TCID/ml (100cp/ml) (IFB) respectively.
-For positive samples, we performed targeted sequencing to compare genetic relatedness between SARS-CoV-2 strains detected in the air and human samples\cite{Goncalves2021}, but we were unable to amplify and sequence any of the gene targets in the bioaerosol samples due to low RNA concentrations.
+For positive samples, we performed targeted sequencing to compare genetic relatedness between SARS-CoV-2 strains detected in the air and human samples\cite{Goncalves2021Microorganisms}, but we were unable to amplify and sequence any of the gene targets in the bioaerosol samples due to low RNA concentrations.
 
@@
 Apart from that there were no major deviations from the SAP regarding the conducted analyses.
-There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in \supp~\zref{sec:transmission-model} to \supp~\zref{sec:env-regression-model}, which provide detailed descriptions of the models used in each analysis.
+There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in Text~\zref{sec:transmission-model} to Text~\zref{sec:env-regression-model} in \supp, which provide detailed descriptions of the models used in each analysis.
 
@@
 
-A detailed description of the model and choice of priors for all model parameters are provided in \supp~\zref{sec:transmission-model}.
+A detailed description of the model and choice of priors for all model parameters are provided in Text~\zref{sec:transmission-model} in \supp.
 Importantly, since multiple pathogens were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the pathogen-specific distributions.
@@
 \noindent Molecular analysis determined which saliva samples were positive for a respiratory pathogen.
-We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~\zref{sec:multinomial-model}) where the expected number of positive samples for a pathogen was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each pathogen.
+We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (Text~\zref{sec:multinomial-model} in \supp) where the expected number of positive samples for a pathogen was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each pathogen.
 The latter was computed as the difference between the number of students per class and the cumulative number of positive samples.
@@
 \noindent We computed the average concentrations per day and compared these between study conditions.
-We estimated the reduction in concentrations using Bayesian log-linear regression models (\supp~\zref{sec:env-regression-model}), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+We estimated the reduction in concentrations using Bayesian log-linear regression models (Text~\zref{sec:env-regression-model} in \supp), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
 \medskip
@@
 
-The study population consisted of 90~students (39~female, 51~male; Table~\ref{tab:cases-overview-school}).
-Of these, 27~students were fully vaccinated and 34~students had recovered from an infection within the last year.
-Over the seven-week study period (3,150 student-days in total), students were absent from school for 644~days (20\% of total) of which 147~days (23\% of absences) were due to isolation related to COVID-19, and 247~(38\% of absences) were due to sickness.
-Overall, there were 35~confirmed and 73~suspected cases of COVID-19, exceeding the number of students in some classes (Table~\zref{tab:cases-overview}).
-This suggests that only a proportion of suspected cases were actual cases of COVID-19 as it is unlikely that students were infected twice.
-We estimated 55 (95\%-CrI 35 to 76) cases of COVID-19 across schools.
-\deleted{(\supp~Fig.~\zref{fig:redcap-supp-total-cases}).}
+The study population consisted of 38~students (19~female, 19~male; Table~\ref{tab:cases-overview-school}).
+Of these, 7~students were vaccinated or recovered from an infection within SARS-CoV-2 in the last four months.
+Over the seven-week study period (1,330 student-days in total), students were absent from school for 220~days (18\% of total) of which 129~days (59\% of absences) were due to sickness.
 
@@
     \centering
-    \caption{Overview of the study population, number of COVID-19 cases, and person-days of absences.}
+    \caption{Overview of the study population and person-days of absences.}
     \label{tab:cases-overview-school}
@@
          &  Class A & Class B & Total \\ \midrule 
-        \textbf{Students} & \textbf{52 (58\%)} & \textbf{38 (42\%)} & \textbf{90 (100\%)} \\
+        \textbf{Students} & \textbf{20 (53\%)} & \textbf{18 (47\%)} & \textbf{38 (100\%)} \\
         \emph{Gender} \\
-        $\drsh$ Female & 20 (38\%) & 19 (50\%) & 39 (\hphantom{0}43\%) \\
-        $\drsh$ Male & 32 (62\%) & 19 (50\%) & 51 (\hphantom{0}57\%) \\
-        \emph{Vaccination status} \\
-        $\drsh$ Vaccinated & 16 (31\%) & 11 (29\%) & 27 (\hphantom{0}30\%) \\
-        $\drsh$ Not vaccinated & 36 (69\%) & 27 (71\%) & 63 (\hphantom{0}70\%) \\
-        \emph{Recovery status} \\
-        $\drsh$ Recovered last year & 25 (48\%) & 9 (24\%) & 34 (\hphantom{0}38\%) \\
-        $\drsh$ Not recovered & 27 (52\%) & 29 (76\%) & 56 (\hphantom{0}62\%) \\
-        & \\
-        \textbf{Absent person-days} & \textbf{334 (52\%)} & \textbf{310 (48\%)} & \textbf{644 (100\%)} \\
-        $\drsh$ Sickness & 95 (28\%) & 152 (49\%) & 247 (\hphantom{0}38\%) \\
-        $\drsh$ Other & 75 (22\%) & 115 (37\%) & 190 (\hphantom{0}30\%) \\
-        & \\
-        \textbf{Respiratory cases} & \textbf{78 (19\%)} & \textbf{30 (13\%)} &  \textbf{108 (100\%)} \\
-        $\drsh$ Epidemiological data & 25 (32\%) & 10 (\hphantom{0}33\%) & 35 (\hphantom{0}32\%) \\
-        $\drsh$ Molecular data & 53 (68\%) & 20 (\hphantom{0}67\%) & 73 (\hphantom{0}68\%) \\
+        $\drsh$ Female & 11 (55\%) & 8 (44\%) & 19 (\hphantom{0}50\%) \\
+        $\drsh$ Male & 9 (45\%) & 10 (56\%) & 19 (\hphantom{0}50\%) \\
+        \emph{Immunity status} \\
+        $\drsh$ Recently vaccinated (or recovered) & 7 (39\%) & 0 (0\%) & 7 (\hphantom{0}18\%) \\
+        $\drsh$ Not recently vaccinated (or recovered) & 11 (61\%) & 20 (100\%) & 31 (\hphantom{0}82\%) \\
+        \textbf{Absent person-days} & \textbf{110 (50\%)} & \textbf{110 (50\%)} & \textbf{220 (100\%)} \\
+        $\drsh$ Sickness & 52 (47\%) & 77 (70\%) & 129 (\hphantom{0}59\%) \\
+        $\drsh$ Other & 58 (53\%) & 33 (30\%) & 91 (\hphantom{0}41\%) \\
         \bottomrule
@@
 
-We analyzed 262 saliva, 130 bioaerosol samples and swabs from the filters of air cleaners (20~swabs per filter) in two classrooms.
-We detected SARS-CoV-2, adenovirus, and influenza virus (Table~\ref{tab:molecular-data} and \supp~Table~\zref{tab:mol-data-1}-\zref{tab:mol-data-2}).
-SARS-CoV-2 made up the vast majority of positive saliva (19 out of 21) and bioaerosol samples (9 out of 10).
-We found four positive air-saliva samples in the same respective week (three SARS-CoV-2 and one adenovirus), suggesting they were paired samples.
-We also detected SARS-CoV-2 and influenza viruses from the HEPA filters of the air cleaners.
-The number of positive saliva and airborne SARS-CoV-2 samples per week varied by study condition (Fig.~\ref{fig:molecular-descriptives}a).
-Without interventions, the proportion of positive samples per week was 11.5\% for saliva and 8.1\% for airborne samples.
-These proportions were lower with mask mandates (Saliva: 5.7\%, Air: 7.1\%) and air cleaners (Saliva: 7.7\%, Air: 5.0\%).
-There were also differences in viral concentration in positive airborne samples (Fig.~\ref{fig:molecular-descriptives}b).
-Without interventions, it was 1.1\,copies per liter per week, which was higher than with mask mandates (0.7\,copies/L per week) and air cleaners (0.1\,copies/L per week).
+We analyzed 448~saliva samples over the study across classes.
+We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenovirus (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~virus~(PIV) (\Cref{fig:molecular-descriptives}a).
+There was an equal number of positive samples with and without air cleaners (\Cref{fig:molecular-descriptives}b).
+Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted risk ratio 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (Figure~\zref{fig:mol-estimation-results-sensitivity} in \supp).
+The distribution of positive samples varied between classes (\Cref{fig:molecular-descriptives}c).
+For example, in the first three study weeks all except for one sample were positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
+The pathogen-specific variations between classes are suggestive of transmission occurring within classes.
+
+We also analyzed 105~bioaerosol samples and detected  two respiratory viruses in the air (1~HRV in class~A and 1~AdV in class~B).
+Furthermore, we detected 1~Flu~B, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
 
@@
     \includegraphics{../../results/mol-data/descriptives.pdf}
-    \caption{\textbf{Analysis of molecular data and comparison by study condition}.
-\textbf{(a)}~Proportion of positive airborne (solid line) and saliva (dashed line) samples (average per week).
-\textbf{(b)}~Viral concentration in positive airborne samples from BioSpot-VIVAS (average per week).}
+    \caption{\textbf{Analysis of molecular data}.
+\textbf{(a)}~Number of positive saliva samples by pathogen.
+\textbf{(b)}~Number of positive saliva samples by study condition.
+\textbf{(c)}~Number of positive saliva samples by study week and class.
+Pathogens: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}
     \label{fig:molecular-descriptives}
@@
 
-Particle concentrations differed by study condition (Fig.~\ref{fig:palas-results}a).
-Aerosol number concentrations were lower with mask mandates (49$\pm$52\,1/cm$^3$) and air cleaners (84$\pm$56\,1/cm$^3$) than without interventions (177$\pm$109\,1/cm$^3$).
-Similarly, particle mass concentrations (\eg PM$_1$) were lower with mask mandates (2.0$\pm$1.6\,$\mu$gm$^{-3}$) and air cleaners (3.8$\pm$2.9\,$\mu$gm$^{-3}$) than without interventions (6.9$\pm$4.1\,$\mu$gm$^{-3}$).
-CO$_2$ levels without interventions (953$\pm$198\,ppm) were slightly lower than with mask mandates (1,155$\pm$237\,ppm) and air cleaners (1,088$\pm$224\,ppm), indicating increased ventilation through outdoor air exchange (\supp~Fig.~\zref{fig:palas-supp-descriptives}), although the proportion of times windows were opened did not change with mask mandates (0.03, 95\%-CrI $-$0.07 to 0.12) or air cleaners (0.00, 95\%-CrI $-$0.11 to 0.11).
-When adjusting for different ventilation rates, the number of students in class, school and weekday effects, the aerosol number concentration decreased by 69\% (95\%-CrI 42\% to 86\%) with mask mandates and by 39\% (95\%-CrI 4\% to 69\%) with air cleaners (Fig.~\ref{fig:palas-results}b and Table~\zref{tab:palas-est-results}).
-The concentration of smaller particles (PM$_1$ and PM$_{2.5}$) was more reduced by mask mandates and the concentration of larger particles (PM$_4$ and PM$_{10}$) was more reduced by air cleaners.
+There was a strong difference in particle concentrations between study conditions (\Cref{fig:palas-results}a).
+Aerosol number concentration with air cleaners were markedly lower (27$\pm$34\,1/cm$^3$) than without air cleaners (95$\pm$81\,1/cm$^3$).
+Similarly, particle mass concentrations (\eg PM$_1$) were markedly lower with air cleaners (1.16$\pm$1.33\,$\mu$gm$^{-3}$) than without air cleaners (4.20$\pm$3.74\,$\mu$gm$^{-3}$).
+There was no considerable difference in other environmental variables (Figure~\zref{fig:env-descriptives-other-vars} in \supp).
+In particular, daily average CO$_2$ levels (air change rates) were 1,769$\pm$391\,ppm (0.74$\pm$0.41\,h$^{-1}$) with air cleaners and 1,636$\pm$341\,ppm (0.68$\pm$0.32\,h$^{-1}$), suggesting that ventilation was comparable between study conditions.
+When adjusting for air change rates and multiple other factors, the aerosol number concentration decreased by 76\% (95\%-CrI 63\% to 86\%) with air cleaners (\Cref{fig:palas-results}b and Table~\zref{tab:palas-est-results} in \supp).
+The decrease in the concentration of larger particles (PM$_{10}$) was larger than the decrease in the concentration of smaller particles (PM$_1$ to PM$_{4}$).
 
@@
     \includegraphics[width=\linewidth]{../../results/env-data/estimation-results-figure.pdf}
-    \caption{\textbf{Analysis of particle concentrations and comparison by study condition}.
+    \caption{\textbf{Analysis of particle concentrations and comparison between study conditions}.
 \textbf{(a)}~Boxplot of the daily average values for aerosol number concentration (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$).
-Results for CO$_2$ and other environmental variables are shown in \supp~\zref{sec:detailed-palas}.
-\textbf{(b)}~Estimated reduction in aerosol number and particle mass concentrations with interventions (posterior mean as dot and 50\%-, 80\%- and 95\%-CrI as lines, respectively).
+\textbf{(b)}~Estimated reduction in aerosol number and particle mass concentrations with air cleaners (posterior mean as dot and 50\%-, 80\%- and 95\%-CrI as lines, respectively).
 }
@@
 
-\subsection{Estimating transmission risk of SARS-CoV-2}
+\subsection{Risk of infection}
 
-The cumulative number of cases of COVID-19 increased considerably in all classes without intervention and the large majority of students has been infected in School~1 by the time air cleaners were installed, except for class~D in School~2 (Fig.~\ref{fig:redcap-results}a).
-Based on our Bayesian transmission model, the probability of getting infected was lower with mask mandates than without interventions (adjusted odds ratio 0.19, 95\%-CrI 0.09 to 0.38) and comparable with air cleaners (1.00, 95\%-CrI 0.15 to 6.51).
-Accordingly, mask mandates were associated with a considerable number of avoided infections (9.98, 95\%-CrI 2.16 to 19.00), but not air cleaners (Fig.~\ref{fig:redcap-results}b).
-As an additional analysis, we used a modified Wells-Riley equation\cite{Rudnick2003} and assumed that the change in the rate of emitted infectious quanta were proportional to the \replaced{estimated}{measured} reduction \added{mean and 95\%-CrI) in the aerosol number concentration, while other parameters were kept constant under all study conditions (\supp~\zref{sec:transmission-risk}).
-Accordingly, the daily risk of infection for a six~hour school day was 1.0\% (95\%-CrI 0.4\% to 1.9\%) with mask mandates and 1.9\% (95\%-CrI 1.0\% to 3.0\%) with air cleaners, compared with a 3.1\% risk \added{for the baseline} without interventions (\supp~Fig.~\zref{fig:results-trans-risk}).
+Overall, there were 35~respiratory cases (absences related to respiratory infections) across classes.
+Incidentally, there were more cases in class~B (21) than in class~A (14), which is the class were more students recently recovered from a SARS-CoV-2 infection or were vaccinated.
+Except for one study week, there was an equal or higher number of cases without air cleaners (\Cref{fig:redcap-results}a).
+Based on our Bayesian latent hierarchical regression model, we estimated a trend towards a lower risk of infection with air cleaners (adjusted risk ratio 0.72, 95\%-Credible Interval 0.44 to 1.15).
+As a result, the estimated number of infections if air cleaners had been installed throughout the study is expected to be 36 (95\%-CrI 21 to 64) compared to 62 (95\%-CrI 29 to 141) infections if they would not have been installed (\Cref{fig:redcap-results}b).
+Detailed estimation results are provided in Text~\zref{sec:detailed-redcap} in \supp.
 
@@
     \includegraphics[width=\linewidth]{../../results/epi-data/avoided-infections.pdf}
-    \caption{\textbf{Analysis of epidemiological data and estimated transmission by study condition using Bayesian hierarchical transmission model}.
-\textbf{(a)}~Estimated mean cumulative number of cases for each school class after probabilistic simulation accounting for uncertainty in suspected cases and the date of symptom onset (\supp~\zref{sec:prob-sim}).
-Dotted red lines indicate the number of students per class.
-A comparison of the estimated number of new cases after probabilistic simulation with the observed number of new confirmed and suspected cases is shown in \supp~Fig.\zref{fig:redcap-supp-descriptives}.
-\textbf{(b)}~Estimated number of avoided infections associated with interventions across schools (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) based on Bayesian hierarchical transmission model.
-Detailed estimation results are shown in \supp~\zref{sec:detailed-redcap}.}
+    \caption{\textbf{Analysis of epidemiological data and estimated relative risk of infection with air cleaners.}.
+\textbf{(a)}~Weekly number of respiratory cases with and without air cleaners.
+\textbf{(b)}~Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if air cleaners had always or never been installed in both classes.}
     \label{fig:redcap-results}
@@
 
-We collected epidemiological, environmental, and molecular data to estimate transmission of respiratory infections in schools and assessed the effectiveness of infection control measures.
-Airborne detection of SARS-CoV-2 documented sustained SARS-CoV-2 transmission.
-Bioaerosol SARS-CoV-2 concentrations with general mask mandates were lower, and SARS-CoV-2 was detected on the filters of air cleaners.
-Aerosol number and particle mass concentrations dropped significantly with both interventions.
-The Bayesian transmission model estimated that mask mandated avoided SARS-CoV-2 infections, but not air cleaners.
+% summary
 
-Our study demonstrated airborne detection of SARS-CoV-2 in schools.
-Although sampling and molecular detection of infectious bioaerosols are challenging and there is no agreed standard\cite{Santarpia2020,Morris2022,Stockwell2019}, it provides important evidence on the airborne transmission of respiratory pathogens.
-So far, viral RNA in airborne samples of SARS-CoV-2 was mainly found in hospitals and healthcare facilities\cite{Shen2021}.
-A related study in two South African schools detected airborne \emph{Mycobacterium tuberculosis}\cite{Bunyasi2022}.
-Tuberculosis was the leading cause of death worldwide due to an infectious disease prior to the COVID-19 pandemic.
-An improved understanding of airborne transmission and the effectiveness of interventions can benefit the control of both infectious diseases\cite{Rieder2021}.
-Our study provides evidence on airborne SARS-CoV-2 viral transmission and the potential effects of interventions in schools based on airborne and saliva samples from students.
-In our study, positive samples mostly pertained to SARS-CoV-2, but we occasionally detected other respiratory viruses such as adenovirus and influenza.
-General public health measures during the study likely suppressed the spread of other respiratory viruses.
-It also must be noted that the detected viral concentrations were low, as shown by high cycle thresholds (CTs) in the RT-qPCR results.
-The molecular presence of other viral pathogens cannot be excluded.
-The low sensitivity of molecular assays to detect airborne pathogens is a common challenge\cite{Morris2022,Stockwell2019}.
+We collected epidemiological, environmental, and molecular data to estimate the risk of infection with respiratory infectious diseases in a Swiss school and assessed the effectiveness of air cleaners as an infection control measure.
+We found a wide range of respiratory viruses in human saliva samples, primarily adenovirus, influenza B, and rhinovirus, but only few viruses were detected in bioaerosol samples and on the filters of air cleaners.
+Aerosol number and particle mass concentrations dropped significantly with air cleaners.
+Bayesian modeling showed conflicting results between the analysis of molecular data and absences related to respiratory infections.
+While the positivity rate of human saliva samples was comparable between study conditions, the relative risk of infection based on absences tended to be lower with air cleaners.
 
-An experimental study demonstrated the effectiveness of infection control measures (universal face mask wearing and air cleaners) using simulated exhaled SARS-CoV-2 bioaerosols in a closed indoor space\cite{Lindsley2021}.
-In contrast, we studied infection control measures in a real-life setting and demonstrated their effectiveness using a multiple-measurement approach, and obtain similar results.
-Our findings also align with existing evidence from population-level studies showing that the incidence of COVID-19 was lower schools with mask use\cite{Heinsohn2022,Gettings2021} and improved ventilation\cite{Gettings2021}.
-Similarly, a field study showed that adequate ventilation was associated with reduced incidence of tuberculosis, a strictly airborne disease, in a university building\cite{Du2020}.
-Altogether, these findings support arguments in favor of multiple intervention strategies to address airborne transmission of respiratory infections in crowded indoor settings\cite{Morawska2021}.
+% comparison with our previous study
 
-Indoor ventilation is one of the key determinants of airborne transmission\cite{Fennelly2020,Wang2021}, but schools are often poorly ventilated\cite{Santamouris2008,DiGilio2021}.
-We showed that air cleaners effectively reduce the concentration of both larger and smaller particles, in line with findings about their effectiveness in hospitals\cite{Morris2022} and simulated indoor environments\cite{Lindsley2021}.
-The detection of viruses on the filters of air cleaners further supports the evidence on the effective removal of bioaerosols.
-However, it was difficult to estimate changes in transmission following air cleaners because they were installed at the end of the study period when presumably a large proportion of students were already infected with SARS-CoV-2.
-In addition, physicochemical properties of aerosols, environmental factors and the distance to infectious people determine the survival of airborne particles and the loss of infectivity over time\cite{Wang2021}.
-Thus, a predominance of close range high particle density aerosol transmission of SARS-CoV-2 in school settings could further explain why air cleaners appeared not effective in preventing transmission.
-Our study used portable, affordable air cleaners that could be implemented at scale, rather than large cleaners\cite{Duill2021}.
-Noise exposure, and a lack of acceptance by teachers\cite{Sanguinetti2022} may still prevent their widespread use.
-Although not specifically assessed, we perceived good acceptance of our air cleaners during the short study period.
-Nevertheless, investments in professional building ventilation systems should be preferred to air cleaners in the long term\cite{Nardell2016}.
+In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners in reducing the transmission of SARS-CoV-2 in two Swiss schools.
+We only found a detectable effect on the risk of infection for mask wearing, but not air cleaners.
+We argued that air cleaners may have had less potential to reduce infections because they were implemented at the end of the study when most students had already been infected with SARS-CoV-2.
+In this study, we used a cross-over design to estimate the effectiveness of air cleaners.
+We found similar reductions in particle concentrations and a tendency that the relative risk of infection was lower with air cleaners.
+To some extent, the differences between the two studies may be explained by different adjustments.
+While we could not adjust for ventilation in our previous study, the estimates of this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see Text~\zref{sec:detailed-molecular} in \supp).
+Furthermore, both studies were conducted during different times and the distribution of respiratory pathogens varied considerably.
+During the COVID-19 pandemic (previous study), we could almost only detect SARS-CoV-2, whereas a range of respiratory pathogens was detected during non-pandemic times (this study).
 
-School closures during the COVID-19 pandemic have been intensely debated as children and adolescents are particularly vulnerable to the negative impact of such interventions on their well-being and mental health\cite{Loades2020}.
-Furthermore, numerous studies have examined the role of children in transmitting SARS-CoV-2\cite{Goldstein2021} and it remains unclear to what extent transmission of SARS-CoV-2 occurs in schools\cite{Ulyte2021Clustering}.
-These findings contrast with studies of influenza viruses showing that school children may drive the seasonal influenza epidemic.
-Community studies in the US demonstrated that influenza transmission rates in children and adolescents were high in schools and that they easily transmit influenza viruses to household members and into their communities\cite{Glezen1978,Shang2018,Fiore2012}.
-Our study suggests that also transmission of SARS-CoV-2 occurs to a considerable extent in schools.
+% why air cleaners may not have been so effective
+
+Despite a tendency to lower the risk of infection, our study indicates that air cleaners were not as effective as other infection control measures.
+Compared to \eg mask wearing, air cleaners cannot influence transmission outside the classrooms.
+Furthermore, air cleaners are unable to prevent close range high particle density aerosol transmission.
+By contrast, face masks can also reduce the generation of airborne infectious particles and can thus mitigate transmission over short ranges.
+Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective in the removal of the larger particles ($>5\mu$m) than the smaller particles ($\leq5\mu$m).
+Yet, a larger proportion of respiratory pathogens are carried in the smaller particles\cite{Fennelly2020}.
+
+% why could we not detect so many bioaerosols
+
+We detected few respiratory pathogens in bioaerosol samples (3~positive bioaoerosl and 50~positive saliva samples), especially when comparing to our previous study\cite{Banholzer2023PLoSMed} (10~positive bioaerosol and 21~positive saliva samples).
+The number of pathogens detected on the filters of the air cleaners was also lower (this study: 4~positive samples in 160~swabs, previous study: 8~positive samples in 80~swabs).
+The bioaerosol sampling devices were the same in both studies and technical issues were not observed.
+It is also unlikely that environmental factors are responsible for the differences.
+Temperature and humidity were similar, and CO$_2$ levels were even higher in this study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
+Therefore, we think that the differences in the number of positive bioaerosols samples between both studies is most likely related to differences in the distribution of respiratory pathogens.
+In this study, we detected a wide range of respiratory pathogens (Flu B, HRV, AdV, etc), as compared to just SARS-CoV-2 in our previous study.
+SARS-CoV-2 may be easier to detect in bioaerosol samples due to differences in the physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity) that influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}.
+In addition, although respiratory pathogens have been detected in bioaerosols (\eg Flu\cite{Bischoff2013JID,Pan2017mSphere}, HRV\cite{Myatt2004AJRCCM,Fabian2011JAMPDD}, AdV\cite{Nguyen2017OFID,Pan2017mSphere}, and CoV\cite{Guo2020EID,Liu2020Nature}), bioaerosol sampling is a challenge\cite{Wang2021 ,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
+
+% transmission within classrooms
+
+The low rate of positive bioaerosol samples also means that it is unclear to what extent students airborne transmission occurred in classrooms.
+It may be possible that students had rather little exposure to respiratory pathogens in schools and acquired their infections elsewhere.
+However, the distribution of saliva samples between classes shows remarkable differences.
+AdV was spreading entirely in class~A in the first three study weeks and only two infections with Flu~B occurred over the study.
+By contrast, Flu~B was spreading throughout the study in class~B and infections with AdV were only detected in the last study week.
+These class-specific patterns in the spread of respiratory pathogens indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.
+
+% limitations
 
@@
 First, aerosol measurements and molecular detection of pathogens in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (human to air, air to human) cannot be determined.
-Second, a comparison of viral concentration between study conditions should be interpreted with care due to the technical limitations of molecular detection in bioaerosol samples, and because the number of possibly infectious (and thus susceptible) students decreased towards the end of the study.
-Third, our epidemiological analysis is based on observational data, thus subject to potential confounding, \eg the incidence of COVID-19 (cases per week) in the community varied over the study period.
-However, levels were high throughout and included two Omicron subvariant peaks.
-Community transmission was also considered in our Bayesian transmission model.
-CO$_2$ levels were not considered in the model but the levels were slightly lower without interventions, suggesting that lower ventilation during intervention phases may have actually reduced the estimated effectiveness of mask mandates and air cleaners.
-Fourth, epidemiological data may not always be complete due to the under-reporting of COVID-19 among absent students.
-We thus used a probabilistic approach to estimate the proportion of suspected cases being actual cases of COVID-19 and the date of symptom onset where it was not reported.
-While this allows us to consider uncertainty in the observed data, the estimated effects of interventions will be less precise as reflected in the larger uncertainty intervals.
+Nevertheless, the pathogen-specific distribution of positive saliva samples between classes indicate potential transmission within classrooms.
+Second, it was not possible to ascertain the specific causes for the absences based on the epidemiological data.
+Therefore, some absences may have falsely been attributed to respiratory infections and we could only approximate the incubation period for each epidemiological case.
+Third, the effectiveness of air cleaners was somewhat limited because students occasionally had lessons outside the classroom.
+This is in contrast to other infection control measures such as face masks that had to be worn in every indoor space during pandemic times.
 
-In conclusion, using epidemiological, environmental, and molecular data, our study shows that considerable transmission of SARS-CoV-2 occurred in the participating schools.
-General face mask wearing were associated with reduced SARS-CoV-2 transmission and prevented infections.
-The effectiveness of interventions was supported by significant decreases in the concentration of aerosols.
-Taken together, our results suggests that infection control measures can reduce the transmission of respiratory infections in school rooms.
-Future studies may use our multiple-measurement approach to assess the effectiveness of infection control measures in reducing the transmission of respiratory infections.
-Ideally, these data should be collected routinely in sentinel schools, thus continuously monitoring transmission risks and alerting health authorities when infection control measures should be taken.
+% conclusion
+
+In conclusion, using epidemiological, environmental, and molecular data, our study shows that considerable transmission of respiratory pathogens occurred in the studied Swiss school.
+Air cleaners significantly reduced aerosol concentrations, but it remains unclear how effectively they remove infectious particles as the number of positive bioaerosol samples was generally too low.
+A tendency towards fewer absences related to respiratory infection during non-pandemic times was estimated.
+Future studies may use our multiple-measurement approach to assess
+whether air cleaners can reduce the risk of airborne transmission and whether their effectiveness varies with the respiratory pathogen that is studied.
 

### 2023-07-05 doc/paper/manuscript.tex
Update Tex - First Draft

@@
 \noindent\textbf{Conclusions:} A wide range of respiratory viruses were detected  in students, but rarely SARS-CoV-2.
-Airborne molecular detection was also rare.
+Airborne molecular detection of respiratory viruses was rare, and when detected, it was not SARS-CoV-2.
 Our results show a reduction in aerosol concentrations with air cleaners and suggest a small benefit of air cleaners in reducing the transmission of airborne respiratory infections in school rooms.
@@
 \noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the Canton of Solothurn, Switzerland, over a seven-week study period from 26~January to 11~March 2023.
-\Cref{fig:study-setup} shows the schematic study setup and \Cref{tab:data} provides an overview of the types of data that were collected in each classroom.
+\Cref{fig:study-setup} shows the schematic study setup and Table~\zref{tab:data} in \supp \, provides an overview of the types of data that were collected in each classroom.
 
@@
 
-\begin{table}[!htpb]
-    \footnotesize
-    \centering
-    \caption{\textbf{Data collection}. Type of data collected, method/device, and frequency in the rooms of classes A and B.}
-    \begin{tabular}{p{3.5cm}p{6cm} p{1cm} p{1cm} p{3cm}}
-    \midrule
-    Type of data & Method / device & Class A & Class B & Frequency \\
-    \midrule
-    \textbf{Molecular data} \\
-    \midrule
-    Bioaerosol sampling & Bioaerosol sampling devices: BioSpot-VIVAS (Aerosol Devices Inc., Ft.
-Collins, Colorado, United States) and Coriolis Micro Air (Bertin Instruments Montigny-le/Bretonneux, France) & x & (x) \newline only Coriolis & daily \\
-    Saliva samples & Saliva samples & x & x & twice per week \\
-    Swabs from air cleaners & Swab samples from Xiaomi HEPA filters (Xiaomi Mi Air Pro 70m2, Shenzhen, China) & x & x & after each intervention phase (both classes) and before the vacation (only class B) \\
-    \midrule
-    \textbf{Environmental data} \\
-    \midrule
-    CO$_2$, aerosol/particle concentrations, humidity, temperature & Air quality device: AQ Guard (Palas GmbH, Karlsruhe, Germany) & x & x & daily by minute \\
-    Coughs & Sounds collected via microphones and coughs detected with a Machine Learning model \cite{Bertschinger2023CBMS} & x & x & daily by seconds \\
-    \midrule
-    \textbf{Epidemiological data} \\
-    \midrule
-    Absences & Survey & x & x & daily \\
-    Student characteristics & Survey & x & x & at study start \\
-    \bottomrule
-    \end{tabular}
-    \label{tab:data}
-\end{table}
-
 \subsection{Study intervention}
@@
     \centering
-    \caption{\textbf{Study design.} Cross-over design where two portable air cleaners (AC) were installed during different time periods in classes A and B over a seven-week study period from 16~January to 11~March 2023, excluding a week of vacation between February 6 and 11.}\label{tab:study_design}
+    \caption{\textbf{Study design.} Cross-over design where two portable air cleaners (AC) were installed during different time periods in classes A and B over a seven-week study period from 16~January to 11~March 2023, excluding a week of vacation between February 6 and 11.
+Swabs from the HEPA filters were taken after each intervention phase as indicated by the ``*''.}\label{tab:study_design}
     \begin{tabular}{l l l l l l l l l}
@@
       \midrule
-      \textbf{A} & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC & \cellcolor{gray!10} None & & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC \\
-      \textbf{B} & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC & & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC & \cellcolor{gray!10} None & \cellcolor{gray!10} None \\
+      \textbf{A} & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{0000}*& \cellcolor{gray!10} None & & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{00009}* \\
+      \textbf{B} & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC\hphantom{0000000}* & & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{0000}*& \cellcolor{gray!10} None & \cellcolor{gray!10} None \\
       \bottomrule
@@
 At the end of the day, samples were transported to the Institute of Infectious Diseases (IFIK) and stored at $-$80°C.
-After each intervention phase (see table xxx), the HEPA filters were carefully removed from the air cleaners, divided into 20~fields, and one swab per field was taken for a total of 20~swabs per filter.\medskip
+After each intervention phase (see \Cref{tab:study_design}), the HEPA filters were carefully removed from the air cleaners, divided into 20~fields, and one swab per field was taken for a total of 20~swabs per filter.\medskip
 
@@
 \textbf{(a)}~Weekly number of respiratory cases with and without air cleaners.
-\textbf{(b)}~Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if air cleaners had always or never been installed in both classes.}
+\textbf{(b)}~Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
     \label{fig:redcap-results}
@@
 We collected epidemiological, environmental, and molecular data to estimate the risk of infection with respiratory infectious diseases in a Swiss school and assessed the effectiveness of air cleaners as an infection control measure.
-We found a wide range of respiratory viruses in human saliva samples, primarily adenovirus, influenza B, and rhinovirus, but only few viruses were detected in bioaerosol samples and on the filters of air cleaners.
+We found a wide range of respiratory viruses in human saliva samples, primarily adenovirus, influenza B, and rhinovirus, but only few non-SARS-CoV-2 respiratory viruses were detected in bioaerosol samples and on the filters of air cleaners.
 Aerosol number and particle mass concentrations dropped significantly with air cleaners.
@@
 
-Despite a tendency to lower the risk of infection, our study indicates that air cleaners were not as effective as other infection control measures.
+Our study indicated that air cleaners can reduce the risk of infection, although we we only were able tho show a trend.
+Other infection control measures, such as universal mask use, appear to be more effective (REFS).
+There may be several reasons for this.
 Compared to \eg mask wearing, air cleaners cannot influence transmission outside the classrooms.
@@
 Therefore, we think that the differences in the number of positive bioaerosols samples between both studies is most likely related to differences in the distribution of respiratory pathogens.
-In this study, we detected a wide range of respiratory pathogens (Flu B, HRV, AdV, etc), as compared to just SARS-CoV-2 in our previous study.
-SARS-CoV-2 may be easier to detect in bioaerosol samples due to differences in the physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity) that influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}.
-In addition, although respiratory pathogens have been detected in bioaerosols (\eg Flu\cite{Bischoff2013JID,Pan2017mSphere}, HRV\cite{Myatt2004AJRCCM,Fabian2011JAMPDD}, AdV\cite{Nguyen2017OFID,Pan2017mSphere}, and CoV\cite{Guo2020EID,Liu2020Nature}), bioaerosol sampling is a challenge\cite{Wang2021 ,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
+Strikingly, we detected a wide range of respiratory pathogens (Flu B, HRV, AdV, etc) in this study, as compared to only SARS-CoV-2 in our previous study.
+The shift in the pattern of respiratory viruses has been observed in other studies (xxx).
+SARS-CoV-2 may be present in higher concentrations in air and may be more easily detected in bioaerosol samples due to differences in physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity) that influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}.
+In addition, although respiratory pathogens have been detected in bioaerosols (\eg Flu\cite{Bischoff2013JID,Pan2017mSphere}, HRV\cite{Myatt2004AJRCCM,Fabian2011JAMPDD}, AdV\cite{Nguyen2017OFID,Pan2017mSphere}, and CoV\cite{Guo2020EID,Liu2020Nature}), bioaerosol sampling remains a technical challenge\cite{Wang2021 ,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
 
@@
 Third, the effectiveness of air cleaners was somewhat limited because students occasionally had lessons outside the classroom.
-This is in contrast to other infection control measures such as face masks that had to be worn in every indoor space during pandemic times.
+This is in contrast to other infection control measures such as face masks that had to be used in every indoor space during pandemic times.
 
@@
 
-In conclusion, using epidemiological, environmental, and molecular data, our study shows that considerable transmission of respiratory pathogens occurred in the studied Swiss school.
-Air cleaners significantly reduced aerosol concentrations, but it remains unclear how effectively they remove infectious particles as the number of positive bioaerosol samples was generally too low.
-A tendency towards fewer absences related to respiratory infection during non-pandemic times was estimated.
-Future studies may use our multiple-measurement approach to assess
-whether air cleaners can reduce the risk of airborne transmission and whether their effectiveness varies with the respiratory pathogen that is studied.
+In conclusion, using epidemiological, environmental, and molecular data, our study shows that considerable transmission of respiratory pathogens other than SARS-CoV-2 occurred in school rooms under non-pandemic conditions.
+Air cleaners significantly reduced aerosol concentrations, but airborne respiratory viruses were rarely detected and if only viruses other than SARS-CoV-2.
+The epidemiologic transmission model estimated a trend toward a reduction in infections with air filters.
+Future studies may use our multiple-measurement approach to assess whether professional building ventilation, which should be preferred over air cleaners in the long term, and other infection control measures reduce the risk of airborne transmission.
+Whether their effectiveness varies with the seasonal distribution of respiratory pathogens under non-pandemic conditions remains to be investigated.
 

### 2023-07-05 doc/paper/supplements.tex
Updates from Overleaf
Update Tex - First Draft

@@
 
-\addto\captionsenglish{\renewcommand{\figurename}{Fig }}
+\addto\captionsenglish{\renewcommand{\figurename}{Figure}}
 
@@
 
-%\renewcommand{\thesection}{Text \arabic{section}}
+\renewcommand{\thesection}{Text \arabic{section}}
 \usepackage{titlesec}
-\titleformat{\section}{\normalfont\Large\bfseries}{Text \thesection.~#1}{1em}{}
-\renewcommand{\thefigure}{\Alph{figure}}
-\renewcommand{\thetable}{\Alph{table}}
+%\titleformat{\section}{\normalfont\Large\bfseries}{Text \thesection.~#1}{1em}{}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
 
@@
 \author[1$\ddag$]{Nicolas Banholzer}
-\author[1$\ddag$]{Kathrin Z\"urcher}
 \author[2]{Philipp Jent}
@@
 \newpage
+
+\section{Data collection}\label{sec:data-collection}
+
+\Cref{tab:data} in \supp provides an overview of the types of data that were collected in each classroom.
+
+\begin{table}[!htpb]
+    \footnotesize
+    \centering
+    \caption{\textbf{Data collection}. Type of data collected, method/device, and frequency in the rooms of classes A and B.}
+    \begin{tabular}{p{3.5cm}p{6cm} p{1cm} p{1cm} p{3cm}}
+    \midrule
+    Type of data & Method / device & Class A & Class B & Frequency \\
+    \midrule
+    \textbf{Molecular data} \\
+    \midrule
+    Bioaerosol sampling & Bioaerosol sampling devices: BioSpot-VIVAS (Aerosol Devices Inc., Ft.
+Collins, Colorado, United States) and Coriolis Micro Air (Bertin Instruments Montigny-le/Bretonneux, France) & x & (x) \newline only Coriolis & daily \\
+    Saliva samples & Saliva samples & x & x & twice per week \\
+    Swabs from air cleaners & Swab samples from Xiaomi HEPA filters (Xiaomi Mi Air Pro 70m2, Shenzhen, China) & x & x & after each intervention phase (both classes) and before the vacation (only class B) \\
+    \midrule
+    \textbf{Environmental data} \\
+    \midrule
+    CO$_2$, aerosol/particle concentrations, humidity, temperature & Air quality device: AQ Guard (Palas GmbH, Karlsruhe, Germany) & x & x & daily by minute \\
+    Coughs & Sounds collected via microphones and coughs detected with a Machine Learning model \cite{Bertschinger2023CBMS} & x & x & daily by seconds \\
+    \midrule
+    \textbf{Epidemiological data} \\
+    \midrule
+    Absences & Survey & x & x & daily \\
+    Student characteristics & Survey & x & x & at study start \\
+    \bottomrule
+    \end{tabular}
+    \label{tab:data}
+\end{table}
+
+\clearpage
 

### 2023-07-05 analysis/env_analysis.Rmd
minor graphical changes

@@
   ggplot() +
-  geom_rect(aes(xmin = .lower, xmax = .upper, ymin = 0.75, ymax = 1.25, fill = `.width`)) +
-  geom_segment(aes(x = aircleaner, xend = aircleaner, y = 0.75, yend = 1.25), size = 1.5 / cm(1)) +
-  geom_point(aes(x = aircleaner, y = 1), size = 3, fill = "white", shape = 21) +
+  geom_rect(aes(xmin = .lower, xmax = .upper, ymin = 75, ymax = 125, fill = `.width`)) +
+  geom_segment(aes(x = aircleaner, xend = aircleaner, y = 75, yend = 125), size = 1.5 / cm(1)) +
+  geom_point(aes(x = aircleaner, y = 100), size = 3, fill = "white", shape = 21) +
   facet_wrap(~ outcome, nrow = 1, labeller = as_labeller(vars_labels_no_units, label_parsed)) +

### 2023-07-05 analysis/mol_analysis.Rmd
minor graphical changes

@@
   theme(panel.grid = element_blank(),
-        panel.spacing.x = unit(2, "cm"),
+        panel.spacing.x = unit(1, "cm"),
         axis.title.x = element_text(color = "white"),
@@
 # combine
-descr <- pos_tot_pl + pos_cond_pl & theme(legend.position = "bottom", legend.key.height = unit(.1, "cm"),
+descr <- pos_tot_pl + plot_spacer() + pos_cond_pl & theme(legend.position = "bottom", legend.key.height = unit(.1, "cm"),
                                           legend.key.width = unit(.3, "cm"))
-descr <- descr + plot_layout(guides = "collect", widths = c(10, 6), ncol = 2) 
+descr <- descr + plot_layout(guides = "collect", widths = c(9.5, 1, 5.5), ncol = 2) 
 descr <- descr + pos_week_pl 
-descr <- descr + plot_layout(nrow = 2, heights = c(8, 8), design = 'AB
-                             CC')
+descr <- descr + plot_layout(nrow = 2, heights = c(8, 8), design = 'ABC
+                             DDD')
 save_plot(descr, pdf_file = "../results/mol-data/descriptives.pdf", eps_file = "../results/mol-data/descriptives.eps", h = 16, w = 16)
-saveRDS(pos_tot_pl + ggtitle(""), file = "../results/mol-data/descriptives-total.rds")
-saveRDS(pos_cond_pl + ggtitle(""), file = "../results/mol-data/descriptives-cond.rds")
-saveRDS(pos_week_pl + ggtitle(""), file = "../results/mol-data/descriptives-week.rds")
+# saveRDS(pos_tot_pl + ggtitle(""), file = "../results/mol-data/descriptives-total.rds")
+# saveRDS(pos_cond_pl + ggtitle(""), file = "../results/mol-data/descriptives-cond.rds")
+# saveRDS(pos_week_pl + ggtitle(""), file = "../results/mol-data/descriptives-week.rds")
 ```

### 2023-07-05 doc/paper/manuscript.tex
Replace pathogen with virus

@@
 \\
-% Study limitations include unknown specific causes of absences related to respiratory infections (i.e which pathogen caused the sickness).
+% Study limitations include unknown specific causes of absences related to respiratory infections (i.e which virus caused the sickness).
 Furthermore, students occasionally had lessons outside the classroom, thus limiting the potential effectiveness of air cleaners.
@@
 The transmission of respiratory infections such as SARS-CoV-2 and influenza presents significant challenges in terms of mitigation and control.
-Person-to-person transmission primarily occurs through the release of respiratory particles containing the respective pathogens.
+Person-to-person transmission primarily occurs through the release of respiratory particles containing the respective viruses.
 The recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
@@
 Furthermore, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}.
-As a result, studies have simulated the potential of air cleaners in reducing indoor transmission risks of SARS-CoV-2\cite{Lindsley2021} and other respiratory pathogens\cite{Cortellessa2023Build}.
+As a result, studies have simulated the potential of air cleaners in reducing indoor transmission risks of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
 However, empirical evidence linking the use of air cleaners with a reduced risk of infection is lacking.
@@
 A detailed description of the model and choice of priors for all model parameters are provided in Text~\zref{sec:transmission-model} in \supp.
-Importantly, since multiple pathogens were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the pathogen-specific distributions.
-The weights were inferred based on the proportion of positive saliva samples for each pathogen per week.
-As a result, the distribution of the incubation period used for modeling could change every week depending on the proportion of pathogens that were detected in the students' saliva.
+Importantly, since multiple viruses were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the virus-specific distributions.
+The weights were inferred based on the proportion of positive saliva samples for each virus per week.
+As a result, the distribution of the incubation period used for modeling could change every week depending on the proportion of viruses that were detected in the students' saliva.
 Moreover, respiratory cases by date of symptom onset were under-reported for weekends and thus a re-weighting scheme was used to account for weekday effects.
@@
 
-\noindent Molecular analysis determined which saliva samples were positive for a respiratory pathogen.
-We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (Text~\zref{sec:multinomial-model} in \supp) where the expected number of positive samples for a pathogen was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each pathogen.
+\noindent Molecular analysis determined which saliva samples were positive for a respiratory virus.
+We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (Text~\zref{sec:multinomial-model} in \supp) where the expected number of positive samples for a virus was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each virus.
 The latter was computed as the difference between the number of students per class and the cumulative number of positive samples.
@@
 For example, in the first three study weeks all except for one sample were positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
-The pathogen-specific variations between classes are suggestive of transmission occurring within classes.
+The virus-specific variations between classes are suggestive of transmission occurring within classes.
 
@@
     \caption{\textbf{Analysis of molecular data}.
-\textbf{(a)}~Number of positive saliva samples by pathogen.
+\textbf{(a)}~Number of positive saliva samples by virus.
 \textbf{(b)}~Number of positive saliva samples by study condition.
 \textbf{(c)}~Number of positive saliva samples by study week and class.
-Pathogens: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}
+Viruses: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}
     \label{fig:molecular-descriptives}
@@
 While we could not adjust for ventilation in our previous study, the estimates of this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see Text~\zref{sec:detailed-molecular} in \supp).
-Furthermore, both studies were conducted during different times and the distribution of respiratory pathogens varied considerably.
-During the COVID-19 pandemic (previous study), we could almost only detect SARS-CoV-2, whereas a range of respiratory pathogens was detected during non-pandemic times (this study).
+Furthermore, both studies were conducted during different times and the distribution of respiratory viruses varied considerably.
+During the COVID-19 pandemic (previous study), we could almost only detect SARS-CoV-2, whereas a range of respiratory viruses was detected during non-pandemic times (this study).
 
@@
 Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective in the removal of the larger particles ($>5\mu$m) than the smaller particles ($\leq5\mu$m).
-Yet, a larger proportion of respiratory pathogens are carried in the smaller particles\cite{Fennelly2020}.
+Yet, a larger proportion of respiratory viruses are carried in the smaller particles\cite{Fennelly2020}.
 
@@
 
-We detected few respiratory pathogens in bioaerosol samples (3~positive bioaoerosl and 50~positive saliva samples), especially when comparing to our previous study\cite{Banholzer2023PLoSMed} (10~positive bioaerosol and 21~positive saliva samples).
-The number of pathogens detected on the filters of the air cleaners was also lower (this study: 4~positive samples in 160~swabs, previous study: 8~positive samples in 80~swabs).
+We detected few respiratory viruses in bioaerosol samples (3~positive bioaoerosl and 50~positive saliva samples), especially when comparing to our previous study\cite{Banholzer2023PLoSMed} (10~positive bioaerosol and 21~positive saliva samples).
+The number of viruses detected on the filters of the air cleaners was also lower (this study: 4~positive samples in 160~swabs, previous study: 8~positive samples in 80~swabs).
 The bioaerosol sampling devices were the same in both studies and technical issues were not observed.
@@
 Temperature and humidity were similar, and CO$_2$ levels were even higher in this study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
-Therefore, we think that the differences in the number of positive bioaerosols samples between both studies is most likely related to differences in the distribution of respiratory pathogens.
-Strikingly, we detected a wide range of respiratory pathogens (Flu B, HRV, AdV, etc) in this study, as compared to only SARS-CoV-2 in our previous study.
+Therefore, we think that the differences in the number of positive bioaerosols samples between both studies is most likely related to differences in the distribution of respiratory viruses.
+Strikingly, we detected a wide range of respiratory viruses (Flu B, HRV, AdV, etc) in this study, as compared to only SARS-CoV-2 in our previous study.
 The shift in the pattern of respiratory viruses has been observed in other studies (xxx).
 SARS-CoV-2 may be present in higher concentrations in air and may be more easily detected in bioaerosol samples due to differences in physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity) that influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}.
-In addition, although respiratory pathogens have been detected in bioaerosols (\eg Flu\cite{Bischoff2013JID,Pan2017mSphere}, HRV\cite{Myatt2004AJRCCM,Fabian2011JAMPDD}, AdV\cite{Nguyen2017OFID,Pan2017mSphere}, and CoV\cite{Guo2020EID,Liu2020Nature}), bioaerosol sampling remains a technical challenge\cite{Wang2021 ,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
+In addition, although respiratory viruses have been detected in bioaerosols (\eg Flu\cite{Bischoff2013JID,Pan2017mSphere}, HRV\cite{Myatt2004AJRCCM,Fabian2011JAMPDD}, AdV\cite{Nguyen2017OFID,Pan2017mSphere}, and CoV\cite{Guo2020EID,Liu2020Nature}), bioaerosol sampling remains a technical challenge\cite{Wang2021 ,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
 
@@
 The low rate of positive bioaerosol samples also means that it is unclear to what extent students airborne transmission occurred in classrooms.
-It may be possible that students had rather little exposure to respiratory pathogens in schools and acquired their infections elsewhere.
+It may be possible that students had rather little exposure to respiratory viruses in schools and acquired their infections elsewhere.
 However, the distribution of saliva samples between classes shows remarkable differences.
@@
 By contrast, Flu~B was spreading throughout the study in class~B and infections with AdV were only detected in the last study week.
-These class-specific patterns in the spread of respiratory pathogens indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.
+These class-specific patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.
 
@@
 Our study has several limitations.
-First, aerosol measurements and molecular detection of pathogens in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (human to air, air to human) cannot be determined.
-Nevertheless, the pathogen-specific distribution of positive saliva samples between classes indicate potential transmission within classrooms.
+First, aerosol measurements and molecular detection of viruses in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (human to air, air to human) cannot be determined.
+Nevertheless, the virus-specific distribution of positive saliva samples between classes indicate potential transmission within classrooms.
 Second, it was not possible to ascertain the specific causes for the absences based on the epidemiological data.
@@
 
-In conclusion, using epidemiological, environmental, and molecular data, our study shows that considerable transmission of respiratory pathogens other than SARS-CoV-2 occurred in school rooms under non-pandemic conditions.
+In conclusion, using epidemiological, environmental, and molecular data, our study shows that considerable transmission of respiratory viruses other than SARS-CoV-2 occurred in school rooms under non-pandemic conditions.
 Air cleaners significantly reduced aerosol concentrations, but airborne respiratory viruses were rarely detected and if only viruses other than SARS-CoV-2.
@@
 Future studies may use our multiple-measurement approach to assess whether professional building ventilation, which should be preferred over air cleaners in the long term, and other infection control measures reduce the risk of airborne transmission.
-Whether their effectiveness varies with the seasonal distribution of respiratory pathogens under non-pandemic conditions remains to be investigated.
+Whether their effectiveness varies with the seasonal distribution of respiratory viruses under non-pandemic conditions remains to be investigated.
 

### 2023-07-05 doc/paper/supplements.tex
Replace pathogen with virus

@@
 \Cref{tab:mol-data-line-list} shows the line list data for the laboratory test results from human saliva samples.
-For each class and test, it shows the date when the test was taken, the test result and (if positive) the pathogen that was detected.
+For each class and test, it shows the date when the test was taken, the test result and (if positive) the virus that was detected.
 Recall that all students in class were sampled twice per week every Tuesday and Thursday.
@@
     \caption[Line list of molecular test results over the study period]{Line list of respiratory cases over the study period.
-Pathogens: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}\label{tab:mol-data-line-list} \\
+Viruses: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}\label{tab:mol-data-line-list} \\
     \toprule
-    Class & Date of test & Weekday & Test result & Pathogen \\
+    Class & Date of test & Weekday & Test result & Virus \\
     \midrule
@@
 
-The pathogen of each respiratory infection could not be identified from the epidemiological data because the students never obtained a laboratory test result.
-As a consequence, different incubation periods need to be considered in $p_{\text{IN}}$, reflecting a combination of the pathogen-specific incubation periods.
-The combination is determined based on the weekly proportion of positive saliva samples for each pathogen found in the molecular analysis.
-Formally, let $p_{\text{IN}}^{v}$ be the distribution for the incubation period of respiratory virus $v$ and let $pp_{vw}$ be the proportion of positive saliva samples in study week $w$, then each week the combined incubation period is computed as the weighted sum of the pathogen-specific incubation periods
+The virus of each respiratory infection could not be identified from the epidemiological data because the students never obtained a laboratory test result.
+As a consequence, different incubation periods need to be considered in $p_{\text{IN}}$, reflecting a combination of the virus-specific incubation periods.
+The combination is determined based on the weekly proportion of positive saliva samples for each virus found in the molecular analysis.
+Formally, let $p_{\text{IN}}^{v}$ be the distribution for the incubation period of respiratory virus $v$ and let $pp_{vw}$ be the proportion of positive saliva samples in study week $w$, then each week the combined incubation period is computed as the weighted sum of the virus-specific incubation periods
 \begin{align}
@@
 
-The prior distributions for the pathogen-specific incubation periods are based on estimates published in the literature\cite{McAloon2020,Lessler2009LancetID}.
+The prior distributions for the virus-specific incubation periods are based on estimates published in the literature\cite{McAloon2020,Lessler2009LancetID}.
 The distributions are shown in \Cref{fig:prior-pin}.
@@
 
-We estimate the pathogen-specific distributions from our data as part of fitting the overall model.
+We estimate the virus-specific distributions from our data as part of fitting the overall model.
 Furthermore, note that $p_\text{IN}$ is discretized via $p_\text{IN}(s) = \int_0^{0.5} p_\text{IN}(\tau) \;\text{d}\tau$ for $s = 0$ and $p_\text{IN}(s) = \int_{s-0.5}^{s+0.5} p_\text{IN}(\tau) \;\text{d}\tau$ for $s > 0$, where $p_\text{IN}(\tau) \sim \textrm{Lognormal}(\mu, \sigma)$ is the density of the lognormal distribution.
@@
     \includegraphics{../../results/epi-data/incubation-periods.pdf}
-    \caption[Choices of priors for the incubation periods]{Choices of priors for the distribution of the pathogen-specific incubation periods.
-Pathogens: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}
+    \caption[Choices of priors for the incubation periods]{Choices of priors for the distribution of the virus-specific incubation periods.
+Viruses: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}
     \label{fig:prior-pin}
@@
 Cases in the first week of the study could indicate infections before the study commenced.
-We will therefore seed our model $2 \cdot \text{m}\,$days before the study start, where $\text{m}$ is the average incubation period of the pathogen with the largest incubation period (\ie adenovirus).
+We will therefore seed our model $2 \cdot \text{m}\,$days before the study start, where $\text{m}$ is the average incubation period of the virus with the largest incubation period (\ie adenovirus).
 The number of infections before the study start will be modeled with an exponential prior
@@
 
-Molecular analysis determined which saliva samples were positive for a respiratory pathogen.
-Let $p = 0, 1, \dots, P$ denote the pathogen, where $0$ refers to negative samples and $P$ is the number of different pathogens detected over the study.
+Molecular analysis determined which saliva samples were positive for a respiratory virus.
+Let $p = 0, 1, \dots, P$ denote the virus, where $0$ refers to negative samples and $P$ is the number of different viruses detected over the study.
 The number of positive bioaerosol and saliva samples $y_p$ in samples of sizes $n$ is analyzed with a Multinomial logistic regression model
@@
 The negative test is set as the reference category.
-Overall variation in positive samples by pathogen will be modeled with $\beta_{0p}$ and the effect of air cleaners will be modeled with $\beta_1$.
-The effect of air cleaners is adjusted for class-specific effects and the decreasing number of susceptibles over time (for each pathogen computed as the number of students minus the total number of positive samples).
+Overall variation in positive samples by virus will be modeled with $\beta_{0p}$ and the effect of air cleaners will be modeled with $\beta_1$.
+The effect of air cleaners is adjusted for class-specific effects and the decreasing number of susceptibles over time (for each virus computed as the number of students minus the total number of positive samples).
 
 Note that we deviated from the statistical analysis plan by estimating $\beta_{0p}$ and $\beta_1$ without a hierarchical prior.
-First, there was considerable variation in the overall positivity rate of each pathogen, so that we decided to use unpooled intercepts $\beta_{0p}$ instead of partially pooled intercepts $\beta_{0[p]}$ for each pathogen.
+First, there was considerable variation in the overall positivity rate of each virus, so that we decided to use unpooled intercepts $\beta_{0p}$ instead of partially pooled intercepts $\beta_{0[p]}$ for each virus.
 Second, there was insufficient variation in the data to inform partially pooled estimates $\beta_{1[p]}$ for the effects of air cleaners.
-Therefore, we decided to only estimate the average effect across pathogens, \ie the completely pooled estimate $\beta_1$.
+Therefore, we decided to only estimate the average effect across viruses, \ie the completely pooled estimate $\beta_1$.
 

### 2023-07-05 analysis/env_analysis.Rmd
minor graphical changes

@@
   scale_x_continuous(limits = c(50, 100), expand = expansion(mult = c(0.1, 0.1)), breaks = c(50, 75, 100)) +
-  labs(x = "Estimated reduction (%)", fill = "CrI", title = "b") +
+  labs(x = "Estimated reduction (%)", y = "Type", fill = "CrI", title = "b") +
   theme_custom() +

### 2023-07-05 analysis/epi_analysis.Rmd
minor graphical changes

@@
   ggplot() +
-  geom_rect(aes(xmin = .lower, xmax = .upper, ymin = 0.75, ymax = 1.25, fill = `.width`)) +
-  geom_segment(aes(x = infections, xend = infections, y = 0.75, yend = 1.25), size = 1.5 / cm(1)) +
-  geom_point(aes(x = infections, y = 1), size = 3, fill = "white", shape = 21) +
+  geom_rect(aes(xmin = .lower, xmax = .upper, ymin = 7.5, ymax = 12.5, fill = `.width`)) +
+  geom_segment(aes(x = infections, xend = infections, y = 7.5, yend = 12.5), size = 1.5 / cm(1)) +
+  geom_point(aes(x = infections, y = 10), size = 3, fill = "white", shape = 21) +
   facet_wrap(~ type, ncol = 1) +
@@
   scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0)), breaks = seq(0, 200, 25)) +
-  labs(x = "Estimated number of infections", fill = "CrI", title = "b") +
+  labs(x = "Estimated number of infections", y = "Scenario", fill = "CrI", title = "b") +
   theme_custom() +
-  theme(axis.title.y = element_blank(),
+  theme(axis.title.y = element_text(color = "white"),
         axis.text.y = element_text(color = "white"),

### 2023-07-05 analysis/mol_analysis.Rmd
minor graphical changes

@@
   scale_x_continuous(breaks = seq(1, 7, 1)) +
-  labs(y = "Number of positive samples", x = "Study week", fill = "Pathogen", title = "c") +
+  labs(y = "Number of positive samples", x = "Study week", fill = "Virus", title = "c") +
   theme_custom() +
@@
         axis.title.x = element_text(color = "white"),
-        legend.position = "none") +
+        legend.position = "none",
+        plot.title.position = "plot") +
   guides(fill = guide_legend(nrow = 1))
@@
   scale_y_continuous(expand = c(0,0)) +
-  labs(y = "Number of positive samples", x = "Air cleaner", fill = "Pathogen", title = "b") +
+  labs(y = "Number of positive samples", x = "Air cleaner", fill = "Virus", title = "b") +
   theme_custom() +
   theme(panel.grid = element_blank(),
-        legend.position = "bottom") +
+        legend.position = "bottom",
+        plot.title.position = "plot") +
   guides(fill = guide_legend(nrow = 1))
@@
   scale_y_continuous(expand = expansion(add = c(0,1))) +
-  labs(y = "Number of positive samples", x = "Pathogen", fill = "Pathogen", title = "a") +
+  labs(y = "Number of positive samples", x = "Virus", fill = "Virus", title = "a") +
   theme_custom() +
   theme(panel.grid = element_blank(),
-        legend.position = "bottom") +
+        legend.position = "bottom",
+        plot.title.position = "plot") +
   guides(fill = guide_legend(nrow = 1))

### 2023-07-05 doc/paper/manuscript.tex
Updates from Overleaf

@@
 Our study indicated that air cleaners can reduce the risk of infection, although we we only were able tho show a trend.
-Other infection control measures, such as universal mask use, appear to be more effective (REFS).
+Other infection control measures, such as universal mask use, appear to be more effective\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021}.
 There may be several reasons for this.
@@
 Strikingly, we detected a wide range of respiratory viruses (Flu B, HRV, AdV, etc) in this study, as compared to only SARS-CoV-2 in our previous study.
-The shift in the pattern of respiratory viruses has been observed in other studies (xxx).
+The shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.
 SARS-CoV-2 may be present in higher concentrations in air and may be more easily detected in bioaerosol samples due to differences in physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity) that influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}.

### 2023-07-10 analysis/env_analysis.Rmd
add study comparison

@@
   mutate_if(is.numeric, round_k) %>%
-  mutate(across(-outcome, tex_sign)) %>%
   arrange(-row_number()) %>%

### 2023-07-10 analysis/study_comaprison.Rmd
add study comparison

---
title: "Study Comparison"
author: "Nicolas Banholzer"
date: "2023-07-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(reshape2)
source("../utils/epi.r")
source("../utils/bayes.r")
source("../utils/plotting.r")
source("../utils/tex.r")
```

## Virus distribution

```{r}
saliva <- data.frame(
    year = c(rep(2022, 6), rep(2023, 6)),
    virus = rep(c("Flu", "HRV", "AdV", "CoV", "MPV", "PIV"), 2),
    n = c(c(1, NA, 1, 19, NA, NA), c(15, 15, 14, 3, 2, 1))
  ) %>%
  group_by(year) %>%
  mutate(p = round(100 * n / sum(n, na.rm = T))) %>%
  ungroup() %>%
  mutate(year = factor(year, levels = c(2022, 2023)),
         virus = factor(virus, levels = c("Flu", "HRV", "AdV", "CoV", "MPV", "PIV"))) %>%
  ggplot(mapping = aes(y = factor(year), x = p, fill = virus)) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE), width = .5) +
  geom_text(mapping = aes(label = p), size = 8 / cm(1), color = "white", position = position_stack(vjust = 0.5, reverse = T)) +
  scale_x_continuous(expand = expansion(mult = c(0,0.025))) +
  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
  labs(x = "Proportion of positive saliva samples (%)", y = "Study year", fill = "Virus", title = "a") +
  theme_custom() +
  theme(legend.position = "top",
        plot.title.position = "plot",
        legend.key.height = unit(.1, "cm"),
        legend.key.width = unit(.3, "cm")) +
  guides(fill = guide_legend(nrow = 1))

saliva
```

## No. of positive samples

```{r}
pos_samples <- data.frame(
  year = factor(c(rep(2022, 3), rep(2023, 3)), levels = c(2022, 2023)),
  type = factor(rep(c("Saliva", "Bioaerosols", "HEPA-filter"), times = 2), levels = c("Saliva", "Bioaerosols", "HEPA-filter")),
  n = c(c(21, 10, 8), c(50, 2, 4))
) %>%
  ggplot(aes(x = n, y = year, pattern = year, color = year, pattern_color = year)) +
  geom_bar_pattern(stat = "identity", fill = "white", pattern_density = 0.1, pattern_spacing = 0.1, width = .75) +
  geom_text(aes(label = n), size = 8 / cm(1), hjust = -1) +
  facet_wrap(~ type, ncol = 1) +
  scale_x_continuous(expand = expansion(add = c(0,9.5))) +
  scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) +
  scale_pattern_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) +
  labs(y = "Study year", x = "Frequency", title = "b") +
  theme_custom() +
  theme(legend.position = "none",
        plot.title.position = "plot")

pos_samples
```

## Particle reduction

```{r}
vars <- c("CO2", "rH", "Temp", "AER", "Cn", "PM1", "PM25", "PM4", "PM10")
vars_labels_no_units <- c("CO[2]", "Relative~humidity", "Temperature", "Air~exchange~rate",
                          "CN", "PM[1]", "PM[2.5]", "PM[4]", "PM[10]")
names(vars_labels_no_units) <- vars

particle <- data.frame(
  year = factor(c(rep(2022, 15), rep(2023, 15)), levels = c(2022, 2023)),
  variable = factor(c(rep(c("Cn", "PM1", "PM25", "PM4", "PM10"), each = 3, times = 2)), levels = c("Cn", "PM1", "PM25", "PM4", "PM10")),
  interval = rep(c("Mean", "Lower", "Upper"), times = 5),
  estimate = c(c(39.24, 3.88, 68.80, 
                 41.05, 6.62, 67.98,
                 45.12, 19.66, 63.16,
                 46.49, 27.26, 61.72,
                 46.60, 29.99, 59.49),
               c(76.83, 63.20, 86.28,
                 76.79, 64.84, 85.46,
                 72.86, 64.24, 79.66,
                 70.77, 63.46, 76.97,
                 99.99, 99.94, 100.00))
) %>%
  dcast(year + variable ~ interval) %>%
  ggplot(aes(y = year, color = year)) +
  facet_wrap(~ variable, nrow = 1, dir = "h", labeller = as_labeller(vars_labels_no_units, label_parsed)) +
  geom_errorbar(mapping = aes(xmin = Lower, xmax = Upper), lineend = "square", size = 1, width = .05) +
  geom_point(mapping = aes(x = Mean), shape = 23, size = 2, fill = "white") +
  scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1)), breaks = c(0, 50, 100), limits = c(0, 100)) +
  labs(y = "Study year", x = "Estimated reduction (%)", title = "c") + 
  theme_custom() +
  theme(legend.position = "none",
        plot.title.position = "plot",
        panel.spacing = unit(.5, "lines"))

particle
```

## Risk ratio

```{r}
rr <- data.frame(
  year = factor(c(rep(2022, 3), rep(2023, 3)), levels = c(2022, 2023)),
  interval = rep(c("Mean", "Lower", "Upper"), times = 2),
  estimate = c(c(1.00, 0.22, 2.067), c(0.72, 0.44, 1.15))
) %>%
  dcast(year ~ interval) %>%
  ggplot(aes(y = year, color = year)) +
  geom_vline(aes(xintercept = 1), linetype = "dashed") +
  geom_errorbar(mapping = aes(xmin = Lower, xmax = Upper), lineend = "square", size = 1, width = .05) +
  geom_point(mapping = aes(x = Mean), shape = 23, size = 2, fill = "white") +
  labs(y = "Study year", x = "Adjusted risk ratio (%)", title = "d") + 
  scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) +
  theme_custom() +
  theme(legend.position = "none",
        plot.title.position = "plot")

rr  
```

## Combined

```{r}
comparison <- grid.arrange(saliva, pos_samples, particle, rr, ncol = 2, nrow = 2, heights = c(6, 4), widths = c(10,6))
ggsave(plot = comparison, file = "../results/comparison.pdf", width = 16 / cm(1), height = 10 / cm(1))
```

### 2023-07-10 doc/paper/manuscript.tex
Updates from Overleaf

@@
 
-\title{\bfseries\LARGE\singlespacing{Transmission of airborne respiratory infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}}
+\title{\bfseries\LARGE\singlespacing{Transmission of airborne respiratory viral infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}}
 
@@
 Infection control measures such as air ventilation or filtration can reduce the risk of airborne transmission.
-The aim of this study was to evaluate the effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory infections in schools using a multiple-measurement approach.
+The aim of this study was to evaluate the effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory viral infections in schools using a multiple-measurement approach.
 \medskip
@@
 The recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
-Unlike larger respiratory droplets that tend to settle quickly, aerosols can remain suspended in the air for several hours and their respiratory plumes can reach more than 2m\cite{Coleman2022,Wang2020,Heneghan2021}, thus exceeding the recommended rules for social distancing by public health organizations\cite{Bazant2021PNAS,Trivedi2021PhyFluids,Lindsley2010CID}.
+Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and their respiratory plumes can reach more than 2 meters\cite{Coleman2022,Wang2020,Heneghan2021}, exceeding the social distancing rules recommended by public health organizations\cite{Bazant2021PNAS,Trivedi2021PhyFluids,Lindsley2010CID}.
 
-The generation, spread, and concentration of bioareosols can be influenced with infection control measures.
-For example, compulsory face mask wearing reduced the transmission of SARS-CoV-2 during the COVID-19 pandemic\cite{Gettings2021,Heinsohn2022,Leech2022PNAS,Mitze2020PNAS}.
-More generally, improved ventilation systems are considered crucial for healthy indoor environments both during pandemic and non-pandemic times\cite{Wang2021,Morawska2021}.
-Portable air-filtration devices (air cleaners) may present another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear.
+The generation, spread, and concentration of bioareosols can be influenced by infection control measures.
+For example, compulsory face mask wearing reduced the transmission of SARS-CoV-2 during the COVID-19 pandemic\cite{Gettings2021,Heinsohn2022,Leech2022PNAS,Mitze2020PNAS,Banholzer2021PLOS}.
+In addition, improved ventilation systems are critical for a healthy indoor environment both during pandemic and non-pandemic times\cite{Wang2021,Morawska2021}.
+Portable HEPA-air filtration devices (air cleaners ) may present another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear.
 Various studies showed that air cleaners reduce the concentration of aerosols and particle matter in indoor air\cite{Park2020Build,Buising2022InfContr,Ren2021Dent,Banholzer2023PLoSMed}.
-These findings suggest potential protection from respiratory infection as long-term exposure to higher aerosol and particle matter mass concentration have been associated with increased risks of respiratory infections\cite{Gordon2014Resp} and, more recently, with the incidence and mortality of COVID-19\cite{Kelly2023Atmos,Veronesi2022OccEnvMed,Travaglio2021EnvPoll,DeAngelis2021EnvRes}.
+These findings suggest potential protection against respiratory infection, as long-term exposure to higher aerosol and particle matter mass concentrations has been associated with increased risk of respiratory infections\cite{Gordon2014Resp} and, more recently, with the incidence and mortality of COVID-19\cite{Kelly2023Atmos,Veronesi2022OccEnvMed,Travaglio2021EnvPoll,DeAngelis2021EnvRes}.
 Furthermore, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}.
-As a result, studies have simulated the potential of air cleaners in reducing indoor transmission risks of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
-However, empirical evidence linking the use of air cleaners with a reduced risk of infection is lacking.
-Thus, it remains unclear whether the reduction in particle concentrations and the removal of bioaerosols from the air translates into lower transmission of respiratory infections.
+Simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission  of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
+We have studied the effectiveness of air cleaners in reducing the transmission of SARS-CoV-2 transmission in schools during the pandemic\cite{Banholzer2023PLoSMed}, but the analysis was limited by the study design (installation of air cleaners at the end of the Omicron wave when most students were already infected).
+In conclusion, empirical evidence linking the use of air cleaners with a reduced risk of infection is lacking.
+It remains unclear whether reducing particle concentrations and removing bioaerosols from the air will translate in reduced transmission of respiratory infections.
 
-We used a multiple-measurement approach to study the transmission of respiratory viruses in school rooms under non-pandemic conditions in the winter of 2023.
-We collected epidemiological (absences probably related to respiratory infections), environmental (CO$_2$, aerosol and particle concentrations), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) over a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
-We used a cross-over study design to assess the effectiveness of air cleaners as an infection control measure.
-For this, we estimated changes in particle concentrations, the positivity rate for respiratory viruses, and the risk of infection.
+We used a multiple-measurement approach to study the transmission of viral respiratory viruses and the effect of aircleaners using a cross-over study design in school rooms under non-pandemic conditions in the winter of 2023.
+We collected epidemiological (absences probably related to respiratory infections), environmental (CO$_2$, aerosol and particle concentrations), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
+We determined changes in particle concentrations, respiratory virus positivity rates, and modeled the risk of infection.
 
@@
 
+\subsection{Comparison with previous study}
+
+In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 Omicron wave.
+\Cref{fig:comparison} shows a comparison of the results.
+The distribution of respiratory infections is different (\Cref{fig:comparison}a) because we almost exclusively detected SARS-CoV-2 in our previous study during the COVID-19 pandemic, as compared to a range of respiratory infections in this study during non-pandemic times.
+Fewer respiratory infections were detected in bioaerosols and on the HEPA-filters of the air cleaners in this study (\Cref{fig:comparison}b).
+The reductions in particle concentrations were similar (\Cref{fig:comparison}c).
+The risk of infection tended to be lower with air cleaners in this study where air cleaners were analyzed using a cross-over design, but not in our previous study where air cleaners were installed only at the end of the study when most students had already been infected with SARS-Cov-2(\Cref{fig:comparison}d).
+
+\begin{figure}[!htpb]
+    \includegraphics[width=\linewidth]{../../results/epi-data/cases_by_week.pdf} 
+    \includegraphics[width=\linewidth]{../../results/epi-data/avoided-infections.pdf}
+    \caption{\textbf{Analysis of epidemiological data and estimated relative risk of infection with air cleaners.}.
+\textbf{(a)}~Weekly number of respiratory cases with and without air cleaners.
+\textbf{(b)}~Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
+    \label{fig:comparison}
+\end{figure}
+
 \FloatBarrier

### 2023-07-10 doc/paper/manuscript.tex
Comparison with previous study

@@
 The distribution of respiratory infections is different (\Cref{fig:comparison}a) because we almost exclusively detected SARS-CoV-2 in our previous study during the COVID-19 pandemic, as compared to a range of respiratory infections in this study during non-pandemic times.
-Fewer respiratory infections were detected in bioaerosols and on the HEPA-filters of the air cleaners in this study (\Cref{fig:comparison}b).
-The reductions in particle concentrations were similar (\Cref{fig:comparison}c).
+Although there were more positive saliva samples in this study, the number of respiratory infections detected in bioaerosols and on the HEPA-filters of the air cleaners was lower (\Cref{fig:comparison}b).
+The reductions in particle concentrations were larger and with lower uncertainty in this study (\Cref{fig:comparison}c).
 The risk of infection tended to be lower with air cleaners in this study where air cleaners were analyzed using a cross-over design, but not in our previous study where air cleaners were installed only at the end of the study when most students had already been infected with SARS-Cov-2(\Cref{fig:comparison}d).
@@
 \begin{figure}[!htpb]
-    \includegraphics[width=\linewidth]{../../results/epi-data/cases_by_week.pdf} 
-    \includegraphics[width=\linewidth]{../../results/epi-data/avoided-infections.pdf}
-    \caption{\textbf{Analysis of epidemiological data and estimated relative risk of infection with air cleaners.}.
-\textbf{(a)}~Weekly number of respiratory cases with and without air cleaners.
-\textbf{(b)}~Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
+    \includegraphics[width=\linewidth]{../../results/comparison.pdf} 
+    \caption{\textbf{Comparison of the results between our previous study in 2022 during the COVID-19 pandemic (blue) and this study in 2023 during non-pandemic times (black).}.
+\textbf{(a)}~Proportion of positive saliva samples and virus.
+Viruses: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).
+\textbf{(b)}~Number of positive saliva, bioaerosol, and HEPA-filter samples.
+\textbf{(c)}~Estimated reduction in aerosol number and particle mass concentrations (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$) with air cleaners (posterior mean as squares and 95\%-CrI as lines).
+\textbf{(d)}~Adjusted infection risk ratio (posterior mean as squares and 95\%-CrI as lines).}
     \label{fig:comparison}
@@
 Temperature and humidity were similar, and CO$_2$ levels were even higher in this study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
-Therefore, we think that the differences in the number of positive bioaerosols samples between both studies is most likely related to differences in the distribution of respiratory viruses.
-Strikingly, we detected a wide range of respiratory viruses (Flu B, HRV, AdV, etc) in this study, as compared to only SARS-CoV-2 in our previous study.
+Lower ventilation was potentially offset by a larger reduction in aerosol and particle concentrations through air cleaners in this study.
+However, it could be that air cleaners removed particles less quickly than ventilation.
+The most striking difference is in the distribution of respiratory viruses.
+Compared to the COVID-19 pandemic, we detected a wide range of respiratory viruses (Flu B, HRV, AdV, etc) in this study.
 The shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.
 SARS-CoV-2 may be present in higher concentrations in air and may be more easily detected in bioaerosol samples due to differences in physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity) that influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}.
-In addition, although respiratory viruses have been detected in bioaerosols (\eg Flu\cite{Bischoff2013JID,Pan2017mSphere}, HRV\cite{Myatt2004AJRCCM,Fabian2011JAMPDD}, AdV\cite{Nguyen2017OFID,Pan2017mSphere}, and CoV\cite{Guo2020EID,Liu2020Nature}), bioaerosol sampling remains a technical challenge\cite{Wang2021 ,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
+In addition, although various respiratory viruses have been detected in bioaerosols (\eg Flu\cite{Bischoff2013JID,Pan2017mSphere}, HRV\cite{Myatt2004AJRCCM,Fabian2011JAMPDD}, AdV\cite{Nguyen2017OFID,Pan2017mSphere}, and CoV\cite{Guo2020EID,Liu2020Nature}), bioaerosol sampling remains a technical challenge\cite{Wang2021 ,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
 

### 2023-08-18 analysis/aud_analysis.Rmd
Adding results from cough analysis

---
title: "Cough Analysis"
author: "Nicolas Banholzer"
date: "2023-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(reshape2)
library(tidybayes)
library(lubridate)
library(rstanarm)
library(brms)
library(rstan)
source("../utils/epi.r")
source("../utils/bayes.r")
source("../utils/plotting.r")
source("../utils/tex.r")
```

## Data

```{r}
# load cough data 
coughs <- read.csv("../data-clean/coughs.csv", sep = ";") %>%
  mutate(class =  ifelse(class == "B", "A", "B")) %>% # correct wrong label by Simon who put Kontrollklasse 3a=A instead of Kontrollklasse 3a=B
  rename(n_coughs = `coughs.detected`) %>%
  mutate(datetime = as.POSIXct(paste(date, time), format = "%d.%m.%Y %H:%M:%S"),
         date = as.Date(datetime)) %>%
  dplyr::select(class, date, datetime, n_coughs)

# load time in the class room
time_in_room <- readRDS("../data-clean/time-in-room.rds") %>%
  dplyr::select(class, datetime, no_palas) %>%
  group_by(class) %>%
  complete(datetime = full_seq(datetime, 60)) %>%
  fill(no_palas, .direction = "down") %>%
  ungroup()

# filter time in room
coughs <- coughs %>%
  left_join(time_in_room) %>%
  filter(!no_palas)

# summarize coughs
coughs <- coughs %>%
  group_by(class, date) %>%
  summarize(n_coughs = sum(n_coughs),
            TT = n()) %>%
  ungroup()

# load epi data 
epi_data <- readRDS("../data-clean/epidemiological.rds") %>%
  dplyr::select(class, date, weekday, n_present, new_cases, aircleaner) %>%
  group_by(class) %>%
  arrange(date) %>%
  mutate(cases = cumsum(new_cases)) %>%
  ungroup() %>%
  dplyr::select(-new_cases)

# load env data
env_data <- readRDS("../data-clean/environmental.rds") %>%
  dplyr::select(class, date, AER)

# merge all together
coughs <- coughs %>%
  left_join(epi_data) %>%
  left_join(env_data)
```

## Descriptives

```{r}
# raw data
coughs %>%
  ggplot(aes(x = date, y = n_coughs / TT)) +
  geom_point(aes(color = aircleaner)) +
  geom_smooth() +
  facet_wrap(~ class)
hist(coughs$n_coughs/coughs$TT)

# global average
coughs %>%
  group_by(aircleaner) %>%
  summarize(sum(n_coughs) / sum(TT))

# weekly coughs per min by class and condition
coughs_pl <- coughs %>%
  mutate(week = isoweek(date)) %>%
  filter(week != 6) %>%
  mutate(week = ifelse(week < 6, week - min(week) + 1, week - min(week))) %>%
  group_by(class, week) %>%
  summarize(coughs_per_min = sum(n_coughs) / sum(TT),
            aircleaner = first(aircleaner)) %>%
  ungroup() %>%
  mutate(aircleaner = factor(aircleaner, levels = c("Yes", "No")),
         class = factor(class, levels = c("A", "B"))) %>%
  arrange(class, aircleaner) %>%
  ggplot(aes(x = factor(week), y = coughs_per_min, fill = aircleaner, pattern = class)) +
  geom_bar_pattern(stat = "identity", color = "black", pattern_fill = "white", pattern_angle = 45,
                   pattern_density = 0.25, pattern_spacing = 0.05, width = .5) +
  labs(x = "Study week", y = "Number of coughs per min", fill = "Air cleaner", pattern = "Class") +
  scale_fill_manual(values = rev(wes_palette("Royal1")[1:2])) +
  scale_x_discrete(breaks = 1:7) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = seq(0, 10, 2)) +
  scale_pattern_manual(values = c(A = "stripe", B = "none")) +
  theme_custom() +
  theme(legend.position = "top",
        legend.direction = "horizontal",
        legend.key.width = unit(0.5, "cm"),
        legend.key.height = unit(0.3, "cm")) +
        #plot.title.position = "plot") +
  guides(fill = guide_legend(nrow = 1, override.aes = list(pattern = "none", order = 2)),
         pattern = guide_legend(nrow = 1, override.aes = list(fill = "white"), order = 1))

coughs_pl

save_plot(coughs_pl, pdf_file = "../results/cough-data/coughs-frequency-per-week.pdf", eps_file = "../results/cough-data/coughs-frequency-per-week.eps", w = 12, h = 8)
```

## Model

### Main

```{r}
# prepare stan data
sdl <- list()
coughs_A <- coughs %>% filter(class == "A") %>% arrange(date)
coughs_B <- coughs %>% filter(class == "B") %>% arrange(date)
sdl$Da <- nrow(coughs_A)
sdl$Db <- nrow(coughs_B)
sdl$coughsA <- coughs_A$n_coughs
sdl$coughsB <- coughs_B$n_coughs
sdl$tA <- coughs_A$TT
sdl$tB <- coughs_B$TT
sdl$airCleanerA <- ifelse(coughs_A$aircleaner == 'Yes', 1, 0)
sdl$airCleanerB <- ifelse(coughs_B$aircleaner == 'Yes', 1, 0)
sdl$aerA <- (coughs_A$AER - mean(coughs$AER)) 
sdl$aerB <- (coughs_B$AER - mean(coughs$AER))
sdl$nA <- (coughs_A$n_present - mean(coughs$n_present)) 
sdl$nB <- (coughs_B$n_present - mean(coughs$n_present)) 
sdl$casesA <- (coughs_A$cases - mean(coughs$cases)) 
sdl$casesB <- (coughs_B$cases - mean(coughs$cases)) 
sdl$weekdayA <- coughs_A %>%
  mutate(weekday = factor(weekday, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))) %>%
  dplyr::select(date, weekday) %>%
  dcast(date ~ weekday) %>%
  mutate(across(-date, ~ ifelse(is.na(.x), 0, 1))) %>%
  dplyr::select(-date, -Monday) %>%
  as.matrix()
sdl$weekdayB <- coughs_B %>%
  mutate(weekday = factor(weekday, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))) %>%
  dplyr::select(date, weekday) %>%
  dcast(date ~ weekday) %>%
  mutate(across(-date, ~ ifelse(is.na(.x), 0, 1))) %>%
  dplyr::select(-date, -Monday) %>%
  as.matrix()
sdl$s_x <- c(sd(ifelse(coughs$aircleaner == "Yes", 1, 0)), # aircleaner
             sd(ifelse(coughs$aircleaner == "Yes", 1, 0)), # class
             apply(rbind(sdl$weekdayA, sdl$weekdayB), 2, sd),
             sd(coughs$n_present),
             sd(coughs$AER),
             sd(coughs$cases))

# model
stan_file <- "../models/negbinom-coughs-model.stan"
coughsMod <- stan(stan_file, data = sdl, seed = 12345, cores = 4)

# results
raw_res <- summary(coughsMod, probs = c(0.025, 0.5, 0.975), pars = c("phi", "beta0", "beta1", "beta2", "beta3", "beta4", "beta5", "beta6"))$summary %>%
  data.frame() %>%
  mutate_all(round, 2) 
raw_res

exp(raw_res["beta1",c("mean", "X2.5.", "X97.5.")])
exp(raw_res["beta2",c("mean", "X2.5.", "X97.5.")])
```

### 2023-08-18 analysis/epi_analysis.Rmd
Adding results from cough analysis

@@
   theme_bw2() +
-  theme(plot.title = element_text(hjust = 0), legend.position = "none")
+  theme(plot.title = element_text(hjust = 0), legend.position = "none",
+        panel.spacing = unit(15, units = "pt"), 
+        plot.margin = margin(r = 10))
 prior_pin_pl

### 2023-08-18 analysis/mol_analysis.Rmd
Adding results from cough analysis

@@
          variable = ifelse(variable == "SARS-CoV-2", "CoV", variable),
-         variable = factor(variable, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")))
+         variable = factor(variable, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")),
+         class = paste("Class", class)) %>%
+  arrange(class, week, desc(variable))
 
 pos_week_pl <- ggplot(mapping = aes(x = week, y = n)) +
-  geom_bar(data = df_pos %>% mutate(class = paste("Class", class)), mapping = aes(fill = variable),
-           stat = "identity") +
-  geom_text(data = subset(df_pos, n > 0) %>% mutate(class = paste("Class", class)), mapping = aes(label = n), 
+  facet_wrap(~ class) +
+  geom_bar(data = df_pos, mapping = aes(fill = variable),
+           stat = "identity", position = "stack") +
+  geom_text(data = subset(df_pos, n > 0), mapping = aes(label = n), 
             size = 8 / cm(1), position = position_stack(vjust = 0.5), color = "white") +
-  facet_wrap(~ class) +
   geom_text(
@@
   ) +
-
   scale_fill_manual(values = wes_palette("IsleofDogs1")) +

### 2023-08-18 doc/paper/supplements.tex
Updates from Overleaf

@@
 \title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
-Transmission of airborne respiratory infections with and without air cleaners in a Swiss school during non-pandemic conditions: A modeling study of epidemiological, environmental, and molecular data}}
+Transmission of respiratory viral infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}}
 
 % author list
-\author[1$\ddag$]{Nicolas Banholzer}
-\author[2]{Philipp Jent}
-\author[3]{Pascal Bittel}
-\author[3]{Lavinia Furrer}
+\author[1$\ddag$,2]{Nicolas Banholzer}
+\author[2,3]{Philipp Jent}
+\author[2,4]{Pascal Bittel}
+\author[4]{Lavinia Furrer}
+\author[4]{Alban Ramette}
+\author[4]{Ronald Dijkman}
+\author[5]{Jenna Kelly}
+\author[1]{Kathrin Zürcher}
 \author[1]{Matthias Egger}
-\author[4]{Tina Hascher}
-\author[1*]{Lukas Fenner}
+\author[2,6]{Tina Hascher}
+\author[1*,2]{Lukas Fenner}
 
 \affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
-\affil[2]{Department of Infectious Diseases, Inselspital, Bern University Hospital, University of Bern, Bern, Switzerland}
-\affil[3]{Institute of Infectious Diseases, University of Bern, Bern, Switzerland}
-\affil[4]{Institute of Educational Science, University of Bern, Bern, Switzerland}
+\affil[2]{Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland}
+\affil[3]{Department of Infectious Diseases, Inselspital, Bern University Hospital, University of Bern, Bern, Switzerland}
+\affil[4]{Institute of Infectious Diseases, University of Bern, Bern, Switzerland}
+\affil[5]{Interfaculty Bioinformatics Unit, University of Bern, Bern, Switzerland}
+\affil[6]{Institute of Educational Science, University of Bern, Bern, Switzerland}
 
@@
 
+\section{Modeling changes in cough frequency}\label{sec:aud-regression-model}
+
+We analyze the daily number of coughs $Y$ with a Negative Binomial regression model
+\begin{align*}
+    Y | \beta_0, \beta_1, \dots, \beta_6 &\sim \text{Negative-Binomial}(\mu,\phi) \\
+    \log \mu &= \log T + \beta_0 + \beta_1 \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Weekday} + \beta_4 \cdot \text{Students} + \beta_5 \cdot \text{Ventilation} + \beta_6 \cdot \text{Cases} \\
+    \beta_0 &\sim \text{Student-t}_5(0, 2.5) \\
+    \beta_1, \dots, \beta_6 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) \\
+    \frac{1}{\sqrt{\phi}} &\sim \text{Half-Normal}(0,1),
+\end{align*}
+where $T$ is the daily duration that students were in the classroom and $\phi$ is the parameter modeling over-dispersion in count data.
+
+\clearpage
+
 \section{Detailed results for risk of infection model}\label{sec:detailed-redcap}

### 2023-08-18 doc/paper/manuscript.tex

@@
 
-\title{\bfseries\LARGE\singlespacing{Transmission of airborne respiratory viral infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}}
+\title{\bfseries\LARGE\singlespacing{Transmission of respiratory viral infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}}
 
@@
 % author list
-\author[1$\ddag$]{Nicolas Banholzer}
-\author[2]{Philipp Jent}
-\author[3]{Pascal Bittel}
-\author[3]{Lavinia Furrer}
+\author[1$\ddag$,2]{Nicolas Banholzer}
+\author[2,3]{Philipp Jent}
+\author[2,4]{Pascal Bittel}
+\author[4]{Lavinia Furrer}
+\author[4]{Alban Ramette}
+\author[4]{Ronald Dijkman}
+\author[5]{Jenna Kelly}
 \author[1]{Kathrin Zürcher}
 \author[1]{Matthias Egger}
-\author[4]{Tina Hascher}
-\author[1*]{Lukas Fenner}
+\author[2,6]{Tina Hascher}
+\author[1*,2]{Lukas Fenner}
 
 \affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
-\affil[2]{Department of Infectious Diseases, Inselspital, Bern University Hospital, University of Bern, Bern, Switzerland}
-\affil[3]{Institute of Infectious Diseases, University of Bern, Bern, Switzerland}
-\affil[4]{Institute of Educational Science, University of Bern, Bern, Switzerland}
+\affil[2]{Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland}
+\affil[3]{Department of Infectious Diseases, Inselspital, Bern University Hospital, University of Bern, Bern, Switzerland}
+\affil[4]{Institute of Infectious Diseases, University of Bern, Bern, Switzerland}
+\affil[5]{Interfaculty Bioinformatics Unit, University of Bern, Bern, Switzerland}
+\affil[6]{Institute of Educational Science, University of Bern, Bern, Switzerland}
 
@@
 \noindent\textbf{Background:} Airborne respiratory infections can be transmitted in crowded indoor settings, including schools.
-Infection control measures such as air ventilation or filtration can reduce the risk of airborne transmission.
-The aim of this study was to evaluate the effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory viral infections in schools using a multiple-measurement approach.
+Infection control measures such as air ventilation or filtration are supposed to reduce the risk of airborne transmission.
+The aim of this study was to evaluate the real-life effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory viral infections in schools using a multiple-measurement approach.
 \medskip
@@
 The estimated risk of infection tended to be lower with air cleaners (adjusted risk ratio 0.72, 95\%-Credible Interval 0.44 to 1.15).
-\\
+\medskip
+
 % Study limitations include unknown specific causes of absences related to respiratory infections (i.e which virus caused the sickness).
 Furthermore, students occasionally had lessons outside the classroom, thus limiting the potential effectiveness of air cleaners.
-\medskip %\vspace{-1.3em}
 
-\noindent\textbf{Conclusions:} A wide range of respiratory viruses were detected  in students, but rarely SARS-CoV-2.
-Airborne molecular detection of respiratory viruses was rare, and when detected, it was not SARS-CoV-2.
-Our results show a reduction in aerosol concentrations with air cleaners and suggest a small benefit of air cleaners in reducing the transmission of airborne respiratory infections in school rooms.
+\noindent\textbf{Conclusions:} A wide range of non-SARS-CoV-2 respiratory viruses were detected in students, but airborne molecular detection was rare.
+SARS-CoV-2 was rarely detected in saliva and bioaerosols.
+Molecular transmission networks documented sustained transmission of non-SARS-CoV-2 seasonal viruses in school rooms despite negative airborne detection, suggesting the need for prolonged close contact.
+Air cleaners showed a small, potential benefit in reducing airborne transmission.
 
@@
 
+\subsection{Molecular genotyping}
+
+For adenovirus-positive samples (saliva, bioaerosol, and air filter samples), we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2020ClinicEpi}.
+The phylogenetic trees were constructed using the Neighbor-Joining method based on evolutionary distances computed using the number of differences method.
+The molecular transmission network for influenza B was determined by whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
+
+\subsection{Audio recordings and cough detection}
+
+We installed one microphone (RØDE NT-USB; Sydney, Australia) in each classroom to continuously record sounds.
+We detected the number of coughs per minute using an AI algorithm based on a convolutional neural network\cite{Bertschinger2023IEEE}, which was recently trained and validated with audio data from the waiting room of primary care clinic\cite{Zurcher2022}.
+
 \subsection{Statistical analyses and modeling}
@@
 Apart from that there were no major deviations from the SAP regarding the conducted analyses.
-There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in Text~\zref{sec:transmission-model} to Text~\zref{sec:env-regression-model} in \supp, which provide detailed descriptions of the models used in each analysis.
+There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in Supplementary Text~\zref{sec:transmission-model} to Supplementary Text~\zref{sec:env-regression-model} in \supp, which provide detailed descriptions of the models used in each analysis.
 
@@
 
-A detailed description of the model and choice of priors for all model parameters are provided in Text~\zref{sec:transmission-model} in \supp.
+A detailed description of the model and choice of priors for all model parameters are provided in Supplementary Text~\zref{sec:transmission-model} in \supp.
 Importantly, since multiple viruses were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the virus-specific distributions.
@@
 \noindent Molecular analysis determined which saliva samples were positive for a respiratory virus.
-We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (Text~\zref{sec:multinomial-model} in \supp) where the expected number of positive samples for a virus was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each virus.
+We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (Supplementary Text~\zref{sec:multinomial-model} in \supp) where the expected number of positive samples for a virus was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each virus.
 The latter was computed as the difference between the number of students per class and the cumulative number of positive samples.
 As a sensitivity analysis, we lagged the test results by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday), to consider the unknown delay when students were infected and tested positive.
+\medskip
 
@@
 \noindent We computed the average concentrations per day and compared these between study conditions.
-We estimated the reduction in concentrations using Bayesian log-linear regression models (Text~\zref{sec:env-regression-model} in \supp), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+We estimated the reduction in concentrations using Bayesian log-linear regression models (Supplementary Text~\zref{sec:env-regression-model} in \supp), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
 \medskip
+
+\noindent\textbf{Number of detected coughs} \smallskip
+
+We computed the daily number of detected coughs and compared it between study conditions.
+We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model (Supplementary Text~\zref{sec:aud-regression-model}) with time in class as the model offset, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
 
@@
 As a result, the estimated number of infections if air cleaners had been installed throughout the study is expected to be 36 (95\%-CrI 21 to 64) compared to 62 (95\%-CrI 29 to 141) infections if they would not have been installed (\Cref{fig:redcap-results}b).
-Detailed estimation results are provided in Text~\zref{sec:detailed-redcap} in \supp.
+Detailed estimation results are provided in Supplementary Text~\zref{sec:detailed-redcap} in \supp.
 
@@
 \end{figure}
+
+\subsection{Detected coughs}
+
+Slightly fewer coughs were detected during study weeks where air cleaners were installed (\Cref{fig:coughing}).
+On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners (adjusted risk ratio 0.93, 95\%-CrI 0.85 to 1.02).
+Significantly more coughs were detected in class~B (adjusted risk ratio 2.16, 95\%-CrI 1.90 to 2.46), suggesting a link with the observed variation in virus-specific transmission between classes (\Cref{fig:molecular-descriptives}c).
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{}
+    \caption{Caption}
+    \label{fig:coughing}
+\end{figure}
+
+\subsection{Molecular transmission networks for adenovirus and influenza B}
+
+xxxxxx
 
@@
 
-We collected epidemiological, environmental, and molecular data to estimate the risk of infection with respiratory infectious diseases in a Swiss school and assessed the effectiveness of air cleaners as an infection control measure.
+We collected epidemiological, environmental, and molecular data to estimate the risk of infection with respiratory viruses in a Swiss school and assessed the effectiveness of air cleaners as an infection control measure.
 We found a wide range of respiratory viruses in human saliva samples, primarily adenovirus, influenza B, and rhinovirus, but only few non-SARS-CoV-2 respiratory viruses were detected in bioaerosol samples and on the filters of air cleaners.
-Aerosol number and particle mass concentrations dropped significantly with air cleaners.
-Bayesian modeling showed conflicting results between the analysis of molecular data and absences related to respiratory infections.
-While the positivity rate of human saliva samples was comparable between study conditions, the relative risk of infection based on absences tended to be lower with air cleaners.
+Aerosol number and particle mass concentrations decreased significantly with air cleaners and Bayesian modeling based on the absences of students showed a small reduction in the relative risk of infection with air cleaners.
+Reduced intensity of symptoms (coughing) in school, as determined by acoustic recording, indicates small prevention of symptomatic infections by air cleaners.
 
@@
 We argued that air cleaners may have had less potential to reduce infections because they were implemented at the end of the study when most students had already been infected with SARS-CoV-2.
-In this study, we used a cross-over design to estimate the effectiveness of air cleaners.
+In this study, we used a cross-over design to estimate the effectiveness of air cleaners again, but this time under non-pandemic conditions.
 We found similar reductions in particle concentrations and a tendency that the relative risk of infection was lower with air cleaners.
 To some extent, the differences between the two studies may be explained by different adjustments.
-While we could not adjust for ventilation in our previous study, the estimates of this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see Text~\zref{sec:detailed-molecular} in \supp).
-Furthermore, both studies were conducted during different times and the distribution of respiratory viruses varied considerably.
-During the COVID-19 pandemic (previous study), we could almost only detect SARS-CoV-2, whereas a range of respiratory viruses was detected during non-pandemic times (this study).
-
-% why air cleaners may not have been so effective
-
-Our study indicated that air cleaners can reduce the risk of infection, although we we only were able tho show a trend.
-Other infection control measures, such as universal mask use, appear to be more effective\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021}.
-There may be several reasons for this.
-Compared to \eg mask wearing, air cleaners cannot influence transmission outside the classrooms.
-Furthermore, air cleaners are unable to prevent close range high particle density aerosol transmission.
-By contrast, face masks can also reduce the generation of airborne infectious particles and can thus mitigate transmission over short ranges.
-Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective in the removal of the larger particles ($>5\mu$m) than the smaller particles ($\leq5\mu$m).
-Yet, a larger proportion of respiratory viruses are carried in the smaller particles\cite{Fennelly2020}.
+While we could not adjust for ventilation in our previous study, the estimates of this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see Supplementary Text~\zref{sec:detailed-molecular} in \supp).
+Furthermore, the distribution of respiratory viruses in this study was very different from the distribution in our previous study.
 
@@
 
-We detected few respiratory viruses in bioaerosol samples (3~positive bioaoerosl and 50~positive saliva samples), especially when comparing to our previous study\cite{Banholzer2023PLoSMed} (10~positive bioaerosol and 21~positive saliva samples).
-The number of viruses detected on the filters of the air cleaners was also lower (this study: 4~positive samples in 160~swabs, previous study: 8~positive samples in 80~swabs).
-The bioaerosol sampling devices were the same in both studies and technical issues were not observed.
-It is also unlikely that environmental factors are responsible for the differences.
-Temperature and humidity were similar, and CO$_2$ levels were even higher in this study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
-Lower ventilation was potentially offset by a larger reduction in aerosol and particle concentrations through air cleaners in this study.
-However, it could be that air cleaners removed particles less quickly than ventilation.
-The most striking difference is in the distribution of respiratory viruses.
-Compared to the COVID-19 pandemic, we detected a wide range of respiratory viruses (Flu B, HRV, AdV, etc) in this study.
+We detected fewer respiratory viruses in bioaerosol samples compared to our previous study\cite{Banholzer2023PLoSMed} (1~sample of adenovirus and 1~sample of rhinovirus in this study vs 10~samples of SARS-CoV-2 in the previous study), especially in relation to the number of positive saliva samples (50~samples in this study vs 21~in the previous study).
+Similarly, the number of viruses detected on the filters of the air cleaners was also lower in this study (4~positive samples in 160~swabs vs 8~positive samples in 80~swabs in the previous study).
+Importantly, there was a clear shift in the distribution of respiratory viruses in the current study compared to the previous study in 2022 during the SARS-CoV-2 Omicron variant.
+Whereas in 2022 we detected almost exclusively SARS-CoV-2 in human saliva, the most common respiratory viruses in 2023 were adenovirus, influenza~B, and rhinoviruses (only three SARS-CoV-2-positive samples).
 The shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.
-SARS-CoV-2 may be present in higher concentrations in air and may be more easily detected in bioaerosol samples due to differences in physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity) that influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}.
-In addition, although various respiratory viruses have been detected in bioaerosols (\eg Flu\cite{Bischoff2013JID,Pan2017mSphere}, HRV\cite{Myatt2004AJRCCM,Fabian2011JAMPDD}, AdV\cite{Nguyen2017OFID,Pan2017mSphere}, and CoV\cite{Guo2020EID,Liu2020Nature}), bioaerosol sampling remains a technical challenge\cite{Wang2021 ,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
+xxxxx
+Technical reasons for these differences are unlikely.
+The bioaerosol sampling devices and laboratory procedures were the same in both studies and no technical issues were observed.
+It is also unlikely that environmental factors inhibited airborne detection.
+Temperature and humidity were similar, and CO$_2$ levels were even higher in this study (1,702$\pm$370\,ppm vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic, although lower ventilation was partially offset by a greater reduction in aerosol and particle concentrations from air cleaners.
+However, the removal of particles through air cleaners may be slower than through ventilation.
+
+% bioaerosol sampling in more depth
+
+we speculate that the difference in airborne molecular detection of respiratory viruses is due to the characteristics of the respiratory viruses.
+Differences in physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity), which influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}, could make it easier to detect SARS-CoV-2 in the air.
+The reason is that viral loads do often not exceed the detection limits of existing devices for bioaerosol sampling, which remains a challenge\cite{Wang2021,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
+For example, one study was unable to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.
+Nevertheless, previous studies show that also influenza\cite{Bischoff2013JID,Pan2017mSphere}, rhinovirus\cite{Myatt2004AJRCCM}, and adenovirus\cite{Nguyen2017OFID,Pan2017mSphere} can be detected in airborne samples.
+The ability to detect pathogens in bioaerosols could also be related to patient characteristics of the study participants.
+Multiple studies suggest an association between the generation of pathogenic aerosols and infectiousness\cite{Leung2020NatMed,Bischoff2013JID,Escombe2008PLoSMed}.
+Therefore, studies including more infectious patients (intentionally or unintentionally) will probably have a higher chance of detecting respiratory viruses in the air.
 
@@
 
+Molecular transmission networks based on targeted sequencing or whole genome sequencing documented transmission in the classrooms.
+xxxx (influenza b, adeno, etc.).
 The low rate of positive bioaerosol samples also means that it is unclear to what extent students airborne transmission occurred in classrooms.
@@
 However, the distribution of saliva samples between classes shows remarkable differences.
-AdV was spreading entirely in class~A in the first three study weeks and only two infections with Flu~B occurred over the study.
-By contrast, Flu~B was spreading throughout the study in class~B and infections with AdV were only detected in the last study week.
-These class-specific patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.
+Adenovirus was spreading entirely in class~A in the first three study weeks and only two infections with influenza~B occurred over the study.
+By contrast, influenza~B was spreading throughout the study in class~B and infections with adenovirus were only detected in the last study week.
+Adenovirus infections often tend to be mild\cite{Kunz2010CIDR} and less associated with coughing than influenza\cite{Ma2018RMV}, which is in line with the lower frequency of detected coughs in class~B.
+Altogether, the class-specific patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.
+
+% effects of air cleaners
+
+Our study indicated a small, potential benefit of air cleaners, but other infection control measures such as universal mask wearing appear to be more effective\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Dharmadhikari2012AJRCCM,Leung2020NatMed,Milton2013PLoSPathogens}.
+There may be several reasons for this.
+Compared to \eg mask wearing, air cleaners cannot influence transmission outside the classrooms.
+Furthermore, previous work suggests that prolonged close contact may be needed for transmission\cite{Leung2020NatMed,Brankston2007LancetIDw}.
+However, compared to face masks, air cleaners are unable to prevent such close range high particle density aerosol transmission.
+Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective in the removal of larger particles ($>5\mu$m).
+Yet, a larger proportion of respiratory viruses are carried in the smaller particles ($\leq5\mu$m)\cite{Fennelly2020}.
+Finally, classroom activity, airflow and other unobserved, confounding factors make it difficult to assess the effects of air cleaners.
+Nevertheless, it is important to assess their effectiveness in a real-life setting, and compare with the hypothetical impact suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
 
@@
 In conclusion, using epidemiological, environmental, and molecular data, our study shows that considerable transmission of respiratory viruses other than SARS-CoV-2 occurred in school rooms under non-pandemic conditions.
-Air cleaners significantly reduced aerosol concentrations, but airborne respiratory viruses were rarely detected and if only viruses other than SARS-CoV-2.
-The epidemiologic transmission model estimated a trend toward a reduction in infections with air filters.
+Air cleaners significantly reduced aerosol concentrations, but airborne respiratory viruses were rarely detected, suggesting that prolonged close contact may be required for transmission.
+The epidemiologic transmission model estimated a tendency towards lower transmission with air filters.
 Future studies may use our multiple-measurement approach to assess whether professional building ventilation, which should be preferred over air cleaners in the long term, and other infection control measures reduce the risk of airborne transmission.
-Whether their effectiveness varies with the seasonal distribution of respiratory viruses under non-pandemic conditions remains to be investigated.
+Whether their effectiveness varies with the seasonal distribution of respiratory viruses under non-pandemic conditions also remains to be investigated.
 
@@
 We would like to thank the schools, teachers, and students participating in the study.
-We are also grateful to the Educational Department of the Canton of Solothurn for their support throughout the study.
+We are grateful to the Educational Department of the Canton of Solothurn for their support throughout the study.
+We also would like to thank the biosecurity team at the Institute for Infectious Diseases of the University of Bern for their assistance with the bioaerosol devices (Kathrina Summermatter, Julia Feldmann, and Monika Gsell).
 Finally, we are indebted to the student assistants who helped with the data collection at the schools.

### 2023-08-22 analysis/aud_analysis.Rmd
Add association between coughing and molecular tests|

@@
 
+# load mol data
+mol_data <- readRDS("../data-clean/molecular.rds") %>%
+  dplyr::select(-weekday, -aircleaner, -N, -n_class)
+
 # merge all together
@@
                    pattern_density = 0.25, pattern_spacing = 0.05, width = .5) +
-  labs(x = "Study week", y = "Number of coughs per min", fill = "Air cleaner", pattern = "Class") +
+  labs(x = "Study week", y = "Number of coughs per min", fill = "Air cleaner", pattern = "Class", title = "A") +
   scale_fill_manual(values = rev(wes_palette("Royal1")[1:2])) +
@@
         legend.key.width = unit(0.5, "cm"),
-        legend.key.height = unit(0.3, "cm")) +
-        #plot.title.position = "plot") +
+        legend.key.height = unit(0.3, "cm"),
+        plot.title.position = "plot") +
   guides(fill = guide_legend(nrow = 1, override.aes = list(pattern = "none", order = 2)),
@@
 ```
+
+### Association with molecular data
+
+```{r}
+# subset data
+coughs_TueThu <- mol_data %>%
+  left_join(dplyr::select(coughs, class, date, n_coughs, TT)) %>%
+  left_join(epi_data) %>%
+  left_join(env_data) %>%
+  dplyr::select(-Neg) %>%
+  left_join(coughs) %>%
+  filter(!is.na(n_coughs))
+
+# create stan data
+sdl_acm <- list(
+  N = nrow(coughs_TueThu),
+  C = 2,
+  V = 6,
+  coughs = coughs_TueThu$n_coughs,
+  TT = coughs_TueThu$TT,
+  viruses = as.matrix(dplyr::select(coughs_TueThu, `Flu B`, HRV, AdV, `SARS-CoV-2`, MPV, PIV))
+)
+sdl_acm$s_v <- mean(apply(sdl_acm$viruses, 2, sd))
+
+# model fit
+stan_file_acm <- "../models/negbinom-cough-molecular-model.stan"
+coughsMolMod <- stan(stan_file_acm, data = sdl_acm, seed = 12345, cores = 4)
+
+# results
+raw_res_acm <- summary(coughsMolMod, probs = c(0.025, 0.5, 0.975), pars = c("phi", "beta0", "theta", "theta_v", "tau"))$summary %>%
+  data.frame() %>%
+  mutate_all(round, 2) 
+raw_res_acm
+
+# plot
+viruses <- c("Average", "Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")
+names(viruses) <- 0:6
+draws_theta <- coughsMolMod %>% spread_draws(theta_v[v]) %>%
+  mutate(v = recode(v, !!! viruses),
+         v = factor(v, levels = viruses),
+         rr = exp(theta_v))
+
+draws_theta %>%
+  dplyr::select(v, rr) %>%
+  group_by(v) %>%
+  mean_qi() %>%
+  ungroup() %>%
+  mutate_if(is.numeric, round, 2)
+
+theta_pl <- draws_theta %>%
+  ggplot(aes(y = v, x = rr, color = v)) +
+  geom_vline(aes(xintercept = 1), linetype = "dotted", color = "red") +
+  stat_pointinterval(shape = 21, fill = "white") +
+  scale_color_manual(values = wes_palette("IsleofDogs1")) +
+  labs(x = "Relative risk ratio", title = "B") +
+  theme_custom() +
+  theme(axis.title.y = element_blank(), legend.position = "none",
+        plot.title.position = "plot")
+
+theta_pl
+save_plot(theta_pl, pdf_file = "../results/cough-data/coughs-virus-association.pdf", eps_file = "../results/cough-data/coughs-virus-association.eps", w = 12, h = 8)
+```

### 2023-08-22 doc/paper/manuscript.tex

@@
 
-% \begin{information}\normalfont
+\begin{information}\normalfont
+
+xxxx
+
 % \noindent\textbf{Running head}: SARS-COV-2 transmission in schools and effect of air cleaners
@@
 % \par
-% \end{information}
+\end{information}
 
@@
 We installed one microphone (RØDE NT-USB; Sydney, Australia) in each classroom to continuously record sounds.
-We detected the number of coughs per minute using an AI algorithm based on a convolutional neural network\cite{Bertschinger2023IEEE}, which was recently trained and validated with audio data from the waiting room of primary care clinic\cite{Zurcher2022}.
+We detected the number of coughs per minute using an AI algorithm based on a convolutional neural network\cite{Bertschinger2023IEEE}, which was recently trained and validated with audio data from the waiting room of a primary care clinic\cite{Zurcher2022}.
 
@@
 
-We computed the daily number of detected coughs and compared it between study conditions.
+\noindent We computed the daily number of detected coughs and compared it between study conditions.
 We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model (Supplementary Text~\zref{sec:aud-regression-model}) with time in class as the model offset, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+\medskip
 
@@
     \centering
-    \includegraphics{}
-    \caption{Caption}
+    \includegraphics{results/cough-data/coughs-frequency-per-week.pdf}
+    \caption{Daily average number of detected coughs per minute by study week, condition and class.}
     \label{fig:coughing}
@@
 xxxxx
-Technical reasons for these differences are unlikely.
+
+% other reasons
+
+Technical reasons for the differences in airborne detection are unlikely.
 The bioaerosol sampling devices and laboratory procedures were the same in both studies and no technical issues were observed.
-It is also unlikely that environmental factors inhibited airborne detection.
-Temperature and humidity were similar, and CO$_2$ levels were even higher in this study (1,702$\pm$370\,ppm vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic, although lower ventilation was partially offset by a greater reduction in aerosol and particle concentrations from air cleaners.
-However, the removal of particles through air cleaners may be slower than through ventilation.
+Instead, airborne detection could have been influenced by environmental factors.
+Although temperature (between 19 and 25$^{\circ}$C) and relative humidity (between 25 and 50\%) were similar in both studies, the same levels can have a different effect on airborne survival of respiratory viruses.
+For example, for rhinovirus and adenovirus, survival increases with relative humidity, while the reverse relationship has been documented for influenza and SARS-CoV-2\cite{Tellier2009JTRSI,Ahlawat2020AAQR,Biryukov2020mS,Karim1985CJM,Davis1971AM}.
+A change in ventilation conditions could have had a profound effect, but CO$_2$ levels were even higher in this study (1,702$\pm$370\,ppm vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
+Although lower ventilation was partially offset by a greater reduction in aerosol and particle concentrations from air cleaners, the removal of particles through air cleaners may be slower than through ventilation.
 
@@
 
-we speculate that the difference in airborne molecular detection of respiratory viruses is due to the characteristics of the respiratory viruses.
+As a consequence, we speculate that the difference in airborne molecular detection of respiratory viruses is due to the characteristics of the respiratory viruses.
 Differences in physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity), which influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}, could make it easier to detect SARS-CoV-2 in the air.
@@
 Compared to \eg mask wearing, air cleaners cannot influence transmission outside the classrooms.
-Furthermore, previous work suggests that prolonged close contact may be needed for transmission\cite{Leung2020NatMed,Brankston2007LancetIDw}.
+Furthermore, previous work suggests that prolonged close contact may be needed for transmission\cite{Leung2020NatMed,Brankston2007LancetID}.
 However, compared to face masks, air cleaners are unable to prevent such close range high particle density aerosol transmission.

### 2023-08-22 analysis/aud_analysis.Rmd
minor graphical changes

@@
                    pattern_density = 0.25, pattern_spacing = 0.05, width = .5) +
-  labs(x = "Study week", y = "Number of coughs per min", fill = "Air cleaner", pattern = "Class", title = "A") +
+  labs(x = "Study week", y = "Number of coughs per min", fill = "Air cleaner", pattern = "Class", title = "a") +
   scale_fill_manual(values = rev(wes_palette("Royal1")[1:2])) +
@@
   scale_color_manual(values = wes_palette("IsleofDogs1")) +
-  labs(x = "Relative risk ratio", title = "B") +
+  labs(x = "Relative risk", title = "b") +
   theme_custom() +

### 2023-08-22 analysis/mol_analysis.Rmd
minor graphical changes

@@
   scale_size_continuous(range = c(1, 9)) +
-  labs(x = "Relative risk ratio", y = "Lag of sampling results", fill = "CrI") +
+  labs(x = "Relative risk", y = "Lag of sampling results", fill = "CrI") +
   theme_custom() +

### 2023-08-22 analysis/study_comaprison.Rmd
minor graphical changes

@@
   geom_point(mapping = aes(x = Mean), shape = 23, size = 2, fill = "white") +
-  labs(y = "Study year", x = "Adjusted risk ratio (%)", title = "d") + 
+  labs(y = "Study year", x = "Adjusted relative risk (%)", title = "d") + 
   scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) +

### 2023-08-22 doc/paper/manuscript.tex
Updates from Overleaf

@@
 Molecular analysis detected multiple viruses in saliva samples throughout the study (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus), but rarely in bioaerosol samples (2/105 positive: 1~rhinovirus and 1~adenovirus) and on the filters of air cleaners (4/160~positive: 1~influenza~B, 1~HRV, 1~adenovirus, and 1~SARS-CoV-2).
-The detection rate of positive saliva samples was comparable between study conditions (adjusted risk ratio 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61).
+The detection rate of positive saliva samples was comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61).
 Daily average CO$_2$ levels were 1,769$\pm$391\,ppm ($\pm$ standard deviation) with air cleaners and 1,636$\pm$341\,ppm without air cleaners, corresponding to similar air change rates (0.74$\pm$0.41\,h$^{-1}$ vs.
@@
 21~absences without air cleaners).
-The estimated risk of infection tended to be lower with air cleaners (adjusted risk ratio 0.72, 95\%-Credible Interval 0.44 to 1.15).
+The estimated risk of infection tended to be lower with air cleaners (adjusted relative risk 0.72, 95\%-Credible Interval 0.44 to 1.15).
 \medskip
@@
 We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model (Supplementary Text~\zref{sec:aud-regression-model}) with time in class as the model offset, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+In addition, we used a Bayesian Negative Binomial hierarchical (random effects) regression model to estimate the association between the number of coughs and the number of positive saliva test results for respiratory viruses.
 \medskip
@@
 There was an equal number of positive samples with and without air cleaners (\Cref{fig:molecular-descriptives}b).
-Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted risk ratio 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (Figure~\zref{fig:mol-estimation-results-sensitivity} in \supp).
+Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (Figure~\zref{fig:mol-estimation-results-sensitivity} in \supp).
 The distribution of positive samples varied between classes (\Cref{fig:molecular-descriptives}c).
@@
 Except for one study week, there was an equal or higher number of cases without air cleaners (\Cref{fig:redcap-results}a).
-Based on our Bayesian latent hierarchical regression model, we estimated a trend towards a lower risk of infection with air cleaners (adjusted risk ratio 0.72, 95\%-Credible Interval 0.44 to 1.15).
+Based on our Bayesian latent hierarchical regression model, we estimated a trend towards a lower risk of infection with air cleaners (adjusted relative risk 0.72, 95\%-Credible Interval 0.44 to 1.15).
 As a result, the estimated number of infections if air cleaners had been installed throughout the study is expected to be 36 (95\%-CrI 21 to 64) compared to 62 (95\%-CrI 29 to 141) infections if they would not have been installed (\Cref{fig:redcap-results}b).
@@
     \includegraphics[width=\linewidth]{../../results/epi-data/avoided-infections.pdf}
-    \caption{\textbf{Analysis of epidemiological data and estimated relative risk of infection with air cleaners.}.
+    \caption{\textbf{Analysis of epidemiological data and estimated number of infections with and without air cleaners.}.
 \textbf{(a)}~Weekly number of respiratory cases with and without air cleaners.
@@
 
-Slightly fewer coughs were detected during study weeks where air cleaners were installed (\Cref{fig:coughing}).
-On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners (adjusted risk ratio 0.93, 95\%-CrI 0.85 to 1.02).
-Significantly more coughs were detected in class~B (adjusted risk ratio 2.16, 95\%-CrI 1.90 to 2.46), suggesting a link with the observed variation in virus-specific transmission between classes (\Cref{fig:molecular-descriptives}c).
+Slightly fewer coughs were detected during study weeks where air cleaners were installed (\Cref{fig:coughing}a).
+On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners (adjusted relative risk 0.93, 95\%-CrI 0.85 to 1.02).
+Significantly more coughs were detected in class~B (adjusted relative risk 2.16, 95\%-CrI 1.90 to 2.46), suggesting a link with the observed variation in virus-specific transmission between classes (\Cref{fig:molecular-descriptives}c).
+This link can also be seen in the association between the frequency of coughing and the number of positive saliva tests for respiratory viruses (\Cref{fig:coughing}b).
+Coughing was more frequent for Flu~B (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent with AdV (0.87, 95\%-CrI 0.49 to 1.12).
 
@@
     \includegraphics{results/cough-data/coughs-frequency-per-week.pdf}
-    \caption{Daily average number of detected coughs per minute by study week, condition and class.}
+    \includegraphics{results/cough-data/coughs-virus-association.pdf}
+    \caption{\textbf{Analysis of coughing data.} \textbf{(a)}~Daily average number of detected coughs per minute by study week, condition and class.
+\textbf{(b)}~Estimated association (relative risk; posterior mean as dot, 50\%, 80\%, and 95\%-CrI as lines) between cough frequency and number of positive saliva test results by respiratory virus.}
     \label{fig:coughing}
@@
 \textbf{(c)}~Estimated reduction in aerosol number and particle mass concentrations (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$) with air cleaners (posterior mean as squares and 95\%-CrI as lines).
-\textbf{(d)}~Adjusted infection risk ratio (posterior mean as squares and 95\%-CrI as lines).}
+\textbf{(d)}~Adjusted relative risk of infection (posterior mean as squares and 95\%-CrI as lines).}
     \label{fig:comparison}
@@
 By contrast, influenza~B was spreading throughout the study in class~B and infections with adenovirus were only detected in the last study week.
-Adenovirus infections often tend to be mild\cite{Kunz2010CIDR} and less associated with coughing than influenza\cite{Ma2018RMV}, which is in line with the lower frequency of detected coughs in class~B.
+Adenovirus infections often tend to be mild\cite{Kunz2010CIDR} and less associated with coughing than influenza\cite{Ma2018RMV}, which is in line with the comparably lower frequency of detected coughs in class~B.
 Altogether, the class-specific patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.

### 2023-08-22 doc/paper/supplements.tex
Updates from Overleaf

@@
 
+In addition, we analyze the association between the daily number of coughs $Y$ and the number of positive saliva test resutls for respiratory viruses with a Negative Binomial hierarchical regression model
+\begin{align*}
+    Y | \beta_0, \theta_{[v]}, \theta, \tau &\sim \text{Negative-Binomial}(\mu,\phi) \\
+    \log \mu &= \log T + \beta_0 + \sum_v \theta_{[v]} \cdot \text{Saliva}_v \\
+    \beta_0 &\sim \text{Student-t}_5(0, 2.5) \\
+    \theta_{[\text{Flu~B}]}, \dots, \theta_{[\text{PIV}]} &\sim \text{Normal}\left(\theta, \tau\right) \\
+    \theta &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{v}}\right) \\
+    \tau &\sim \text{Half-Normal}\left(0, 1\right) \\
+    \frac{1}{\sqrt{\phi}} &\sim \text{Half-Normal}(0,1),
+\end{align*}
+where $\theta_{[v]}$ is the partially pooled estimate for respiratory virus $v \in$ (Flu~B, HRV, AdV, CoV, MPV, PIV), $\tau$ estimates variation between viruses, and $s_v$ is the average standard deviation across the count variables of the respiratory viruses.
+
 \clearpage
@@
     \caption[Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples]{Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples.
-Shown is the adjusted relative risk ratio (posterior mean as dot and 50\%-, 80\%, and 95\%-CrIs as lines) for air cleaners by lag of the molecular test results.}
+Shown is the adjusted relative risk (posterior mean as dot and 50\%-, 80\%, and 95\%-CrIs as lines) for air cleaners by lag of the molecular test results.}
     \label{fig:mol-estimation-results-sensitivity}

### 2023-10-03 analysis/mol_analysis.Rmd
add transmission chain plot

@@
 library(lubridate)
+library(scales)
 library(rstan)
@@
 df_ll <- readRDS("../data-clean/molecular-long.rds")
+df_network <- rbind(readxl::read_excel("../data-clean/transmission-network.xlsx", sheet = "A"),
+                    readxl::read_excel("../data-clean/transmission-network.xlsx", sheet = "B"))  
 ```
@@
              hline.after = NULL)
+```
+
+### Transmission network
+
+```{r}
+tnet <- df_network %>%
+  mutate(date_infection = as.POSIXct(paste0(date_infection, "00:00:00")),
+         virus = ifelse(virus == "SARS-CoV-2", "CoV", virus),
+         virus = factor(virus, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")),
+         class = paste("Class", class)) %>%
+  group_by(id, class, date_infection, virus, chain) %>%
+  summarize(n = n()) %>%
+  ungroup() %>%
+  mutate(n = factor(n, levels = c(1,2,3))) %>%
+  group_by(class, chain) %>%
+  arrange(date_infection) %>%
+  mutate(from = date_infection + hours(12),
+         to = lead(date_infection) - hours(12)) %>%
+  ungroup()
+
+coord_y_datetime <- function(xlim = NULL, ylim = NULL, expand = TRUE) {
+  if (!is.null(ylim)) {
+    ylim <- lubridate::as_datetime(ylim)
+  }
+  ggplot2::coord_cartesian(xlim = xlim, ylim = ylim, expand = expand)
+}
+
+tnet_pl <- tnet %>%
+  ggplot(aes(y = date_infection, x = virus, fill = virus, size = n)) +
+  facet_wrap(~ class, ncol = 2, scales = "free_x") +
+  geom_point(shape = 21, color = "black") +
+  geom_segment(aes(y = from, yend = to, xend = virus), size = 1,
+               arrow = arrow(length = unit(0.2, "cm")), color = "black") +
+  labs(size = "No. of infections", fill = "Virus", title = "c") +
+  coord_y_datetime(ylim = c(max(tnet$date_infection), min(tnet$date_infection))) +
+  scale_y_datetime(date_breaks = "1 week", date_labels = "%b %d") +
+  scale_size_discrete(range = c(4,8)) +
+  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
+  theme_custom() +
+  theme(legend.position = "none",
+        plot.title.position = "plot",
+        axis.title = element_blank(),
+        #axis.text.x = element_blank(),
+        #axis.ticks.x = element_blank(),
+        panel.spacing = unit(1, "cm"),
+        panel.grid.major.x = element_line(),
+        panel.grid.minor.y = element_blank(),
+        strip.text.x = element_text(size = 10, margin = margin(b = 10))) #+
+  #guides(fill = "none")
+
+#save_plot(tnet_pl, "../results/mol-data/network-plot.pdf", w = 16, h = 20)
 ```
@@
 descr <- descr + plot_layout(guides = "collect", widths = c(9.5, 1, 5.5), ncol = 2) 
-descr <- descr + pos_week_pl 
-descr <- descr + plot_layout(nrow = 2, heights = c(8, 8), design = 'ABC
+descr <- descr + tnet_pl 
+descr <- descr + plot_layout(nrow = 2, heights = c(4, 16), design = 'ABC
                              DDD')
-save_plot(descr, pdf_file = "../results/mol-data/descriptives.pdf", eps_file = "../results/mol-data/descriptives.eps", h = 16, w = 16)
+save_plot(descr, pdf_file = "../results/mol-data/descriptives.pdf", eps_file = "../results/mol-data/descriptives.eps", h = 24, w = 16)
 # saveRDS(pos_tot_pl + ggtitle(""), file = "../results/mol-data/descriptives-total.rds")

### 2023-10-03 preprocessing/prep_palas.r

@@ -187,6 +187,65 @@ aer %>%
   geom_point() +
   geom_smooth()
 
+# check entire day
+df %>%
+  mutate(hh = hour(datetime),
+         mm_n = minute(datetime) / 60,
+         hhmm_n = hh + mm_n,
+         class = paste("Class", class)) %>%
+  ggplot(aes(x = hhmm_n, y = CO2, group = date)) +
+  facet_wrap(~class, scales = "free_x") +
+  geom_line(color = "black", alpha = .5) +
+  geom_vline(aes(xintercept = 7.3), color = "red", linetype = "dashed") +
+  annotate("text", x = 8, y = 3600, color = "red", label = "School start", size = 10 / cm(1), hjust = 0) +
+  scale_x_continuous(expand = c(0,0), breaks = seq(0, 24, 4)) +
+  scale_y_continuous(expand = c(0,0), breaks = seq(500, 4000, 500)) +
+  labs(y = expression("CO"[2]*" (ppm)")) +
+  theme_bw() +
+  theme(axis.title.x = element_blank())
+ggsave("results/co2-full-time.png", width = 16 / cm(1), height = 8 / cm(1))
+
+# full day AER
+vent_dat_full <- df %>%
+  group_by(class, date) %>%
+  slice(1, n()) %>%
+  dplyr::select(class, date, datetime) %>%
+  mutate(datetime = map(date, function(x) seq.POSIXt(first(datetime), last(datetime), by = "1 min"))) %>%
+  unnest(datetime) %>%
+  left_join(df %>% dplyr::select(class, datetime, CO2, no_palas)) %>%
+  mutate(no_palas = ifelse(is.na(no_palas), T, no_palas)) %>%
+  group_by(class, date) %>%
+  arrange(datetime) %>%
+  mutate(CO2 = zoo::na.approx(CO2, na.rm = F, maxgap = Inf)) %>%
+  tidyr::fill(CO2, .direction = "downup") %>%
+  ungroup() %>%
+  left_join(epi_dat %>% dplyr::select(date, class, n_present)) %>%
+  mutate(n_present = ifelse(no_palas, 0, n_present)) %>%
+  group_by(class, date) %>%
+  arrange(datetime) %>%
+  mutate(n_present_imp = linear_impute(n_present, no_palas)) %>%
+  ungroup() %>% 
+  mutate(n_present_imp = ifelse(!no_palas, n_present_imp + 1, n_present_imp),
[…]

### 2023-10-03 doc/paper/manuscript.tex

@@
 In addition, we used a Bayesian Negative Binomial hierarchical (random effects) regression model to estimate the association between the number of coughs and the number of positive saliva test results for respiratory viruses.
+For this analysis, we only used the days were saliva testing was performed and coughs could be detected because people were in the classroom (27~days in total).
 \medskip
@@
 Instead, airborne detection could have been influenced by environmental factors.
-Although temperature (between 19 and 25$^{\circ}$C) and relative humidity (between 25 and 50\%) were similar in both studies, the same levels can have a different effect on airborne survival of respiratory viruses.
+Although temperature (overall between 19 and 25$^{\circ}$C) and relative humidity (overall between 25 and 50\%) were similar in both studies, the same levels can have a different effect on airborne survival of respiratory viruses.
 For example, for rhinovirus and adenovirus, survival increases with relative humidity, while the reverse relationship has been documented for influenza and SARS-CoV-2\cite{Tellier2009JTRSI,Ahlawat2020AAQR,Biryukov2020mS,Karim1985CJM,Davis1971AM}.
@@
 Compared to \eg mask wearing, air cleaners cannot influence transmission outside the classrooms.
-Furthermore, previous work suggests that prolonged close contact may be needed for transmission\cite{Leung2020NatMed,Brankston2007LancetID}.
+Furthermore, previous work suggests that prolonged close contact may be needed for transmission\cite{Leung2020NatMed,Brankston2007LancetID}, or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
 However, compared to face masks, air cleaners are unable to prevent such close range high particle density aerosol transmission.

### 2023-10-03 analysis/mol_analysis.Rmd
revise transmission chain plot

@@
                              DDD')
-save_plot(descr, pdf_file = "../results/mol-data/descriptives.pdf", eps_file = "../results/mol-data/descriptives.eps", h = 24, w = 16)
+save_plot(descr, pdf_file = "../results/mol-data/descriptives.pdf", eps_file = "../results/mol-data/descriptives.eps", h = 22, w = 16)
 # saveRDS(pos_tot_pl + ggtitle(""), file = "../results/mol-data/descriptives-total.rds")

### 2023-10-03 doc/paper/manuscript.tex

@@
 Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (Figure~\zref{fig:mol-estimation-results-sensitivity} in \supp).
-The distribution of positive samples varied between classes (\Cref{fig:molecular-descriptives}c).
+The distribution of positive samples varied between classes.
 For example, in the first three study weeks all except for one sample were positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
 The virus-specific variations between classes are suggestive of transmission occurring within classes.
+Possible transmission chains within classes are shown in \Cref{fig:molecular-descriptives}c.
 
@@
 \textbf{(b)}~Number of positive saliva samples by study condition.
-\textbf{(c)}~Number of positive saliva samples by study week and class.
+\textbf{(c)}~Possible transmission chains within classes.
+Direction of transmission as arrows.
+Number of infections per day as dot sizes.
 Viruses: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}
@@
 \end{figure}
-
-\subsection{Molecular transmission networks for adenovirus and influenza B}
-
-xxxxxx
 

### 2023-10-03 analysis/mol_analysis.Rmd
revise transmission chain plot

@@
              hline.after = NULL)
-```
-
-### Transmission network
-
-```{r}
-tnet <- df_network %>%
-  mutate(date_infection = as.POSIXct(paste0(date_infection, "00:00:00")),
-         virus = ifelse(virus == "SARS-CoV-2", "CoV", virus),
-         virus = factor(virus, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")),
-         class = paste("Class", class)) %>%
-  group_by(id, class, date_infection, virus, chain) %>%
-  summarize(n = n()) %>%
-  ungroup() %>%
-  mutate(n = factor(n, levels = c(1,2,3))) %>%
-  group_by(class, chain) %>%
-  arrange(date_infection) %>%
-  mutate(from = date_infection + hours(12),
-         to = lead(date_infection) - hours(12)) %>%
-  ungroup()
-
-coord_y_datetime <- function(xlim = NULL, ylim = NULL, expand = TRUE) {
-  if (!is.null(ylim)) {
-    ylim <- lubridate::as_datetime(ylim)
-  }
-  ggplot2::coord_cartesian(xlim = xlim, ylim = ylim, expand = expand)
-}
-
-tnet_pl <- tnet %>%
-  ggplot(aes(y = date_infection, x = virus, fill = virus, size = n)) +
-  facet_wrap(~ class, ncol = 2, scales = "free_x") +
-  geom_point(shape = 21, color = "black") +
-  geom_segment(aes(y = from, yend = to, xend = virus), size = 1,
-               arrow = arrow(length = unit(0.2, "cm")), color = "black") +
-  labs(size = "No. of infections", fill = "Virus", title = "c") +
-  coord_y_datetime(ylim = c(max(tnet$date_infection), min(tnet$date_infection))) +
-  scale_y_datetime(date_breaks = "1 week", date_labels = "%b %d") +
-  scale_size_discrete(range = c(4,8)) +
-  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
-  theme_custom() +
-  theme(legend.position = "none",
-        plot.title.position = "plot",
-        axis.title = element_blank(),
-        #axis.text.x = element_blank(),
-        #axis.ticks.x = element_blank(),
-        panel.spacing = unit(1, "cm"),
-        panel.grid.major.x = element_line(),
-        panel.grid.minor.y = element_blank(),
-        strip.text.x = element_text(size = 10, margin = margin(b = 10))) #+
-  #guides(fill = "none")
-
-#save_plot(tnet_pl, "../results/mol-data/network-plot.pdf", w = 16, h = 20)
 ```
@@
 pos_tot_pl
+```
+
+### Transmission network
+
+```{r}
+tnet <- df_network %>%
+  mutate(date_infection = as.POSIXct(paste0(date_infection, "00:00:00")),
+         virus = ifelse(virus == "SARS-CoV-2", "CoV", virus),
+         virus = factor(virus, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")),
+         class = paste("Class", class)) %>%
+  group_by(id, class, date_infection, virus, chain) %>%
+  summarize(n = n()) %>%
+  ungroup() %>%
+  mutate(n = factor(n, levels = c(1,2,3))) %>%
+  group_by(class, chain) %>%
+  arrange(date_infection) %>%
+  mutate(from = date_infection + hours(12),
+         to = lead(date_infection) - hours(12)) %>%
+  ungroup()
+
+coord_y_datetime <- function(xlim = NULL, ylim = NULL, expand = TRUE) {
+  if (!is.null(ylim)) {
+    ylim <- lubridate::as_datetime(ylim)
+  }
+  ggplot2::coord_cartesian(xlim = xlim, ylim = ylim, expand = expand)
+}
+
+tnet_pl <- tnet %>%
+  ggplot(aes(y = date_infection, x = virus, fill = virus, size = n)) +
+  facet_wrap(~ class, ncol = 2, scales = "free_x") +
+  geom_point(shape = 21, color = "black") +
+  geom_segment(aes(y = from, yend = to, xend = virus), size = 1,
+               arrow = arrow(length = unit(0.2, "cm")), color = "black") +
+  labs(size = "No. of infections", fill = "Virus", title = "c") +
+  coord_y_datetime(ylim = c(max(tnet$date_infection), min(tnet$date_infection))) +
+  scale_y_datetime(date_breaks = "1 week", date_labels = "%b %d") +
+  scale_size_discrete(range = c(4,8)) +
+  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
+  theme_custom() +
+  theme(legend.position = "none",
+        plot.title.position = "plot",
+        axis.title = element_blank(),
+        axis.text.x = element_blank(),
+        axis.ticks.x = element_blank(),
+        panel.spacing = unit(1, "cm"),
+        panel.grid.major.x = element_line(),
+        panel.grid.minor.y = element_blank(),
+        strip.text.x = element_text(size = 10, margin = margin(b = 10))) #+
+  #guides(fill = "none")
+
+#save_plot(tnet_pl, "../results/mol-data/network-plot.pdf", w = 16, h = 20)
 ```

### 2023-10-03 doc/paper/manuscript.tex
Updates from Overleaf

@@
     \caption{\textbf{Analysis of molecular data}.
-\textbf{(a)}~Number of positive saliva samples by virus.
-\textbf{(b)}~Number of positive saliva samples by study condition.
-\textbf{(c)}~Possible transmission chains within classes.
-Direction of transmission as arrows.
-Number of infections per day as dot sizes.
-Viruses: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}
+\textbf{(a,b)}~No.
+of positive saliva samples by virus and study condition.
+\textbf{(c)}~Daily no.
+of infections (1 to 3 as dot sizes) and possible transmission chains (arrows).
+Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
     \label{fig:molecular-descriptives}

### 2023-10-03 analysis/mol_analysis.Rmd
revise transmission chain plot

@@
 
+```{r}
+# combine
+descr <- pos_tot_pl + plot_spacer() + pos_cond_pl & theme(legend.position = "bottom", legend.key.height = unit(.1, "cm"),
+                                          legend.key.width = unit(.3, "cm"))
+descr <- descr + plot_layout(guides = "collect", widths = c(9.5, 1, 5.5), ncol = 2) 
+descr <- descr + pos_week_pl 
+descr <- descr + plot_layout(nrow = 2, heights = c(8, 8), design = 'ABC
+                             DDD')
+save_plot(descr, pdf_file = "../results/mol-data/descriptives.pdf", eps_file = "../results/mol-data/descriptives.eps", h = 16, w = 16)
+# saveRDS(pos_tot_pl + ggtitle(""), file = "../results/mol-data/descriptives-total.rds")
+# saveRDS(pos_cond_pl + ggtitle(""), file = "../results/mol-data/descriptives-cond.rds")
+# saveRDS(pos_week_pl + ggtitle(""), file = "../results/mol-data/descriptives-week.rds")
+```
+
 ### Transmission network
@@
   theme_custom() +
-  theme(legend.position = "none",
+  theme(legend.position = "top",
+        legend.direction = "horizontal",
+        legend.box = "vertical",
         plot.title.position = "plot",
         axis.title = element_blank(),
-        axis.text.x = element_blank(),
-        axis.ticks.x = element_blank(),
+        #axis.text.x = element_blank(),
+        #axis.ticks.x = element_blank(),
         panel.spacing = unit(1, "cm"),
@@
         panel.grid.minor.y = element_blank(),
-        strip.text.x = element_text(size = 10, margin = margin(b = 10))) #+
-  #guides(fill = "none")
+        strip.text.x = element_text(size = 10, margin = margin(b = 10))) +
+  guides(fill = "none",
+         size = guide_legend(nrow = 1, order = 1))
 
-#save_plot(tnet_pl, "../results/mol-data/network-plot.pdf", w = 16, h = 20)
-```
-
-```{r}
-# combine
-descr <- pos_tot_pl + plot_spacer() + pos_cond_pl & theme(legend.position = "bottom", legend.key.height = unit(.1, "cm"),
-                                          legend.key.width = unit(.3, "cm"))
-descr <- descr + plot_layout(guides = "collect", widths = c(9.5, 1, 5.5), ncol = 2) 
-descr <- descr + tnet_pl 
-descr <- descr + plot_layout(nrow = 2, heights = c(4, 16), design = 'ABC
-                             DDD')
-save_plot(descr, pdf_file = "../results/mol-data/descriptives.pdf", eps_file = "../results/mol-data/descriptives.eps", h = 22, w = 16)
-# saveRDS(pos_tot_pl + ggtitle(""), file = "../results/mol-data/descriptives-total.rds")
-# saveRDS(pos_cond_pl + ggtitle(""), file = "../results/mol-data/descriptives-cond.rds")
-# saveRDS(pos_week_pl + ggtitle(""), file = "../results/mol-data/descriptives-week.rds")
+save_plot(tnet_pl, pdf_file = "../results/mol-data/network-plot.pdf", eps_file = "../results/mol-data/network-plot.eps", w = 16, h = 22)
 ```

### 2023-10-03 analysis/mol_analysis.Rmd
revise transmission chain plot

@@
                arrow = arrow(length = unit(0.2, "cm")), color = "black") +
-  labs(size = "No. of infections", fill = "Virus", title = "c") +
+  labs(size = "Number of infections", fill = "Virus") +
   coord_y_datetime(ylim = c(max(tnet$date_infection), min(tnet$date_infection))) +
@@
         legend.box = "vertical",
-        plot.title.position = "plot",
+        #plot.title.position = "plot",
         axis.title = element_blank(),
@@
 
-save_plot(tnet_pl, pdf_file = "../results/mol-data/network-plot.pdf", eps_file = "../results/mol-data/network-plot.eps", w = 16, h = 22)
+save_plot(tnet_pl, pdf_file = "../results/mol-data/network-plot.pdf", eps_file = "../results/mol-data/network-plot.eps", w = 16, h = 21)
 ```

### 2023-10-03 doc/paper/manuscript.tex
revise transmission chain plot

@@
 Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (Figure~\zref{fig:mol-estimation-results-sensitivity} in \supp).
-The distribution of positive samples varied between classes.
+The distribution of positive samples varied between classes (\Cref{fig:molecular-descriptives}c).
 For example, in the first three study weeks all except for one sample were positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
 The virus-specific variations between classes are suggestive of transmission occurring within classes.
-Possible transmission chains within classes are shown in \Cref{fig:molecular-descriptives}c.
+Possible transmission chains within classes are shown in \Cref{fig:molecular-network}.
 
@@
     \caption{\textbf{Analysis of molecular data}.
-\textbf{(a,b)}~No.
-of positive saliva samples by virus and study condition.
-\textbf{(c)}~Daily no.
-of infections (1 to 3 as dot sizes) and possible transmission chains (arrows).
+Number of positive saliva samples by \textbf{(a)}~virus, \textbf{(b)}~study condition, and \textbf{(c)}~study week and class.
 Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
     \label{fig:molecular-descriptives}
+\end{figure}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{../../results/mol-data/network-plot.pdf}
+    \caption{\textbf{Transmission network}.
+Daily number of positive saliva samples (dot sizes) and possible transmission chains within classes (directed arrows).
+Positive samples are linked if they are of the same virus and less than 1~week apart.
+Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
+    \label{fig:molecular-network}
 \end{figure}

### 2023-10-03 analysis/mol_analysis.Rmd
revise transmission chain plot

@@
   theme_custom() +
-  theme(legend.position = "top",
+  theme(legend.position = "bottom",
         legend.direction = "horizontal",

### 2023-10-03 doc/paper/manuscript.tex
Updates from Overleaf

@@
 The virus-specific variations between classes are suggestive of transmission occurring within classes.
-Possible transmission chains within classes are shown in \Cref{fig:molecular-network}.
+Accordingly, we have manually labeled possible transmission chains within classes as shown in \Cref{fig:molecular-network}.
 
@@
 The shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.
-xxxxx
 
@@
 
-Molecular transmission networks based on targeted sequencing or whole genome sequencing documented transmission in the classrooms.
-xxxx (influenza b, adeno, etc.).
 The low rate of positive bioaerosol samples also means that it is unclear to what extent students airborne transmission occurred in classrooms.
@@
 However, the distribution of saliva samples between classes shows remarkable differences.
-Adenovirus was spreading entirely in class~A in the first three study weeks and only two infections with influenza~B occurred over the study.
+For example, adenovirus was spreading entirely in class~A in the first three study weeks and only two infections with influenza~B occurred over the study.
 By contrast, influenza~B was spreading throughout the study in class~B and infections with adenovirus were only detected in the last study week.

### 2023-10-03 analysis/epi_analysis.Rmd
revise prior pin plot

--- 2023-07-05 analysis/epi_analysis.Rmd
@@
   scale_fill_brewer() + 
-  facet_wrap(~ virus) +
-  labs(x = "s (delay in days)", y = expression(p[IN]*" (%)")) +
+  facet_wrap(~ virus, scales = "free_x") +
+  labs(x = "s (delay in days)", y = expression(p[IN]*" (%)"), fill = "Prior distribution (CrI)") +
   scale_x_continuous(expand = c(0,0), breaks = seq(0, 10, 2)) +
   scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
-  theme_bw2() +
-  theme(plot.title = element_text(hjust = 0), legend.position = "none")
+  theme_custom() +
+  theme(plot.title = element_text(hjust = 0), legend.position = "top",
+        panel.spacing = unit(25, units = "pt"), 
+        plot.margin = margin(r = 10))
 prior_pin_pl
 
-save_plot(prior_pin_pl, pdf_file = "../results/epi-data/incubation-periods.pdf", w = 16, h = 10)
+save_plot(prior_pin_pl, pdf_file = "../results/epi-data/incubation-periods.pdf", w = 16, h = 11)
 ```

### 2023-10-03 doc/paper/supplements.tex

@@
     \caption[Line list of molecular test results over the study period]{Line list of respiratory cases over the study period.
-Viruses: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}\label{tab:mol-data-line-list} \\
+Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}\label{tab:mol-data-line-list} \\
     \toprule
@@
     \caption[Choices of priors for the incubation periods]{Choices of priors for the distribution of the virus-specific incubation periods.
-Viruses: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}
+Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
     \label{fig:prior-pin}

### 2023-10-04 analysis/aud_analysis.Rmd
changes to plots

@@
 coughs_pl <- coughs %>%
-  mutate(week = isoweek(date)) %>%
-  filter(week != 6) %>%
-  mutate(week = ifelse(week < 6, week - min(week) + 1, week - min(week))) %>%
-  group_by(class, week) %>%
-  summarize(coughs_per_min = sum(n_coughs) / sum(TT),
-            aircleaner = first(aircleaner)) %>%
+  group_by(class, aircleaner) %>%
+  summarize(coughs_per_min = sum(n_coughs) / sum(TT)) %>%
   ungroup() %>%
-  mutate(aircleaner = factor(aircleaner, levels = c("Yes", "No")),
+  mutate(aircleaner = factor(aircleaner, levels = c("No", "Yes")),
          class = factor(class, levels = c("A", "B"))) %>%
   arrange(class, aircleaner) %>%
-  ggplot(aes(x = factor(week), y = coughs_per_min, fill = aircleaner, pattern = class)) +
-  geom_bar_pattern(stat = "identity", color = "black", pattern_fill = "white", pattern_angle = 45,
+  ggplot(aes(x = aircleaner, y = coughs_per_min, fill = aircleaner)) +
+  geom_bar_pattern(mapping = aes(pattern = class),
+                   stat = "identity", position = position_dodge(width = .66), color = "black", pattern_fill = "white", pattern_angle = 45,
                    pattern_density = 0.25, pattern_spacing = 0.05, width = .5) +
+  geom_text(data = coughs %>% group_by(aircleaner) %>% summarize(coughs_per_min = sum(n_coughs) / sum(TT)) %>% ungroup() %>%
+              mutate(y = 4.5),
+            mapping = aes(y = y, label = paste("Mean =", round(coughs_per_min, 1))),
+            size = 8 / cm(1), vjust = -1) +
   labs(x = "Study week", y = "Number of coughs per min", fill = "Air cleaner", pattern = "Class", title = "a") +
-  scale_fill_manual(values = rev(wes_palette("Royal1")[1:2])) +
-  scale_x_discrete(breaks = 1:7) +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = seq(0, 10, 2)) +
+  scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.1)), breaks = seq(0, 10, 2)) +
   scale_pattern_manual(values = c(A = "stripe", B = "none")) +
@@
         plot.title.position = "plot") +
-  guides(fill = guide_legend(nrow = 1, override.aes = list(pattern = "none", order = 2)),
+  guides(fill = "none",
          pattern = guide_legend(nrow = 1, override.aes = list(fill = "white"), order = 1))
@@
 
-save_plot(coughs_pl, pdf_file = "../results/cough-data/coughs-frequency-per-week.pdf", eps_file = "../results/cough-data/coughs-frequency-per-week.eps", w = 12, h = 8)
+save_plot(coughs_pl, 
+          pdf_file = "../results/cough-data/coughs-frequency-by-condition.pdf", 
+          eps_file = "../results/cough-data/coughs-frequency-by-condition.eps", 
+          w = 6, h = 8)
 ```
@@
 theta_pl
-save_plot(theta_pl, pdf_file = "../results/cough-data/coughs-virus-association.pdf", eps_file = "../results/cough-data/coughs-virus-association.eps", w = 12, h = 8)
+save_plot(theta_pl, 
+          pdf_file = "../results/cough-data/coughs-virus-association.pdf", 
+          eps_file = "../results/cough-data/coughs-virus-association.eps", 
+          w = 10, h = 8)
 ```

### 2023-10-04 analysis/epi_analysis.Rmd
changes to plots

@@
 cases_by_week_pl <- epi_data %>%
-  mutate(week = isoweek(date)) %>%
-  filter(week != 6) %>%
-  mutate(week = ifelse(week < 6, week - min(week) + 1, week - min(week))) %>%
-  group_by(class, week) %>%
-  summarize(cases = sum(new_cases),
-            aircleaner = first(aircleaner)) %>%
+  group_by(class, aircleaner) %>%
+  summarize(cases = sum(new_cases)) %>%
   ungroup() %>%
-  mutate(aircleaner = factor(aircleaner, levels = c("Yes", "No")),
+  mutate(aircleaner = factor(aircleaner, levels = c("No", "Yes")),
          class = factor(class, levels = c("A", "B"))) %>%
   arrange(class, aircleaner) %>%
-  ggplot(aes(x = factor(week), y = cases, fill = aircleaner, pattern = class)) +
-  geom_bar_pattern(stat = "identity", color = "black", pattern_fill = "white", pattern_angle = 45,
+  ggplot(aes(x = aircleaner, y = cases, fill = aircleaner)) +
+  geom_bar_pattern(mapping = aes(pattern = class),
+                   stat = "identity", color = "black", pattern_fill = "white", pattern_angle = 45,
                    pattern_density = 0.25, pattern_spacing = 0.05, width = .5) +
-  labs(x = "Study week", y = "Number of cases", fill = "Air cleaner", pattern = "Class", title = "a") +
-  scale_fill_manual(values = rev(wes_palette("Royal1")[1:2])) +
-  scale_x_discrete(breaks = 1:7) +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = seq(0, 10, 2)) +
+  geom_text(data = epi_data %>% group_by(aircleaner) %>% summarize(cases = sum(new_cases)) %>% ungroup(),
+            mapping = aes(label = cases),
+            size = 8 / cm(1), vjust = -1) +
+  labs(x = "Air cleaner", y = "Number of cases", fill = "Air cleaner", pattern = "Class", title = "a") +
+  scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.1)), breaks = scales::pretty_breaks(4)) +
   scale_pattern_manual(values = c(A = "stripe", B = "none")) +
   theme_custom() +
-  theme(legend.position = c(0.13, 0.8),
+  theme(legend.position = "top",
+        legend.box = "horizontal",
+        legend.direction = "horizontal",
         legend.key.width = unit(0.5, "cm"),
@@
         plot.title.position = "plot") +
-  guides(fill = guide_legend(nrow = 1, override.aes = list(pattern = "none", order = 2)),
+  guides(fill = "none",
          pattern = guide_legend(nrow = 1, override.aes = list(fill = "white"), order = 1))
 
+cases_by_week_pl
+
 save_plot(cases_by_week_pl,
-          pdf_file = "../results/epi-data/cases_by_week.pdf",
-          eps_file = "../results/epi-data/cases_by_week.eps",
-          w = 16, h = 8) 
+          pdf_file = "../results/epi-data/cases_by_condition.pdf",
+          eps_file = "../results/epi-data/cases_by_condition.eps",
+          w = 6, h = 8) 
 ```
@@
 arr <- spread_draws(latInfMod, `beta[1]`) %>%
-  rename(effect = `beta[1]`) %>%
+  rename(effect = `beta[1]`)
+
+sum(arr$effect<0) / nrow(arr) * 100
+
+arr <- arr %>%
   dplyr::select(effect) %>%
@@
           eps_file = "../results/epi-data/avoided_infections.eps",
-          w = 16, h = 8)
+          w = 10, h = 8)
 ```

### 2023-10-04 analysis/mol_analysis.Rmd
changes to plots

--- 2023-10-03 analysis/mol_analysis.Rmd
@@
 
+mirror_scales <- list(scale_x_discrete(limits = rev), scale_x_discrete())
+
 tnet_pl <- tnet %>%
@@
   facet_wrap(~ class, ncol = 2, scales = "free_x") +
+  ggh4x::facetted_pos_scales(x = mirror_scales) +
   geom_point(shape = 21, color = "black") +

### 2023-10-04 analysis/study_comaprison.Rmd
changes to plots

--- 2023-07-10 analysis/study_comaprison.Rmd
@@
   scale_pattern_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) +
-  labs(y = "Study year", x = "Frequency", title = "b") +
+  labs(y = "Study year", x = "Frequency of positive samples", title = "b") +
   theme_custom() +
@@
   scale_x_continuous(expand = expansion(mult = c(0, 0.1)), breaks = c(0, 50, 100), limits = c(0, 100)) +
-  labs(y = "Study year", x = "Estimated reduction (%)", title = "c") + 
+  labs(y = "Study year", x = "Estimated reduction in particle concentration (%)", title = "c") + 
   theme_custom() +
@@
   geom_point(mapping = aes(x = Mean), shape = 23, size = 2, fill = "white") +
-  labs(y = "Study year", x = "Adjusted risk ratio (%)", title = "d") + 
+  labs(y = "Study year", x = "ARR of infection with air cleaners (%)", title = "d") + 
   scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) +

### 2023-10-04 doc/paper/manuscript.tex

@@
 \author[2,4]{Pascal Bittel}
+\author[1]{Kathrin Zürcher}
 \author[4]{Lavinia Furrer}
-\author[4]{Alban Ramette}
-\author[4]{Ronald Dijkman}
-\author[5]{Jenna Kelly}
-\author[1]{Kathrin Zürcher}
+\author[1]{Simon Bertschinger}
+\author[2,4]{Alban Ramette}
+\author[2,4]{Ronald Dijkman}
+\author[2,5]{Jenna Kelly}
 \author[1]{Matthias Egger}
@@
 \affil[3]{Department of Infectious Diseases, Inselspital, Bern University Hospital, University of Bern, Bern, Switzerland}
-\affil[4]{Institute of Infectious Diseases, University of Bern, Bern, Switzerland}
+\affil[4]{Institute for Infectious Diseases, University of Bern, Bern, Switzerland}
 \affil[5]{Interfaculty Bioinformatics Unit, University of Bern, Bern, Switzerland}
@@
 Infection control measures such as air ventilation or filtration are supposed to reduce the risk of airborne transmission.
-The aim of this study was to evaluate the real-life effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory viral infections in schools using a multiple-measurement approach.
+The aim of this study was to evaluate the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory viral infections in schools using a multiple-measurement approach.
 \medskip
 
-\noindent\textbf{Methods and Findings:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), and molecular data (bioaerosol and saliva samples) over seven weeks from 16 January to 11 March 2023 in a secondary school (n=38 students in two classes) in Switzerland.
+\noindent\textbf{Methods:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), audio (recordings to detect coughing), and molecular data (bioaerosol and saliva samples) over seven weeks from 16 January to 11 March 2023 in a secondary school (38 students in two classes) in Switzerland.
 We performed laboratory and molecular analyses to detect a panel of 19 respiratory viruses.
 We installed air cleaners in a cross-over design and compared positive samples and particle concentrations between weeks with and without air filters.
-We modeled the number of new infections using a Bayesian negative binomial regression model, adjusting for the incubation period, the number of recent and cumulative infections, the ventilation rate, the number of students in class, and community transmission.
-\\
-Molecular analysis detected multiple viruses in saliva samples throughout the study (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus), but rarely in bioaerosol samples (2/105 positive: 1~rhinovirus and 1~adenovirus) and on the filters of air cleaners (4/160~positive: 1~influenza~B, 1~HRV, 1~adenovirus, and 1~SARS-CoV-2).
+We modeled the number of new infections using a Bayesian negative binomial regression model, adjusting for incubation periods, number of recent and cumulative infections, ventilation rate, number of students in class, and community transmission.
+\medskip
+
+\noindent\textbf{Results:} Molecular analysis detected multiple viruses in saliva samples throughout the study (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus), but rarely in bioaerosol samples (2/105 positive: 1~rhinovirus and 1~adenovirus) and on the filters of air cleaners (4/160~positive: 1~influenza~B, 1~HRV, 1~adenovirus, and 1~SARS-CoV-2).
+Nevertheless, spatiotemporal analysis of saliva samples suggested transmission within classrooms.
 The detection rate of positive saliva samples was comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61).
@@
 Daily average aerosol number concentrations were 27$\pm$34\,1/cm$^3$ with and 95$\pm$81\,1/cm$^3$ without air cleaners, corresponding to an estimated adjusted decrease of 77\% (95\%-CrI 63\% to 86\%).
-Absences related to respiratory infections were less frequent with air cleaners in place (13~absences with vs.
+Slightly less coughs  were detected with air cleaners (2.5~coughs/min with vs.
+3.1~coughs/min without air cleaners).
+Absences related to respiratory infections were less frequent with air cleaners (13~absences with vs.
 21~absences without air cleaners).
@@
 
-\noindent\textbf{Conclusions:} A wide range of non-SARS-CoV-2 respiratory viruses were detected in students, but airborne molecular detection was rare.
-SARS-CoV-2 was rarely detected in saliva and bioaerosols.
-Molecular transmission networks documented sustained transmission of non-SARS-CoV-2 seasonal viruses in school rooms despite negative airborne detection, suggesting the need for prolonged close contact.
-Air cleaners showed a small, potential benefit in reducing airborne transmission.
+\noindent\textbf{Conclusions:} A wide range of respiratory viruses were detected in students' saliva.
+Despite infrequent airborne molecular detection, transmission of respiratory viruses appeared likely in school rooms.
+A comparison with previous work indicates that respiratory viruses other than SARS-CoV-2 are less detectable in the air, but more work is needed to examine and explain heterogeneity in airborne transmission among respiratory viruses.
+Air cleaners improved air quality (lower particle concentrations), were associated with slightly lower symptom intensity (less coughing), and showed a small, potential benefit in reducing respiratory infections (fewer illness-related absences).
 
@@
 
-The transmission of respiratory infections such as SARS-CoV-2 and influenza presents significant challenges in terms of mitigation and control.
-Person-to-person transmission primarily occurs through the release of respiratory particles containing the respective viruses.
-The recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
-Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and their respiratory plumes can reach more than 2 meters\cite{Coleman2022,Wang2020,Heneghan2021}, exceeding the social distancing rules recommended by public health organizations\cite{Bazant2021PNAS,Trivedi2021PhyFluids,Lindsley2010CID}.
+The transmission of respiratory infections such as SARS-CoV-2 and influenza is difficult to mitigate and control.
+Person-to-person transmission occurs primarily through the release of respiratory particles containing the respective viruses.
+Recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
+Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and their respiratory plumes can reach more than 2m\cite{Coleman2022,Wang2020,Heneghan2021}, exceeding the social distancing rules recommended by public health organizations\cite{Bazant2021PNAS,Trivedi2021PhyFluids,Lindsley2010CID}.
 
 The generation, spread, and concentration of bioareosols can be influenced by infection control measures.
-For example, compulsory face mask wearing reduced the transmission of SARS-CoV-2 during the COVID-19 pandemic\cite{Gettings2021,Heinsohn2022,Leech2022PNAS,Mitze2020PNAS,Banholzer2021PLOS}.
-In addition, improved ventilation systems are critical for a healthy indoor environment both during pandemic and non-pandemic times\cite{Wang2021,Morawska2021}.
-Portable HEPA-air filtration devices (air cleaners ) may present another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear.
-Various studies showed that air cleaners reduce the concentration of aerosols and particle matter in indoor air\cite{Park2020Build,Buising2022InfContr,Ren2021Dent,Banholzer2023PLoSMed}.
-These findings suggest potential protection against respiratory infection, as long-term exposure to higher aerosol and particle matter mass concentrations has been associated with increased risk of respiratory infections\cite{Gordon2014Resp} and, more recently, with the incidence and mortality of COVID-19\cite{Kelly2023Atmos,Veronesi2022OccEnvMed,Travaglio2021EnvPoll,DeAngelis2021EnvRes}.
-Furthermore, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}.
+For example, mandatory face mask use reduced transmission of SARS-CoV-2 during the COVID-19 pandemic\cite{Gettings2021,Heinsohn2022,Leech2022PNAS,Mitze2020PNAS,Banholzer2021PLOS}.
+In addition, improved ventilation systems are critical for a healthy indoor environment both in pandemic and non-pandemic times\cite{Wang2021,Morawska2021}.
+Portable HEPA-air filtration devices (air cleaners) may be another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear.
+Several studies have shown that air cleaners reduce the concentration of aerosols and particle matter in indoor air\cite{Park2020Build,Buising2022InfContr,Ren2021Dent,Banholzer2023PLoSMed}.
+These findings suggest potential protection against respiratory infections, as long-term exposure to higher aerosol and particle matter mass concentrations has been associated with increased risk of respiratory infections\cite{Gordon2014Resp} and, more recently, with the incidence and mortality of COVID-19\cite{Kelly2023Atmos,Veronesi2022OccEnvMed,Travaglio2021EnvPoll,DeAngelis2021EnvRes}.
+In addition, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}.
 Simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission  of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
-We have studied the effectiveness of air cleaners in reducing the transmission of SARS-CoV-2 transmission in schools during the pandemic\cite{Banholzer2023PLoSMed}, but the analysis was limited by the study design (installation of air cleaners at the end of the Omicron wave when most students were already infected).
-In conclusion, empirical evidence linking the use of air cleaners with a reduced risk of infection is lacking.
-It remains unclear whether reducing particle concentrations and removing bioaerosols from the air will translate in reduced transmission of respiratory infections.
+We have studied the effectiveness of air cleaners in reducing transmission of SARS-CoV-2 in schools during the pandemic\cite{Banholzer2023PLoSMed}, but the analysis was limited by the study design (installation of air cleaners at the end of the omicron wave when most students were already infected).
+To date, empirical evidence linking the use of air cleaners to a reduced risk of infection is lacking.
+It remains unclear whether reducing particle concentrations and removing bioaerosols from the air will result in reduced transmission of respiratory infections.
 
-We used a multiple-measurement approach to study the transmission of viral respiratory viruses and the effect of aircleaners using a cross-over study design in school rooms under non-pandemic conditions in the winter of 2023.
-We collected epidemiological (absences probably related to respiratory infections), environmental (CO$_2$, aerosol and particle concentrations), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
+We used a multiple-measurement approach to study the transmission of viral respiratory viruses and the effect of air cleaners in school rooms.
+We employed a cross-over study design under non-pandemic conditions in the winter of 2023.
+We collected epidemiological (absences likely related to respiratory infections), environmental (CO$_2$, aerosol and particle concentrations), audio (recordings to detect coughing), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
 We determined changes in particle concentrations, respiratory virus positivity rates, and modeled the risk of infection.
@@
 
-\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the Canton of Solothurn, Switzerland, over a seven-week study period from 26~January to 11~March 2023.
-\Cref{fig:study-setup} shows the schematic study setup and Table~\zref{tab:data} in \supp \, provides an overview of the types of data that were collected in each classroom.
+\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from 26~January to 11~March 2023.
+\Cref{fig:study-setup} shows the schematic study setup and Table~\zref{tab:data} in \supp \, provides an overview of the types of data collected in each classroom.
 
@@
 Schematic study setup of the classrooms.
-One air cleaner was placed in the front and the other in the back of classrooms.
-All devices were placed at the head level of students when they were seated.
-Both classrooms were not equipped with an active HVAC (Heating, Ventilation, Air conditioning) system, but were ventilated using passive window ventilation.
+One air cleaner was placed in the front and one in the back of the classrooms.
+All devices were placed at the head level of the students when they were seated.
+Both classrooms were not equipped with an active HVAC (Heating, Ventilation, Air conditioning) system, but were ventilated naturally by opening windows.
 }
@@
 
-\noindent We used a cross-over design to study the effectiveness of air cleaners, thereby distinguishing two study conditions in each class: weeks with and weeks without air cleaners being installed in each classroom (\Cref{tab:study_design}).
+\noindent We employed a cross-over design to study the effectiveness of air cleaners, distinguishing two study conditions in each class: weeks with and weeks without air cleaners being installed in each classroom (\Cref{tab:study_design}).
 Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China; approx.
@@
 According to the manufacturer, these air cleaners run at a clean air delivery rate of 2$\,\times\,$600 m$^{3}$/h.
-Based on our own experiments conducted in the classrooms, we measured a lower clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h.
+We measured a lower clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h when testing the devices in an empty classroom.
 Window ventilation was done at the discretion of the teachers, as there were no specific recommendations from the national health authorities at that time.
@@
     \centering
-    \caption{\textbf{Study design.} Cross-over design where two portable air cleaners (AC) were installed during different time periods in classes A and B over a seven-week study period from 16~January to 11~March 2023, excluding a week of vacation between February 6 and 11.
-Swabs from the HEPA filters were taken after each intervention phase as indicated by the ``*''.}\label{tab:study_design}
+    \caption{\textbf{Study design.} Cross-over design describing when portable air cleaners (AC) were installed in the rooms of classes A and B during a seven-week study period from 16~January to 11~March 2023, excluding a week of vacation from February 6 to 11.
+A ``*'' indicates when swabs from the HEPA filters were taken, i.e.
+after each intervention phase and before the vacation.}\label{tab:study_design}
     \begin{tabular}{l l l l l l l l l}
@@
 \noindent At the beginning of the study, we collected aggregated data on age, sex and COVID-19 vaccination and recovery status in the participating classes.
-Daily we collected data on each absent student, \ie the absence and return date as well as the reason for the absence (sickness, career days,  or other).
-For sickness-related absences, we collected the symptoms and the symptom onset date.
-We also collected data on laboratory tests but none of the students reported a test result over the study.
-Based on the epidemiological data, we defined a case of respiratory infection as an absence where the student reported a sickness with at least one of the following symptoms: fever, coughing, tiredness, loss of test or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, stomach.
-Students were asked to report the first date when they experienced symptoms, \eg a student absent on Monday can also report Saturday or Sunday as the symptom onset date.
-We will denote respiratory cases by their symptom onset date, which usually corresponded to the absence date, unless the student went to school while having symptoms.
-Informed consent of one student could not be obtained and the sole absence of this student was considered as unrelated to a respiratory infection.
+On a daily basis, we collected data on each absent student, \ie the absence and return dates as well as the reason for the absence (illness, career days, or other).
+For absences due to illness, we collected symptoms and the date of symptom onset.
+We also collected data on laboratory tests but none of the students reported a test result during the study.
+Based on the epidemiological data, we defined a case of respiratory infection as an absence in which the student reported an illness with at least one of the following symptoms: fever, cough, fatigue, loss of taste or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, and stomach.
+Students were asked to report the first date they experienced symptoms, so that a student absent on Monday could report Saturday or Sunday as the day when symptoms began to be felt.
+We will refer respiratory cases by their dates of symptom onset, which usually corresponded to the absence date, unless the student attended the school while experiencing symptoms.
+Informed consent for one student could not be obtained and the single absence of this student was considered unrelated to a respiratory infection.
 Reports about absences were entered electronically into a REDcap database\cite{Harris2009,Harris2019}.
@@
 In addition, both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below).
-The saliva samples were transported to the laboratory and stored at $-$80°C until further processed\cite{Galar2021,To2019,Huber2021}.
+Saliva samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Galar2021,To2019,Huber2021}.
 The line list data is shown in Table~\zref{tab:mol-data-line-list} in \supp.
@@
 \noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute.
-This device has been used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}.
-A transient mass balance model was used to estimate the air change rate from indoor CO$_2$ levels\cite{Batterman2017IJERPH}, assuming that the classrooms were fully occupied during lessons and that room occupancy increased (decreased) linearly 10~minutes before (after) the start (end) of a lesson.
-Particle concentrations were filtered according to the times students were in the classroom, which were monitored together with the times during which windows were opened.
+This device was used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}.
+A transient mass balance model was used to estimate the air change rate from indoor CO$_2$ levels\cite{Batterman2017IJERPH}, assuming that classrooms were fully occupied during lessons and that room occupancy increased (decreased) linearly 10~minutes before (after) the start (end) of a lesson.
+Particle concentrations were filtered according to the times when students were in the classroom, which were monitored together with the times during which windows were opened.
 \medskip
@@
 
-\noindent We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France) , running at 200\,l/min and collecting into 15\,mL PBS, as previously reported in clinic settings\cite{Moore2021}.
-The Coriolis Micro Air only operated shortly before and during break times to reduce noise exposure (approximately 60\,min/day).
+\noindent We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL PBS, as previously reported in clinic settings\cite{Moore2021}.
+The Coriolis Micro Air could only operate shortly before and during break times to prevent unacceptable noise exposure (approximately 60\,min/day).
 In one class, we also used the BioSpot-VIVAS device sampler (Aerosol Devices Inc., Ft.
 Collins, CO, USA).
-This device samples airborne virus particles into a viral transport medium (VTM) using a laminar flow water-based condensation method\cite{Pan2016JAM,Lednicky2016AST}.
-BioSpot-VIVAS operated throughout lessons.
-The removable parts of both sampling devices were regularly autoclaved to avoid contamination.
+This device samples airborne virus particles into a viral transport medium (VTM) using a water-based laminar flow condensation method\cite{Pan2016JAM,Lednicky2016AST}.
+BioSpot-VIVAS could operate throughout lessons.
+Removable parts of both sampling devices were regularly autoclaved to prevent contamination.
 At the end of the day, samples were transported to the Institute of Infectious Diseases (IFIK) and stored at $-$80°C.
@@
 The human saliva samples were directly analyzed without prior filtration.
-The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19~respiratory viruses: SARS-CoV-2, Influenza A virus, Influenza B virus, Human respiratory syncytial virus A/B (RSV), Human metapneumovirus (MPV), Human adenovirus A/B/C/D/E/F (AdV), Human rhinovirus A/B/C (HRV), and Human parainfluenza virus 1/2/3/4 (PIV) in one single reaction.
+The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19~respiratory viruses: SARS-CoV-2 (CoV), Influenza A virus (Flu A), Influenza B virus (Flu B), Human respiratory syncytial virus A/B (RSV), Human metapneumovirus (MPV), Human adenovirus A/B/C/D/E/F (AdV), Human rhinovirus A/B/C (HRV), and Human parainfluenza virus 1/2/3/4 (PIV) in one single reaction.
 %/Viral genome load (VGL) of specimens was quantified (copies/L) by curve extrapolation derived from standardized ATCC quantitative genomic standard dilutions (ATCC-VR-3347D, LGC Standards, LGC Group UK).
@@
 The Limit of detection (LOD) of the assay SARS-CoV-2, IFA/IFB is 7.5x10$^0$ TCID/ml (SARS-CoV-2), 3.7x10$^{-1}$ TCID/ml (IFA), and 4.8x100 TCID/ml (100cp/ml) (IFB) respectively.
-For positive samples, we performed targeted sequencing to compare genetic relatedness between SARS-CoV-2 strains detected in the air and human samples\cite{Goncalves2021Microorganisms}, but we were unable to amplify and sequence any of the gene targets in the bioaerosol samples due to low RNA concentrations.
 
@@
 
-For adenovirus-positive samples (saliva, bioaerosol, and air filter samples), we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2020ClinicEpi}.
-The phylogenetic trees were constructed using the Neighbor-Joining method based on evolutionary distances computed using the number of differences method.
-The molecular transmission network for influenza B was determined by whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
+We performed molecular genotyping for positive samples of AdV and Flu~B.
+For AdV (saliva, bioaerosol, and air filter samples), we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2020ClinicEpi,Akello2021SciRep}.
+For Flu~B, we tried to determine the molecular transmission network using whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
 
@@
 
-We installed one microphone (RØDE NT-USB; Sydney, Australia) in each classroom to continuously record sounds.
+We installed one portable audio recorder (ZOOM H6; New York, USA) in each classroom to record sounds continuously.
 We detected the number of coughs per minute using an AI algorithm based on a convolutional neural network\cite{Bertschinger2023IEEE}, which was recently trained and validated with audio data from the waiting room of a primary care clinic\cite{Zurcher2022}.
@@
 All statistical analyses were described in the statistical analysis plan (SAP)\cite{Banholzer2023SAP}, which was published before analyses began.
-In contrast to the SAP, the number of positive bioaerosol samples and the viral load concentration of positive samples could not be analyzed because there were too few positive samples over the study.
+In contrast to the SAP, the number of positive bioaerosol samples and the viral load concentration of positive samples could not be analyzed because there were too few positive samples during the study.
 Apart from that there were no major deviations from the SAP regarding the conducted analyses.
@@
 \end{align}
-where $F_{js}$ is the number of infections in the previous week (a proxy for the number of infectious students), $N_{js}$ is the cumulative number of infections, $\beta_0$ is the infection rate without air cleaners, and $\beta_1$ is the effect of air cleaners.
-Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the community (\ie the Canton of Solothurn).
+where $F_{js}$ is the number of infections in the previous week (a proxy for the number of infectious students), $N_{js}$ is the cumulative number of infections (a proxy for the number of susceptible students), $\beta_0$ is the infection rate without air cleaners, and $\beta_1$ is the effect of air cleaners.
+Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the community (\ie the canton of Solothurn).
 
@@
 We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (Supplementary Text~\zref{sec:multinomial-model} in \supp) where the expected number of positive samples for a virus was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each virus.
-The latter was computed as the difference between the number of students per class and the cumulative number of positive samples.
-As a sensitivity analysis, we lagged the test results by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday), to consider the unknown delay when students were infected and tested positive.
+The number of susceptibles was computed as the difference between the number of students per class and the cumulative number of positive samples.
+As a sensitivity analysis, we lagged the test results once by one test date (\ie from Tuesday to last Thursday and from Thursday to last Tuesday) and once by one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday), to consider the unknown delay when students were infected and tested positive.
 \medskip
@@
 
-\noindent We computed the average concentrations per day and compared these between study conditions.
-We estimated the reduction in concentrations using Bayesian log-linear regression models (Supplementary Text~\zref{sec:env-regression-model} in \supp), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+\noindent We computed the average particle concentrations per day and compared them between study conditions.
+We estimated the reduction in particle concentrations using Bayesian log-linear regression models (Supplementary Text~\zref{sec:env-regression-model} in \supp), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
 \medskip
@@
 In addition, we used a Bayesian Negative Binomial hierarchical (random effects) regression model to estimate the association between the number of coughs and the number of positive saliva test results for respiratory viruses.
-For this analysis, we only used the days were saliva testing was performed and coughs could be detected because people were in the classroom (27~days in total).
+For this analysis, we only used the days were saliva testing was performed (i.e.
+every Tuesday and Thursday, 27~days in total).
 \medskip
@@
 Each model is estimated with 4 Markov chains and 2,000 iterations of which the first 1,000 iterations are discarded as part of the warm-up.
-Estimation power is evaluated via the effective sample size ESS and convergence of the Markov chains is evaluated with the Gelman-Rubin convergence diagnostic ($\hat{R}$).
+Estimation power is evaluated via the effective sample size (ESS) and convergence of the Markov chains is evaluated with the Gelman-Rubin convergence diagnostic ($\hat{R}$).
 If not stated otherwise, we report posterior means and credible intervals (CrIs) based on the 50\%, 80\%, and 95\% quantiles of the posterior samples, respectively.
@@
 
-\noindent The Ethics Committee of the Canton of Bern, Switzerland, approved the study (reference no.
+\noindent The Ethics Committee of the canton of Bern, Switzerland, approved the study (reference no.
 2021-02377).
@@
 Of these, 7~students were vaccinated or recovered from an infection within SARS-CoV-2 in the last four months.
-Over the seven-week study period (1,330 student-days in total), students were absent from school for 220~days (18\% of total) of which 129~days (59\% of absences) were due to sickness.
+During the seven-week study period (total of 1,330 student-days), students were absent from school for 220~days (18\% of the total) of which 129~days (59\% of absences) were due to illness.
 
@@
 
-\subsection{Molecular analyses}
+\subsection{Molecular analyses of saliva and air samples}
 
@@
 The distribution of positive samples varied between classes (\Cref{fig:molecular-descriptives}c).
-For example, in the first three study weeks all except for one sample were positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
-The virus-specific variations between classes are suggestive of transmission occurring within classes.
-Accordingly, we have manually labeled possible transmission chains within classes as shown in \Cref{fig:molecular-network}.
-
+For example, during the first three study weeks, all but one sample was positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
 We also analyzed 105~bioaerosol samples and detected  two respiratory viruses in the air (1~HRV in class~A and 1~AdV in class~B).
-Furthermore, we detected 1~Flu~B, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
+Finally, we detected 1~Flu~B, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
 
@@
 
+\subsection{Analyses of transmission networks and molecular genotyping}
+
+The virus-specific variation between classes and the temporal distribution of positive samples suggest that transmission occurred within classes.
+To illustrate possible transmission chains, we linked positive saliva samples within classes that belong to the same virus and are less than 1~week apart.
+Based on this spatiotemporal analysis of students' saliva samples, we ascertained 10~transmission chains, which are depicted in \Cref{fig:molecular-network}.
+The longest transmission chains possibly occurred in January and refer to the cluster of AdV infections in class~A and Flu~B infections in class~B.
+We performed molecular genotyping of saliva, bioaerosol, and air filter samples to verify our suggested transmission networks.
+However, molecular genotyping was not successful because we were unable to amplify and sequence any of the gene targets.
+
 \begin{figure}[!htpb]
@@
     \includegraphics{../../results/mol-data/network-plot.pdf}
-    \caption{\textbf{Transmission network}.
+    \caption{\textbf{Transmission network of respiratory viruses: spatiotemporal analysis of students' saliva samples}.
 Daily number of positive saliva samples (dot sizes) and possible transmission chains within classes (directed arrows).
-Positive samples are linked if they are of the same virus and less than 1~week apart.
+Positive samples are linked if they belong to the same virus and are less than 1~week apart.
 Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
@@
 When adjusting for air change rates and multiple other factors, the aerosol number concentration decreased by 76\% (95\%-CrI 63\% to 86\%) with air cleaners (\Cref{fig:palas-results}b and Table~\zref{tab:palas-est-results} in \supp).
-The decrease in the concentration of larger particles (PM$_{10}$) was larger than the decrease in the concentration of smaller particles (PM$_1$ to PM$_{4}$).
+The decrease in the concentration of larger particles (PM$_{10}$) was greater than the decrease in the concentration of smaller particles (PM$_1$ to PM$_{4}$).
 
@@
 Overall, there were 35~respiratory cases (absences related to respiratory infections) across classes.
-Incidentally, there were more cases in class~B (21) than in class~A (14), which is the class were more students recently recovered from a SARS-CoV-2 infection or were vaccinated.
+Incidentally, there were more cases in class~B (21) than in class~A (14), which is incidentally also the class where more students recently recovered from a SARS-CoV-2 infection or were vaccinated.
 Except for one study week, there was an equal or higher number of cases without air cleaners (\Cref{fig:redcap-results}a).
 Based on our Bayesian latent hierarchical regression model, we estimated a trend towards a lower risk of infection with air cleaners (adjusted relative risk 0.72, 95\%-Credible Interval 0.44 to 1.15).
-As a result, the estimated number of infections if air cleaners had been installed throughout the study is expected to be 36 (95\%-CrI 21 to 64) compared to 62 (95\%-CrI 29 to 141) infections if they would not have been installed (\Cref{fig:redcap-results}b).
-Detailed estimation results are provided in Supplementary Text~\zref{sec:detailed-redcap} in \supp.
+As a result, the estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\Cref{fig:redcap-results}b).
+The posterior probability that air cleaners reduced the risk of infection was 90.8\%.
+Detailed estimations are provided in the Supplementary Text~\zref{sec:detailed-redcap} in \supp.
 
 \begin{figure}[!htpb]
-    \includegraphics[width=\linewidth]{../../results/epi-data/cases_by_week.pdf} 
-    \includegraphics[width=\linewidth]{../../results/epi-data/avoided-infections.pdf}
+    \includegraphics{../../results/epi-data/cases_by_condition.pdf} 
+    \includegraphics{../../results/epi-data/avoided-infections.pdf}
     \caption{\textbf{Analysis of epidemiological data and estimated number of infections with and without air cleaners.}.
@@
 
-Slightly fewer coughs were detected during study weeks where air cleaners were installed (\Cref{fig:coughing}a).
+Slightly fewer coughs were detected in audio recordings during study weeks where air cleaners were installed (\Cref{fig:coughing}a).
 On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners (adjusted relative risk 0.93, 95\%-CrI 0.85 to 1.02).
 Significantly more coughs were detected in class~B (adjusted relative risk 2.16, 95\%-CrI 1.90 to 2.46), suggesting a link with the observed variation in virus-specific transmission between classes (\Cref{fig:molecular-descriptives}c).
-This link can also be seen in the association between the frequency of coughing and the number of positive saliva tests for respiratory viruses (\Cref{fig:coughing}b).
-Coughing was more frequent for Flu~B (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent with AdV (0.87, 95\%-CrI 0.49 to 1.12).
+As suspected, the frequency of coughing was associated with the number of positive saliva samples by respiratory virus (\Cref{fig:coughing}b).
+Coughing was more frequent during Flu~B positivity (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent during AdV positivity (0.87, 95\%-CrI 0.49 to 1.12).
 
@@
     \includegraphics[width=\linewidth]{../../results/comparison.pdf} 
-    \caption{\textbf{Comparison of the results between our previous study in 2022 during the COVID-19 pandemic (blue) and this study in 2023 during non-pandemic times (black).}.
-\textbf{(a)}~Proportion of positive saliva samples and virus.
-Viruses: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).
-\textbf{(b)}~Number of positive saliva, bioaerosol, and HEPA-filter samples.
+    \caption{\textbf{Comparison of the results between our previous study in 2022 during the COVID-19 pandemic (blue) and this study in 2023 during non-pandemic times (black).} \textbf{(a)}~Proportion of positive saliva samples by virus.
+Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.
 \textbf{(c)}~Estimated reduction in aerosol number and particle mass concentrations (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$) with air cleaners (posterior mean as squares and 95\%-CrI as lines).
-\textbf{(d)}~Adjusted relative risk of infection (posterior mean as squares and 95\%-CrI as lines).}
+\textbf{(d)}~Adjusted relative risk (ARR) of infection with air cleaners (posterior mean as squares and 95\%-CrI as lines).}
     \label{fig:comparison}
@@
 
-We collected epidemiological, environmental, and molecular data to estimate the risk of infection with respiratory viruses in a Swiss school and assessed the effectiveness of air cleaners as an infection control measure.
-We found a wide range of respiratory viruses in human saliva samples, primarily adenovirus, influenza B, and rhinovirus, but only few non-SARS-CoV-2 respiratory viruses were detected in bioaerosol samples and on the filters of air cleaners.
-Aerosol number and particle mass concentrations decreased significantly with air cleaners and Bayesian modeling based on the absences of students showed a small reduction in the relative risk of infection with air cleaners.
-Reduced intensity of symptoms (coughing) in school, as determined by acoustic recording, indicates small prevention of symptomatic infections by air cleaners.
+We collected epidemiological, environmental, and molecular data to estimate the risk of respiratory virus infection in a Swiss school and assessed the effectiveness of air cleaners as an infection control measure.
+We found a wide range of respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus, but very few respiratory viruses were detected in bioaerosol samples and on the filters of the air cleaners.
+Aerosol number and particle mass concentrations decreased significantly with air cleaners and Bayesian modeling based on the absences of students due to illness showed a small, potential reduction in the relative risk of infection with air cleaners.
+Reduced coughing could indicate that air cleaners prevent some symptomatic infections.
 
@@
 In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners in reducing the transmission of SARS-CoV-2 in two Swiss schools.
-We only found a detectable effect on the risk of infection for mask wearing, but not air cleaners.
-We argued that air cleaners may have had less potential to reduce infections because they were implemented at the end of the study when most students had already been infected with SARS-CoV-2.
+We found a detectable effect on the risk of infection only for mask wearing, but not air cleaners.
+We argued that air cleaners may have had less potential to reduce infections because they were introduced at the end of the study, when most students were already infected with SARS-CoV-2.
 In this study, we used a cross-over design to estimate the effectiveness of air cleaners again, but this time under non-pandemic conditions.
-We found similar reductions in particle concentrations and a tendency that the relative risk of infection was lower with air cleaners.
+We found similar reductions in particle concentrations and a tendency for the relative risk of infection to be lower with air cleaners.
 To some extent, the differences between the two studies may be explained by different adjustments.
-While we could not adjust for ventilation in our previous study, the estimates of this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see Supplementary Text~\zref{sec:detailed-molecular} in \supp).
-Furthermore, the distribution of respiratory viruses in this study was very different from the distribution in our previous study.
+While we could not adjust for ventilation in our previous study, the estimates in this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see Supplementary Text~\zref{sec:detailed-molecular} in \supp).
+Furthermore, the distribution of respiratory viruses in this study was very different from that in our previous study.
 
@@
 Similarly, the number of viruses detected on the filters of the air cleaners was also lower in this study (4~positive samples in 160~swabs vs 8~positive samples in 80~swabs in the previous study).
-Importantly, there was a clear shift in the distribution of respiratory viruses in the current study compared to the previous study in 2022 during the SARS-CoV-2 Omicron variant.
+Importantly, there was a clear shift in the distribution of respiratory viruses in the current study compared to the previous study in 2022 during the SARS-CoV-2 omicron variant.
 Whereas in 2022 we detected almost exclusively SARS-CoV-2 in human saliva, the most common respiratory viruses in 2023 were adenovirus, influenza~B, and rhinoviruses (only three SARS-CoV-2-positive samples).
@@
 Technical reasons for the differences in airborne detection are unlikely.
-The bioaerosol sampling devices and laboratory procedures were the same in both studies and no technical issues were observed.
+The bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
 Instead, airborne detection could have been influenced by environmental factors.
-Although temperature (overall between 19 and 25$^{\circ}$C) and relative humidity (overall between 25 and 50\%) were similar in both studies, the same levels can have a different effect on airborne survival of respiratory viruses.
-For example, for rhinovirus and adenovirus, survival increases with relative humidity, while the reverse relationship has been documented for influenza and SARS-CoV-2\cite{Tellier2009JTRSI,Ahlawat2020AAQR,Biryukov2020mS,Karim1985CJM,Davis1971AM}.
-A change in ventilation conditions could have had a profound effect, but CO$_2$ levels were even higher in this study (1,702$\pm$370\,ppm vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
-Although lower ventilation was partially offset by a greater reduction in aerosol and particle concentrations from air cleaners, the removal of particles through air cleaners may be slower than through ventilation.
+Although temperature (overall between 19 and 25$^{\circ}$C) and relative humidity (overall between 25 and 50\%) were similar in both studies, the same levels may have different effects on the airborne survival of respiratory viruses.
+For example, airborne survival of rhinovirus and adenovirus increases with relative humidity, whereas the opposite relationship has been documented for influenza and SARS-CoV-2\cite{Tellier2009JTRSI,Ahlawat2020AAQR,Biryukov2020mS,Karim1985CJM,Davis1971AM}.
+A change in ventilation conditions could also have had a profound effect, but CO$_2$ levels were higher compared to the previous study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
+Although lower ventilation was partially offset by a greater reduction in aerosol and particle concentrations from air cleaners, the removal of particles by air cleaners may be slower than by natural ventilation.
 
@@
 
-As a consequence, we speculate that the difference in airborne molecular detection of respiratory viruses is due to the characteristics of the respiratory viruses.
-Differences in physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity), which influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}, could make it easier to detect SARS-CoV-2 in the air.
+We speculate that the difference in airborne molecular detection between 2022 (pandemic) and 2023 (non-pandemic) of respiratory viruses is due to the different characteristics of the respiratory viruses.
+Differences in the physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity), which influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}, may facilitate the detection of SARS-CoV-2 in the air.
 The reason is that viral loads do often not exceed the detection limits of existing devices for bioaerosol sampling, which remains a challenge\cite{Wang2021,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
-For example, one study was unable to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.
-Nevertheless, previous studies show that also influenza\cite{Bischoff2013JID,Pan2017mSphere}, rhinovirus\cite{Myatt2004AJRCCM}, and adenovirus\cite{Nguyen2017OFID,Pan2017mSphere} can be detected in airborne samples.
-The ability to detect pathogens in bioaerosols could also be related to patient characteristics of the study participants.
-Multiple studies suggest an association between the generation of pathogenic aerosols and infectiousness\cite{Leung2020NatMed,Bischoff2013JID,Escombe2008PLoSMed}.
-Therefore, studies including more infectious patients (intentionally or unintentionally) will probably have a higher chance of detecting respiratory viruses in the air.
+For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.
+However, several studies showed that influenza\cite{Bischoff2013JID,Pan2017mSphere}, rhinovirus\cite{Myatt2004AJRCCM}, and adenovirus\cite{Nguyen2017OFID,Pan2017mSphere} can also be detected in airborne samples.
+Finally, the ability to detect pathogens in bioaerosols may be related to patient characteristics of the study participants.
+Several studies suggested an association between the generation of pathogenic aerosols and infectiousness\cite{Leung2020NatMed,Bischoff2013JID,Escombe2008PLoSMed}.
+Therefore, studies that include more infectious patients (intentionally or unintentionally) are likely to have a higher chance of detecting respiratory viruses in the air.
 
@@
 
-The low rate of positive bioaerosol samples also means that it is unclear to what extent students airborne transmission occurred in classrooms.
-It may be possible that students had rather little exposure to respiratory viruses in schools and acquired their infections elsewhere.
+The low rate of positive bioaerosol samples also means that it is unclear to what extent airborne transmission occurred in classrooms.
+It may be possible that students had relatively little exposure to respiratory viruses in schools and acquired their infections elsewhere.
 However, the distribution of saliva samples between classes shows remarkable differences.
-For example, adenovirus was spreading entirely in class~A in the first three study weeks and only two infections with influenza~B occurred over the study.
-By contrast, influenza~B was spreading throughout the study in class~B and infections with adenovirus were only detected in the last study week.
-Adenovirus infections often tend to be mild\cite{Kunz2010CIDR} and less associated with coughing than influenza\cite{Ma2018RMV}, which is in line with the comparably lower frequency of detected coughs in class~B.
-Altogether, the class-specific patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.
+For example, adenovirus spread entirely in class~A during the first three study weeks and only two infections with influenza~B occurred over the study.
+In contrast, influenza~B spread throughout the study in class~B, and adenovirus infections were  detected only in the last week of the study.
+Adenovirus infections tend to be mild\cite{Kunz2010CIDR} and less associated with cough than influenza\cite{Ma2018RMV}, which is consistent with the comparatively lower frequency of detected coughs in class~B.
+Taken together, the class-specific, spatiotemporal patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.
 
@@
 
-Our study indicated a small, potential benefit of air cleaners, but other infection control measures such as universal mask wearing appear to be more effective\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Dharmadhikari2012AJRCCM,Leung2020NatMed,Milton2013PLoSPathogens}.
-There may be several reasons for this.
-Compared to \eg mask wearing, air cleaners cannot influence transmission outside the classrooms.
-Furthermore, previous work suggests that prolonged close contact may be needed for transmission\cite{Leung2020NatMed,Brankston2007LancetID}, or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
-However, compared to face masks, air cleaners are unable to prevent such close range high particle density aerosol transmission.
-Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective in the removal of larger particles ($>5\mu$m).
-Yet, a larger proportion of respiratory viruses are carried in the smaller particles ($\leq5\mu$m)\cite{Fennelly2020}.
-Finally, classroom activity, airflow and other unobserved, confounding factors make it difficult to assess the effects of air cleaners.
-Nevertheless, it is important to assess their effectiveness in a real-life setting, and compare with the hypothetical impact suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
+Our study indicated a small, potential benefit of air cleaners.
+There may be several reasons for this relatively small effect.
+Compared to \eg universal mask wearing, which has been shown to be effective\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Dharmadhikari2012AJRCCM,Leung2020NatMed,Milton2013PLoSPathogens}, air cleaners cannot affect transmission outside the classroom.
+Furthermore, previous work suggests that prolonged close contact may be necessary for transmission\cite{Leung2020NatMed,Brankston2007LancetID}, or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
+However, compared to face masks, air cleaners are not able to prevent transmission by close range, high particle density.
+Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective at removing larger particles ($>5\mu$m).
+However, a greater proportion of respiratory viruses are carried in the smaller particles, which are also more relevant for transmission ($\leq5\mu$m)\cite{Fennelly2020}.
+Finally, classroom activity, airflow and other unobserved, confounding factors make it difficult to evaluate the effects of air cleaners.
+Nevertheless, it is important to assess their effectiveness in a real-world setting and to compare them with the hypothetical impact suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
 
@@
 Our study has several limitations.
-First, aerosol measurements and molecular detection of viruses in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (human to air, air to human) cannot be determined.
-Nevertheless, the virus-specific distribution of positive saliva samples between classes indicate potential transmission within classrooms.
-Second, it was not possible to ascertain the specific causes for the absences based on the epidemiological data.
+First, aerosol measurements and molecular detection of viruses in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (person to air, air to person) cannot be determined.
+However, the virus-specific distribution of positive saliva samples between classes (spatiotemporal pattern) may indicate transmission within classrooms.
+Second, the specific causes for the absences could not be determined from the epidemiological data.
 Therefore, some absences may have falsely been attributed to respiratory infections and we could only approximate the incubation period for each epidemiological case.
 Third, the effectiveness of air cleaners was somewhat limited because students occasionally had lessons outside the classroom.
-This is in contrast to other infection control measures such as face masks that had to be used in every indoor space during pandemic times.
+This is in contrast to other infection control measures such as face masks, which had to be used in every indoor space during the pandemic.
 
@@
 
-In conclusion, using epidemiological, environmental, and molecular data, our study shows that considerable transmission of respiratory viruses other than SARS-CoV-2 occurred in school rooms under non-pandemic conditions.
-Air cleaners significantly reduced aerosol concentrations, but airborne respiratory viruses were rarely detected, suggesting that prolonged close contact may be required for transmission.
-The epidemiologic transmission model estimated a tendency towards lower transmission with air filters.
-Future studies may use our multiple-measurement approach to assess whether professional building ventilation, which should be preferred over air cleaners in the long term, and other infection control measures reduce the risk of airborne transmission.
-Whether their effectiveness varies with the seasonal distribution of respiratory viruses under non-pandemic conditions also remains to be investigated.
+In conclusion, using epidemiological, environmental, and molecular data, our study shows that transmission of respiratory viruses likely occurred in school rooms under non-pandemic conditions when public health measures were lifted.
+Air cleaners improved air quality by significantly reducing aerosol and particle concentrations.
+Based on our transmission model using epidemiological data, the risk of infection tends to be lower with air filters.
+Airborne respiratory viruses were rarely detected, suggesting that prolonged close contact may be required for transmission.
+Future studies could investigate the relative importance of prolonged close contact and long range transmission.
 
@@
 We would like to thank the schools, teachers, and students participating in the study.
-We are grateful to the Educational Department of the Canton of Solothurn for their support throughout the study.
-We also would like to thank the biosecurity team at the Institute for Infectious Diseases of the University of Bern for their assistance with the bioaerosol devices (Kathrina Summermatter, Julia Feldmann, and Monika Gsell).
-Finally, we are indebted to the student assistants who helped with the data collection at the schools.
+We are grateful to the Educational Department of the canton of Solothurn for their support throughout the study.
+We also would like to thank the biosecurity team at the Institute for Infectious Diseases of the University of Bern for their assistance with the bioaerosol devices (Julia Feldmann, Monika Gsell, Kathrina Summermatter).
+Finally, we are indebted to the student assistants (Khadija Ahmed, Marie-Joséphine Brancato, Michelle Bürki, Santiago Martinez, Moric Toszeghi, Sylvain Wasmer) who helped with the data collection in the schools.
 

### 2023-10-06 analysis/mol_analysis.Rmd
update network plot with air and filter samples

@@
 df_network <- rbind(readxl::read_excel("../data-clean/transmission-network.xlsx", sheet = "A"),
-                    readxl::read_excel("../data-clean/transmission-network.xlsx", sheet = "B"))  
+                    readxl::read_excel("../data-clean/transmission-network.xlsx", sheet = "B")) %>%
+  mutate(date_infection = gsub("2022", "2023", as.character(date_infection)))
+
+# manually from file because it is only a few samples
+air_samples <- matrix(
+  c("B", "25-01-2023", "HRV", "Air",
+    "A", "20-02-2023", "AdV", "Air",
+    "B", "01-02-2023", "Flu B", "Filter", # we use the middle of the week(s) when the swabs were taken
+    "A", "05-03-2023", "CoV", "Filter",
+    "A", "05-03-2023", "AdV", "Filter",
+    "A", "05-03-2023", "HRV", "Filter"),
+  ncol = 4, byrow = T
+) %>%
+  data.frame() %>%
+  set_names(c("class", "date", "virus", "type")) %>%
+  mutate(date = as.Date(date, format = "%d-%m-%Y"))
 ```
@@
 tnet_pl <- tnet %>%
-  ggplot(aes(y = date_infection, x = virus, fill = virus, size = n)) +
+  mutate(type = factor("Saliva", levels = c("Saliva", "Air", "Filter"))) %>%
+  ggplot(aes(y = date_infection, x = virus, size = n, shape = type)) +
   facet_wrap(~ class, ncol = 2, scales = "free_x") +
   ggh4x::facetted_pos_scales(x = mirror_scales) +
-  geom_point(shape = 21, color = "black") +
+  geom_point(data = air_samples %>% 
+               mutate(date_infection = as.POSIXct(paste0(date, "00:00:00", format = "%d-%m-%Y %H:%M:%S")),
+                      n = factor(1, levels = c(1,2,3)),
+                      class = factor(paste("Class", class), levels = c("Class A", "Class B")),
+                      type = factor(type, levels = c("Saliva", "Air", "Filter")),
+                      virus = factor(virus, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV"))),
+             mapping = aes(color = virus), fill = "white",
+             position = position_dodge(width = -.6, preserve = "single")) +
+  geom_point(mapping = aes(fill = virus), color = "black") +
   geom_segment(aes(y = from, yend = to, xend = virus), size = 1,
               arrow = arrow(length = unit(0.2, "cm")), color = "black") +
-  labs(size = "Number of infections", fill = "Virus") +
+  labs(size = "Number of infections", fill = "Virus", shape = "") +
   coord_y_datetime(ylim = c(max(tnet$date_infection), min(tnet$date_infection))) +
@@
   scale_fill_manual(values = wes_palette("IsleofDogs1")) +
+  scale_color_manual(values = wes_palette("IsleofDogs1")) +
+  scale_shape_manual(values = c("Saliva" = 21, "Air" = 22, "Filter" = 23),
+                     breaks = c("Saliva", "Air", "Filter")) +
   theme_custom() +
@@
         legend.direction = "horizontal",
-        legend.box = "vertical",
+        legend.box = "horizontal",
         #plot.title.position = "plot",
@@
   guides(fill = "none",
-         size = guide_legend(nrow = 1, order = 1))
+         color = "none",
+         shape = guide_legend(order = 1),
+         size = guide_legend(nrow = 1, order = 2))
 

### 2023-10-06 doc/paper/manuscript.tex
Update final draft of manuscript based on feedback from Lukas

@@
 \author[2,4]{Ronald Dijkman}
-\author[2,5]{Jenna Kelly}
 \author[1]{Matthias Egger}
-\author[2,6]{Tina Hascher}
+\author[2,5]{Tina Hascher}
 \author[1*,2]{Lukas Fenner}
@@
 \affil[4]{Institute for Infectious Diseases, University of Bern, Bern, Switzerland}
-\affil[5]{Interfaculty Bioinformatics Unit, University of Bern, Bern, Switzerland}
-\affil[6]{Institute of Educational Science, University of Bern, Bern, Switzerland}
+\affil[5]{Institute of Educational Science, University of Bern, Bern, Switzerland}
 
@@
 
-%TC:newcounter abst Words in abstract
-%TC:envir abstract [] abst
+%TC:ignore
 \begin{abstract}\normalfont
@@
 
-\noindent\textbf{Conclusions:} A wide range of respiratory viruses were detected in students' saliva.
-Despite infrequent airborne molecular detection, transmission of respiratory viruses appeared likely in school rooms.
-A comparison with previous work indicates that respiratory viruses other than SARS-CoV-2 are less detectable in the air, but more work is needed to examine and explain heterogeneity in airborne transmission among respiratory viruses.
-Air cleaners improved air quality (lower particle concentrations), were associated with slightly lower symptom intensity (less coughing), and showed a small, potential benefit in reducing respiratory infections (fewer illness-related absences).
+\noindent\textbf{Conclusions:} During this non-pandemic period, a wide range of non-SARS-CoV-2 respiratory viruses were detected in students' saliva, but airborne molecular detection was rare.
+However, transmission of these respiratory viruses seemed likely in schools, suggesting that prolonged close contact is required for transmission.SARS-CoV-2 was very rarely detected in saliva and air.
+Air cleaners improved air quality, were associated with lower symptom intensity, and showed a small, potential benefit in reducing respiratory infections.
 
@@
 \end{abstract}
-
-%TC:ignore
 
@@
 \vspace{2em}
-
-%\noindent\textbf{Word count:} \abstractcount{manuscript}words (max. 500)
 
@@
 
-%TC:break main
 \section{Introduction}
@@
 Simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission  of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
-We have studied the effectiveness of air cleaners in reducing transmission of SARS-CoV-2 in schools during the pandemic\cite{Banholzer2023PLoSMed}, but the analysis was limited by the study design (installation of air cleaners at the end of the omicron wave when most students were already infected).
+We have studied the effectiveness of air cleaners in reducing transmission of SARS-CoV-2 in schools during the pandemic\cite{Banholzer2023PLoSMed}, but the analysis was limited by the study design (installation of air cleaners at the end of the Omicron wave when most students were already infected).
 To date, empirical evidence linking the use of air cleaners to a reduced risk of infection is lacking.
@@
 According to the manufacturer, these air cleaners run at a clean air delivery rate of 2$\,\times\,$600 m$^{3}$/h.
-We measured a lower clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h when testing the devices in an empty classroom.
+% We measured a lower clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h when testing the devices in an empty classroom.
 Window ventilation was done at the discretion of the teachers, as there were no specific recommendations from the national health authorities at that time.
@@
 We performed molecular genotyping for positive samples of AdV and Flu~B.
-For AdV (saliva, bioaerosol, and air filter samples), we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2020ClinicEpi,Akello2021SciRep}.
+For AdV (saliva, bioaerosol, and air filter samples), we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}.
 For Flu~B, we tried to determine the molecular transmission network using whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
@@
 There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in Supplementary Text~\zref{sec:transmission-model} to Supplementary Text~\zref{sec:env-regression-model} in \supp, which provide detailed descriptions of the models used in each analysis.
+\medskip
 
@@
 The code is available from \url{https://osf.io/38j9g}.
-
-% \subsection{Additional data collection}
-
-% \noindent We also collected data about the students' subjective experiences of well-being and emotions using standardized daily online questionnaires and a pre-post questionnaire\cite{Brandenberger2018}.
-These data are not part of the current analysis.
 
@@
 
-The virus-specific variation between classes and the temporal distribution of positive samples suggest that transmission occurred within classes.
 To illustrate possible transmission chains, we linked positive saliva samples within classes that belong to the same virus and are less than 1~week apart.
-Based on this spatiotemporal analysis of students' saliva samples, we ascertained 10~transmission chains, which are depicted in \Cref{fig:molecular-network}.
+Based on this spatiotemporal analysis of student saliva samples, we identified 10~possible transmission chains, which are shown in \Cref{fig:molecular-network}.
 The longest transmission chains possibly occurred in January and refer to the cluster of AdV infections in class~A and Flu~B infections in class~B.
-We performed molecular genotyping of saliva, bioaerosol, and air filter samples to verify our suggested transmission networks.
-However, molecular genotyping was not successful because we were unable to amplify and sequence any of the gene targets.
+We performed molecular genotyping of saliva, bioaerosol, and air filter samples to verify our proposed transmission networks.
+However, molecular genotyping was not successful because we were unable to amplify and sequence any of the gene targets.The virus-specific variation between classes and the temporal distribution of positive samples suggest that transmission occurred within classes.
 
@@
     \caption{\textbf{Transmission network of respiratory viruses: spatiotemporal analysis of students' saliva samples}.
-Daily number of positive saliva samples (dot sizes) and possible transmission chains within classes (directed arrows).
+Daily number of positive saliva samples (colored circles) and possible transmission chains within classes (directed arrows).
 Positive samples are linked if they belong to the same virus and are less than 1~week apart.
@@
 As a result, the estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\Cref{fig:redcap-results}b).
-The posterior probability that air cleaners reduced the risk of infection was 90.8\%.
+The posterior probability that air cleaners reduced the risk of infection was 90.8\%, also suggesting a beneficial effect of air cleaners.
 Detailed estimations are provided in the Supplementary Text~\zref{sec:detailed-redcap} in \supp.
@@
     \centering
-    \includegraphics{results/cough-data/coughs-frequency-per-week.pdf}
+    \includegraphics{results/cough-data/coughs-frequency-by-condition.pdf}
     \includegraphics{results/cough-data/coughs-virus-association.pdf}
@@
 Similarly, the number of viruses detected on the filters of the air cleaners was also lower in this study (4~positive samples in 160~swabs vs 8~positive samples in 80~swabs in the previous study).
-Importantly, there was a clear shift in the distribution of respiratory viruses in the current study compared to the previous study in 2022 during the SARS-CoV-2 omicron variant.
+Importantly, there was a clear shift in the distribution of respiratory viruses in the current study compared to the previous study in 2022 during the SARS-CoV-2 Omicron variant.
 Whereas in 2022 we detected almost exclusively SARS-CoV-2 in human saliva, the most common respiratory viruses in 2023 were adenovirus, influenza~B, and rhinoviruses (only three SARS-CoV-2-positive samples).
@@
 
-Technical reasons for the differences in airborne detection are unlikely.
+% Technical reasons for the differences in airborne detection are unlikely.
 The bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
@@
 
-We speculate that the difference in airborne molecular detection between 2022 (pandemic) and 2023 (non-pandemic) of respiratory viruses is due to the different characteristics of the respiratory viruses.
+We speculate that the difference in airborne molecular detection of respiratory viruses between 2022 (pandemic) and 2023 (non-pandemic) is due to the different characteristics of the respiratory viruses.
 Differences in the physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity), which influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}, may facilitate the detection of SARS-CoV-2 in the air.
@@
 
-In conclusion, using epidemiological, environmental, and molecular data, our study shows that transmission of respiratory viruses likely occurred in school rooms under non-pandemic conditions when public health measures were lifted.
+In conclusion, using epidemiological, environmental, and molecular data, our study showed that a wide range of non-SARS-CoV-2, but rarely SARS-CoV-2, was detected in students under non-pandemic conditions when public health measures were lifted, and that transmission of these respiratory viruses likely occurred in schools.
+Airborne detection of non-SARS-CoV-2 respiratory viruses was rare, suggesting that prolonged close contact may be required for transmission.
+Comparison with previous work during the pandemic, which showed frequent airborne SARS-CoV-2, suggests that respiratory viruses other than SARS-CoV-2 may be less detectable in the air.
 Air cleaners improved air quality by significantly reducing aerosol and particle concentrations.
 Based on our transmission model using epidemiological data, the risk of infection tends to be lower with air filters.
-Airborne respiratory viruses were rarely detected, suggesting that prolonged close contact may be required for transmission.
-Future studies could investigate the relative importance of prolonged close contact and long range transmission.
+Future studies should investigate the relative importance of prolonged close contact and long range transmission of respiratory viruses in schools and other congregate settings.
 

### 2023-10-06 doc/paper/manuscript.tex
Finalized draft of long manuscript

@@
 
-\noindent\textbf{Conclusions:} During this non-pandemic period, a wide range of non-SARS-CoV-2 respiratory viruses were detected in students' saliva, but airborne molecular detection was rare.
-However, transmission of these respiratory viruses seemed likely in schools, suggesting that prolonged close contact is required for transmission.SARS-CoV-2 was very rarely detected in saliva and air.
+\noindent\textbf{Conclusions:} During a non-pandemic period of winter 2023, a wide range of non-SARS-CoV-2 respiratory viruses were detected in the saliva of students in a Swiss school.
+Although airborne molecular detection was rare, transmission of these respiratory viruses seemed likely in schools, suggesting that prolonged close contact is required for transmission.
 Air cleaners improved air quality, were associated with lower symptom intensity, and showed a small, potential benefit in reducing respiratory infections.
@@
 
-To illustrate possible transmission chains, we linked positive saliva samples within classes that belong to the same virus and are less than 1~week apart.
-Based on this spatiotemporal analysis of student saliva samples, we identified 10~possible transmission chains, which are shown in \Cref{fig:molecular-network}.
+To illustrate possible transmission chains, we linked positive saliva samples within classes that belong to the same virus and were less than 1~week apart.
+Based on this spatiotemporal analysis of student saliva samples, we identified10~possible transmission chains, which are shown in \Cref{fig:molecular-network}.
 The longest transmission chains possibly occurred in January and refer to the cluster of AdV infections in class~A and Flu~B infections in class~B.
-We performed molecular genotyping of saliva, bioaerosol, and air filter samples to verify our proposed transmission networks.
-However, molecular genotyping was not successful because we were unable to amplify and sequence any of the gene targets.The virus-specific variation between classes and the temporal distribution of positive samples suggest that transmission occurred within classes.
+We also performed molecular genotyping of saliva, bioaerosol, and air filter samples to verify our proposed transmission networks.
+However, molecular genotyping was not successful because we were unable to amplify and sequence any of the gene targets.Nevertheless, the virus-specific variation between classes and the temporal distribution of positive samples suggest that transmission occurred within classes.
 
@@
 Positive samples are linked if they belong to the same virus and are less than 1~week apart.
+Positive samples from the air and filters as blank squares.
 Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
@@
 
-% Technical reasons for the differences in airborne detection are unlikely.
+Technical reasons for the differences in airborne detection are unlikely.
 The bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.

### 2023-10-06 doc/paper/supplements.tex
Finalized draft of long manuscript

@@
 \author[2,4]{Pascal Bittel}
+\author[1]{Kathrin Zürcher}
 \author[4]{Lavinia Furrer}
-\author[4]{Alban Ramette}
-\author[4]{Ronald Dijkman}
-\author[5]{Jenna Kelly}
-\author[1]{Kathrin Zürcher}
+\author[1]{Simon Bertschinger}
+\author[2,4]{Alban Ramette}
+\author[2,4]{Ronald Dijkman}
 \author[1]{Matthias Egger}
-\author[2,6]{Tina Hascher}
+\author[2,5]{Tina Hascher}
 \author[1*,2]{Lukas Fenner}
@@
 \affil[3]{Department of Infectious Diseases, Inselspital, Bern University Hospital, University of Bern, Bern, Switzerland}
-\affil[4]{Institute of Infectious Diseases, University of Bern, Bern, Switzerland}
-\affil[5]{Interfaculty Bioinformatics Unit, University of Bern, Bern, Switzerland}
-\affil[6]{Institute of Educational Science, University of Bern, Bern, Switzerland}
+\affil[4]{Institute for Infectious Diseases, University of Bern, Bern, Switzerland}
+\affil[5]{Institute of Educational Science, University of Bern, Bern, Switzerland}
 

### 2023-10-06 analysis/mol_analysis.Rmd
update network plot with air and filter samples

@@
     "A", "20-02-2023", "AdV", "Air",
-    "B", "01-02-2023", "Flu B", "Filter", # we use the middle of the week(s) when the swabs were taken
-    "A", "05-03-2023", "CoV", "Filter",
-    "A", "05-03-2023", "AdV", "Filter",
-    "A", "05-03-2023", "HRV", "Filter"),
+    "B", "01-02-2023", "Flu B", "Filter", # we set the dates for the filter swabs to align with saliva
+    "A", "27-02-2023", "CoV", "Filter",
+    "A", "03-03-2023", "AdV", "Filter",
+    "A", "08-03-2023", "HRV", "Filter"),
   ncol = 4, byrow = T

### 2023-10-06 doc/paper/manuscript-lancet.tex
Updates from Overleaf

--- 2023-10-06 doc/paper/manuscript.tex
@@
 \usepackage{import}
-\usepackage{main}
+\usepackage{main-lancet}
 
@@
 \begin{abstract}\normalfont
-\noindent\textbf{Background:} Airborne respiratory infections can be transmitted in crowded indoor settings, including schools.
-Infection control measures such as air ventilation or filtration are supposed to reduce the risk of airborne transmission.
-The aim of this study was to evaluate the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory viral infections in schools using a multiple-measurement approach.
+\noindent\textbf{Background:} Infection control measures should reduce airborne transmission.
+We aimed to evaluate the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory viral infections in schools using a multiple-measurement approach.
 \medskip
 
-\noindent\textbf{Methods:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), audio (recordings to detect coughing), and molecular data (bioaerosol and saliva samples) over seven weeks from 16 January to 11 March 2023 in a secondary school (38 students in two classes) in Switzerland.
-We performed laboratory and molecular analyses to detect a panel of 19 respiratory viruses.
-We installed air cleaners in a cross-over design and compared positive samples and particle concentrations between weeks with and without air filters.
-We modeled the number of new infections using a Bayesian negative binomial regression model, adjusting for incubation periods, number of recent and cumulative infections, ventilation rate, number of students in class, and community transmission.
+\noindent\textbf{Methods:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), audio (recordings detecting coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2023 in two classes of a Swiss secondary school.
+Using a cross-over study design, we compared particle concentrations, coughing, and the risk of infection with and without air cleaners.
 \medskip
 
-\noindent\textbf{Results:} Molecular analysis detected multiple viruses in saliva samples throughout the study (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus), but rarely in bioaerosol samples (2/105 positive: 1~rhinovirus and 1~adenovirus) and on the filters of air cleaners (4/160~positive: 1~influenza~B, 1~HRV, 1~adenovirus, and 1~SARS-CoV-2).
+\noindent\textbf{Results:} Molecular analysis detected multiple viruses in saliva (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza), but rarely in bioaerosol samples (2/105 positive) and on the filters of air cleaners (4/160~positive).
 Nevertheless, spatiotemporal analysis of saliva samples suggested transmission within classrooms.
-The detection rate of positive saliva samples was comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61).
-Daily average CO$_2$ levels were 1,769$\pm$391\,ppm ($\pm$ standard deviation) with air cleaners and 1,636$\pm$341\,ppm without air cleaners, corresponding to similar air change rates (0.74$\pm$0.41\,h$^{-1}$ vs.
-0.68$\pm$0.32\,h$^{-1}$).
-Daily average aerosol number concentrations were 27$\pm$34\,1/cm$^3$ with and 95$\pm$81\,1/cm$^3$ without air cleaners, corresponding to an estimated adjusted decrease of 77\% (95\%-CrI 63\% to 86\%).
-Slightly less coughs  were detected with air cleaners (2.5~coughs/min with vs.
-3.1~coughs/min without air cleaners).
-Absences related to respiratory infections were less frequent with air cleaners (13~absences with vs.
-21~absences without air cleaners).
-The estimated risk of infection tended to be lower with air cleaners (adjusted relative risk 0.72, 95\%-Credible Interval 0.44 to 1.15).
+Daily average CO$_2$ levels with and without air cleaners were similar (1,769$\pm$391\,ppm vs.
+1,636$\pm$341\,ppm).
+Particle concentrations decreased by 77\% (95\%-credible interval [CrI] 63\% to 86\%), coughing was less frequent ($-$0.5~coughs/min), and the estimated risk of infection likely lower with air cleaners (adjusted relative risk 0.72, 95\%-CrI 0.44 to 1.15).
 \medskip
 
-% Study limitations include unknown specific causes of absences related to respiratory infections (i.e which virus caused the sickness).
-Furthermore, students occasionally had lessons outside the classroom, thus limiting the potential effectiveness of air cleaners.
-
-\noindent\textbf{Conclusions:} During a non-pandemic period of winter 2023, a wide range of non-SARS-CoV-2 respiratory viruses were detected in the saliva of students in a Swiss school.
-Although airborne molecular detection was rare, transmission of these respiratory viruses seemed likely in schools, suggesting that prolonged close contact is required for transmission.
-Air cleaners improved air quality, were associated with lower symptom intensity, and showed a small, potential benefit in reducing respiratory infections.
+\noindent\textbf{Conclusions:} A wide range of respiratory viruses were detected in students' saliva.
+Although airborne detection was rare, transmission seemed likely in schools, suggesting that prolonged close contact is required for transmission.
+Air cleaners improved air quality, were associated with lower symptom intensity, and showed a small, potential benefit in reducing infections.
 
@@
 Recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
-Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and their respiratory plumes can reach more than 2m\cite{Coleman2022,Wang2020,Heneghan2021}, exceeding the social distancing rules recommended by public health organizations\cite{Bazant2021PNAS,Trivedi2021PhyFluids,Lindsley2010CID}.
+Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and travel long distances\cite{Coleman2022,Wang2020,Heneghan2021}.
 
-The generation, spread, and concentration of bioareosols can be influenced by infection control measures.
-For example, mandatory face mask use reduced transmission of SARS-CoV-2 during the COVID-19 pandemic\cite{Gettings2021,Heinsohn2022,Leech2022PNAS,Mitze2020PNAS,Banholzer2021PLOS}.
-In addition, improved ventilation systems are critical for a healthy indoor environment both in pandemic and non-pandemic times\cite{Wang2021,Morawska2021}.
+Improved ventilation systems are critical for a healthy indoor environment and can reduce the risk of respiratory transmission\cite{Wang2021,Morawska2021}.
 Portable HEPA-air filtration devices (air cleaners) may be another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear.
-Several studies have shown that air cleaners reduce the concentration of aerosols and particle matter in indoor air\cite{Park2020Build,Buising2022InfContr,Ren2021Dent,Banholzer2023PLoSMed}.
-These findings suggest potential protection against respiratory infections, as long-term exposure to higher aerosol and particle matter mass concentrations has been associated with increased risk of respiratory infections\cite{Gordon2014Resp} and, more recently, with the incidence and mortality of COVID-19\cite{Kelly2023Atmos,Veronesi2022OccEnvMed,Travaglio2021EnvPoll,DeAngelis2021EnvRes}.
+Several studies have shown that air cleaners reduce particle concentrations in indoor air\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed}, which is associated with a reduced risk of respiratory infection and mortality\cite{Gordon2014Resp,Kelly2023Atmos,DeAngelis2021EnvRes}.
 In addition, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}.
 Simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission  of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
-We have studied the effectiveness of air cleaners in reducing transmission of SARS-CoV-2 in schools during the pandemic\cite{Banholzer2023PLoSMed}, but the analysis was limited by the study design (installation of air cleaners at the end of the Omicron wave when most students were already infected).
-To date, empirical evidence linking the use of air cleaners to a reduced risk of infection is lacking.
-It remains unclear whether reducing particle concentrations and removing bioaerosols from the air will result in reduced transmission of respiratory infections.
+%We have studied the effectiveness of air cleaners in reducing transmission of SARS-CoV-2 in schools during the pandemic\cite{Banholzer2023PLoSMed}, but the analysis was limited by the study design (installation of air cleaners at the end of the Omicron wave when most students were already infected).
+To date, it remains unclear whether reducing particle concentrations and removing bioaerosols from the air will result in reduced transmission of respiratory infections.
 
@@
 We employed a cross-over study design under non-pandemic conditions in the winter of 2023.
-We collected epidemiological (absences likely related to respiratory infections), environmental (CO$_2$, aerosol and particle concentrations), audio (recordings to detect coughing), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
+We collected epidemiological (absences likely related to respiratory infections), environmental (CO$_2$, aerosol and particle concentrations), audio (recordings detecting coughing), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
 We determined changes in particle concentrations, respiratory virus positivity rates, and modeled the risk of infection.
@@
 \noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from 26~January to 11~March 2023.
-\Cref{fig:study-setup} shows the schematic study setup and Table~\zref{tab:data} in \supp \, provides an overview of the types of data collected in each classroom.
+\cref{fig:study-setup} shows the schematic study setup and \supp~table~\zref{tab:data} provides an overview of the types of data collected in each classroom.
 
@@
 
-\noindent We employed a cross-over design to study the effectiveness of air cleaners, distinguishing two study conditions in each class: weeks with and weeks without air cleaners being installed in each classroom (\Cref{tab:study_design}).
+\noindent We employed a cross-over design to study the effectiveness of air cleaners, distinguishing two study conditions in each class: weeks with and weeks without air cleaners being installed in each classroom (\cref{tab:study_design}).
 Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China; approx.
@@
 For absences due to illness, we collected symptoms and the date of symptom onset.
-We also collected data on laboratory tests but none of the students reported a test result during the study.
-Based on the epidemiological data, we defined a case of respiratory infection as an absence in which the student reported an illness with at least one of the following symptoms: fever, cough, fatigue, loss of taste or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, and stomach.
-Students were asked to report the first date they experienced symptoms, so that a student absent on Monday could report Saturday or Sunday as the day when symptoms began to be felt.
-We will refer respiratory cases by their dates of symptom onset, which usually corresponded to the absence date, unless the student attended the school while experiencing symptoms.
-Informed consent for one student could not be obtained and the single absence of this student was considered unrelated to a respiratory infection.
-Reports about absences were entered electronically into a REDcap database\cite{Harris2009,Harris2019}.
-The line list data is shown in Table~\zref{tab:epi-data-line-list} in \supp.
-
-In addition, both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below).
-Saliva samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Galar2021,To2019,Huber2021}.
-The line list data is shown in Table~\zref{tab:mol-data-line-list} in \supp.
+Based on the epidemiological data, we defined a case of respiratory infection as an absence in which the student reported an illness with at least one respiratory symptoms (\supp~text~\zref{sec:case-data}).
+The line list data is shown in \supp~table~\zref{tab:epi-data-line-list}.
 \medskip
 
-\noindent\textbf{Environmental data} \smallskip
+\noindent\textbf{Molecular data} \smallskip
 
-\noindent \emph{CO$_2$ and particle measurements}
+\noindent \emph{Saliva sampling} \smallskip
 
-\noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute.
-This device was used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}.
-A transient mass balance model was used to estimate the air change rate from indoor CO$_2$ levels\cite{Batterman2017IJERPH}, assuming that classrooms were fully occupied during lessons and that room occupancy increased (decreased) linearly 10~minutes before (after) the start (end) of a lesson.
-Particle concentrations were filtered according to the times when students were in the classroom, which were monitored together with the times during which windows were opened.
+\noindent Both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below).
+Saliva samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Galar2021,To2019,Huber2021}.
+The line list data is shown in \supp~table~\zref{tab:mol-data-line-list}.
 \medskip
 
-\noindent \emph{Bioaerosol sampling and HEPA filters of the air cleaners}
+\noindent \emph{Bioaerosol sampling and HEPA filters of the air cleaners} \smallskip
 
@@
 At the end of the day, samples were transported to the Institute of Infectious Diseases (IFIK) and stored at $-$80°C.
-After each intervention phase (see \Cref{tab:study_design}), the HEPA filters were carefully removed from the air cleaners, divided into 20~fields, and one swab per field was taken for a total of 20~swabs per filter.\medskip
+After each intervention phase (see \cref{tab:study_design}), the HEPA filters were carefully removed from the air cleaners, divided into 20~fields, and one swab per field was taken for a total of 20~swabs per filter.
+\medskip
+
+\noindent\textbf{Environmental data} \smallskip
+
+\noindent \emph{CO$_2$ and particle measurements}
+
+\noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute.
+This device was used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}.
+A transient mass balance model was used to estimate the air change rate from indoor CO$_2$ levels\cite{Batterman2017IJERPH}.%, assuming that classrooms were fully occupied during lessons and that room occupancy increased (decreased) linearly 10~minutes before (after) the start (end) of a lesson.
+Particle concentrations were filtered according to the times when students were in the classroom. %, which were monitored together with the times during which windows were opened.
 
@@
 The human saliva samples were directly analyzed without prior filtration.
-The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19~respiratory viruses: SARS-CoV-2 (CoV), Influenza A virus (Flu A), Influenza B virus (Flu B), Human respiratory syncytial virus A/B (RSV), Human metapneumovirus (MPV), Human adenovirus A/B/C/D/E/F (AdV), Human rhinovirus A/B/C (HRV), and Human parainfluenza virus 1/2/3/4 (PIV) in one single reaction.
+The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19~respiratory viruses, including SARS-CoV-2 (CoV), influenza (Flu), respiratory syncytial (RSV), metapneumovirus (MPV), adenovirus (AdV), rhinovirus (HRV), and parainfluenza (PIV).
 %/Viral genome load (VGL) of specimens was quantified (copies/L) by curve extrapolation derived from standardized ATCC quantitative genomic standard dilutions (ATCC-VR-3347D, LGC Standards, LGC Group UK).
@@
 We installed one portable audio recorder (ZOOM H6; New York, USA) in each classroom to record sounds continuously.
-We detected the number of coughs per minute using an AI algorithm based on a convolutional neural network\cite{Bertschinger2023IEEE}, which was recently trained and validated with audio data from the waiting room of a primary care clinic\cite{Zurcher2022}.
+We detected the number of coughs per minute using an AI algorithm\cite{Bertschinger2023IEEE}.
 
@@
 
-All statistical analyses were described in the statistical analysis plan (SAP)\cite{Banholzer2023SAP}, which was published before analyses began.
+All statistical analyses were described in the statistical analysis plan (SAP)\cite{Banholzer2023SAP}, which was published ahead of analyses.
 In contrast to the SAP, the number of positive bioaerosol samples and the viral load concentration of positive samples could not be analyzed because there were too few positive samples during the study.
 Apart from that there were no major deviations from the SAP regarding the conducted analyses.
-There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in Supplementary Text~\zref{sec:transmission-model} to Supplementary Text~\zref{sec:env-regression-model} in \supp, which provide detailed descriptions of the models used in each analysis.
+There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in \supp~text~\zref{sec:transmission-model}-\zref{sec:env-regression-model}, which provide detailed descriptions of the models used in each analysis.
 \medskip
@@
 
-A detailed description of the model and choice of priors for all model parameters are provided in Supplementary Text~\zref{sec:transmission-model} in \supp.
+A detailed description of the model and choice of priors for all model parameters are provided in \supp~text~\zref{sec:transmission-model}.
 Importantly, since multiple viruses were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the virus-specific distributions.
@@
 \noindent Molecular analysis determined which saliva samples were positive for a respiratory virus.
-We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (Supplementary Text~\zref{sec:multinomial-model} in \supp) where the expected number of positive samples for a virus was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each virus.
+We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~text~\zref{sec:multinomial-model}) where the expected number of positive samples for a virus was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each virus.
 The number of susceptibles was computed as the difference between the number of students per class and the cumulative number of positive samples.
@@
 \noindent We computed the average particle concentrations per day and compared them between study conditions.
-We estimated the reduction in particle concentrations using Bayesian log-linear regression models (Supplementary Text~\zref{sec:env-regression-model} in \supp), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+We estimated the reduction in particle concentrations using Bayesian log-linear regression models (\supp~text~\zref{sec:env-regression-model}), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
 \medskip
@@
 \noindent We computed the daily number of detected coughs and compared it between study conditions.
-We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model (Supplementary Text~\zref{sec:aud-regression-model}) with time in class as the model offset, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model (\supp~text~\zref{sec:aud-regression-model}) with time in class as the model offset, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
 In addition, we used a Bayesian Negative Binomial hierarchical (random effects) regression model to estimate the association between the number of coughs and the number of positive saliva test results for respiratory viruses.
@@
 We analyzed 448~saliva samples over the study across classes.
-We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenovirus (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~virus~(PIV) (\Cref{fig:molecular-descriptives}a).
-There was an equal number of positive samples with and without air cleaners (\Cref{fig:molecular-descriptives}b).
-Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (Figure~\zref{fig:mol-estimation-results-sensitivity} in \supp).
-The distribution of positive samples varied between classes (\Cref{fig:molecular-descriptives}c).
+We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenovirus (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~virus~(PIV) (\cref{fig:molecular-descriptives}a).
+There was an equal number of positive samples with and without air cleaners (\cref{fig:molecular-descriptives}b).
+Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
+The distribution of positive samples varied between classes (\cref{fig:molecular-descriptives}c).
 For example, during the first three study weeks, all but one sample was positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
@@
 To illustrate possible transmission chains, we linked positive saliva samples within classes that belong to the same virus and were less than 1~week apart.
-Based on this spatiotemporal analysis of student saliva samples, we identified10~possible transmission chains, which are shown in \Cref{fig:molecular-network}.
+Based on this spatiotemporal analysis of student saliva samples, we identified10~possible transmission chains, which are shown in \cref{fig:molecular-network}.
 The longest transmission chains possibly occurred in January and refer to the cluster of AdV infections in class~A and Flu~B infections in class~B.
@@
 Positive samples are linked if they belong to the same virus and are less than 1~week apart.
-Positive samples from the air and filters as blank squares.
+Positive samples from the air and filters as blank squares aligned.
 Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
@@
 
-There was a strong difference in particle concentrations between study conditions (\Cref{fig:palas-results}a).
+There was a strong difference in particle concentrations between study conditions (\cref{fig:palas-results}a).
 Aerosol number concentration with air cleaners were markedly lower (27$\pm$34\,1/cm$^3$) than without air cleaners (95$\pm$81\,1/cm$^3$).
 Similarly, particle mass concentrations (\eg PM$_1$) were markedly lower with air cleaners (1.16$\pm$1.33\,$\mu$gm$^{-3}$) than without air cleaners (4.20$\pm$3.74\,$\mu$gm$^{-3}$).
-There was no considerable difference in other environmental variables (Figure~\zref{fig:env-descriptives-other-vars} in \supp).
+There was no considerable difference in other environmental variables (\supp~figure~\zref{fig:env-descriptives-other-vars}).
 In particular, daily average CO$_2$ levels (air change rates) were 1,769$\pm$391\,ppm (0.74$\pm$0.41\,h$^{-1}$) with air cleaners and 1,636$\pm$341\,ppm (0.68$\pm$0.32\,h$^{-1}$), suggesting that ventilation was comparable between study conditions.
-When adjusting for air change rates and multiple other factors, the aerosol number concentration decreased by 76\% (95\%-CrI 63\% to 86\%) with air cleaners (\Cref{fig:palas-results}b and Table~\zref{tab:palas-est-results} in \supp).
+When adjusting for air change rates and multiple other factors, the aerosol number concentration decreased by 76\% (95\%-CrI 63\% to 86\%) with air cleaners (\cref{fig:palas-results}b and \supp~table~\zref{tab:palas-est-results}).
 The decrease in the concentration of larger particles (PM$_{10}$) was greater than the decrease in the concentration of smaller particles (PM$_1$ to PM$_{4}$).
@@
 Incidentally, there were more cases in class~B (21) than in class~A (14), which is incidentally also the class where more students recently recovered from a SARS-CoV-2 infection or were vaccinated.
-Except for one study week, there was an equal or higher number of cases without air cleaners (\Cref{fig:redcap-results}a).
+Except for one study week, there was an equal or higher number of cases without air cleaners (\cref{fig:redcap-results}a).
 Based on our Bayesian latent hierarchical regression model, we estimated a trend towards a lower risk of infection with air cleaners (adjusted relative risk 0.72, 95\%-Credible Interval 0.44 to 1.15).
-As a result, the estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\Cref{fig:redcap-results}b).
+As a result, the estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\cref{fig:redcap-results}b).
 The posterior probability that air cleaners reduced the risk of infection was 90.8\%, also suggesting a beneficial effect of air cleaners.
-Detailed estimations are provided in the Supplementary Text~\zref{sec:detailed-redcap} in \supp.
+Detailed estimations are provided in the \supp~text~\zref{sec:detailed-redcap}.
 
@@
 
-Slightly fewer coughs were detected in audio recordings during study weeks where air cleaners were installed (\Cref{fig:coughing}a).
-On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners (adjusted relative risk 0.93, 95\%-CrI 0.85 to 1.02).
-Significantly more coughs were detected in class~B (adjusted relative risk 2.16, 95\%-CrI 1.90 to 2.46), suggesting a link with the observed variation in virus-specific transmission between classes (\Cref{fig:molecular-descriptives}c).
-As suspected, the frequency of coughing was associated with the number of positive saliva samples by respiratory virus (\Cref{fig:coughing}b).
+Slightly fewer coughs were detected in audio recordings during study weeks where air cleaners were installed (\cref{fig:coughing}a).
+On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners ($\Delta$ $-$0.6~coughs/min; adjusted relative risk 0.93, 95\%-CrI 0.85 to 1.02).
+Significantly more coughs were detected in class~B (adjusted relative risk 2.16, 95\%-CrI 1.90 to 2.46), suggesting a link with the observed variation in virus-specific transmission between classes (\cref{fig:molecular-descriptives}c).
+As suspected, the frequency of coughing was associated with the number of positive saliva samples by respiratory virus (\cref{fig:coughing}b).
 Coughing was more frequent during Flu~B positivity (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent during AdV positivity (0.87, 95\%-CrI 0.49 to 1.12).
@@
 In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 Omicron wave.
-\Cref{fig:comparison} shows a comparison of the results.
-The distribution of respiratory infections is different (\Cref{fig:comparison}a) because we almost exclusively detected SARS-CoV-2 in our previous study during the COVID-19 pandemic, as compared to a range of respiratory infections in this study during non-pandemic times.
-Although there were more positive saliva samples in this study, the number of respiratory infections detected in bioaerosols and on the HEPA-filters of the air cleaners was lower (\Cref{fig:comparison}b).
-The reductions in particle concentrations were larger and with lower uncertainty in this study (\Cref{fig:comparison}c).
-The risk of infection tended to be lower with air cleaners in this study where air cleaners were analyzed using a cross-over design, but not in our previous study where air cleaners were installed only at the end of the study when most students had already been infected with SARS-Cov-2(\Cref{fig:comparison}d).
+\cref{fig:comparison} shows a comparison of the results.
+The distribution of respiratory infections is different (\cref{fig:comparison}a) because we almost exclusively detected SARS-CoV-2 in our previous study during the COVID-19 pandemic, as compared to a range of respiratory infections in this study during non-pandemic times.
+Although there were more positive saliva samples in this study, the number of respiratory infections detected in bioaerosols and on the HEPA-filters of the air cleaners was lower (\cref{fig:comparison}b).
+The reductions in particle concentrations were larger and with lower uncertainty in this study (\cref{fig:comparison}c).
+The risk of infection tended to be lower with air cleaners in this study where air cleaners were analyzed using a cross-over design, but not in our previous study where air cleaners were installed only at the end of the study when most students had already been infected with SARS-Cov-2(\cref{fig:comparison}d).
 
@@
 To some extent, the differences between the two studies may be explained by different adjustments.
-While we could not adjust for ventilation in our previous study, the estimates in this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see Supplementary Text~\zref{sec:detailed-molecular} in \supp).
+While we could not adjust for ventilation in our previous study, the estimates in this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see \supp~text~\zref{sec:detailed-molecular}).
 Furthermore, the distribution of respiratory viruses in this study was very different from that in our previous study.

### 2023-10-06 doc/paper/supplements.tex
Updates from Overleaf

@@
 
+Reports about absences were entered electronically into a REDcap database\cite{Harris2009,Harris2019}.
 \Cref{tab:epi-data-line-list} shows the line list data for respiratory cases based on epidemiological data.
 For each class and case, it shows the date when the student was absent, s/he reported symptoms, and returned to school.
-Recall that a case of respiratory infection was defined as an absence where the student reported a sickness with at least one of the following symptoms: fever, coughing, tiredness, loss of test or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, stomach.
+We also collected data on laboratory tests but none of the students reported a test result during the study.
+A case of respiratory infection was defined as an absence where the student reported a sickness with at least one of the following symptoms: fever, coughing, tiredness, loss of test or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, stomach.
+Students were asked to report the first date they experienced symptoms, so that a student absent on Monday could report Saturday or Sunday as the day when symptoms began to be felt.
+We will refer respiratory cases by their dates of symptom onset, which usually corresponded to the absence date, unless the student attended the school while experiencing symptoms.
+Informed consent for one student could not be obtained and the single absence of this student was considered unrelated to a respiratory infection.
 

### 2023-10-16 analysis/aud_analysis.Rmd
update plots

--- 2023-08-22 analysis/aud_analysis.Rmd
@@
 coughs_pl <- coughs %>%
-  mutate(week = isoweek(date)) %>%
-  filter(week != 6) %>%
-  mutate(week = ifelse(week < 6, week - min(week) + 1, week - min(week))) %>%
-  group_by(class, week) %>%
-  summarize(coughs_per_min = sum(n_coughs) / sum(TT),
-            aircleaner = first(aircleaner)) %>%
+  group_by(aircleaner) %>%
+  summarize(coughs_per_min = sum(n_coughs) / sum(TT)) %>%
   ungroup() %>%
-  mutate(aircleaner = factor(aircleaner, levels = c("Yes", "No")),
-         class = factor(class, levels = c("A", "B"))) %>%
-  arrange(class, aircleaner) %>%
-  ggplot(aes(x = factor(week), y = coughs_per_min, fill = aircleaner, pattern = class)) +
-  geom_bar_pattern(stat = "identity", color = "black", pattern_fill = "white", pattern_angle = 45,
-                   pattern_density = 0.25, pattern_spacing = 0.05, width = .5) +
-  labs(x = "Study week", y = "Number of coughs per min", fill = "Air cleaner", pattern = "Class", title = "a") +
-  scale_fill_manual(values = rev(wes_palette("Royal1")[1:2])) +
-  scale_x_discrete(breaks = 1:7) +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = seq(0, 10, 2)) +
-  scale_pattern_manual(values = c(A = "stripe", B = "none")) +
+  mutate(aircleaner = factor(aircleaner, levels = c("No", "Yes"))) %>%
+  ggplot(aes(x = aircleaner, y = coughs_per_min, fill = aircleaner)) +
+  geom_bar(stat = "identity", color = "black") +
+  geom_text(mapping = aes(label = round(coughs_per_min, 1)),
+            size = 8 / cm(1), vjust = -1) +
+  labs(x = "Study week", y = "Number of coughs per min", fill = "Air cleaner", title = "d") +
+  scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
   theme_custom() +
-  theme(legend.position = "top",
+  theme(legend.position = "none",
         legend.direction = "horizontal",
@@
         plot.title.position = "plot") +
-  guides(fill = guide_legend(nrow = 1, override.aes = list(pattern = "none", order = 2)),
-         pattern = guide_legend(nrow = 1, override.aes = list(fill = "white"), order = 1))
+  guides(fill = "none")
 
@@
 
-save_plot(coughs_pl, pdf_file = "../results/cough-data/coughs-frequency-per-week.pdf", eps_file = "../results/cough-data/coughs-frequency-per-week.eps", w = 12, h = 8)
+save_plot(coughs_pl, 
+          pdf_file = "../results/cough-data/coughs-frequency-by-condition.pdf", 
+          eps_file = "../results/cough-data/coughs-frequency-by-condition.eps", 
+          w = 8, h = 8)
 ```
@@
   scale_color_manual(values = wes_palette("IsleofDogs1")) +
-  labs(x = "Relative risk", title = "b") +
+  labs(x = "Relative risk") +
   theme_custom() +
@@
 theta_pl
-save_plot(theta_pl, pdf_file = "../results/cough-data/coughs-virus-association.pdf", eps_file = "../results/cough-data/coughs-virus-association.eps", w = 12, h = 8)
+
+save_plot(theta_pl, 
+          pdf_file = "../results/cough-data/coughs-virus-association.pdf", 
+          eps_file = "../results/cough-data/coughs-virus-association.eps", 
+          w = 12, h = 8)
 ```

### 2023-10-16 analysis/env_analysis.Rmd
update network plot with air and filter samples
update plots

@@
 
+tt_co2 <- t.test(env_epi_data$CO2[env_epi_data$aircleaner=="Yes"], env_epi_data$CO2[env_epi_data$aircleaner=="No"])
+
 # by air cleaner
@@
 ## Boxplots
+
+```{r}
+# only particle number concentration
+boxpl_cn <- env_epi_data %>%
+  dplyr::select(class, date, aircleaner, Cn) %>%
+  ggplot(aes(x = aircleaner, fill = aircleaner, y = Cn)) +
+  geom_boxplot2() +
+  #geom_jitter(alpha = .3, width = .2, color = "grey") +
+  scale_fill_manual(values = wes_palette("Royal1")) +
+  scale_shape_manual(values = c(2,20)) +
+  scale_y_continuous(expand = c(0,0), limits = c(0, NA)) +
+  labs(title = "b", y = expression("Aerosol number concentration (1/cm"^3*")"), x = "Air cleaner") +
+  theme_custom() +
+  theme(legend.position = "none",
+        plot.title.position = "plot") +
+  guides(color = "none")
+
+boxpl_cn
+
+save_plot(boxpl_cn, 
+          pdf_file = "../results/env-data/aerosol-number-boxplot.pdf", 
+          eps_file = "../results/env-data/aerosol-number-boxplot.eps", 
+          w = 8, h = 8)
+```
 
@@
   guides(color = "none")
+
 save_plot(boxpl_pc, pdf_file = "../results/env-data/particles-boxplot.pdf", eps_file = "../results/env-data/particles-boxplot.eps", w = 16, h = 8)

### 2023-10-16 analysis/epi_analysis.Rmd
update plots

@@
 # number of cases by week
-cases_by_week_pl <- epi_data %>%
-  group_by(class, aircleaner) %>%
+cases_by_cond_pl <- epi_data %>%
+  group_by(aircleaner) %>%
   summarize(cases = sum(new_cases)) %>%
   ungroup() %>%
-  mutate(aircleaner = factor(aircleaner, levels = c("No", "Yes")),
-         class = factor(class, levels = c("A", "B"))) %>%
-  arrange(class, aircleaner) %>%
+  mutate(aircleaner = factor(aircleaner, levels = c("No", "Yes"))) %>%
   ggplot(aes(x = aircleaner, y = cases, fill = aircleaner)) +
-  geom_bar_pattern(mapping = aes(pattern = class),
-                   stat = "identity", color = "black", pattern_fill = "white", pattern_angle = 45,
-                   pattern_density = 0.25, pattern_spacing = 0.05, width = .5) +
-  geom_text(data = epi_data %>% group_by(aircleaner) %>% summarize(cases = sum(new_cases)) %>% ungroup(),
-            mapping = aes(label = cases),
+  geom_bar(stat = "identity", color = "black") +
+  geom_text(mapping = aes(label = cases),
             size = 8 / cm(1), vjust = -1) +
-  labs(x = "Air cleaner", y = "Number of cases", fill = "Air cleaner", pattern = "Class", title = "a") +
+  labs(x = "Air cleaner", y = "Number of cases", fill = "Air cleaner", title = "c") +
   scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
   scale_y_continuous(expand = expansion(mult = c(0, 0.1)), breaks = scales::pretty_breaks(4)) +
-  scale_pattern_manual(values = c(A = "stripe", B = "none")) +
   theme_custom() +
-  theme(legend.position = "top",
-        legend.box = "horizontal",
-        legend.direction = "horizontal",
-        legend.key.width = unit(0.5, "cm"),
-        legend.key.height = unit(0.3, "cm"),
-        plot.title.position = "plot") +
-  guides(fill = "none",
-         pattern = guide_legend(nrow = 1, override.aes = list(fill = "white"), order = 1))
+  theme(plot.title.position = "plot") +
+  guides(fill = "none")
 
-cases_by_week_pl
+cases_by_cond_pl
 
-save_plot(cases_by_week_pl,
+save_plot(cases_by_cond_pl,
           pdf_file = "../results/epi-data/cases_by_condition.pdf",
           eps_file = "../results/epi-data/cases_by_condition.eps",
-          w = 6, h = 8) 
+          w = 8, h = 8) 
 ```
@@
   scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0)), breaks = seq(0, 200, 25)) +
-  labs(x = "Estimated number of infections", y = "Scenario", fill = "CrI", title = "b") +
+  labs(x = "Estimated number of infections", y = "Scenario", fill = "CrI") +
   theme_custom() +
@@
           eps_file = "../results/epi-data/avoided_infections.eps",
-          w = 10, h = 8)
+          w = 12, h = 8)
 ```

### 2023-10-16 analysis/mol_analysis.Rmd
update plots

@@
   arrange(class, week, desc(variable))
+```
 
+```{r}
 pos_week_pl <- ggplot(mapping = aes(x = week, y = n)) +
@@
   scale_x_continuous(breaks = seq(1, 7, 1)) +
-  labs(y = "Number of positive samples", x = "Study week", fill = "Virus", title = "c") +
+  labs(y = "Number of positive samples", x = "Study week", fill = "Virus") +
   theme_custom() +
@@
   guides(fill = guide_legend(nrow = 1))
+
 pos_week_pl
@@
 df_pos_stud <- df_pos %>%
-  group_by(aircleaner, variable) %>%
+  group_by(aircleaner) %>%
   summarize(n = sum(n)) %>%
-  ungroup()
-pos_cond_pl <- ggplot(mapping = aes(x = aircleaner, y = n, fill = variable)) +
-  geom_bar(data = df_pos_stud, stat = "identity", position = "stack") +
+  ungroup() %>%
+  mutate(aircleaner = factor(aircleaner, levels = c("No", "Yes")))
+
+pos_cond_pl <- ggplot(mapping = aes(x = aircleaner, y = n, fill = aircleaner)) +
+  geom_bar(data = df_pos_stud, stat = "identity", position = "stack", color = "black") +
   geom_text(data = subset(df_pos_stud, n != 0), mapping = aes(label = n), 
-            size = 8 / cm(1), position = position_stack(vjust = 0.5), color = "white") +
-  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
-  scale_y_continuous(expand = c(0,0)) +
-  labs(y = "Number of positive samples", x = "Air cleaner", fill = "Virus", title = "b") +
+            size = 8 / cm(1), vjust = -1) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
+  scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
+  labs(y = "Number of positive samples", x = "Air cleaner", fill = "Virus", title = "a") +
   theme_custom() +
   theme(panel.grid = element_blank(),
-        legend.position = "bottom",
-        plot.title.position = "plot") +
+        plot.title.position = "plot",
+        legend.position = "top",
+        legend.title = element_blank(),
+        legend.key.width = unit(.3, "cm"),
+        legend.key.height = unit(.1, "cm")) +
   guides(fill = guide_legend(nrow = 1))
+
 pos_cond_pl
+
+save_plot(pos_cond_pl, 
+          "../results/mol-data/saliva-by-study-condition.pdf",
+          "../results/mol-data/saliva-by-study-condition.eps",
+          w = 8, h = 8)
 ```
@@
 
-pos_tot_pl <- df_pos_tot %>%
-  ggplot(aes(x = variable, y = n, fill = variable)) +
-  geom_bar(stat = "identity") +
-  geom_text(aes(label = n, y = n + 1), size = 8 / cm(1)) +
+pos_tot_pl <- ggplot(mapping = aes(y = n, x = 1, fill = variable)) +
+  geom_bar(data = df_pos_tot, stat = "identity", position = position_stack(reverse = T)) +
+  geom_text(data = subset(df_pos_tot, n != 0), mapping = aes(label = n), 
+            size = 8 / cm(1), position = position_stack(vjust = 0.5, reverse = T), color = "white") +
   scale_fill_manual(values = wes_palette("IsleofDogs1")) +
-  scale_y_continuous(expand = expansion(add = c(0,1))) +
-  labs(y = "Number of positive samples", x = "Virus", fill = "Virus", title = "a") +
+  scale_y_continuous(expand = c(0,0)) +
+  scale_x_continuous(expand = expansion(mult = c(0.05,0))) +
+  coord_flip() +
+  labs(y = "Number of positive samples", fill = "Virus", title = "a") +
   theme_custom() +
   theme(panel.grid = element_blank(),
-        legend.position = "bottom",
-        plot.title.position = "plot") +
+        plot.title.position = "plot",
+        legend.position = "top",
+        legend.title = element_blank(),
+        legend.key.width = unit(.3, "cm"),
+        legend.key.height = unit(.1, "cm"),
+        axis.title.y = element_blank(),
+        axis.line.y = element_blank(),
+        axis.text.y = element_blank(),
+        axis.ticks.y = element_blank()) +
   guides(fill = guide_legend(nrow = 1))
+
 pos_tot_pl
+
+save_plot(pos_tot_pl, 
+          "../results/mol-data/saliva-distribution.pdf",
+          "../results/mol-data/saliva-distribution.eps",
+          w = 16, h = 3.5)
 ```
 
-```{r}
-# combine
-descr <- pos_tot_pl + plot_spacer() + pos_cond_pl & theme(legend.position = "bottom", legend.key.height = unit(.1, "cm"),
-                                          legend.key.width = unit(.3, "cm"))
-descr <- descr + plot_layout(guides = "collect", widths = c(9.5, 1, 5.5), ncol = 2) 
-descr <- descr + pos_week_pl 
-descr <- descr + plot_layout(nrow = 2, heights = c(8, 8), design = 'ABC
-                             DDD')
-save_plot(descr, pdf_file = "../results/mol-data/descriptives.pdf", eps_file = "../results/mol-data/descriptives.eps", h = 16, w = 16)
-# saveRDS(pos_tot_pl + ggtitle(""), file = "../results/mol-data/descriptives-total.rds")
-# saveRDS(pos_cond_pl + ggtitle(""), file = "../results/mol-data/descriptives-cond.rds")
-# saveRDS(pos_week_pl + ggtitle(""), file = "../results/mol-data/descriptives-week.rds")
-```
+<!-- ```{r} -->
+<!-- # combine -->
+<!-- descr <- pos_tot_pl + plot_spacer() + pos_cond_pl & theme(legend.position = "bottom", legend.key.height = unit(.1, "cm"), -->
+<!--                                           legend.key.width = unit(.3, "cm")) -->
+<!-- descr <- descr + plot_layout(guides = "collect", widths = c(9.5, 1, 5.5), ncol = 2)  -->
+<!-- descr <- descr + pos_week_pl  -->
+<!-- descr <- descr + plot_layout(nrow = 2, heights = c(8, 8), design = 'ABC -->
+<!--                              DDD') -->
+<!-- save_plot(descr, pdf_file = "../results/mol-data/descriptives.pdf", eps_file = "../results/mol-data/descriptives.eps", h = 16, w = 16) -->
+<!-- # saveRDS(pos_tot_pl + ggtitle(""), file = "../results/mol-data/descriptives-total.rds") -->
+<!-- # saveRDS(pos_cond_pl + ggtitle(""), file = "../results/mol-data/descriptives-cond.rds") -->
+<!-- # saveRDS(pos_week_pl + ggtitle(""), file = "../results/mol-data/descriptives-week.rds") -->
+<!-- ``` -->
 
@@
               arrow = arrow(length = unit(0.2, "cm")), color = "black") +
-  labs(size = "Number of infections", fill = "Virus", shape = "") +
+  labs(size = "Number of infections", fill = "Virus", shape = "", title = "b") +
   coord_y_datetime(ylim = c(max(tnet$date_infection), min(tnet$date_infection))) +
@@
   theme_custom() +
-  theme(legend.position = "bottom",
+  theme(legend.position = "top",
         legend.direction = "horizontal",
         legend.box = "horizontal",
-        #plot.title.position = "plot",
+        plot.title.position = "plot",
         axis.title = element_blank(),
@@
 
-save_plot(tnet_pl, pdf_file = "../results/mol-data/network-plot.pdf", eps_file = "../results/mol-data/network-plot.eps", w = 16, h = 21)
+save_plot(tnet_pl, 
+          pdf_file = "../results/mol-data/network-plot.pdf", 
+          eps_file = "../results/mol-data/network-plot.eps", 
+          w = 16, h = 16.5)
 ```

### 2023-10-16 utils/plotting.R

@@ -4,6 +4,7 @@ require(gridExtra)
 require(GGally)
 require(Cairo)
 require(gg.layers)
+require(kunstomverse)
 require(wesanderson) # https://github.com/karthik/wesanderson
 
 # ggplot theme for Nature style

### 2023-10-16 doc/comparison-mcid12/manuscript.tex
Updates from Overleaf

--- 2023-10-06 doc/paper/manuscript-lancet.tex
@@
-\PassOptionsToPackage{top=3cm,left=3cm,right=3cm,bottom=3cm}{geometry}
-\documentclass[fleqn,11pt]{wlscirep}
-
-\usepackage{import}
-\usepackage{main-lancet}
-
-\renewcommand{\paragraph}[1]{\vspace{0.3cm}\noindent\underline{\emph{#1}}\hfill\noindent}
-
-% word count
-% \newcommand{\maincount}[1]{%
-%   \immediate\write18{texcount -1 -sum=1 -merge -q -nobib #1.tex > #1-words.sum}%
-%   \input{#1-words.sum}%
-% }
-
-% \newcommand{\abstractcount}[1]{%
-%   \immediate\write18{texcount -template="{abst}" #1.tex > #1-words.sum}%
-%   \input{#1-words.sum}%
-% }
-
-\begin{document}
-
-\doublespacing
-
-\title{\bfseries\LARGE\singlespacing{Transmission of respiratory viral infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}}
-
-% Alternative title: SARS-CoV-2 transmission in schools based on molecular and epidemiological data: potential effects of mask wearing and air cleaners
-
-% author list
-\author[1$\ddag$,2]{Nicolas Banholzer}
-\author[2,3]{Philipp Jent}
-\author[2,4]{Pascal Bittel}
-\author[1]{Kathrin Zürcher}
-\author[4]{Lavinia Furrer}
-\author[1]{Simon Bertschinger}
-\author[2,4]{Alban Ramette}
-\author[2,4]{Ronald Dijkman}
-\author[1]{Matthias Egger}
-\author[2,5]{Tina Hascher}
-\author[1*,2]{Lukas Fenner}
-
-\affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
-\affil[2]{Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland}
-\affil[3]{Department of Infectious Diseases, Inselspital, Bern University Hospital, University of Bern, Bern, Switzerland}
-\affil[4]{Institute for Infectious Diseases, University of Bern, Bern, Switzerland}
-\affil[5]{Institute of Educational Science, University of Bern, Bern, Switzerland}
-
-\affil[*]{Corresponding author: lukas.fenner@unibe.ch }
-
-\vspace{1em}
-
-\begin{information}\normalfont
-
-xxxx
-
-% \noindent\textbf{Running head}: SARS-COV-2 transmission in schools and effect of air cleaners
-
-% %\noindent\textbf{Subject categorization}: 6.20  Indoor Air; 10.11 Pediatrics: Respiratory Infections
-
-% \noindent\textbf{Word count}: \maincount{manuscript}words, abstract \abstractcount{manuscript}words (max. 500), title 157 chars (max. 200)
-
-% %\noindent\textbf{Inserts:} 2 tables, 6 figures, 44 references
-
-% \noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.
-
-% \vspace{1em}
-
-% \noindent\textbf{Funding}
-
-% \noindent This study is funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
-NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
-ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498).
-\medskip
-
-% \noindent\textbf{Contributions}
-
-% \noindent Conception and design: NB, LF.
-Epidemiological and environmental data collection: NB, PJ, TS, LF.
-Laboratory data collection: PB, LFu.
-Additional data collection: TH.
-Statistical analysis: NB, KZ.
-Genomic analysis: LB, LFu.
-Paper draft: NB, LF, ME.
-All authors reviewed and approved the final version of the manuscript.
-
-% \par
-\end{information}
-
-%TC:ignore
-\begin{abstract}\normalfont
-\noindent\textbf{Background:} Infection control measures should reduce airborne transmission.
-We aimed to evaluate the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory viral infections in schools using a multiple-measurement approach.
-\medskip
-
-\noindent\textbf{Methods:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), audio (recordings detecting coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2023 in two classes of a Swiss secondary school.
-Using a cross-over study design, we compared particle concentrations, coughing, and the risk of infection with and without air cleaners.
-\medskip
-
-\noindent\textbf{Results:} Molecular analysis detected multiple viruses in saliva (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza), but rarely in bioaerosol samples (2/105 positive) and on the filters of air cleaners (4/160~positive).
-Nevertheless, spatiotemporal analysis of saliva samples suggested transmission within classrooms.
-Daily average CO$_2$ levels with and without air cleaners were similar (1,769$\pm$391\,ppm vs.
-1,636$\pm$341\,ppm).
-Particle concentrations decreased by 77\% (95\%-credible interval [CrI] 63\% to 86\%), coughing was less frequent ($-$0.5~coughs/min), and the estimated risk of infection likely lower with air cleaners (adjusted relative risk 0.72, 95\%-CrI 0.44 to 1.15).
-\medskip
-
-\noindent\textbf{Conclusions:} A wide range of respiratory viruses were detected in students' saliva.
-Although airborne detection was rare, transmission seemed likely in schools, suggesting that prolonged close contact is required for transmission.
-Air cleaners improved air quality, were associated with lower symptom intensity, and showed a small, potential benefit in reducing infections.
-
-\par
-\end{abstract}
-
-\flushbottom
-\maketitle
-\setcounter{page}{1}
-\thispagestyle{fancy}
-
-\vspace{2em}
-
-\vspace{0.5em}
-
-\noindent\textbf{Keywords:} schools, air cleaner, respiratory viruses, airborne transmission, molecular detection
-% maximum of 3-5 keywords
-\newpage
-
-\sloppy
-\raggedbottom
-%TC:endignore
-
-\newpage
-
-\section{Introduction}
-
-The transmission of respiratory infections such as SARS-CoV-2 and influenza is difficult to mitigate and control.
-Person-to-person transmission occurs primarily through the release of respiratory particles containing the respective viruses.
-Recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
-Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and travel long distances\cite{Coleman2022,Wang2020,Heneghan2021}.
-
-Improved ventilation systems are critical for a healthy indoor environment and can reduce the risk of respiratory transmission\cite{Wang2021,Morawska2021}.
-Portable HEPA-air filtration devices (air cleaners) may be another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear.
-Several studies have shown that air cleaners reduce particle concentrations in indoor air\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed}, which is associated with a reduced risk of respiratory infection and mortality\cite{Gordon2014Resp,Kelly2023Atmos,DeAngelis2021EnvRes}.
-In addition, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}.
-Simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission  of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
-%We have studied the effectiveness of air cleaners in reducing transmission of SARS-CoV-2 in schools during the pandemic\cite{Banholzer2023PLoSMed}, but the analysis was limited by the study design (installation of air cleaners at the end of the Omicron wave when most students were already infected).
-To date, it remains unclear whether reducing particle concentrations and removing bioaerosols from the air will result in reduced transmission of respiratory infections.
-
-We used a multiple-measurement approach to study the transmission of viral respiratory viruses and the effect of air cleaners in school rooms.
-We employed a cross-over study design under non-pandemic conditions in the winter of 2023.
-We collected epidemiological (absences likely related to respiratory infections), environmental (CO$_2$, aerosol and particle concentrations), audio (recordings detecting coughing), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
-We determined changes in particle concentrations, respiratory virus positivity rates, and modeled the risk of infection.
-
-\newpage
-
-\section{Methods}
-
-\subsection{Study setting and design}
-
-\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from 26~January to 11~March 2023.
-\cref{fig:study-setup} shows the schematic study setup and \supp~table~\zref{tab:data} provides an overview of the types of data collected in each classroom.
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{../study_setting.pdf}
-    \caption{\textbf{Study setting}.
-Schematic study setup of the classrooms.
-One air cleaner was placed in the front and one in the back of the classrooms.
-All devices were placed at the head level of the students when they were seated.
-Both classrooms were not equipped with an active HVAC (Heating, Ventilation, Air conditioning) system, but were ventilated naturally by opening windows.
-}
-    \label{fig:study-setup}
-\end{figure}
-
-\subsection{Study intervention}
-
-\noindent We employed a cross-over design to study the effectiveness of air cleaners, distinguishing two study conditions in each class: weeks with and weeks without air cleaners being installed in each classroom (\cref{tab:study_design}).
-Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China; approx.
-USD\,250 per device).
-According to the manufacturer, these air cleaners run at a clean air delivery rate of 2$\,\times\,$600 m$^{3}$/h.
-% We measured a lower clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h when testing the devices in an empty classroom.
-Window ventilation was done at the discretion of the teachers, as there were no specific recommendations from the national health authorities at that time.
-
-\begin{table}[!htpb]
-    \footnotesize
-    \centering
-    \caption{\textbf{Study design.} Cross-over design describing when portable air cleaners (AC) were installed in the rooms of classes A and B during a seven-week study period from 16~January to 11~March 2023, excluding a week of vacation from February 6 to 11.
-A ``*'' indicates when swabs from the HEPA filters were taken, i.e.
-after each intervention phase and before the vacation.}\label{tab:study_design}
-    \begin{tabular}{l l l l l l l l l}
-    \toprule
-      & \textbf{Week 1} & \textbf{Week 2} & \textbf{Week 3} & Vacation & \textbf{Week 4} & \textbf{Week 5} & \textbf{Week 6} & \textbf{Week 7} \\
-      & Jan 16-21 & Jan 23-28 & Jan 30-Feb 3 & Feb 6-11 & Feb 13-18 & Feb 20-25 & Feb 27-March 3 & March 6-11 \\
-      \midrule
-      \textbf{A} & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{0000}*& \cellcolor{gray!10} None & & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{00009}* \\
-      \textbf{B} & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC\hphantom{0000000}* & & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{0000}*& \cellcolor{gray!10} None & \cellcolor{gray!10} None \\
-      \bottomrule
-    \end{tabular}
-\end{table}
-
-\subsection{Data collection}
-
-\noindent\textbf{Epidemiological data} \smallskip
-
-\noindent At the beginning of the study, we collected aggregated data on age, sex and COVID-19 vaccination and recovery status in the participating classes.
-On a daily basis, we collected data on each absent student, \ie the absence and return dates as well as the reason for the absence (illness, career days, or other).
-For absences due to illness, we collected symptoms and the date of symptom onset.
-Based on the epidemiological data, we defined a case of respiratory infection as an absence in which the student reported an illness with at least one respiratory symptoms (\supp~text~\zref{sec:case-data}).
-The line list data is shown in \supp~table~\zref{tab:epi-data-line-list}.
-\medskip
-
-\noindent\textbf{Molecular data} \smallskip
-
-\noindent \emph{Saliva sampling} \smallskip
-
-\noindent Both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below).
-Saliva samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Galar2021,To2019,Huber2021}.
-The line list data is shown in \supp~table~\zref{tab:mol-data-line-list}.
-\medskip
-
-\noindent \emph{Bioaerosol sampling and HEPA filters of the air cleaners} \smallskip
-
-\noindent We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL PBS, as previously reported in clinic settings\cite{Moore2021}.
-The Coriolis Micro Air could only operate shortly before and during break times to prevent unacceptable noise exposure (approximately 60\,min/day).
-In one class, we also used the BioSpot-VIVAS device sampler (Aerosol Devices Inc., Ft.
-Collins, CO, USA).
-This device samples airborne virus particles into a viral transport medium (VTM) using a water-based laminar flow condensation method\cite{Pan2016JAM,Lednicky2016AST}.
-BioSpot-VIVAS could operate throughout lessons.
-Removable parts of both sampling devices were regularly autoclaved to prevent contamination.
-At the end of the day, samples were transported to the Institute of Infectious Diseases (IFIK) and stored at $-$80°C.
-After each intervention phase (see \cref{tab:study_design}), the HEPA filters were carefully removed from the air cleaners, divided into 20~fields, and one swab per field was taken for a total of 20~swabs per filter.
-\medskip
-
-\noindent\textbf{Environmental data} \smallskip
-
-\noindent \emph{CO$_2$ and particle measurements}
-
-\noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute.
-This device was used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}.
-A transient mass balance model was used to estimate the air change rate from indoor CO$_2$ levels\cite{Batterman2017IJERPH}.%, assuming that classrooms were fully occupied during lessons and that room occupancy increased (decreased) linearly 10~minutes before (after) the start (end) of a lesson.
-Particle concentrations were filtered according to the times when students were in the classroom. %, which were monitored together with the times during which windows were opened.
-
-\subsection{Laboratory and molecular analyses}\label{sec:mol_analyses}
-
-\noindent Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL.
-The human saliva samples were directly analyzed without prior filtration.
-The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19~respiratory viruses, including SARS-CoV-2 (CoV), influenza (Flu), respiratory syncytial (RSV), metapneumovirus (MPV), adenovirus (AdV), rhinovirus (HRV), and parainfluenza (PIV).
-%/Viral genome load (VGL) of specimens was quantified (copies/L) by curve extrapolation derived from standardized ATCC quantitative genomic standard dilutions (ATCC-VR-3347D, LGC Standards, LGC Group UK).
-For the BioSpot-VIVAS air samples, the air concentration of the DNA was exactly calculated according to the airflow influx into the sampling device.
-%/
-The Limit of detection (LOD) of the assay SARS-CoV-2, IFA/IFB is 7.5x10$^0$ TCID/ml (SARS-CoV-2), 3.7x10$^{-1}$ TCID/ml (IFA), and 4.8x100 TCID/ml (100cp/ml) (IFB) respectively.
-
-\subsection{Molecular genotyping}
-
-We performed molecular genotyping for positive samples of AdV and Flu~B.
-For AdV (saliva, bioaerosol, and air filter samples), we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}.
-For Flu~B, we tried to determine the molecular transmission network using whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
-
-\subsection{Audio recordings and cough detection}
-
-We installed one portable audio recorder (ZOOM H6; New York, USA) in each classroom to record sounds continuously.
-We detected the number of coughs per minute using an AI algorithm\cite{Bertschinger2023IEEE}.
-
-\subsection{Statistical analyses and modeling}
-
-All statistical analyses were described in the statistical analysis plan (SAP)\cite{Banholzer2023SAP}, which was published ahead of analyses.
-In contrast to the SAP, the number of positive bioaerosol samples and the viral load concentration of positive samples could not be analyzed because there were too few positive samples during the study.
-Apart from that there were no major deviations from the SAP regarding the conducted analyses.
-There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in \supp~text~\zref{sec:transmission-model}-\zref{sec:env-regression-model}, which provide detailed descriptions of the models used in each analysis.
-\medskip
-
-\noindent\textbf{Risk of infection} \smallskip
-
-\noindent We estimated the relative risk of infection with air cleaners using a Bayesian latent hierarchical regression model.
-The observed outcome of this model is the number of new respiratory cases $C$ (absences related to respiratory infections by date of symptom onset) on day $t$ in class $j$, which are modeled with a Negative Binomial distribution.
-The expected number of new cases is the weighted sum of the number of new infections $I_{js}$ (unobserved/latent outcome) in the previous days $s<t$, with the weights corresponding to the probability distribution of the incubation period.
-The number of new infections is related to the presence of air cleaners using a regression-like log-link
-\begin{align}
-    \log I_{js} = \log F_{js} - \log N_{js} + \beta_0 + \beta_1 \cdot \text{AirCleaner}_{js},
-\end{align}
-where $F_{js}$ is the number of infections in the previous week (a proxy for the number of infectious students), $N_{js}$ is the cumulative number of infections (a proxy for the number of susceptible students), $\beta_0$ is the infection rate without air cleaners, and $\beta_1$ is the effect of air cleaners.
-Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the community (\ie the canton of Solothurn).
-
-A detailed description of the model and choice of priors for all model parameters are provided in \supp~text~\zref{sec:transmission-model}.
-Importantly, since multiple viruses were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the virus-specific distributions.
-The weights were inferred based on the proportion of positive saliva samples for each virus per week.
-As a result, the distribution of the incubation period used for modeling could change every week depending on the proportion of viruses that were detected in the students' saliva.
-Moreover, respiratory cases by date of symptom onset were under-reported for weekends and thus a re-weighting scheme was used to account for weekday effects.
-\medskip
-
-\noindent\textbf{Number of positive saliva samples} \smallskip
-
-\noindent Molecular analysis determined which saliva samples were positive for a respiratory virus.
-We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~text~\zref{sec:multinomial-model}) where the expected number of positive samples for a virus was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each virus.
-The number of susceptibles was computed as the difference between the number of students per class and the cumulative number of positive samples.
-As a sensitivity analysis, we lagged the test results once by one test date (\ie from Tuesday to last Thursday and from Thursday to last Tuesday) and once by one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday), to consider the unknown delay when students were infected and tested positive.
-\medskip
-
-\noindent\textbf{Aerosol and particle concentrations} \smallskip
-
-\noindent We computed the average particle concentrations per day and compared them between study conditions.
-We estimated the reduction in particle concentrations using Bayesian log-linear regression models (\supp~text~\zref{sec:env-regression-model}), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
-\medskip
-
-\noindent\textbf{Number of detected coughs} \smallskip
-
-\noindent We computed the daily number of detected coughs and compared it between study conditions.
-We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model (\supp~text~\zref{sec:aud-regression-model}) with time in class as the model offset, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
-In addition, we used a Bayesian Negative Binomial hierarchical (random effects) regression model to estimate the association between the number of coughs and the number of positive saliva test results for respiratory viruses.
-For this analysis, we only used the days were saliva testing was performed (i.e.
-every Tuesday and Thursday, 27~days in total).
-\medskip
-
-\noindent\textbf{Software and model estimation} \smallskip
-
-\noindent All analyses were performed in R software (version 4.2.0)\cite{RCoreTeam2022} and modeling in Stan (version 2.21.0)\cite{Carpenter2017}.
-Model parameters were estimated with a Bayesian approach.
-Specifically, Markov chain Monte Carlo (MCMC) sampling was used as implemented by the Hamiltonian Monte Carlo algorithm with the No-U-Turn Sampler (NUTS)\cite{Hoffman2014}.
-Each model is estimated with 4 Markov chains and 2,000 iterations of which the first 1,000 iterations are discarded as part of the warm-up.
-Estimation power is evaluated via the effective sample size (ESS) and convergence of the Markov chains is evaluated with the Gelman-Rubin convergence diagnostic ($\hat{R}$).
-If not stated otherwise, we report posterior means and credible intervals (CrIs) based on the 50\%, 80\%, and 95\% quantiles of the posterior samples, respectively.
-The code is available from \url{https://osf.io/38j9g}.
-
-\subsection{Ethics statement}
-
-\noindent The Ethics Committee of the canton of Bern, Switzerland, approved the study (reference no.
-2021-02377).
-For the saliva samples, we included all students who were willing to participate and obtained written informed consent from their caregivers.
-
-\newpage
-
-\section{Results}
-
-The study population consisted of 38~students (19~female, 19~male; Table~\ref{tab:cases-overview-school}).
-Of these, 7~students were vaccinated or recovered from an infection within SARS-CoV-2 in the last four months.
-During the seven-week study period (total of 1,330 student-days), students were absent from school for 220~days (18\% of the total) of which 129~days (59\% of absences) were due to illness.
-
-\begin{table}[!htpb]
-    \centering
-    \caption{Overview of the study population and person-days of absences.}
-    \label{tab:cases-overview-school}
-    \footnotesize
-    \begin{tabular}{l r r r}
-    \toprule
-         &  Class A & Class B & Total \\ \midrule 
-        \textbf{Students} & \textbf{20 (53\%)} & \textbf{18 (47\%)} & \textbf{38 (100\%)} \\
-        \emph{Gender} \\
-        $\drsh$ Female & 11 (55\%) & 8 (44\%) & 19 (\hphantom{0}50\%) \\
-        $\drsh$ Male & 9 (45\%) & 10 (56\%) & 19 (\hphantom{0}50\%) \\
-        \emph{Immunity status} \\
-        $\drsh$ Recently vaccinated (or recovered) & 7 (39\%) & 0 (0\%) & 7 (\hphantom{0}18\%) \\
-        $\drsh$ Not recently vaccinated (or recovered) & 11 (61\%) & 20 (100\%) & 31 (\hphantom{0}82\%) \\
-        \textbf{Absent person-days} & \textbf{110 (50\%)} & \textbf{110 (50\%)} & \textbf{220 (100\%)} \\
-        $\drsh$ Sickness & 52 (47\%) & 77 (70\%) & 129 (\hphantom{0}59\%) \\
-        $\drsh$ Other & 58 (53\%) & 33 (30\%) & 91 (\hphantom{0}41\%) \\
-        \bottomrule
-    \end{tabular} 
-\end{table}
-
-\subsection{Molecular analyses of saliva and air samples}
-
-We analyzed 448~saliva samples over the study across classes.
-We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenovirus (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~virus~(PIV) (\cref{fig:molecular-descriptives}a).
-There was an equal number of positive samples with and without air cleaners (\cref{fig:molecular-descriptives}b).
-Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
-The distribution of positive samples varied between classes (\cref{fig:molecular-descriptives}c).
-For example, during the first three study weeks, all but one sample was positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
-We also analyzed 105~bioaerosol samples and detected  two respiratory viruses in the air (1~HRV in class~A and 1~AdV in class~B).
-Finally, we detected 1~Flu~B, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{../../results/mol-data/descriptives.pdf}
-    \caption{\textbf{Analysis of molecular data}.
-Number of positive saliva samples by \textbf{(a)}~virus, \textbf{(b)}~study condition, and \textbf{(c)}~study week and class.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
-    \label{fig:molecular-descriptives}
-\end{figure}
-
-\subsection{Analyses of transmission networks and molecular genotyping}
-
-To illustrate possible transmission chains, we linked positive saliva samples within classes that belong to the same virus and were less than 1~week apart.
-Based on this spatiotemporal analysis of student saliva samples, we identified10~possible transmission chains, which are shown in \cref{fig:molecular-network}.
-The longest transmission chains possibly occurred in January and refer to the cluster of AdV infections in class~A and Flu~B infections in class~B.
-We also performed molecular genotyping of saliva, bioaerosol, and air filter samples to verify our proposed transmission networks.
-However, molecular genotyping was not successful because we were unable to amplify and sequence any of the gene targets.Nevertheless, the virus-specific variation between classes and the temporal distribution of positive samples suggest that transmission occurred within classes.
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{../../results/mol-data/network-plot.pdf}
-    \caption{\textbf{Transmission network of respiratory viruses: spatiotemporal analysis of students' saliva samples}.
-Daily number of positive saliva samples (colored circles) and possible transmission chains within classes (directed arrows).
-Positive samples are linked if they belong to the same virus and are less than 1~week apart.
-Positive samples from the air and filters as blank squares aligned.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
-    \label{fig:molecular-network}
-\end{figure}
-
-\subsection{Analysis of aerosol and particle concentrations}
-
-There was a strong difference in particle concentrations between study conditions (\cref{fig:palas-results}a).
-Aerosol number concentration with air cleaners were markedly lower (27$\pm$34\,1/cm$^3$) than without air cleaners (95$\pm$81\,1/cm$^3$).
-Similarly, particle mass concentrations (\eg PM$_1$) were markedly lower with air cleaners (1.16$\pm$1.33\,$\mu$gm$^{-3}$) than without air cleaners (4.20$\pm$3.74\,$\mu$gm$^{-3}$).
-There was no considerable difference in other environmental variables (\supp~figure~\zref{fig:env-descriptives-other-vars}).
-In particular, daily average CO$_2$ levels (air change rates) were 1,769$\pm$391\,ppm (0.74$\pm$0.41\,h$^{-1}$) with air cleaners and 1,636$\pm$341\,ppm (0.68$\pm$0.32\,h$^{-1}$), suggesting that ventilation was comparable between study conditions.
-When adjusting for air change rates and multiple other factors, the aerosol number concentration decreased by 76\% (95\%-CrI 63\% to 86\%) with air cleaners (\cref{fig:palas-results}b and \supp~table~\zref{tab:palas-est-results}).
-The decrease in the concentration of larger particles (PM$_{10}$) was greater than the decrease in the concentration of smaller particles (PM$_1$ to PM$_{4}$).
-
-\begin{figure}[!htpb]
-\centering
-    \includegraphics[width=\linewidth]{../../results/env-data/particles-boxplot.pdf}
-    \includegraphics[width=\linewidth]{../../results/env-data/estimation-results-figure.pdf}
-    \caption{\textbf{Analysis of particle concentrations and comparison between study conditions}.
-\textbf{(a)}~Boxplot of the daily average values for aerosol number concentration (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$).
-\textbf{(b)}~Estimated reduction in aerosol number and particle mass concentrations with air cleaners (posterior mean as dot and 50\%-, 80\%- and 95\%-CrI as lines, respectively).
-}
-    \label{fig:palas-results}
-\end{figure}
-
-\subsection{Risk of infection}
-
-Overall, there were 35~respiratory cases (absences related to respiratory infections) across classes.
-Incidentally, there were more cases in class~B (21) than in class~A (14), which is incidentally also the class where more students recently recovered from a SARS-CoV-2 infection or were vaccinated.
-Except for one study week, there was an equal or higher number of cases without air cleaners (\cref{fig:redcap-results}a).
-Based on our Bayesian latent hierarchical regression model, we estimated a trend towards a lower risk of infection with air cleaners (adjusted relative risk 0.72, 95\%-Credible Interval 0.44 to 1.15).
-As a result, the estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\cref{fig:redcap-results}b).
-The posterior probability that air cleaners reduced the risk of infection was 90.8\%, also suggesting a beneficial effect of air cleaners.
-Detailed estimations are provided in the \supp~text~\zref{sec:detailed-redcap}.
-
-\begin{figure}[!htpb]
-    \includegraphics{../../results/epi-data/cases_by_condition.pdf} 
-    \includegraphics{../../results/epi-data/avoided-infections.pdf}
-    \caption{\textbf{Analysis of epidemiological data and estimated number of infections with and without air cleaners.}.
-\textbf{(a)}~Weekly number of respiratory cases with and without air cleaners.
-\textbf{(b)}~Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
-    \label{fig:redcap-results}
-\end{figure}
-
-\subsection{Detected coughs}
-
-Slightly fewer coughs were detected in audio recordings during study weeks where air cleaners were installed (\cref{fig:coughing}a).
-On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners ($\Delta$ $-$0.6~coughs/min; adjusted relative risk 0.93, 95\%-CrI 0.85 to 1.02).
-Significantly more coughs were detected in class~B (adjusted relative risk 2.16, 95\%-CrI 1.90 to 2.46), suggesting a link with the observed variation in virus-specific transmission between classes (\cref{fig:molecular-descriptives}c).
-As suspected, the frequency of coughing was associated with the number of positive saliva samples by respiratory virus (\cref{fig:coughing}b).
-Coughing was more frequent during Flu~B positivity (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent during AdV positivity (0.87, 95\%-CrI 0.49 to 1.12).
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{results/cough-data/coughs-frequency-by-condition.pdf}
-    \includegraphics{results/cough-data/coughs-virus-association.pdf}
-    \caption{\textbf{Analysis of coughing data.} \textbf{(a)}~Daily average number of detected coughs per minute by study week, condition and class.
-\textbf{(b)}~Estimated association (relative risk; posterior mean as dot, 50\%, 80\%, and 95\%-CrI as lines) between cough frequency and number of positive saliva test results by respiratory virus.}
-    \label{fig:coughing}
-\end{figure}
-
-\subsection{Comparison with previous study}
-
-In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 Omicron wave.
-\cref{fig:comparison} shows a comparison of the results.
-The distribution of respiratory infections is different (\cref{fig:comparison}a) because we almost exclusively detected SARS-CoV-2 in our previous study during the COVID-19 pandemic, as compared to a range of respiratory infections in this study during non-pandemic times.
-Although there were more positive saliva samples in this study, the number of respiratory infections detected in bioaerosols and on the HEPA-filters of the air cleaners was lower (\cref{fig:comparison}b).
-The reductions in particle concentrations were larger and with lower uncertainty in this study (\cref{fig:comparison}c).
-The risk of infection tended to be lower with air cleaners in this study where air cleaners were analyzed using a cross-over design, but not in our previous study where air cleaners were installed only at the end of the study when most students had already been infected with SARS-Cov-2(\cref{fig:comparison}d).
-
-\begin{figure}[!htpb]
-    \includegraphics[width=\linewidth]{../../results/comparison.pdf} 
-    \caption{\textbf{Comparison of the results between our previous study in 2022 during the COVID-19 pandemic (blue) and this study in 2023 during non-pandemic times (black).} \textbf{(a)}~Proportion of positive saliva samples by virus.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.
-\textbf{(c)}~Estimated reduction in aerosol number and particle mass concentrations (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$) with air cleaners (posterior mean as squares and 95\%-CrI as lines).
-\textbf{(d)}~Adjusted relative risk (ARR) of infection with air cleaners (posterior mean as squares and 95\%-CrI as lines).}
-    \label{fig:comparison}
-\end{figure}
-
-\FloatBarrier
-
-\newpage
-
 \section{Discussion}
 
-% summary
-
-We collected epidemiological, environmental, and molecular data to estimate the risk of respiratory virus infection in a Swiss school and assessed the effectiveness of air cleaners as an infection control measure.
-We found a wide range of respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus, but very few respiratory viruses were detected in bioaerosol samples and on the filters of the air cleaners.
-Aerosol number and particle mass concentrations decreased significantly with air cleaners and Bayesian modeling based on the absences of students due to illness showed a small, potential reduction in the relative risk of infection with air cleaners.
-Reduced coughing could indicate that air cleaners prevent some symptomatic infections.
-
-% comparison with our previous study
-
-In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners in reducing the transmission of SARS-CoV-2 in two Swiss schools.
-We found a detectable effect on the risk of infection only for mask wearing, but not air cleaners.
-We argued that air cleaners may have had less potential to reduce infections because they were introduced at the end of the study, when most students were already infected with SARS-CoV-2.
-In this study, we used a cross-over design to estimate the effectiveness of air cleaners again, but this time under non-pandemic conditions.
-We found similar reductions in particle concentrations and a tendency for the relative risk of infection to be lower with air cleaners.
-To some extent, the differences between the two studies may be explained by different adjustments.
-While we could not adjust for ventilation in our previous study, the estimates in this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see \supp~text~\zref{sec:detailed-molecular}).
 Furthermore, the distribution of respiratory viruses in this study was very different from that in our previous study.
@@
 
-% transmission within classrooms
+\section{Comparison with previous study}\label{sec:comparison-prev-study}
 
-The low rate of positive bioaerosol samples also means that it is unclear to what extent airborne transmission occurred in classrooms.
-It may be possible that students had relatively little exposure to respiratory viruses in schools and acquired their infections elsewhere.
-However, the distribution of saliva samples between classes shows remarkable differences.
-For example, adenovirus spread entirely in class~A during the first three study weeks and only two infections with influenza~B occurred over the study.
-In contrast, influenza~B spread throughout the study in class~B, and adenovirus infections were  detected only in the last week of the study.
-Adenovirus infections tend to be mild\cite{Kunz2010CIDR} and less associated with cough than influenza\cite{Ma2018RMV}, which is consistent with the comparatively lower frequency of detected coughs in class~B.
-Taken together, the class-specific, spatiotemporal patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.
+% descriptive results
 
-% effects of air cleaners
+In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 Omicron wave.
+\cref{fig:comparison} shows a comparison of the results.
+The distribution of respiratory infections is different (\cref{fig:comparison}a) because we almost exclusively detected SARS-CoV-2 in our previous study during the COVID-19 pandemic, as compared to a range of respiratory infections in this study during non-pandemic times.
+Although there were more positive saliva samples in this study, the number of respiratory infections detected in bioaerosols and on the HEPA-filters of the air cleaners was lower (\cref{fig:comparison}b).
+The reductions in particle concentrations were larger and with lower uncertainty in this study (\cref{fig:comparison}c).
+The risk of infection tended to be lower with air cleaners in this study where air cleaners were analyzed using a cross-over design, but not in our previous study where air cleaners were installed only at the end of the study when most students had already been infected with SARS-Cov-2(\cref{fig:comparison}d).
 
-Our study indicated a small, potential benefit of air cleaners.
-There may be several reasons for this relatively small effect.
-Compared to \eg universal mask wearing, which has been shown to be effective\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Dharmadhikari2012AJRCCM,Leung2020NatMed,Milton2013PLoSPathogens}, air cleaners cannot affect transmission outside the classroom.
-Furthermore, previous work suggests that prolonged close contact may be necessary for transmission\cite{Leung2020NatMed,Brankston2007LancetID}, or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
-However, compared to face masks, air cleaners are not able to prevent transmission by close range, high particle density.
-Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective at removing larger particles ($>5\mu$m).
-However, a greater proportion of respiratory viruses are carried in the smaller particles, which are also more relevant for transmission ($\leq5\mu$m)\cite{Fennelly2020}.
-Finally, classroom activity, airflow and other unobserved, confounding factors make it difficult to evaluate the effects of air cleaners.
-Nevertheless, it is important to assess their effectiveness in a real-world setting and to compare them with the hypothetical impact suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
+% summary
 
-% limitations
+To summarize, we found similar reductions in particle concentrations and a tendency for the relative risk of infection to be lower with air cleaners.
+To some extent, the differences between the two studies may be explained by different adjustments.
+While we could not adjust for ventilation in our previous study, the estimates in this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see \Cref{sec:detailed-molecular}).
+Furthermore, the distribution of respiratory viruses in this study was very different from that in our previous study and molecular airborne detection was lower.
 
-Our study has several limitations.
-First, aerosol measurements and molecular detection of viruses in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (person to air, air to person) cannot be determined.
-However, the virus-specific distribution of positive saliva samples between classes (spatiotemporal pattern) may indicate transmission within classrooms.
-Second, the specific causes for the absences could not be determined from the epidemiological data.
-Therefore, some absences may have falsely been attributed to respiratory infections and we could only approximate the incubation period for each epidemiological case.
-Third, the effectiveness of air cleaners was somewhat limited because students occasionally had lessons outside the classroom.
-This is in contrast to other infection control measures such as face masks, which had to be used in every indoor space during the pandemic.
+\begin{figure}[!htpb]
+    \includegraphics[width=\linewidth]{../../results/comparison.pdf} 
+    \caption{\textbf{Comparison of the results between our previous study in 2022 during the COVID-19 pandemic (blue) and this study in 2023 during non-pandemic times (black).} \textbf{(a)}~Proportion of positive saliva samples by virus.
+Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.
+\textbf{(c)}~Estimated reduction in aerosol number and particle mass concentrations (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$) with air cleaners (posterior mean as squares and 95\%-CrI as lines).
+\textbf{(d)}~Adjusted relative risk (ARR) of infection with air cleaners (posterior mean as squares and 95\%-CrI as lines).}
+    \label{fig:comparison}
+\end{figure}
 
-% conclusion
+% unlikely reasons for less airborne detection
 
-In conclusion, using epidemiological, environmental, and molecular data, our study showed that a wide range of non-SARS-CoV-2, but rarely SARS-CoV-2, was detected in students under non-pandemic conditions when public health measures were lifted, and that transmission of these respiratory viruses likely occurred in schools.
-Airborne detection of non-SARS-CoV-2 respiratory viruses was rare, suggesting that prolonged close contact may be required for transmission.
-Comparison with previous work during the pandemic, which showed frequent airborne SARS-CoV-2, suggests that respiratory viruses other than SARS-CoV-2 may be less detectable in the air.
-Air cleaners improved air quality by significantly reducing aerosol and particle concentrations.
-Based on our transmission model using epidemiological data, the risk of infection tends to be lower with air filters.
-Future studies should investigate the relative importance of prolonged close contact and long range transmission of respiratory viruses in schools and other congregate settings.
+Technical reasons for the differences in airborne detection are unlikely.
+The bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
+Instead, airborne detection could have been influenced by environmental factors.
+Although temperature (overall between 19 and 25$^{\circ}$C) and relative humidity (overall between 25 and 50\%) were similar in both studies, the same levels may have different effects on the airborne survival of respiratory viruses.
+For example, airborne survival of rhinovirus and adenovirus increases with relative humidity, whereas the opposite relationship has been documented for influenza and SARS-CoV-2\cite{Tellier2009JTRSI,Ahlawat2020AAQR,Biryukov2020mS,Karim1985CJM,Davis1971AM}.
+A change in ventilation conditions could also have had a profound effect, but CO$_2$ levels were higher compared to the previous study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
+Although lower ventilation was partially offset by a greater reduction in aerosol and particle concentrations from air cleaners, the removal of particles by air cleaners may be slower than by natural ventilation.
 
-\newpage
+% likely reason for less airborne detection
 
-%TC:ignore
-\section*{Acknowledgements}
-We would like to thank the schools, teachers, and students participating in the study.
-We are grateful to the Educational Department of the canton of Solothurn for their support throughout the study.
-We also would like to thank the biosecurity team at the Institute for Infectious Diseases of the University of Bern for their assistance with the bioaerosol devices (Julia Feldmann, Monika Gsell, Kathrina Summermatter).
-Finally, we are indebted to the student assistants (Khadija Ahmed, Marie-Joséphine Brancato, Michelle Bürki, Santiago Martinez, Moric Toszeghi, Sylvain Wasmer) who helped with the data collection in the schools.
-
-\bibliography{references.bib}
-%TC:endignore
-
-\end{document}
+More likely, SARS-CoV-2 may be easier to detect in the air due to physicochemical properties of virus-laden aerosols, so that its viral loads more often exceed the detection limits of existing devices for bioaerosol sampling,  which remains a challenge\cite{Wang2021,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
+For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.
+However, several studies showed that influenza\cite{Bischoff2013JID,Pan2017mSphere}, rhinovirus\cite{Myatt2004AJRCCM}, and adenovirus\cite{Nguyen2017OFID,Pan2017mSphere} can also be detected in airborne samples.

### 2023-10-16 doc/paper/supplements.tex
Updates from Overleaf

@@
 \author[1]{Matthias Egger}
-\author[2,5]{Tina Hascher}
+\author[5]{Tina Hascher}
 \author[1*,2]{Lukas Fenner}
@@
 
-\clearpage
-
 \section{Detailed results for risk of infection model}\label{sec:detailed-redcap}
@@
 
+Our estimation results imply a small, potential benefit of air cleaners.
+Subject to considerable uncertainty, \Cref{fig:avoided-infections} shows the estimated number number of infections that would have occurred if air cleaners had been installed throughout the study (scenario~1) or not been installed (scenario~2).
+
+\begin{figure}[!htpb]
+    \includegraphics{../../results/epi-data/avoided-infections.pdf}
+    \caption[Estimated number of infections with and without air cleaners]{Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
+    \label{fig:avoided-infections}
+\end{figure}
+
 \clearpage
@@
 
-In the main paper, we compared particle concentrations between study conditions.
-\Cref{fig:env-descriptives-other-vars} compares additional environmental variables, \ie CO$_2$, air change rate, relative humidity, and temperature.
-There are no considerable differences between in any of the variables between study conditions.
-Furthermore, numerical estimation results for the reduction in particle concentrations with air cleaners are shown in \Cref{tab:palas-est-results}.
+There was a strong difference in particle concentrations between study conditions (\Cref{fig:palas-results}a).
+When adjusting for air change rates and multiple other factors, the aerosol number concentration decreased by 76\% (95\%-CrI 63\% to 86\%) with air cleaners (\Cref{fig:palas-results}b and \Cref{tab:palas-est-results}).
+The decrease in the concentration of larger particles (PM$_{10}$) was greater than the decrease in the concentration of smaller particles (PM$_1$ to PM$_{4}$).
 
-\begin{figure}[!htb]
+\begin{figure}[!htpb]
 \centering
-    \includegraphics[width=\linewidth]{../../results/env-data/otherVars-boxplot.pdf}
-    \caption[Boxplot of environmental variables by school and study condition]{Boxplot for the daily average values of each environmental variable by school and study condition.}
-    \label{fig:env-descriptives-other-vars}
+    \includegraphics[width=\linewidth]{../../results/env-data/particles-boxplot.pdf}
+    \includegraphics[width=\linewidth]{../../results/env-data/estimation-results-figure.pdf}
+    \caption[Analysis of particle concentrations and comparison between study conditions]{\textbf{Analysis of particle concentrations and comparison between study conditions}.
+\textbf{(a)}~Boxplot of the daily average values for aerosol number concentration (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$).
+\textbf{(b)}~Estimated reduction in aerosol number and particle mass concentrations with air cleaners (posterior mean as dot and 50\%-, 80\%- and 95\%-CrI as lines, respectively).
+}
+    \label{fig:palas-results}
 \end{figure}
@@
     \end{tabular}
+\end{table}
 
-\end{table}
+In the main paper, we compared particle concentrations between study conditions.
+\Cref{fig:env-descriptives-other-vars} compares additional environmental variables, \ie CO$_2$, air change rate, relative humidity, and temperature.
+There are no considerable differences between in any of the variables between study conditions.
+Furthermore, numerical estimation results for the reduction in particle concentrations with air cleaners are shown in \Cref{tab:palas-est-results}.
+
+\begin{figure}[!htb]
+\centering
+    \includegraphics[width=\linewidth]{../../results/env-data/otherVars-boxplot.pdf}
+    \caption[Boxplot of environmental variables by study condition]{Boxplot for the daily average values of each environmental variable by study condition.}
+    \label{fig:env-descriptives-other-vars}
+\end{figure}
+
+\clearpage
+
+\section{Detailed results for cough frequency}\label{sec:detailed-results-coughing}
+
+Differences in the frequency of coughs between classes suggest an association with class-specific virus patterns.
+\Cref{fig:coughing-association} shows the relative risk of cough for each respiratory virus.
+Cough appeared more frequent for Flu~B and less frequent for AdV.
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/cough-data/coughs-virus-association.pdf}
+    \caption[Association between cough frequency and virus-specific positivity rate]{Estimated association (relative risk; posterior mean as dot, 50\%, 80\%, and 95\%-CrI as lines) between cough frequency and number of positive saliva test results by respiratory virus.
+Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
+    \label{fig:coughing-association}
+\end{figure}
 

### 2023-10-16 doc/paper/manuscript-lancet.tex

@@
 
-xxxx
-
-% \noindent\textbf{Running head}: SARS-COV-2 transmission in schools and effect of air cleaners
-
-% %\noindent\textbf{Subject categorization}: 6.20  Indoor Air; 10.11 Pediatrics: Respiratory Infections
-
-% \noindent\textbf{Word count}: \maincount{manuscript}words, abstract \abstractcount{manuscript}words (max. 500), title 157 chars (max. 200)
-
-% %\noindent\textbf{Inserts:} 2 tables, 6 figures, 44 references
+\noindent \textcolor{red}{[list of appendices to be added here later; including STROBE]}
 
 % \noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.
-
-% \vspace{1em}
-
-% \noindent\textbf{Funding}
-
-% \noindent This study is funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
-NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
-ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498).
-\medskip
-
-% \noindent\textbf{Contributions}
-
-% \noindent Conception and design: NB, LF.
-Epidemiological and environmental data collection: NB, PJ, TS, LF.
-Laboratory data collection: PB, LFu.
-Additional data collection: TH.
-Statistical analysis: NB, KZ.
-Genomic analysis: LB, LFu.
-Paper draft: NB, LF, ME.
-All authors reviewed and approved the final version of the manuscript.
 
@@
 \noindent\textbf{Background:} Infection control measures should reduce airborne transmission.
-We aimed to evaluate the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory viral infections in schools using a multiple-measurement approach.
+We aimed to evaluate the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing transmission of respiratory viral infections in schools using a multiple-measurement approach.
 \medskip
 
-\noindent\textbf{Methods:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), audio (recordings detecting coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2023 in two classes of a Swiss secondary school.
+\noindent\textbf{Methods:} We collected epidemiological data (absences related to respiratory infections), environmental data (CO$_2$ and particle concentrations), audio recordings (detecting coughs), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2023 in two classes of a Swiss secondary school.
 Using a cross-over study design, we compared particle concentrations, coughing, and the risk of infection with and without air cleaners.
@@
 \noindent\textbf{Results:} Molecular analysis detected multiple viruses in saliva (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza), but rarely in bioaerosol samples (2/105 positive) and on the filters of air cleaners (4/160~positive).
-Nevertheless, spatiotemporal analysis of saliva samples suggested transmission within classrooms.
+Spatiotemporal analysis of positive saliva samples suggested within-classroom transmission.
 Daily average CO$_2$ levels with and without air cleaners were similar (1,769$\pm$391\,ppm vs.
 1,636$\pm$341\,ppm).
-Particle concentrations decreased by 77\% (95\%-credible interval [CrI] 63\% to 86\%), coughing was less frequent ($-$0.5~coughs/min), and the estimated risk of infection likely lower with air cleaners (adjusted relative risk 0.72, 95\%-CrI 0.44 to 1.15).
+Particle concentrations decreased by 77\% (95\%-credible interval [CrI] 63\% to 86\%), coughs were less frequent ($-$0.5~coughs/min), and the estimated risk of infection likely lower with air cleaners (adjusted relative risk 0.72, 95\%-CrI 0.44 to 1.15).
 \medskip
 
-\noindent\textbf{Conclusions:} A wide range of respiratory viruses were detected in students' saliva.
+\noindent\textbf{Conclusions:} A wide range of non-SARS-CoV-2 respiratory viruses were detected in student saliva.
 Although airborne detection was rare, transmission seemed likely in schools, suggesting that prolonged close contact is required for transmission.
@@
 \maketitle
-\setcounter{page}{1}
-\thispagestyle{fancy}
 
@@
 % maximum of 3-5 keywords
-\newpage
 
+\thispagestyle{empty}
 \sloppy
 \raggedbottom
-%TC:endignore
 
@@
 
+\noindent\textbf{\Large{Research in context}} \medskip
+
+\noindent \textbf{Evidence before this study} \smallskip
+
+\noindent \textcolor{red}{[Search and search terms to be added]} Several studies have shown that air cleaners reduce particle concentrations in indoor air.
+Another study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward.
+A simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission.
+A real-world study during the pandemic showed a trend toward a reduction of the infection risk.
+\medskip
+
+\noindent \textbf{Added value of this study} \smallskip
+
+\noindent We showed the transmission dynamics in classrooms of a secondary school under non-pandemic conditions.
+Although airborne detection was rare, transmission seemed likely in schools, suggesting that prolonged close contact is required for transmission.
+Air cleaners improved air quality, were associated with lower symptom intensity, and showed a small, potential benefit in reducing infections.\medskip
+
+\noindent \textbf{Implications of all the available evidence} \smallskip
+
+\noindent Good ventilation improves indoor air quality and can reduce the risk of infection under non-pandemic conditions.
+Greater emphasis should be given to ventilation systems in schools in the long term.
+Seasonal respiratory viruses other than SARS-CoV-2 may require prolonged close contact.
+Future studies should examine the relative importance of prolonged close contact and long-distance transmission of respiratory viruses in schools and other congregate settings.
+
+\thispagestyle{empty}
+\sloppy
+\raggedbottom
+
+\newpage
+%TC:endignore
+
+\setcounter{page}{1}
+
 \section{Introduction}
 
-The transmission of respiratory infections such as SARS-CoV-2 and influenza is difficult to mitigate and control.
-Person-to-person transmission occurs primarily through the release of respiratory particles containing the respective viruses.
-Recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
+Transmission of respiratory infections such as SARS-CoV-2 and influenza are difficult to mitigate and control.
+Person-to-person transmission occurs primarily through the release of respiratory particles containing the viruses.
+Recent focus has been on small respiratory particles called aerosols ($\leq$ 5$\mu$m), which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
 Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and travel long distances\cite{Coleman2022,Wang2020,Heneghan2021}.
@@
 
-We used a multiple-measurement approach to study the transmission of viral respiratory viruses and the effect of air cleaners in school rooms.
-We employed a cross-over study design under non-pandemic conditions in the winter of 2023.
-We collected epidemiological (absences likely related to respiratory infections), environmental (CO$_2$, aerosol and particle concentrations), audio (recordings detecting coughing), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
+We used a multiple-measurement approach to study the transmission of respiratory viruses and the effect of air cleaners in schools.
+We used a cross-over study design under non-pandemic conditions in the winter of 2023.
+We collected epidemiological data (absences likely related to respiratory infections), environmental data (CO$_2$, aerosol and particle concentrations), audio recordings (detecting coughs), and molecular data (detection of viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
 We determined changes in particle concentrations, respiratory virus positivity rates, and modeled the risk of infection.
@@
 
-\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from 26~January to 11~March 2023.
-\cref{fig:study-setup} shows the schematic study setup and \supp~table~\zref{tab:data} provides an overview of the types of data collected in each classroom.
+\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, during seven weeks from 26~January to 11~March 2023.
+Figure~\ref{fig:study-setup} shows the schematic study setup and \supp~table~\zref{tab:data} provides an overview of the types of data collected.
 
@@
 
-\noindent We employed a cross-over design to study the effectiveness of air cleaners, distinguishing two study conditions in each class: weeks with and weeks without air cleaners being installed in each classroom (\cref{tab:study_design}).
-Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China; approx.
-USD\,250 per device).
-According to the manufacturer, these air cleaners run at a clean air delivery rate of 2$\,\times\,$600 m$^{3}$/h.
+\noindent We used a cross-over design to study the effectiveness of air cleaners, with two study conditions: weeks with and weeks without air cleaners being installed in each classroom (\cref{tab:study_design}).
+Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China).
+According to the manufacturer, these air cleaners run at clean air delivery rates of 2$\,\times\,$600 m$^{3}$/h.
 % We measured a lower clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h when testing the devices in an empty classroom.
-Window ventilation was done at the discretion of the teachers, as there were no specific recommendations from the national health authorities at that time.
+Window ventilation was done at the discretion of the teachers, as there were no specific recommendations from the national health authorities at the time.
 
@@
 
-\noindent At the beginning of the study, we collected aggregated data on age, sex and COVID-19 vaccination and recovery status in the participating classes.
-On a daily basis, we collected data on each absent student, \ie the absence and return dates as well as the reason for the absence (illness, career days, or other).
-For absences due to illness, we collected symptoms and the date of symptom onset.
-Based on the epidemiological data, we defined a case of respiratory infection as an absence in which the student reported an illness with at least one respiratory symptoms (\supp~text~\zref{sec:case-data}).
+\noindent At the beginning of the study, we collected aggregated data on age, sex, and COVID-19 vaccination and recovery status in the participating classes.
+On a daily basis, we collected data on each absent student, \ie the absence and return dates and reason for the absence (illness, career days, or other).
+For absences due to illness, we recorded symptoms and the date of symptom onset.
+We defined a case of respiratory infection as an absence in which the student reported an illness with at least one respiratory symptom (\supp~text~\zref{sec:case-data}).
 The line list data is shown in \supp~table~\zref{tab:epi-data-line-list}.
@@
 
-\noindent \emph{Saliva sampling} \smallskip
+\noindent \emph{Saliva} \smallskip
 
-\noindent Both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below).
+\noindent Both classes participated in repetitive bi-weekly testing (Tuesdays and Thursdays; see \nameref{sec:mol_analyses} below).
 Saliva samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Galar2021,To2019,Huber2021}.
@@
 
-\noindent \emph{Bioaerosol sampling and HEPA filters of the air cleaners} \smallskip
+\noindent \emph{Bioaerosol and HEPA filters of the air cleaners} \smallskip
 
-\noindent We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL PBS, as previously reported in clinic settings\cite{Moore2021}.
+\noindent We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL PBS.
 The Coriolis Micro Air could only operate shortly before and during break times to prevent unacceptable noise exposure (approximately 60\,min/day).
 In one class, we also used the BioSpot-VIVAS device sampler (Aerosol Devices Inc., Ft.
-Collins, CO, USA).
-This device samples airborne virus particles into a viral transport medium (VTM) using a water-based laminar flow condensation method\cite{Pan2016JAM,Lednicky2016AST}.
-BioSpot-VIVAS could operate throughout lessons.
-Removable parts of both sampling devices were regularly autoclaved to prevent contamination.
+Collins, CO, USA) \cite{Pan2016JAM,Lednicky2016AST}.
+BioSpot-VIVAS operated throughout the lessons.
+Removable parts of both sampling devices were regularly autoclaved.
 At the end of the day, samples were transported to the Institute of Infectious Diseases (IFIK) and stored at $-$80°C.
@@
 \noindent Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL.
-The human saliva samples were directly analyzed without prior filtration.
+The human saliva samples were analyzed directly without prior filtration.
 The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19~respiratory viruses, including SARS-CoV-2 (CoV), influenza (Flu), respiratory syncytial (RSV), metapneumovirus (MPV), adenovirus (AdV), rhinovirus (HRV), and parainfluenza (PIV).
@@
 For the BioSpot-VIVAS air samples, the air concentration of the DNA was exactly calculated according to the airflow influx into the sampling device.
-%/
 The Limit of detection (LOD) of the assay SARS-CoV-2, IFA/IFB is 7.5x10$^0$ TCID/ml (SARS-CoV-2), 3.7x10$^{-1}$ TCID/ml (IFA), and 4.8x100 TCID/ml (100cp/ml) (IFB) respectively.
@@
 
-We performed molecular genotyping for positive samples of AdV and Flu~B.
-For AdV (saliva, bioaerosol, and air filter samples), we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}.
+We performed molecular genotyping for positive samples for AdV and Flu~B.
+(saliva, bioaerosol, air filter samples).
+For AdV, we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}.
 For Flu~B, we tried to determine the molecular transmission network using whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
 
-\subsection{Audio recordings and cough detection}
+\subsection{Cough detection}
 
-We installed one portable audio recorder (ZOOM H6; New York, USA) in each classroom to record sounds continuously.
-We detected the number of coughs per minute using an AI algorithm\cite{Bertschinger2023IEEE}.
+We installed a portable audio recorder (ZOOM H6; New York, USA) in each classroom to record sounds continuously.
+We determined the number of coughs per minute using an AI algorithm\cite{Bertschinger2023IEEE}.
 
@@
 
-All statistical analyses were described in the statistical analysis plan (SAP)\cite{Banholzer2023SAP}, which was published ahead of analyses.
-In contrast to the SAP, the number of positive bioaerosol samples and the viral load concentration of positive samples could not be analyzed because there were too few positive samples during the study.
-Apart from that there were no major deviations from the SAP regarding the conducted analyses.
-There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in \supp~text~\zref{sec:transmission-model}-\zref{sec:env-regression-model}, which provide detailed descriptions of the models used in each analysis.
+All statistical analyses were described in the statistical analysis plan (SAP)\cite{Banholzer2023SAP}, which was published prior to the analyses.
+Contrary to the SAP, bioaerosol samples and viral load concentrations could not be analyzed because there were too few positive samples.
+Otherwise, there were no major deviations from the SAP.
+Minor deviations are documented in \supp~text~\zref{sec:transmission-model}-\zref{sec:env-regression-model}, including detailed descriptions of the models.
 \medskip
@@
 
-\noindent We estimated the relative risk of infection with air cleaners using a Bayesian latent hierarchical regression model.
+\noindent We estimated the relative risk of infection with air cleaners using a Bayesian latent variable hierarchical regression model.
 The observed outcome of this model is the number of new respiratory cases $C$ (absences related to respiratory infections by date of symptom onset) on day $t$ in class $j$, which are modeled with a Negative Binomial distribution.
@@
 Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the community (\ie the canton of Solothurn).
-
 A detailed description of the model and choice of priors for all model parameters are provided in \supp~text~\zref{sec:transmission-model}.
-Importantly, since multiple viruses were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the virus-specific distributions.
+%Importantly, since multiple viruses were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the virus-specific distributions.
 The weights were inferred based on the proportion of positive saliva samples for each virus per week.
 As a result, the distribution of the incubation period used for modeling could change every week depending on the proportion of viruses that were detected in the students' saliva.
-Moreover, respiratory cases by date of symptom onset were under-reported for weekends and thus a re-weighting scheme was used to account for weekday effects.
+Moreover, a re-weighting scheme was used to account for weekday effects.
 \medskip
@@
 
-\noindent Molecular analysis determined which saliva samples were positive for a respiratory virus.
-We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~text~\zref{sec:multinomial-model}) where the expected number of positive samples for a virus was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each virus.
-The number of susceptibles was computed as the difference between the number of students per class and the cumulative number of positive samples.
+\noindent We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~text~\zref{sec:multinomial-model}) and linked the expected number of positive samples to the presence of air cleaners.
+\medskip % The number of susceptibles was computed as the difference between the number of students per class and the cumulative number of positive samples.
 As a sensitivity analysis, we lagged the test results once by one test date (\ie from Tuesday to last Thursday and from Thursday to last Tuesday) and once by one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday), to consider the unknown delay when students were infected and tested positive.
-\medskip
 
@@
 \noindent We computed the average particle concentrations per day and compared them between study conditions.
-We estimated the reduction in particle concentrations using Bayesian log-linear regression models (\supp~text~\zref{sec:env-regression-model}), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+We estimated the reduction in particle concentrations using Bayesian log-linear regression models (\supp~text~\zref{sec:env-regression-model}).
+\medskip %, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+
+\noindent\textbf{Number of coughs} \smallskip
+
+\noindent We computed the daily number of detected coughs and compared it between study conditions.
+We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model (\supp~text~\zref{sec:aud-regression-model}).
+%with time in class as the model offset, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+In addition, we used a Bayesian Negative Binomial hierarchical (random effects) regression model to estimate the association between the number of coughs and the number of positive saliva test results.
+For this analysis, we used only days on which saliva samples were collected (Tuesdays and Thursdays, for a total of 27~days).
 \medskip
 
-\noindent\textbf{Number of detected coughs} \smallskip
-
-\noindent We computed the daily number of detected coughs and compared it between study conditions.
-We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model (\supp~text~\zref{sec:aud-regression-model}) with time in class as the model offset, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
-In addition, we used a Bayesian Negative Binomial hierarchical (random effects) regression model to estimate the association between the number of coughs and the number of positive saliva test results for respiratory viruses.
-For this analysis, we only used the days were saliva testing was performed (i.e.
-every Tuesday and Thursday, 27~days in total).
-\medskip
-
-\noindent\textbf{Software and model estimation} \smallskip
+\noindent\textbf{Software} \smallskip
 
 \noindent All analyses were performed in R software (version 4.2.0)\cite{RCoreTeam2022} and modeling in Stan (version 2.21.0)\cite{Carpenter2017}.
-Model parameters were estimated with a Bayesian approach.
-Specifically, Markov chain Monte Carlo (MCMC) sampling was used as implemented by the Hamiltonian Monte Carlo algorithm with the No-U-Turn Sampler (NUTS)\cite{Hoffman2014}.
-Each model is estimated with 4 Markov chains and 2,000 iterations of which the first 1,000 iterations are discarded as part of the warm-up.
+Model parameters were estimated with a Bayesian approach, using the Hamiltonian Monte Carlo algorithm with the No-U-Turn Sampler (NUTS)\cite{Hoffman2014}.
+%Each model is estimated with 4 Markov chains and 2,000 iterations of which the first 1,000 iterations are discarded as part of the warm-up.
 Estimation power is evaluated via the effective sample size (ESS) and convergence of the Markov chains is evaluated with the Gelman-Rubin convergence diagnostic ($\hat{R}$).
@@
 The study population consisted of 38~students (19~female, 19~male; Table~\ref{tab:cases-overview-school}).
-Of these, 7~students were vaccinated or recovered from an infection within SARS-CoV-2 in the last four months.
+Seven~students had been vaccinated or recovered from a SARS-CoV-2 infection within the last four months.
 During the seven-week study period (total of 1,330 student-days), students were absent from school for 220~days (18\% of the total) of which 129~days (59\% of absences) were due to illness.
@@
 
-We analyzed 448~saliva samples over the study across classes.
+We analyzed a total of 448~saliva samples.
 We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenovirus (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~virus~(PIV) (\cref{fig:molecular-descriptives}a).
-There was an equal number of positive samples with and without air cleaners (\cref{fig:molecular-descriptives}b).
-Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
-The distribution of positive samples varied between classes (\cref{fig:molecular-descriptives}c).
+The distribution of positive saliva samples varied between classes (\cref{fig:molecular-descriptives}b).
 For example, during the first three study weeks, all but one sample was positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
-We also analyzed 105~bioaerosol samples and detected  two respiratory viruses in the air (1~HRV in class~A and 1~AdV in class~B).
-Finally, we detected 1~Flu~B, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
+To illustrate possible transmission chains within classes, we linked positive saliva samples of the same virus that were less than one week apart.
+Based on this spatiotemporal analysis, we identified 10~possible transmission chains (\cref{fig:molecular-network}).
+The longest transmission chains possibly occurred in January and refer to a cluster of AdV infections in class~A and Flu~B infections in class~B.
+
+Molecular genotyping to verify the proposed transmission network was not successful because we were unable to amplify and sequence any of the gene targets.
+Furthermore, the rate of positive bioaerosol samples was low.
+We analyzed 105~bioaerosol samples and detected only two respiratory viruses in the air (1~HRV in class~A and 1~AdV in class~B).
+Similarly, we detected only 1~Flu~B, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
+However, the virus-specific variation between classes and the temporal distribution of positive samples still suggest that transmission occurred within classes.
 
@@
     \includegraphics{../../results/mol-data/descriptives.pdf}
-    \caption{\textbf{Analysis of molecular data}.
-Number of positive saliva samples by \textbf{(a)}~virus, \textbf{(b)}~study condition, and \textbf{(c)}~study week and class.
+    \caption{\textbf{Molecular detection of respiratory viruses and transmission network based on spatiotemporal analysis of students' saliva samples}.
+\textbf{(a)}~Number of positive saliva samples by virus.
+\textbf{(b)}~Daily number of positive saliva samples (colored circles) and possible transmission chains within classes (directed arrows).
+Positive samples are linked if they belong to the same virus and are less than 1~week apart.
+Positive samples from the air and filters as blank squares aligned.
 Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
@@
 
-\subsection{Analyses of transmission networks and molecular genotyping}
+\subsection{Risk of infection with and without air cleaners}
 
-To illustrate possible transmission chains, we linked positive saliva samples within classes that belong to the same virus and were less than 1~week apart.
-Based on this spatiotemporal analysis of student saliva samples, we identified10~possible transmission chains, which are shown in \cref{fig:molecular-network}.
-The longest transmission chains possibly occurred in January and refer to the cluster of AdV infections in class~A and Flu~B infections in class~B.
-We also performed molecular genotyping of saliva, bioaerosol, and air filter samples to verify our proposed transmission networks.
-However, molecular genotyping was not successful because we were unable to amplify and sequence any of the gene targets.Nevertheless, the virus-specific variation between classes and the temporal distribution of positive samples suggest that transmission occurred within classes.
+There was an equal number of positive saliva samples with and without air cleaners (\cref{fig:infection-risk}b).
+The positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-CrI 0.49 to 1.61), and was not sensitive to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
+
+Particle concentrations differed considerably between study conditions (\cref{fig:infection-risk}a).
+Aerosol number concentration with air cleaners were markedly lower (27$\pm$34\,1/cm$^3$) than without air cleaners (95$\pm$81\,1/cm$^3$), corresponding to an adjusted decrease of 76\% (95\%-CrI 63\% to 86\%).
+Similarly, particle mass concentrations were markedly lower with air cleaners and the decrease was greater for larger particles (PM$_{10}$) than for smaller particles (PM$_1$ to PM$_{4}$) (\supp~figure~\zref{fig:palas-results} and \supp~table~\zref{tab:palas-est-results}).
+There was no considerable difference in other environmental variables (\supp~figure~\zref{fig:env-descriptives-other-vars}).
+In particular, daily average CO$_2$ levels were 1,769$\pm$391\,ppm with and 1,636$\pm$341\,ppm without air cleaners, suggesting that ventilation was comparable between study conditions.
+
+Absences related to respiratory infections were less frequent with air cleaners (\cref{fig:infection-risk}c).
+Overall, there were 22~respiratory cases without and 13~cases with air cleaners.
+Based on our Bayesian latent variable hierarchical regression model, the posterior probability that air cleaners reduced the risk of infection was 90.8\%, suggesting a beneficial effect of air cleaners.
+However, the decrease in the risk of infection was small (adjusted relative risk 0.72, 95\%-CrI 0.44 to 1.15).
+As a result, the estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\supp~\zref{fig:avoided-infections}).
+Detailed estimations are provided in the \supp~text~\zref{sec:detailed-redcap}.
+
+Slightly fewer coughs were detected in audio recordings with air cleaners (\cref{fig:coughing}a).
+On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners ($\Delta=-$0.6~coughs/min; adjusted relative risk 0.93, 95\%-CrI 0.85 to 1.02).
+Significantly more coughs were detected in the room of class~B (adjusted relative risk 2.16, 95\%-CrI 1.90 to 2.46), suggesting an association with the observed variation in virus-specific transmission between classes.
+Indeed, cough was more frequent with Flu~B positivity (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent with AdV positivity (0.87, 95\%-CrI 0.49 to 1.12) (\cref{fig:coughing-association}b).
 
@@
 \centering
-    \includegraphics{../../results/mol-data/network-plot.pdf}
-    \caption{\textbf{Transmission network of respiratory viruses: spatiotemporal analysis of students' saliva samples}.
-Daily number of positive saliva samples (colored circles) and possible transmission chains within classes (directed arrows).
-Positive samples are linked if they belong to the same virus and are less than 1~week apart.
-Positive samples from the air and filters as blank squares aligned.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
-    \label{fig:molecular-network}
-\end{figure}
-
-\subsection{Analysis of aerosol and particle concentrations}
-
-There was a strong difference in particle concentrations between study conditions (\cref{fig:palas-results}a).
-Aerosol number concentration with air cleaners were markedly lower (27$\pm$34\,1/cm$^3$) than without air cleaners (95$\pm$81\,1/cm$^3$).
-Similarly, particle mass concentrations (\eg PM$_1$) were markedly lower with air cleaners (1.16$\pm$1.33\,$\mu$gm$^{-3}$) than without air cleaners (4.20$\pm$3.74\,$\mu$gm$^{-3}$).
-There was no considerable difference in other environmental variables (\supp~figure~\zref{fig:env-descriptives-other-vars}).
-In particular, daily average CO$_2$ levels (air change rates) were 1,769$\pm$391\,ppm (0.74$\pm$0.41\,h$^{-1}$) with air cleaners and 1,636$\pm$341\,ppm (0.68$\pm$0.32\,h$^{-1}$), suggesting that ventilation was comparable between study conditions.
-When adjusting for air change rates and multiple other factors, the aerosol number concentration decreased by 76\% (95\%-CrI 63\% to 86\%) with air cleaners (\cref{fig:palas-results}b and \supp~table~\zref{tab:palas-est-results}).
-The decrease in the concentration of larger particles (PM$_{10}$) was greater than the decrease in the concentration of smaller particles (PM$_1$ to PM$_{4}$).
-
-\begin{figure}[!htpb]
-\centering
-    \includegraphics[width=\linewidth]{../../results/env-data/particles-boxplot.pdf}
-    \includegraphics[width=\linewidth]{../../results/env-data/estimation-results-figure.pdf}
-    \caption{\textbf{Analysis of particle concentrations and comparison between study conditions}.
-\textbf{(a)}~Boxplot of the daily average values for aerosol number concentration (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$).
-\textbf{(b)}~Estimated reduction in aerosol number and particle mass concentrations with air cleaners (posterior mean as dot and 50\%-, 80\%- and 95\%-CrI as lines, respectively).
-}
+    \includegraphics{../../results/mol-data/saliva-by-condition.pdf}
+    \includegraphics{../../results/env-data/estimation-results-figure.pdf}
+    \includegraphics{../../results/epi-data/cases_by_condition.pdf}
+    \includegraphics{results/cough-data/coughs-frequency-by-condition.pdf}
+    \caption{\textbf{Risk of infection by study condition}.
+\textbf{(a)}~Number of positive saliva samples.
+\textbf{(b)}~Daily average aerosol number concentrations (CN in 1/cm$^3$) as boxplots.
+\textbf{(c)}~Number of respiratory cases.
+\textbf{(d)}~Daily average number of detected coughs per minute.}
     \label{fig:palas-results}
-\end{figure}
-
-\subsection{Risk of infection}
-
-Overall, there were 35~respiratory cases (absences related to respiratory infections) across classes.
-Incidentally, there were more cases in class~B (21) than in class~A (14), which is incidentally also the class where more students recently recovered from a SARS-CoV-2 infection or were vaccinated.
-Except for one study week, there was an equal or higher number of cases without air cleaners (\cref{fig:redcap-results}a).
-Based on our Bayesian latent hierarchical regression model, we estimated a trend towards a lower risk of infection with air cleaners (adjusted relative risk 0.72, 95\%-Credible Interval 0.44 to 1.15).
-As a result, the estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\cref{fig:redcap-results}b).
-The posterior probability that air cleaners reduced the risk of infection was 90.8\%, also suggesting a beneficial effect of air cleaners.
-Detailed estimations are provided in the \supp~text~\zref{sec:detailed-redcap}.
-
-\begin{figure}[!htpb]
-    \includegraphics{../../results/epi-data/cases_by_condition.pdf} 
-    \includegraphics{../../results/epi-data/avoided-infections.pdf}
-    \caption{\textbf{Analysis of epidemiological data and estimated number of infections with and without air cleaners.}.
-\textbf{(a)}~Weekly number of respiratory cases with and without air cleaners.
-\textbf{(b)}~Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
-    \label{fig:redcap-results}
-\end{figure}
-
-\subsection{Detected coughs}
-
-Slightly fewer coughs were detected in audio recordings during study weeks where air cleaners were installed (\cref{fig:coughing}a).
-On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners ($\Delta$ $-$0.6~coughs/min; adjusted relative risk 0.93, 95\%-CrI 0.85 to 1.02).
-Significantly more coughs were detected in class~B (adjusted relative risk 2.16, 95\%-CrI 1.90 to 2.46), suggesting a link with the observed variation in virus-specific transmission between classes (\cref{fig:molecular-descriptives}c).
-As suspected, the frequency of coughing was associated with the number of positive saliva samples by respiratory virus (\cref{fig:coughing}b).
-Coughing was more frequent during Flu~B positivity (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent during AdV positivity (0.87, 95\%-CrI 0.49 to 1.12).
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{results/cough-data/coughs-frequency-by-condition.pdf}
-    \includegraphics{results/cough-data/coughs-virus-association.pdf}
-    \caption{\textbf{Analysis of coughing data.} \textbf{(a)}~Daily average number of detected coughs per minute by study week, condition and class.
-\textbf{(b)}~Estimated association (relative risk; posterior mean as dot, 50\%, 80\%, and 95\%-CrI as lines) between cough frequency and number of positive saliva test results by respiratory virus.}
-    \label{fig:coughing}
-\end{figure}
-
-\subsection{Comparison with previous study}
-
-In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 Omicron wave.
-\cref{fig:comparison} shows a comparison of the results.
-The distribution of respiratory infections is different (\cref{fig:comparison}a) because we almost exclusively detected SARS-CoV-2 in our previous study during the COVID-19 pandemic, as compared to a range of respiratory infections in this study during non-pandemic times.
-Although there were more positive saliva samples in this study, the number of respiratory infections detected in bioaerosols and on the HEPA-filters of the air cleaners was lower (\cref{fig:comparison}b).
-The reductions in particle concentrations were larger and with lower uncertainty in this study (\cref{fig:comparison}c).
-The risk of infection tended to be lower with air cleaners in this study where air cleaners were analyzed using a cross-over design, but not in our previous study where air cleaners were installed only at the end of the study when most students had already been infected with SARS-Cov-2(\cref{fig:comparison}d).
-
-\begin{figure}[!htpb]
-    \includegraphics[width=\linewidth]{../../results/comparison.pdf} 
-    \caption{\textbf{Comparison of the results between our previous study in 2022 during the COVID-19 pandemic (blue) and this study in 2023 during non-pandemic times (black).} \textbf{(a)}~Proportion of positive saliva samples by virus.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.
-\textbf{(c)}~Estimated reduction in aerosol number and particle mass concentrations (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$) with air cleaners (posterior mean as squares and 95\%-CrI as lines).
-\textbf{(d)}~Adjusted relative risk (ARR) of infection with air cleaners (posterior mean as squares and 95\%-CrI as lines).}
-    \label{fig:comparison}
 \end{figure}
@@
 
-We collected epidemiological, environmental, and molecular data to estimate the risk of respiratory virus infection in a Swiss school and assessed the effectiveness of air cleaners as an infection control measure.
+We used a multiple-measurement approach to estimate the risk of respiratory virus infection in a Swiss school and assessed the effectiveness of air cleaners.
 We found a wide range of respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus, but very few respiratory viruses were detected in bioaerosol samples and on the filters of the air cleaners.
-Aerosol number and particle mass concentrations decreased significantly with air cleaners and Bayesian modeling based on the absences of students due to illness showed a small, potential reduction in the relative risk of infection with air cleaners.
+Aerosol number and particle mass concentrations decreased significantly with air cleaners and Bayesian modeling based on epidemiological data showed a small, potential reduction in the relative risk of infection with air cleaners.
 Reduced coughing could indicate that air cleaners prevent some symptomatic infections.
 
-% comparison with our previous study
+% comparison with previous study
 
-In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners in reducing the transmission of SARS-CoV-2 in two Swiss schools.
-We found a detectable effect on the risk of infection only for mask wearing, but not air cleaners.
+In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during winter 2022.
+Our previous study was conducted during the SARS-CoV-2 Omicron wave and most detected viruses were SARS-CoV-2.
+We found a detectable effect on the risk of SARS-CoV-2 infection only for mask wearing, but not air cleaners.
 We argued that air cleaners may have had less potential to reduce infections because they were introduced at the end of the study, when most students were already infected with SARS-CoV-2.
-In this study, we used a cross-over design to estimate the effectiveness of air cleaners again, but this time under non-pandemic conditions.
-We found similar reductions in particle concentrations and a tendency for the relative risk of infection to be lower with air cleaners.
-To some extent, the differences between the two studies may be explained by different adjustments.
-While we could not adjust for ventilation in our previous study, the estimates in this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see \supp~text~\zref{sec:detailed-molecular}).
-Furthermore, the distribution of respiratory viruses in this study was very different from that in our previous study.
+In this study, we used a cross-over design and estimated the effectiveness of air cleaners during winter 2023.
+This time, we detected a range of respiratory viruses, which consisted primarily of adenovirus, influenza~B and rhinovirus, with only three positive SARS-CoV-2 saliva samples.
+A similar shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.
+In contrast to our previous study, we found a small, potential benefit of air cleaners on the risk of respiratory infection.
 
-% why could we not detect so many bioaerosols
+% discussing the effect of air cleaners
 
-We detected fewer respiratory viruses in bioaerosol samples compared to our previous study\cite{Banholzer2023PLoSMed} (1~sample of adenovirus and 1~sample of rhinovirus in this study vs 10~samples of SARS-CoV-2 in the previous study), especially in relation to the number of positive saliva samples (50~samples in this study vs 21~in the previous study).
-Similarly, the number of viruses detected on the filters of the air cleaners was also lower in this study (4~positive samples in 160~swabs vs 8~positive samples in 80~swabs in the previous study).
-Importantly, there was a clear shift in the distribution of respiratory viruses in the current study compared to the previous study in 2022 during the SARS-CoV-2 Omicron variant.
-Whereas in 2022 we detected almost exclusively SARS-CoV-2 in human saliva, the most common respiratory viruses in 2023 were adenovirus, influenza~B, and rhinoviruses (only three SARS-CoV-2-positive samples).
-The shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.
+There may be several reasons for the relatively small effect of air cleaners.
+Compared to \eg universal mask wearing, which has been shown to be effective\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Leung2020NatMed,Milton2013PLoSPathogens}, air cleaners cannot affect transmission outside the classroom.
+Furthermore, previous work suggests that prolonged close contact may be necessary for transmission\cite{Leung2020NatMed,Brankston2007LancetID}, or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
+However, compared to face masks, air cleaners are not able to prevent transmission by close range, high particle density.
+Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective at removing larger particles ($>5\mu$m).
+However, a greater proportion of respiratory viruses are carried in the smaller particles, which are also more relevant for transmission ($\leq5\mu$m)\cite{Fennelly2020}.
+Finally, classroom activity, airflow and other unobserved, confounding factors make it difficult to evaluate the effects of air cleaners.
+Nevertheless, it is important to assess their effectiveness in a real-world setting and to compare them with the hypothetical impact suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
 
-% other reasons
+% transmission within classroom
 
-Technical reasons for the differences in airborne detection are unlikely.
-The bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
-Instead, airborne detection could have been influenced by environmental factors.
-Although temperature (overall between 19 and 25$^{\circ}$C) and relative humidity (overall between 25 and 50\%) were similar in both studies, the same levels may have different effects on the airborne survival of respiratory viruses.
-For example, airborne survival of rhinovirus and adenovirus increases with relative humidity, whereas the opposite relationship has been documented for influenza and SARS-CoV-2\cite{Tellier2009JTRSI,Ahlawat2020AAQR,Biryukov2020mS,Karim1985CJM,Davis1971AM}.
-A change in ventilation conditions could also have had a profound effect, but CO$_2$ levels were higher compared to the previous study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
-Although lower ventilation was partially offset by a greater reduction in aerosol and particle concentrations from air cleaners, the removal of particles by air cleaners may be slower than by natural ventilation.
-
-% bioaerosol sampling in more depth
-
-We speculate that the difference in airborne molecular detection of respiratory viruses between 2022 (pandemic) and 2023 (non-pandemic) is due to the different characteristics of the respiratory viruses.
-Differences in the physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity), which influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}, may facilitate the detection of SARS-CoV-2 in the air.
-The reason is that viral loads do often not exceed the detection limits of existing devices for bioaerosol sampling, which remains a challenge\cite{Wang2021,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
-For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.
-However, several studies showed that influenza\cite{Bischoff2013JID,Pan2017mSphere}, rhinovirus\cite{Myatt2004AJRCCM}, and adenovirus\cite{Nguyen2017OFID,Pan2017mSphere} can also be detected in airborne samples.
-Finally, the ability to detect pathogens in bioaerosols may be related to patient characteristics of the study participants.
-Several studies suggested an association between the generation of pathogenic aerosols and infectiousness\cite{Leung2020NatMed,Bischoff2013JID,Escombe2008PLoSMed}.
-Therefore, studies that include more infectious patients (intentionally or unintentionally) are likely to have a higher chance of detecting respiratory viruses in the air.
-
-% transmission within classrooms
-
-The low rate of positive bioaerosol samples also means that it is unclear to what extent airborne transmission occurred in classrooms.
+We detected only few respiratory viruses in bioaerosol samples (1~sample of adenovirus and 1~sample of rhinovirus ) and on the filters of the air cleaners (4~positive samples in 160~swabs) in this study.
+The low rate of positive bioaerosol samples means that it is unclear to what extent airborne transmission occurred in classrooms.
 It may be possible that students had relatively little exposure to respiratory viruses in schools and acquired their infections elsewhere.
@@
 Taken together, the class-specific, spatiotemporal patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.
-
-% effects of air cleaners
-
-Our study indicated a small, potential benefit of air cleaners.
-There may be several reasons for this relatively small effect.
-Compared to \eg universal mask wearing, which has been shown to be effective\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Dharmadhikari2012AJRCCM,Leung2020NatMed,Milton2013PLoSPathogens}, air cleaners cannot affect transmission outside the classroom.
-Furthermore, previous work suggests that prolonged close contact may be necessary for transmission\cite{Leung2020NatMed,Brankston2007LancetID}, or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
-However, compared to face masks, air cleaners are not able to prevent transmission by close range, high particle density.
-Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective at removing larger particles ($>5\mu$m).
-However, a greater proportion of respiratory viruses are carried in the smaller particles, which are also more relevant for transmission ($\leq5\mu$m)\cite{Fennelly2020}.
-Finally, classroom activity, airflow and other unobserved, confounding factors make it difficult to evaluate the effects of air cleaners.
-Nevertheless, it is important to assess their effectiveness in a real-world setting and to compare them with the hypothetical impact suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
 
@@
 Based on our transmission model using epidemiological data, the risk of infection tends to be lower with air filters.
-Future studies should investigate the relative importance of prolonged close contact and long range transmission of respiratory viruses in schools and other congregate settings.
+Future studies should investigate the relative importance of prolonged close contact and long-range transmission of respiratory viruses in schools and other congregate settings.
 
@@
 %TC:ignore
+
+\section*{Funding}
+This study is funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
+NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
+ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498).
+\medskip
+
+\section*{Author contributions}
+Conception and design: NB, KZ, LF, PB, PJ, TS.
+Epidemiological and environmental data collection: NB, PJ, KZ, TS, LF.
+Laboratory data collection: PB, LFu.
+Additional data collection: TH.
+Cough detection: SB.
+Molecular genotyping: RD, AR, PB, LFu.
+Statistical analysis: NB.
+Paper draft: NB, LF, ME.
+All authors reviewed and approved the final version of the manuscript.
+
 \section*{Acknowledgements}

### 2023-10-16 analysis/aud_analysis.Rmd
update plots

@@
   ggplot(aes(x = aircleaner, y = coughs_per_min, fill = aircleaner)) +
-  geom_bar(stat = "identity", color = "black") +
+  geom_bar(stat = "identity", color = "black", width = .4) +
   geom_text(mapping = aes(label = round(coughs_per_min, 1)),
@@
   scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
   theme_custom() +
@@
           eps_file = "../results/cough-data/coughs-frequency-by-condition.eps", 
-          w = 8, h = 8)
+          w = 6, h = 6)
 ```

### 2023-10-16 analysis/env_analysis.Rmd
update plots

@@
   ggplot(aes(x = aircleaner, fill = aircleaner, y = Cn)) +
-  geom_boxplot2() +
+  geom_boxplot2(width.errorbar = .3, width = .4) +
   #geom_jitter(alpha = .3, width = .2, color = "grey") +
@@
   scale_y_continuous(expand = c(0,0), limits = c(0, NA)) +
-  labs(title = "b", y = expression("Aerosol number concentration (1/cm"^3*")"), x = "Air cleaner") +
+  labs(title = "b", y = expression("Aerosol concentration (1/cm"^3*")"), x = "Air cleaner") +
   theme_custom() +
@@
           eps_file = "../results/env-data/aerosol-number-boxplot.eps", 
-          w = 8, h = 8)
+          w = 6, h = 6)
 ```

### 2023-10-16 analysis/epi_analysis.Rmd
update plots

@@
   ggplot(aes(x = aircleaner, y = cases, fill = aircleaner)) +
-  geom_bar(stat = "identity", color = "black") +
+  geom_bar(stat = "identity", color = "black", width = .4) +
   geom_text(mapping = aes(label = cases),
@@
   scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.1)), breaks = scales::pretty_breaks(4)) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.2)), breaks = scales::pretty_breaks(4)) +
   theme_custom() +
@@
           eps_file = "../results/epi-data/cases_by_condition.eps",
-          w = 8, h = 8) 
+          w = 6, h = 6) 
 ```

### 2023-10-16 analysis/mol_analysis.Rmd
update plots

@@
 pos_cond_pl <- ggplot(mapping = aes(x = aircleaner, y = n, fill = aircleaner)) +
-  geom_bar(data = df_pos_stud, stat = "identity", position = "stack", color = "black") +
+  geom_bar(data = df_pos_stud, stat = "identity", position = "stack", color = "black", width = .4) +
   geom_text(data = subset(df_pos_stud, n != 0), mapping = aes(label = n), 
             size = 8 / cm(1), vjust = -1) +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
   scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
@@
         plot.title.position = "plot",
-        legend.position = "top",
-        legend.title = element_blank(),
-        legend.key.width = unit(.3, "cm"),
-        legend.key.height = unit(.1, "cm")) +
+        legend.position = "none") +
   guides(fill = guide_legend(nrow = 1))
@@
           "../results/mol-data/saliva-by-study-condition.eps",
-          w = 8, h = 8)
+          w = 6, h = 6)
 ```
@@
 pos_tot_pl <- ggplot(mapping = aes(y = n, x = 1, fill = variable)) +
-  geom_bar(data = df_pos_tot, stat = "identity", position = position_stack(reverse = T)) +
-  geom_text(data = subset(df_pos_tot, n != 0), mapping = aes(label = n), 
-            size = 8 / cm(1), position = position_stack(vjust = 0.5, reverse = T), color = "white") +
+  geom_bar(data = df_pos_tot, stat = "identity", position = position_stack(), width = .5) +
+  geom_text(data = df_pos_tot %>% mutate(n_lab = paste0(variable, ": ", n)), 
+            mapping = aes(label = n_lab), 
+            size = 8 / cm(1), position = position_stack(vjust = 0.5), color = "white") +
   scale_fill_manual(values = wes_palette("IsleofDogs1")) +
+  scale_color_manual(values = wes_palette("IsleofDogs1")) +
   scale_y_continuous(expand = c(0,0)) +
-  scale_x_continuous(expand = expansion(mult = c(0.05,0))) +
-  coord_flip() +
+  scale_x_continuous(expand = expansion(mult = c(0.1,0.3))) +
   labs(y = "Number of positive samples", fill = "Virus", title = "a") +
@@
         plot.title.position = "plot",
-        legend.position = "top",
-        legend.title = element_blank(),
-        legend.key.width = unit(.3, "cm"),
-        legend.key.height = unit(.1, "cm"),
-        axis.title.y = element_blank(),
-        axis.line.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank()) +
+        legend.position = "none",
+        # legend.title = element_blank(),
+        # legend.key.width = unit(.3, "cm"),
+        # legend.key.height = unit(.1, "cm"),
+        axis.title.x = element_blank(),
+        axis.line.x = element_blank(),
+        axis.text.x = element_blank(),
+        axis.ticks.x = element_blank()) +
   guides(fill = guide_legend(nrow = 1))
@@
           "../results/mol-data/saliva-distribution.eps",
-          w = 16, h = 3.5)
+          w = 3.25, h = 21)
 ```
@@
           eps_file = "../results/mol-data/network-plot.eps", 
-          w = 16, h = 16.5)
+          w = 12.75, h = 21)
 ```

### 2023-10-16 doc/paper/manuscript-lancet.tex

@@
 We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenovirus (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~virus~(PIV) (\cref{fig:molecular-descriptives}a).
-The distribution of positive saliva samples varied between classes (\cref{fig:molecular-descriptives}b).
+The distribution of positive saliva samples varied between classes.
 For example, during the first three study weeks, all but one sample was positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
 To illustrate possible transmission chains within classes, we linked positive saliva samples of the same virus that were less than one week apart.
-Based on this spatiotemporal analysis, we identified 10~possible transmission chains (\cref{fig:molecular-network}).
+Based on this spatiotemporal analysis, we identified 10~possible transmission chains (\cref{fig:molecular-descriptives}b).
 The longest transmission chains possibly occurred in January and refer to a cluster of AdV infections in class~A and Flu~B infections in class~B.
@@
     \centering
-    \includegraphics{../../results/mol-data/descriptives.pdf}
+    \includegraphics{../../results/mol-data/saliva-distribution.pdf}
+    \includegraphics{../../results/mol-data/network-plot.pdf}
     \caption{\textbf{Molecular detection of respiratory viruses and transmission network based on spatiotemporal analysis of students' saliva samples}.
@@
 
-Slightly fewer coughs were detected in audio recordings with air cleaners (\cref{fig:coughing}a).
+Slightly fewer coughs were detected in audio recordings with air cleaners (\cref{fig:infection-risk}d).
 On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners ($\Delta=-$0.6~coughs/min; adjusted relative risk 0.93, 95\%-CrI 0.85 to 1.02).
 Significantly more coughs were detected in the room of class~B (adjusted relative risk 2.16, 95\%-CrI 1.90 to 2.46), suggesting an association with the observed variation in virus-specific transmission between classes.
-Indeed, cough was more frequent with Flu~B positivity (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent with AdV positivity (0.87, 95\%-CrI 0.49 to 1.12) (\cref{fig:coughing-association}b).
+Indeed, cough was more frequent with Flu~B positivity (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent with AdV positivity (0.87, 95\%-CrI 0.49 to 1.12) (\supp~figure~\zref{fig:coughing-association}b).
 
@@
 \centering
-    \includegraphics{../../results/mol-data/saliva-by-condition.pdf}
-    \includegraphics{../../results/env-data/estimation-results-figure.pdf}
+    \includegraphics{../../results/mol-data/saliva-by-study-condition.pdf}
+    \includegraphics{../../results/env-data/aerosol-number-boxplot.pdf}
     \includegraphics{../../results/epi-data/cases_by_condition.pdf}
-    \includegraphics{results/cough-data/coughs-frequency-by-condition.pdf}
+    \includegraphics{../../results/cough-data/coughs-frequency-by-condition.pdf}
     \caption{\textbf{Risk of infection by study condition}.
 \textbf{(a)}~Number of positive saliva samples.
-\textbf{(b)}~Daily average aerosol number concentrations (CN in 1/cm$^3$) as boxplots.
+\textbf{(b)}~Daily average aerosol number concentrations as boxplots.
 \textbf{(c)}~Number of respiratory cases.
 \textbf{(d)}~Daily average number of detected coughs per minute.}
-    \label{fig:palas-results}
+    \label{fig:infection-risk}
 \end{figure}

### 2023-10-16 analysis/aud_analysis.Rmd
update plots

@@
             size = 8 / cm(1), vjust = -1) +
-  labs(x = "Study week", y = "Number of coughs per min", fill = "Air cleaner", title = "d") +
+  labs(x = "Air cleaner", y = "Number of coughs per min", fill = "Air cleaner", title = "d") +
   scale_fill_manual(values = wes_palette("Royal1")[1:2]) +

### 2023-10-16 analysis/epi_analysis.Rmd
update plots

@@
             size = 8 / cm(1), vjust = -1) +
-  labs(x = "Air cleaner", y = "Number of cases", fill = "Air cleaner", title = "c") +
+  labs(x = "Air cleaner", y = "Number of respiratory cases", fill = "Air cleaner", title = "c") +
   scale_fill_manual(values = wes_palette("Royal1")[1:2]) +

### 2023-10-16 analysis/mol_analysis.Rmd
update plots

@@
 pos_tot_pl <- ggplot(mapping = aes(y = n, x = 1, fill = variable)) +
-  geom_bar(data = df_pos_tot, stat = "identity", position = position_stack(), width = .5) +
+  geom_bar(data = df_pos_tot, stat = "identity", position = position_stack(reverse = T), width = .5) +
   geom_text(data = df_pos_tot %>% mutate(n_lab = paste0(variable, ": ", n)), 
             mapping = aes(label = n_lab), 
-            size = 8 / cm(1), position = position_stack(vjust = 0.5), color = "white") +
+            size = 8 / cm(1), position = position_stack(vjust = 0.5, reverse = T), color = "white") +
   scale_fill_manual(values = wes_palette("IsleofDogs1")) +
@@
           "../results/mol-data/saliva-distribution.eps",
-          w = 3.25, h = 21)
+          w = 3.25, h = 20)
 ```
@@
              mapping = aes(color = virus), fill = "white",
-             position = position_dodge(width = -.6, preserve = "single")) +
+             position = position_dodge(width = -.85, preserve = "single")) +
   geom_point(mapping = aes(fill = virus), color = "black") +
@@
         axis.title = element_blank(),
-        #axis.text.x = element_blank(),
-        #axis.ticks.x = element_blank(),
-        panel.spacing = unit(1, "cm"),
+        panel.spacing = unit(.5, "cm"),
         panel.grid.major.x = element_line(),
@@
           eps_file = "../results/mol-data/network-plot.eps", 
-          w = 12.75, h = 21)
+          w = 12.25, h = 20)
 ```

### 2023-10-16 doc/paper/manuscript-lancet.tex

@@
     \centering
-    \includegraphics{../../results/mol-data/saliva-distribution.pdf}
+    \includegraphics{../../results/mol-data/saliva-distribution.pdf}\hspace{.5cm}
     \includegraphics{../../results/mol-data/network-plot.pdf}
@@
 \centering
-    \includegraphics{../../results/mol-data/saliva-by-study-condition.pdf}
+    \includegraphics{../../results/mol-data/saliva-by-study-condition.pdf}\hspace{1.5cm}
     \includegraphics{../../results/env-data/aerosol-number-boxplot.pdf}
-    \includegraphics{../../results/epi-data/cases_by_condition.pdf}
+    \includegraphics{../../results/epi-data/cases_by_condition.pdf}\hspace{1.5cm}
     \includegraphics{../../results/cough-data/coughs-frequency-by-condition.pdf}

### 2023-10-17 analysis/aud_analysis.Rmd
add posterior probabilities to plots

--- 2023-10-16 analysis/aud_analysis.Rmd
@@
 coughs_pl <- coughs %>%
-  group_by(aircleaner) %>%
-  summarize(coughs_per_min = sum(n_coughs) / sum(TT)) %>%
-  ungroup() %>%
+  mutate(coughs_per_min = n_coughs / TT) %>%
   mutate(aircleaner = factor(aircleaner, levels = c("No", "Yes"))) %>%
-  ggplot(aes(x = aircleaner, y = coughs_per_min, fill = aircleaner)) +
-  geom_bar(stat = "identity", color = "black") +
-  geom_text(mapping = aes(label = round(coughs_per_min, 1)),
-            size = 8 / cm(1), vjust = -1) +
-  labs(x = "Study week", y = "Number of coughs per min", fill = "Air cleaner", title = "d") +
-  scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
+  ggplot(aes(x = aircleaner, fill = aircleaner, y = coughs_per_min)) +
+  geom_boxplot2(width.errorbar = .3, width = .4) +
+  ggsignif::geom_signif(annotations = "PP(Yes<No) = 93%", y_position = 6.25, xmin = 1, xmax = 2, 
+                        textsize = 8 / cm(1), vjust = -.5) +
+  scale_fill_manual(values = wes_palette("Royal1")) +
+  scale_y_continuous(limits = c(0, NA), expand = expansion(add = c(0, .66))) +
+  labs(title = "b", y = "Number of coughs per min", x = "Air cleaner") +
   theme_custom() +
   theme(legend.position = "none",
-        legend.direction = "horizontal",
-        legend.key.width = unit(0.5, "cm"),
-        legend.key.height = unit(0.3, "cm"),
-        plot.title.position = "plot") +
-  guides(fill = "none")
+        plot.title.position = "plot")
 
@@
           eps_file = "../results/cough-data/coughs-frequency-by-condition.eps", 
-          w = 8, h = 8)
+          w = 7.25, h = 6)
 ```
@@
 exp(raw_res["beta2",c("mean", "X2.5.", "X97.5.")])
+
+# posterior probability
+coughsMod %>%
+  spread_draws(beta1) %>%
+  summarize(round(sum(beta1 < 0) / n() * 100))
 ```

### 2023-10-17 analysis/env_analysis.Rmd
add posterior probabilities to plots

--- 2023-10-16 analysis/env_analysis.Rmd
@@
   ggplot(aes(x = aircleaner, fill = aircleaner, y = Cn)) +
-  geom_boxplot2() +
-  #geom_jitter(alpha = .3, width = .2, color = "grey") +
+  geom_boxplot2(width.errorbar = .3, width = .4) +
+  ggsignif::geom_signif(annotations = "PP(Yes<No) = 100%", y_position = 240, xmin = 1, xmax = 2, 
+                        textsize = 8 / cm(1), vjust = -.5) +
   scale_fill_manual(values = wes_palette("Royal1")) +
-  scale_shape_manual(values = c(2,20)) +
-  scale_y_continuous(expand = c(0,0), limits = c(0, NA)) +
-  labs(title = "b", y = expression("Aerosol number concentration (1/cm"^3*")"), x = "Air cleaner") +
+  scale_y_continuous(expand = expansion(add = c(0, 30)), limits = c(0, NA)) +
+  labs(title = "b", y = expression("Aerosol concentration (1/cm"^3*")"), x = "Air cleaner") +
   theme_custom() +
   theme(legend.position = "none",
-        plot.title.position = "plot") +
-  guides(color = "none")
+        plot.title.position = "plot")
 
@@
           eps_file = "../results/env-data/aerosol-number-boxplot.eps", 
-          w = 8, h = 8)
+          w = 7.25, h = 6)
 ```
@@
   mutate(outcome = factor(rep(particle_vars, each = nrow(draws[[i]])), levels = rev(particle_vars))) %>%
-  set_names(c("aircleaner", "outcome")) %>%
+  set_names(c("aircleaner", "outcome"))
+
+# posterior probability
+draws_filt %>%
+  filter(outcome == "Cn") %>%
+  summarize(round(sum(aircleaner < 0) / n() * 100))
+
+# adjusted decrease
+draws_filt <- draws_filt %>%
   mutate(aircleaner = (-1) * 100 * (exp(aircleaner) - 1))

### 2023-10-17 analysis/epi_analysis.Rmd
add posterior probabilities to plots

--- 2023-10-16 analysis/epi_analysis.Rmd
@@
   ggplot(aes(x = aircleaner, y = cases, fill = aircleaner)) +
-  geom_bar(stat = "identity", color = "black") +
-  geom_text(mapping = aes(label = cases),
-            size = 8 / cm(1), vjust = -1) +
-  labs(x = "Air cleaner", y = "Number of cases", fill = "Air cleaner", title = "c") +
+  geom_bar(stat = "identity", color = "black", width = .4) +
+  geom_text(mapping = aes(label = cases), size = 8 / cm(1), vjust = -1) +
+  ggsignif::geom_signif(annotations = "PP(Yes<No) = 91%", y_position = 27, xmin = 1, xmax = 2, 
+                        textsize = 8 / cm(1), vjust = -.5, tip_length = .075) +
+  labs(x = "Air cleaner", y = "Number of respiratory cases", fill = "Air cleaner", title = "c") +
   scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
@@
           eps_file = "../results/epi-data/cases_by_condition.eps",
-          w = 8, h = 8) 
+          w = 7.25, h = 6) 
 ```

### 2023-10-17 analysis/mol_analysis.Rmd
add posterior probabilities to plots

@@
   group_by(aircleaner) %>%
-  summarize(n = sum(n)) %>%
+  summarize(n_pos = sum(n)) %>%
   ungroup() %>%
@@
 
-pos_cond_pl <- ggplot(mapping = aes(x = aircleaner, y = n, fill = aircleaner)) +
-  geom_bar(data = df_pos_stud, stat = "identity", position = "stack", color = "black", width = .4) +
-  geom_text(data = subset(df_pos_stud, n != 0), mapping = aes(label = n), 
-            size = 8 / cm(1), vjust = -1) +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
+pos_cond_pl <- df_pos_stud %>%
+  ggplot(aes(x = aircleaner, y = n_pos)) +
+  geom_bar(mapping = aes(fill = aircleaner), stat = "identity", color = "black", width = .4) +
+  geom_text(mapping = aes(label = n_pos), size = 8 / cm(1), vjust = -1) +
+  geom_segment(data = data.frame(x = 1, xend = 1, y = 30, yend = 29), mapping = aes(x = x, xend = xend, y = y, yend = yend), color = "white") +
+  ggsignif::geom_signif(annotations = "PP(Yes<No) = 65%", y_position = 28, xmin = 1, xmax = 2, 
+                        textsize = 8 / cm(1), vjust = -.5) +
+  labs(x = "Air cleaner", y = "Number of positive saliva samples", title = "a") +
   scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
-  labs(y = "Number of positive samples", x = "Air cleaner", fill = "Virus", title = "a") +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.1)), breaks = scales::pretty_breaks(4)) +
   theme_custom() +
-  theme(panel.grid = element_blank(),
-        plot.title.position = "plot",
-        legend.position = "none") +
-  guides(fill = guide_legend(nrow = 1))
+  theme(plot.title.position = "plot") +
+  guides(fill = "none")
 
@@
           "../results/mol-data/saliva-by-study-condition.eps",
-          w = 6, h = 6)
+          w = 7.25, h = 6)
 ```
@@
   mean_qi()
+
+# posterior probability
+spread_draws(multLogitMod, beta1) %>%
+  summarize(round(sum(beta1 < 0) / n() * 100))
 ```

### 2023-10-17 doc/paper/manuscript-lancet.tex
Updates from Overleaf

@@
 Estimation power is evaluated via the effective sample size (ESS) and convergence of the Markov chains is evaluated with the Gelman-Rubin convergence diagnostic ($\hat{R}$).
-If not stated otherwise, we report posterior means and credible intervals (CrIs) based on the 50\%, 80\%, and 95\% quantiles of the posterior samples, respectively.
-The code is available from \url{https://osf.io/38j9g}.
+If not stated otherwise, we report the posterior probability (PP), mean, and 95\%credible interval (CrI) of the posterior samples. The code is available from \url{https://osf.io/38j9g}.
 
@@
 Overall, there were 22~respiratory cases without and 13~cases with air cleaners.
-Based on our Bayesian latent variable hierarchical regression model, the posterior probability that air cleaners reduced the risk of infection was 90.8\%, suggesting a beneficial effect of air cleaners.
+Based on our Bayesian latent variable hierarchical regression model, the posterior probability that air cleaners reduced the risk of infection was 91\%, suggesting a beneficial effect of air cleaners.
 However, the decrease in the risk of infection was small (adjusted relative risk 0.72, 95\%-CrI 0.44 to 1.15).
@@
 \centering
-    \includegraphics{../../results/mol-data/saliva-by-study-condition.pdf}\hspace{1.5cm}
+    \includegraphics{../../results/mol-data/saliva-by-study-condition.pdf}\hspace{.5cm}
     \includegraphics{../../results/env-data/aerosol-number-boxplot.pdf}
-    \includegraphics{../../results/epi-data/cases_by_condition.pdf}\hspace{1.5cm}
+    \includegraphics{../../results/epi-data/cases_by_condition.pdf}\hspace{.5cm}
     \includegraphics{../../results/cough-data/coughs-frequency-by-condition.pdf}
@@
 \textbf{(c)}~Number of respiratory cases.
-\textbf{(d)}~Daily average number of detected coughs per minute.}
+\textbf{(d)}~Daily average number of detected coughs per minute.
+``PP(Yes<No)'' is the posterior probability that the outcome is lower with air cleaners, which is computed based on the posterior samples of the Bayesian models adjusted for observed confounders.}
     \label{fig:infection-risk}

### 2023-10-17 analysis/aud_analysis.Rmd
add posterior probabilities to plots

@@
   scale_y_continuous(limits = c(0, NA), expand = expansion(add = c(0, .66))) +
-  labs(title = "b", y = "Number of coughs per min", x = "Air cleaner") +
+  labs(title = "d", y = "Number of coughs per min", x = "Air cleaner") +
   theme_custom() +

### 2023-10-17 analysis/mol_analysis.Rmd
add posterior probabilities to plots

@@
   geom_text(mapping = aes(label = n_pos), size = 8 / cm(1), vjust = -1) +
-  geom_segment(data = data.frame(x = 1, xend = 1, y = 30, yend = 29), mapping = aes(x = x, xend = xend, y = y, yend = yend), color = "white") +
-  ggsignif::geom_signif(annotations = "PP(Yes<No) = 65%", y_position = 28, xmin = 1, xmax = 2, 
-                        textsize = 8 / cm(1), vjust = -.5) +
+  geom_segment(data = data.frame(x = .5, xend = .5, y = 30, yend = 29), mapping = aes(x = x, xend = xend, y = y, yend = yend), color = "white") +
+  ggsignif::geom_signif(annotations = "PP(Yes<No) = 65%", y_position = 30, xmin = 1, xmax = 2, 
+                        textsize = 8 / cm(1), vjust = -.5, tip_length = .1) +
   labs(x = "Air cleaner", y = "Number of positive saliva samples", title = "a") +

### 2023-10-17 doc/paper/manuscript-lancet.tex

@@
 There was an equal number of positive saliva samples with and without air cleaners (\cref{fig:infection-risk}b).
-The positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-CrI 0.49 to 1.61), and was not sensitive to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
+The posterior probability that fewer samples were positive with air cleaners was only 65\%.
+As such, the positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-CrI 0.49 to 1.61), and was not sensitive to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
 

### 2023-10-17 analysis/mol_analysis.Rmd
add posterior probabilities to plots

@@
   scale_x_continuous(expand = expansion(mult = c(0.1,0.3))) +
-  labs(y = "Number of positive samples", fill = "Virus", title = "a") +
+  labs(y = "Number of positive saliva samples", fill = "Virus", title = "a") +
   theme_custom() +
@@
           "../results/mol-data/saliva-distribution.eps",
-          w = 3.25, h = 20)
+          w = 3.25, h = 19)
 ```
@@
           eps_file = "../results/mol-data/network-plot.eps", 
-          w = 12.25, h = 20)
+          w = 12.25, h = 19)
 ```

### 2023-10-17 analysis/mol_analysis.Rmd
add posterior probabilities to plots

@@
           "../results/mol-data/saliva-distribution.eps",
-          w = 3.25, h = 19)
+          w = 3.25, h = 20)
 ```
@@
           eps_file = "../results/mol-data/network-plot.eps", 
-          w = 12.25, h = 19)
+          w = 12.25, h = 20)
 ```

### 2023-10-17 analysis/mol_analysis.Rmd
add posterior probabilities to plots

@@
           "../results/mol-data/saliva-distribution.eps",
-          w = 3.25, h = 20)
+          w = 3.25, h = 21)
 ```
@@
           eps_file = "../results/mol-data/network-plot.eps", 
-          w = 12.25, h = 20)
+          w = 12.25, h = 21)
 ```

### 2023-10-17 analysis/mol_analysis.Rmd
add posterior probabilities to plots

@@
           "../results/mol-data/saliva-distribution.eps",
-          w = 3.25, h = 21)
+          w = 3.25, h = 20.5)
 ```
@@
           eps_file = "../results/mol-data/network-plot.eps", 
-          w = 12.25, h = 21)
+          w = 12.25, h = 20.5)
 ```

### 2023-10-17 analysis/aud_analysis.Rmd
rename flu b to ifb

@@
 coughs %>%
+  mutate(coughs_per_min = n_coughs / TT) %>%
   group_by(aircleaner) %>%
-  summarize(sum(n_coughs) / sum(TT))
+  summarize(mean(coughs_per_min),
+            sd(coughs_per_min)) %>%
+  mutate_if(is.numeric, round, 1)
 
@@
   TT = coughs_TueThu$TT,
-  viruses = as.matrix(dplyr::select(coughs_TueThu, `Flu B`, HRV, AdV, `SARS-CoV-2`, MPV, PIV))
+  viruses = as.matrix(dplyr::select(coughs_TueThu, `IFB`, HRV, AdV, `CoV`, MPV, PIV))
 )
@@
 # plot
-viruses <- c("Average", "Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")
+viruses <- c("Average", "IFB", "HRV", "AdV", "CoV", "MPV", "PIV")
 names(viruses) <- 0:6

### 2023-10-17 analysis/epi_analysis.Rmd
rename flu b to ifb

@@
 prior_pin_par <- data.frame(
-  virus = c("AdV", "Flu B", "PIV", "HRV", "CoV", "MPV"),
+  virus = c("AdV", "IFB", "PIV", "HRV", "CoV", "MPV"),
   mum = c(p_in_mu_m, mean(p_in_mu_m)), 
@@
 ) %>%
-  mutate(virus = factor(virus, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV"))) %>%
+  mutate(virus = factor(virus, levels = c("IFB", "HRV", "AdV", "CoV", "MPV", "PIV"))) %>%
   arrange(virus)
@@
          week = ifelse(week > 4, week - 1, week))  %>%
-  dplyr::select(week, class, AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`) %>%
+  dplyr::select(week, class, AdV, `IFB`, HRV, MPV, PIV, `CoV`) %>%
   melt(c("week", "class")) %>%
@@
   mutate(variable = as.character(variable), 
-         variable = ifelse(variable == "SARS-CoV-2", "CoV", variable),
-         variable = factor(variable, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")))
+         variable = factor(variable, levels = c("IFB", "HRV", "AdV", "CoV", "MPV", "PIV")))
 mol_class_week_extend <- epi_data %>%
@@
   ggtitle("a | Number of positive saliva samples by study condition.",
-          subtitle = "Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza (PIV).")
+          subtitle = "Influenza B (IFB), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza (PIV).")
 boxplot_env <- readRDS("../results/env-data/particles-boxplot.rds") +
@@
                       "$\\vartheta_{\\text{Friday}}$",
-                      "$\\mu_{p_{\\text{IN}}^{\\text{Flu B}}}$",
+                      "$\\mu_{p_{\\text{IN}}^{\\text{IFB}}}$",
                       "$\\mu_{p_{\\text{IN}}^{\\text{HRV}}}$",
@@
                       "$\\mu_{p_{\\text{IN}}^{\\text{PIV}}}$",
-                      "$\\sigma_{p_{\\text{IN}}^{\\text{Flu B}}}$",
+                      "$\\sigma_{p_{\\text{IN}}^{\\text{IFB}}}$",
                       "$\\sigma_{p_{\\text{IN}}^{\\text{HRV}}}$",

### 2023-10-17 analysis/mol_analysis.Rmd
rename flu b to ifb

@@
     "A", "20-02-2023", "AdV", "Air",
-    "B", "01-02-2023", "Flu B", "Filter", # we set the dates for the filter swabs to align with saliva
+    "B", "01-02-2023", "IFB", "Filter", # we set the dates for the filter swabs to align with saliva
     "A", "27-02-2023", "CoV", "Filter",
@@
          week = ifelse(week > 4, week - 1, week)) %>%
-  dplyr::select(aircleaner, week, class, AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`) %>%
+  dplyr::select(aircleaner, week, class, AdV, `IFB`, HRV, MPV, PIV, `CoV`) %>%
   melt(c("aircleaner", "week", "class")) %>%
@@
   mutate(variable = as.character(variable), 
-         variable = ifelse(variable == "SARS-CoV-2", "CoV", variable),
-         variable = factor(variable, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")),
+         variable = factor(variable, levels = c("IFB", "HRV", "AdV", "CoV", "MPV", "PIV")),
          class = paste("Class", class)) %>%
@@
          virus = ifelse(virus == "SARS-CoV-2", "CoV", virus),
-         virus = factor(virus, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV")),
+         virus = factor(virus, levels = c("IFB", "HRV", "AdV", "CoV", "MPV", "PIV")),
          class = paste("Class", class)) %>%
@@
                       type = factor(type, levels = c("Saliva", "Air", "Filter")),
-                      virus = factor(virus, levels = c("Flu B", "HRV", "AdV", "CoV", "MPV", "PIV"))),
+                      virus = factor(virus, levels = c("IFB", "HRV", "AdV", "CoV", "MPV", "PIV"))),
              mapping = aes(color = virus), fill = "white",
@@
 
+tnet_pl
+
 save_plot(tnet_pl, 
@@
     arrange(class, date) %>%
-    dplyr::select(Neg, AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`) %>%
+    dplyr::select(Neg, AdV, `IFB`, HRV, MPV, PIV, `CoV`) %>%
     as.matrix()
@@
     arrange(date) %>%
-    mutate(across(c(AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`), ~ n_class - cumsum(.x)),
-           across(c(AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`), ~ (.x - mean(.x)) / sd(.x)),
+    mutate(across(c(AdV, `IFB`, HRV, MPV, PIV, `CoV`), ~ n_class - cumsum(.x)),
+           across(c(AdV, `IFB`, HRV, MPV, PIV, `CoV`), ~ (.x - mean(.x)) / sd(.x)),
            PIV = 0) %>% # no variation in PIV
@@
     arrange(class, date) %>%
-    dplyr::select(AdV, `Flu B`, HRV, MPV, PIV, `SARS-CoV-2`)
+    dplyr::select(AdV, `IFB`, HRV, MPV, PIV, `CoV`)
 

### 2023-10-17 preprocessing/prep_molecular.R

@@ -14,13 +14,14 @@ df <- read_xlsx("data-raw/molecular/ZF Resultate Airchecker 2023.xlsx", sheet = 
          virus = `Auto   Interpretation`) %>%
   dplyr::select(class, date, virus) %>% # We set A as the "Studienklasse" and B as the "Kontrollklasse" --> 3a is B and 3b is A
   mutate(class = ifelse(class == "3a", "B", "A"),
-         date = as.Date(gsub("2022", "2023", as.character(date))))
+         date = as.Date(gsub("2022", "2023", as.character(date))),
+         virus = ifelse(virus == "SARS-CoV-2", "CoV", virus),
+         virus = ifelse(virus == "Flu B", "IFB", virus))
 
 # line list
 
 long_list <- df %>%
-  mutate(result = ifelse(virus == "-", "Negative", "Positive"),
-         virus = ifelse(virus == "SARS-CoV-2", "CoV", virus)) %>%
+  mutate(result = ifelse(virus == "-", "Negative", "Positive")) %>%
   dplyr::select(class, date, result, virus) %>%
   arrange(class, virus) %>%
   ungroup()
@@ -48,7 +49,7 @@ df <- expanded_df %>%
 # total count
 df_wide <- reshape2::dcast(df, class + date ~ virus) %>%
   rename(Neg = `-`) %>%
-  mutate(N = Neg + AdV + `Flu B` + HRV + MPV + PIV + `SARS-CoV-2`)
+  mutate(N = Neg + AdV + IFB + HRV + MPV + PIV + CoV)
 
 # add additional info
 df_wide <- df_wide %>%

### 2023-10-17 doc/paper/manuscript-lancet.tex
Updates from Overleaf

--- 2023-10-17 doc/paper/manuscript-lancet.tex
@@
 
-\title{\bfseries\LARGE\singlespacing{Transmission of respiratory viral infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}}
+\title{\bfseries\LARGE\singlespacing{Impact of air cleaners in schoolson respiratory viral infections: A modeling study using epidemiological, environmental, and molecular data}}
 
@@
 \author[1]{Simon Bertschinger}
+\author[5]{Ernest Weingartner}
 \author[2,4]{Alban Ramette}
@@
 \author[1]{Matthias Egger}
-\author[5]{Tina Hascher}
+\author[6]{Tina Hascher}
 \author[1*,2]{Lukas Fenner}
@@
 \affil[4]{Institute for Infectious Diseases, University of Bern, Bern, Switzerland}
-\affil[5]{Institute of Educational Science, University of Bern, Bern, Switzerland}
+\affil[5]{School of Engineering, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
+\affil[6]{Institute of Educational Science, University of Bern, Bern, Switzerland}
 
@@
 \begin{abstract}\normalfont
-\noindent\textbf{Background:} Infection control measures should reduce airborne transmission.
-We aimed to evaluate the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing transmission of respiratory viral infections in schools using a multiple-measurement approach.
+\noindent\textbf{Background:} Infection control measures can reduce airborne transmission.
+Using a multiple-measurement approach, we aimed to evaluate the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing transmission of respiratory viral infections in schools.
 \medskip
 
-\noindent\textbf{Methods:} We collected epidemiological data (absences related to respiratory infections), environmental data (CO$_2$ and particle concentrations), audio recordings (detecting coughs), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2023 in two classes of a Swiss secondary school.
+\noindent\textbf{Methods:} We collected epidemiological data (absences related to respiratory infections), environmental data (CO$_2$ and particle concentrations), audio recordings (coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2023 in two classes of a Swiss secondary school.
 Using a cross-over study design, we compared particle concentrations, coughing, and the risk of infection with and without air cleaners.
@@
 Spatiotemporal analysis of positive saliva samples suggested within-classroom transmission.
-Daily average CO$_2$ levels with and without air cleaners were similar (1,769$\pm$391\,ppm vs.
-1,636$\pm$341\,ppm).
-Particle concentrations decreased by 77\% (95\%-credible interval [CrI] 63\% to 86\%), coughs were less frequent ($-$0.5~coughs/min), and the estimated risk of infection likely lower with air cleaners (adjusted relative risk 0.72, 95\%-CrI 0.44 to 1.15).
+Testing positive was equally likely with air cleaners (adjusted relative risk [ARR] 0.93 95\%-credible interval [CrI] 0.49 to 1.61).
+CO$_2$ levels were similar (1,769$\pm$391\,ppm with vs.
+1,636$\pm$341\,ppm without air cleaners), yet particle concentrations decreased by 77\%, (95\%-CrI 63\% to 86\%) and coughing was slightly less frequent (ARR 0.93, 95\%-CrI 0.85 to 1.02).
+Infection risk was also lower (posterior probability 91\%), but the association was small (ARR 0.72, 95\%-CrI 0.44 to 1.15).
 \medskip
@@
 Although airborne detection was rare, transmission seemed likely in schools, suggesting that prolonged close contact is required for transmission.
-Air cleaners improved air quality, were associated with lower symptom intensity, and showed a small, potential benefit in reducing infections.
+Air cleaners improved air quality, were associated with lower symptom intensity, and showed a potential benefit in reducing infections.
 
@@
 Improved ventilation systems are critical for a healthy indoor environment and can reduce the risk of respiratory transmission\cite{Wang2021,Morawska2021}.
-Portable HEPA-air filtration devices (air cleaners) may be another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear.
+Portable HEPA-air filtration devices (air cleaners) may be another cost-effective alternative to upgrading ventilation systems.
+Still, the effects of air cleaners on the transmission of respiratory infections are less clear.
 Several studies have shown that air cleaners reduce particle concentrations in indoor air\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed}, which is associated with a reduced risk of respiratory infection and mortality\cite{Gordon2014Resp,Kelly2023Atmos,DeAngelis2021EnvRes}.
@@
 Simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission  of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
-%We have studied the effectiveness of air cleaners in reducing transmission of SARS-CoV-2 in schools during the pandemic\cite{Banholzer2023PLoSMed}, but the analysis was limited by the study design (installation of air cleaners at the end of the Omicron wave when most students were already infected).
-To date, it remains unclear whether reducing particle concentrations and removing bioaerosols from the air will result in reduced transmission of respiratory infections.
+To date, it remains unclear whether reducing particle concentrations and removing bioaerosols from the air will reduce transmission of respiratory infections.
 
-We used a multiple-measurement approach to study the transmission of respiratory viruses and the effect of air cleaners in schools.
-We used a cross-over study design under non-pandemic conditions in the winter of 2023.
-We collected epidemiological data (absences likely related to respiratory infections), environmental data (CO$_2$, aerosol and particle concentrations), audio recordings (detecting coughs), and molecular data (detection of viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
-We determined changes in particle concentrations, respiratory virus positivity rates, and modeled the risk of infection.
+We used a multiple-measurement approach to study the transmission of respiratory viruses under non-pandemic conditions and the effect of air cleaners in schools with a cross-over study design in the winter of 2023.
+We collected epidemiological data (absences likely related to respiratory infections), environmental data (CO$_2$, particle concentrations), audio recordings (coughing), and molecular data (detection of viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
+We determined changes in absences related to respiratory infections, positivity rates of saliva samples, particle concentrations, and coughing.
 
@@
 
-\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, during seven weeks from 26~January to 11~March 2023.
-Figure~\ref{fig:study-setup} shows the schematic study setup and \supp~table~\zref{tab:data} provides an overview of the types of data collected.
+\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, for seven weeks from January~16 to March~11, 2023.
+Figure~\ref{fig:study-setup} shows the schematic study setup, and \supp~table~\zref{tab:data} provides an overview of the types of data collected.
 
@@
 All devices were placed at the head level of the students when they were seated.
-Both classrooms were not equipped with an active HVAC (Heating, Ventilation, Air conditioning) system, but were ventilated naturally by opening windows.
+Neither classrooms was equipped with an active HVAC (Heating, Ventilation, Air conditioning) system, but they were ventilated naturally by opening windows.
 }
@@
 According to the manufacturer, these air cleaners run at clean air delivery rates of 2$\,\times\,$600 m$^{3}$/h.
-% We measured a lower clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h when testing the devices in an empty classroom.
 Window ventilation was done at the discretion of the teachers, as there were no specific recommendations from the national health authorities at the time.
@@
     \centering
-    \caption{\textbf{Study design.} Cross-over design describing when portable air cleaners (AC) were installed in the rooms of classes A and B during a seven-week study period from 16~January to 11~March 2023, excluding a week of vacation from February 6 to 11.
-A ``*'' indicates when swabs from the HEPA filters were taken, i.e.
-after each intervention phase and before the vacation.}\label{tab:study_design}
+    \caption{\textbf{Study design.} Cross-over design describing when portable air cleaners (AC) were installed in the rooms of classes A and B during a seven-week study period from January~16 to March~11, 2023, excluding a week of vacation from February 6 to 11.}\label{tab:study_design}
     \begin{tabular}{l l l l l l l l l}
@@
       & \textbf{Week 1} & \textbf{Week 2} & \textbf{Week 3} & Vacation & \textbf{Week 4} & \textbf{Week 5} & \textbf{Week 6} & \textbf{Week 7} \\
-      & Jan 16-21 & Jan 23-28 & Jan 30-Feb 3 & Feb 6-11 & Feb 13-18 & Feb 20-25 & Feb 27-March 3 & March 6-11 \\
+      & Jan 16-21 & Jan 23-28 & Jan 30-Feb 3 & Feb 6-11 & Feb 13-18 & Feb 20-25 & Feb 27-Mar 3 & Mar 6-11 \\
       \midrule
@@
       \bottomrule
+      \multicolumn{9}{l}{\scriptsize *~Swabs from the HEPA filters taken after each intervention phase and before the vacation.}
     \end{tabular}
@@
 
-\noindent At the beginning of the study, we collected aggregated data on age, sex, and COVID-19 vaccination and recovery status in the participating classes.
-On a daily basis, we collected data on each absent student, \ie the absence and return dates and reason for the absence (illness, career days, or other).
+\noindent At the beginning of the study, we collected aggregated data on age, sex, COVID-19 vaccination, and COVID-19 recovery status in the participating classes.
+Daily, we collected data on each absent student, \ie the absence and return dates and reason for the absence (illness, career days, or other).
 For absences due to illness, we recorded symptoms and the date of symptom onset.
@@
 
-\noindent \emph{Saliva} \smallskip
-
 \noindent Both classes participated in repetitive bi-weekly testing (Tuesdays and Thursdays; see \nameref{sec:mol_analyses} below).
@@
 The line list data is shown in \supp~table~\zref{tab:mol-data-line-list}.
-\medskip
-
-\noindent \emph{Bioaerosol and HEPA filters of the air cleaners} \smallskip
-
-\noindent We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL PBS.
-The Coriolis Micro Air could only operate shortly before and during break times to prevent unacceptable noise exposure (approximately 60\,min/day).
+Furthermore, We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL Phosphate-Buffered Saline (PBS).
+The Coriolis Micro Air ran shortly before and during break times to prevent unacceptable noise exposure (approximately 60\,min/day).
 In one class, we also used the BioSpot-VIVAS device sampler (Aerosol Devices Inc., Ft.
@@
 BioSpot-VIVAS operated throughout the lessons.
-Removable parts of both sampling devices were regularly autoclaved.
+The removable parts of both sampling devices were regularly autoclaved.
 At the end of the day, samples were transported to the Institute of Infectious Diseases (IFIK) and stored at $-$80°C.
-After each intervention phase (see \cref{tab:study_design}), the HEPA filters were carefully removed from the air cleaners, divided into 20~fields, and one swab per field was taken for a total of 20~swabs per filter.
+Finally, we collected swabs from the HEPA filters of the air cleaners after each intervention phase (see \cref{tab:study_design}).
+The HEPA filters were removed from the air cleaners and divided into 20~fields.
+One swab per field was then taken for a total of 20~swabs per filter.
 \medskip
@@
 
-\noindent \emph{CO$_2$ and particle measurements}
-
 \noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute.
 This device was used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}.
-A transient mass balance model was used to estimate the air change rate from indoor CO$_2$ levels\cite{Batterman2017IJERPH}.%, assuming that classrooms were fully occupied during lessons and that room occupancy increased (decreased) linearly 10~minutes before (after) the start (end) of a lesson.
-Particle concentrations were filtered according to the times when students were in the classroom. %, which were monitored together with the times during which windows were opened.
 
@@
 The human saliva samples were analyzed directly without prior filtration.
-The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19~respiratory viruses, including SARS-CoV-2 (CoV), influenza (Flu), respiratory syncytial (RSV), metapneumovirus (MPV), adenovirus (AdV), rhinovirus (HRV), and parainfluenza (PIV).
-%/Viral genome load (VGL) of specimens was quantified (copies/L) by curve extrapolation derived from standardized ATCC quantitative genomic standard dilutions (ATCC-VR-3347D, LGC Standards, LGC Group UK).
-For the BioSpot-VIVAS air samples, the air concentration of the DNA was exactly calculated according to the airflow influx into the sampling device.
-The Limit of detection (LOD) of the assay SARS-CoV-2, IFA/IFB is 7.5x10$^0$ TCID/ml (SARS-CoV-2), 3.7x10$^{-1}$ TCID/ml (IFA), and 4.8x100 TCID/ml (100cp/ml) (IFB) respectively.
+The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 8~respiratory viruses, including SARS-CoV-2 (CoV), influenza (IF), respiratory syncytial (RSV), metapneumovirus (MPV), adenovirus (AdV), rhinovirus (HRV), and parainfluenza (PIV).
 
@@
 
-We performed molecular genotyping for positive samples for AdV and Flu~B.
-(saliva, bioaerosol, air filter samples).
+We performed molecular genotyping for positive saliva, bioaerosol, air filter samples of AdV and IF.
 For AdV, we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}.
-For Flu~B, we tried to determine the molecular transmission network using whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
+For IF, we tried to determine the molecular transmission network using whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
 
@@
 A detailed description of the model and choice of priors for all model parameters are provided in \supp~text~\zref{sec:transmission-model}.
-%Importantly, since multiple viruses were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the virus-specific distributions.
-The weights were inferred based on the proportion of positive saliva samples for each virus per week.
-As a result, the distribution of the incubation period used for modeling could change every week depending on the proportion of viruses that were detected in the students' saliva.
-Moreover, a re-weighting scheme was used to account for weekday effects.
-\medskip
 
@@
 
-\noindent We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~text~\zref{sec:multinomial-model}) and linked the expected number of positive samples to the presence of air cleaners.
-\medskip % The number of susceptibles was computed as the difference between the number of students per class and the cumulative number of positive samples.
-As a sensitivity analysis, we lagged the test results once by one test date (\ie from Tuesday to last Thursday and from Thursday to last Tuesday) and once by one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday), to consider the unknown delay when students were infected and tested positive.
+\noindent We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~text~\zref{sec:multinomial-model}) and linked the expected number of positive samples to the presence of air cleaners, adjusting for the cumulative number of positive tests.
 
@@
 \noindent We computed the average particle concentrations per day and compared them between study conditions.
-We estimated the reduction in particle concentrations using Bayesian log-linear regression models (\supp~text~\zref{sec:env-regression-model}).
-\medskip %, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+We estimated the reduction in particle concentrations using Bayesian log-linear regression models, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases (\supp~text~\zref{sec:env-regression-model}).
+\medskip
 
@@
 \noindent We computed the daily number of detected coughs and compared it between study conditions.
-We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model (\supp~text~\zref{sec:aud-regression-model}).
-%with time in class as the model offset, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model, using time in class as the model offset and adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases (\supp~text~\zref{sec:aud-regression-model}).
 In addition, we used a Bayesian Negative Binomial hierarchical (random effects) regression model to estimate the association between the number of coughs and the number of positive saliva test results.
@@
 Model parameters were estimated with a Bayesian approach, using the Hamiltonian Monte Carlo algorithm with the No-U-Turn Sampler (NUTS)\cite{Hoffman2014}.
-%Each model is estimated with 4 Markov chains and 2,000 iterations of which the first 1,000 iterations are discarded as part of the warm-up.
-Estimation power is evaluated via the effective sample size (ESS) and convergence of the Markov chains is evaluated with the Gelman-Rubin convergence diagnostic ($\hat{R}$).
-If not stated otherwise, we report the posterior probability (PP), mean, and 95\%credible interval (CrI) of the posterior samples. The code is available from \url{https://osf.io/38j9g}.
+If not stated otherwise, we report for each outcome the posterior probability (PP) that the adjusted relative risk (ARR) was lower with air cleaners (ARR<1).
+ARR is reported with mean and 95\%credible interval (CrI) based on 4,000 samples from the posterior distribution of each model.
+The code is available from \url{https://osf.io/38j9g}.
 
@@
     \footnotesize
+    \renewcommand{\arraystretch}{1.5}
     \begin{tabular}{l r r r}
@@
 We analyzed a total of 448~saliva samples.
-We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenovirus (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~virus~(PIV) (\cref{fig:molecular-descriptives}a).
+We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenoviruses (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~viruses~(PIV) (\cref{fig:molecular-descriptives}a).
 The distribution of positive saliva samples varied between classes.
-For example, during the first three study weeks, all but one sample was positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
+For example, during the first three study weeks, all but one sample was positive for AdV in class~A, while the vast majority of positive IFB samples were found in class~B.
 To illustrate possible transmission chains within classes, we linked positive saliva samples of the same virus that were less than one week apart.
 Based on this spatiotemporal analysis, we identified 10~possible transmission chains (\cref{fig:molecular-descriptives}b).
-The longest transmission chains possibly occurred in January and refer to a cluster of AdV infections in class~A and Flu~B infections in class~B.
+The longest potential transmission chains occurred in January, referring to a cluster of AdV infections in class~A and IFB infections in class~B.
 
-Molecular genotyping to verify the proposed transmission network was not successful because we were unable to amplify and sequence any of the gene targets.
+Molecular genotyping to verify the proposed transmission network was unsuccessful because we could not amplify and sequence any of the gene targets.
 Furthermore, the rate of positive bioaerosol samples was low.
 We analyzed 105~bioaerosol samples and detected only two respiratory viruses in the air (1~HRV in class~A and 1~AdV in class~B).
-Similarly, we detected only 1~Flu~B, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
+Similarly, we detected only 1~IFB, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
 However, the virus-specific variation between classes and the temporal distribution of positive samples still suggest that transmission occurred within classes.
@@
 
-There was an equal number of positive saliva samples with and without air cleaners (\cref{fig:infection-risk}b).
-The positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-CrI 0.49 to 1.61), and was not sensitive to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
+There was an equal number of 25~positive saliva samples in both study conditions (\cref{fig:infection-risk}b).
+Based on our Bayesian Multinomial logistic regression model, the positivity rate remained comparable between study conditions (ARR 0.93, 95\%-CrI 0.49 to 1.61; PP 61\%).
+The estimated ARR was not sensitive to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
 
-Particle concentrations differed considerably between study conditions (\cref{fig:infection-risk}a).
-Aerosol number concentration with air cleaners were markedly lower (27$\pm$34\,1/cm$^3$) than without air cleaners (95$\pm$81\,1/cm$^3$), corresponding to an adjusted decrease of 76\% (95\%-CrI 63\% to 86\%).
-Similarly, particle mass concentrations were markedly lower with air cleaners and the decrease was greater for larger particles (PM$_{10}$) than for smaller particles (PM$_1$ to PM$_{4}$) (\supp~figure~\zref{fig:palas-results} and \supp~table~\zref{tab:palas-est-results}).
-There was no considerable difference in other environmental variables (\supp~figure~\zref{fig:env-descriptives-other-vars}).
+The aerosol number concentration was 95$\pm$81\,1/cm$^3$ without and 27$\pm$34\,1/cm$^3$ with air cleaners (\cref{fig:infection-risk}a).
+The adjusted decrease in the aerosol concentration with air cleaners was 76\% (95\%-CrI 63\% to 86\%; PP 100\%).
+This decrease was greater for larger (PM$_{10}$) than for smaller (PM$_1$ to PM$_{4}$) particle matter mass concentrations  (\supp~figure~\zref{fig:palas-results} and \supp~table~\zref{tab:palas-est-results}).
+There was little change in other environmental variables (\supp~figure~\zref{fig:env-descriptives-other-vars}).
 In particular, daily average CO$_2$ levels were 1,769$\pm$391\,ppm with and 1,636$\pm$341\,ppm without air cleaners, suggesting that ventilation was comparable between study conditions.
 
-Absences related to respiratory infections were less frequent with air cleaners (\cref{fig:infection-risk}c).
-Overall, there were 22~respiratory cases without and 13~cases with air cleaners.
-Based on our Bayesian latent variable hierarchical regression model, the posterior probability that air cleaners reduced the risk of infection was 91\%, suggesting a beneficial effect of air cleaners.
-However, the decrease in the risk of infection was small (adjusted relative risk 0.72, 95\%-CrI 0.44 to 1.15).
-As a result, the estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\supp~\zref{fig:avoided-infections}).
+Absences related to respiratory infections were 22~respiratory cases without and 13~cases with air cleaners (\cref{fig:infection-risk}c).
+Our Bayesian latent variable hierarchical regression model suggested that air cleaners reduced the risk of infection (PP 91\%).
+However, the reduction was small and subject to uncertainty (ARR 0.72, 95\%-CrI 0.44 to 1.15).
+The estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\supp~\zref{fig:avoided-infections}).
 Detailed estimations are provided in the \supp~text~\zref{sec:detailed-redcap}.
 
-Slightly fewer coughs were detected in audio recordings with air cleaners (\cref{fig:infection-risk}d).
-On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners ($\Delta=-$0.6~coughs/min; adjusted relative risk 0.93, 95\%-CrI 0.85 to 1.02).
-Significantly more coughs were detected in the room of class~B (adjusted relative risk 2.16, 95\%-CrI 1.90 to 2.46), suggesting an association with the observed variation in virus-specific transmission between classes.
-Indeed, cough was more frequent with Flu~B positivity (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent with AdV positivity (0.87, 95\%-CrI 0.49 to 1.12) (\supp~figure~\zref{fig:coughing-association}b).
+We detected 3.1$\pm$1.2~coughs/min in audio recordings without and 2.6$\pm$1.1~coughs/min with air cleaners (\cref{fig:infection-risk}d).
+Based on our Bayesian Negative Binomial regression model, the risk of coughing was slightly lower with air cleaners (ARR 0.93, 95\%-CrI 0.85 to 1.02; PP 93\%).
+Significantly more coughs were detected in the room of class~B (ARR 2.16, 95\%-CrI 1.90 to 2.46), suggesting an association with the observed variation in virus-specific transmission between classes.
+Indeed, cough was more frequent with IFB positivity (ARR 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent with AdV positivity (ARR 0.87, 95\%-CrI 0.49 to 1.12) (\supp~figure~\zref{fig:coughing-association}b).
 
@@
 \textbf{(d)}~Daily average number of detected coughs per minute.
-``PP(Yes<No)'' is the posterior probability that the outcome is lower with air cleaners, which is computed based on the posterior samples of the Bayesian models adjusted for observed confounders.}
+The posterior probability (PP) for each outcome being lower with air cleaners (Yes<No) was computed based on the posterior samples of each Bayesian model used to estimate the effect of air cleaners after adjusting for confounders.}
     \label{fig:infection-risk}
@@
 We found a wide range of respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus, but very few respiratory viruses were detected in bioaerosol samples and on the filters of the air cleaners.
-Aerosol number and particle mass concentrations decreased significantly with air cleaners and Bayesian modeling based on epidemiological data showed a small, potential reduction in the relative risk of infection with air cleaners.
+Aerosol number and particle mass concentrations decreased significantly with air cleaners and Bayesian modeling based on epidemiological data showed a potential reduction in the relative risk of infection with air cleaners.
 Reduced coughing could indicate that air cleaners prevent some symptomatic infections.
@@
 Nevertheless, it is important to assess their effectiveness in a real-world setting and to compare them with the hypothetical impact suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
+
+% economic and psychological impact of air cleaners
+% tbd
+% psychological part from Tina
 

### 2023-10-17 doc/paper/supplements.tex
Updates from Overleaf

@@
     \beta_0 &\sim \text{Student-t}_5(0, 2.5) \\
-    \theta_{[\text{Flu~B}]}, \dots, \theta_{[\text{PIV}]} &\sim \text{Normal}\left(\theta, \tau\right) \\
+    \theta_{[\text{IFB}]}, \dots, \theta_{[\text{PIV}]} &\sim \text{Normal}\left(\theta, \tau\right) \\
     \theta &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{v}}\right) \\
@@
 \end{align*}
-where $\theta_{[v]}$ is the partially pooled estimate for respiratory virus $v \in$ (Flu~B, HRV, AdV, CoV, MPV, PIV), $\tau$ estimates variation between viruses, and $s_v$ is the average standard deviation across the count variables of the respiratory viruses.
+where $\theta_{[v]}$ is the partially pooled estimate for respiratory virus $v \in$ (IFB, HRV, AdV, CoV, MPV, PIV), $\tau$ estimates variation between viruses, and $s_v$ is the average standard deviation across the count variables of the respiratory viruses.
 
@@
 \Cref{fig:env-descriptives-other-vars} compares additional environmental variables, \ie CO$_2$, air change rate, relative humidity, and temperature.
+The air change rate was estimated from indoor CO$_2$ levels using a transient mass balance model\cite{Batterman2017IJERPH}.
 There are no considerable differences between in any of the variables between study conditions.
@@
 \Cref{fig:coughing-association} shows the relative risk of cough for each respiratory virus.
-Cough appeared more frequent for Flu~B and less frequent for AdV.
+Cough appeared more frequent for IFB and less frequent for AdV.
 

### 2023-10-19 analysis/aud_analysis.Rmd
change color in plots and reorder

@@
                         textsize = 8 / cm(1), vjust = -.5) +
-  scale_fill_manual(values = wes_palette("Royal1")) +
+  scale_fill_manual(values = bluegrey) +
   scale_y_continuous(limits = c(0, NA), expand = expansion(add = c(0, .66))) +

### 2023-10-19 analysis/env_analysis.Rmd
change color in plots and reorder

@@
                         textsize = 8 / cm(1), vjust = -.5) +
-  scale_fill_manual(values = wes_palette("Royal1")) +
+  scale_fill_manual(values = bluegrey) +
   scale_y_continuous(expand = expansion(add = c(0, 30)), limits = c(0, NA)) +
-  labs(title = "b", y = expression("Aerosol concentration (1/cm"^3*")"), x = "Air cleaner") +
+  labs(title = "c", y = expression("Aerosol concentration (1/cm"^3*")"), x = "Air cleaner") +
   theme_custom() +
@@
   geom_jitter(aes(shape = Class), alpha = .3, width = .2) +
-  scale_color_manual(values = wes_palette("Royal1")) +
+  scale_color_manual(values = bluegrey) +
   scale_shape_manual(values = c(2,20)) +
@@
   guides(color = "none")
+
+boxpl_pc
 
@@
   guides(color = "none")
+
+boxpl_oth
+
 save_plot(boxpl_oth, pdf_file = "../results/env-data/otherVars-boxplot.pdf", w = 16, h = 8)
@@
   facet_wrap(~ outcome, nrow = 1, labeller = as_labeller(vars_labels_no_units, label_parsed)) +
-  scale_fill_manual(values = wes_palette("Royal1")[c(3,4,2)]) +
+  scale_fill_brewer() +
   scale_y_continuous(expand = expansion(mult = c(0.25, 0.25))) +
@@
         panel.spacing = unit(20, units = "pt"))
+
+effects_env_main
+
 save_plot(effects_env_main, 

### 2023-10-19 analysis/epi_analysis.Rmd
change color in plots and reorder

@@
                         textsize = 8 / cm(1), vjust = -.5, tip_length = .075) +
-  labs(x = "Air cleaner", y = "Number of respiratory cases", fill = "Air cleaner", title = "c") +
-  scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
+  labs(x = "Air cleaner", y = "Number of respiratory cases", fill = "Air cleaner", title = "a") +
+  scale_fill_manual(values = bluegrey) +
   scale_y_continuous(expand = expansion(mult = c(0, 0.1)), breaks = scales::pretty_breaks(4)) +
@@
   facet_wrap(~ type, ncol = 1) +
-  scale_fill_manual(values = wes_palette("Royal1")[c(3,4,2)]) +
+  scale_fill_brewer() +
   scale_y_continuous(expand = expansion(mult = c(0.25, 0.25))) +
@@
         strip.text.x = element_text(margin = margin(b = 5)))
+
+avoid_infections_pl
 
@@
                   mapping = aes(y = estimated_cases),
-                  .width = c(0.5, 0.8, 0.95), color = "brown") +
+                  .width = c(0.5, 0.8, 0.95), color = "darkblue") +
   geom_line(data = fitted_cases %>% group_by(week) %>% slice(1),
             mapping = aes(y = true_cases, color = "Observed"), size = 1.5) +
-  scale_fill_manual(values = wes_palette("Royal1")[c(3,4,2)]) +
+  scale_fill_brewer() +
   scale_x_continuous(expand = c(0,0)) +
@@
   guides(color = guide_legend(title = element_blank()))
+
 fitted_cases_pl

### 2023-10-19 analysis/mol_analysis.Rmd
change color in plots and reorder

@@
                         textsize = 8 / cm(1), vjust = -.5, tip_length = .1) +
-  labs(x = "Air cleaner", y = "Number of positive saliva samples", title = "a") +
-  scale_fill_manual(values = wes_palette("Royal1")[1:2]) +
+  labs(x = "Air cleaner", y = "Number of positive saliva samples", title = "b") +
+  scale_fill_manual(values = bluegrey) +
   scale_y_continuous(expand = expansion(mult = c(0, 0.1)), breaks = scales::pretty_breaks(4)) +

### 2023-10-19 utils/plotting.R

@@ -7,6 +7,9 @@ require(gg.layers)
 require(kunstomverse)
 require(wesanderson) # https://github.com/karthik/wesanderson
 
+# default colors
+bluegrey <- c(wes_palette("Royal1")[1], wes_palette("Darjeeling1")[5])
+
 # ggplot theme for Nature style
 text_size = 8
 update_geom_defaults("text", list(size = text_size))

### 2023-10-19 doc/paper/manuscript-lancet.tex
Updates from Overleaf

@@
 
-\title{\bfseries\LARGE\singlespacing{Impact of air cleaners in schoolson respiratory viral infections: A modeling study using epidemiological, environmental, and molecular data}}
+\title{\bfseries\LARGE\singlespacing{Impact of air cleaners in schools on respiratory viral infections: A modeling study using epidemiological, environmental, and molecular data}}
 
@@
 
-\noindent\textbf{Methods:} We collected epidemiological data (absences related to respiratory infections), environmental data (CO$_2$ and particle concentrations), audio recordings (coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2023 in two classes of a Swiss secondary school.
+\noindent\textbf{Methods:} We collected epidemiological data (absences related to respiratory infections), environmental data (CO$_2$ and particle concentrations), audio recordings (coughs), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2023 in two classes of a Swiss secondary school.
 Using a cross-over study design, we compared particle concentrations, coughing, and the risk of infection with and without air cleaners.
@@
 
-\noindent\textbf{Results:} Molecular analysis detected multiple viruses in saliva (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza), but rarely in bioaerosol samples (2/105 positive) and on the filters of air cleaners (4/160~positive).
+\noindent\textbf{Results:} Molecular analysis detected several viruses in saliva (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza), but rarely in bioaerosol samples (2/105 positive) and on filters of air cleaners (4/160~positive).
 Spatiotemporal analysis of positive saliva samples suggested within-classroom transmission.
-Testing positive was equally likely with air cleaners (adjusted relative risk [ARR] 0.93 95\%-credible interval [CrI] 0.49 to 1.61).
+Positive saliva tests were equally likely with air cleaners (adjusted relative risk [ARR] 0.93 95\%-credible interval [CrI] 0.49 to 1.61).
 CO$_2$ levels were similar (1,769$\pm$391\,ppm with vs.
-1,636$\pm$341\,ppm without air cleaners), yet particle concentrations decreased by 77\%, (95\%-CrI 63\% to 86\%) and coughing was slightly less frequent (ARR 0.93, 95\%-CrI 0.85 to 1.02).
-Infection risk was also lower (posterior probability 91\%), but the association was small (ARR 0.72, 95\%-CrI 0.44 to 1.15).
+1,636$\pm$341\,ppm without air cleaners), but particle concentrations decreased by 77\%, (95\%-CrI 63\% to 86\%) and coughing was slightly less frequent (ARR 0.93, 95\%-CrI 0.85 to 1.02).
+Infection risk was also lower (posterior probability 91\%), but the association was small (ARR 0.73, 95\%-CrI 0.44 to 1.18).
 \medskip
@@
 
-\noindent \textcolor{red}{[Search and search terms to be added]} Several studies have shown that air cleaners reduce particle concentrations in indoor air.
-Another study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward.
-A simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission.
-A real-world study during the pandemic showed a trend toward a reduction of the infection risk.
+\noindent We searched Scopus on October 19, 2023, for all publications from inception using the terms ("air filt" OR "air clean") AND ("infection" OR "transmission").
+Several studies have shown that air cleaners reduce particle concentrations in indoor air.
+Studies also showed that air cleaners effectively removed SARS-CoV-2 bioaerosols.
+One study reported lower incidence of SARS-CoV-2 in schools using air filtration devices.
+While simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission, a real-world study during the pandemic was inconclusive.
 \medskip
@@
 
-\noindent We showed the transmission dynamics in classrooms of a secondary school under non-pandemic conditions.
-Although airborne detection was rare, transmission seemed likely in schools, suggesting that prolonged close contact is required for transmission.
-Air cleaners improved air quality, were associated with lower symptom intensity, and showed a small, potential benefit in reducing infections.\medskip
+\noindent To date, it remains unclear whether the reduction in particle concentrations with air cleaners also translates into lower risks of respiratory infection.
+This study used a multiple-measurement approach to estimate the risk of respiratory virus infection in a Swiss school and assessed the effectiveness of air cleaners in a cross-over study design.
+Schools are particularly important because students spend much of their time indoors and may introduce infections from and into their household.
+We collected molecular, epidemiological, environmental, and audio data over seven weeks during non-pandemic conditions in winter 2023 in a Swiss school.
+We found a wide range of respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus.
+Although airborne detection was rare, spatiotemporal analysis of students' saliva samples indicated that transmission seemed likely in school rooms, suggesting that prolonged close contact is required for transmission.
+Air cleaners improved air quality (lower particle concentrations), were associated with lower symptom intensity (fewer coughs), and showed a potential benefit in reducing cases (fewer absences related to respiratory infections).\medskip
 
@@
 
-Improved ventilation systems are critical for a healthy indoor environment and can reduce the risk of respiratory transmission\cite{Wang2021,Morawska2021}.
+Improved ventilation systems are critical for a healthy indoor environment, especially in schools where students spend much of their time indoors during the week, and can reduce the risk of respiratory transmission\cite{Wang2021,Morawska2021}.
 Portable HEPA-air filtration devices (air cleaners) may be another cost-effective alternative to upgrading ventilation systems.
@@
 Several studies have shown that air cleaners reduce particle concentrations in indoor air\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed}, which is associated with a reduced risk of respiratory infection and mortality\cite{Gordon2014Resp,Kelly2023Atmos,DeAngelis2021EnvRes}.
-In addition, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}.
+In addition, recent studies showed that air cleaners effectively removed SARS-CoV-2 from the air\cite{Morris2022,Ueki2022mSphere,Myers2022IndoorAir,Rodriguez2021SOTE}.
 Simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission  of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
-To date, it remains unclear whether reducing particle concentrations and removing bioaerosols from the air will reduce transmission of respiratory infections.
+One study reported lower incidence of SARS-CoV-2 in US elementary schools using air filtration methods\cite{Gettings2021}.
+To date, it remains unclear whether reducing particle concentrations and removing bioaerosols from the air will reduce indoor transmission of respiratory infections.
 
 We used a multiple-measurement approach to study the transmission of respiratory viruses under non-pandemic conditions and the effect of air cleaners in schools with a cross-over study design in the winter of 2023.
-We collected epidemiological data (absences likely related to respiratory infections), environmental data (CO$_2$, particle concentrations), audio recordings (coughing), and molecular data (detection of viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
+We collected epidemiological data (absences likely related to respiratory infections), environmental data (CO$_2$, particle concentrations), audio recordings (coughs), and molecular data (detection of viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
 We determined changes in absences related to respiratory infections, positivity rates of saliva samples, particle concentrations, and coughing.
@@
 Furthermore, We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL Phosphate-Buffered Saline (PBS).
-The Coriolis Micro Air ran shortly before and during break times to prevent unacceptable noise exposure (approximately 60\,min/day).
+The Coriolis Micro Air ran shortly before and during break times (approximately 60\,min/day) to prevent unacceptable noise exposure.
 In one class, we also used the BioSpot-VIVAS device sampler (Aerosol Devices Inc., Ft.
@@
 The HEPA filters were removed from the air cleaners and divided into 20~fields.
-One swab per field was then taken for a total of 20~swabs per filter.
+One PBS-moistened swab per field was then taken for a total of 20~swabs per filter.
 \medskip
@@
 The human saliva samples were analyzed directly without prior filtration.
-The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 8~respiratory viruses, including SARS-CoV-2 (CoV), influenza (IF), respiratory syncytial (RSV), metapneumovirus (MPV), adenovirus (AdV), rhinovirus (HRV), and parainfluenza (PIV).
+The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19 major respiratory viruses and viral subtypes, including SARS-CoV-2 (CoV), influenza (IF), respiratory syncytial (RSV), metapneumovirus (MPV), adenovirus (AdV), rhinovirus (HRV), and parainfluenza (PIV).
 
@@
 A detailed description of the model and choice of priors for all model parameters are provided in \supp~text~\zref{sec:transmission-model}.
+\medskip
 
@@
 \noindent We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~text~\zref{sec:multinomial-model}) and linked the expected number of positive samples to the presence of air cleaners, adjusting for the cumulative number of positive tests.
+\medskip
 
@@
 We analyzed a total of 448~saliva samples.
-We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenoviruses (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~viruses~(PIV) (\cref{fig:molecular-descriptives}a).
+We detected 15~IF~B~(IFB), 15~HRV, 14~AdV, 3~CoV, 2~MPV, and 1~PIV (\cref{fig:molecular-descriptives}a).
 The distribution of positive saliva samples varied between classes.
@@
 Positive samples from the air and filters as blank squares aligned.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
+IFB: influenza~B, HRV: human rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: human metapneumovirus, PIV: parainfluenza virus.}
     \label{fig:molecular-descriptives}
@@
 
+Absences related to respiratory infections were 22~respiratory cases without and 13~cases with air cleaners (\cref{fig:infection-risk}a).
+Our Bayesian latent variable hierarchical regression model suggested that air cleaners reduced the risk of infection (PP 91\%).
+However, the reduction was small and subject to uncertainty (ARR 0.73, 95\%-CrI 0.44 to 1.18).
+The estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\supp~\zref{fig:avoided-infections}).
+Detailed estimation results are provided in the \supp~text~\zref{sec:detailed-redcap}.
+
 There was an equal number of 25~positive saliva samples in both study conditions (\cref{fig:infection-risk}b).
-Based on our Bayesian Multinomial logistic regression model, the positivity rate remained comparable between study conditions (ARR 0.93, 95\%-CrI 0.49 to 1.61; PP 61\%).
+Based on our Bayesian Multinomial logistic regression model, the positivity rate remained comparable between study conditions (ARR 0.93, 95\%-CrI 0.49 to 1.61; PP 65\%).
 The estimated ARR was not sensitive to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
+Nevertheless, these estimates are subject to unmeasured confounding because the delay may be sample-specific and repeated positive tests from the same person could not be precluded because testing was anonymized.
 
-The aerosol number concentration was 95$\pm$81\,1/cm$^3$ without and 27$\pm$34\,1/cm$^3$ with air cleaners (\cref{fig:infection-risk}a).
+The aerosol number concentration was 95$\pm$81\,1/cm$^3$ without and 27$\pm$34\,1/cm$^3$ with air cleaners (\cref{fig:infection-risk}c).
 The adjusted decrease in the aerosol concentration with air cleaners was 76\% (95\%-CrI 63\% to 86\%; PP 100\%).
@@
 In particular, daily average CO$_2$ levels were 1,769$\pm$391\,ppm with and 1,636$\pm$341\,ppm without air cleaners, suggesting that ventilation was comparable between study conditions.
-
-Absences related to respiratory infections were 22~respiratory cases without and 13~cases with air cleaners (\cref{fig:infection-risk}c).
-Our Bayesian latent variable hierarchical regression model suggested that air cleaners reduced the risk of infection (PP 91\%).
-However, the reduction was small and subject to uncertainty (ARR 0.72, 95\%-CrI 0.44 to 1.15).
-The estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\supp~\zref{fig:avoided-infections}).
-Detailed estimations are provided in the \supp~text~\zref{sec:detailed-redcap}.
 
@@
 \centering
-    \includegraphics{../../results/mol-data/saliva-by-study-condition.pdf}\hspace{.5cm}
-    \includegraphics{../../results/env-data/aerosol-number-boxplot.pdf}
     \includegraphics{../../results/epi-data/cases_by_condition.pdf}\hspace{.5cm}
+    \includegraphics{../../results/mol-data/saliva-by-study-condition.pdf}
+    \includegraphics{../../results/env-data/aerosol-number-boxplot.pdf}\hspace{.5cm}
     \includegraphics{../../results/cough-data/coughs-frequency-by-condition.pdf}
     \caption{\textbf{Risk of infection by study condition}.
-\textbf{(a)}~Number of positive saliva samples.
-\textbf{(b)}~Daily average aerosol number concentrations as boxplots.
-\textbf{(c)}~Number of respiratory cases.
-\textbf{(d)}~Daily average number of detected coughs per minute.
-The posterior probability (PP) for each outcome being lower with air cleaners (Yes<No) was computed based on the posterior samples of each Bayesian model used to estimate the effect of air cleaners after adjusting for confounders.}
+Comparison of raw outcomes with and without air cleaners.
+Atop of each plot the posterior probability (PP) is shown that the outcome was lower with air cleaners (Yes<No), which was computed based on the posterior samples of the Bayesian model used to estimate the effect of air cleaners after adjusting for observed confounders.
+\textbf{(a)}~Number of respiratory cases.
+\textbf{(b)}~Number of positive saliva samples.
+\textbf{(c)}~Daily average aerosol number concentrations as boxplots.
+\textbf{(d)}~Daily average number of detected coughs per minute as boxplots.
+}
     \label{fig:infection-risk}

### 2023-10-19 doc/paper/supplements.tex
Updates from Overleaf

@@
     \caption[Line list of molecular test results over the study period]{Line list of respiratory cases over the study period.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}\label{tab:mol-data-line-list} \\
+IFB: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}\label{tab:mol-data-line-list} \\
     \toprule
@@
     \caption[Choices of priors for the incubation periods]{Choices of priors for the distribution of the virus-specific incubation periods.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
+IFB: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
     \label{fig:prior-pin}
@@
     \caption[Association between cough frequency and virus-specific positivity rate]{Estimated association (relative risk; posterior mean as dot, 50\%, 80\%, and 95\%-CrI as lines) between cough frequency and number of positive saliva test results by respiratory virus.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
+IFB: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
     \label{fig:coughing-association}

### 2023-11-03 doc/paper/supplements.tex
Updates from Overleaf

@@
 \title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
-Transmission of respiratory viral infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}}
+Impact of air cleaners in schools on respiratory viral infections: \\ A modeling study using epidemiological, environmental, and molecular data}}
 
@@
 \author[1]{Simon Bertschinger}
+\author[5]{Ernest Weingartner}
 \author[2,4]{Alban Ramette}
-\author[2,4]{Ronald Dijkman}
 \author[1]{Matthias Egger}
-\author[5]{Tina Hascher}
+\author[6]{Tina Hascher}
 \author[1*,2]{Lukas Fenner}
@@
 \affil[4]{Institute for Infectious Diseases, University of Bern, Bern, Switzerland}
-\affil[5]{Institute of Educational Science, University of Bern, Bern, Switzerland}
+\affil[5]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
+\affil[6]{Institute of Educational Science, University of Bern, Bern, Switzerland}
 
@@
 \newpage
+
+\section{Experimental determination of the clean air delivery rate}\label{sec:cadr}
+
+The Clean Air Delivery Rate (CADR) is typically used as a measure to quantify the cleaning efficiency of air purifiers\cite{Nelson1993IndoorEnv}.
+The CADR is expressed in m$^3$/s and describes the volumetric flow rate of particle-free, clean air that an air purifier delivers into the room.
+The CADR is equal to the product of the filtration efficiency and the volumetric flow rate passing through the device.
+
+We followed the approach by K\"{u}pper et al.\cite{Kuepper20199AAQR} to determine the CADR value of the air purifier.
+It is based on the release of fine aerosols ($D < 1 \mu m$) in a closed, unventilated room and the subsequent measurement of the decay rates $k$ of the particle concentration $N(t)$ caused by the air cleaner, \ie
+\begin{align}
+    N(t) = N_0 \cdot e^{-k\cdot t}
+\end{align}
+Since the surfaces in the room (\eg walls and furniture) are also sinks for aerosol particles, measurements were carried out with the air cleaner turned off ($k_\mathrm{background}$) and on ($k_\mathrm{cleaner}$).
+The CADR can then be derived as
+\begin{align}
+    \mathrm{CADR} = V \cdot k_\mathrm{cleaner} - V \cdot k_\mathrm{background},
+\end{align}
+where $V = 233 m^3$ is the volume of the classroom.
+
+We ensured that the particle concentrations in the experiments were not too high, thereby preventing bias due to coagulation (\ie the formation of a few large particles from many small particles).
+At the beginning of an experiment, fine aerosol particles were released for a few minutes by ultrasonic nebulization of an aqueous NaCl solution of 5\,g/l.
+The generated dehydrated salt particles have a broad size distribution in the range of 0.02-1\,$\mu$m.
+Decay rates were measured using 11~fine dust sensors (SPS30, Sensirion AG), which were placed at head height of the students.
+These sensors measured the number concentration of individual particles by counting the scattered light pulses they produce when passing through a laser beam.
+Because of their scattering properties, these sensors measure the number of particles larger than $\sim$0.3\,$\mu$m.
+
+Figure~\ref{fig:experiment} shows the particle number concentration with and without without the air purifier. The exponential decrease is clearly visible as a linear curve in the logarithmic plot.
+
+\begin{figure}
+    \centering
+    %\includegraphics{}
+    \caption{Measured concentration curves with the air purifier on and off.
+The red dot shows the average concentration measured by the 11 distributed sensors of particles with $0.3\mu m < D < 1 \mu m$.
+The standard deviation is shown in gray as an error bar.
+The increase in concentration at the beginning of these experiments is due to the release of the NaCl particles during approximately 15-20\,min.
+The inhomogeneity of the concentration during the nebulization period is seen in larger error bars.
+Afterwards, the aerosol source is turned off, the concentration homogenizes quickly, and the concentration drop follows an exponential decay, which is used for the CADR calculation.}
+    \label{fig:experiment}
+\end{figure}
+
+Based on these experiments, we were able to determine the net CADR values of the air purifier as 420\,m$^3$/h with an uncertainty of $\pm$10\% due to systematic differences in the measurement setup and measurement errors of the instruments.
+Our estimated CADR is lower than the CADR reported by the manufacturer of 600-700 m³/h.
+The reported CADR was determined with comparably large particles (pollen in the super micrometer size range).
+Thus, we explain the lower filtration efficiency determined here with the fact that CADR is strongly dependent on the particle size, \ie large particles are easier to filter than fine aerosols.
 

### 2023-11-03 doc/paper/manuscript-lancet.tex

@@
 \author[2,4]{Alban Ramette}
-\author[2,4]{Ronald Dijkman}
 \author[1]{Matthias Egger}
@@
 \affil[4]{Institute for Infectious Diseases, University of Bern, Bern, Switzerland}
-\affil[5]{School of Engineering, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
+\affil[5]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
 \affil[6]{Institute of Educational Science, University of Bern, Bern, Switzerland}
@@
 Spatiotemporal analysis of positive saliva samples suggested within-classroom transmission.
-Positive saliva tests were equally likely with air cleaners (adjusted relative risk [ARR] 0.93 95\%-credible interval [CrI] 0.49 to 1.61).
+Positive saliva tests were equally likely with air cleaners than without (adjusted relative risk [ARR] 0.93 95\%-credible interval [CrI] 0.49 to 1.61).
 CO$_2$ levels were similar (1,769$\pm$391\,ppm with vs.
@@
 \noindent\textbf{Conclusions:} A wide range of non-SARS-CoV-2 respiratory viruses were detected in student saliva.
-Although airborne detection was rare, transmission seemed likely in schools, suggesting that prolonged close contact is required for transmission.
+Although airborne detection was rare, transmission seemed likely to be occurring in schools, suggesting that close contact is required for transmission of most respiratory viruses.
 Air cleaners improved air quality, were associated with lower symptom intensity, and showed a potential benefit in reducing infections.
@@
 Studies also showed that air cleaners effectively removed SARS-CoV-2 bioaerosols.
-One study reported lower incidence of SARS-CoV-2 in schools using air filtration devices.
+One study reported a lower incidence of SARS-CoV-2 in schools using different ventilation strategies, with or without additional air filtration devices.
 While simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission, a real-world study during the pandemic was inconclusive.
@@
 
-\noindent Good ventilation improves indoor air quality and can reduce the risk of infection under non-pandemic conditions.
+\noindent Good ventilation can improve indoor air quality and reduce the risk of infection under non-pandemic conditions.
 Greater emphasis should be given to ventilation systems in schools in the long term.
-Seasonal respiratory viruses other than SARS-CoV-2 may require prolonged close contact.
-Future studies should examine the relative importance of prolonged close contact and long-distance transmission of respiratory viruses in schools and other congregate settings.
+Seasonal respiratory viruses other than SARS-CoV-2 may require  close contact for transmission.
+Future studies should examine the relative importance of close contact vs long-distance transmission of respiratory viruses in schools and other congregate settings.
 
@@
 Person-to-person transmission occurs primarily through the release of respiratory particles containing the viruses.
-Recent focus has been on small respiratory particles called aerosols ($\leq$ 5$\mu$m), which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
+Recent focus has been on small respiratory particles called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
 Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and travel long distances\cite{Coleman2022,Wang2020,Heneghan2021}.
@@
 Several studies have shown that air cleaners reduce particle concentrations in indoor air\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed}, which is associated with a reduced risk of respiratory infection and mortality\cite{Gordon2014Resp,Kelly2023Atmos,DeAngelis2021EnvRes}.
-In addition, recent studies showed that air cleaners effectively removed SARS-CoV-2 from the air\cite{Morris2022,Ueki2022mSphere,Myers2022IndoorAir,Rodriguez2021SOTE}.
-Simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission  of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
-One study reported lower incidence of SARS-CoV-2 in US elementary schools using air filtration methods\cite{Gettings2021}.
+In line with this, a population-level study reported a lower incidence of SARS-CoV-2 in US elementary schools using different ventilation strategies, with or without additional air filtration devices\cite{Gettings2021}.
+Although some studies could not find an association between particle concentrations and viral RNA load in airborne samples\cite{Nor2021SciRep,Hanna2023PONE}, recent studies showed that air cleaners effectively removed SARS-CoV-2 from the air\cite{Morris2022,Ueki2022mSphere,Myers2022IndoorAir,Rodriguez2021SOTE}.
+Simulation studies have further demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
+However, most simulation studies assume that the detection of RNA equals transmissible virus, despite recent data showing a relevant loss of viral infectivity in respiratory particles over time\cite{Oswin2022PNAS}.
 To date, it remains unclear whether reducing particle concentrations and removing bioaerosols from the air will reduce indoor transmission of respiratory infections.
 
-We used a multiple-measurement approach to study the transmission of respiratory viruses under non-pandemic conditions and the effect of air cleaners in schools with a cross-over study design in the winter of 2023.
-We collected epidemiological data (absences likely related to respiratory infections), environmental data (CO$_2$, particle concentrations), audio recordings (coughs), and molecular data (detection of viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
+We used a multiple-measurement approach to study the transmission of respiratory viruses under non-pandemic conditions and the effect of air cleaners in schools with a cross-over study design in the winter of 2022/2023.
+We collected epidemiological data (absences likely related to respiratory infections), environmental data (CO$_2$, particle concentrations), audio recordings (coughs), and molecular data (detection of viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2023 in two classes of a secondary school in Switzerland.
 We determined changes in absences related to respiratory infections, positivity rates of saliva samples, particle concentrations, and coughing.
@@
 All devices were placed at the head level of the students when they were seated.
-Neither classrooms was equipped with an active HVAC (Heating, Ventilation, Air conditioning) system, but they were ventilated naturally by opening windows.
+Both classrooms lacked an active HVAC (Heating, Ventilation, Air conditioning) system, but they were ventilated naturally by opening windows.
 }
@@
 According to the manufacturer, these air cleaners run at clean air delivery rates of 2$\,\times\,$600 m$^{3}$/h.
+When testing the devices in an empty classroom with sub-micrometer sized particles, we measured a lower effective clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h (\supp~text~\zref{sec:cadr}).
 Window ventilation was done at the discretion of the teachers, as there were no specific recommendations from the national health authorities at the time.
@@
     \centering
-    \caption{\textbf{Study design.} Cross-over design describing when portable air cleaners (AC) were installed in the rooms of classes A and B during a seven-week study period from January~16 to March~11, 2023, excluding a week of vacation from February 6 to 11.}\label{tab:study_design}
+    \caption{\textbf{Study design.} Cross-over design describing when portable air cleaners (PAC) were installed in the rooms of classes A and B during a seven-week study period from January~16 to March~11, 2023, excluding a week of vacation from February 6 to 11.}\label{tab:study_design}
     \begin{tabular}{l l l l l l l l l}
@@
       \midrule
-      \textbf{A} & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{0000}*& \cellcolor{gray!10} None & & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{00009}* \\
-      \textbf{B} & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC\hphantom{0000000}* & & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{0000}*& \cellcolor{gray!10} None & \cellcolor{gray!10} None \\
+      \textbf{A} & \cellcolor{gray!50} PAC & \cellcolor{gray!50} PAC\hphantom{000}*& \cellcolor{gray!10} None & & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} PAC & \cellcolor{gray!50} PAC\hphantom{0000}* \\
+      \textbf{B} & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} PAC\hphantom{00000}* & & \cellcolor{gray!50} PAC & \cellcolor{gray!50} PAC\hphantom{000}*& \cellcolor{gray!10} None & \cellcolor{gray!10} None \\
       \bottomrule
@@
 
-\noindent Both classes participated in repetitive bi-weekly testing (Tuesdays and Thursdays; see \nameref{sec:mol_analyses} below).
-Saliva samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Galar2021,To2019,Huber2021}.
+\noindent Both classes participated in repetitive bi-weekly saliva testing (Tuesdays and Thursdays; see \nameref{sec:mol_analyses} below).
+Samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Galar2021,To2019,Huber2021}.
 The line list data is shown in \supp~table~\zref{tab:mol-data-line-list}.
-Furthermore, We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL Phosphate-Buffered Saline (PBS).
+Furthermore, we collected airborne respiratory viruses in both classrooms with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL Phosphate-Buffered Saline (PBS).
 The Coriolis Micro Air ran shortly before and during break times (approximately 60\,min/day) to prevent unacceptable noise exposure.
-In one class, we also used the BioSpot-VIVAS device sampler (Aerosol Devices Inc., Ft.
+In one class, we also sampled with the BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft.
 Collins, CO, USA) \cite{Pan2016JAM,Lednicky2016AST}.
@@
 The HEPA filters were removed from the air cleaners and divided into 20~fields.
-One PBS-moistened swab per field was then taken for a total of 20~swabs per filter.
+One sterile PBS-moistened swab per field was then taken for a total of 20~swabs per filter.
 \medskip
@@
 We performed molecular genotyping for positive saliva, bioaerosol, air filter samples of AdV and IF.
-For AdV, we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}.
+For AdV, we used a PCR procedure to amplify three hypervariable regions (hexon, fiber, penton) of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}.
 For IF, we tried to determine the molecular transmission network using whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
@@
 We determined the number of coughs per minute using an AI algorithm\cite{Bertschinger2023IEEE}.
+A recent study has shown a significant correlation between coughing and the positivity rate of virus detection in airborne samples\cite{Hanna2023PONE}.
 
@@
 We analyzed a total of 448~saliva samples.
-We detected 15~IF~B~(IFB), 15~HRV, 14~AdV, 3~CoV, 2~MPV, and 1~PIV (\cref{fig:molecular-descriptives}a).
+We detected 15~influenza B~(IFB), 15~rhinovirus~(HRV), 14~adenovirus~(AdV), 3~SARS-CoV-2~(CoV), 2~metapneumovirus~(MPV), and 1~parainfluenza virus~(PIV), respectively (\cref{fig:molecular-descriptives}a).
 The distribution of positive saliva samples varied between classes.
@@
 Furthermore, the rate of positive bioaerosol samples was low.
-We analyzed 105~bioaerosol samples and detected only two respiratory viruses in the air (1~HRV in class~A and 1~AdV in class~B).
+We analyzed 105~bioaerosol samples and detected only in two of them viral RNA (1~HRV in class~A and 1~AdV in class~B).
 Similarly, we detected only 1~IFB, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
@@
 
-We used a multiple-measurement approach to estimate the risk of respiratory virus infection in a Swiss school and assessed the effectiveness of air cleaners.
-We found a wide range of respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus, but very few respiratory viruses were detected in bioaerosol samples and on the filters of the air cleaners.
-Aerosol number and particle mass concentrations decreased significantly with air cleaners and Bayesian modeling based on epidemiological data showed a potential reduction in the relative risk of infection with air cleaners.
-Reduced coughing could indicate that air cleaners prevent some symptomatic infections.
+We used a multiple-measurement approach to estimate the risk of respiratory virus infection in a Swiss school and to assess the effectiveness of air cleaners.
+We found a wide range of respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus, but very few viral RNA were detected in bioaerosol samples and on the filters of the air cleaners.
+Aerosol number and particle mass concentrations decreased significantly with air cleaners, and Bayesian modeling based on epidemiological data showed a potential reduction in the relative risk of infection with air cleaners.
+Reduced coughing may indicate that air cleaners prevented symptomatic infections.
 
@@
 
-In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during winter 2022.
-Our previous study was conducted during the SARS-CoV-2 Omicron wave and most detected viruses were SARS-CoV-2.
-We found a detectable effect on the risk of SARS-CoV-2 infection only for mask wearing, but not air cleaners.
+We estimated the effectiveness of air cleaners during the winter of 2022/2023 in a cross-over design.
+We detected a range of respiratory virus in students' saliva, mainly adenovirus, influenza~B and rhinovirus, with only three positive SARS-CoV-2 saliva samples.
+A similar shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.
+We found a potential benefit of air cleaners on the risk of respiratory infection.
+In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 omicron wave in the winter of 2021/2022.
+At that time, we detected almost exclusively SARS-CoV-2 in the students' saliva.
+We found a detectable effect on the risk of SARS-CoV-2 infection only for mask wearing, but not for air cleaners.
 We argued that air cleaners may have had less potential to reduce infections because they were introduced at the end of the study, when most students were already infected with SARS-CoV-2.
-In this study, we used a cross-over design and estimated the effectiveness of air cleaners during winter 2023.
-This time, we detected a range of respiratory viruses, which consisted primarily of adenovirus, influenza~B and rhinovirus, with only three positive SARS-CoV-2 saliva samples.
-A similar shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.
-In contrast to our previous study, we found a small, potential benefit of air cleaners on the risk of respiratory infection.
 
@@
 
-There may be several reasons for the relatively small effect of air cleaners.
-Compared to \eg universal mask wearing, which has been shown to be effective\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Leung2020NatMed,Milton2013PLoSPathogens}, air cleaners cannot affect transmission outside the classroom.
-Furthermore, previous work suggests that prolonged close contact may be necessary for transmission\cite{Leung2020NatMed,Brankston2007LancetID}, or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
-However, compared to face masks, air cleaners are not able to prevent transmission by close range, high particle density.
-Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective at removing larger particles ($>5\mu$m).
+We showed in a real-world setting that air cleaners can reduce the risk of infection in schools and that they also, consistent with previous  studies\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed}, improve indoor air quality.
+Although the risk reduction effect estimated in this study is rather small, the cumulative impact may be relevant because minimizing the risk of transmission of respiratory infections in schools is an important public health consideration\cite{Beale2023JOMT} and because prevention and control measures in schools, such as air cleaners, must be adequate and acceptable\cite{WHO2020SchoolMeasures}.
+There are several reasons why the effect of air cleaners is likely smaller compared to \eg universal mask wearing, which has been shown to be a very effective infection control measure\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Leung2020NatMed,Milton2013PLoSPathogens}.
+Unlike masks, air cleaners cannot prevent transmission outside the classroom and, importantly, transmission at close range, with a high particle density.
+Prolonged close contact may be necessary for transmission\cite{Leung2020NatMed,Brankston2007LancetID} or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
+Evidence from this and our previous study\cite{Banholzer2023PLoSMed} further suggests that air cleaners are more effective at removing larger particles ($>5\mu$m), which also explains the difference between our measured and the manufacturer's reported clean air delivery rate (see \supp~text~\zref{sec:cadr}).
 However, a greater proportion of respiratory viruses are carried in the smaller particles, which are also more relevant for transmission ($\leq5\mu$m)\cite{Fennelly2020}.
-Finally, classroom activity, airflow and other unobserved, confounding factors make it difficult to evaluate the effects of air cleaners.
-Nevertheless, it is important to assess their effectiveness in a real-world setting and to compare them with the hypothetical impact suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
+Finally, classroom activity, airflow and other unobserved, confounding factors make it difficult to evaluate the effects of air cleaners on transmission.
+Nevertheless, it is important to evaluate their effectiveness in a real-world setting and to compare them with the hypothetical effects suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
 
 % economic and psychological impact of air cleaners
-% tbd
-% psychological part from Tina
+
+The beneficial effects of air cleaners on indoor air quality and transmission come at a reasonable cost.
+The portable air cleaners used in our study cost approximately USD\,250 per unit.
+Their operating cost effectiveness in providing clean air could be even higher than that of a ventilation system when compared in parallel using the same air delivery ratings\cite{Noh2016EnBuild}.
+Therefore, air cleaners could be a cost-effective public health measure, particularly during pandemics or epidemics when there is greater exposure to respiratory infections.
+In addition, air cleaners can have a psychological impact.
+On the one hand, they can give students and teachers a sense of support during difficult times.
+For example, it has been shown that getting infected was one of the main worries for students during the pandemic\cite{Yuerekli2022IJERPH}.
+A safe and supportive environment is a key determinant for students' success and well-being in school\cite{Kutsyuruba2015RevEduc}.
+On the other hand, air cleaners may not be well accepted by teachers due to noise pollution, limited space, technical issues, and maintenance requirements \cite{Sanguinetti2022IndoorAir}.
+Therefore, investments in professional building ventilation systems are still preferred in the long term\cite{Nardell2016}.
 
@@
 
-We detected only few respiratory viruses in bioaerosol samples (1~sample of adenovirus and 1~sample of rhinovirus ) and on the filters of the air cleaners (4~positive samples in 160~swabs) in this study.
-The low rate of positive bioaerosol samples means that it is unclear to what extent airborne transmission occurred in classrooms.
-It may be possible that students had relatively little exposure to respiratory viruses in schools and acquired their infections elsewhere.
+We detected only few respiratory viruses in bioaerosol samples (1~sample of adenovirus and 1~sample of rhinovirus) and on the filters of the air cleaners (4~positive samples in 160~swabs) in this study.
+The low rate of positive bioaerosol samples may indicate that it is unlikely that airborne transmission occurred to a considerable extent in the classrooms, \ie it is possible that students had relatively little exposure to respiratory viruses at school and acquired their infections elsewhere.
 However, the distribution of saliva samples between classes shows remarkable differences.
@@
 Our study has several limitations.
-First, aerosol measurements and molecular detection of viruses in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (person to air, air to person) cannot be determined.
+First, aerosol measurements and molecular detection of virus in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (person to air, air to person) cannot be determined.
 However, the virus-specific distribution of positive saliva samples between classes (spatiotemporal pattern) may indicate transmission within classrooms.
-Second, the specific causes for the absences could not be determined from the epidemiological data.
-Therefore, some absences may have falsely been attributed to respiratory infections and we could only approximate the incubation period for each epidemiological case.
+Second, the specific causes of school absences could not accurately be determined from the epidemiological data.
+Therefore, some absences may have been incorrectly attributed to respiratory infections, and we could only approximate the incubation period for each epidemiological case.
 Third, the effectiveness of air cleaners was somewhat limited because students occasionally had lessons outside the classroom.
@@
 In conclusion, using epidemiological, environmental, and molecular data, our study showed that a wide range of non-SARS-CoV-2, but rarely SARS-CoV-2, was detected in students under non-pandemic conditions when public health measures were lifted, and that transmission of these respiratory viruses likely occurred in schools.
-Airborne detection of non-SARS-CoV-2 respiratory viruses was rare, suggesting that prolonged close contact may be required for transmission.
-Comparison with previous work during the pandemic, which showed frequent airborne SARS-CoV-2, suggests that respiratory viruses other than SARS-CoV-2 may be less detectable in the air.
+Airborne detection of non-SARS-CoV-2 respiratory viruses was rare, suggesting that that respiratory viruses other than SARS-CoV-2 may be less detectable in the air and that close contact may be required for transmission in most cases.
 Air cleaners improved air quality by significantly reducing aerosol and particle concentrations.
 Based on our transmission model using epidemiological data, the risk of infection tends to be lower with air filters.
-Future studies should investigate the relative importance of prolonged close contact and long-range transmission of respiratory viruses in schools and other congregate settings.
+Future studies should further investigate the relative importance of prolonged close contact versus long-range transmission of respiratory viruses in schools and other congregate settings.
 
@@
 \section*{Funding}
-This study is funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
+This study was funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
 NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
@@
 Conception and design: NB, KZ, LF, PB, PJ, TS.
-Epidemiological and environmental data collection: NB, PJ, KZ, TS, LF.
+Epidemiological and environmental data collection: EW, NB, PJ, KZ, TS, LF.
 Laboratory data collection: PB, LFu.
@@
 Cough detection: SB.
-Molecular genotyping: RD, AR, PB, LFu.
+Molecular genotyping: AR, PB, LFu.
 Statistical analysis: NB.
@@
 We are grateful to the Educational Department of the canton of Solothurn for their support throughout the study.
-We also would like to thank the biosecurity team at the Institute for Infectious Diseases of the University of Bern for their assistance with the bioaerosol devices (Julia Feldmann, Monika Gsell, Kathrina Summermatter).
+We would also like to thank Ronald Dijkman for the influenza molecular genotyping and the biosecurity team (Julia Feldmann, Monika Gsell, Kathrina Summermatter) from the Institute for Infectious Diseases at the University of Bern for their assistance with the bioaerosol devices.
 Finally, we are indebted to the student assistants (Khadija Ahmed, Marie-Joséphine Brancato, Michelle Bürki, Santiago Martinez, Moric Toszeghi, Sylvain Wasmer) who helped with the data collection in the schools.

### 2023-11-03 doc/paper/supplements.tex
Updates from Overleaf

@@
 
-\section{Experimental determination of the clean air delivery rate}\label{sec:cadr}
+\section{Experimental determination of the CADR}\label{sec:cadr}
 
-The Clean Air Delivery Rate (CADR) is typically used as a measure to quantify the cleaning efficiency of air purifiers\cite{Nelson1993IndoorEnv}.
-The CADR is expressed in m$^3$/s and describes the volumetric flow rate of particle-free, clean air that an air purifier delivers into the room.
+The clean air delivery rate (CADR) is typically used to quantify the cleaning efficiency of air purifiers\cite{Nelson1993IndoorEnv}.
+It is expressed in m$^3$/s and describes the volumetric flow rate of particle-free, clean air that an air purifier delivers into the room.
 The CADR is equal to the product of the filtration efficiency and the volumetric flow rate passing through the device.
 
-We followed the approach by K\"{u}pper et al.\cite{Kuepper20199AAQR} to determine the CADR value of the air purifier.
-It is based on the release of fine aerosols ($D < 1 \mu m$) in a closed, unventilated room and the subsequent measurement of the decay rates $k$ of the particle concentration $N(t)$ caused by the air cleaner, \ie
+We followed the approach by K\"{u}pper et al.\cite{Kuepper20199AAQR} to determine the CADR of the air purifiers used in this study.
+The approach is based on the release of fine aerosols (particle diameter $D$<1$\mu$m) in a closed, unventilated room by measuring the decay rate $k$ of the particle concentration $N(t)$ caused by the air cleaner, \ie
 \begin{align}
@@
 \end{align}
-Since the surfaces in the room (\eg walls and furniture) are also sinks for aerosol particles, measurements were carried out with the air cleaner turned off ($k_\mathrm{background}$) and on ($k_\mathrm{cleaner}$).
-The CADR can then be derived as
+Since the surfaces in the room (\eg walls and furniture) are also sinks for aerosol particles, measurements were carried out with ($k_\mathrm{cleaner}$) and without air cleaners($k_\mathrm{background}$).
+Based on that, we can compute the CADR as
 \begin{align}
@@
 We ensured that the particle concentrations in the experiments were not too high, thereby preventing bias due to coagulation (\ie the formation of a few large particles from many small particles).
-At the beginning of an experiment, fine aerosol particles were released for a few minutes by ultrasonic nebulization of an aqueous NaCl solution of 5\,g/l.
-The generated dehydrated salt particles have a broad size distribution in the range of 0.02-1\,$\mu$m.
+At the beginning of each experiment, fine aerosol particles were released for a few minutes by ultrasonic nebulization of an aqueous NaCl solution of 5\,g/l.
+The generated dehydrated salt particles have a broad size distribution in the range of 0.02$-$1.0$\mu$m.
 Decay rates were measured using 11~fine dust sensors (SPS30, Sensirion AG), which were placed at head height of the students.
@@
 
-Figure~\ref{fig:experiment} shows the particle number concentration with and without without the air purifier. The exponential decrease is clearly visible as a linear curve in the logarithmic plot.
+Figure~\ref{fig:experiment} shows the particle number concentration with and without without the air purifier.
+The exponential decrease is clearly visible as a linear curve in the logarithmic plot.
+Based on these experiments, we were able to determine the net CADR values of the air purifier as 420\,m$^3$/h with an uncertainty of $\pm$10\% due to systematic differences in the setup and measurement errors of the instruments.
+Our estimated CADR is lower than the CADR reported by the manufacturer of 600m$^3$/h, most likely because the reported CADR was determined with comparably large particles (pollen in the super micrometer size range).
+Thus, we explain the lower filtration efficiency determined here with the fact that CADR is strongly dependent on the particle size, \ie large particles are easier to filter than fine aerosols.
 
@@
     \centering
-    %\includegraphics{}
+    \includegraphics[width=\linewidth]{doc/paper/experiment-particle-concentration-decay.png}
     \caption{Measured concentration curves with the air purifier on and off.
@@
 
-Based on these experiments, we were able to determine the net CADR values of the air purifier as 420\,m$^3$/h with an uncertainty of $\pm$10\% due to systematic differences in the measurement setup and measurement errors of the instruments.
-Our estimated CADR is lower than the CADR reported by the manufacturer of 600-700 m³/h.
-The reported CADR was determined with comparably large particles (pollen in the super micrometer size range).
-Thus, we explain the lower filtration efficiency determined here with the fact that CADR is strongly dependent on the particle size, \ie large particles are easier to filter than fine aerosols.
+\clearpage
 

### 2023-11-06 analysis/aud_analysis.Rmd
change order of figure

@@
   geom_boxplot2(width.errorbar = .3, width = .4) +
-  ggsignif::geom_signif(annotations = "PP(Yes<No) = 93%", y_position = 6.25, xmin = 1, xmax = 2, 
+  ggsignif::geom_signif(annotations = "Posterior prob. = 93%", y_position = 6.25, xmin = 1, xmax = 2, 
                         textsize = 8 / cm(1), vjust = -.5) +
@@
   scale_y_continuous(limits = c(0, NA), expand = expansion(add = c(0, .66))) +
-  labs(title = "d", y = "Number of coughs per min", x = "Air cleaner") +
+  labs(title = "c", y = "Number of coughs per min", x = "Air cleaner") +
   theme_custom() +

### 2023-11-06 analysis/env_analysis.Rmd
change order of figure

@@
   geom_boxplot2(width.errorbar = .3, width = .4) +
-  ggsignif::geom_signif(annotations = "PP(Yes<No) = 100%", y_position = 240, xmin = 1, xmax = 2, 
+  ggsignif::geom_signif(annotations = "Posterior prob. = 100%", y_position = 240, xmin = 1, xmax = 2, 
                         textsize = 8 / cm(1), vjust = -.5) +
@@
   scale_y_continuous(expand = expansion(add = c(0, 30)), limits = c(0, NA)) +
-  labs(title = "c", y = expression("Aerosol concentration (1/cm"^3*")"), x = "Air cleaner") +
+  labs(title = "a", y = expression("Aerosol concentration (1/cm"^3*")"), x = "Air cleaner") +
   theme_custom() +

### 2023-11-06 analysis/epi_analysis.Rmd
change order of figure

@@
   geom_text(mapping = aes(label = cases), size = 8 / cm(1), vjust = -1) +
-  ggsignif::geom_signif(annotations = "PP(Yes<No) = 91%", y_position = 27, xmin = 1, xmax = 2, 
+  ggsignif::geom_signif(annotations = "Posterior prob. = 91%", y_position = 27, xmin = 1, xmax = 2, 
                         textsize = 8 / cm(1), vjust = -.5, tip_length = .075) +

### 2023-11-06 analysis/mol_analysis.Rmd
change order of figure

@@
   geom_segment(data = data.frame(x = .5, xend = .5, y = 30, yend = 29), mapping = aes(x = x, xend = xend, y = y, yend = yend), color = "white") +
-  ggsignif::geom_signif(annotations = "PP(Yes<No) = 65%", y_position = 30, xmin = 1, xmax = 2, 
+  ggsignif::geom_signif(annotations = "Posterior prob. = 65%", y_position = 30, xmin = 1, xmax = 2, 
                         textsize = 8 / cm(1), vjust = -.5, tip_length = .1) +
-  labs(x = "Air cleaner", y = "Number of positive saliva samples", title = "b") +
+  labs(x = "Air cleaner", y = "Number of positive saliva samples", title = "d") +
   scale_fill_manual(values = bluegrey) +

### 2023-11-06 doc/paper/manuscript-lancet.tex
Updates from Overleaf

@@
 
-\title{\bfseries\LARGE\singlespacing{Impact of air cleaners in schools on respiratory viral infections: A modeling study using epidemiological, environmental, and molecular data}}
+\title{\bfseries\LARGE\singlespacing{Air cleaners and respiratory infections in schools: A modeling study using epidemiological, environmental, and molecular data}}
 
@@
 \author[2,4]{Alban Ramette}
-\author[1]{Matthias Egger}
-\author[6]{Tina Hascher}
+\author[1,6,7]{Matthias Egger}
+\author[8]{Tina Hascher}
 \author[1*,2]{Lukas Fenner}
@@
 \affil[5]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
-\affil[6]{Institute of Educational Science, University of Bern, Bern, Switzerland}
+\affil[6]{Population Health Sciences, University of Bristol, Bristol, UK}
+\affil[7]{Centre for Infectious Disease Epidemiology and Research, University of Cape Town, Cape Town, South Africa}
+\affil[8]{Institute of Educational Science, University of Bern, Bern, Switzerland}
 
@@
 \begin{abstract}\normalfont
-\noindent\textbf{Background:} Infection control measures can reduce airborne transmission.
-Using a multiple-measurement approach, we aimed to evaluate the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing transmission of respiratory viral infections in schools.
-\medskip
+\noindent\textbf{Background:} Infection control measures can reduce airborne transmission of respiratory infections.
+Using a multiple-measurement approach, we examined the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in a school setting.\medskip
 
-\noindent\textbf{Methods:} We collected epidemiological data (absences related to respiratory infections), environmental data (CO$_2$ and particle concentrations), audio recordings (coughs), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2023 in two classes of a Swiss secondary school.
-Using a cross-over study design, we compared particle concentrations, coughing, and the risk of infection with and without air cleaners.
-\medskip
+\noindent\textbf{Methods:} We collected epidemiological data (absences related to respiratory infections), environmental data (CO$_2$ and particle concentrations), audio recordings (coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2022/2023 in two Swiss secondary school classes.
+Using a cross-over study design, we compared particle concentrations, coughing, and the risk of infection with vs without air cleaners.\medskip
 
-\noindent\textbf{Results:} Molecular analysis detected several viruses in saliva (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza), but rarely in bioaerosol samples (2/105 positive) and on filters of air cleaners (4/160~positive).
-Spatiotemporal analysis of positive saliva samples suggested within-classroom transmission.
-Positive saliva tests were equally likely with air cleaners than without (adjusted relative risk [ARR] 0.93 95\%-credible interval [CrI] 0.49 to 1.61).
-CO$_2$ levels were similar (1,769$\pm$391\,ppm with vs.
-1,636$\pm$341\,ppm without air cleaners), but particle concentrations decreased by 77\%, (95\%-CrI 63\% to 86\%) and coughing was slightly less frequent (ARR 0.93, 95\%-CrI 0.85 to 1.02).
-Infection risk was also lower (posterior probability 91\%), but the association was small (ARR 0.73, 95\%-CrI 0.44 to 1.18).
-\medskip
+\noindent\textbf{Findings:} All 38~students (age 13$-$15 years) participated.
+With air cleaners, mean particle concentration decreased by 77\% (95\% credible interval 63\%$-$86\%).
+There were no differences in CO$_2$ levels.
+Absences related to respiratory infections were 13 with vs 22 without air cleaners, and the risk of infection was slightly reduced (relative risk 0.73, 95\% credible interval 0.44$-$1.18), with a 91\% posterior probability for a reduction.
+Coughing also tended to be less frequent (posterior probability 93\%).
+Molecular analysis detected mainly non-SARS-CoV-2 viruses in saliva (50/448 positive).
+The detection rate was similar with vs without air cleaners.
+Positive samples in bioaerosols (2/105 positive) and HEPA-filters (4/160~positive) were rare, but spatiotemporal analysis of saliva samples suggested within-classroom transmission.\medskip
 
-\noindent\textbf{Conclusions:} A wide range of non-SARS-CoV-2 respiratory viruses were detected in student saliva.
-Although airborne detection was rare, transmission seemed likely to be occurring in schools, suggesting that close contact is required for transmission of most respiratory viruses.
-Air cleaners improved air quality, were associated with lower symptom intensity, and showed a potential benefit in reducing infections.
+\noindent\textbf{Interpretation:} Air cleaners improved air quality, were associated with less coughing, and showed a potential benefit in reducing respiratory infections.
+Future studies should examine their cost-effectiveness.
+Airborne detection of non-SARS-CoV-2 viruses was rare, necessitating further research on the importance of close contact vs long-range airborne transmission.\medskip
+
+\noindent\textbf{Funding:} This study was funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
+NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
+ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498).
 
@@
 One study reported a lower incidence of SARS-CoV-2 in schools using different ventilation strategies, with or without additional air filtration devices.
-While simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission, a real-world study during the pandemic was inconclusive.
+Several simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission.
+One real-world study during the COVID-19 pandemic was inconclusive regarding the effectiveness of air cleaners in a school setting.
 \medskip
@@
 This study used a multiple-measurement approach to estimate the risk of respiratory virus infection in a Swiss school and assessed the effectiveness of air cleaners in a cross-over study design.
-Schools are particularly important because students spend much of their time indoors and may introduce infections from and into their household.
-We collected molecular, epidemiological, environmental, and audio data over seven weeks during non-pandemic conditions in winter 2023 in a Swiss school.
-We found a wide range of respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus.
-Although airborne detection was rare, spatiotemporal analysis of students' saliva samples indicated that transmission seemed likely in school rooms, suggesting that prolonged close contact is required for transmission.
-Air cleaners improved air quality (lower particle concentrations), were associated with lower symptom intensity (fewer coughs), and showed a potential benefit in reducing cases (fewer absences related to respiratory infections).\medskip
+Schools are particularly important because students spend much of their time indoors and may introduce infections from and into their households.
+We collected molecular, epidemiological, environmental, and audio data over seven weeks during non-pandemic conditions in winter 2022/2023 in a Swiss school.
+Air cleaners in classrooms improved air quality (lower particle concentrations), were associated with fewer symptoms (less coughing), and showed a potential benefit in reducing absences related to respiratory infections.
+We found many respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus.
+Although airborne detection of viruses was rare, spatiotemporal analysis of students' saliva samples indicated that transmission was likely in school rooms.
+\medskip
 
@@
 
-\noindent Good ventilation can improve indoor air quality and reduce the risk of infection under non-pandemic conditions.
-Greater emphasis should be given to ventilation systems in schools in the long term.
-Seasonal respiratory viruses other than SARS-CoV-2 may require  close contact for transmission.
-Future studies should examine the relative importance of close contact vs long-distance transmission of respiratory viruses in schools and other congregate settings.
+\noindent Air cleaners have a tangible positive impact on air quality, significantly reducing particle concentrations.
+While the individual-level reduction in the risk of respiratory infections due to air cleaners may be relatively modest, the collective benefit at the population level, in terms of reducing illnesses and preventing school absences, is likely to be substantial.
+Future research should examine the cost-effectiveness and feasibility of implementing air cleaners in school and other communal settings.
 
@@
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 Person-to-person transmission occurs primarily through the release of respiratory particles containing the viruses.
-Recent focus has been on small respiratory particles called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
+Recently, the focus has been on small respiratory particles called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
 Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and travel long distances\cite{Coleman2022,Wang2020,Heneghan2021}.
@@
 In line with this, a population-level study reported a lower incidence of SARS-CoV-2 in US elementary schools using different ventilation strategies, with or without additional air filtration devices\cite{Gettings2021}.
-Although some studies could not find an association between particle concentrations and viral RNA load in airborne samples\cite{Nor2021SciRep,Hanna2023PONE}, recent studies showed that air cleaners effectively removed SARS-CoV-2 from the air\cite{Morris2022,Ueki2022mSphere,Myers2022IndoorAir,Rodriguez2021SOTE}.
+Although some studies could not find an association between particle concentrations and viral RNA load in airborne samples\cite{Nor2021SciRep,Hanna2023PONE}, recent studies showed that air cleaners effectively removed SARS-CoV-2 from the indoor air\cite{Morris2022,Ueki2022mSphere,Myers2022IndoorAir,Rodriguez2021SOTE}.
 Simulation studies have further demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
 However, most simulation studies assume that the detection of RNA equals transmissible virus, despite recent data showing a relevant loss of viral infectivity in respiratory particles over time\cite{Oswin2022PNAS}.
-To date, it remains unclear whether reducing particle concentrations and removing bioaerosols from the air will reduce indoor transmission of respiratory infections.
+To date, it remains unclear whether reducing particle concentrations and removing bioaerosols will reduce indoor transmission of respiratory infections.
 
@@
 
-\section{Methods}
+\section*{Methods}
 
-\subsection{Study setting and design}
+\subsection*{Study setting and design}
 
 \noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, for seven weeks from January~16 to March~11, 2023.
-Figure~\ref{fig:study-setup} shows the schematic study setup, and \supp~table~\zref{tab:data} provides an overview of the types of data collected.
+Figure~\ref{fig:study-setup} shows the schematic study setup.
 
@@
 
-\subsection{Study intervention}
+\subsection*{Study intervention}
 
@@
 When testing the devices in an empty classroom with sub-micrometer sized particles, we measured a lower effective clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h (\supp~text~\zref{sec:cadr}).
-Window ventilation was done at the discretion of the teachers, as there were no specific recommendations from the national health authorities at the time.
+Window ventilation was done at the discretion of the teachers.
 
@@
     \centering
-    \caption{\textbf{Study design.} Cross-over design describing when portable air cleaners (PAC) were installed in the rooms of classes A and B during a seven-week study period from January~16 to March~11, 2023, excluding a week of vacation from February 6 to 11.}\label{tab:study_design}
+    \caption{\textbf{Cross-over study design.} Description of when portable air cleaners (PAC) were installed in the rooms of classes A and B during a seven-week study period from January~16 to March~11, 2023, excluding a week of vacation from February 6 to 11.}\label{tab:study_design}
     \begin{tabular}{l l l l l l l l l}
@@
 
-\subsection{Data collection}
+\subsection*{Data collection}
+
+An overview of the collected data types is provided in \supp~table~\zref{tab:data}. \medskip
 
@@
 We defined a case of respiratory infection as an absence in which the student reported an illness with at least one respiratory symptom (\supp~text~\zref{sec:case-data}).
-The line list data is shown in \supp~table~\zref{tab:epi-data-line-list}.
+The list of absences by date of symptom onset, absence and return is shown in \supp~table~\zref{tab:epi-data-line-list}.
 \medskip
@@
 Samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Galar2021,To2019,Huber2021}.
-The line list data is shown in \supp~table~\zref{tab:mol-data-line-list}.
+The list of positive samples is shown in \supp~table~\zref{tab:mol-data-line-list}.
 Furthermore, we collected airborne respiratory viruses in both classrooms with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL Phosphate-Buffered Saline (PBS).
-The Coriolis Micro Air ran shortly before and during break times (approximately 60\,min/day) to prevent unacceptable noise exposure.
+The Coriolis Micro Air ran shortly before and during break times (approximately 60\,min/day) to minimize noise.
 In one class, we also sampled with the BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft.
@@
 This device was used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}.
+The device did \emph{not} alert teachers when CO$_2$ levels or particle concentrations were too high.
 
-\subsection{Laboratory and molecular analyses}\label{sec:mol_analyses}
+\subsection*{Laboratory and molecular analyses}\label{sec:mol_analyses}
 
@@
 
-\subsection{Molecular genotyping}
+\subsection*{Molecular genotyping}
 
-We performed molecular genotyping for positive saliva, bioaerosol, air filter samples of AdV and IF.
-For AdV, we used a PCR procedure to amplify three hypervariable regions (hexon, fiber, penton) of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}.
-For IF, we tried to determine the molecular transmission network using whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
+We performed molecular genotyping for positive saliva, bioaerosol, air filter samples of adenovirus and influenza.
+For adenovirus, we used a PCR procedure to amplify three hypervariable regions (hexon, fiber, penton) of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}.
+For influenza, we tried to determine the molecular transmission network using whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
 
-\subsection{Cough detection}
+\subsection*{Cough detection}
 
@@
 
-\subsection{Statistical analyses and modeling}
+\subsection*{Statistical analyses and modeling}
 
-All statistical analyses were described in the statistical analysis plan (SAP)\cite{Banholzer2023SAP}, which was published prior to the analyses.
+All statistical analyses were described in a statistical analysis plan (SAP)\cite{Banholzer2023SAP}, which was published prior to the analyses.
 Contrary to the SAP, bioaerosol samples and viral load concentrations could not be analyzed because there were too few positive samples.
@@
 The expected number of new cases is the weighted sum of the number of new infections $I_{js}$ (unobserved/latent outcome) in the previous days $s<t$, with the weights corresponding to the probability distribution of the incubation period.
-The number of new infections is related to the presence of air cleaners using a regression-like log-link
+The number of new infections is related to the presence of air cleaners as follows
 \begin{align}
@@
 where $F_{js}$ is the number of infections in the previous week (a proxy for the number of infectious students), $N_{js}$ is the cumulative number of infections (a proxy for the number of susceptible students), $\beta_0$ is the infection rate without air cleaners, and $\beta_1$ is the effect of air cleaners.
-Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the community (\ie the canton of Solothurn).
+Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the canton.
 A detailed description of the model and choice of priors for all model parameters are provided in \supp~text~\zref{sec:transmission-model}.
@@
 
-\noindent\textbf{Number of positive saliva samples} \smallskip
+\noindent\textbf{Saliva samples} \smallskip
 
@@
 
-\noindent\textbf{Aerosol and particle concentrations} \smallskip
+\noindent\textbf{Particle concentrations} \smallskip
 
-\noindent We computed the average particle concentrations per day and compared them between study conditions.
+\noindent We computed mean particle concentrations per day and compared them between study conditions.
 We estimated the reduction in particle concentrations using Bayesian log-linear regression models, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases (\supp~text~\zref{sec:env-regression-model}).
@@
 
-\noindent\textbf{Number of coughs} \smallskip
+\noindent\textbf{Coughing} \smallskip
 
@@
 
-\noindent\textbf{Software} \smallskip
+\noindent\textbf{Software}\smallskip
 
@@
 Model parameters were estimated with a Bayesian approach, using the Hamiltonian Monte Carlo algorithm with the No-U-Turn Sampler (NUTS)\cite{Hoffman2014}.
-If not stated otherwise, we report for each outcome the posterior probability (PP) that the adjusted relative risk (ARR) was lower with air cleaners (ARR<1).
-ARR is reported with mean and 95\%credible interval (CrI) based on 4,000 samples from the posterior distribution of each model.
+For each outcome, we report the posterior probability of a reduction with air cleaners.
+The estimated reduction is reported with the posterior mean and 95\%credible interval (CrI).
 The code is available from \url{https://osf.io/38j9g}.
 
-\subsection{Ethics statement}
+\subsection*{Ethics statement}
 
@@
 
-\section{Results}
+\section*{Results}
 
@@
 
-\subsection{Molecular analyses of saliva and air samples}
+\subsection*{Air quality}
+
+The mean aerosol number concentration was 95\,1/cm$^3$ (standard deviation [SD] 81\,1/cm$^3$) without and 27\,1/cm$^3$ (SD 34\,1/cm$^3$) with air cleaners (\cref{fig:infection-risk}a).
+The Bayesian regression model suggested a clear reduction in the aerosol concentration with air cleaners, with a posterior probability of 100\%.
+The estimated decrease was 76\% (95\%-CrI 63\% to 86\%), which was greater for larger than for smaller particles (\supp~figure~\zref{fig:palas-results}).
+There was little change in other environmental variables (\supp~figure~\zref{fig:env-descriptives-other-vars}).
+In particular, daily average CO$_2$ levels were 1,769\,ppm (SD 391\,ppm) with and 1,636\,ppm (SD 341\,ppm) without air cleaners, suggesting that ventilation was comparable between study conditions.
+
+\begin{figure}[!htpb]
+\centering
+    \includegraphics{../../results/env-data/aerosol-number-boxplot.pdf}\hspace{.5cm}
+    \includegraphics{../../results/mol-data/saliva-by-study-condition.pdf}
+    \includegraphics{../../results/cough-data/coughs-frequency-by-condition.pdf}\hspace{.5cm}
+    \includegraphics{../../results/epi-data/cases_by_condition.pdf}
+    \caption{\textbf{Comparison of outcomes with vs without air cleaners.} At the top of each plot, the posterior probability for a reduction with air cleaners is shown, based on the Bayesian model.
+\textbf{(a)}~Daily average aerosol number concentrations as boxplots.
+\textbf{(b)}~Number of respiratory cases.
+\textbf{(c)}~Daily average number of detected coughs per minute as boxplots.
+\textbf{(d)}~Number of positive saliva samples.}
+    \label{fig:infection-risk}
+\end{figure}
+
+\subsection*{Risk of infection}
+
+Absences related to respiratory infections included 22~cases without vs 13~cases with air cleaners (\cref{fig:infection-risk}b).
+The Bayesian latent variable hierarchical regression model suggested that air cleaners reduced the risk of infection, with an estimated posterior probability of 91\%.
+The adjusted relative risk of infection with air cleaners was 0.73 (95\%-CrI 0.44 to 1.18).
+Based on that, the estimated number of infections would have been 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed.
+Detailed estimation results are provided in the \supp~text~\zref{sec:detailed-redcap}.
+
+\subsection*{Coughing}
+
+On average, we detected 3.1\,coughs/min (SD 1.2\,coughs/min) without vs 2.6\,coughs/min (SD 1.1\,coughs/min) with air cleaners (\cref{fig:infection-risk}c).
+The Bayesian model suggested that coughing was less frequent with air cleaners, with a posterior probability of 93\%.
+The adjused relative risk of coughing with air cleaners was 0.93 (95\%-CrI 0.85 to 1.02).
+Coughing was also associated with virus-specific transmission (\supp~figure~\zref{fig:coughing-association}).
+
+\subsection*{Molecular analyses}
 
 We analyzed a total of 448~saliva samples.
-We detected 15~influenza B~(IFB), 15~rhinovirus~(HRV), 14~adenovirus~(AdV), 3~SARS-CoV-2~(CoV), 2~metapneumovirus~(MPV), and 1~parainfluenza virus~(PIV), respectively (\cref{fig:molecular-descriptives}a).
+We detected 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus, respectively (\cref{fig:molecular-descriptives}a).
 The distribution of positive saliva samples varied between classes.
-For example, during the first three study weeks, all but one sample was positive for AdV in class~A, while the vast majority of positive IFB samples were found in class~B.
+For example, during the first three study weeks, all but one sample was positive for adenvovirus in class~A, while the vast majority of positive influenza~B samples were found in class~B.
 To illustrate possible transmission chains within classes, we linked positive saliva samples of the same virus that were less than one week apart.
 Based on this spatiotemporal analysis, we identified 10~possible transmission chains (\cref{fig:molecular-descriptives}b).
-The longest potential transmission chains occurred in January, referring to a cluster of AdV infections in class~A and IFB infections in class~B.
+The longest potential transmission chains occurred in January, referring to a cluster of adenovirus infections in class~A and influenza~B infections in class~B.
 
@@
 Furthermore, the rate of positive bioaerosol samples was low.
-We analyzed 105~bioaerosol samples and detected only in two of them viral RNA (1~HRV in class~A and 1~AdV in class~B).
-Similarly, we detected only 1~IFB, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
-However, the virus-specific variation between classes and the temporal distribution of positive samples still suggest that transmission occurred within classes.
+We analyzed 105~bioaerosol samples and detected in two of them viral RNA (1~rhinovirus in class~A and 1~adenovirus in class~B).
+Similarly, we detected 1~influenza~B, 1~rhinovirus, 1~adenovirus, and 1~SARS-CoV-2 in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
+
+There was an equal number of 25~positive saliva samples in both study conditions (\cref{fig:infection-risk}d).
+Based on our Bayesian Multinomial logistic regression model, the positivity rate remained comparable between study conditions (ARR 0.93, 95\%-CrI 0.49 to 1.61; PP 65\%).
+The estimated ARR was not sensitive to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
+Nevertheless, these estimates are subject to unmeasured confounding because the delay may be sample-specific and repeated positive tests from the same person could not be precluded because testing was anonymized.
 
@@
 
-\subsection{Risk of infection with and without air cleaners}
-
-Absences related to respiratory infections were 22~respiratory cases without and 13~cases with air cleaners (\cref{fig:infection-risk}a).
-Our Bayesian latent variable hierarchical regression model suggested that air cleaners reduced the risk of infection (PP 91\%).
-However, the reduction was small and subject to uncertainty (ARR 0.73, 95\%-CrI 0.44 to 1.18).
-The estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\supp~\zref{fig:avoided-infections}).
-Detailed estimation results are provided in the \supp~text~\zref{sec:detailed-redcap}.
-
-There was an equal number of 25~positive saliva samples in both study conditions (\cref{fig:infection-risk}b).
-Based on our Bayesian Multinomial logistic regression model, the positivity rate remained comparable between study conditions (ARR 0.93, 95\%-CrI 0.49 to 1.61; PP 65\%).
-The estimated ARR was not sensitive to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
-Nevertheless, these estimates are subject to unmeasured confounding because the delay may be sample-specific and repeated positive tests from the same person could not be precluded because testing was anonymized.
-
-The aerosol number concentration was 95$\pm$81\,1/cm$^3$ without and 27$\pm$34\,1/cm$^3$ with air cleaners (\cref{fig:infection-risk}c).
-The adjusted decrease in the aerosol concentration with air cleaners was 76\% (95\%-CrI 63\% to 86\%; PP 100\%).
-This decrease was greater for larger (PM$_{10}$) than for smaller (PM$_1$ to PM$_{4}$) particle matter mass concentrations  (\supp~figure~\zref{fig:palas-results} and \supp~table~\zref{tab:palas-est-results}).
-There was little change in other environmental variables (\supp~figure~\zref{fig:env-descriptives-other-vars}).
-In particular, daily average CO$_2$ levels were 1,769$\pm$391\,ppm with and 1,636$\pm$341\,ppm without air cleaners, suggesting that ventilation was comparable between study conditions.
-
-We detected 3.1$\pm$1.2~coughs/min in audio recordings without and 2.6$\pm$1.1~coughs/min with air cleaners (\cref{fig:infection-risk}d).
-Based on our Bayesian Negative Binomial regression model, the risk of coughing was slightly lower with air cleaners (ARR 0.93, 95\%-CrI 0.85 to 1.02; PP 93\%).
-Significantly more coughs were detected in the room of class~B (ARR 2.16, 95\%-CrI 1.90 to 2.46), suggesting an association with the observed variation in virus-specific transmission between classes.
-Indeed, cough was more frequent with IFB positivity (ARR 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent with AdV positivity (ARR 0.87, 95\%-CrI 0.49 to 1.12) (\supp~figure~\zref{fig:coughing-association}b).
-
-\begin{figure}[!htpb]
-\centering
-    \includegraphics{../../results/epi-data/cases_by_condition.pdf}\hspace{.5cm}
-    \includegraphics{../../results/mol-data/saliva-by-study-condition.pdf}
-    \includegraphics{../../results/env-data/aerosol-number-boxplot.pdf}\hspace{.5cm}
-    \includegraphics{../../results/cough-data/coughs-frequency-by-condition.pdf}
-    \caption{\textbf{Risk of infection by study condition}.
-Comparison of raw outcomes with and without air cleaners.
-Atop of each plot the posterior probability (PP) is shown that the outcome was lower with air cleaners (Yes<No), which was computed based on the posterior samples of the Bayesian model used to estimate the effect of air cleaners after adjusting for observed confounders.
-\textbf{(a)}~Number of respiratory cases.
-\textbf{(b)}~Number of positive saliva samples.
-\textbf{(c)}~Daily average aerosol number concentrations as boxplots.
-\textbf{(d)}~Daily average number of detected coughs per minute as boxplots.
-}
-    \label{fig:infection-risk}
-\end{figure}
-
 \FloatBarrier
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 
-\section*{Funding}
-This study was funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
-NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
-ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498).
-\medskip
-
 \section*{Author contributions}

### 2023-11-06 doc/paper/supplements.tex

@@
 \title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
-Impact of air cleaners in schools on respiratory viral infections: \\ A modeling study using epidemiological, environmental, and molecular data}}
+Air cleaners and respiratory infections in schools: \\ A modeling study using epidemiological, environmental, and molecular data}}
 

### 2023-11-06 analysis/epi_analysis.Rmd
minor fix

@@
                         textsize = 8 / cm(1), vjust = -.5, tip_length = .075) +
-  labs(x = "Air cleaner", y = "Number of respiratory cases", fill = "Air cleaner", title = "a") +
+  labs(x = "Air cleaner", y = "Number of respiratory cases", fill = "Air cleaner", title = "b") +
   scale_fill_manual(values = bluegrey) +

### 2023-11-06 doc/paper/manuscript-lancet.tex
Updates from Overleaf

@@
 
-\noindent\textbf{Methods:} We collected epidemiological data (absences related to respiratory infections), environmental data (CO$_2$ and particle concentrations), audio recordings (coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2022/2023 in two Swiss secondary school classes.
+\noindent\textbf{Methods:} We collected environmental data (CO$_2$ and particle concentrations), epidemiological data (absences related to respiratory infections), audio recordings (coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2022/2023 in two Swiss secondary school classes.
 Using a cross-over study design, we compared particle concentrations, coughing, and the risk of infection with vs without air cleaners.\medskip
@@
 
-\noindent\textbf{Funding:} This study was funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
+\noindent\textbf{Funding:} Multidisciplinary Center for Infectious Diseases, University of Bern, Switzerland.
+
+%This study was funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
 NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
@@
 
+\noindent\textbf{Environmental data} \smallskip
+
+\noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute.
+This device was used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}.
+The device did \emph{not} alert teachers when CO$_2$ levels or particle concentrations were too high.\medskip
+
 \noindent\textbf{Epidemiological data} \smallskip
@@
 We defined a case of respiratory infection as an absence in which the student reported an illness with at least one respiratory symptom (\supp~text~\zref{sec:case-data}).
-The list of absences by date of symptom onset, absence and return is shown in \supp~table~\zref{tab:epi-data-line-list}.
-\medskip
+The list of absences by date of symptom onset, absence and return is shown in \supp~table~\zref{tab:epi-data-line-list}.\medskip
 
@@
 \medskip
-
-\noindent\textbf{Environmental data} \smallskip
-
-\noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute.
-This device was used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}.
-The device did \emph{not} alert teachers when CO$_2$ levels or particle concentrations were too high.
 
@@
 
+\noindent\textbf{Particle concentrations} \smallskip
+
+\noindent We computed mean particle concentrations per day and compared them between study conditions.
+We estimated the reduction in particle concentrations using Bayesian log-linear regression models, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases (\supp~text~\zref{sec:env-regression-model}).
+\medskip
+
 \noindent\textbf{Risk of infection} \smallskip
@@
 
-\noindent\textbf{Saliva samples} \smallskip
-
-\noindent We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~text~\zref{sec:multinomial-model}) and linked the expected number of positive samples to the presence of air cleaners, adjusting for the cumulative number of positive tests.
-\medskip
-
-\noindent\textbf{Particle concentrations} \smallskip
-
-\noindent We computed mean particle concentrations per day and compared them between study conditions.
-We estimated the reduction in particle concentrations using Bayesian log-linear regression models, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases (\supp~text~\zref{sec:env-regression-model}).
-\medskip
-
 \noindent\textbf{Coughing} \smallskip
@@
 For this analysis, we used only days on which saliva samples were collected (Tuesdays and Thursdays, for a total of 27~days).
+\medskip
+
+\noindent\textbf{Saliva samples} \smallskip
+
+\noindent We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~text~\zref{sec:multinomial-model}) and linked the expected number of positive samples to the presence of air cleaners, adjusting for the cumulative number of positive tests.
 \medskip
@@
 For the saliva samples, we included all students who were willing to participate and obtained written informed consent from their caregivers.
+
+\subsection*{Role of the funding source}
+
+The funder of the study had no role in study design, data collection, data analysis, data interpretation, or writing of the report.
 
@@
 
-The mean aerosol number concentration was 95\,1/cm$^3$ (standard deviation [SD] 81\,1/cm$^3$) without and 27\,1/cm$^3$ (SD 34\,1/cm$^3$) with air cleaners (\cref{fig:infection-risk}a).
+The mean aerosol number concentration was 95\,1/cm$^3$ (standard deviation [SD] 81\,1/cm$^3$) without vs 27\,1/cm$^3$ (SD 34\,1/cm$^3$) with air cleaners (\cref{fig:infection-risk}a).
 The Bayesian regression model suggested a clear reduction in the aerosol concentration with air cleaners, with a posterior probability of 100\%.
@@
 There was little change in other environmental variables (\supp~figure~\zref{fig:env-descriptives-other-vars}).
-In particular, daily average CO$_2$ levels were 1,769\,ppm (SD 391\,ppm) with and 1,636\,ppm (SD 341\,ppm) without air cleaners, suggesting that ventilation was comparable between study conditions.
+In particular, daily mean CO$_2$ levels were 1,769\,ppm (SD 391\,ppm) with and 1,636\,ppm (SD 341\,ppm) without air cleaners, suggesting that ventilation was comparable between study conditions.
 
@@
     \includegraphics{../../results/env-data/aerosol-number-boxplot.pdf}\hspace{.5cm}
+    \includegraphics{../../results/epi-data/cases_by_condition.pdf}
+    \includegraphics{../../results/cough-data/coughs-frequency-by-condition.pdf}\hspace{.5cm}
     \includegraphics{../../results/mol-data/saliva-by-study-condition.pdf}
-    \includegraphics{../../results/cough-data/coughs-frequency-by-condition.pdf}\hspace{.5cm}
-    \includegraphics{../../results/epi-data/cases_by_condition.pdf}
     \caption{\textbf{Comparison of outcomes with vs without air cleaners.} At the top of each plot, the posterior probability for a reduction with air cleaners is shown, based on the Bayesian model.
@@
 Absences related to respiratory infections included 22~cases without vs 13~cases with air cleaners (\cref{fig:infection-risk}b).
-The Bayesian latent variable hierarchical regression model suggested that air cleaners reduced the risk of infection, with an estimated posterior probability of 91\%.
+The Bayesian latent variable hierarchical regression model suggested that air cleaners reduced the risk of infection, with a posterior probability of 91\%.
 The adjusted relative risk of infection with air cleaners was 0.73 (95\%-CrI 0.44 to 1.18).
-Based on that, the estimated number of infections would have been 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed.
 Detailed estimation results are provided in the \supp~text~\zref{sec:detailed-redcap}.
@@
 The Bayesian model suggested that coughing was less frequent with air cleaners, with a posterior probability of 93\%.
-The adjused relative risk of coughing with air cleaners was 0.93 (95\%-CrI 0.85 to 1.02).
+The adjusted relative risk of coughing with air cleaners was 0.93 (95\%-CrI 0.85 to 1.02).
 Coughing was also associated with virus-specific transmission (\supp~figure~\zref{fig:coughing-association}).
 
-\subsection*{Molecular analyses}
+\subsection*{Saliva samples}
 
@@
 We detected 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus, respectively (\cref{fig:molecular-descriptives}a).
+There were 25~positive saliva samples in both study conditions (\cref{fig:infection-risk}d) and, based on the Bayesian model, the posterior probability that air cleaners reduced the positivity rate was 65\%, with an adjusted relative risk of 0.93 (95\%-CrI 0.49 to 1.61; PP 65\%).
+This estimate was not sensitive to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
+
+\subsection*{Transmission patterns}
+
 The distribution of positive saliva samples varied between classes.
-For example, during the first three study weeks, all but one sample was positive for adenvovirus in class~A, while the vast majority of positive influenza~B samples were found in class~B.
+For example, all but one sample was positive for adenvovirus in class~A during the first three study weeks, while the vast majority of positive influenza~B samples were found in class~B.
 To illustrate possible transmission chains within classes, we linked positive saliva samples of the same virus that were less than one week apart.
@@
 The longest potential transmission chains occurred in January, referring to a cluster of adenovirus infections in class~A and influenza~B infections in class~B.
-
 Molecular genotyping to verify the proposed transmission network was unsuccessful because we could not amplify and sequence any of the gene targets.
@@
 
-There was an equal number of 25~positive saliva samples in both study conditions (\cref{fig:infection-risk}d).
-Based on our Bayesian Multinomial logistic regression model, the positivity rate remained comparable between study conditions (ARR 0.93, 95\%-CrI 0.49 to 1.61; PP 65\%).
-The estimated ARR was not sensitive to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}).
-Nevertheless, these estimates are subject to unmeasured confounding because the delay may be sample-specific and repeated positive tests from the same person could not be precluded because testing was anonymized.
+%Nevertheless, these estimates are subject to unmeasured confounding because the delay may be sample-specific and repeated positive tests from the same person could not be precluded because testing was anonymized.
 
@@
 
-We used a multiple-measurement approach to estimate the risk of respiratory virus infection in a Swiss school and to assess the effectiveness of air cleaners.
-We found a wide range of respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus, but very few viral RNA were detected in bioaerosol samples and on the filters of the air cleaners.
-Aerosol number and particle mass concentrations decreased significantly with air cleaners, and Bayesian modeling based on epidemiological data showed a potential reduction in the relative risk of infection with air cleaners.
-Reduced coughing may indicate that air cleaners prevented symptomatic infections.
+We used a multiple-measurement approach within a cross-over design to estimate the risk of respiratory virus infection in a Swiss school and to assess the effectiveness of air cleaners.
+We found a wide range of respiratory viruses in saliva samples, mainly adenovirus, influenza B, and rhinovirus, but very few viral RNA were detected in bioaerosol samples and on the filters of the air cleaners.
+Aerosol number and particle mass concentrations decreased significantly with air cleaners, and Bayesian modeling based on epidemiological data indicated a reduction in the relative risk of infection with air cleaners.
+Coughing was reduced, compatible with air cleaners preventing some symptomatic infections.
 
@@
 We found a potential benefit of air cleaners on the risk of respiratory infection.
-In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 omicron wave in the winter of 2021/2022.
+In a previous, similar study in the same setting\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 omicron wave in the winter of 2021/2022.
 At that time, we detected almost exclusively SARS-CoV-2 in the students' saliva.
-We found a detectable effect on the risk of SARS-CoV-2 infection only for mask wearing, but not for air cleaners.
-We argued that air cleaners may have had less potential to reduce infections because they were introduced at the end of the study, when most students were already infected with SARS-CoV-2.
+We found a reduction in the risk of SARS-CoV-2 infection for mask wearing, but not for air cleaners, possibly because the air cleaners were introduced only at the end of the study, when most students were already infected with SARS-CoV-2.
 
@@
 
-We showed in a real-world setting that air cleaners can reduce the risk of infection in schools and that they also, consistent with previous  studies\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed}, improve indoor air quality.
-Although the risk reduction effect estimated in this study is rather small, the cumulative impact may be relevant because minimizing the risk of transmission of respiratory infections in schools is an important public health consideration\cite{Beale2023JOMT} and because prevention and control measures in schools, such as air cleaners, must be adequate and acceptable\cite{WHO2020SchoolMeasures}.
-There are several reasons why the effect of air cleaners is likely smaller compared to \eg universal mask wearing, which has been shown to be a very effective infection control measure\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Leung2020NatMed,Milton2013PLoSPathogens}.
-Unlike masks, air cleaners cannot prevent transmission outside the classroom and, importantly, transmission at close range, with a high particle density.
-Prolonged close contact may be necessary for transmission\cite{Leung2020NatMed,Brankston2007LancetID} or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
+It is well documented that air cleaners improve indoor air quality\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed}.
+There are several reasons why the effect of air cleaners is probably smaller than universal mask wearing, which has been shown to be a very effective infection control measure\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Leung2020NatMed,Milton2013PLoSPathogens}.
+Unlike masks, air cleaners cannot prevent transmission outside the classroom or transmission due to close range, high particle density.
+Prolonged and close contact may be necessary for transmission of some respiratory viruses\cite{Leung2020NatMed,Brankston2007LancetID} or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
 Evidence from this and our previous study\cite{Banholzer2023PLoSMed} further suggests that air cleaners are more effective at removing larger particles ($>5\mu$m), which also explains the difference between our measured and the manufacturer's reported clean air delivery rate (see \supp~text~\zref{sec:cadr}).
-However, a greater proportion of respiratory viruses are carried in the smaller particles, which are also more relevant for transmission ($\leq5\mu$m)\cite{Fennelly2020}.
-Finally, classroom activity, airflow and other unobserved, confounding factors make it difficult to evaluate the effects of air cleaners on transmission.
-Nevertheless, it is important to evaluate their effectiveness in a real-world setting and to compare them with the hypothetical effects suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
+However, many respiratory viruses are carried in smaller particles, which are more relevant for transmission ($\leq5\mu$m)\cite{Fennelly2020}.
+Finally, classroom activity, airflow and other unobserved, confounding factors make it challenging to evaluate the effects of air cleaners on transmission in real-world settings.
+Nevertheless, it is important to evaluate their effectiveness in such settings and to compare them with the hypothetical effects suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
+
+%Although the risk reduction effect estimated in this study is rather small, the cumulative impact may be relevant because minimizing the risk of transmission of respiratory infections in schools is an important public health consideration\cite{Beale2023JOMT} and because prevention and control measures in schools, such as air cleaners, must be adequate and acceptable\cite{WHO2020SchoolMeasures}.
 
@@
 A safe and supportive environment is a key determinant for students' success and well-being in school\cite{Kutsyuruba2015RevEduc}.
-On the other hand, air cleaners may not be well accepted by teachers due to noise pollution, limited space, technical issues, and maintenance requirements \cite{Sanguinetti2022IndoorAir}.
+On the other hand, air cleaners may not be well accepted by teachers and authorities due to noise, limited space, technical issues, and the required maintenance\cite{Sanguinetti2022IndoorAir}.
 Therefore, investments in professional building ventilation systems are still preferred in the long term\cite{Nardell2016}.
@@
 
-We detected only few respiratory viruses in bioaerosol samples (1~sample of adenovirus and 1~sample of rhinovirus) and on the filters of the air cleaners (4~positive samples in 160~swabs) in this study.
+We detected only few respiratory viruses in bioaerosol samples (1~sample of adenovirus and 1~sample of rhinovirus) and on the filters of the air cleaners (4~positive samples in 160~swabs).
 The low rate of positive bioaerosol samples may indicate that it is unlikely that airborne transmission occurred to a considerable extent in the classrooms, \ie it is possible that students had relatively little exposure to respiratory viruses at school and acquired their infections elsewhere.
-However, the distribution of saliva samples between classes shows remarkable differences.
-For example, adenovirus spread entirely in class~A during the first three study weeks and only two infections with influenza~B occurred over the study.
+However, the distribution of positive saliva samples markedly differed between classes.
+Adenovirus spread in class~A during the first three study weeks, with only two infections of influenza~B over the study period.
 In contrast, influenza~B spread throughout the study in class~B, and adenovirus infections were detected only in the last week of the study.
-Adenovirus infections tend to be mild\cite{Kunz2010CIDR} and less associated with cough than influenza\cite{Ma2018RMV}, which is consistent with the comparatively lower frequency of detected coughs in class~B.
-Taken together, the class-specific, spatiotemporal patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.
+Adenovirus infections tend to be mild\cite{Kunz2010CIDR} and less frequently associated with cough than influenza\cite{Ma2018RMV}, which is consistent with the comparatively lower frequency of coughing in class~B.
+Taken together, the class-specific, spatiotemporal patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred within the classrooms.
 
@@
 Our study has several limitations.
-First, aerosol measurements and molecular detection of virus in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (person to air, air to person) cannot be determined.
-However, the virus-specific distribution of positive saliva samples between classes (spatiotemporal pattern) may indicate transmission within classrooms.
-Second, the specific causes of school absences could not accurately be determined from the epidemiological data.
-Therefore, some absences may have been incorrectly attributed to respiratory infections, and we could only approximate the incubation period for each epidemiological case.
-Third, the effectiveness of air cleaners was somewhat limited because students occasionally had lessons outside the classroom.
-This is in contrast to other infection control measures such as face masks, which had to be used in every indoor space during the pandemic.
+Aerosol measurements and molecular detection of viruses in bioaerosol samples document exposure, but not transmission and the direction of transmission (person to air, air to person) cannot be determined.
+Further, the reasons for school absences were self-reported by students and some absences may have been incorrectly attributed to respiratory infections.
+In addition, we could only approximate the incubation period for each epidemiological case.
+Finally, the study was conducted in two classrooms of one Swiss school and although results are likely to apply to many settings in Switzerland and other European countries, they will not be applicable to settings in the global South.
 
@@
 
-In conclusion, using epidemiological, environmental, and molecular data, our study showed that a wide range of non-SARS-CoV-2, but rarely SARS-CoV-2, was detected in students under non-pandemic conditions when public health measures were lifted, and that transmission of these respiratory viruses likely occurred in schools.
-Airborne detection of non-SARS-CoV-2 respiratory viruses was rare, suggesting that that respiratory viruses other than SARS-CoV-2 may be less detectable in the air and that close contact may be required for transmission in most cases.
-Air cleaners improved air quality by significantly reducing aerosol and particle concentrations.
-Based on our transmission model using epidemiological data, the risk of infection tends to be lower with air filters.
-Future studies should further investigate the relative importance of prolonged close contact versus long-range transmission of respiratory viruses in schools and other congregate settings.
+In conclusion, using epidemiological, environmental, and molecular data, our study showed that a wide range of respiratory viruses, but rarely SARS-CoV-2, was detected in students under non-pandemic conditions when public health measures were lifted, and that transmission of these respiratory viruses likely occurred in schools.
+Airborne detection of non-SARS-CoV-2 respiratory viruses was rare, suggesting that that respiratory viruses other than SARS-CoV-2 may be more difficult to detect in the air and that close contact may be required for transmission in most cases.
+Air cleaners improved air quality by significantly reducing particle concentrations.
+The reduction in risk of respiratory infections conferred by air cleaners may be modest at the individual level, however, the benefit at the population level in terms of illness and school absences prevented will likely be important.
+Future studies should investigate the relative importance of prolonged close contact and long-range transmission of respiratory viruses in schools and other congregate settings, and examine the cost-effectiveness of using air cleaners in these settings.
 

### 2023-11-06 doc/paper/supplements.tex

@@
 
-Our estimation results imply a small, potential benefit of air cleaners.
+% Our estimation results imply a small, potential benefit of air cleaners.
 Subject to considerable uncertainty, \Cref{fig:avoided-infections} shows the estimated number number of infections that would have occurred if air cleaners had been installed throughout the study (scenario~1) or not been installed (scenario~2).
 
-\begin{figure}[!htpb]
-    \includegraphics{../../results/epi-data/avoided-infections.pdf}
-    \caption[Estimated number of infections with and without air cleaners]{Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
-    \label{fig:avoided-infections}
-\end{figure}
+% \begin{figure}[!htpb]
+%     \includegraphics{../../results/epi-data/avoided-infections.pdf}
+%     \caption[Estimated number of infections with and without air cleaners]{Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
+%     \label{fig:avoided-infections}
+% \end{figure}
 

### 2023-11-07 analysis/epi_analysis.Rmd
excluding vacation and weekends from computing avoided infections

@@
 avoid_infections <- spread_draws(latInfMod, mu_cases_ac[day,class,type]) %>%
+  mutate(day = day - stan_data$S) %>%
+  filter(day > 0) %>%
+  left_join(data.frame(day = 1:length(stan_data$weekend),
+                       weekend = stan_data$weekend,
+                       vacation = stan_data$vacation)) %>%
+  filter(weekend == 0,
+         vacation == 0) %>%
   group_by(.draw, type) %>%
@@
   filter(type != types[1])
+
+avoid_infections
 

### 2023-11-07 doc/paper/supplements.tex
Updates from Overleaf

@@
 
-\title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
+\title{\LARGE\singlespacing{\textbf{Appendix} \\ \medskip
 Air cleaners and respiratory infections in schools: \\ A modeling study using epidemiological, environmental, and molecular data}}
@@
 
-The clean air delivery rate (CADR) is typically used to quantify the cleaning efficiency of air purifiers\cite{Nelson1993IndoorEnv}.
-It is expressed in m$^3$/s and describes the volumetric flow rate of particle-free, clean air that an air purifier delivers into the room.
+The clean air delivery rate (CADR) is typically used to quantify the cleaning efficiency of air cleaners.\cite{Nelson1993IndoorEnv} It is expressed in m$^3$/s and describes the volumetric flow rate of particle-free, clean air that an air cleaner delivers into the room.
 The CADR is equal to the product of the filtration efficiency and the volumetric flow rate passing through the device.
 
-We followed the approach by K\"{u}pper et al.\cite{Kuepper20199AAQR} to determine the CADR of the air purifiers used in this study.
-The approach is based on the release of fine aerosols (particle diameter $D$<1$\mu$m) in a closed, unventilated room by measuring the decay rate $k$ of the particle concentration $N(t)$ caused by the air cleaner, \ie
+We followed the approach by K\"{u}pper et al.
+to determine the CADR.\cite{Kuepper20199AAQR} The approach is based on the release of fine aerosols (particle diameter $D$<1$\mu$m) in a closed, unventilated room by measuring the decay rate $k$ of the particle concentration $N(t)$ caused by the air cleaner, \ie
 \begin{align}
@@
 
-Figure~\ref{fig:experiment} shows the particle number concentration with and without without the air purifier.
+Figure~\ref{fig:experiment} shows the particle number concentration with and without without the air cleaner.
 The exponential decrease is clearly visible as a linear curve in the logarithmic plot.
-Based on these experiments, we were able to determine the net CADR values of the air purifier as 420\,m$^3$/h with an uncertainty of $\pm$10\% due to systematic differences in the setup and measurement errors of the instruments.
+Based on these experiments, we were able to determine the net CADR values of the air cleaner as 420\,m$^3$/h with an uncertainty of $\pm$10\% due to systematic differences in the setup and measurement errors of the instruments.
 Our estimated CADR is lower than the CADR reported by the manufacturer of 600m$^3$/h, most likely because the reported CADR was determined with comparably large particles (pollen in the super micrometer size range).
@@
     \includegraphics[width=\linewidth]{doc/paper/experiment-particle-concentration-decay.png}
-    \caption{Measured concentration curves with the air purifier on and off.
-The red dot shows the average concentration measured by the 11 distributed sensors of particles with $0.3\mu m < D < 1 \mu m$.
+    \caption[Measured concentration curves with and without air cleaner]{\textbf{Measured concentration curves with and without air cleaner.} The red dot shows the average concentration measured by the 11~distributed sensors of particles with $0.3\mu m < D < 1 \mu m$.
 The standard deviation is shown in gray as an error bar.
@@
     \centering
-    \caption{\textbf{Data collection}. Type of data collected, method/device, and frequency in the rooms of classes A and B.}
+    \caption[Overview of collected data]{\textbf{Overview of collected data}. Type of data collected, method/device, and frequency in the rooms of classes A and B.}
     \begin{tabular}{p{3.5cm}p{6cm} p{1cm} p{1cm} p{3cm}}
@@
     CO$_2$, aerosol/particle concentrations, humidity, temperature & Air quality device: AQ Guard (Palas GmbH, Karlsruhe, Germany) & x & x & daily by minute \\
-    Coughs & Sounds collected via microphones and coughs detected with a Machine Learning model \cite{Bertschinger2023CBMS} & x & x & daily by seconds \\
+    Coughs & Sounds collected via microphones and coughs detected with a Machine Learning model\cite{Bertschinger2023CBMS} & x & x & daily by seconds \\
     \midrule
@@
 
-Reports about absences were entered electronically into a REDcap database\cite{Harris2009,Harris2019}.
-\Cref{tab:epi-data-line-list} shows the line list data for respiratory cases based on epidemiological data.
+Reports about absences were entered electronically into a REDcap database.\cite{Harris2009,Harris2019} \Cref{tab:epi-data-line-list} shows the line list data for respiratory cases based on epidemiological data.
 For each class and case, it shows the date when the student was absent, s/he reported symptoms, and returned to school.
@@
 \end{longtable}}
+
+\clearpage
+
+\section{Modeling changes in particle concentrations}\label{sec:env-regression-model}
+
+We only analyze particle concentration data of the time periods during which students were in the classroom.
+Particle concentrations $Y$ on day $t$ in class $j$ are summarized daily with the mean across these time periods.
+The change in aerosol number and particle matter mass concentrations is then estimated using Bayesian log-linear regression models
+\begin{align*}
+    \log Y | \beta_0, \beta_1, \dots, \beta_6 &\sim \text{Normal}(\mu,\sigma) \\
+    \mu &= \beta_0 + \beta_1 \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Weekday} + \beta_4 \cdot \text{Students} \\
+    &\hphantom{= } + \beta_5 \cdot \text{AirChangeRate} + \beta_6 \cdot \text{Cases} \\
+    \beta_0 &\sim \text{Student-t}_5(\mu_y, 2.5s_{y}) \\
+    \beta_1, \dots, \beta_6 &\sim \text{Student-t}_5\left(0, 2.5\frac{s_{y}}{s_{x}}\right) \\
+    \sigma &\sim \text{Exponential}\left(\frac{1}{s_{y}}\right)~.
+\end{align*}
+The effect of air cleaners is adjusted for class- and weekday-specific effects, the number of students in school, the air change rate, and the cumulative number of cases related to respiratory infections.
+A log transform is applied to all continuous input variables.
 
@@
 
-The prior distributions for the virus-specific incubation periods are based on estimates published in the literature\cite{McAloon2020,Lessler2009LancetID}.
-The distributions are shown in \Cref{fig:prior-pin}.
+The prior distributions for the virus-specific incubation periods are based on estimates published in the literature.\cite{McAloon2020,Lessler2009LancetID} The distributions are shown in \Cref{fig:prior-pin}.
 Since we could not obtain prior estimates for the incubation period of metapneumovirus (MPV) from the literature, we instead formed a distribution by using the equally weighted average of the parameters from the other distributions.
@@
 
-We use weakly informative priors for all modeling parameters following recommendations on the choice of priors \cite{Gelman2008StatMed,Gelman2008StatAnnals,Gelman2020RegOther,Stan2020Priors,Gabry2023Priors}.
-The continuous adjustment variables are standardized to have zero mean and a standard deviation of 0.5.
+We use weakly informative priors for all modeling parameters following recommendations on the choice of priors.\cite{Gelman2008StatMed,Gelman2008StatAnnals,Gelman2020RegOther,Stan2020Priors,Gabry2023Priors} The continuous adjustment variables are standardized to have zero mean and a standard deviation of 0.5.
 \begin{align*}
@@
 where Student-t$_5$ is a Student-t distribution with 5 degrees of freedom.
-
-\clearpage
-
-\section{Modeling positivity rate of human saliva samples}\label{sec:multinomial-model}
-
-Molecular analysis determined which saliva samples were positive for a respiratory virus.
-Let $p = 0, 1, \dots, P$ denote the virus, where $0$ refers to negative samples and $P$ is the number of different viruses detected over the study.
-The number of positive bioaerosol and saliva samples $y_p$ in samples of sizes $n$ is analyzed with a Multinomial logistic regression model
-\begin{align*}
-    y_0, y_1, \cdots, y_P | n, \beta_0, \beta_1, \beta_2, \beta_3 &\sim \text{Multinomial-Logit}(\theta_0, \theta_1, \dots \theta_P) \\
-    \theta &= \text{softmax}(\mu) \\
-    \mu &= \begin{cases}
-                0, & \text{if } p=0, \\
-                \beta_{0p} + \beta_{1} \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Susceptibles}_p, & \text{if } p>0.
-            \end{cases} \\
-    \beta_{0p} &\sim \text{Normal}(0, 2.5) \quad \forall p \\
-    \beta_1, \beta_2, \beta_3 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) 
-\end{align*}
-where $\text{softmax}(\mu) = \exp(\mu)/\sum\exp(\mu)$ and $s_{x}$ is the empirical standard deviation of each input variable.
-The negative test is set as the reference category.
-Overall variation in positive samples by virus will be modeled with $\beta_{0p}$ and the effect of air cleaners will be modeled with $\beta_1$.
-The effect of air cleaners is adjusted for class-specific effects and the decreasing number of susceptibles over time (for each virus computed as the number of students minus the total number of positive samples).
-
-Note that we deviated from the statistical analysis plan by estimating $\beta_{0p}$ and $\beta_1$ without a hierarchical prior.
-First, there was considerable variation in the overall positivity rate of each virus, so that we decided to use unpooled intercepts $\beta_{0p}$ instead of partially pooled intercepts $\beta_{0[p]}$ for each virus.
-Second, there was insufficient variation in the data to inform partially pooled estimates $\beta_{1[p]}$ for the effects of air cleaners.
-Therefore, we decided to only estimate the average effect across viruses, \ie the completely pooled estimate $\beta_1$.
-
-\section{Modeling changes in particle concentrations}\label{sec:env-regression-model}
-
-We only analyze particle concentration data of the time periods during which students were in the classroom.
-Particle concentrations $Y$ on day $t$ in class $j$ are summarized daily with the mean across these time periods.
-The change in aerosol number and particle matter mass concentrations is then estimated using Bayesian log-linear regression models
-\begin{align*}
-    \log Y | \beta_0, \beta_1, \dots, \beta_6 &\sim \text{Normal}(\mu,\sigma) \\
-    \mu &= \beta_0 + \beta_1 \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Weekday} + \beta_4 \cdot \text{Students} \\
-    &\hphantom{= } + \beta_5 \cdot \text{AirChangeRate} + \beta_6 \cdot \text{Cases} \\
-    \beta_0 &\sim \text{Student-t}_5(\mu_y, 2.5s_{y}) \\
-    \beta_1, \dots, \beta_6 &\sim \text{Student-t}_5\left(0, 2.5\frac{s_{y}}{s_{x}}\right) \\
-    \sigma &\sim \text{Exponential}\left(\frac{1}{s_{y}}\right)~.
-\end{align*}
-The effect of air cleaners is adjusted for class- and weekday-specific effects, the number of students in school, the air change rate, and the cumulative number of cases related to respiratory infections.
-A log transform is applied to all continuous input variables.
 
@@
 
-\section{Detailed results for risk of infection model}\label{sec:detailed-redcap}
+\section{Modeling positivity rate of human saliva samples}\label{sec:multinomial-model}
 
-Table~\ref{tab:estimation-results} presents the posterior mean, credible intervals and model diagnostics for all model parameters.
-See the main paper for a discussion of the effects of interventions.
-Here we briefly discuss some of the additional model parameters.
+Molecular analysis determined which saliva samples were positive for a respiratory virus.
+Let $p = 0, 1, \dots, P$ denote the virus, where $0$ refers to negative samples and $P$ is the number of different viruses detected over the study.
+The number of positive bioaerosol and saliva samples $y_p$ in samples of sizes $n$ is analyzed with a Multinomial logistic regression model
+\begin{align*}
+    y_0, y_1, \cdots, y_P | n, \beta_0, \beta_1, \beta_2, \beta_3 &\sim \text{Multinomial-Logit}(\theta_0, \theta_1, \dots \theta_P) \\
+    \theta &= \text{softmax}(\mu) \\
+    \mu &= \begin{cases}
+                0, & \text{if } p=0, \\
+                \beta_{0p} + \beta_{1} \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Susceptibles}_p, & \text{if } p>0.
+            \end{cases} \\
+    \beta_{0p} &\sim \text{Normal}(0, 2.5) \quad \forall p \\
+    \beta_1, \beta_2, \beta_3 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) 
+\end{align*}
+where $\text{softmax}(\mu) = \exp(\mu)/\sum\exp(\mu)$ and $s_{x}$ is the empirical standard deviation of each input variable.
+The negative test is set as the reference category.
+Overall variation in positive samples by virus will be modeled with $\beta_{0p}$ and the effect of air cleaners will be modeled with $\beta_1$.
+The effect of air cleaners is adjusted for class-specific effects and the decreasing number of susceptibles over time (for each virus computed as the number of students minus the total number of positive samples).
 
-The effective sample size (ESS) and the Gelman-Rubin convergence diagnostic ($\hat{R}$) indicate good estimation power. It further suggests that the Markov chains converged.
-
-The estimate for $\beta_2$ is positive, suggesting a higher risk of infection in class $B$, which is in line with the higher number of cases in this class.
-The estimate for $\beta_3$ is positive, suggesting a higher risk of infection as more students were in class.
-The estimate for $\beta_4$ is negative, suggesting a lower risk of infection with higher air change rates, in line with the intuition that the risk of infection is lower when the classroom is better ventilated.
-The estimate for $\beta_5$ and $\beta_6$ are both positive although only the former is distinguishable from zero, suggesting that higher transmission in the community was also associated with a higher risk of infection in the school.
-
-The overdispersion parameter ($\phi$) cannot be precisely estimated, but the credible interval suggests rather small overdispersion ($\phi \rightarrow \infty$).
-The school-free effect $\omega$, the weekday weights $\vartheta$, and the parameters for the distribution of the incubation period $\mu_{p_{\text{IN}}}$ and $\sigma_{p_{\text{IN}}}$ are all close to their priors, suggest that these parameters are barely informed by date but more by our prior assumptions.
-
-\begin{table}[!htpb]
-    \caption[Estimation results from infection risk model]{Estimation results from infection risk model for the number of new respiratory cases.}
-    \label{tab:estimation-results}
-    \footnotesize
-    \centering
-    \begin{tabular}{lrrrrr}
-    \toprule
-    Parameter & Mean & Lower 95\%-CrI & Upper 95\%-CrI & $\hat{R}$ & ESS \\
-    \midrule
-    \input ../../results/epi-data/estimation-results
-    \bottomrule
-    \multicolumn{6}{p{13cm}}{\scriptsize
-        ESS is the effective sample size, \ie the number of independent MCMC samples with estimation power equivalent to the total number of autocorrelated samples\cite{Stan2022}, and $\hat{R}$ is the Gelman-Rubin convergence diagnostic \cite{Gelman1992}.
-Low ESS or $\hat{R}$ or $\hat{R}>1.10$ indicate bad convergence of the model \cite{Gelman2013}.}
-    \end{tabular}
-\end{table}
-
-Our model accounts for the delay from infection to case confirmation and allows transmission to change only at the dates of interventions.
-It thus reflects overall trends in transmission during study conditions rather than day-to-day variation.
-To evaluate how well our model fits these trends, \Cref{fig:coverage} compares the model-estimated expected number of cases  with the observed number cases only on a weekly basis and across classes.
-Overall the estimates are in relatively good agreement.
-The 95\%-CrIs of the model-estimates always include the observed cases.
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{../../results/epi-data/model-fit.pdf}
-    \caption[Comparison of model-estimates with observed case data]{Model-estimated number of new respiratory cases (posterior mean as red line and 50\%-, 80\%, and 95\%-CrIs as shaded areas) and the observed number of new respiratory cases (black line) across classes by study week.}
-    \label{fig:coverage}
-\end{figure}
-
-% Our estimation results imply a small, potential benefit of air cleaners.
-Subject to considerable uncertainty, \Cref{fig:avoided-infections} shows the estimated number number of infections that would have occurred if air cleaners had been installed throughout the study (scenario~1) or not been installed (scenario~2).
-
-% \begin{figure}[!htpb]
-%     \includegraphics{../../results/epi-data/avoided-infections.pdf}
-%     \caption[Estimated number of infections with and without air cleaners]{Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
-%     \label{fig:avoided-infections}
-% \end{figure}
-
-\clearpage
-
-\section{Detailed results for positivity rate in saliva samples}
-\label{sec:detailed-molecular}
-
-There can be unknown delay between when students were infected and when they tested positive.
-Therefore, we performed a sensitivity analysis where we lagged the molecular test results (number of positive saliva samples) by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday).
-The results are shown in \Cref{fig:mol-estimation-results-sensitivity}.
-The estimated effect of air cleaners remains indistinguishable from zero regardless of the considered lag.
-
-\begin{figure}[!htb]
-\centering
-    \includegraphics[width=10cm]{../../results/mol-data/model-results.pdf}
-    \caption[Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples]{Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples.
-Shown is the adjusted relative risk (posterior mean as dot and 50\%-, 80\%, and 95\%-CrIs as lines) for air cleaners by lag of the molecular test results.}
-    \label{fig:mol-estimation-results-sensitivity}
-\end{figure}
+Note that we deviated from the statistical analysis plan by estimating $\beta_{0p}$ and $\beta_1$ without a hierarchical prior.
+First, there was considerable variation in the overall positivity rate of each virus, so that we decided to use unpooled intercepts $\beta_{0p}$ instead of partially pooled intercepts $\beta_{0[p]}$ for each virus.
+Second, there was insufficient variation in the data to inform partially pooled estimates $\beta_{1[p]}$ for the effects of air cleaners.
+Therefore, we decided to only estimate the average effect across viruses, \ie the completely pooled estimate $\beta_1$.
 
@@
 \Cref{fig:env-descriptives-other-vars} compares additional environmental variables, \ie CO$_2$, air change rate, relative humidity, and temperature.
-The air change rate was estimated from indoor CO$_2$ levels using a transient mass balance model\cite{Batterman2017IJERPH}.
-There are no considerable differences between in any of the variables between study conditions.
+The air change rate was estimated from indoor CO$_2$ levels using a transient mass balance model.\cite{Batterman2017IJERPH} There are no considerable differences between in any of the variables between study conditions.
 Furthermore, numerical estimation results for the reduction in particle concentrations with air cleaners are shown in \Cref{tab:palas-est-results}.
@@
 \end{figure}
+
+\clearpage
+
+\section{Detailed results for risk of infection model}\label{sec:detailed-redcap}
+
+Table~\ref{tab:estimation-results} presents the posterior mean, credible intervals and model diagnostics for all model parameters.
+See the main paper for a discussion of the effects of interventions.
+Here we briefly discuss some of the additional model parameters.
+
+The effective sample size (ESS) and the Gelman-Rubin convergence diagnostic ($\hat{R}$) indicate good estimation power. It further suggests that the Markov chains converged.
+
+The estimate for $\beta_2$ is positive, suggesting a higher risk of infection in class $B$, which is in line with the higher number of cases in this class.
+The estimate for $\beta_3$ is positive, suggesting a higher risk of infection as more students were in class.
+The estimate for $\beta_4$ is negative, suggesting a lower risk of infection with higher air change rates, in line with the intuition that the risk of infection is lower when the classroom is better ventilated.
+The estimate for $\beta_5$ and $\beta_6$ are both positive although only the former is distinguishable from zero, suggesting that higher transmission in the community was also associated with a higher risk of infection in the school.
+
+The overdispersion parameter ($\phi$) cannot be precisely estimated, but the credible interval suggests rather small overdispersion ($\phi \rightarrow \infty$).
+The school-free effect $\omega$, the weekday weights $\vartheta$, and the parameters for the distribution of the incubation period $\mu_{p_{\text{IN}}}$ and $\sigma_{p_{\text{IN}}}$ are all close to their priors, suggest that these parameters are barely informed by date but more by our prior assumptions.
+
+\begin{table}[!htpb]
+    \caption[Estimation results from infection risk model]{Estimation results from infection risk model for the number of new respiratory cases.}
+    \label{tab:estimation-results}
+    \footnotesize
+    \centering
+    \begin{tabular}{lrrrrr}
+    \toprule
+    Parameter & Mean & Lower 95\%-CrI & Upper 95\%-CrI & $\hat{R}$ & ESS \\
+    \midrule
+    \input ../../results/epi-data/estimation-results
+    \bottomrule
+    \multicolumn{6}{p{13cm}}{\scriptsize
+        ESS is the effective sample size, \ie the number of independent MCMC samples with estimation power equivalent to the total number of autocorrelated samples,\cite{Stan2022} and $\hat{R}$ is the Gelman-Rubin convergence diagnostic.\cite{Gelman1992} Low ESS or $\hat{R}$ or $\hat{R}>1.10$ indicate bad convergence of the model.\cite{Gelman2013}}
+    \end{tabular}
+\end{table}
+
+Our model accounts for the delay from infection to case confirmation and allows transmission to change only at the dates of interventions.
+It thus reflects overall trends in transmission during study conditions rather than day-to-day variation.
+To evaluate how well our model fits these trends, \Cref{fig:coverage} compares the model-estimated expected number of cases  with the observed number cases only on a weekly basis and across classes.
+Overall the estimates are in relatively good agreement.
+The 95\%-CrIs of the model-estimates always include the observed cases.
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{../../results/epi-data/model-fit.pdf}
+    \caption[Comparison of model-estimates with observed case data]{Model-estimated number of new respiratory cases (posterior mean as red line and 50\%-, 80\%, and 95\%-CrIs as shaded areas) and the observed number of new respiratory cases (black line) across classes by study week.}
+    \label{fig:coverage}
+\end{figure}
+
+% Our estimation results imply a small, potential benefit of air cleaners.
+Subject to considerable uncertainty, \Cref{fig:avoided-infections} shows the estimated number number of infections that would have occurred if air cleaners had been installed throughout the study (scenario~1) or not been installed (scenario~2).
+
+% \begin{figure}[!htpb]
+%     \includegraphics{../../results/epi-data/avoided-infections.pdf}
+%     \caption[Estimated number of infections with and without air cleaners]{Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
+%     \label{fig:avoided-infections}
+% \end{figure}
 
@@
 
+\section{Detailed results for positivity rate in saliva samples}
+\label{sec:detailed-molecular}
+
+There can be unknown delay between when students were infected and when they tested positive.
+Therefore, we performed a sensitivity analysis where we lagged the molecular test results (number of positive saliva samples) by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday).
+The results are shown in \Cref{fig:mol-estimation-results-sensitivity}.
+The estimated effect of air cleaners remains indistinguishable from zero regardless of the considered lag.
+\begin{figure}[!htb]
+\centering
+    \includegraphics[width=10cm]{../../results/mol-data/model-results.pdf}
+    \caption[Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples]{Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples.
+Shown is the adjusted relative risk (posterior mean as dot and 50\%-, 80\%, and 95\%-CrIs as lines) for air cleaners by lag of the molecular test results.}
+    \label{fig:mol-estimation-results-sensitivity}
+\end{figure}
+
+\clearpage
+
 \bibliography{references.bib}

### 2023-11-07 doc/paper/manuscript.tex
finalised manuscript

--- 2023-10-06 doc/paper/manuscript.tex
@@
 
-\title{\bfseries\LARGE\singlespacing{Transmission of respiratory viral infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}}
+\title{\bfseries\LARGE\singlespacing{Air cleaners and respiratory infections in schools: A modeling study using epidemiological, environmental, and molecular data}}
 
@@
 \author[1]{Simon Bertschinger}
+\author[5]{Ernest Weingartner}
 \author[2,4]{Alban Ramette}
-\author[2,4]{Ronald Dijkman}
-\author[1]{Matthias Egger}
-\author[2,5]{Tina Hascher}
+\author[1,6,7]{Matthias Egger}
+\author[8]{Tina Hascher}
 \author[1*,2]{Lukas Fenner}
@@
 \affil[4]{Institute for Infectious Diseases, University of Bern, Bern, Switzerland}
-\affil[5]{Institute of Educational Science, University of Bern, Bern, Switzerland}
+\affil[5]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
+\affil[6]{Population Health Sciences, University of Bristol, Bristol, UK}
+\affil[7]{Centre for Infectious Disease Epidemiology and Research, University of Cape Town, Cape Town, South Africa}
+\affil[8]{Institute of Educational Science, University of Bern, Bern, Switzerland}
 
@@
 
-\vspace{1em}
+\vspace{2em}
 
 \begin{information}\normalfont
-
-xxxx
-
-% \noindent\textbf{Running head}: SARS-COV-2 transmission in schools and effect of air cleaners
-
-% %\noindent\textbf{Subject categorization}: 6.20  Indoor Air; 10.11 Pediatrics: Respiratory Infections
-
-% \noindent\textbf{Word count}: \maincount{manuscript}words, abstract \abstractcount{manuscript}words (max. 500), title 157 chars (max. 200)
-
-% %\noindent\textbf{Inserts:} 2 tables, 6 figures, 44 references
-
-% \noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.
-
-% \vspace{1em}
-
-% \noindent\textbf{Funding}
-
-% \noindent This study is funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
-NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
-ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498).
-\medskip
-
-% \noindent\textbf{Contributions}
-
-% \noindent Conception and design: NB, LF.
-Epidemiological and environmental data collection: NB, PJ, TS, LF.
-Laboratory data collection: PB, LFu.
-Additional data collection: TH.
-Statistical analysis: NB, KZ.
-Genomic analysis: LB, LFu.
-Paper draft: NB, LF, ME.
-All authors reviewed and approved the final version of the manuscript.
-
+\textbf{Word count:} Summary 276 words, Research in context 359 words, Main text 3,492 words \\
+\textbf{Display items:} Tables 2, Figures 3 \\
+\textbf{Supporting information:} STROBE checklist, Appendix with additional text, tables, and figures.
 % \par
@@
 \begin{abstract}\normalfont
-\noindent\textbf{Background:} Airborne respiratory infections can be transmitted in crowded indoor settings, including schools.
-Infection control measures such as air ventilation or filtration are supposed to reduce the risk of airborne transmission.
-The aim of this study was to evaluate the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory viral infections in schools using a multiple-measurement approach.
-\medskip
+\noindent\textbf{Background:} Infection control measures can reduce airborne transmission of respiratory infections.
+Using a multiple-measurement approach, we examined the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in a school setting.\medskip
 
-\noindent\textbf{Methods:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), audio (recordings to detect coughing), and molecular data (bioaerosol and saliva samples) over seven weeks from 16 January to 11 March 2023 in a secondary school (38 students in two classes) in Switzerland.
-We performed laboratory and molecular analyses to detect a panel of 19 respiratory viruses.
-We installed air cleaners in a cross-over design and compared positive samples and particle concentrations between weeks with and without air filters.
-We modeled the number of new infections using a Bayesian negative binomial regression model, adjusting for incubation periods, number of recent and cumulative infections, ventilation rate, number of students in class, and community transmission.
-\medskip
+\noindent\textbf{Methods:} We collected environmental data (CO$_2$ and particle concentrations), epidemiological data (absences related to respiratory infections), audio recordings (coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2022/2023 in two Swiss secondary school classes.
+Using a cross-over study design, we compared particle concentrations, coughing, and the risk of infection with vs without air cleaners.\medskip
 
-\noindent\textbf{Results:} Molecular analysis detected multiple viruses in saliva samples throughout the study (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus), but rarely in bioaerosol samples (2/105 positive: 1~rhinovirus and 1~adenovirus) and on the filters of air cleaners (4/160~positive: 1~influenza~B, 1~HRV, 1~adenovirus, and 1~SARS-CoV-2).
-Nevertheless, spatiotemporal analysis of saliva samples suggested transmission within classrooms.
-The detection rate of positive saliva samples was comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61).
-Daily average CO$_2$ levels were 1,769$\pm$391\,ppm ($\pm$ standard deviation) with air cleaners and 1,636$\pm$341\,ppm without air cleaners, corresponding to similar air change rates (0.74$\pm$0.41\,h$^{-1}$ vs.
-0.68$\pm$0.32\,h$^{-1}$).
-Daily average aerosol number concentrations were 27$\pm$34\,1/cm$^3$ with and 95$\pm$81\,1/cm$^3$ without air cleaners, corresponding to an estimated adjusted decrease of 77\% (95\%-CrI 63\% to 86\%).
-Slightly less coughs  were detected with air cleaners (2.5~coughs/min with vs.
-3.1~coughs/min without air cleaners).
-Absences related to respiratory infections were less frequent with air cleaners (13~absences with vs.
-21~absences without air cleaners).
-The estimated risk of infection tended to be lower with air cleaners (adjusted relative risk 0.72, 95\%-Credible Interval 0.44 to 1.15).
-\medskip
+\noindent\textbf{Findings:} All 38~students (age 13$-$15 years) participated.
+With air cleaners, mean particle concentration decreased by 77\% (95\% credible interval 63\%$-$86\%).
+There were no differences in CO$_2$ levels.
+Absences related to respiratory infections were 22~without vs 13~with air cleaners.
+Bayesian modeling suggested a reduced risk of infection, with a posterior probability of 91\% and a relative risk of 0.73 (95\% credible interval 0.44$-$1.18).
+Coughing also tended to be less frequent (posterior probability 93\%).
+Molecular analysis detected mainly non-SARS-CoV-2 viruses in saliva (50/448 positive), but not in bioaerosol samples (2/105 positive) or HEPA-filters (4/160).
+The detection rate was similar with vs without air cleaners.
+Spatiotemporal analysis of positive saliva samples identified several likely transmissions.\medskip
 
-% Study limitations include unknown specific causes of absences related to respiratory infections (i.e which virus caused the sickness).
-Furthermore, students occasionally had lessons outside the classroom, thus limiting the potential effectiveness of air cleaners.
+\noindent\textbf{Interpretation:} Air cleaners improved air quality, showed a potential benefit in reducing respiratory infections, and were associated with less coughing.
+Airborne detection of non-SARS-CoV-2 viruses was rare, suggesting that these viruses may be more difficult to detect in the air.
+Future studies should examine the importance of close contact and long-range transmission in schools and other congregate settings, and the cost-effectiveness of using air cleaners.\medskip
 
-\noindent\textbf{Conclusions:} During this non-pandemic period, a wide range of non-SARS-CoV-2 respiratory viruses were detected in students' saliva, but airborne molecular detection was rare.
-However, transmission of these respiratory viruses seemed likely in schools, suggesting that prolonged close contact is required for transmission.SARS-CoV-2 was very rarely detected in saliva and air.
-Air cleaners improved air quality, were associated with lower symptom intensity, and showed a small, potential benefit in reducing respiratory infections.
+\noindent\textbf{Funding:} Multidisciplinary Center for Infectious Diseases, Switzerland.
+
+%This study was funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
+NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
+ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498).
 
@@
 
-\section{Introduction}
+\noindent\textbf{\Large{Research in context}} \medskip
 
-The transmission of respiratory infections such as SARS-CoV-2 and influenza is difficult to mitigate and control.
-Person-to-person transmission occurs primarily through the release of respiratory particles containing the respective viruses.
-Recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
-Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and their respiratory plumes can reach more than 2m\cite{Coleman2022,Wang2020,Heneghan2021}, exceeding the social distancing rules recommended by public health organizations\cite{Bazant2021PNAS,Trivedi2021PhyFluids,Lindsley2010CID}.
+\noindent \textbf{Evidence before this study} \smallskip
 
-The generation, spread, and concentration of bioareosols can be influenced by infection control measures.
-For example, mandatory face mask use reduced transmission of SARS-CoV-2 during the COVID-19 pandemic\cite{Gettings2021,Heinsohn2022,Leech2022PNAS,Mitze2020PNAS,Banholzer2021PLOS}.
-In addition, improved ventilation systems are critical for a healthy indoor environment both in pandemic and non-pandemic times\cite{Wang2021,Morawska2021}.
-Portable HEPA-air filtration devices (air cleaners) may be another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear.
-Several studies have shown that air cleaners reduce the concentration of aerosols and particle matter in indoor air\cite{Park2020Build,Buising2022InfContr,Ren2021Dent,Banholzer2023PLoSMed}.
-These findings suggest potential protection against respiratory infections, as long-term exposure to higher aerosol and particle matter mass concentrations has been associated with increased risk of respiratory infections\cite{Gordon2014Resp} and, more recently, with the incidence and mortality of COVID-19\cite{Kelly2023Atmos,Veronesi2022OccEnvMed,Travaglio2021EnvPoll,DeAngelis2021EnvRes}.
-In addition, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}.
-Simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission  of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.
-We have studied the effectiveness of air cleaners in reducing transmission of SARS-CoV-2 in schools during the pandemic\cite{Banholzer2023PLoSMed}, but the analysis was limited by the study design (installation of air cleaners at the end of the Omicron wave when most students were already infected).
-To date, empirical evidence linking the use of air cleaners to a reduced risk of infection is lacking.
-It remains unclear whether reducing particle concentrations and removing bioaerosols from the air will result in reduced transmission of respiratory infections.
+\noindent We searched Scopus on October 19, 2023, for all publications from inception using the terms ("air filt" OR "air clean") AND ("infection" OR "transmission").
+Several studies have shown that air cleaners reduce particle concentrations in indoor air.
+Studies also showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in hospitals and other indoor settings.
+One study reported a lower incidence of SARS-CoV-2 in schools using different ventilation strategies, with or without additional air filtration devices.
+Several simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission.
+One real-world study during the COVID-19 pandemic was inconclusive regarding the effectiveness of air cleaners in a school setting.
+\medskip
 
-We used a multiple-measurement approach to study the transmission of viral respiratory viruses and the effect of air cleaners in school rooms.
-We employed a cross-over study design under non-pandemic conditions in the winter of 2023.
-We collected epidemiological (absences likely related to respiratory infections), environmental (CO$_2$, aerosol and particle concentrations), audio (recordings to detect coughing), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland.
-We determined changes in particle concentrations, respiratory virus positivity rates, and modeled the risk of infection.
+\noindent \textbf{Added value of this study} \smallskip
+
+\noindent To date, it remains unclear whether the reduction in particle concentrations with air cleaners also translates into lower risks of respiratory infection.
+This study used a multiple-measurement approach to estimate the risk of respiratory virus infection in a Swiss school and assessed the effectiveness of air cleaners in a cross-over study design.
+Schools are particularly important because students spend much of their time indoors and may introduce infections from and into their households.
+We collected epidemiological data, environmental data, audio recordings, and molecular data over seven weeks during non-pandemic conditions in winter 2022/2023 in a Swiss school.
+Air cleaners in classrooms improved air quality (lower particle concentrations), were associated with fewer symptoms (less coughing), and showed a potential benefit in reducing absences related to respiratory infections.
+We found many respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus.
+Although airborne detection of viruses was rare, spatiotemporal analysis of students' saliva samples indicated that transmission was likely in school rooms.
+\medskip
+
+\noindent \textbf{Implications of all the available evidence} \smallskip
+
+\noindent Air cleaners have a tangible positive impact on air quality, significantly reducing particle concentrations.
+While the individual-level reduction in the risk of respiratory infections due to air cleaners may be relatively modest, the collective benefit at the population level, in terms of reducing illnesses and preventing school absences, is likely to be substantial.
+Future research should examine the cost-effectiveness and feasibility of implementing air cleaners in school and other communal settings.
+
+\thispagestyle{empty}
+\sloppy
+\raggedbottom
 
 \newpage
+%TC:endignore
 
-\section{Methods}
+\setcounter{page}{1}
 
-\subsection{Study setting and design}
+\section*{Introduction}
 
-\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from 26~January to 11~March 2023.
-\Cref{fig:study-setup} shows the schematic study setup and Table~\zref{tab:data} in \supp \, provides an overview of the types of data collected in each classroom.
+Transmission of respiratory infections such as SARS-CoV-2 and influenza are difficult to mitigate and control.
+Person-to-person transmission occurs primarily through the release of respiratory particles containing the viruses.
+Recently, the focus has been on small respiratory particles called aerosols, which have been found to carry the majority of viruses during respiratory activities.\cite{Fennelly2020} Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and travel long distances.\cite{Wang2020}
 
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{../study_setting.pdf}
-    \caption{\textbf{Study setting}.
-Schematic study setup of the classrooms.
-One air cleaner was placed in the front and one in the back of the classrooms.
-All devices were placed at the head level of the students when they were seated.
-Both classrooms were not equipped with an active HVAC (Heating, Ventilation, Air conditioning) system, but were ventilated naturally by opening windows.
-}
-    \label{fig:study-setup}
-\end{figure}
+Improved ventilation systems are critical for a healthy indoor environment and can reduce the risk of respiratory transmission,\cite{Wang2021,Morawska2021} especially in schools where students spend much of their time indoors during the week.
+Portable HEPA-air filtration devices (air cleaners) may be another cost-effective alternative to upgrading ventilation systems.
+Still, the effects of air cleaners on the transmission of respiratory infections are less clear.
+Several studies have shown that air cleaners reduce particle concentrations in indoor air,\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed} which is associated with a reduced risk of respiratory infection and mortality.\cite{Gordon2014Resp,Kelly2023Atmos,DeAngelis2021EnvRes} In line with this, a population-level study reported a lower incidence of SARS-CoV-2 in US elementary schools using different ventilation strategies, with or without additional air filtration devices.\cite{Gettings2021} Although some studies could not find an association between particle concentrations and viral RNA load in airborne samples,\cite{Nor2021SciRep,Hanna2023PONE} recent studies showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in hospitals and other indoor settings.\cite{Morris2022,Ueki2022mSphere,Myers2022IndoorAir} Simulation studies have further demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission of SARS-CoV-2,\cite{Lindsley2021} and other respiratory viruses.\cite{Cortellessa2023Build} However, most simulation studies assume that the detection of RNA equals transmissible virus, despite recent data showing a relevant loss of viral infectivity in respiratory particles over time.\cite{Oswin2022PNAS} To date, it remains unclear whether reducing particle concentrations and removing bioaerosols will reduce indoor transmission of respiratory infections.
 
-\subsection{Study intervention}
+We used a multiple-measurement approach to study the transmission of respiratory viruses under non-pandemic conditions and the effect of air cleaners in schools with a cross-over study design in the winter of 2022/2023.
+We collected environmental data (CO$_2$, particle concentrations), epidemiological data (absences likely related to respiratory infections), audio recordings (coughing), and molecular data (detection of viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2023 in two classes of a secondary school in Switzerland.
+We determined changes in particle concentrations, absences related to respiratory infections, coughing, and the positivity rate of saliva samples.
 
-\noindent We employed a cross-over design to study the effectiveness of air cleaners, distinguishing two study conditions in each class: weeks with and weeks without air cleaners being installed in each classroom (\Cref{tab:study_design}).
-Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China; approx.
-USD\,250 per device).
-According to the manufacturer, these air cleaners run at a clean air delivery rate of 2$\,\times\,$600 m$^{3}$/h.
-% We measured a lower clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h when testing the devices in an empty classroom.
-Window ventilation was done at the discretion of the teachers, as there were no specific recommendations from the national health authorities at that time.
+\section*{Methods}
 
-\begin{table}[!htpb]
-    \footnotesize
-    \centering
-    \caption{\textbf{Study design.} Cross-over design describing when portable air cleaners (AC) were installed in the rooms of classes A and B during a seven-week study period from 16~January to 11~March 2023, excluding a week of vacation from February 6 to 11.
-A ``*'' indicates when swabs from the HEPA filters were taken, i.e.
-after each intervention phase and before the vacation.}\label{tab:study_design}
-    \begin{tabular}{l l l l l l l l l}
-    \toprule
-      & \textbf{Week 1} & \textbf{Week 2} & \textbf{Week 3} & Vacation & \textbf{Week 4} & \textbf{Week 5} & \textbf{Week 6} & \textbf{Week 7} \\
-      & Jan 16-21 & Jan 23-28 & Jan 30-Feb 3 & Feb 6-11 & Feb 13-18 & Feb 20-25 & Feb 27-March 3 & March 6-11 \\
-      \midrule
-      \textbf{A} & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{0000}*& \cellcolor{gray!10} None & & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{00009}* \\
-      \textbf{B} & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC\hphantom{0000000}* & & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{0000}*& \cellcolor{gray!10} None & \cellcolor{gray!10} None \\
-      \bottomrule
-    \end{tabular}
-\end{table}
+\subsection*{Study setting and design}
 
-\subsection{Data collection}
+\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, for seven weeks from January~16 to March~11, 2023.
+Figure~\ref{fig:study-setup} shows the schematic study setup.
+This study is reported as per the Strengthening the Reporting of Observational Studies in Epidemiology guideline (see STROBE checklist).
+
+\subsection*{Study intervention}
+
+\noindent We used a cross-over design to study the effectiveness of air cleaners (\cref{tab:study_design}).
+Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China).
+According to the manufacturer, these air cleaners run at clean air delivery rates of 2$\,\times\,$600 m$^{3}$/h.
+When testing the devices in an empty classroom with sub-micrometer sized particles, we measured a lower effective clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h (\supp~pp~\zpageref{sec:cadr}$-$5).
+
+\subsection*{Data collection}
+
+An overview of the collected data is provided in \supp~p~\zpageref{tab:data}. \medskip
+
+\noindent\textbf{Environmental data} \smallskip
+
+\noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute.\cite{Banholzer2023PLoSMed} The devices did \emph{not} alert teachers when CO$_2$ levels were too high.\medskip
 
@@
 
-\noindent At the beginning of the study, we collected aggregated data on age, sex and COVID-19 vaccination and recovery status in the participating classes.
-On a daily basis, we collected data on each absent student, \ie the absence and return dates as well as the reason for the absence (illness, career days, or other).
-For absences due to illness, we collected symptoms and the date of symptom onset.
-We also collected data on laboratory tests but none of the students reported a test result during the study.
-Based on the epidemiological data, we defined a case of respiratory infection as an absence in which the student reported an illness with at least one of the following symptoms: fever, cough, fatigue, loss of taste or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, and stomach.
-Students were asked to report the first date they experienced symptoms, so that a student absent on Monday could report Saturday or Sunday as the day when symptoms began to be felt.
-We will refer respiratory cases by their dates of symptom onset, which usually corresponded to the absence date, unless the student attended the school while experiencing symptoms.
-Informed consent for one student could not be obtained and the single absence of this student was considered unrelated to a respiratory infection.
-Reports about absences were entered electronically into a REDcap database\cite{Harris2009,Harris2019}.
-The line list data is shown in Table~\zref{tab:epi-data-line-list} in \supp.
+\noindent At the beginning of the study, we collected aggregated data on age, sex, COVID-19 vaccination, and COVID-19 recovery status in the participating classes.
+Daily, we collected data on each absent student, \ie the absence and return dates and reason for the absence (illness, career days, or other).
+For absences due to illness, we recorded symptoms and the date of symptom onset.
+We defined a case of respiratory infection as an absence in which the student reported an illness with at least one respiratory symptom (\supp~p~\zpageref{sec:case-data}).
+The list of absences by date of symptom onset, absence and return, is shown in \supp~pp~\zpageref{tab:epi-data-line-list}$-$8.\medskip
 
-In addition, both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below).
-Saliva samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Galar2021,To2019,Huber2021}.
-The line list data is shown in Table~\zref{tab:mol-data-line-list} in \supp.
+\subsection*{Audio recordings and cough detection}
+
+We installed a portable audio recorder (ZOOM H6; New York, USA) in each classroom to record sounds continuously.
+We determined the number of coughs per minute using an AI algorithm.\cite{Bertschinger2023IEEE} A recent study showed a significant correlation between coughing and the positivity rate of virus detection in airborne samples.\cite{Hanna2023PONE}
+
+\noindent\textbf{Molecular data analyses} \smallskip
+
+\noindent Both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) saliva testing.
+Samples were transported to the laboratory and stored at $-$80°C until further processing.\cite{Huber2021} The list of positive samples is shown in \supp~pp~\zpageref{tab:mol-data-line-list}$-$21.
+Furthermore, we collected airborne respiratory viruses in both classrooms with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL Phosphate-Buffered Saline.
+The Coriolis Micro Air ran shortly before and during break times (approximately 60\,min/day) to minimize noise.
+In one class, we also sampled with the BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft.
+Collins, CO, USA).\cite{Lednicky2016AST} BioSpot-VIVAS operated throughout the lessons.
+The removable parts of both sampling devices were regularly autoclaved.
+At the end of the day, samples were transported to the Institute of Infectious Diseases and stored at $-$80°C.
+Finally, we collected swabs from the HEPA filters of the air cleaners after each intervention phase (see \cref{tab:study_design}).
+The HEPA filters were removed from the air cleaners and divided into 20~fields.
+One sterile Phosphate-Buffered Saline-moistened swab per field was then taken for a total of 20~swabs per filter.
+
+Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL.
+The human saliva samples were analyzed directly without prior filtration.
+The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19 major respiratory viruses and viral subtypes, including SARS-CoV-2 (CoV), influenza (IFA/IFB), respiratory syncytial (RSV), metapneumovirus (MPV), adenovirus (AdV), rhinovirus (HRV), and parainfluenza (PIV).
+
+We also performed molecular genotyping for positive saliva, bioaerosol, air filter samples of adenovirus and influenza.
+For adenovirus, we used a PCR procedure to amplify three hypervariable regions (hexon, fiber, penton) of the viral genome and then sequenced the PCR products.\cite{Akello2021SciRep} For influenza, we tried to determine the molecular transmission network using whole genome sequencing.\cite{Kelly2022FrontiersImmuno}
+
+\subsection*{Statistical analyses and modeling}
+
+All statistical analyses were described in a statistical analysis plan,\cite{Banholzer2023SAP} which was published prior to the analyses.
+Bioaerosol samples and viral load concentrations could not be analyzed because there were too few positive samples.
+Otherwise, there were no major deviations from the statistical analysis plan.
+Minor deviations are documented in \supp~pp~\zpageref{sec:env-regression-model}$-$\zpageref{sec:multinomial-model}, including detailed descriptions of the models.
 \medskip
 
-\noindent\textbf{Environmental data} \smallskip
+\noindent\textbf{Particle concentrations} \smallskip
 
-\noindent \emph{CO$_2$ and particle measurements}
-
-\noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute.
-This device was used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}.
-A transient mass balance model was used to estimate the air change rate from indoor CO$_2$ levels\cite{Batterman2017IJERPH}, assuming that classrooms were fully occupied during lessons and that room occupancy increased (decreased) linearly 10~minutes before (after) the start (end) of a lesson.
-Particle concentrations were filtered according to the times when students were in the classroom, which were monitored together with the times during which windows were opened.
-\medskip
-
-\noindent \emph{Bioaerosol sampling and HEPA filters of the air cleaners}
-
-\noindent We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL PBS, as previously reported in clinic settings\cite{Moore2021}.
-The Coriolis Micro Air could only operate shortly before and during break times to prevent unacceptable noise exposure (approximately 60\,min/day).
-In one class, we also used the BioSpot-VIVAS device sampler (Aerosol Devices Inc., Ft.
-Collins, CO, USA).
-This device samples airborne virus particles into a viral transport medium (VTM) using a water-based laminar flow condensation method\cite{Pan2016JAM,Lednicky2016AST}.
-BioSpot-VIVAS could operate throughout lessons.
-Removable parts of both sampling devices were regularly autoclaved to prevent contamination.
-At the end of the day, samples were transported to the Institute of Infectious Diseases (IFIK) and stored at $-$80°C.
-After each intervention phase (see \Cref{tab:study_design}), the HEPA filters were carefully removed from the air cleaners, divided into 20~fields, and one swab per field was taken for a total of 20~swabs per filter.\medskip
-
-\subsection{Laboratory and molecular analyses}\label{sec:mol_analyses}
-
-\noindent Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL.
-The human saliva samples were directly analyzed without prior filtration.
-The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19~respiratory viruses: SARS-CoV-2 (CoV), Influenza A virus (Flu A), Influenza B virus (Flu B), Human respiratory syncytial virus A/B (RSV), Human metapneumovirus (MPV), Human adenovirus A/B/C/D/E/F (AdV), Human rhinovirus A/B/C (HRV), and Human parainfluenza virus 1/2/3/4 (PIV) in one single reaction.
-%/Viral genome load (VGL) of specimens was quantified (copies/L) by curve extrapolation derived from standardized ATCC quantitative genomic standard dilutions (ATCC-VR-3347D, LGC Standards, LGC Group UK).
-For the BioSpot-VIVAS air samples, the air concentration of the DNA was exactly calculated according to the airflow influx into the sampling device.
-%/
-The Limit of detection (LOD) of the assay SARS-CoV-2, IFA/IFB is 7.5x10$^0$ TCID/ml (SARS-CoV-2), 3.7x10$^{-1}$ TCID/ml (IFA), and 4.8x100 TCID/ml (100cp/ml) (IFB) respectively.
-
-\subsection{Molecular genotyping}
-
-We performed molecular genotyping for positive samples of AdV and Flu~B.
-For AdV (saliva, bioaerosol, and air filter samples), we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}.
-For Flu~B, we tried to determine the molecular transmission network using whole genome sequencing\cite{Kelly2022FrontiersImmuno}.
-
-\subsection{Audio recordings and cough detection}
-
-We installed one portable audio recorder (ZOOM H6; New York, USA) in each classroom to record sounds continuously.
-We detected the number of coughs per minute using an AI algorithm based on a convolutional neural network\cite{Bertschinger2023IEEE}, which was recently trained and validated with audio data from the waiting room of a primary care clinic\cite{Zurcher2022}.
-
-\subsection{Statistical analyses and modeling}
-
-All statistical analyses were described in the statistical analysis plan (SAP)\cite{Banholzer2023SAP}, which was published before analyses began.
-In contrast to the SAP, the number of positive bioaerosol samples and the viral load concentration of positive samples could not be analyzed because there were too few positive samples during the study.
-Apart from that there were no major deviations from the SAP regarding the conducted analyses.
-There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in Supplementary Text~\zref{sec:transmission-model} to Supplementary Text~\zref{sec:env-regression-model} in \supp, which provide detailed descriptions of the models used in each analysis.
+\noindent We compared daily mean particle concentrations between study conditions.
+We estimated the reduction in particle concentrations with air cleaners using Bayesian log-linear regression models, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases (\supp~p~\zpageref{sec:env-regression-model}).
 \medskip
@@
 
-\noindent We estimated the relative risk of infection with air cleaners using a Bayesian latent hierarchical regression model.
+\noindent We estimated the relative risk of infection with air cleaners using a Bayesian latent variable hierarchical regression model.
 The observed outcome of this model is the number of new respiratory cases $C$ (absences related to respiratory infections by date of symptom onset) on day $t$ in class $j$, which are modeled with a Negative Binomial distribution.
 The expected number of new cases is the weighted sum of the number of new infections $I_{js}$ (unobserved/latent outcome) in the previous days $s<t$, with the weights corresponding to the probability distribution of the incubation period.
-The number of new infections is related to the presence of air cleaners using a regression-like log-link
+The number of new infections is related to the presence of air cleaners as follows:
 \begin{align}
@@
 where $F_{js}$ is the number of infections in the previous week (a proxy for the number of infectious students), $N_{js}$ is the cumulative number of infections (a proxy for the number of susceptible students), $\beta_0$ is the infection rate without air cleaners, and $\beta_1$ is the effect of air cleaners.
-Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the community (\ie the canton of Solothurn).
-
-A detailed description of the model and choice of priors for all model parameters are provided in Supplementary Text~\zref{sec:transmission-model} in \supp.
-Importantly, since multiple viruses were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the virus-specific distributions.
-The weights were inferred based on the proportion of positive saliva samples for each virus per week.
-As a result, the distribution of the incubation period used for modeling could change every week depending on the proportion of viruses that were detected in the students' saliva.
-Moreover, respiratory cases by date of symptom onset were under-reported for weekends and thus a re-weighting scheme was used to account for weekday effects.
+Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the canton.
+A detailed description of the model and choice of priors for all model parameters are provided in \supp~pp~\zpageref{sec:transmission-model}$-$27.
 \medskip
 
-\noindent\textbf{Number of positive saliva samples} \smallskip
+\noindent\textbf{Coughing} \smallskip
 
-\noindent Molecular analysis determined which saliva samples were positive for a respiratory virus.
-We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (Supplementary Text~\zref{sec:multinomial-model} in \supp) where the expected number of positive samples for a virus was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each virus.
-The number of susceptibles was computed as the difference between the number of students per class and the cumulative number of positive samples.
-As a sensitivity analysis, we lagged the test results once by one test date (\ie from Tuesday to last Thursday and from Thursday to last Tuesday) and once by one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday), to consider the unknown delay when students were infected and tested positive.
+\noindent We compared the daily number of detected coughs between study conditions.
+We estimated the reduction in the number of coughs with air cleaners using a Bayesian Negative Binomial regression model, using time in class as the model offset and adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+In addition, we used a hierarchical regression model to estimate the association between the number of coughs and the virus-specific number of positive saliva samples.
+For this analysis, we used only days on which saliva samples were collected (Tuesdays and Thursdays, for a total of 27~days).
+A detailed description of these models is provided in \supp~p~\zpageref{sec:aud-regression-model}.
 \medskip
 
-\noindent\textbf{Aerosol and particle concentrations} \smallskip
+\noindent\textbf{Saliva samples} \smallskip
 
-\noindent We computed the average particle concentrations per day and compared them between study conditions.
-We estimated the reduction in particle concentrations using Bayesian log-linear regression models (Supplementary Text~\zref{sec:env-regression-model} in \supp), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
+\noindent We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~p~\zpageref{sec:multinomial-model}) and linked the expected number of positive samples to the presence of air cleaners, adjusting for the cumulative number of positive tests.
 \medskip
 
-\noindent\textbf{Number of detected coughs} \smallskip
+\noindent\textbf{Software}\smallskip
 
-\noindent We computed the daily number of detected coughs and compared it between study conditions.
-We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model (Supplementary Text~\zref{sec:aud-regression-model}) with time in class as the model offset, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
-In addition, we used a Bayesian Negative Binomial hierarchical (random effects) regression model to estimate the association between the number of coughs and the number of positive saliva test results for respiratory viruses.
-For this analysis, we only used the days were saliva testing was performed (i.e.
-every Tuesday and Thursday, 27~days in total).
-\medskip
+\noindent All analyses were performed in R software (version 4.2.0).
+Model parameters were estimated in in Stan (version 2.21.0) with a Bayesian approach, using the Hamiltonian Monte Carlo algorithm with the No-U-Turn Sampler (NUTS).\cite{Hoffman2014} For each outcome, we report the posterior probability of a reduction with air cleaners.
+The estimated reduction is reported with the posterior mean and 95\%credible interval (CrI).
+Preprocessed data and code is available from \url{https://osf.io/38j9g}.
 
-\noindent\textbf{Software and model estimation} \smallskip
-
-\noindent All analyses were performed in R software (version 4.2.0)\cite{RCoreTeam2022} and modeling in Stan (version 2.21.0)\cite{Carpenter2017}.
-Model parameters were estimated with a Bayesian approach.
-Specifically, Markov chain Monte Carlo (MCMC) sampling was used as implemented by the Hamiltonian Monte Carlo algorithm with the No-U-Turn Sampler (NUTS)\cite{Hoffman2014}.
-Each model is estimated with 4 Markov chains and 2,000 iterations of which the first 1,000 iterations are discarded as part of the warm-up.
-Estimation power is evaluated via the effective sample size (ESS) and convergence of the Markov chains is evaluated with the Gelman-Rubin convergence diagnostic ($\hat{R}$).
-If not stated otherwise, we report posterior means and credible intervals (CrIs) based on the 50\%, 80\%, and 95\% quantiles of the posterior samples, respectively.
-The code is available from \url{https://osf.io/38j9g}.
-
-\subsection{Ethics statement}
+\subsection*{Ethics statement}
 
@@
 
+\subsection*{Role of the funding source}
+
+The funder of the study had no role in study design, data collection, data analysis, data interpretation, or writing of the report.
+
 \newpage
 
-\section{Results}
+\section*{Results}
 
-The study population consisted of 38~students (19~female, 19~male; Table~\ref{tab:cases-overview-school}).
-Of these, 7~students were vaccinated or recovered from an infection within SARS-CoV-2 in the last four months.
+The study population consisted of 38~students (age 13$-$15 years, 19/19~female/male; \cref{tab:cases-overview-school}).
+Seven~students had been vaccinated or recovered from a SARS-CoV-2 infection within the last four months.
 During the seven-week study period (total of 1,330 student-days), students were absent from school for 220~days (18\% of the total) of which 129~days (59\% of absences) were due to illness.
+
+\subsection*{Air quality}
+
+The mean aerosol number concentration was 95\,1/cm$^3$ (standard deviation [SD] 81\,1/cm$^3$) without vs 27\,1/cm$^3$ (SD 34\,1/cm$^3$) with air cleaners (\cref{fig:infection-risk}a).
+The Bayesian regression model suggested a clear reduction in the aerosol concentration with air cleaners, with a posterior probability of 100\%.
+The model-estimated decrease was 76\% (95\%-CrI 63\% to 86\%), which was greater for larger (PM$_{10}$) than for smaller (PM$_{1-4}$) particles (\supp~p~\zpageref{fig:palas-results}).
+Daily mean CO$_2$ levels were comparable between study conditions (1,636\,ppm (SD 341\,ppm) without vs 1,769\,ppm (SD 391\,ppm) with air cleaners).
+There was little change in other environmental variables (\supp~p~\zpageref{fig:env-descriptives-other-vars}).
+
+\subsection*{Risk of infection}
+
+Absences related to respiratory infections included 22~cases without vs 13~cases with air cleaners (\cref{fig:infection-risk}b).
+The Bayesian latent variable hierarchical regression model suggested that air cleaners reduced the risk of infection, with a posterior probability of 91\%.
+The adjusted relative risk of infection with air cleaners was 0.73 (95\%-CrI 0.44 to 1.18).
+The estimated number of respiratory infections in school would have been 19 (95\%-CrI 9 to 37) if air cleaners had been installed throughout the study period, compared to 36 (95\%-CrI 12 to 92) infections if air cleaners had not been installed.
+Detailed estimation results are provided in \supp~pp~\zpageref{sec:detailed-redcap}$-$34.
+
+\subsection*{Coughing}
+
+On average, we detected 3.1\,coughs/min (SD 1.2\,coughs/min) without vs 2.6\,coughs/min (SD 1.1\,coughs/min) with air cleaners (\cref{fig:infection-risk}c).
+The Bayesian model suggested that coughing was less frequent with air cleaners, with a posterior probability of 93\%.
+The adjusted relative risk of coughing with air cleaners was 0.93 (95\%-CrI 0.85 to 1.02).
+Coughing was associated with virus-specific transmission (\supp~p~\zpageref{fig:coughing-association}).
+
+\subsection*{Saliva samples}
+
+We analyzed a total of 448~saliva samples.
+We detected 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus, respectively (\cref{fig:molecular-descriptives}a).
+There were 25~positive saliva samples in both study conditions (\cref{fig:infection-risk}d) and, based on the Bayesian model, the posterior probability that air cleaners reduced the positivity rate was 65\%, with an adjusted relative risk of 0.93 (95\%-CrI 0.49 to 1.61).
+The estimated relative risk was not sensitive to infection to testing delays (\supp~p~\zpageref{fig:mol-estimation-results-sensitivity}).
+
+\subsection*{Transmission patterns}
+
+The distribution of positive saliva samples varied between classes.
+For example, all but one sample was positive for adenvovirus in class~A during the first three study weeks, while the vast majority of positive influenza~B samples were found in class~B.
+To illustrate possible transmission chains within classes, we linked positive saliva samples of the same virus that were less than one week apart.
+Based on this spatiotemporal analysis, we identified 10~possible transmission chains (\cref{fig:molecular-descriptives}b).
+The longest potential transmission chains occurred in January, referring to a cluster of adenovirus infections in class~A and influenza~B infections in class~B.
+Molecular genotyping to verify the proposed transmission network was unsuccessful because we could not amplify and sequence any of the gene targets.
+Furthermore, the rate of positive bioaerosol samples was low.
+We also analyzed 105~bioaerosol samples and detected in two of them viral RNA (1~rhinovirus in class~A and 1~adenovirus in class~B).
+Similarly, we detected 1~influenza~B, 1~rhinovirus, 1~adenovirus, and 1~SARS-CoV-2 in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
+
+%Nevertheless, these estimates are subject to unmeasured confounding because the delay may be sample-specific and repeated positive tests from the same person could not be precluded because testing was anonymized.
+
+\section*{Discussion}
+
+% summary
+
+We used a multiple-measurement approach within a cross-over study design to estimate the risk of respiratory virus infection in a Swiss school and to assess the effectiveness of air cleaners.
+We found a wide range of respiratory viruses in saliva samples, mainly adenovirus, influenza B, and rhinovirus, but very few viral RNA were detected in bioaerosol samples and on the filters of the air cleaners.
+Particle mass concentrations decreased significantly with air cleaners, and Bayesian modeling based on epidemiological data indicated a reduction in the relative risk of infection with air cleaners.
+Coughing was reduced, compatible with air cleaners preventing some symptomatic infections.
+
+% comparison with previous study
+
+We detected a range of respiratory virus in students' saliva, mainly adenovirus, influenza~B and rhinovirus, with only three positive SARS-CoV-2 saliva samples.
+In a previous, similar study in the same setting, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 omicron wave in the winter of 2021/2022 and detected almost exclusively SARS-CoV-2 in the students' saliva.\cite{Banholzer2023PLoSMed} A similar shift in the pattern of respiratory viruses has been observed in other studies.\cite{Nygaard2023Lancet,Sauteur2022EuroSurv} We found a reduction in the risk of SARS-CoV-2 infection for mask wearing, but not for air cleaners, possibly because the air cleaners were introduced only at the end of the study, when most students were already infected with SARS-CoV-2.
+
+% discussing the effect of air cleaners
+
+It is well documented that air cleaners improve indoor air quality.\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed} There are several reasons why the effect of air cleaners is probably smaller than universal mask wearing, which has been shown to be a very effective infection control measure.\cite{Banholzer2023PLoSMed,Gettings2021,Leung2020NatMed,Milton2013PLoSPathogens} Unlike masks, air cleaners cannot prevent transmission outside the classroom or transmission due to close range, high particle density.
+Prolonged and close contact may be necessary for transmission of some respiratory viruses,\cite{Leung2020NatMed,Brankston2007LancetID} or make transmission more likely despite prior vaccination or infection.\cite{Lind2023NatCommun} Our results further suggests that air cleaners are more effective at removing larger particles ($>5\mu$m), which also explains the difference between our measured and the manufacturer's reported clean air delivery rate (see \supp~pp~\zpageref{sec:cadr}$-$5).
+However, many respiratory viruses are carried in smaller particles, which are more relevant for transmission ($\leq5\mu$m).\cite{Fennelly2020} Finally, classroom activity, airflow and other unobserved, confounding factors make it challenging to evaluate the effects of air cleaners on transmission in real-world settings.
+Nevertheless, it is important to evaluate their effectiveness in such settings and to compare them with the hypothetical effects suggested by simulation studies.\cite{Lindsley2021,Cortellessa2023Build}
+
+%Although the risk reduction effect estimated in this study is rather small, the cumulative impact may be relevant because minimizing the risk of transmission of respiratory infections in schools is an important public health consideration\cite{Beale2023JOMT} and because prevention and control measures in schools, such as air cleaners, must be adequate and acceptable\cite{WHO2020SchoolMeasures}.
+
+% economic and psychological impact of air cleaners
+
+The beneficial effects of air cleaners on indoor air quality and transmission come at a reasonable cost.
+The portable air cleaners used in our study cost approximately USD\,250 per unit.
+Their operating cost effectiveness in providing clean air could be even higher than that of a ventilation system when compared in parallel using the same air delivery ratings.\cite{Noh2016EnBuild} Therefore, air cleaners could be a cost-effective public health measure, particularly during pandemics or epidemics when there is greater exposure to respiratory infections.
+In addition, air cleaners can have a psychological impact.
+On the one hand, they can give students and teachers a sense of support during difficult times.
+For example, it has been shown that becoming infected was one of the main concerns for students during the pandemic.\cite{Yuerekli2022IJERPH} A safe and supportive environment is a key determinant of student success and well-being in school.\cite{Kutsyuruba2015RevEduc} On the other hand, air cleaners may not be well accepted by teachers and authorities due to noise, limited space, technical issues, and the required maintenance.\cite{Sanguinetti2022IndoorAir} Therefore, investments in professional building ventilation systems are still preferred in the long run.\cite{Nardell2016}
+
+% transmission within classroom
+
+We detected only few respiratory viruses in bioaerosol samples (1~sample of adenovirus and 1~sample of rhinovirus) and on the filters of the air cleaners (4~positive samples in 160~swabs).
+The low rate of positive bioaerosol samples may indicate that it is unlikely that airborne transmission occurred in classrooms.
+It may be possible that students had relatively little exposure to respiratory viruses at school and acquired their infections elsewhere.
+However, the distribution of positive saliva samples markedly differed between classes.
+Adenovirus spread in class~A during the first three study weeks, with only two infections of influenza~B over the study period.
+In contrast, influenza~B spread throughout the study in class~B, and adenovirus infections were detected only in the last week of the study.
+Furthermore, adenovirus infections tend to be mild,\cite{Kunz2010CIDR} and less frequently associated with cough than influenza,\cite{Ma2018RMV} which is consistent with the comparatively lower frequency of coughing in class~B.
+Taken together, the class-specific, spatiotemporal patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred within the classrooms.
+
+% limitations
+
+Our study has limitations.
+Aerosol measurements and molecular detection of viruses in bioaerosol samples document exposure, but not transmission and the direction of transmission (person to air, air to person) cannot be determined.
+Further, the reasons for school absences were self-reported by students and some absences may have been incorrectly attributed to respiratory infections.
+In addition, we could only approximate the incubation period for each epidemiological case.
+Finally, the study was conducted in two classrooms of one Swiss school and although results are likely to apply to many settings in Switzerland and other European countries, they will not be applicable to settings in the global South.
+
+% conclusion
+
+In conclusion, using epidemiological, environmental, and molecular data, our study showed that a wide range of respiratory viruses, but rarely SARS-CoV-2, was detected in students under non-pandemic conditions when public health measures were lifted, and that transmission of these respiratory viruses likely occurred in schools.
+Airborne detection of non-SARS-CoV-2 respiratory viruses was rare, suggesting that respiratory viruses other than SARS-CoV-2 may be more difficult to detect in the air and that prolonged close contact may be required for transmission.
+Air cleaners improved air quality by significantly reducing particle concentrations.
+The reduction in risk of respiratory infections conferred by air cleaners may be modest at the individual level, however, the benefit at the population level in terms of illness and school absences prevented will likely be important.
+Future studies should examine the relative importance of prolonged close contact and long-range transmission of respiratory viruses in schools and other congregate settings, and examine the cost-effectiveness of using air cleaners in these settings.
+
+\newpage
+
+%TC:ignore
+
+\section*{Author contributions}
+Conception and design: NB, KZ, LF, PB, PJ, TS.
+Epidemiological and environmental data collection: EW, NB, PJ, KZ, TS, LF.
+Laboratory data collection: PB, LFu.
+Additional data collection: TH.
+Cough detection: SB.
+Molecular genotyping: AR, PB, LFu.
+Statistical analysis: NB.
+Paper draft: NB, LF, ME.
+All authors reviewed and approved the final version of the manuscript.
+
+\section*{Acknowledgements}
+We would like to thank the schools, teachers, and students participating in the study.
+We are grateful to the Educational Department of the canton of Solothurn for their support throughout the study.
+We would also like to thank Ronald Dijkman for the influenza molecular genotyping and the biosecurity team (Julia Feldmann, Monika Gsell, Kathrina Summermatter) from the Institute for Infectious Diseases at the University of Bern for their assistance with the bioaerosol devices.
+Finally, we are indebted to the student assistants (Khadija Ahmed, Marie-Joséphine Brancato, Michelle Bürki, Santiago Martinez, Moric Toszeghi, Sylvain Wasmer) who helped with the data collection in the schools.
+
+\section*{Data sharing statement}
+Preprocessed data and code is available from \url{https://osf.io/38j9g}.
+
+\bibliography{references.bib}
+%TC:endignore
+
+\clearpage
+
+\begin{table}[!htpb]
+    \footnotesize
+    \centering
+    \caption{\textbf{Cross-over study design.} Description of when portable air cleaners (PAC) were installed in the rooms of classes A and B during a seven-week study period from January~16 to March~11, 2023, excluding a week of vacation from February 6 to 11.}\label{tab:study_design}
+    \begin{tabular}{l l l l l l l l l}
+    \toprule
+      & \textbf{Week 1} & \textbf{Week 2} & \textbf{Week 3} & Vacation & \textbf{Week 4} & \textbf{Week 5} & \textbf{Week 6} & \textbf{Week 7} \\
+      & Jan 16-21 & Jan 23-28 & Jan 30-Feb 3 & Feb 6-11 & Feb 13-18 & Feb 20-25 & Feb 27-Mar 3 & Mar 6-11 \\
+      \midrule
+      \textbf{A} & \cellcolor{gray!50} PAC & \cellcolor{gray!50} PAC\hphantom{000}*& \cellcolor{gray!10} None & & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} PAC & \cellcolor{gray!50} PAC\hphantom{0000}* \\
+      \textbf{B} & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} PAC\hphantom{00000}* & & \cellcolor{gray!50} PAC & \cellcolor{gray!50} PAC\hphantom{000}*& \cellcolor{gray!10} None & \cellcolor{gray!10} None \\
+      \bottomrule
+      \multicolumn{9}{l}{\scriptsize *~Swabs from the HEPA filters taken after each intervention phase and before the vacation.}
+    \end{tabular}
+\end{table}
+
+\clearpage
 
@@
 
-\subsection{Molecular analyses of saliva and air samples}
-
-We analyzed 448~saliva samples over the study across classes.
-We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenovirus (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~virus~(PIV) (\Cref{fig:molecular-descriptives}a).
-There was an equal number of positive samples with and without air cleaners (\Cref{fig:molecular-descriptives}b).
-Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (Figure~\zref{fig:mol-estimation-results-sensitivity} in \supp).
-The distribution of positive samples varied between classes (\Cref{fig:molecular-descriptives}c).
-For example, during the first three study weeks, all but one sample was positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B.
-We also analyzed 105~bioaerosol samples and detected  two respiratory viruses in the air (1~HRV in class~A and 1~AdV in class~B).
-Finally, we detected 1~Flu~B, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
+\clearpage
 
@@
     \centering
-    \includegraphics{../../results/mol-data/descriptives.pdf}
-    \caption{\textbf{Analysis of molecular data}.
-Number of positive saliva samples by \textbf{(a)}~virus, \textbf{(b)}~study condition, and \textbf{(c)}~study week and class.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
+    \includegraphics{../study_setting.pdf}
+    \caption{\textbf{Study setting}.
+Schematic study setup of the classrooms.
+One air cleaner was placed in the front and one in the back of the classrooms.
+All devices were placed at the head level of the students when they were seated.
+Both classrooms lacked an active HVAC (Heating, Ventilation, Air conditioning) system, but they were ventilated naturally by opening windows at the discretion of the teachers.
+}
+    \label{fig:study-setup}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[!htpb]
+\centering
+    \includegraphics{../../results/env-data/aerosol-number-boxplot.pdf}\hspace{.5cm}
+    \includegraphics{../../results/epi-data/cases_by_condition.pdf}
+    \includegraphics{../../results/cough-data/coughs-frequency-by-condition.pdf}\hspace{.5cm}
+    \includegraphics{../../results/mol-data/saliva-by-study-condition.pdf}
+    \caption{\textbf{Comparison of outcomes with vs without air cleaners.} At the top of each plot, the posterior probability for a reduction with air cleaners is shown, based on the Bayesian model.
+\textbf{(a)}~Daily average aerosol number concentrations as boxplots.
+\textbf{(b)}~Number of respiratory cases.
+\textbf{(c)}~Daily average number of detected coughs per minute as boxplots.
+\textbf{(d)}~Number of positive saliva samples.}
+    \label{fig:infection-risk}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{../../results/mol-data/saliva-distribution.pdf}\hspace{.5cm}
+    \includegraphics{../../results/mol-data/network-plot.pdf}
+    \caption{\textbf{Molecular detection of respiratory viruses and transmission network based on spatiotemporal analysis of students' saliva samples}.
+\textbf{(a)}~Number of positive saliva samples by virus.
+\textbf{(b)}~Daily number of positive saliva samples (colored circles) and possible transmission chains within classes (directed arrows).
+Positive samples are linked if they belong to the same virus and are less than 1~week apart.
+Positive samples from the air and filters as blank squares aligned.
+IFB: influenza~B, HRV: human rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: human metapneumovirus, PIV: parainfluenza virus.}
     \label{fig:molecular-descriptives}
@@
 
-\subsection{Analyses of transmission networks and molecular genotyping}
-
-To illustrate possible transmission chains, we linked positive saliva samples within classes that belong to the same virus and are less than 1~week apart.
-Based on this spatiotemporal analysis of student saliva samples, we identified 10~possible transmission chains, which are shown in \Cref{fig:molecular-network}.
-The longest transmission chains possibly occurred in January and refer to the cluster of AdV infections in class~A and Flu~B infections in class~B.
-We performed molecular genotyping of saliva, bioaerosol, and air filter samples to verify our proposed transmission networks.
-However, molecular genotyping was not successful because we were unable to amplify and sequence any of the gene targets.The virus-specific variation between classes and the temporal distribution of positive samples suggest that transmission occurred within classes.
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{../../results/mol-data/network-plot.pdf}
-    \caption{\textbf{Transmission network of respiratory viruses: spatiotemporal analysis of students' saliva samples}.
-Daily number of positive saliva samples (colored circles) and possible transmission chains within classes (directed arrows).
-Positive samples are linked if they belong to the same virus and are less than 1~week apart.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
-    \label{fig:molecular-network}
-\end{figure}
-
-\subsection{Analysis of aerosol and particle concentrations}
-
-There was a strong difference in particle concentrations between study conditions (\Cref{fig:palas-results}a).
-Aerosol number concentration with air cleaners were markedly lower (27$\pm$34\,1/cm$^3$) than without air cleaners (95$\pm$81\,1/cm$^3$).
-Similarly, particle mass concentrations (\eg PM$_1$) were markedly lower with air cleaners (1.16$\pm$1.33\,$\mu$gm$^{-3}$) than without air cleaners (4.20$\pm$3.74\,$\mu$gm$^{-3}$).
-There was no considerable difference in other environmental variables (Figure~\zref{fig:env-descriptives-other-vars} in \supp).
-In particular, daily average CO$_2$ levels (air change rates) were 1,769$\pm$391\,ppm (0.74$\pm$0.41\,h$^{-1}$) with air cleaners and 1,636$\pm$341\,ppm (0.68$\pm$0.32\,h$^{-1}$), suggesting that ventilation was comparable between study conditions.
-When adjusting for air change rates and multiple other factors, the aerosol number concentration decreased by 76\% (95\%-CrI 63\% to 86\%) with air cleaners (\Cref{fig:palas-results}b and Table~\zref{tab:palas-est-results} in \supp).
-The decrease in the concentration of larger particles (PM$_{10}$) was greater than the decrease in the concentration of smaller particles (PM$_1$ to PM$_{4}$).
-
-\begin{figure}[!htpb]
-\centering
-    \includegraphics[width=\linewidth]{../../results/env-data/particles-boxplot.pdf}
-    \includegraphics[width=\linewidth]{../../results/env-data/estimation-results-figure.pdf}
-    \caption{\textbf{Analysis of particle concentrations and comparison between study conditions}.
-\textbf{(a)}~Boxplot of the daily average values for aerosol number concentration (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$).
-\textbf{(b)}~Estimated reduction in aerosol number and particle mass concentrations with air cleaners (posterior mean as dot and 50\%-, 80\%- and 95\%-CrI as lines, respectively).
-}
-    \label{fig:palas-results}
-\end{figure}
-
-\subsection{Risk of infection}
-
-Overall, there were 35~respiratory cases (absences related to respiratory infections) across classes.
-Incidentally, there were more cases in class~B (21) than in class~A (14), which is incidentally also the class where more students recently recovered from a SARS-CoV-2 infection or were vaccinated.
-Except for one study week, there was an equal or higher number of cases without air cleaners (\Cref{fig:redcap-results}a).
-Based on our Bayesian latent hierarchical regression model, we estimated a trend towards a lower risk of infection with air cleaners (adjusted relative risk 0.72, 95\%-Credible Interval 0.44 to 1.15).
-As a result, the estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\Cref{fig:redcap-results}b).
-The posterior probability that air cleaners reduced the risk of infection was 90.8\%, also suggesting a beneficial effect of air cleaners.
-Detailed estimations are provided in the Supplementary Text~\zref{sec:detailed-redcap} in \supp.
-
-\begin{figure}[!htpb]
-    \includegraphics{../../results/epi-data/cases_by_condition.pdf} 
-    \includegraphics{../../results/epi-data/avoided-infections.pdf}
-    \caption{\textbf{Analysis of epidemiological data and estimated number of infections with and without air cleaners.}.
-\textbf{(a)}~Weekly number of respiratory cases with and without air cleaners.
-\textbf{(b)}~Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
-    \label{fig:redcap-results}
-\end{figure}
-
-\subsection{Detected coughs}
-
-Slightly fewer coughs were detected in audio recordings during study weeks where air cleaners were installed (\Cref{fig:coughing}a).
-On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners (adjusted relative risk 0.93, 95\%-CrI 0.85 to 1.02).
-Significantly more coughs were detected in class~B (adjusted relative risk 2.16, 95\%-CrI 1.90 to 2.46), suggesting a link with the observed variation in virus-specific transmission between classes (\Cref{fig:molecular-descriptives}c).
-As suspected, the frequency of coughing was associated with the number of positive saliva samples by respiratory virus (\Cref{fig:coughing}b).
-Coughing was more frequent during Flu~B positivity (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent during AdV positivity (0.87, 95\%-CrI 0.49 to 1.12).
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{results/cough-data/coughs-frequency-by-condition.pdf}
-    \includegraphics{results/cough-data/coughs-virus-association.pdf}
-    \caption{\textbf{Analysis of coughing data.} \textbf{(a)}~Daily average number of detected coughs per minute by study week, condition and class.
-\textbf{(b)}~Estimated association (relative risk; posterior mean as dot, 50\%, 80\%, and 95\%-CrI as lines) between cough frequency and number of positive saliva test results by respiratory virus.}
-    \label{fig:coughing}
-\end{figure}
-
-\subsection{Comparison with previous study}
-
-In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 Omicron wave.
-\Cref{fig:comparison} shows a comparison of the results.
-The distribution of respiratory infections is different (\Cref{fig:comparison}a) because we almost exclusively detected SARS-CoV-2 in our previous study during the COVID-19 pandemic, as compared to a range of respiratory infections in this study during non-pandemic times.
-Although there were more positive saliva samples in this study, the number of respiratory infections detected in bioaerosols and on the HEPA-filters of the air cleaners was lower (\Cref{fig:comparison}b).
-The reductions in particle concentrations were larger and with lower uncertainty in this study (\Cref{fig:comparison}c).
-The risk of infection tended to be lower with air cleaners in this study where air cleaners were analyzed using a cross-over design, but not in our previous study where air cleaners were installed only at the end of the study when most students had already been infected with SARS-Cov-2(\Cref{fig:comparison}d).
-
-\begin{figure}[!htpb]
-    \includegraphics[width=\linewidth]{../../results/comparison.pdf} 
-    \caption{\textbf{Comparison of the results between our previous study in 2022 during the COVID-19 pandemic (blue) and this study in 2023 during non-pandemic times (black).} \textbf{(a)}~Proportion of positive saliva samples by virus.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.
-\textbf{(c)}~Estimated reduction in aerosol number and particle mass concentrations (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$) with air cleaners (posterior mean as squares and 95\%-CrI as lines).
-\textbf{(d)}~Adjusted relative risk (ARR) of infection with air cleaners (posterior mean as squares and 95\%-CrI as lines).}
-    \label{fig:comparison}
-\end{figure}
-
-\FloatBarrier
-
-\newpage
-
-\section{Discussion}
-
-% summary
-
-We collected epidemiological, environmental, and molecular data to estimate the risk of respiratory virus infection in a Swiss school and assessed the effectiveness of air cleaners as an infection control measure.
-We found a wide range of respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus, but very few respiratory viruses were detected in bioaerosol samples and on the filters of the air cleaners.
-Aerosol number and particle mass concentrations decreased significantly with air cleaners and Bayesian modeling based on the absences of students due to illness showed a small, potential reduction in the relative risk of infection with air cleaners.
-Reduced coughing could indicate that air cleaners prevent some symptomatic infections.
-
-% comparison with our previous study
-
-In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners in reducing the transmission of SARS-CoV-2 in two Swiss schools.
-We found a detectable effect on the risk of infection only for mask wearing, but not air cleaners.
-We argued that air cleaners may have had less potential to reduce infections because they were introduced at the end of the study, when most students were already infected with SARS-CoV-2.
-In this study, we used a cross-over design to estimate the effectiveness of air cleaners again, but this time under non-pandemic conditions.
-We found similar reductions in particle concentrations and a tendency for the relative risk of infection to be lower with air cleaners.
-To some extent, the differences between the two studies may be explained by different adjustments.
-While we could not adjust for ventilation in our previous study, the estimates in this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see Supplementary Text~\zref{sec:detailed-molecular} in \supp).
-Furthermore, the distribution of respiratory viruses in this study was very different from that in our previous study.
-
-% why could we not detect so many bioaerosols
-
-We detected fewer respiratory viruses in bioaerosol samples compared to our previous study\cite{Banholzer2023PLoSMed} (1~sample of adenovirus and 1~sample of rhinovirus in this study vs 10~samples of SARS-CoV-2 in the previous study), especially in relation to the number of positive saliva samples (50~samples in this study vs 21~in the previous study).
-Similarly, the number of viruses detected on the filters of the air cleaners was also lower in this study (4~positive samples in 160~swabs vs 8~positive samples in 80~swabs in the previous study).
-Importantly, there was a clear shift in the distribution of respiratory viruses in the current study compared to the previous study in 2022 during the SARS-CoV-2 Omicron variant.
-Whereas in 2022 we detected almost exclusively SARS-CoV-2 in human saliva, the most common respiratory viruses in 2023 were adenovirus, influenza~B, and rhinoviruses (only three SARS-CoV-2-positive samples).
-The shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.
-
-% other reasons
-
-% Technical reasons for the differences in airborne detection are unlikely.
-The bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
-Instead, airborne detection could have been influenced by environmental factors.
-Although temperature (overall between 19 and 25$^{\circ}$C) and relative humidity (overall between 25 and 50\%) were similar in both studies, the same levels may have different effects on the airborne survival of respiratory viruses.
-For example, airborne survival of rhinovirus and adenovirus increases with relative humidity, whereas the opposite relationship has been documented for influenza and SARS-CoV-2\cite{Tellier2009JTRSI,Ahlawat2020AAQR,Biryukov2020mS,Karim1985CJM,Davis1971AM}.
-A change in ventilation conditions could also have had a profound effect, but CO$_2$ levels were higher compared to the previous study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
-Although lower ventilation was partially offset by a greater reduction in aerosol and particle concentrations from air cleaners, the removal of particles by air cleaners may be slower than by natural ventilation.
-
-% bioaerosol sampling in more depth
-
-We speculate that the difference in airborne molecular detection of respiratory viruses between 2022 (pandemic) and 2023 (non-pandemic) is due to the different characteristics of the respiratory viruses.
-Differences in the physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity), which influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}, may facilitate the detection of SARS-CoV-2 in the air.
-The reason is that viral loads do often not exceed the detection limits of existing devices for bioaerosol sampling, which remains a challenge\cite{Wang2021,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
-For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.
-However, several studies showed that influenza\cite{Bischoff2013JID,Pan2017mSphere}, rhinovirus\cite{Myatt2004AJRCCM}, and adenovirus\cite{Nguyen2017OFID,Pan2017mSphere} can also be detected in airborne samples.
-Finally, the ability to detect pathogens in bioaerosols may be related to patient characteristics of the study participants.
-Several studies suggested an association between the generation of pathogenic aerosols and infectiousness\cite{Leung2020NatMed,Bischoff2013JID,Escombe2008PLoSMed}.
-Therefore, studies that include more infectious patients (intentionally or unintentionally) are likely to have a higher chance of detecting respiratory viruses in the air.
-
-% transmission within classrooms
-
-The low rate of positive bioaerosol samples also means that it is unclear to what extent airborne transmission occurred in classrooms.
-It may be possible that students had relatively little exposure to respiratory viruses in schools and acquired their infections elsewhere.
-However, the distribution of saliva samples between classes shows remarkable differences.
-For example, adenovirus spread entirely in class~A during the first three study weeks and only two infections with influenza~B occurred over the study.
-In contrast, influenza~B spread throughout the study in class~B, and adenovirus infections were  detected only in the last week of the study.
-Adenovirus infections tend to be mild\cite{Kunz2010CIDR} and less associated with cough than influenza\cite{Ma2018RMV}, which is consistent with the comparatively lower frequency of detected coughs in class~B.
-Taken together, the class-specific, spatiotemporal patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.
-
-% effects of air cleaners
-
-Our study indicated a small, potential benefit of air cleaners.
-There may be several reasons for this relatively small effect.
-Compared to \eg universal mask wearing, which has been shown to be effective\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Dharmadhikari2012AJRCCM,Leung2020NatMed,Milton2013PLoSPathogens}, air cleaners cannot affect transmission outside the classroom.
-Furthermore, previous work suggests that prolonged close contact may be necessary for transmission\cite{Leung2020NatMed,Brankston2007LancetID}, or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
-However, compared to face masks, air cleaners are not able to prevent transmission by close range, high particle density.
-Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective at removing larger particles ($>5\mu$m).
-However, a greater proportion of respiratory viruses are carried in the smaller particles, which are also more relevant for transmission ($\leq5\mu$m)\cite{Fennelly2020}.
-Finally, classroom activity, airflow and other unobserved, confounding factors make it difficult to evaluate the effects of air cleaners.
-Nevertheless, it is important to assess their effectiveness in a real-world setting and to compare them with the hypothetical impact suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.
-
-% limitations
-
-Our study has several limitations.
-First, aerosol measurements and molecular detection of viruses in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (person to air, air to person) cannot be determined.
-However, the virus-specific distribution of positive saliva samples between classes (spatiotemporal pattern) may indicate transmission within classrooms.
-Second, the specific causes for the absences could not be determined from the epidemiological data.
-Therefore, some absences may have falsely been attributed to respiratory infections and we could only approximate the incubation period for each epidemiological case.
-Third, the effectiveness of air cleaners was somewhat limited because students occasionally had lessons outside the classroom.
-This is in contrast to other infection control measures such as face masks, which had to be used in every indoor space during the pandemic.
-
-% conclusion
-
-In conclusion, using epidemiological, environmental, and molecular data, our study showed that a wide range of non-SARS-CoV-2, but rarely SARS-CoV-2, was detected in students under non-pandemic conditions when public health measures were lifted, and that transmission of these respiratory viruses likely occurred in schools.
-Airborne detection of non-SARS-CoV-2 respiratory viruses was rare, suggesting that prolonged close contact may be required for transmission.
-Comparison with previous work during the pandemic, which showed frequent airborne SARS-CoV-2, suggests that respiratory viruses other than SARS-CoV-2 may be less detectable in the air.
-Air cleaners improved air quality by significantly reducing aerosol and particle concentrations.
-Based on our transmission model using epidemiological data, the risk of infection tends to be lower with air filters.
-Future studies should investigate the relative importance of prolonged close contact and long range transmission of respiratory viruses in schools and other congregate settings.
-
-\newpage
-
-%TC:ignore
-\section*{Acknowledgements}
-We would like to thank the schools, teachers, and students participating in the study.
-We are grateful to the Educational Department of the canton of Solothurn for their support throughout the study.
-We also would like to thank the biosecurity team at the Institute for Infectious Diseases of the University of Bern for their assistance with the bioaerosol devices (Julia Feldmann, Monika Gsell, Kathrina Summermatter).
-Finally, we are indebted to the student assistants (Khadija Ahmed, Marie-Joséphine Brancato, Michelle Bürki, Santiago Martinez, Moric Toszeghi, Sylvain Wasmer) who helped with the data collection in the schools.
-
-\bibliography{references.bib}
-%TC:endignore
-
 \end{document}

### 2023-11-07 doc/paper/old-manuscript-long.tex
finalised manuscript

\PassOptionsToPackage{top=3cm,left=3cm,right=3cm,bottom=3cm}{geometry}
\documentclass[fleqn,11pt]{wlscirep}

\usepackage{import}
\usepackage{main}

\renewcommand{\paragraph}[1]{\vspace{0.3cm}\noindent\underline{\emph{#1}}\hfill\noindent}

% word count
% \newcommand{\maincount}[1]{%
%   \immediate\write18{texcount -1 -sum=1 -merge -q -nobib #1.tex > #1-words.sum}%
%   \input{#1-words.sum}%
% }

% \newcommand{\abstractcount}[1]{%
%   \immediate\write18{texcount -template="{abst}" #1.tex > #1-words.sum}%
%   \input{#1-words.sum}%
% }

\begin{document}

\doublespacing

\title{\bfseries\LARGE\singlespacing{Transmission of respiratory viral infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}}

% Alternative title: SARS-CoV-2 transmission in schools based on molecular and epidemiological data: potential effects of mask wearing and air cleaners

% author list
\author[1$\ddag$,2]{Nicolas Banholzer}
\author[2,3]{Philipp Jent}
\author[2,4]{Pascal Bittel}
\author[1]{Kathrin Zürcher}
\author[4]{Lavinia Furrer}
\author[1]{Simon Bertschinger}
\author[2,4]{Alban Ramette}
\author[2,4]{Ronald Dijkman}
\author[1]{Matthias Egger}
\author[2,5]{Tina Hascher}
\author[1*,2]{Lukas Fenner}

\affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
\affil[2]{Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland}
\affil[3]{Department of Infectious Diseases, Inselspital, Bern University Hospital, University of Bern, Bern, Switzerland}
\affil[4]{Institute for Infectious Diseases, University of Bern, Bern, Switzerland}
\affil[5]{Institute of Educational Science, University of Bern, Bern, Switzerland}

\affil[*]{Corresponding author: lukas.fenner@unibe.ch }

\vspace{1em}

\begin{information}\normalfont

xxxx

% \noindent\textbf{Running head}: SARS-COV-2 transmission in schools and effect of air cleaners

% %\noindent\textbf{Subject categorization}: 6.20  Indoor Air; 10.11 Pediatrics: Respiratory Infections

% \noindent\textbf{Word count}: \maincount{manuscript}words, abstract \abstractcount{manuscript}words (max. 500), title 157 chars (max. 200)

% %\noindent\textbf{Inserts:} 2 tables, 6 figures, 44 references

% \noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.

% \vspace{1em}

% \noindent\textbf{Funding}

% \noindent This study is funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland. NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05. ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498). \medskip

% \noindent\textbf{Contributions}

% \noindent Conception and design: NB, LF. Epidemiological and environmental data collection: NB, PJ, TS, LF. Laboratory data collection: PB, LFu. Additional data collection: TH. Statistical analysis: NB, KZ. Genomic analysis: LB, LFu. Paper draft: NB, LF, ME. All authors reviewed and approved the final version of the manuscript.

% \par
\end{information}

%TC:ignore
\begin{abstract}\normalfont
\noindent\textbf{Background:} Airborne respiratory infections can be transmitted in crowded indoor settings, including schools. Infection control measures such as air ventilation or filtration are supposed to reduce the risk of airborne transmission. The aim of this study was to evaluate the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in reducing the transmission of respiratory viral infections in schools using a multiple-measurement approach. \medskip

\noindent\textbf{Methods:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), audio (recordings to detect coughing), and molecular data (bioaerosol and saliva samples) over seven weeks from 16 January to 11 March 2023 in a secondary school (38 students in two classes) in Switzerland. We performed laboratory and molecular analyses to detect a panel of 19 respiratory viruses. We installed air cleaners in a cross-over design and compared positive samples and particle concentrations between weeks with and without air filters. We modeled the number of new infections using a Bayesian negative binomial regression model, adjusting for incubation periods, number of recent and cumulative infections, ventilation rate, number of students in class, and community transmission. \medskip

\noindent\textbf{Results:} Molecular analysis detected multiple viruses in saliva samples throughout the study (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus), but rarely in bioaerosol samples (2/105 positive: 1~rhinovirus and 1~adenovirus) and on the filters of air cleaners (4/160~positive: 1~influenza~B, 1~HRV, 1~adenovirus, and 1~SARS-CoV-2). Nevertheless, spatiotemporal analysis of saliva samples suggested transmission within classrooms. The detection rate of positive saliva samples was comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61). Daily average CO$_2$ levels were 1,769$\pm$391\,ppm ($\pm$ standard deviation) with air cleaners and 1,636$\pm$341\,ppm without air cleaners, corresponding to similar air change rates (0.74$\pm$0.41\,h$^{-1}$ vs. 0.68$\pm$0.32\,h$^{-1}$). Daily average aerosol number concentrations were 27$\pm$34\,1/cm$^3$ with and 95$\pm$81\,1/cm$^3$ without air cleaners, corresponding to an estimated adjusted decrease of 77\% (95\%-CrI 63\% to 86\%). Slightly less coughs  were detected with air cleaners (2.5~coughs/min with vs. 3.1~coughs/min without air cleaners). Absences related to respiratory infections were less frequent with air cleaners (13~absences with vs. 21~absences without air cleaners). The estimated risk of infection tended to be lower with air cleaners (adjusted relative risk 0.72, 95\%-Credible Interval 0.44 to 1.15). \medskip

% Study limitations include unknown specific causes of absences related to respiratory infections (i.e which virus caused the sickness). Furthermore, students occasionally had lessons outside the classroom, thus limiting the potential effectiveness of air cleaners.

\noindent\textbf{Conclusions:} During a non-pandemic period of winter 2023, a wide range of non-SARS-CoV-2 respiratory viruses were detected in the saliva of students in a Swiss school. Although airborne molecular detection was rare, transmission of these respiratory viruses seemed likely in schools, suggesting that prolonged close contact is required for transmission. Air cleaners improved air quality, were associated with lower symptom intensity, and showed a small, potential benefit in reducing respiratory infections.

\par
\end{abstract}

\flushbottom
\maketitle
\setcounter{page}{1}
\thispagestyle{fancy}

\vspace{2em}

\vspace{0.5em}

\noindent\textbf{Keywords:} schools, air cleaner, respiratory viruses, airborne transmission, molecular detection
% maximum of 3-5 keywords
\newpage

\sloppy
\raggedbottom
%TC:endignore

\newpage

\section{Introduction}

The transmission of respiratory infections such as SARS-CoV-2 and influenza is difficult to mitigate and control. Person-to-person transmission occurs primarily through the release of respiratory particles containing the respective viruses. Recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}. Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and their respiratory plumes can reach more than 2m\cite{Coleman2022,Wang2020,Heneghan2021}, exceeding the social distancing rules recommended by public health organizations\cite{Bazant2021PNAS,Trivedi2021PhyFluids,Lindsley2010CID}.

The generation, spread, and concentration of bioareosols can be influenced by infection control measures. For example, mandatory face mask use reduced transmission of SARS-CoV-2 during the COVID-19 pandemic\cite{Gettings2021,Heinsohn2022,Leech2022PNAS,Mitze2020PNAS,Banholzer2021PLOS}. In addition, improved ventilation systems are critical for a healthy indoor environment both in pandemic and non-pandemic times\cite{Wang2021,Morawska2021}. Portable HEPA-air filtration devices (air cleaners) may be another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear. Several studies have shown that air cleaners reduce the concentration of aerosols and particle matter in indoor air\cite{Park2020Build,Buising2022InfContr,Ren2021Dent,Banholzer2023PLoSMed}. These findings suggest potential protection against respiratory infections, as long-term exposure to higher aerosol and particle matter mass concentrations has been associated with increased risk of respiratory infections\cite{Gordon2014Resp} and, more recently, with the incidence and mortality of COVID-19\cite{Kelly2023Atmos,Veronesi2022OccEnvMed,Travaglio2021EnvPoll,DeAngelis2021EnvRes}. In addition, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}. Simulation studies have demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission  of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}.  We have studied the effectiveness of air cleaners in reducing transmission of SARS-CoV-2 in schools during the pandemic\cite{Banholzer2023PLoSMed}, but the analysis was limited by the study design (installation of air cleaners at the end of the Omicron wave when most students were already infected). To date, empirical evidence linking the use of air cleaners to a reduced risk of infection is lacking. It remains unclear whether reducing particle concentrations and removing bioaerosols from the air will result in reduced transmission of respiratory infections.

We used a multiple-measurement approach to study the transmission of viral respiratory viruses and the effect of air cleaners in school rooms. We employed a cross-over study design under non-pandemic conditions in the winter of 2023. We collected epidemiological (absences likely related to respiratory infections), environmental (CO$_2$, aerosol and particle concentrations), audio (recordings to detect coughing), and molecular data (detection of respiratory viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2022 in two classes of a secondary school in Switzerland. We determined changes in particle concentrations, respiratory virus positivity rates, and modeled the risk of infection.

\newpage

\section{Methods}

\subsection{Study setting and design}

\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from 26~January to 11~March 2023. \Cref{fig:study-setup} shows the schematic study setup and \supp~table~\zref{tab:data} provides an overview of the types of data collected in each classroom.

\begin{figure}[!htpb]
    \centering
    \includegraphics{../study_setting.pdf}
    \caption{\textbf{Study setting}. Schematic study setup of the classrooms. One air cleaner was placed in the front and one in the back of the classrooms. All devices were placed at the head level of the students when they were seated. Both classrooms were not equipped with an active HVAC (Heating, Ventilation, Air conditioning) system, but were ventilated naturally by opening windows. }
    \label{fig:study-setup}
\end{figure}

\subsection{Study intervention}

\noindent We employed a cross-over design to study the effectiveness of air cleaners, distinguishing two study conditions in each class: weeks with and weeks without air cleaners being installed in each classroom (\Cref{tab:study_design}). Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China; approx. USD\,250 per device). According to the manufacturer, these air cleaners run at a clean air delivery rate of 2$\,\times\,$600 m$^{3}$/h. % We measured a lower clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h when testing the devices in an empty classroom. 
Window ventilation was done at the discretion of the teachers, as there were no specific recommendations from the national health authorities at that time.

\begin{table}[!htpb]
    \footnotesize
    \centering
    \caption{\textbf{Study design.} Cross-over design describing when portable air cleaners (AC) were installed in the rooms of classes A and B during a seven-week study period from 16~January to 11~March 2023, excluding a week of vacation from February 6 to 11. A ``*'' indicates when swabs from the HEPA filters were taken, i.e. after each intervention phase and before the vacation.}\label{tab:study_design}
    \begin{tabular}{l l l l l l l l l}
    \toprule
      & \textbf{Week 1} & \textbf{Week 2} & \textbf{Week 3} & Vacation & \textbf{Week 4} & \textbf{Week 5} & \textbf{Week 6} & \textbf{Week 7} \\
      & Jan 16-21 & Jan 23-28 & Jan 30-Feb 3 & Feb 6-11 & Feb 13-18 & Feb 20-25 & Feb 27-March 3 & March 6-11 \\
      \midrule
      \textbf{A} & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{0000}*& \cellcolor{gray!10} None & & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{00009}* \\
      \textbf{B} & \cellcolor{gray!10} None & \cellcolor{gray!10} None & \cellcolor{gray!50} AC\hphantom{0000000}* & & \cellcolor{gray!50} AC & \cellcolor{gray!50} AC\hphantom{0000}*& \cellcolor{gray!10} None & \cellcolor{gray!10} None \\
      \bottomrule
    \end{tabular}
\end{table}

\subsection{Data collection}

\noindent\textbf{Epidemiological data} \smallskip

\noindent At the beginning of the study, we collected aggregated data on age, sex and COVID-19 vaccination and recovery status in the participating classes. On a daily basis, we collected data on each absent student, \ie the absence and return dates as well as the reason for the absence (illness, career days, or other). For absences due to illness, we collected symptoms and the date of symptom onset. We also collected data on laboratory tests but none of the students reported a test result during the study. Based on the epidemiological data, we defined a case of respiratory infection as an absence in which the student reported an illness with at least one of the following symptoms: fever, cough, fatigue, loss of taste or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, and stomach. Students were asked to report the first date they experienced symptoms, so that a student absent on Monday could report Saturday or Sunday as the day when symptoms began to be felt. We will refer respiratory cases by their dates of symptom onset, which usually corresponded to the absence date, unless the student attended the school while experiencing symptoms. Informed consent for one student could not be obtained and the single absence of this student was considered unrelated to a respiratory infection. Reports about absences were entered electronically into a REDcap database\cite{Harris2009,Harris2019}. The line list data is shown in \supp~table~\zref{tab:epi-data-line-list}.

In addition, both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) testing for a panel of respiratory infections (see \nameref{sec:mol_analyses} below). Saliva samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Galar2021,To2019,Huber2021}. The line list data is shown in \supp~table~\zref{tab:mol-data-line-list}. \medskip

\noindent\textbf{Environmental data} \smallskip

\noindent \emph{CO$_2$ and particle measurements}

\noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute. This device was used in previous work\cite{DiGilio2021,Duill2021,Banholzer2023PLoSMed}. A transient mass balance model was used to estimate the air change rate from indoor CO$_2$ levels\cite{Batterman2017IJERPH}, assuming that classrooms were fully occupied during lessons and that room occupancy increased (decreased) linearly 10~minutes before (after) the start (end) of a lesson. Particle concentrations were filtered according to the times when students were in the classroom, which were monitored together with the times during which windows were opened. \medskip

\noindent \emph{Bioaerosol sampling and HEPA filters of the air cleaners}

\noindent We collected airborne respiratory viruses in both classrooms with a bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL PBS, as previously reported in clinic settings\cite{Moore2021}. The Coriolis Micro Air could only operate shortly before and during break times to prevent unacceptable noise exposure (approximately 60\,min/day). In one class, we also used the BioSpot-VIVAS device sampler (Aerosol Devices Inc., Ft. Collins, CO, USA). This device samples airborne virus particles into a viral transport medium (VTM) using a water-based laminar flow condensation method\cite{Pan2016JAM,Lednicky2016AST}. BioSpot-VIVAS could operate throughout lessons. Removable parts of both sampling devices were regularly autoclaved to prevent contamination. At the end of the day, samples were transported to the Institute of Infectious Diseases (IFIK) and stored at $-$80°C. After each intervention phase (see \Cref{tab:study_design}), the HEPA filters were carefully removed from the air cleaners, divided into 20~fields, and one swab per field was taken for a total of 20~swabs per filter.\medskip

\subsection{Laboratory and molecular analyses}\label{sec:mol_analyses}

\noindent Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL. The human saliva samples were directly analyzed without prior filtration. The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19~respiratory viruses: SARS-CoV-2 (CoV), Influenza A virus (Flu A), Influenza B virus (Flu B), Human respiratory syncytial virus A/B (RSV), Human metapneumovirus (MPV), Human adenovirus A/B/C/D/E/F (AdV), Human rhinovirus A/B/C (HRV), and Human parainfluenza virus 1/2/3/4 (PIV) in one single reaction. %/Viral genome load (VGL) of specimens was quantified (copies/L) by curve extrapolation derived from standardized ATCC quantitative genomic standard dilutions (ATCC-VR-3347D, LGC Standards, LGC Group UK). For the BioSpot-VIVAS air samples, the air concentration of the DNA was exactly calculated according to the airflow influx into the sampling device. %/ 
The Limit of detection (LOD) of the assay SARS-CoV-2, IFA/IFB is 7.5x10$^0$ TCID/ml (SARS-CoV-2), 3.7x10$^{-1}$ TCID/ml (IFA), and 4.8x100 TCID/ml (100cp/ml) (IFB) respectively.

\subsection{Molecular genotyping}

We performed molecular genotyping for positive samples of AdV and Flu~B. For AdV (saliva, bioaerosol, and air filter samples), we used a PCR procedure to amplify three hypervariable regions of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}. For Flu~B, we tried to determine the molecular transmission network using whole genome sequencing\cite{Kelly2022FrontiersImmuno}.

\subsection{Audio recordings and cough detection}

We installed one portable audio recorder (ZOOM H6; New York, USA) in each classroom to record sounds continuously. We detected the number of coughs per minute using an AI algorithm based on a convolutional neural network\cite{Bertschinger2023IEEE}, which was recently trained and validated with audio data from the waiting room of a primary care clinic\cite{Zurcher2022}.

\subsection{Statistical analyses and modeling}

All statistical analyses were described in the statistical analysis plan (SAP)\cite{Banholzer2023SAP}, which was published before analyses began. In contrast to the SAP, the number of positive bioaerosol samples and the viral load concentration of positive samples could not be analyzed because there were too few positive samples during the study. Apart from that there were no major deviations from the SAP regarding the conducted analyses. There were a few minor deviations from the SAP (\eg small changes in prior distributions or model specifications), which are documented in \supp~text~\zref{sec:transmission-model}-\zref{sec:env-regression-model}, which provide detailed descriptions of the models used in each analysis. \medskip

\noindent\textbf{Risk of infection} \smallskip

\noindent We estimated the relative risk of infection with air cleaners using a Bayesian latent hierarchical regression model. The observed outcome of this model is the number of new respiratory cases $C$ (absences related to respiratory infections by date of symptom onset) on day $t$ in class $j$, which are modeled with a Negative Binomial distribution. The expected number of new cases is the weighted sum of the number of new infections $I_{js}$ (unobserved/latent outcome) in the previous days $s<t$, with the weights corresponding to the probability distribution of the incubation period. The number of new infections is related to the presence of air cleaners using a regression-like log-link
\begin{align}
    \log I_{js} = \log F_{js} - \log N_{js} + \beta_0 + \beta_1 \cdot \text{AirCleaner}_{js},
\end{align}
where $F_{js}$ is the number of infections in the previous week (a proxy for the number of infectious students), $N_{js}$ is the cumulative number of infections (a proxy for the number of susceptible students), $\beta_0$ is the infection rate without air cleaners, and $\beta_1$ is the effect of air cleaners. Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the community (\ie the canton of Solothurn).

A detailed description of the model and choice of priors for all model parameters are provided in \supp~text~\zref{sec:transmission-model}. Importantly, since multiple viruses were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the virus-specific distributions. The weights were inferred based on the proportion of positive saliva samples for each virus per week. As a result, the distribution of the incubation period used for modeling could change every week depending on the proportion of viruses that were detected in the students' saliva. Moreover, respiratory cases by date of symptom onset were under-reported for weekends and thus a re-weighting scheme was used to account for weekday effects. \medskip

\noindent\textbf{Number of positive saliva samples} \smallskip

\noindent Molecular analysis determined which saliva samples were positive for a respiratory virus. We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~text~\zref{sec:multinomial-model}) where the expected number of positive samples for a virus was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each virus. The number of susceptibles was computed as the difference between the number of students per class and the cumulative number of positive samples. As a sensitivity analysis, we lagged the test results once by one test date (\ie from Tuesday to last Thursday and from Thursday to last Tuesday) and once by one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday), to consider the unknown delay when students were infected and tested positive. \medskip

\noindent\textbf{Aerosol and particle concentrations} \smallskip

\noindent We computed the average particle concentrations per day and compared them between study conditions. We estimated the reduction in particle concentrations using Bayesian log-linear regression models (\supp~text~\zref{sec:env-regression-model}), adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases. \medskip

\noindent\textbf{Number of detected coughs} \smallskip

\noindent We computed the daily number of detected coughs and compared it between study conditions. We estimated the reduction in the number of coughs using a Bayesian Negative Binomial regression model (\supp~text~\zref{sec:aud-regression-model}) with time in class as the model offset, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases. In addition, we used a Bayesian Negative Binomial hierarchical (random effects) regression model to estimate the association between the number of coughs and the number of positive saliva test results for respiratory viruses. For this analysis, we only used the days were saliva testing was performed (i.e. every Tuesday and Thursday, 27~days in total).  \medskip

\noindent\textbf{Software and model estimation} \smallskip

\noindent All analyses were performed in R software (version 4.2.0)\cite{RCoreTeam2022} and modeling in Stan (version 2.21.0)\cite{Carpenter2017}. Model parameters were estimated with a Bayesian approach. Specifically, Markov chain Monte Carlo (MCMC) sampling was used as implemented by the Hamiltonian Monte Carlo algorithm with the No-U-Turn Sampler (NUTS)\cite{Hoffman2014}. Each model is estimated with 4 Markov chains and 2,000 iterations of which the first 1,000 iterations are discarded as part of the warm-up. Estimation power is evaluated via the effective sample size (ESS) and convergence of the Markov chains is evaluated with the Gelman-Rubin convergence diagnostic ($\hat{R}$). If not stated otherwise, we report posterior means and credible intervals (CrIs) based on the 50\%, 80\%, and 95\% quantiles of the posterior samples, respectively. The code is available from \url{https://osf.io/38j9g}.

\subsection{Ethics statement}

\noindent The Ethics Committee of the canton of Bern, Switzerland, approved the study (reference no. 2021-02377). For the saliva samples, we included all students who were willing to participate and obtained written informed consent from their caregivers.

\newpage

\section{Results}

The study population consisted of 38~students (19~female, 19~male; Table~\ref{tab:cases-overview-school}). Of these, 7~students were vaccinated or recovered from an infection within SARS-CoV-2 in the last four months. During the seven-week study period (total of 1,330 student-days), students were absent from school for 220~days (18\% of the total) of which 129~days (59\% of absences) were due to illness.

\begin{table}[!htpb]
    \centering
    \caption{Overview of the study population and person-days of absences.}
    \label{tab:cases-overview-school}
    \footnotesize
    \begin{tabular}{l r r r}
    \toprule
         &  Class A & Class B & Total \\ \midrule 
        \textbf{Students} & \textbf{20 (53\%)} & \textbf{18 (47\%)} & \textbf{38 (100\%)} \\
        \emph{Gender} \\
        $\drsh$ Female & 11 (55\%) & 8 (44\%) & 19 (\hphantom{0}50\%) \\
        $\drsh$ Male & 9 (45\%) & 10 (56\%) & 19 (\hphantom{0}50\%) \\
        \emph{Immunity status} \\
        $\drsh$ Recently vaccinated (or recovered) & 7 (39\%) & 0 (0\%) & 7 (\hphantom{0}18\%) \\
        $\drsh$ Not recently vaccinated (or recovered) & 11 (61\%) & 20 (100\%) & 31 (\hphantom{0}82\%) \\
        \textbf{Absent person-days} & \textbf{110 (50\%)} & \textbf{110 (50\%)} & \textbf{220 (100\%)} \\
        $\drsh$ Sickness & 52 (47\%) & 77 (70\%) & 129 (\hphantom{0}59\%) \\
        $\drsh$ Other & 58 (53\%) & 33 (30\%) & 91 (\hphantom{0}41\%) \\
        \bottomrule
    \end{tabular} 
\end{table}

\subsection{Molecular analyses of saliva and air samples}

We analyzed 448~saliva samples over the study across classes. We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenovirus (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~virus~(PIV) (\Cref{fig:molecular-descriptives}a). There was an equal number of positive samples with and without air cleaners (\Cref{fig:molecular-descriptives}b). Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted relative risk 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (\supp~figure~\zref{fig:mol-estimation-results-sensitivity}). The distribution of positive samples varied between classes (\Cref{fig:molecular-descriptives}c). For example, during the first three study weeks, all but one sample was positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B. We also analyzed 105~bioaerosol samples and detected  two respiratory viruses in the air (1~HRV in class~A and 1~AdV in class~B). Finally, we detected 1~Flu~B, 1~HRV, 1~AdV, and 1~CoV in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).

\begin{figure}[!htpb]
    \centering
    \includegraphics{../../results/mol-data/descriptives.pdf}
    \caption{\textbf{Analysis of molecular data}. Number of positive saliva samples by \textbf{(a)}~virus, \textbf{(b)}~study condition, and \textbf{(c)}~study week and class. Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
    \label{fig:molecular-descriptives}
\end{figure}

\subsection{Analyses of transmission networks and molecular genotyping}

To illustrate possible transmission chains, we linked positive saliva samples within classes that belong to the same virus and were less than 1~week apart. Based on this spatiotemporal analysis of student saliva samples, we identified10~possible transmission chains, which are shown in \Cref{fig:molecular-network}. The longest transmission chains possibly occurred in January and refer to the cluster of AdV infections in class~A and Flu~B infections in class~B. We also performed molecular genotyping of saliva, bioaerosol, and air filter samples to verify our proposed transmission networks. However, molecular genotyping was not successful because we were unable to amplify and sequence any of the gene targets.Nevertheless, the virus-specific variation between classes and the temporal distribution of positive samples suggest that transmission occurred within classes.

\begin{figure}[!htpb]
    \centering
    \includegraphics{../../results/mol-data/network-plot.pdf}
    \caption{\textbf{Transmission network of respiratory viruses: spatiotemporal analysis of students' saliva samples}. Daily number of positive saliva samples (colored circles) and possible transmission chains within classes (directed arrows). Positive samples are linked if they belong to the same virus and are less than 1~week apart. Positive samples from the air and filters as blank squares. Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
    \label{fig:molecular-network}
\end{figure}

\subsection{Analysis of aerosol and particle concentrations}

There was a strong difference in particle concentrations between study conditions (\Cref{fig:palas-results}a). Aerosol number concentration with air cleaners were markedly lower (27$\pm$34\,1/cm$^3$) than without air cleaners (95$\pm$81\,1/cm$^3$). Similarly, particle mass concentrations (\eg PM$_1$) were markedly lower with air cleaners (1.16$\pm$1.33\,$\mu$gm$^{-3}$) than without air cleaners (4.20$\pm$3.74\,$\mu$gm$^{-3}$). There was no considerable difference in other environmental variables (\supp~figure~\zref{fig:env-descriptives-other-vars}). In particular, daily average CO$_2$ levels (air change rates) were 1,769$\pm$391\,ppm (0.74$\pm$0.41\,h$^{-1}$) with air cleaners and 1,636$\pm$341\,ppm (0.68$\pm$0.32\,h$^{-1}$), suggesting that ventilation was comparable between study conditions. When adjusting for air change rates and multiple other factors, the aerosol number concentration decreased by 76\% (95\%-CrI 63\% to 86\%) with air cleaners (\Cref{fig:palas-results}b and \supp~table~\zref{tab:palas-est-results}). The decrease in the concentration of larger particles (PM$_{10}$) was greater than the decrease in the concentration of smaller particles (PM$_1$ to PM$_{4}$).

\begin{figure}[!htpb]
\centering
    \includegraphics[width=\linewidth]{../../results/env-data/particles-boxplot.pdf}
    \includegraphics[width=\linewidth]{../../results/env-data/estimation-results-figure.pdf}
    \caption{\textbf{Analysis of particle concentrations and comparison between study conditions}. \textbf{(a)}~Boxplot of the daily average values for aerosol number concentration (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$). \textbf{(b)}~Estimated reduction in aerosol number and particle mass concentrations with air cleaners (posterior mean as dot and 50\%-, 80\%- and 95\%-CrI as lines, respectively). }
    \label{fig:palas-results}
\end{figure}

\subsection{Risk of infection}

Overall, there were 35~respiratory cases (absences related to respiratory infections) across classes. Incidentally, there were more cases in class~B (21) than in class~A (14), which is incidentally also the class where more students recently recovered from a SARS-CoV-2 infection or were vaccinated. Except for one study week, there was an equal or higher number of cases without air cleaners (\Cref{fig:redcap-results}a). Based on our Bayesian latent hierarchical regression model, we estimated a trend towards a lower risk of infection with air cleaners (adjusted relative risk 0.72, 95\%-Credible Interval 0.44 to 1.15). As a result, the estimated number of infections would be 36 (95\%-CrI 21 to 64) if air cleaners had been installed throughout the study period, compared to 62 (95\%-CrI 29 to 141) infections if air cleaners had not been installed (\Cref{fig:redcap-results}b). The posterior probability that air cleaners reduced the risk of infection was 90.8\%, also suggesting a beneficial effect of air cleaners. Detailed estimations are provided in the \supp~text~\zref{sec:detailed-redcap}.

\begin{figure}[!htpb]
    \includegraphics{../../results/epi-data/cases_by_condition.pdf} 
    \includegraphics{../../results/epi-data/avoided-infections.pdf}
    \caption{\textbf{Analysis of epidemiological data and estimated number of infections with and without air cleaners.}. \textbf{(a)}~Weekly number of respiratory cases with and without air cleaners. \textbf{(b)}~Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
    \label{fig:redcap-results}
\end{figure}

\subsection{Detected coughs}

Slightly fewer coughs were detected in audio recordings during study weeks where air cleaners were installed (\Cref{fig:coughing}a). On average, there were 2.5~coughs/min with and 3.1~coughs/min without air cleaners (adjusted relative risk 0.93, 95\%-CrI 0.85 to 1.02). Significantly more coughs were detected in class~B (adjusted relative risk 2.16, 95\%-CrI 1.90 to 2.46), suggesting a link with the observed variation in virus-specific transmission between classes (\Cref{fig:molecular-descriptives}c). As suspected, the frequency of coughing was associated with the number of positive saliva samples by respiratory virus (\Cref{fig:coughing}b). Coughing was more frequent during Flu~B positivity (relative risk 1.25, 95\%-CrI 1.03 to 1.53) and slightly less frequent during AdV positivity (0.87, 95\%-CrI 0.49 to 1.12).

\begin{figure}[!htpb]
    \centering
    \includegraphics{results/cough-data/coughs-frequency-by-condition.pdf}
    \includegraphics{results/cough-data/coughs-virus-association.pdf}
    \caption{\textbf{Analysis of coughing data.} \textbf{(a)}~Daily average number of detected coughs per minute by study week, condition and class. \textbf{(b)}~Estimated association (relative risk; posterior mean as dot, 50\%, 80\%, and 95\%-CrI as lines) between cough frequency and number of positive saliva test results by respiratory virus.}
    \label{fig:coughing}
\end{figure}

\subsection{Comparison with previous study}

In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 Omicron wave. \Cref{fig:comparison} shows a comparison of the results. The distribution of respiratory infections is different (\Cref{fig:comparison}a) because we almost exclusively detected SARS-CoV-2 in our previous study during the COVID-19 pandemic, as compared to a range of respiratory infections in this study during non-pandemic times. Although there were more positive saliva samples in this study, the number of respiratory infections detected in bioaerosols and on the HEPA-filters of the air cleaners was lower (\Cref{fig:comparison}b).
The reductions in particle concentrations were larger and with lower uncertainty in this study (\Cref{fig:comparison}c). The risk of infection tended to be lower with air cleaners in this study where air cleaners were analyzed using a cross-over design, but not in our previous study where air cleaners were installed only at the end of the study when most students had already been infected with SARS-Cov-2(\Cref{fig:comparison}d).

\begin{figure}[!htpb]
    \includegraphics[width=\linewidth]{../../results/comparison.pdf} 
    \caption{\textbf{Comparison of the results between our previous study in 2022 during the COVID-19 pandemic (blue) and this study in 2023 during non-pandemic times (black).} \textbf{(a)}~Proportion of positive saliva samples by virus. Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus. \textbf{(c)}~Estimated reduction in aerosol number and particle mass concentrations (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$) with air cleaners (posterior mean as squares and 95\%-CrI as lines). \textbf{(d)}~Adjusted relative risk (ARR) of infection with air cleaners (posterior mean as squares and 95\%-CrI as lines).}
    \label{fig:comparison}
\end{figure}

\FloatBarrier

\newpage

\section{Discussion}

% summary

We collected epidemiological, environmental, and molecular data to estimate the risk of respiratory virus infection in a Swiss school and assessed the effectiveness of air cleaners as an infection control measure. We found a wide range of respiratory viruses in human saliva samples, mainly adenovirus, influenza B, and rhinovirus, but very few respiratory viruses were detected in bioaerosol samples and on the filters of the air cleaners. Aerosol number and particle mass concentrations decreased significantly with air cleaners and Bayesian modeling based on the absences of students due to illness showed a small, potential reduction in the relative risk of infection with air cleaners. Reduced coughing could indicate that air cleaners prevent some symptomatic infections.

% comparison with our previous study

In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners in reducing the transmission of SARS-CoV-2 in two Swiss schools. We found a detectable effect on the risk of infection only for mask wearing, but not air cleaners. We argued that air cleaners may have had less potential to reduce infections because they were introduced at the end of the study, when most students were already infected with SARS-CoV-2. In this study, we used a cross-over design to estimate the effectiveness of air cleaners again, but this time under non-pandemic conditions. We found similar reductions in particle concentrations and a tendency for the relative risk of infection to be lower with air cleaners. To some extent, the differences between the two studies may be explained by different adjustments. While we could not adjust for ventilation in our previous study, the estimates in this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see \supp~text~\zref{sec:detailed-molecular}).

% transmission within classrooms

The low rate of positive bioaerosol samples also means that it is unclear to what extent airborne transmission occurred in classrooms. It may be possible that students had relatively little exposure to respiratory viruses in schools and acquired their infections elsewhere. However, the distribution of saliva samples between classes shows remarkable differences. For example, adenovirus spread entirely in class~A during the first three study weeks and only two infections with influenza~B occurred over the study. In contrast, influenza~B spread throughout the study in class~B, and adenovirus infections were  detected only in the last week of the study. Adenovirus infections tend to be mild\cite{Kunz2010CIDR} and less associated with cough than influenza\cite{Ma2018RMV}, which is consistent with the comparatively lower frequency of detected coughs in class~B. Taken together, the class-specific, spatiotemporal patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.

% effects of air cleaners

Our study indicated a small, potential benefit of air cleaners. There may be several reasons for this relatively small effect. Compared to \eg universal mask wearing, which has been shown to be effective\cite{Banholzer2023PLoSMed,Heinsohn2022,Gettings2021,Dharmadhikari2012AJRCCM,Leung2020NatMed,Milton2013PLoSPathogens}, air cleaners cannot affect transmission outside the classroom. Furthermore, previous work suggests that prolonged close contact may be necessary for transmission\cite{Leung2020NatMed,Brankston2007LancetID}, or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}. However, compared to face masks, air cleaners are not able to prevent transmission by close range, high particle density. Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective at removing larger particles ($>5\mu$m). However, a greater proportion of respiratory viruses are carried in the smaller particles, which are also more relevant for transmission ($\leq5\mu$m)\cite{Fennelly2020}. Finally, classroom activity, airflow and other unobserved, confounding factors make it difficult to evaluate the effects of air cleaners. Nevertheless, it is important to assess their effectiveness in a real-world setting and to compare them with the hypothetical impact suggested by simulation studies\cite{Lindsley2021,Cortellessa2023Build}.

% limitations

Our study has several limitations. First, aerosol measurements and molecular detection of viruses in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (person to air, air to person) cannot be determined. However, the virus-specific distribution of positive saliva samples between classes (spatiotemporal pattern) may indicate transmission within classrooms. Second, the specific causes for the absences could not be determined from the epidemiological data. Therefore, some absences may have falsely been attributed to respiratory infections and we could only approximate the incubation period for each epidemiological case. Third, the effectiveness of air cleaners was somewhat limited because students occasionally had lessons outside the classroom. This is in contrast to other infection control measures such as face masks, which had to be used in every indoor space during the pandemic.

% conclusion

In conclusion, using epidemiological, environmental, and molecular data, our study showed that a wide range of non-SARS-CoV-2, but rarely SARS-CoV-2, was detected in students under non-pandemic conditions when public health measures were lifted, and that transmission of these respiratory viruses likely occurred in schools. Airborne detection of non-SARS-CoV-2 respiratory viruses was rare, suggesting that prolonged close contact may be required for transmission. Comparison with previous work during the pandemic, which showed frequent airborne SARS-CoV-2, suggests that respiratory viruses other than SARS-CoV-2 may be less detectable in the air. Air cleaners improved air quality by significantly reducing aerosol and particle concentrations. Based on our transmission model using epidemiological data, the risk of infection tends to be lower with air filters. Future studies should investigate the relative importance of prolonged close contact and long range transmission of respiratory viruses in schools and other congregate settings.

\newpage

%TC:ignore
\section*{Acknowledgements}
We would like to thank the schools, teachers, and students participating in the study. We are grateful to the Educational Department of the canton of Solothurn for their support throughout the study. We also would like to thank the biosecurity team at the Institute for Infectious Diseases of the University of Bern for their assistance with the bioaerosol devices (Julia Feldmann, Monika Gsell, Kathrina Summermatter). Finally, we are indebted to the student assistants (Khadija Ahmed, Marie-Joséphine Brancato, Michelle Bürki, Santiago Martinez, Moric Toszeghi, Sylvain Wasmer) who helped with the data collection in the schools.

\bibliography{references.bib}
%TC:endignore

\end{document}

### 2023-12-04 analysis/aud_analysis.Rmd
comparison with mcid12 doc style files

@@
           w = 7.25, h = 6)
+
+ggsave(plot = coughs %>%
+            mutate(coughs_per_min = n_coughs / TT) %>%
+            mutate(aircleaner = factor(aircleaner, levels = c("No", "Yes"))) %>%
+            ggplot(aes(x = aircleaner, fill = aircleaner, y = coughs_per_min)) +
+            geom_boxplot2(width.errorbar = .3, width = .4) +
+            ggsignif::geom_signif(annotations = "Posterior prob. = 93%", y_position = 6.25, xmin = 1, xmax = 2, 
+                                  textsize = 40 / cm(1), vjust = -.5) +
+            scale_fill_manual(values = c(bluegrey[1], "#E4003C")) +
+            scale_y_continuous(limits = c(0, NA), expand = expansion(add = c(0, .66))) +
+            labs(y = "Number of coughs per min", x = "Air cleaner") +
+            theme_custom() +
+            theme(legend.position = "none",
+                  axis.text = element_text(size = 40),
+                  axis.title = element_text(size = 40)), 
+          filename = "../results/cough-data/coughs-frequency-by-condition-large.png", 
+          width = 4 * 7.25 / cm(1), height = 4 * 6  / cm(1))
 ```

### 2023-12-04 analysis/env_analysis.Rmd
comparison with mcid12 doc style files

@@
           w = 7.25, h = 6)
+
+ggsave(plot = env_epi_data %>%
+  dplyr::select(class, date, aircleaner, Cn) %>%
+  ggplot(aes(x = aircleaner, fill = aircleaner, y = Cn)) +
+  geom_boxplot2(width.errorbar = .3, width = .4) +
+  ggsignif::geom_signif(annotations = "Posterior prob. = 100%", y_position = 240, xmin = 1, xmax = 2, 
+                        textsize = 40 / cm(1), vjust = -.5) +
+  scale_fill_manual(values = c(bluegrey[1], "#E4003C")) +
+  scale_y_continuous(expand = expansion(add = c(0, 30)), limits = c(0, NA)) +
+  labs(title = "a", y = expression("Aerosol concentration (1/cm"^3*")"), x = "Air cleaner") +
+  theme_custom() +
+  theme(legend.position = "none",
+        plot.title.position = "plot") +
+  theme(plot.title.position = "plot",
+        axis.text = element_text(size = 40),
+        axis.title = element_text(size = 40)) +
+  guides(fill = "none"),
+  filename = "../results/env-data/aerosol-number-boxplot-large.png",
+  width = 4 * 7.25 / cm(1), height = 4 * 6 / cm(1) )
 ```

### 2023-12-04 analysis/epi_analysis.Rmd
minor fix
comparison with mcid12 doc style files

@@
           w = 7.25, h = 6)
+
+ggsave(plot = epi_data %>%
+  group_by(aircleaner) %>%
+  summarize(cases = sum(new_cases)) %>%
+  ungroup() %>%
+  mutate(aircleaner = factor(aircleaner, levels = c("No", "Yes"))) %>%
+  ggplot(aes(x = aircleaner, y = cases, fill = aircleaner)) +
+  geom_bar(stat = "identity", color = "black", width = .4) +
+  geom_text(mapping = aes(label = cases), size = 40 / cm(1), vjust = -1) +
+  ggsignif::geom_signif(annotations = "Posterior prob. = 91%", y_position = 27, xmin = 1, xmax = 2, 
+                        textsize = 40 / cm(1), vjust = -.5, tip_length = .075) +
+  labs(x = "Air cleaner", y = "Number of respiratory cases", fill = "Air cleaner", title = "b") +
+  scale_fill_manual(values = c(bluegrey[1], "#E4003C")) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.1)), breaks = scales::pretty_breaks(4)) +
+  theme_custom() +
+  theme(plot.title.position = "plot") +
+  guides(fill = "none") +
+    theme(legend.position = "none",
+                  axis.text = element_text(size = 40),
+                  axis.title = element_text(size = 40)),
+  filename = "../results/epi-data/cases_by_condition_large.png",
+  width = 4 * 7.25 / cm(1), height = 4 * 6 / cm(1))
 ```
@@
 
-avoid_infections
+avoid_infections %>%
+  mutate_if(is.numeric, round, 0)
 

### 2023-12-04 analysis/mol_analysis.Rmd
comparison with mcid12 doc style files

@@
           w = 7.25, h = 6)
+
+ggsave(plot = pos_cond_pl <- df_pos_stud %>%
+  ggplot(aes(x = aircleaner, y = n_pos)) +
+  geom_bar(mapping = aes(fill = aircleaner), stat = "identity", color = "black", width = .4) +
+  geom_text(mapping = aes(label = n_pos), size = 40 / cm(1), vjust = -1) +
+  geom_segment(data = data.frame(x = .5, xend = .5, y = 30, yend = 29), 
+               mapping = aes(x = x, xend = xend, y = y, yend = yend), color = "white") +
+  ggsignif::geom_signif(annotations = "Posterior prob. = 65%", y_position = 30, xmin = 1, xmax = 2, 
+                        textsize = 40 / cm(1), vjust = -.5, tip_length = .1) +
+  labs(x = "Air cleaner", y = "Number of positive saliva samples") +
+  scale_fill_manual(values = c(bluegrey[1], "#E4003C")) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.1)), breaks = scales::pretty_breaks(4)) +
+  theme_custom() +
+  theme(plot.title.position = "plot",
+        axis.text = element_text(size = 40),
+        axis.title = element_text(size = 40)) +
+  guides(fill = "none"),
+  filename = "../results/mol-data/saliva-by-study-condition-large.png",
+  width = 4 * 7.25 / cm(1), height = 4 * 6 / cm(1))
 ```

### 2023-12-04 doc/paper/manuscript.tex
Updates from Overleaf

@@
 
-\title{\bfseries\LARGE\singlespacing{Air cleaners and respiratory infections in schools: A modeling study using epidemiological, environmental, and molecular data}}
+\title{\bfseries\LARGE\singlespacing{Estimating the effectiveness of air cleaners in preventing respiratory infections in a school setting using a multiple-measurement approach}}
 
-% Alternative title: SARS-CoV-2 transmission in schools based on molecular and epidemiological data: potential effects of mask wearing and air cleaners
+% Alternative title: Air cleaners and respiratory infections in schools: A modeling study using epidemiological, environmental, and molecular data
 
 % author list
-\author[1$\ddag$,2]{Nicolas Banholzer}
+\author[1,2]{Nicolas Banholzer}
 \author[2,3]{Philipp Jent}
@@
 \author[1,6,7]{Matthias Egger}
-\author[8]{Tina Hascher}
+\author[2,8]{Tina Hascher}
 \author[1*,2]{Lukas Fenner}
@@
 
-\affil[*]{Corresponding author: lukas.fenner@unibe.ch }
+\affil[*]{Corresponding author: lukas.fenner@unibe.ch}
 
@@
 
+%TC:ignore
+
 \begin{information}\normalfont
-\textbf{Word count:} Summary 276 words, Research in context 359 words, Main text 3,492 words \\
+\textbf{Word count:} Abstract 252, Manuscript 3,244 \\
 \textbf{Display items:} Tables 2, Figures 3 \\
-\textbf{Supporting information:} STROBE checklist, Appendix with additional text, tables, and figures.
+\textbf{References:} 37 \\
+\textbf{Supporting information:} STROBE checklist, Supplementary Data with texts, tables, and figures
 % \par
@@
 
-%TC:ignore
 \begin{abstract}\normalfont
-\noindent\textbf{Background:} Infection control measures can reduce airborne transmission of respiratory infections.
-Using a multiple-measurement approach, we examined the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in a school setting.\medskip
+% \noindent\textbf{Background:} Using a multiple-measurement approach, we examined the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in a school setting.\medskip %  Infection control measures can reduce airborne transmission of respiratory infections.
 
-\noindent\textbf{Methods:} We collected environmental data (CO$_2$ and particle concentrations), epidemiological data (absences related to respiratory infections), audio recordings (coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2022/2023 in two Swiss secondary school classes.
+% \noindent\textbf{Methods:} We collected environmental (CO$_2$, particle concentrations), epidemiological (absences related to respiratory infections), audio (coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2022/2023 in two Swiss secondary school classes.
 Using a cross-over study design, we compared particle concentrations, coughing, and the risk of infection with vs without air cleaners.\medskip
 
-\noindent\textbf{Findings:} All 38~students (age 13$-$15 years) participated.
+% \noindent\textbf{Results:} All 38~students (age 13$-$15 years) participated.
 With air cleaners, mean particle concentration decreased by 77\% (95\% credible interval 63\%$-$86\%).
@@
 Coughing also tended to be less frequent (posterior probability 93\%).
-Molecular analysis detected mainly non-SARS-CoV-2 viruses in saliva (50/448 positive), but not in bioaerosol samples (2/105 positive) or HEPA-filters (4/160).
+Molecular analysis detected mainly non-SARS-CoV-2 viruses in saliva (50/448 positive), but not in bioaerosols (2/105 positive) or HEPA-filters (4/160).
 The detection rate was similar with vs without air cleaners.
@@
 
-\noindent\textbf{Interpretation:} Air cleaners improved air quality, showed a potential benefit in reducing respiratory infections, and were associated with less coughing.
+% \noindent\textbf{Conclusions:} Air cleaners improved air quality, showed a potential benefit in reducing respiratory infections, and were associated with less coughing.
 Airborne detection of non-SARS-CoV-2 viruses was rare, suggesting that these viruses may be more difficult to detect in the air.
-Future studies should examine the importance of close contact and long-range transmission in schools and other congregate settings, and the cost-effectiveness of using air cleaners.\medskip
+Future studies should examine the importance of close contact and long-range transmission, and the cost-effectiveness of using air cleaners.\medskip
 
-\noindent\textbf{Funding:} Multidisciplinary Center for Infectious Diseases, Switzerland.
-
-%This study was funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
-NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
-ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498).
+\noindent Infection control measures can reduce airborne transmission of respiratory infections.
+Using a multiple-measurement approach, we examined the real-world effectiveness of portable HEPA-air filtration devices (air cleaners) in a school setting.
+We collected environmental (CO$_2$, particle concentrations), epidemiological (absences related to respiratory infections), audio (coughing), and molecular data (bioaerosol and saliva samples) over seven weeks during winter 2022/2023 in two Swiss secondary school classes.
+Using a cross-over study design, we compared particle concentrations, coughing, and the risk of infection with vs without air cleaners.
+With air cleaners, mean particle concentration decreased by 77\% (95\% credible interval 63\%$-$86\%).
+There were no differences in CO$_2$ levels.
+Among 38~students, absences related to respiratory infections were 22~without vs 13~with air cleaners.
+Bayesian modeling suggested a reduced risk of infection, with a posterior probability of 91\% and a relative risk of 0.73 (95\% credible interval 0.44$-$1.18).
+Coughing also tended to be less frequent (posterior probability 93\%).
+Molecular analysis detected mainly non-SARS-CoV-2 viruses in saliva (50/448 positive), but not in bioaerosols (2/105 positive) or HEPA-filters (4/160).
+The detection rate was similar with vs without air cleaners.
+Spatiotemporal analysis of positive saliva samples identified several likely transmissions.
+In conclusion, air cleaners improved air quality, showed a potential benefit in reducing respiratory infections, and were associated with less coughing.
+Airborne detection of non-SARS-CoV-2 viruses was rare, suggesting that these viruses may be more difficult to detect in the air.
+Future studies should examine the importance of close contact and long-range transmission, and the cost-effectiveness of using air cleaners.\medskip
 
@@
 
-\newpage
+% \newpage
 
-\noindent\textbf{\Large{Research in context}} \medskip
+% \noindent\textbf{\Large{Research in context}} \medskip
 
-\noindent \textbf{Evidence before this study} \smallskip
+% \noindent \textbf{Evidence before this study} \smallskip
 
-\noindent We searched Scopus on October 19, 2023, for all publications from inception using the terms ("air filt" OR "air clean") AND ("infection" OR "transmission").
+% \noindent We searched Scopus on October 19, 2023, for all publications from inception using the terms ("air filt" OR "air clean") AND ("infection" OR "transmission").
 Several studies have shown that air cleaners reduce particle concentrations in indoor air.
@@
 
-\noindent \textbf{Added value of this study} \smallskip
+% \noindent \textbf{Added value of this study} \smallskip
 
-\noindent To date, it remains unclear whether the reduction in particle concentrations with air cleaners also translates into lower risks of respiratory infection.
+% \noindent To date, it remains unclear whether the reduction in particle concentrations with air cleaners also translates into lower risks of respiratory infection.
 This study used a multiple-measurement approach to estimate the risk of respiratory virus infection in a Swiss school and assessed the effectiveness of air cleaners in a cross-over study design.
@@
 
-\noindent \textbf{Implications of all the available evidence} \smallskip
+% \noindent \textbf{Implications of all the available evidence} \smallskip
 
-\noindent Air cleaners have a tangible positive impact on air quality, significantly reducing particle concentrations.
+% \noindent Air cleaners have a tangible positive impact on air quality, significantly reducing particle concentrations.
 While the individual-level reduction in the risk of respiratory infections due to air cleaners may be relatively modest, the collective benefit at the population level, in terms of reducing illnesses and preventing school absences, is likely to be substantial.
@@
 
-\thispagestyle{empty}
-\sloppy
-\raggedbottom
+% \thispagestyle{empty}
+% \sloppy
+% \raggedbottom
 
@@
 Person-to-person transmission occurs primarily through the release of respiratory particles containing the viruses.
-Recently, the focus has been on small respiratory particles called aerosols, which have been found to carry the majority of viruses during respiratory activities.\cite{Fennelly2020} Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and travel long distances.\cite{Wang2020}
+Recently, the focus has been on small respiratory particles called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}.
+Unlike larger respiratory droplets, which tend to settle quickly, aerosols can remain suspended in the air for several hours and travel long distances\cite{Wang2020}.
 
-Improved ventilation systems are critical for a healthy indoor environment and can reduce the risk of respiratory transmission,\cite{Wang2021,Morawska2021} especially in schools where students spend much of their time indoors during the week.
-Portable HEPA-air filtration devices (air cleaners) may be another cost-effective alternative to upgrading ventilation systems.
-Still, the effects of air cleaners on the transmission of respiratory infections are less clear.
-Several studies have shown that air cleaners reduce particle concentrations in indoor air,\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed} which is associated with a reduced risk of respiratory infection and mortality.\cite{Gordon2014Resp,Kelly2023Atmos,DeAngelis2021EnvRes} In line with this, a population-level study reported a lower incidence of SARS-CoV-2 in US elementary schools using different ventilation strategies, with or without additional air filtration devices.\cite{Gettings2021} Although some studies could not find an association between particle concentrations and viral RNA load in airborne samples,\cite{Nor2021SciRep,Hanna2023PONE} recent studies showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in hospitals and other indoor settings.\cite{Morris2022,Ueki2022mSphere,Myers2022IndoorAir} Simulation studies have further demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission of SARS-CoV-2,\cite{Lindsley2021} and other respiratory viruses.\cite{Cortellessa2023Build} However, most simulation studies assume that the detection of RNA equals transmissible virus, despite recent data showing a relevant loss of viral infectivity in respiratory particles over time.\cite{Oswin2022PNAS} To date, it remains unclear whether reducing particle concentrations and removing bioaerosols will reduce indoor transmission of respiratory infections.
+Improved ventilation systems are critical for a healthy indoor environment and can reduce the risk of respiratory transmission\cite{Wang2021,Morawska2021}, especially in schools where students spend most of their time indoors during the week.
+Portable HEPA-air filtration devices (air cleaners) may be another cost-effective alternative to upgrading ventilation systems, but their impact on respiratory viral transmission is less clear.
+A population-level study reported a lower incidence of SARS-CoV-2 in US elementary schools using different ventilation strategies, including air filtration devices\cite{Gettings2021}.
+While several studies showed that air cleaners reduce particle concentrations\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed}, an association with viral RNA load in airborne samples could not be found\cite{Nor2021SciRep,Hanna2023PONE}, although recent studies showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in hospitals and other indoor settings\cite{Morris2022,Ueki2022mSphere,Myers2022IndoorAir}.
+Simulation studies have further demonstrated the efficacy of air cleaners in reducing the risk of indoor transmission of SARS-CoV-2\cite{Lindsley2021}, and other respiratory viruses\cite{Cortellessa2023Build}.
+However, most simulation studies assume that the detection of RNA equals transmissible virus, despite recent data showing a relevant loss of viral infectivity in respiratory particles over time\cite{Oswin2022PNAS}.
+To date, it remains unclear whether reducing particle concentrations and removing bioaerosols will reduce indoor transmission of respiratory infections.
 
-We used a multiple-measurement approach to study the transmission of respiratory viruses under non-pandemic conditions and the effect of air cleaners in schools with a cross-over study design in the winter of 2022/2023.
-We collected environmental data (CO$_2$, particle concentrations), epidemiological data (absences likely related to respiratory infections), audio recordings (coughing), and molecular data (detection of viruses in bioaerosol and human saliva samples) during a seven-week study period from January to March 2023 in two classes of a secondary school in Switzerland.
-We determined changes in particle concentrations, absences related to respiratory infections, coughing, and the positivity rate of saliva samples.
+% Several studies showed that air cleaners reduce particle concentrations in indoor air.\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed} which is associated with a reduced risk of respiratory infection and mortality.\cite{Gordon2014Resp,Kelly2023Atmos,DeAngelis2021EnvRes} In line with this,
 
-\section*{Methods}
+We used a multiple-measurement approach to study transmission of respiratory viruses under non-pandemic conditions and the effect of air cleaners in a school setting with a cross-over study design in the winter of 2022/2023.
+We collected environmental data (CO$_2$, particle concentrations), epidemiological data (absences likely related to respiratory infections), audio recordings (coughing), and molecular data (detection of viruses in bioaerosol and saliva samples) during a seven-week study period from January to March 2023 in two Swiss secondary school classes.
+We determined changes in particle concentrations, absences related to respiratory infections, coughing, and the rate of positive saliva samples.
 
-\subsection*{Study setting and design}
-
-\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, for seven weeks from January~16 to March~11, 2023.
-Figure~\ref{fig:study-setup} shows the schematic study setup.
-This study is reported as per the Strengthening the Reporting of Observational Studies in Epidemiology guideline (see STROBE checklist).
-
-\subsection*{Study intervention}
-
-\noindent We used a cross-over design to study the effectiveness of air cleaners (\cref{tab:study_design}).
-Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China).
-According to the manufacturer, these air cleaners run at clean air delivery rates of 2$\,\times\,$600 m$^{3}$/h.
-When testing the devices in an empty classroom with sub-micrometer sized particles, we measured a lower effective clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h (\supp~pp~\zpageref{sec:cadr}$-$5).
-
-\subsection*{Data collection}
-
-An overview of the collected data is provided in \supp~p~\zpageref{tab:data}. \medskip
-
-\noindent\textbf{Environmental data} \smallskip
-
-\noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number concentrations (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute.\cite{Banholzer2023PLoSMed} The devices did \emph{not} alert teachers when CO$_2$ levels were too high.\medskip
-
-\noindent\textbf{Epidemiological data} \smallskip
-
-\noindent At the beginning of the study, we collected aggregated data on age, sex, COVID-19 vaccination, and COVID-19 recovery status in the participating classes.
-Daily, we collected data on each absent student, \ie the absence and return dates and reason for the absence (illness, career days, or other).
-For absences due to illness, we recorded symptoms and the date of symptom onset.
-We defined a case of respiratory infection as an absence in which the student reported an illness with at least one respiratory symptom (\supp~p~\zpageref{sec:case-data}).
-The list of absences by date of symptom onset, absence and return, is shown in \supp~pp~\zpageref{tab:epi-data-line-list}$-$8.\medskip
-
-\subsection*{Audio recordings and cough detection}
-
-We installed a portable audio recorder (ZOOM H6; New York, USA) in each classroom to record sounds continuously.
-We determined the number of coughs per minute using an AI algorithm.\cite{Bertschinger2023IEEE} A recent study showed a significant correlation between coughing and the positivity rate of virus detection in airborne samples.\cite{Hanna2023PONE}
-
-\noindent\textbf{Molecular data analyses} \smallskip
-
-\noindent Both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) saliva testing.
-Samples were transported to the laboratory and stored at $-$80°C until further processing.\cite{Huber2021} The list of positive samples is shown in \supp~pp~\zpageref{tab:mol-data-line-list}$-$21.
-Furthermore, we collected airborne respiratory viruses in both classrooms with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL Phosphate-Buffered Saline.
-The Coriolis Micro Air ran shortly before and during break times (approximately 60\,min/day) to minimize noise.
-In one class, we also sampled with the BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft.
-Collins, CO, USA).\cite{Lednicky2016AST} BioSpot-VIVAS operated throughout the lessons.
-The removable parts of both sampling devices were regularly autoclaved.
-At the end of the day, samples were transported to the Institute of Infectious Diseases and stored at $-$80°C.
-Finally, we collected swabs from the HEPA filters of the air cleaners after each intervention phase (see \cref{tab:study_design}).
-The HEPA filters were removed from the air cleaners and divided into 20~fields.
-One sterile Phosphate-Buffered Saline-moistened swab per field was then taken for a total of 20~swabs per filter.
-
-Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL.
-The human saliva samples were analyzed directly without prior filtration.
-The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19 major respiratory viruses and viral subtypes, including SARS-CoV-2 (CoV), influenza (IFA/IFB), respiratory syncytial (RSV), metapneumovirus (MPV), adenovirus (AdV), rhinovirus (HRV), and parainfluenza (PIV).
-
-We also performed molecular genotyping for positive saliva, bioaerosol, air filter samples of adenovirus and influenza.
-For adenovirus, we used a PCR procedure to amplify three hypervariable regions (hexon, fiber, penton) of the viral genome and then sequenced the PCR products.\cite{Akello2021SciRep} For influenza, we tried to determine the molecular transmission network using whole genome sequencing.\cite{Kelly2022FrontiersImmuno}
-
-\subsection*{Statistical analyses and modeling}
-
-All statistical analyses were described in a statistical analysis plan,\cite{Banholzer2023SAP} which was published prior to the analyses.
-Bioaerosol samples and viral load concentrations could not be analyzed because there were too few positive samples.
-Otherwise, there were no major deviations from the statistical analysis plan.
-Minor deviations are documented in \supp~pp~\zpageref{sec:env-regression-model}$-$\zpageref{sec:multinomial-model}, including detailed descriptions of the models.
-\medskip
-
-\noindent\textbf{Particle concentrations} \smallskip
-
-\noindent We compared daily mean particle concentrations between study conditions.
-We estimated the reduction in particle concentrations with air cleaners using Bayesian log-linear regression models, adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases (\supp~p~\zpageref{sec:env-regression-model}).
-\medskip
-
-\noindent\textbf{Risk of infection} \smallskip
-
-\noindent We estimated the relative risk of infection with air cleaners using a Bayesian latent variable hierarchical regression model.
-The observed outcome of this model is the number of new respiratory cases $C$ (absences related to respiratory infections by date of symptom onset) on day $t$ in class $j$, which are modeled with a Negative Binomial distribution.
-The expected number of new cases is the weighted sum of the number of new infections $I_{js}$ (unobserved/latent outcome) in the previous days $s<t$, with the weights corresponding to the probability distribution of the incubation period.
-The number of new infections is related to the presence of air cleaners as follows:
-\begin{align}
-    \log I_{js} = \log F_{js} - \log N_{js} + \beta_0 + \beta_1 \cdot \text{AirCleaner}_{js},
-\end{align}
-where $F_{js}$ is the number of infections in the previous week (a proxy for the number of infectious students), $N_{js}$ is the cumulative number of infections (a proxy for the number of susceptible students), $\beta_0$ is the infection rate without air cleaners, and $\beta_1$ is the effect of air cleaners.
-Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the canton.
-A detailed description of the model and choice of priors for all model parameters are provided in \supp~pp~\zpageref{sec:transmission-model}$-$27.
-\medskip
-
-\noindent\textbf{Coughing} \smallskip
-
-\noindent We compared the daily number of detected coughs between study conditions.
-We estimated the reduction in the number of coughs with air cleaners using a Bayesian Negative Binomial regression model, using time in class as the model offset and adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases.
-In addition, we used a hierarchical regression model to estimate the association between the number of coughs and the virus-specific number of positive saliva samples.
-For this analysis, we used only days on which saliva samples were collected (Tuesdays and Thursdays, for a total of 27~days).
-A detailed description of these models is provided in \supp~p~\zpageref{sec:aud-regression-model}.
-\medskip
-
-\noindent\textbf{Saliva samples} \smallskip
-
-\noindent We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~p~\zpageref{sec:multinomial-model}) and linked the expected number of positive samples to the presence of air cleaners, adjusting for the cumulative number of positive tests.
-\medskip
-
-\noindent\textbf{Software}\smallskip
-
-\noindent All analyses were performed in R software (version 4.2.0).
-Model parameters were estimated in in Stan (version 2.21.0) with a Bayesian approach, using the Hamiltonian Monte Carlo algorithm with the No-U-Turn Sampler (NUTS).\cite{Hoffman2014} For each outcome, we report the posterior probability of a reduction with air cleaners.
-The estimated reduction is reported with the posterior mean and 95\%credible interval (CrI).
-Preprocessed data and code is available from \url{https://osf.io/38j9g}.
-
-\subsection*{Ethics statement}
-
-\noindent The Ethics Committee of the canton of Bern, Switzerland, approved the study (reference no.
-2021-02377).
-For the saliva samples, we included all students who were willing to participate and obtained written informed consent from their caregivers.
-
-\subsection*{Role of the funding source}
-
-The funder of the study had no role in study design, data collection, data analysis, data interpretation, or writing of the report.
-
-\newpage
+\clearpage
 
@@
 
-The study population consisted of 38~students (age 13$-$15 years, 19/19~female/male; \cref{tab:cases-overview-school}).
+The study population consisted of 38~students (age 13$-$15 years, 19/19~female/male; \Cref{tab:cases-overview-school}).
 Seven~students had been vaccinated or recovered from a SARS-CoV-2 infection within the last four months.
@@
 
-The mean aerosol number concentration was 95\,1/cm$^3$ (standard deviation [SD] 81\,1/cm$^3$) without vs 27\,1/cm$^3$ (SD 34\,1/cm$^3$) with air cleaners (\cref{fig:infection-risk}a).
+The mean aerosol number concentration was 95\,1/cm$^3$ (standard deviation [SD] 81\,1/cm$^3$) without vs 27\,1/cm$^3$ (SD 34\,1/cm$^3$) with air cleaners (\Cref{fig:infection-risk}a).
 The Bayesian regression model suggested a clear reduction in the aerosol concentration with air cleaners, with a posterior probability of 100\%.
-The model-estimated decrease was 76\% (95\%-CrI 63\% to 86\%), which was greater for larger (PM$_{10}$) than for smaller (PM$_{1-4}$) particles (\supp~p~\zpageref{fig:palas-results}).
+The model-estimated decrease was 76\% (95\%-CrI 63\% to 86\%), which was greater for larger (PM$_{10}$) than for smaller (PM$_{1-4}$) particles (\supp~Figure~\zref{fig:palas-results}).
 Daily mean CO$_2$ levels were comparable between study conditions (1,636\,ppm (SD 341\,ppm) without vs 1,769\,ppm (SD 391\,ppm) with air cleaners).
-There was little change in other environmental variables (\supp~p~\zpageref{fig:env-descriptives-other-vars}).
+There was little change in other environmental variables (\supp~Figure~\zref{fig:env-descriptives-other-vars}).
 
@@
 
-Absences related to respiratory infections included 22~cases without vs 13~cases with air cleaners (\cref{fig:infection-risk}b).
+Absences related to respiratory infections included 22~cases without vs 13~cases with air cleaners (\Cref{fig:infection-risk}b).
 The Bayesian latent variable hierarchical regression model suggested that air cleaners reduced the risk of infection, with a posterior probability of 91\%.
@@
 The estimated number of respiratory infections in school would have been 19 (95\%-CrI 9 to 37) if air cleaners had been installed throughout the study period, compared to 36 (95\%-CrI 12 to 92) infections if air cleaners had not been installed.
-Detailed estimation results are provided in \supp~pp~\zpageref{sec:detailed-redcap}$-$34.
+Detailed estimation results are provided in \supp~Text~\zref{sec:detailed-redcap}.
 
@@
 
-On average, we detected 3.1\,coughs/min (SD 1.2\,coughs/min) without vs 2.6\,coughs/min (SD 1.1\,coughs/min) with air cleaners (\cref{fig:infection-risk}c).
+On average, we detected 3.1\,coughs/min (SD 1.2\,coughs/min) without vs 2.6\,coughs/min (SD 1.1\,coughs/min) with air cleaners (\Cref{fig:infection-risk}c).
 The Bayesian model suggested that coughing was less frequent with air cleaners, with a posterior probability of 93\%.
 The adjusted relative risk of coughing with air cleaners was 0.93 (95\%-CrI 0.85 to 1.02).
-Coughing was associated with virus-specific transmission (\supp~p~\zpageref{fig:coughing-association}).
+Coughing was associated with virus-specific transmission (\supp~Figure~\zref{fig:coughing-association}).
 
@@
 We analyzed a total of 448~saliva samples.
-We detected 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus, respectively (\cref{fig:molecular-descriptives}a).
-There were 25~positive saliva samples in both study conditions (\cref{fig:infection-risk}d) and, based on the Bayesian model, the posterior probability that air cleaners reduced the positivity rate was 65\%, with an adjusted relative risk of 0.93 (95\%-CrI 0.49 to 1.61).
-The estimated relative risk was not sensitive to infection to testing delays (\supp~p~\zpageref{fig:mol-estimation-results-sensitivity}).
+We detected 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus, respectively (\Cref{fig:molecular-descriptives}a).
+There were 25~positive saliva samples in both study conditions (\Cref{fig:infection-risk}d) and, based on the Bayesian model, the posterior probability that air cleaners reduced the positivity rate was 65\%, with an adjusted relative risk of 0.93 (95\%-CrI 0.49 to 1.61).
+The estimated relative risk was not sensitive to infection to testing delays (\supp~Figure~\zref{fig:mol-estimation-results-sensitivity}).
 
@@
 To illustrate possible transmission chains within classes, we linked positive saliva samples of the same virus that were less than one week apart.
-Based on this spatiotemporal analysis, we identified 10~possible transmission chains (\cref{fig:molecular-descriptives}b).
+Based on this spatiotemporal analysis, we identified 10~possible transmission chains (\Cref{fig:molecular-descriptives}b).
 The longest potential transmission chains occurred in January, referring to a cluster of adenovirus infections in class~A and influenza~B infections in class~B.
 Molecular genotyping to verify the proposed transmission network was unsuccessful because we could not amplify and sequence any of the gene targets.
-Furthermore, the rate of positive bioaerosol samples was low.
 We also analyzed 105~bioaerosol samples and detected in two of them viral RNA (1~rhinovirus in class~A and 1~adenovirus in class~B).
@@
 %Nevertheless, these estimates are subject to unmeasured confounding because the delay may be sample-specific and repeated positive tests from the same person could not be precluded because testing was anonymized.
+
+\clearpage
 
@@
 We detected a range of respiratory virus in students' saliva, mainly adenovirus, influenza~B and rhinovirus, with only three positive SARS-CoV-2 saliva samples.
-In a previous, similar study in the same setting, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 omicron wave in the winter of 2021/2022 and detected almost exclusively SARS-CoV-2 in the students' saliva.\cite{Banholzer2023PLoSMed} A similar shift in the pattern of respiratory viruses has been observed in other studies.\cite{Nygaard2023Lancet,Sauteur2022EuroSurv} We found a reduction in the risk of SARS-CoV-2 infection for mask wearing, but not for air cleaners, possibly because the air cleaners were introduced only at the end of the study, when most students were already infected with SARS-CoV-2.
+In a previous, similar study in the same setting, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 omicron wave in the winter of 2021/2022 and detected almost exclusively SARS-CoV-2 in the students' saliva\cite{Banholzer2023PLoSMed}.
+A similar shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.
+We found a reduction in the risk of SARS-CoV-2 infection for mask wearing, but not for air cleaners, possibly because the air cleaners were introduced only at the end of the study, when most students were already infected with SARS-CoV-2.
 
@@
 
-It is well documented that air cleaners improve indoor air quality.\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed} There are several reasons why the effect of air cleaners is probably smaller than universal mask wearing, which has been shown to be a very effective infection control measure.\cite{Banholzer2023PLoSMed,Gettings2021,Leung2020NatMed,Milton2013PLoSPathogens} Unlike masks, air cleaners cannot prevent transmission outside the classroom or transmission due to close range, high particle density.
-Prolonged and close contact may be necessary for transmission of some respiratory viruses,\cite{Leung2020NatMed,Brankston2007LancetID} or make transmission more likely despite prior vaccination or infection.\cite{Lind2023NatCommun} Our results further suggests that air cleaners are more effective at removing larger particles ($>5\mu$m), which also explains the difference between our measured and the manufacturer's reported clean air delivery rate (see \supp~pp~\zpageref{sec:cadr}$-$5).
-However, many respiratory viruses are carried in smaller particles, which are more relevant for transmission ($\leq5\mu$m).\cite{Fennelly2020} Finally, classroom activity, airflow and other unobserved, confounding factors make it challenging to evaluate the effects of air cleaners on transmission in real-world settings.
-Nevertheless, it is important to evaluate their effectiveness in such settings and to compare them with the hypothetical effects suggested by simulation studies.\cite{Lindsley2021,Cortellessa2023Build}
-
-%Although the risk reduction effect estimated in this study is rather small, the cumulative impact may be relevant because minimizing the risk of transmission of respiratory infections in schools is an important public health consideration\cite{Beale2023JOMT} and because prevention and control measures in schools, such as air cleaners, must be adequate and acceptable\cite{WHO2020SchoolMeasures}.
+It is well documented that air cleaners improve indoor air quality\cite{Park2020Build,Buising2022InfContr,Banholzer2023PLoSMed}.
+There are several reasons why the effect of air cleaners is probably smaller than universal mask wearing, which has been shown to be a very effective infection control measure\cite{Banholzer2023PLoSMed,Gettings2021,Leung2020NatMed,Milton2013PLoSPathogens}.
+Unlike masks, air cleaners cannot prevent transmission outside the classroom or transmission due to close range, high particle density.
+Prolonged and close contact may be necessary for transmission of some respiratory viruses\cite{Leung2020NatMed,Brankston2007LancetID}, or make transmission more likely despite prior vaccination or infection\cite{Lind2023NatCommun}.
+Our results further suggests that air cleaners are more effective at removing larger particles ($>5\mu$m), which also explains the difference between our measured and the manufacturer's reported clean air delivery rate (see \supp~Figure~\zref{sec:cadr}).
+However, many respiratory viruses are carried in smaller particles, which are more relevant for transmission ($\leq5\mu$m)\cite{Fennelly2020}.
+Finally, classroom activity, airflow and other unobserved, confounding factors make it challenging to evaluate the effects of air cleaners on transmission in real-world settings.
 
 % economic and psychological impact of air cleaners
+% further potential concent: 
+% Although the risk reduction effect estimated in this study is rather small, the cumulative impact may be relevant because minimizing the risk of transmission of respiratory infections in schools is an important public health consideration\cite{Beale2023JOMT} and because prevention and control measures in schools, such as air cleaners, must be adequate and acceptable\cite{WHO2020SchoolMeasures}.
 
+% short:
+%The beneficial effects of air cleaners on indoor air quality and transmission come at a reasonable cost.
+The portable air cleaners used in our study cost approximately USD\,250 per unit.
+Their operating cost-effectiveness in providing clean air could be even higher than that of a ventilation system when compared in parallel using the same air delivery ratings.\cite{Noh2016EnBuild} Therefore, air cleaners could be a cost-effective public health measure, particularly during pandemics or epidemics when there is greater exposure to respiratory infections, and greater concern of becoming infected.\cite{Yuerekli2022IJERPH} However, their acceptance may be hindered by noise, space limitations, technical issues, and maintenance requirements.\cite{Sanguinetti2022IndoorAir} Therefore, investments in professional building ventilation systems are still preferred in the long run.\cite{Nardell2016}
+
+% long:
 The beneficial effects of air cleaners on indoor air quality and transmission come at a reasonable cost.
 The portable air cleaners used in our study cost approximately USD\,250 per unit.
-Their operating cost effectiveness in providing clean air could be even higher than that of a ventilation system when compared in parallel using the same air delivery ratings.\cite{Noh2016EnBuild} Therefore, air cleaners could be a cost-effective public health measure, particularly during pandemics or epidemics when there is greater exposure to respiratory infections.
+Their operating cost-effectiveness in providing clean air could be even higher than that of a ventilation system when compared in parallel using the same air delivery ratings\cite{Noh2016EnBuild}.
+Therefore, air cleaners could be a cost-effective public health measure, particularly during pandemics or epidemics when there is greater exposure to respiratory infections.
 In addition, air cleaners can have a psychological impact.
 On the one hand, they can give students and teachers a sense of support during difficult times.
-For example, it has been shown that becoming infected was one of the main concerns for students during the pandemic.\cite{Yuerekli2022IJERPH} A safe and supportive environment is a key determinant of student success and well-being in school.\cite{Kutsyuruba2015RevEduc} On the other hand, air cleaners may not be well accepted by teachers and authorities due to noise, limited space, technical issues, and the required maintenance.\cite{Sanguinetti2022IndoorAir} Therefore, investments in professional building ventilation systems are still preferred in the long run.\cite{Nardell2016}
+For example, it has been shown that becoming infected was one of the main concerns for students during the pandemic\cite{Yuerekli2022IJERPH}.
+A safe and supportive environment is a key determinant of student success and well-being in school\cite{Kutsyuruba2015RevEduc}.
+On the other hand, air cleaners may not be well accepted by teachers and authorities due to noise, limited space, technical issues, and the required maintenance\cite{Sanguinetti2022IndoorAir}.
+Therefore, investments in professional building ventilation systems are still preferred in the long run\cite{Nardell2016}.
 
@@
 In contrast, influenza~B spread throughout the study in class~B, and adenovirus infections were detected only in the last week of the study.
-Furthermore, adenovirus infections tend to be mild,\cite{Kunz2010CIDR} and less frequently associated with cough than influenza,\cite{Ma2018RMV} which is consistent with the comparatively lower frequency of coughing in class~B.
-Taken together, the class-specific, spatiotemporal patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred within the classrooms.
+Furthermore, adenovirus infections tend to be mild\cite{Kunz2010CIDR}, and less frequently associated with cough than influenza\cite{Ma2018RMV},, consistent with the comparatively lower frequency of coughing in class~B.
+Taken together, the class-specific, spatiotemporal patterns indicate that transmission of respiratory infections may have occurred within the classrooms.
 
@@
 In addition, we could only approximate the incubation period for each epidemiological case.
-Finally, the study was conducted in two classrooms of one Swiss school and although results are likely to apply to many settings in Switzerland and other European countries, they will not be applicable to settings in the global South.
+Finally, although the study results are likely to apply to many settings in Switzerland and other European countries, they will not be applicable to settings in the global South.
 
@@
 
+% new:
+% In conclusion, a wide range of respiratory viruses, but rarely SARS-CoV-2, was detected in students under non-pandemic conditions when public health measures were lifted.
+Airborne detection was rare, suggesting that respiratory viruses other than SARS-CoV-2 may be more difficult to detect and that prolonged close contact may be required for transmission.
+The risk reduction of respiratory infections conferred by air cleaners may be modest at the individual level, but the benefit at the population level in terms of illness and  absences prevented is likely to be important.
+Future studies should examine the cost-effectiveness of using air cleaners in congregate settings.
+
+% old:
 In conclusion, using epidemiological, environmental, and molecular data, our study showed that a wide range of respiratory viruses, but rarely SARS-CoV-2, was detected in students under non-pandemic conditions when public health measures were lifted, and that transmission of these respiratory viruses likely occurred in schools.
@@
 
-\newpage
+\clearpage
+
+\section*{Methods}
+
+This study is reported as per the Strengthening the Reporting of Observational Studies in Epidemiology guideline (see STROBE checklist).
+
+\subsection*{Study setting and design}
+
+\noindent We collected data in two classrooms of a secondary school (age of students 14-17~years) in the canton of Solothurn, Switzerland, for seven weeks from January~16 to March~11, 2023.
+Figure~\ref{fig:study-setup} shows the schematic study setup.
+
+\subsection*{Study intervention}
+
+\noindent We used a cross-over design to study the effectiveness of air cleaners (\Cref{tab:study_design}).
+Air cleaners refer to commercially available portable HEPA-filtration devices (Xiaomi Mi Air Pro 70m², Shenzhen, China).
+According to the manufacturer, these air cleaners run at clean air delivery rates of 2$\,\times\,$600 m$^{3}$/h.
+When testing the devices in an empty classroom with sub-micrometer sized particles, we measured a lower effective clean air delivery rate of 2$\,\times\,$420 m$^{3}$/h (\supp~Text~\zref{sec:cadr}).
+
+\subsection*{Data collection}
+
+An overview of the collected data is provided in \supp~Table~\zref{tab:data}. \medskip
+
+\noindent\textbf{Environmental data} \smallskip
+
+\noindent An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, aerosol number (particle diameter between 175~nm to 20~$\mu$m) and particle mass concentrations (PM; PM$_1$, PM$_{2.5}$, PM$_4$, PM$_{10}$) by minute\cite{Banholzer2023PLoSMed}.
+The devices did \emph{not} alert teachers when CO$_2$ levels were too high.
+\medskip %
+
+\noindent\textbf{Epidemiological data} \smallskip
+
+\noindent At study start, we collected aggregated data on age, sex, COVID-19 vaccination and recovery status in the participating classes.
+Daily, we collected data on each absent student, \ie the absence and return dates and reason for the absence (illness, career days, or other).
+For absences due to illness, we recorded symptoms and the date of symptom onset.
+We defined a case of respiratory infection as an absence in which the student reported an illness with at least one respiratory symptom (\supp~Text~\zref{sec:case-data}).
+All absences are listed in \supp~Table~\zref{tab:epi-data-line-list}.\medskip %
+
+\subsection*{Audio recordings and cough detection}
+
+We installed portable audio recorders (ZOOM H6; New York, USA) to record sounds continuously.
+We determined the number of coughs per minute using an AI algorithm\cite{Bertschinger2023IEEE}.
+A recent study showed a significant correlation between coughing and airborne viral detection\cite{Hanna2023PONE}.\medskip
+
+\noindent\textbf{Molecular data analyses} \smallskip
+
+\noindent Both classes participated in repetitive bi-weekly (Tuesdays and Thursdays) saliva testing.
+Samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Huber2021}.
+All positive samples are listed in \supp~Table~\zref{tab:mol-data-line-list}.
+Furthermore, we collected airborne respiratory viruses in both classrooms with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France), running at 200\,l/min and collecting into 15\,mL Phosphate-Buffered Saline.
+The Coriolis Micro Air ran shortly before and during break times (approximately 60\,min/day) to minimize noise.
+In one class, we also sampled with the BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft.
+Collins, CO, USA)\cite{Lednicky2016AST}, which operated throughout lessons.
+The removable parts were regularly autoclaved.
+At the end of the day, samples were transported to the Institute of Infectious Diseases and stored at $-$80°C.
+Finally, we collected swabs from the air cleaners' HEPA filters after each intervention phase (see \Cref{tab:study_design}).
+The HEPA filters were removed and divided into 20~fields.
+One sterile Phosphate-Buffered Saline-moistened swab per field was then taken for a total of 20~swabs per filter.
+
+Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL.
+Saliva samples were analyzed directly without prior filtration.
+The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19 major respiratory viruses and viral subtypes, including SARS-CoV-2, influenza, respiratory syncytial, metapneumovirus, adenovirus, rhinovirus, and parainfluenza.
+
+We also performed molecular genotyping for positive saliva, bioaerosol, and air filter samples of adenovirus and influenza\cite{Kelly2022FrontiersImmuno}.
+For adenovirus, we used a PCR procedure to amplify three hypervariable regions (hexon, fiber, penton) of the viral genome and then sequenced the PCR products\cite{Akello2021SciRep}.
+For influenza, we tried to determine the molecular transmission network using whole genome sequencing.
+
+\subsection*{Statistical analyses and modeling}
+
+All statistical analyses were described in a statistical analysis plan\cite{Banholzer2023SAP}.
+Bioaerosol samples and viral load concentrations could not be analyzed because there were too few positive samples.
+Further minor deviations from the statistical analysis plan are documented in \supp~Texts~\zref{sec:env-regression-model}$-$\zref{sec:multinomial-model}, including detailed descriptions of the models.
+\medskip
+
+\noindent\textbf{Particle concentrations} \smallskip
+
+\noindent We compared daily mean particle concentrations between study conditions.
+We estimated the reduction in particle concentrations with air cleaners using Bayesian log-linear regression models, adjusting for adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases (\supp~Text~\zref{sec:env-regression-model}).\medskip %
+
+\noindent\textbf{Risk of infection} \smallskip
+
+\noindent We estimated the relative risk of infection with air cleaners using a Bayesian latent variable regression model (\supp~Text~\zref{sec:transmission-model}).
+The number of new respiratory cases $C$ (observed absences related to respiratory infections by date of symptom onset) on day $t$ in class $j$ are modeled with a Negative Binomial distribution.
+The expected number of new cases is the weighted sum of the number of new infections $I_{js}$ (latent variable) in the previous days $s<t$, with the weights corresponding to the probability distribution of the incubation period.
+The number of new infections is related to the presence of air cleaners as follows:
+\begin{align}
+    \log I_{js} = \log F_{js} - \log N_{js} + \beta_0 + \beta_1 \cdot \text{AirCleaner}_{js},
+\end{align}
+where $F_{js}$ is the number of infections in the previous week (a proxy for the number of infectious students), $N_{js}$ is the cumulative number of infections (a proxy for the number of susceptible students), $\beta_0$ is the infection rate without air cleaners, and $\beta_1$ is the effect of air cleaners.
+Furthermore, the effect of air cleaners is adjusted for class-specific effects, the number of students in class, the daily air change rate, and the weekly positivity rate for COVID-19 and the consultations for influenza-like illnesses in the canton.
+\medskip
+
+\noindent\textbf{Coughing} \smallskip
+
+\noindent We estimated the reduction in the daily number of coughs with air cleaners using a Bayesian Negative Binomial regression model, using time in class as the model offset and adjusting for adjusting for class and weekday effects, the number of students in class, the air change rate, and the cumulative number of respiratory cases (\supp~Text~\zref{sec:aud-regression-model}).
+In addition, we estimated the association between the number of coughs and the virus-specific number of positive saliva samples, using only the days when saliva samples were collected (Tuesdays and Thursdays, for a total of 27~days).\medskip %
+
+\noindent\textbf{Saliva samples} \smallskip
+
+\noindent We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (\supp~Text~\zref{sec:multinomial-model}) and linked the expected number of positive samples to the presence of air cleaners, adjusting for the cumulative number of positive tests.
+\medskip
+
+\noindent\textbf{Software}\smallskip
+
+\noindent All analyses were performed in R software (version 4.2.0) and model parameters estimated in Stan (version 2.21.0)\cite{Carpenter2017,Hoffman2014}.
+For each outcome, we report the posterior probability of a reduction with air cleaners.
+The estimated reduction is reported with the posterior mean and 95\%credible interval (CrI).
+Preprocessed data and code is available from \url{https://osf.io/38j9g}.
+
+\subsection*{Ethics statement}
+
+\noindent The Ethics Committee of the canton of Bern, Switzerland, approved the study (reference no.
+2021-02377).
+For the saliva samples, we included all students who were willing to participate and obtained written informed consent from their caregivers.
+
+\clearpage
 
 %TC:ignore
+
+\section*{Acknowledgements}
+We would like to thank the schools, teachers, and students participating in the study.
+We are grateful to the Educational Department of the canton of Solothurn for their support throughout the study.
+We would also like to thank Ronald Dijkman for the influenza molecular genotyping and the biosecurity team (Julia Feldmann, Monika Gsell, Kathrina Summermatter) from the Institute for Infectious Diseases at the University of Bern for their assistance with the bioaerosol devices.
+Finally, we are indebted to the student assistants (Khadija Ahmed, Marie-Joséphine Brancato, Michelle Bürki, Santiago Martinez, Moric Toszeghi, Sylvain Wasmer) who helped with the data collection in the schools.
+
+\section*{Data availability}
+Preprocessed data and code are available from \url{https://osf.io/38j9g}.
+
+\section*{Funding}
+This study was funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
+NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
+ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498).
+The funder of the study had no role in study design, data collection, data analysis, data interpretation, or writing of the report.
 
@@
 
-\section*{Acknowledgements}
-We would like to thank the schools, teachers, and students participating in the study.
-We are grateful to the Educational Department of the canton of Solothurn for their support throughout the study.
-We would also like to thank Ronald Dijkman for the influenza molecular genotyping and the biosecurity team (Julia Feldmann, Monika Gsell, Kathrina Summermatter) from the Institute for Infectious Diseases at the University of Bern for their assistance with the bioaerosol devices.
-Finally, we are indebted to the student assistants (Khadija Ahmed, Marie-Joséphine Brancato, Michelle Bürki, Santiago Martinez, Moric Toszeghi, Sylvain Wasmer) who helped with the data collection in the schools.
-
-\section*{Data sharing statement}
-Preprocessed data and code is available from \url{https://osf.io/38j9g}.
+\clearpage
 

### 2023-12-04 doc/paper/supplements.tex
Updates from Overleaf

@@
 
-\title{\LARGE\singlespacing{\textbf{Appendix} \\ \medskip
-Air cleaners and respiratory infections in schools: \\ A modeling study using epidemiological, environmental, and molecular data}}
+\title{\LARGE\singlespacing{\textbf{Supplmenetary Material} \\ \medskip
+Air cleaners and respiratory infections in schools: \\ A multi-measurement observational study}}
+
+% long: Air cleaners and respiratory infections in schools: \\ A modeling study using epidemiological, environmental, and molecular data
 
@@
 
-The clean air delivery rate (CADR) is typically used to quantify the cleaning efficiency of air cleaners.\cite{Nelson1993IndoorEnv} It is expressed in m$^3$/s and describes the volumetric flow rate of particle-free, clean air that an air cleaner delivers into the room.
+The clean air delivery rate (CADR) is typically used to quantify the cleaning efficiency of air cleaners\cite{Nelson1993IndoorEnv}.
+It is expressed in m$^3$/s and describes the volumetric flow rate of particle-free, clean air that an air cleaner delivers into the room.
 The CADR is equal to the product of the filtration efficiency and the volumetric flow rate passing through the device.
@@
 We followed the approach by K\"{u}pper et al.
-to determine the CADR.\cite{Kuepper20199AAQR} The approach is based on the release of fine aerosols (particle diameter $D$<1$\mu$m) in a closed, unventilated room by measuring the decay rate $k$ of the particle concentration $N(t)$ caused by the air cleaner, \ie
+to determine the CADR\cite{Kuepper20199AAQR}.
+The approach is based on the release of fine aerosols (particle diameter $D$<1$\mu$m) in a closed, unventilated room by measuring the decay rate $k$ of the particle concentration $N(t)$ caused by the air cleaner, \ie
 \begin{align}
@@
 
-Reports about absences were entered electronically into a REDcap database.\cite{Harris2009,Harris2019} \Cref{tab:epi-data-line-list} shows the line list data for respiratory cases based on epidemiological data.
+Reports about absences were entered electronically into a REDcap database\cite{Harris2009,Harris2019} \Cref{tab:epi-data-line-list}.
+shows the line list data for respiratory cases based on epidemiological data.
 For each class and case, it shows the date when the student was absent, s/he reported symptoms, and returned to school.
@@
 
-The prior distributions for the virus-specific incubation periods are based on estimates published in the literature.\cite{McAloon2020,Lessler2009LancetID} The distributions are shown in \Cref{fig:prior-pin}.
+The prior distributions for the virus-specific incubation periods are based on estimates published in the literature\cite{McAloon2020,Lessler2009LancetID}.
+The distributions are shown in \Cref{fig:prior-pin}.
 Since we could not obtain prior estimates for the incubation period of metapneumovirus (MPV) from the literature, we instead formed a distribution by using the equally weighted average of the parameters from the other distributions.
@@
 
-We use weakly informative priors for all modeling parameters following recommendations on the choice of priors.\cite{Gelman2008StatMed,Gelman2008StatAnnals,Gelman2020RegOther,Stan2020Priors,Gabry2023Priors} The continuous adjustment variables are standardized to have zero mean and a standard deviation of 0.5.
+We use weakly informative priors for all modeling parameters following recommendations on the choice of priors\cite{Gelman2008StatMed,Gelman2008StatAnnals,Gelman2020RegOther,Stan2020Priors,Gabry2023Priors}.
+The continuous adjustment variables are standardized to have zero mean and a standard deviation of 0.5.
 \begin{align*}
@@
 \Cref{fig:env-descriptives-other-vars} compares additional environmental variables, \ie CO$_2$, air change rate, relative humidity, and temperature.
-The air change rate was estimated from indoor CO$_2$ levels using a transient mass balance model.\cite{Batterman2017IJERPH} There are no considerable differences between in any of the variables between study conditions.
+The air change rate was estimated from indoor CO$_2$ levels using a transient mass balance model\cite{Batterman2017IJERPH}.
+There are no considerable differences between in any of the variables between study conditions.
 Furthermore, numerical estimation results for the reduction in particle concentrations with air cleaners are shown in \Cref{tab:palas-est-results}.
@@
     \multicolumn{6}{p{13cm}}{\scriptsize
-        ESS is the effective sample size, \ie the number of independent MCMC samples with estimation power equivalent to the total number of autocorrelated samples,\cite{Stan2022} and $\hat{R}$ is the Gelman-Rubin convergence diagnostic.\cite{Gelman1992} Low ESS or $\hat{R}$ or $\hat{R}>1.10$ indicate bad convergence of the model.\cite{Gelman2013}}
+        ESS is the effective sample size, \ie the number of independent MCMC samples with estimation power equivalent to the total number of autocorrelated samples\cite{Stan2022}, and $\hat{R}$ is the Gelman-Rubin convergence diagnostic\cite{Gelman1992}.
+Low ESS or $\hat{R}$ or $\hat{R}>1.10$ indicate bad convergence of the model\cite{Gelman2013}.}
     \end{tabular}

### 2023-12-04 doc/comparison-mcid12/manuscript.tex

@@
-\section{Discussion}
+\PassOptionsToPackage{top=3cm,left=3cm,right=3cm,bottom=3cm}{geometry}
+\documentclass[fleqn,11pt]{wlscirep}
+
+\usepackage{import}
+\usepackage{../paper/main-lancet}
+
+\renewcommand{\paragraph}[1]{\vspace{0.3cm}\noindent\underline{\emph{#1}}\hfill\noindent}
+
+\begin{document}
+
+\doublespacing
+
+\title{\bfseries\LARGE\singlespacing{Brief report: SARS-CoV-2 more frequently detected in indoor air than other respiratory viruses}}
+
+% author list
+\author[1,2]{Nicolas Banholzer}
+\author[2,3]{Philipp Jent}
+\author[2,4]{Pascal Bittel}
+\author[1]{Kathrin Zürcher}
+\author[4]{Lavinia Furrer}
+\author[1]{Simon Bertschinger}
+\author[5]{Ernest Weingartner}
+\author[2,4]{Alban Ramette}
+\author[1,6,7]{Matthias Egger}
+\author[2,8]{Tina Hascher}
+\author[1*,2]{Lukas Fenner}
+
+\affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
+\affil[2]{Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland}
+\affil[3]{Department of Infectious Diseases, Inselspital, Bern University Hospital, University of Bern, Bern, Switzerland}
+\affil[4]{Institute for Infectious Diseases, University of Bern, Bern, Switzerland}
+\affil[5]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
+\affil[6]{Population Health Sciences, University of Bristol, Bristol, UK}
+\affil[7]{Centre for Infectious Disease Epidemiology and Research, University of Cape Town, Cape Town, South Africa}
+\affil[8]{Institute of Educational Science, University of Bern, Bern, Switzerland}
+
+\affil[*]{Corresponding author: lukas.fenner@unibe.ch}
+
+\vspace{2em}
+
+%TC:ignore
+
+\begin{information}\normalfont
+\textbf{Word count:} Abstract 252, Manuscript 3,244 \\
+\textbf{Display items:} Figure 1 \\
+\textbf{References:} 37 \\
+\textbf{Supporting information:} STROBE checklist, Supplementary Data with texts, tables, and figures
+% \par
+\end{information}
+
+\begin{abstract}\normalfont
+\noindent We compared air, air filters, and saliva samples from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave) and winter 2022/2023 (after the COVID-19 pandemic) in a Swiss school setting.
+Whereas during the pandemic we detected almost exclusively SARS-CoV-2 in human saliva (19~samples of SARS-CoV-2, 2~samples of non-SARS-CoV-2), the most common respiratory viruses after the pandemic were adenovirus, influenza~B, and rhinoviruses (3~samples of SARS-CoV-2, 47~samples of non-SARS-CoV-2).
+Despite the same study setting and environmental conditions, airborne detection of SARS-CoV-2 was relatively more frequent than the detection of non-SARS-CoV-2 viruses (10~samples of SARS-CoV-2 in 2021/2022 vs 2~samples of non-SARS-CoV-2 in 2022/2023).
+Our comparison suggests that respiratory viruses other than SARS-CoV-2 may be less detectable in the air and that close contact is required for transmission.
+\medskip
+
+\par
+\end{abstract}
+
+\flushbottom
+\maketitle
+
+\vspace{2em}
+
+\vspace{0.5em}
+
+\noindent\textbf{Keywords:} respiratory viruses, airborne transmission, molecular detection
+% maximum of 3-5 keywords
+
+\thispagestyle{empty}
+\sloppy
+\raggedbottom
+
+\newpage
+%TC:endignore
+
+\setcounter{page}{1}
+
+%\section*{Introduction}
+
+\section*{Methods}
+
+Data was collected in two classes of two secondary schools in winter 2021/2022 and two classrooms of one secondary school in winter 2022/2023 (age of students 14-17~years) in the canton of Solothurn, Switzerland, both over a seven-week study period from end of January to beginning of March.
+In winter 2021/2022, a general mask mandate was in place for the first 3~study weeks, followed by 2~weeks without intervention, and another 2~weeks of air cleaners installed in two classrooms.
+In winter 2022/2023, air cleaners were installed for half of the study period following a cross-over design.
+An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, temperature, and humidity in all classrooms.
+
+School classes participated in repetitive weekly (winter 2021/2022) and bi-weekly (winter 2022/2023) testing for a panel of respiratory infections.
+Airborne respiratory viruses were collected in each classroom with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France) and a BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft.
+Collins, CO, USA)\cite{Lednicky2016AST}.
+The removable parts were regularly autoclaved.
+Finally, swabs from the air cleaners' HEPA filters were collected at the end of the study (winter 2021/2022) and after each intervention phase (winter 2022/2023).
+The HEPA filters were removed and divided into 20~fields.
+One sterile Phosphate-Buffered Saline-moistened swab per field was then taken for a total of 20~swabs per filter.
+Saliva and airborne samples were transported to the Institute of Infectious Diseases and stored at $-$80°C until further processing\cite{Huber2021}.
+
+Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL.
+Saliva samples were analyzed directly without prior filtration.
+The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19 major respiratory viruses and viral subtypes, including SARS-CoV-2, influenza, respiratory syncytial, metapneumovirus, adenovirus, rhinovirus, and parainfluenza.
+
+Descriptive statistics were utilized to depict differences in the type, number, and proportion of respiratory viruses between winter 2021/2022 and winter 2022/2023.
+
+\section*{Results}
+
+\section*{Discussion}
 
@@
 However, several studies showed that influenza\cite{Bischoff2013JID,Pan2017mSphere}, rhinovirus\cite{Myatt2004AJRCCM}, and adenovirus\cite{Nguyen2017OFID,Pan2017mSphere} can also be detected in airborne samples.
+
+\bibliography{../paper/references.bib}
+
+\end{document}

### 2023-12-04 analysis/study_comaprison.Rmd
comparison figure new

--- 2023-07-10 analysis/study_comaprison.Rmd
@@
 saliva <- data.frame(
-    year = c(rep(2022, 6), rep(2023, 6)),
+    year_num = c(rep(2022, 6), rep(2023, 6)),
     virus = rep(c("Flu", "HRV", "AdV", "CoV", "MPV", "PIV"), 2),
@@
   ) %>%
+  mutate(year = ifelse(year_num == 2022, "Winter 2021/2022", "Winter 2022/2023"),
+         year = factor(year, levels = c("Winter 2021/2022", "Winter 2022/2023")),
+         virus = factor(virus, levels = c("Flu", "HRV", "AdV", "CoV", "MPV", "PIV"))) %>%
+  filter(!is.na(n)) %>%
   group_by(year) %>%
-  mutate(p = round(100 * n / sum(n, na.rm = T))) %>%
+  arrange(virus) %>%
+  mutate(p = n / sum(n, na.rm = T),
+         ymax = cumsum(p),
+         ymin = lag(ymax, n = 1),
+         ymin = ifelse(is.na(ymin), 0, ymin),
+         labelPosition = (ymax + ymin) / 2,
+         label = paste0(virus, ": ", n)) %>%
   ungroup() %>%
-  mutate(year = factor(year, levels = c(2022, 2023)),
-         virus = factor(virus, levels = c("Flu", "HRV", "AdV", "CoV", "MPV", "PIV"))) %>%
-  ggplot(mapping = aes(y = factor(year), x = p, fill = virus)) +
-  geom_bar(stat = "identity", position = position_stack(reverse = TRUE), width = .5) +
-  geom_text(mapping = aes(label = p), size = 8 / cm(1), color = "white", position = position_stack(vjust = 0.5, reverse = T)) +
-  scale_x_continuous(expand = expansion(mult = c(0,0.025))) +
-  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
-  labs(x = "Proportion of positive saliva samples (%)", y = "Study year", fill = "Virus", title = "a") +
-  theme_custom() +
-  theme(legend.position = "top",
+  mutate(labelPosition = ifelse(year == "Winter 2022/2023" & virus == "MPV", labelPosition - .015, labelPosition),
+         labelPosition = ifelse(year == "Winter 2021/2022" & virus == "Flu", ymin, labelPosition),
+         labelPosition = ifelse(year == "Winter 2021/2022" & virus == "AdV", ymax, labelPosition))
+
+saliva_pl <- saliva %>%
+  ggplot() +
+  facet_wrap(~ year_num, nrow = 1) +
+  geom_rect(mapping = aes(ymax = ymax, ymin = ymin, xmax = 4, xmin = 3, fill = virus)) +
+  geomtextpath::coord_curvedpolar(theta = "y", start = 1.9*pi) +
+  geom_text(mapping = aes(x = 5.25, y=labelPosition, label=label, color=virus), size = 8/cm(1)) +
+  geom_text(data = data.frame(x = 1, y = 1, year_num = c(2022, 2023), label = c(2022, 2023)), mapping = aes(x = x, label = label, y = y), 
+            size = 8/cm(1), fontface = 2) +
+  scale_x_continuous(limits = c(1, 5.25), expand = c(0,0)) +
+  scale_y_continuous(expand = c(0,0),  position = "top") +
+  scale_fill_manual(values = RColorBrewer::brewer.pal(9, "BuPu")[4:9]) +
+  scale_color_manual(values = RColorBrewer::brewer.pal(9, "BuPu")[4:9]) +
+  ggtitle("a") +
+  theme_void() +
+  theme(legend.position = "none",
+        plot.title = element_text(size = 10, face = 2),
         plot.title.position = "plot",
         legend.key.height = unit(.1, "cm"),
-        legend.key.width = unit(.3, "cm")) +
-  guides(fill = guide_legend(nrow = 1))
+        legend.key.width = unit(.3, "cm"),
+        strip.text = element_blank()) +
+  guides(fill = guide_legend(nrow = 1),
+         color = "none")
 
-saliva
+saliva_pl
 ```
@@
 pos_samples <- data.frame(
-  year = factor(c(rep(2022, 3), rep(2023, 3)), levels = c(2022, 2023)),
+  year_num = factor(c(rep(2022, 3), rep(2023, 3)), levels = c(2022, 2023)),
   type = factor(rep(c("Saliva", "Bioaerosols", "HEPA-filter"), times = 2), levels = c("Saliva", "Bioaerosols", "HEPA-filter")),
-  n = c(c(21, 10, 8), c(50, 2, 4))
+  n_pos = c(c(21, 10, 8), c(50, 2, 4)),
+  n_tot = c(c(262, 130, 40), c(448, 105, 160))
 ) %>%
-  ggplot(aes(x = n, y = year, pattern = year, color = year, pattern_color = year)) +
-  geom_bar_pattern(stat = "identity", fill = "white", pattern_density = 0.1, pattern_spacing = 0.1, width = .75) +
-  geom_text(aes(label = n), size = 8 / cm(1), hjust = -1) +
-  facet_wrap(~ type, ncol = 1) +
-  scale_x_continuous(expand = expansion(add = c(0,9.5))) +
-  scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) +
-  scale_pattern_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) +
-  labs(y = "Study year", x = "Frequency", title = "b") +
-  theme_custom() +
-  theme(legend.position = "none",
-        plot.title.position = "plot")
+  mutate(n_neg = n_tot - n_pos,
+         year = ifelse(year_num == 2022, "Winter 2021/2022", "Winter 2022/2023"),
+         year = factor(year, levels = c("Winter 2021/2022", "Winter 2022/2023"))) %>%
+  dplyr::select(-n_tot,-n_neg) %>%
+  reshape2::melt(c("year_num", "year", "type"))
 
-pos_samples
-```
-
-## Particle reduction
-
-```{r}
-vars <- c("CO2", "rH", "Temp", "AER", "Cn", "PM1", "PM25", "PM4", "PM10")
-vars_labels_no_units <- c("CO[2]", "Relative~humidity", "Temperature", "Air~exchange~rate",
-                          "CN", "PM[1]", "PM[2.5]", "PM[4]", "PM[10]")
-names(vars_labels_no_units) <- vars
-
-particle <- data.frame(
-  year = factor(c(rep(2022, 15), rep(2023, 15)), levels = c(2022, 2023)),
-  variable = factor(c(rep(c("Cn", "PM1", "PM25", "PM4", "PM10"), each = 3, times = 2)), levels = c("Cn", "PM1", "PM25", "PM4", "PM10")),
-  interval = rep(c("Mean", "Lower", "Upper"), times = 5),
-  estimate = c(c(39.24, 3.88, 68.80, 
-                 41.05, 6.62, 67.98,
-                 45.12, 19.66, 63.16,
-                 46.49, 27.26, 61.72,
-                 46.60, 29.99, 59.49),
-               c(76.83, 63.20, 86.28,
-                 76.79, 64.84, 85.46,
-                 72.86, 64.24, 79.66,
-                 70.77, 63.46, 76.97,
-                 99.99, 99.94, 100.00))
-) %>%
-  dcast(year + variable ~ interval) %>%
-  ggplot(aes(y = year, color = year)) +
-  facet_wrap(~ variable, nrow = 1, dir = "h", labeller = as_labeller(vars_labels_no_units, label_parsed)) +
-  geom_errorbar(mapping = aes(xmin = Lower, xmax = Upper), lineend = "square", size = 1, width = .05) +
-  geom_point(mapping = aes(x = Mean), shape = 23, size = 2, fill = "white") +
-  scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) +
-  scale_x_continuous(expand = expansion(mult = c(0, 0.1)), breaks = c(0, 50, 100), limits = c(0, 100)) +
-  labs(y = "Study year", x = "Estimated reduction (%)", title = "c") + 
+pos_samples_pl <- pos_samples %>%
+  ggplot(aes(y = value, x = year_num, fill = year)) +
+  geom_bar(stat = "identity") +
+  geom_text(aes(label = value, color = year), size = 8 / cm(1), vjust = -1) +
+  facet_wrap(~ type, nrow = 1) +
+  scale_y_continuous(expand = expansion(add = c(0,5))) +
+  scale_fill_manual(values = RColorBrewer::brewer.pal(11, "BrBG")[c(4,8)]) +
+  scale_color_manual(values = RColorBrewer::brewer.pal(11, "BrBG")[c(4,8)]) +
+  labs(y = "Frequency of positive samples", title = "b") +
   theme_custom() +
@@
         plot.title.position = "plot",
-        panel.spacing = unit(.5, "lines"))
+        strip.text = element_text(face = 2, vjust = 2, margin = margin(t = 10, b = 10)),
+        panel.spacing.x = unit(1, "cm"),
+        axis.title.x = element_blank(),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank())
 
-particle
+pos_samples_pl
 ```
 
-## Risk ratio
-
 ```{r}
-rr <- data.frame(
-  year = factor(c(rep(2022, 3), rep(2023, 3)), levels = c(2022, 2023)),
-  interval = rep(c("Mean", "Lower", "Upper"), times = 2),
-  estimate = c(c(1.00, 0.22, 2.067), c(0.72, 0.44, 1.15))
-) %>%
-  dcast(year ~ interval) %>%
-  ggplot(aes(y = year, color = year)) +
-  geom_vline(aes(xintercept = 1), linetype = "dashed") +
-  geom_errorbar(mapping = aes(xmin = Lower, xmax = Upper), lineend = "square", size = 1, width = .05) +
-  geom_point(mapping = aes(x = Mean), shape = 23, size = 2, fill = "white") +
-  labs(y = "Study year", x = "Adjusted risk ratio (%)", title = "d") + 
-  scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) +
-  theme_custom() +
-  theme(legend.position = "none",
-        plot.title.position = "plot")
-
-rr  
+comparison <- grid.arrange(saliva_pl, pos_samples_pl, nrow = 2, heights = c(6, 8), widths = 11)
+ggsave(plot = comparison, file = "../results/comparison.pdf", width = 11.5 / cm(1), height = 14 / cm(1))
 ```
 
-## Combined
+<!-- ## Particle reduction -->
 
-```{r}
-comparison <- grid.arrange(saliva, pos_samples, particle, rr, ncol = 2, nrow = 2, heights = c(6, 4), widths = c(10,6))
-ggsave(plot = comparison, file = "../results/comparison.pdf", width = 16 / cm(1), height = 10 / cm(1))
-```
+<!-- ```{r} -->
+<!-- vars <- c("CO2", "rH", "Temp", "AER", "Cn", "PM1", "PM25", "PM4", "PM10") -->
+<!-- vars_labels_no_units <- c("CO[2]", "Relative~humidity", "Temperature", "Air~exchange~rate", -->
+<!--                           "CN", "PM[1]", "PM[2.5]", "PM[4]", "PM[10]") -->
+<!-- names(vars_labels_no_units) <- vars -->
+
+<!-- particle <- data.frame( -->
+<!--   year = factor(c(rep(2022, 15), rep(2023, 15)), levels = c(2022, 2023)), -->
+<!--   variable = factor(c(rep(c("Cn", "PM1", "PM25", "PM4", "PM10"), each = 3, times = 2)), levels = c("Cn", "PM1", "PM25", "PM4", "PM10")), -->
+<!--   interval = rep(c("Mean", "Lower", "Upper"), times = 5), -->
+<!--   estimate = c(c(39.24, 3.88, 68.80,  -->
+<!--                  41.05, 6.62, 67.98, -->
+<!--                  45.12, 19.66, 63.16, -->
+<!--                  46.49, 27.26, 61.72, -->
+<!--                  46.60, 29.99, 59.49), -->
+<!--                c(76.83, 63.20, 86.28, -->
+<!--                  76.79, 64.84, 85.46, -->
+<!--                  72.86, 64.24, 79.66, -->
+<!--                  70.77, 63.46, 76.97, -->
+<!--                  99.99, 99.94, 100.00)) -->
+<!-- ) %>% -->
+<!--   dcast(year + variable ~ interval) %>% -->
+<!--   ggplot(aes(y = year, color = year)) + -->
+<!--   facet_wrap(~ variable, nrow = 1, dir = "h", labeller = as_labeller(vars_labels_no_units, label_parsed)) + -->
+<!--   geom_errorbar(mapping = aes(xmin = Lower, xmax = Upper), lineend = "square", size = 1, width = .05) + -->
+<!--   geom_point(mapping = aes(x = Mean), shape = 23, size = 2, fill = "white") + -->
+<!--   scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) + -->
+<!--   scale_x_continuous(expand = expansion(mult = c(0, 0.1)), breaks = c(0, 50, 100), limits = c(0, 100)) + -->
+<!--   labs(y = "Study year", x = "Estimated reduction in particle concentration (%)", title = "c") +  -->
+<!--   theme_custom() + -->
+<!--   theme(legend.position = "none", -->
+<!--         plot.title.position = "plot", -->
+<!--         panel.spacing = unit(.5, "lines")) -->
+
+<!-- particle -->
+<!-- ``` -->
+
+<!-- ## Risk ratio -->
+
+<!-- ```{r} -->
+<!-- rr <- data.frame( -->
+<!--   year = factor(c(rep(2022, 3), rep(2023, 3)), levels = c(2022, 2023)), -->
+<!--   interval = rep(c("Mean", "Lower", "Upper"), times = 2), -->
+<!--   estimate = c(c(1.00, 0.22, 2.067), c(0.72, 0.44, 1.15)) -->
+<!-- ) %>% -->
+<!--   dcast(year ~ interval) %>% -->
+<!--   ggplot(aes(y = year, color = year)) + -->
+<!--   geom_vline(aes(xintercept = 1), linetype = "dashed") + -->
+<!--   geom_errorbar(mapping = aes(xmin = Lower, xmax = Upper), lineend = "square", size = 1, width = .05) + -->
+<!--   geom_point(mapping = aes(x = Mean), shape = 23, size = 2, fill = "white") + -->
+<!--   labs(y = "Study year", x = "ARR of infection with air cleaners (%)", title = "d") +  -->
+<!--   scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) + -->
+<!--   theme_custom() + -->
+<!--   theme(legend.position = "none", -->
+<!--         plot.title.position = "plot") -->
+
+<!-- rr   -->
+<!-- ``` -->
+
+<!-- ## Combined  -->
+
+<!-- ```{r} -->
+<!-- comparison <- grid.arrange(saliva, pos_samples, particle, rr, ncol = 2, nrow = 2, heights = c(6, 4), widths = c(10,6)) -->
+<!-- ggsave(plot = comparison, file = "../results/comparison.pdf", width = 16 / cm(1), height = 10 / cm(1)) -->
+<!-- ``` -->

### 2023-12-04 doc/comparison-mcid12/manuscript.tex

@@
 \begin{information}\normalfont
-\textbf{Word count:} Abstract 252, Manuscript 3,244 \\
+\textbf{Word count:} Abstract 126 (max 150), Manuscript xx (max 1,500) \\
 \textbf{Display items:} Figure 1 \\
-\textbf{References:} 37 \\
-\textbf{Supporting information:} STROBE checklist, Supplementary Data with texts, tables, and figures
+\textbf{References:} xx (max 12) \\
+%\textbf{Supporting information:} STROBE checklist, Supplementary Data with texts, tables, and figures
 % \par
@@
 \begin{abstract}\normalfont
-\noindent We compared air, air filters, and saliva samples from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave) and winter 2022/2023 (after the COVID-19 pandemic) in a Swiss school setting.
+\noindent We compare air, air filters, and saliva samples from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave) and winter 2022/2023 (after the COVID-19 pandemic) in a Swiss school setting.
 Whereas during the pandemic we detected almost exclusively SARS-CoV-2 in human saliva (19~samples of SARS-CoV-2, 2~samples of non-SARS-CoV-2), the most common respiratory viruses after the pandemic were adenovirus, influenza~B, and rhinoviruses (3~samples of SARS-CoV-2, 47~samples of non-SARS-CoV-2).
@@
 
-%\section*{Introduction}
+Respiratory viruses, such as COVID-19 and influenza, pose significant challenges in terms of transmission within schools and and other indoor settings\cite{Leung2020}.
+These viruses spread through various routes, including respiratory particles like large droplets and small aerosols.
+Unlike larger droplets that settle quickly, aerosols can remain suspended in the air for extended periods\cite{Wang2020,Morawska2020}, contributing to the overall transmission dynamics.
+Detecting respiratory viruses in the air is challenging\cite{Wang2021,Belser2023PLOSPath}, adding to the complexity of understanding their spread.
+
+Infectious pathogens are primarily found in smaller particles (<5$\mu$m) and the distribution is similar across a range of different pathogens\cite{Fennelly2020}.
+Thus, infectious aerosols have the potential for long-range transmission, but the larger concentration of particles near the infectious person favours short-range transmission\cite{Tellier2009JTRSI,Wang2020,Morawska2020}.
+However, previous work also suggests that prolonged close contact may facilitate transmission\cite{Leung2020NatMed,Brankston2007LancetID,Lind2023NatCommun}.
+The complexity of respiratory virus transmission lies in the interplay between the physicochemical properties of the virus-laden aerosols and environmental conditions\cite{Wang2020,Morawska2020}, making it challenging to devise effective public health strategies for mitigating the spread of respiratory infections in various indoor settings, including schools.
+
+In this study, we compare samples from human saliva, the air, and the filters of air cleaners from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave)\cite{Banholzer2023PLoSMed} and winter 2022/2023 (after the COVID-19 pandemic)\cite{Banholzer2023submitted} in a Swiss school setting.
+We show differences in the distribution of respiratory viruses and emphasise the varying ability to detect these viruses in the air.
 
@@
 
-\bibliography{../paper/references.bib}
+\bibliography{references.bib}
 

### 2023-12-04 analysis/study_comaprison.Rmd
minor color changes to figure

@@
     year_num = c(rep(2022, 6), rep(2023, 6)),
-    virus = rep(c("Flu", "HRV", "AdV", "CoV", "MPV", "PIV"), 2),
+    virus = rep(c("IF", "HRV", "AdV", "CoV", "MPV", "PIV"), 2),
     n = c(c(1, NA, 1, 19, NA, NA), c(15, 15, 14, 3, 2, 1))
@@
          year = factor(year, levels = c("Winter 2021/2022", "Winter 2022/2023")),
-         virus = factor(virus, levels = c("Flu", "HRV", "AdV", "CoV", "MPV", "PIV"))) %>%
+         virus = factor(virus, levels = c("IF", "HRV", "AdV", "CoV", "MPV", "PIV"))) %>%
   filter(!is.na(n)) %>%
@@
   mutate(labelPosition = ifelse(year == "Winter 2022/2023" & virus == "MPV", labelPosition - .015, labelPosition),
-         labelPosition = ifelse(year == "Winter 2021/2022" & virus == "Flu", ymin, labelPosition),
+         labelPosition = ifelse(year == "Winter 2021/2022" & virus == "IF", ymin, labelPosition),
          labelPosition = ifelse(year == "Winter 2021/2022" & virus == "AdV", ymax, labelPosition))
@@
   scale_y_continuous(expand = c(0,0),  position = "top") +
-  scale_fill_manual(values = RColorBrewer::brewer.pal(9, "BuPu")[4:9]) +
-  scale_color_manual(values = RColorBrewer::brewer.pal(9, "BuPu")[4:9]) +
+  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
+  scale_color_manual(values = wes_palette("IsleofDogs1")) +
   ggtitle("a") +
@@
   scale_y_continuous(expand = expansion(add = c(0,5))) +
-  scale_fill_manual(values = RColorBrewer::brewer.pal(11, "BrBG")[c(4,8)]) +
-  scale_color_manual(values = RColorBrewer::brewer.pal(11, "BrBG")[c(4,8)]) +
+  scale_fill_manual(values = RColorBrewer::brewer.pal(9, "PuBu")[c(4,9)]) +
+  scale_color_manual(values = RColorBrewer::brewer.pal(9, "PuBu")[c(4,9)]) +
   labs(y = "Frequency of positive samples", title = "b") +

### 2023-12-04 doc/comparison-mcid12/manuscript.tex
Updates from Overleaf

@@
 Whereas during the pandemic we detected almost exclusively SARS-CoV-2 in human saliva (19~samples of SARS-CoV-2, 2~samples of non-SARS-CoV-2), the most common respiratory viruses after the pandemic were adenovirus, influenza~B, and rhinoviruses (3~samples of SARS-CoV-2, 47~samples of non-SARS-CoV-2).
-Despite the same study setting and environmental conditions, airborne detection of SARS-CoV-2 was relatively more frequent than the detection of non-SARS-CoV-2 viruses (10~samples of SARS-CoV-2 in 2021/2022 vs 2~samples of non-SARS-CoV-2 in 2022/2023).
+Despite the same study setting and similar environmental conditions, airborne detection of SARS-CoV-2 was relatively more frequent than the detection of non-SARS-CoV-2 viruses (10~samples of SARS-CoV-2 in 2021/2022 vs 2~samples of non-SARS-CoV-2 in 2022/2023).
 Our comparison suggests that respiratory viruses other than SARS-CoV-2 may be less detectable in the air and that close contact is required for transmission.
@@
 
-In this study, we compare samples from human saliva, the air, and the filters of air cleaners from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave)\cite{Banholzer2023PLoSMed} and winter 2022/2023 (after the COVID-19 pandemic)\cite{Banholzer2023submitted} in a Swiss school setting.
+In this study, we compare samples from human saliva, the air, and the filters of air cleaners from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave, 2022)\cite{Banholzer2023PLoSMed} and winter 2022/2023 (after the COVID-19 pandemic, 2023)\cite{Banholzer2023submitted} in a Swiss school setting.
 We show differences in the distribution of respiratory viruses and emphasise the varying ability to detect these viruses in the air.
@@
 
-Data was collected in two classes of two secondary schools in winter 2021/2022 and two classrooms of one secondary school in winter 2022/2023 (age of students 14-17~years) in the canton of Solothurn, Switzerland, both over a seven-week study period from end of January to beginning of March.
-In winter 2021/2022, a general mask mandate was in place for the first 3~study weeks, followed by 2~weeks without intervention, and another 2~weeks of air cleaners installed in two classrooms.
-In winter 2022/2023, air cleaners were installed for half of the study period following a cross-over design.
+Data was collected in two classes of two secondary schools in 2022 and two classrooms of one secondary school in winter 2023 (age of students 14-17~years) in the canton of Solothurn, Switzerland, both over a seven-week study period from end of January to beginning of March.
+In 2022, a general mask mandate was in place for the first 3~study weeks, followed by 2~weeks without intervention, and another 2~weeks of air cleaners installed in two classrooms.
+In 2023, air cleaners were installed for half of the study period following a cross-over design.
 An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, temperature, and humidity in all classrooms.
@@
 The removable parts were regularly autoclaved.
-Finally, swabs from the air cleaners' HEPA filters were collected at the end of the study (winter 2021/2022) and after each intervention phase (winter 2022/2023).
+Finally, swabs from the air cleaners' HEPA filters were collected at the end of the study in 2022 and after each intervention phase in 2023.
 The HEPA filters were removed and divided into 20~fields.
@@
 
-Descriptive statistics were utilized to depict differences in the type, number, and proportion of respiratory viruses between winter 2021/2022 and winter 2022/2023.
+Descriptive statistics were utilized to depict differences in the type, number, and proportion of respiratory viruses between 2022 and 2023.
 
@@
 
+In 2022, 44~students from two secondary school classes participated in weekly saliva testing (24~students from school~1 and 20~students from school~2).
+There were 21~positive saliva samples over the study (13~samples in school~1 and 8~samples in school 2): 19~SARS-CoV-2, 1~influenza~A, and 1~adenovirus (\Cref{fig:comparison}a, left).
+There were 10~positive bioaerosol samples (4~samples in school~1 and 6~samples in school~2): 9~SARS-CoV-2 and 1~adenovirus.
+There were 8~positive samples on the HEPA-filters of the air cleaners (2~samples in school~1 and 6~samples in school~2): 6~SARS-CoV-2, 1~influenza~A and 1~adenovirus.
+
+In 2023, 38~students from two classes of secondary school~2 participated in bi-weekly saliva testing.
+There were 50~positive saliva samples over the study: 15~influenza~B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, 1~parainfluenza virus (\Cref{fig:comparison}a, right).
+There were 2~positive bioaerosol samples: 1~rhinovirus and 1~adenovirus.
+There were 4~positive samples on the HEPA-filters of the air cleaners: 1~influenza~B, 1~rhinovirus, 1~adenovirus, and 1~SARS-CoV-2.
+
+In comparison, there was a clear shift in the distribution of respiratory viruses from mainly SARS-CoV-2 in 2022 to mainly non-SARS-CoV-2 viruses in 2023 (\Cref{fig:comparison}a).
+In saliva, 21/262~samples were positive in 2022 compared to 50/448~samples in 2023 (\Cref{fig:comparison}b, left).
+In bioaerosols, 10/130~samples were positive in winter 2022 compared to 2/105~samples in 2023  (\Cref{fig:comparison}b, middle).
+On the HEPA-filters of the air cleaners, 8/40~swabs were positive in 2022 compared to 4/160~swabs in 2023 (\Cref{fig:comparison}b, right).
+
+\begin{figure}
+    \centering
+    \includegraphics{results/comparison.pdf}
+    \caption{Comparison of respiratory viruses in saliva, air, and on the air filters between winter 2021/2022 (2022) and winter 2022/2023 (2023).
+\textbf{(a)}~Distribution of positive saliva samples.
+IF: influenza~A/B, HRV: human rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: human metapneumovirus, PIV: parainfluenza virus.
+\textbf{(b)}~Number of positive saliva, bioaerosol, and HEPA-filter samples.}
+    \label{fig:comparison}
+\end{figure}
+
+% Compare nvironmental conditions
+
 \section*{Discussion}
 
-Furthermore, the distribution of respiratory viruses in this study was very different from that in our previous study.
+% Furthermore, the distribution of respiratory viruses in this study was very different from that in our previous study.
 
-% why could we not detect so many bioaerosols
+% % why could we not detect so many bioaerosols
 
-We detected fewer respiratory viruses in bioaerosol samples compared to our previous study\cite{Banholzer2023PLoSMed} (1~sample of adenovirus and 1~sample of rhinovirus in this study vs 10~samples of SARS-CoV-2 in the previous study), especially in relation to the number of positive saliva samples (50~samples in this study vs 21~in the previous study).
+% We detected fewer respiratory viruses in bioaerosol samples compared to our previous study\cite{Banholzer2023PLoSMed} (1~sample of adenovirus and 1~sample of rhinovirus in this study vs 10~samples of SARS-CoV-2 in the previous study), especially in relation to the number of positive saliva samples (50~samples in this study vs 21~in the previous study).
 Similarly, the number of viruses detected on the filters of the air cleaners was also lower in this study (4~positive samples in 160~swabs vs 8~positive samples in 80~swabs in the previous study).
@@
 
-% other reasons
+% % other reasons
 
-Technical reasons for the differences in airborne detection are unlikely.
+% Technical reasons for the differences in airborne detection are unlikely.
 The bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
@@
 
-% bioaerosol sampling in more depth
+% % bioaerosol sampling in more depth
 
-We speculate that the difference in airborne molecular detection of respiratory viruses between 2022 (pandemic) and 2023 (non-pandemic) is due to the different characteristics of the respiratory viruses.
+% We speculate that the difference in airborne molecular detection of respiratory viruses between 2022 (pandemic) and 2023 (non-pandemic) is due to the different characteristics of the respiratory viruses.
 Differences in the physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity), which influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}, may facilitate the detection of SARS-CoV-2 in the air.
@@
 
-\section{Comparison with previous study}\label{sec:comparison-prev-study}
+% \section{Comparison with previous study}\label{sec:comparison-prev-study}
 
-% descriptive results
+% % descriptive results
 
-In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 Omicron wave.
+% In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 Omicron wave.
 \cref{fig:comparison} shows a comparison of the results.
@@
 Although there were more positive saliva samples in this study, the number of respiratory infections detected in bioaerosols and on the HEPA-filters of the air cleaners was lower (\cref{fig:comparison}b).
-The reductions in particle concentrations were larger and with lower uncertainty in this study (\cref{fig:comparison}c).
+% The reductions in particle concentrations were larger and with lower uncertainty in this study (\cref{fig:comparison}c).
 The risk of infection tended to be lower with air cleaners in this study where air cleaners were analyzed using a cross-over design, but not in our previous study where air cleaners were installed only at the end of the study when most students had already been infected with SARS-Cov-2(\cref{fig:comparison}d).
 
-% summary
+% % summary
 
-To summarize, we found similar reductions in particle concentrations and a tendency for the relative risk of infection to be lower with air cleaners.
+% To summarize, we found similar reductions in particle concentrations and a tendency for the relative risk of infection to be lower with air cleaners.
 To some extent, the differences between the two studies may be explained by different adjustments.
@@
 
-\begin{figure}[!htpb]
-    \includegraphics[width=\linewidth]{../../results/comparison.pdf} 
-    \caption{\textbf{Comparison of the results between our previous study in 2022 during the COVID-19 pandemic (blue) and this study in 2023 during non-pandemic times (black).} \textbf{(a)}~Proportion of positive saliva samples by virus.
+% \begin{figure}[!htpb]
+%     \includegraphics[width=\linewidth]{../../results/comparison.pdf} 
+%     \caption{\textbf{Comparison of the results between our previous study in 2022 during the COVID-19 pandemic (blue) and this study in 2023 during non-pandemic times (black).} \textbf{(a)}~Proportion of positive saliva samples by virus.
 Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.
@@
 \textbf{(d)}~Adjusted relative risk (ARR) of infection with air cleaners (posterior mean as squares and 95\%-CrI as lines).}
-    \label{fig:comparison}
-\end{figure}
+%     \label{fig:comparison}
+% \end{figure}
 
-% unlikely reasons for less airborne detection
+% % unlikely reasons for less airborne detection
 
-Technical reasons for the differences in airborne detection are unlikely.
+% Technical reasons for the differences in airborne detection are unlikely.
 The bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
@@
 
-% likely reason for less airborne detection
+% % likely reason for less airborne detection
 
-More likely, SARS-CoV-2 may be easier to detect in the air due to physicochemical properties of virus-laden aerosols, so that its viral loads more often exceed the detection limits of existing devices for bioaerosol sampling,  which remains a challenge\cite{Wang2021,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
+% More likely, SARS-CoV-2 may be easier to detect in the air due to physicochemical properties of virus-laden aerosols, so that its viral loads more often exceed the detection limits of existing devices for bioaerosol sampling,  which remains a challenge\cite{Wang2021,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
 For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.

### 2023-12-18 analysis/env_analysis.Rmd
revised study comparison

@@
 env_epi_data %>%
-  summarize(across(c(CO2, AER, Cn, PM1), .fns = list(mean = mean, sd = sd), .names = "{.col}_{.fn}")) %>%
-  mutate(across(c(CO2_mean, CO2_sd, Cn_mean, Cn_sd), round),
-         across(c(AER_mean, AER_sd, PM1_mean, PM1_sd), round, 2))
+  summarize(across(c(CO2, AER, Cn, PM1, rH, Temp), .fns = list(mean = mean, sd = sd), .names = "{.col}_{.fn}")) %>%
+  mutate(across(c(CO2_mean, CO2_sd, Cn_mean, Cn_sd, rH_mean, Temp_mean), round),
+         across(c(AER_mean, AER_sd, PM1_mean, PM1_sd, rH_sd, Temp_sd), round))
 

### 2023-12-18 analysis/study_comaprison.Rmd
revised study comparison

@@
 library(reshape2)
+
+library(rstan)
+library(brms)
+library(tidybayes)
+library(modelr)
+
 source("../utils/epi.r")
@@
 
+## Data
+
+```{r}
+# saliva
+saliva_2022 <- readxl::read_xlsx("../data-clean/comparison/mcid-1-molecular-results.xlsx", "Saliva")
+saliva_2023 <- readxl::read_xlsx("../data-clean/comparison/mcid-2-molecular-results.xlsx", "Saliva")
+
+saliva <- rbind(saliva_2022 %>% mutate(year_num = 2022),
+                saliva_2023 %>% mutate(year_num = 2023))
+
+# bioaerosols
+air_2022 <- readxl::read_xlsx("../data-clean/comparison/mcid-1-molecular-results.xlsx", "Bioaerosols")
+air_2023 <- readxl::read_xlsx("../data-clean/comparison/mcid-2-molecular-results.xlsx", "Bioaerosols")
+
+air <- rbind(air_2022 %>% mutate(year_num = 2022),
+             air_2023 %>% mutate(year_num = 2023))
+
+# air filters
+# TODO: just list
+
+# other data
+odf <- readRDS("../data-clean/comparison/merged-data-both-years-all-classes.rds")
+```
+
 ## Virus distribution
@@
 ```{r}
-saliva <- data.frame(
-    year_num = c(rep(2022, 6), rep(2023, 6)),
-    virus = rep(c("IF", "HRV", "AdV", "CoV", "MPV", "PIV"), 2),
-    n = c(c(1, NA, 1, 19, NA, NA), c(15, 15, 14, 3, 2, 1))
-  ) %>%
+saliva_sum <- saliva %>%
+  mutate(IF = IFA + IFB) %>%
+  dplyr::select(year_num, IF, HRV, AdV, CoV, MPV, PIV) %>%
+  melt("year_num") %>%
+  rename(virus = variable,
+         n = value) %>%
+  group_by(year_num, virus) %>%
+  summarize(n = sum(n)) %>%
+  ungroup() %>%
+  filter(n > 0) %>%
   mutate(year = ifelse(year_num == 2022, "Winter 2021/2022", "Winter 2022/2023"),
@@
 
-saliva_pl <- saliva %>%
+saliva_pl <- saliva_sum %>%
   ggplot() +
@@
 
-## No. of positive samples
+## Time distribution
 
 ```{r}
-pos_samples <- data.frame(
-  year_num = factor(c(rep(2022, 3), rep(2023, 3)), levels = c(2022, 2023)),
-  type = factor(rep(c("Saliva", "Bioaerosols", "HEPA-filter"), times = 2), levels = c("Saliva", "Bioaerosols", "HEPA-filter")),
-  n_pos = c(c(21, 10, 8), c(50, 2, 4)),
-  n_tot = c(c(262, 130, 40), c(448, 105, 160))
+saliva_week <- saliva %>%
+  mutate(IF = IFA + IFB,
+         class = ifelse(year_num == 2022, ifelse(class == "D", "B", "A"), class)) %>%
+  dplyr::select(year_num, week, class, intervention, IF, HRV, AdV, CoV, MPV, PIV) %>%
+  melt(c("year_num", "week", "class", "intervention")) %>%
+  rename(virus = variable) %>%
+  group_by(year_num, class, week, intervention, virus) %>%
+  summarize(n = sum(value, na.rm = T)) %>%
+  ungroup()
+
+air_week <- air %>%
+  mutate(IF = IFA + IFB,
+         class = ifelse(year_num == 2022, ifelse(class == "D", "B", "A"), class)) %>%
+  dplyr::select(year_num, week, class, intervention, IF, HRV, AdV, CoV, MPV, PIV) %>%
+  melt(c("year_num", "week", "class", "intervention")) %>%
+  rename(virus = variable) %>%
+  group_by(year_num, class, week, intervention, virus) %>%
+  summarize(n = sum(value, na.rm = T)) %>%
+  ungroup()
+
+saliva_air_week <- rbind(saliva_week %>% mutate(type = "Saliva"),
+                         air_week %>% mutate(type = "Bioaerosol")) %>%
+  slice(rep(1:n(), times = n)) %>%
+  mutate(year = ifelse(year_num == 2022, "Winter 2021/2022", "Winter 2022/2023"),
+         year = factor(year, levels = c("Winter 2021/2022", "Winter 2022/2023")),
+         virus = factor(virus, levels = c("IF", "HRV", "AdV", "CoV", "MPV", "PIV")),
+         type = factor(type, levels = c("Bioaerosol", "Saliva")),
+         type_num = ifelse(type == "Bioaerosol", 1, 2),
+         classroom = factor(paste("Room", class), levels = c("Room A", "Room B")))
+
+backgrd_y <- saliva_air_week %>%
+  group_by(year, classroom, week) %>%
+  summarize(type_num = 1) %>%
+  ungroup() %>%
+  complete(year, classroom, week)
+
+time_pl <- ggplot(mapping = aes(x = week, y = type_num)) +
+  facet_grid(classroom ~ year) +
+  geom_rect(data = backgrd_y,
+            mapping = aes(xmin = week - 0.5, xmax = week + 0.5, ymin = -Inf, ymax = Inf),
+            fill = ifelse(backgrd_y$week %% 2 == 0, "lightgray", "white"),
+            alpha = 0.5) +
+  geom_hline(data = data.frame(y = 1.5), mapping = aes(yintercept = y), linetype = "dashed") +
+  geom_point(data = saliva_air_week %>%
+               mutate(ps = .17) %>%
+               group_by(year, classroom, week, virus, type) %>%
+               mutate(type_num = type_num + cumsum(ps) - .3) %>%
+               ungroup(),
+             mapping = aes(color = virus),
+             position = position_dodge(width = .75)) +
+  scale_color_manual(values = wes_palette("IsleofDogs1")) +
+  scale_x_continuous(breaks = 1:7, limits = c(0.5, 7.5), expand = c(0,0)) +
+  scale_y_continuous(expand = c(0,0), limits = c(0, 3)) +
+  annotate("text", y = 2.8, x = 0.65, label = "Saliva", hjust = 0, size = 8 / cm(1)) +
+  annotate("text", y = 0.2, x = 0.65, label = "Bioaerosol", hjust = 0, size = 8 / cm(1)) +
+  labs(x = "Study week", title = "b") +
+  theme_bw2() +
+  theme(axis.title.y = element_blank(),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        strip.text = element_text(face = 2, size = 8),
+        text = element_text(size = 8),
+        axis.text.y = element_blank(),
+        axis.ticks.y = element_blank()) +
+  guides(shape = "none",
+         color = "none")
+
+time_pl
+```
+
+## Modeling saliva-air pairs
+
+### Checks
+
+Look at the difference in days between positive air and saliva samples. 
+* Are air samples detected later or earlier than a positive saliva samples?
+
+```{r}
+air_pos <- air %>%
+  mutate(IF = IFA + IFB) %>%
+  dplyr::select(-IFA, -IFB) %>%
+  melt(c("year_num", "school", "class", "week", "date", "sampler", "intervention")) %>%
+  filter(!is.na(value), value > 0) %>%
+  mutate(class = ifelse(year_num == 2022, ifelse(class == "A/B", "A", "B"), class)) %>%
+  dplyr::select(-school, -value, -intervention) %>%
+  rename(date_air = date) %>%
+  mutate(sample_id = 1:n())
+
+sal_pos <- saliva %>%
+  mutate(IF = IFA + IFB) %>%
+  dplyr::select(-IFA, -IFB) %>%
+  melt(c("year_num", "school", "class", "week", "date", "intervention")) %>%
+  filter(!is.na(value), value > 0) %>%
+  mutate(class = ifelse(year_num == 2022, ifelse(school == 1, "A", "B"), class)) %>%
+  dplyr::select(-school, -value, -intervention) %>%
+  rename(date_sal = date)
+
+air_sal_pos <- air_pos %>% 
+  dplyr::select(-week) %>%
+  left_join(sal_pos %>% dplyr::select(-week)) %>%
+  mutate(tdays = difftime(date_air, date_sal, units = "days")) %>%
+  group_by(sample_id) %>%
+  arrange(abs(tdays)) %>%
+  slice(1) %>%
+  ungroup()
+
+hist(air_sal_pos$tdays %>% as.numeric, bre = 30)
+```
+
+Consider the issue of the missing VIVAS sampler in class B in 2023.
+* How many weeks were there when VIVAS detected something but not Coriolis?
+
+```{r}
+air_pos_weeks <- air %>%
+  mutate(IF = IFA + IFB) %>%
+  dplyr::select(-IFA, -IFB) %>%
+  melt(c("year_num", "school", "class", "week", "date", "sampler", "intervention")) %>%
+  rename(virus = variable) %>%
+  mutate(class = ifelse(year_num == 2022, ifelse(class == "A/B", "A", "B"), class)) %>%
+  group_by(year_num, class, week, sampler, virus) %>%
+  summarise(detected = ifelse(sum(value, na.rm = T) > 0, 1, 0)) %>%
+  ungroup() %>%
+  group_by(year_num, class, week, virus) %>%
+  filter(any(detected == 1)) %>%
+  ungroup() %>%
+  filter(!(year_num == 2023 & class == "B")) %>%
+  arrange(year_num, class, week, virus, sampler) %>%
+  dcast(year_num + class + week + virus ~ sampler)
+
+table(air_pos_weeks$Coriolis, air_pos_weeks$Vivas)
+
+3 / 7
+```
+
+### Data
+
+```{r}
+# saliva bioaerosol pairs weekly
+pair_df <- rbind(saliva_week %>% mutate(type = "Saliva"),
+                 air_week %>% mutate(type = "Bioaerosol")) %>%
+  group_by(year_num, class, virus, week, intervention, type) %>%
+  summarize(n = sum(n)) %>%
+  ungroup() %>%
+  dcast(year_num + class + virus + week + intervention ~ type) %>%
+  mutate(is_cov = ifelse(virus == "CoV", "yes", "no"),
+         across(c(Bioaerosol, Saliva), ~ ifelse(.x > 0, "yes", "no")),
+         across(c(Bioaerosol, Saliva, is_cov), ~ factor(.x, levels = c("no", "yes"))),
+         year = ifelse(year_num == 2022, "Winter 2021/2022", "Winter 2022/2023"),
+         year = factor(year, levels = c("Winter 2021/2022", "Winter 2022/2023")),
+         virus = factor(virus, levels = c("IF", "HRV", "AdV", "CoV", "MPV", "PIV")),
+         intervention = factor(intervention, levels = c("none", "masks", "air cleaners")))
+
+# add ventilation
+vent <- left_join(
+  air %>% 
+    mutate(date = as.Date(date),
+           class = ifelse(class == "A/B", "A", ifelse(class == "D", "B", class))) %>%
+    group_by(year_num, class, week, date) %>%
+    slice(1) %>%
+    ungroup() %>%
+    dplyr::select(year_num, class, week, date),
+  odf %>%
+    group_by(school, classroom, date) %>%
+    slice(1) %>%
+    ungroup() %>%
+    mutate(year_num = lubridate::year(date),
+           class = ifelse(classroom == "AB", "A",
+                          ifelse(year_num == 2022 & school == "School 2" & classroom == "A", "B", classroom))) %>%
+    dplyr::select(year_num, class, date, co2.max) 
 ) %>%
-  mutate(n_neg = n_tot - n_pos,
-         year = ifelse(year_num == 2022, "Winter 2021/2022", "Winter 2022/2023"),
-         year = factor(year, levels = c("Winter 2021/2022", "Winter 2022/2023"))) %>%
-  dplyr::select(-n_tot,-n_neg) %>%
-  reshape2::melt(c("year_num", "year", "type"))
+  group_by(year_num, class, week) %>%
+  summarize(co2.max = mean(co2.max, na.rm = T)) %>%
+  ungroup() %>%
+  group_by(year_num) %>%
+  mutate(co2.max = ifelse(is.na(co2.max), mean(co2.max, na.rm = T), co2.max)) %>%
+  ungroup() %>%
+  rename(maxCO2 = co2.max)
 
-pos_samples_pl <- pos_samples %>%
-  ggplot(aes(y = value, x = year_num, fill = year)) +
-  geom_bar(stat = "identity") +
-  geom_text(aes(label = value, color = year), size = 8 / cm(1), vjust = -1) +
-  facet_wrap(~ type, nrow = 1) +
-  scale_y_continuous(expand = expansion(add = c(0,5))) +
-  scale_fill_manual(values = RColorBrewer::brewer.pal(9, "PuBu")[c(4,9)]) +
-  scale_color_manual(values = RColorBrewer::brewer.pal(9, "PuBu")[c(4,9)]) +
-  labs(y = "Frequency of positive samples", title = "b") +
-  theme_custom() +
-  theme(legend.position = "none",
-        plot.title.position = "plot",
-        strip.text = element_text(face = 2, vjust = 2, margin = margin(t = 10, b = 10)),
-        panel.spacing.x = unit(1, "cm"),
-        axis.title.x = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank())
+pair_df <- left_join(pair_df, vent)
 
-pos_samples_pl
+# create stan data file
+stanDF <- list(
+  N = nrow(pair_df),
+  K_virus = n_distinct(pair_df$virus),
+  bioaerosol = ifelse(pair_df$Bioaerosol == "yes", 1, 0),
+  saliva = ifelse(pair_df$Saliva == "yes", 1, 0),
+  cov = ifelse(pair_df$is_cov == "yes", 1, 0),
+  virus = as.integer(pair_df$virus),
+  masks = ifelse(pair_df$intervention == "masks", 1, 0),
+  air_cleaners = ifelse(pair_df$intervention == "air cleaners", 1, 0),
+  co2max = (pair_df$maxCO2 - mean(pair_df$maxCO2)) / sd(pair_df$maxCO2) 
+)
+```
+
+### Priors
+
+```{r}
+# add prior scales
+stanDF$s_saliva <- 1 / sd(stanDF$saliva)
+stanDF$s_cov <- 1 / sd(stanDF$cov)
+stanDF$s_masks <- .25 #/ sd(stanDF$masks)
+stanDF$s_air <- .25 #/ sd(stanDF$air_cleaners)
+stanDF$s_co2 <- .25
+stanDF$s_cov <- 1
+
+# add prior skew
+stanDF$t_masks <- .75
+stanDF$t_air <- .75
+stanDF$t_co2 <- .25
+
+# add offset
+stanDF$offsets = ifelse(pair_df$year_num == 2022, ifelse(pair_df$class == "A", log(2), log(1)), 
+                        ifelse(pair_df$class == "A", log(1), log(.5)))
+
+# informative priors
+#' positive effect of saliva based on conditional frequencies
+
+table(pair_df$Saliva, pair_df$Bioaerosol) 
+# 5/3 odds of positive bioaerosol for positive saliva
+stanDF$m_saliva <- sum(pair_df$Saliva[pair_df$Bioaerosol == "yes"] == "yes") / sum(pair_df$Saliva[pair_df$Bioaerosol == "yes"] == "no")
+
+#' 80% prior prob. for a negative effect of interventions
+#' smaller effects sizes expected for interventions
+
+stanDF$m_masks <- optim(par = 0, 
+                 fn = function(l, s, k, p) {
+                   q <- qasym_laplace(p, l, s, k)
+                   abs(q)
+                   }, 
+                 s = stanDF$s_masks, k = stanDF$t_masks, p = 0.9)$par[1] 
+stanDF$m_air <- optim(par = 0, 
+                 fn = function(l, s, k, p) {
+                   q <- qasym_laplace(p, l, s, k)
+                   abs(q)
+                   }, 
+                 s = stanDF$s_air, k = stanDF$t_air, p = 0.9)$par[1]
+
+#' same prior for co2 as for intervention effects, but positive
+
+stanDF$m_co2 <- optim(par = 0, 
+                 fn = function(l, s, k, p) {
+                   q <- qasym_laplace(p, l, s, k)
+                   abs(q)
+                   }, 
+                 s = stanDF$s_co2, k = stanDF$t_co2, p = 0.1)$par[1]
+
+#' uninformative prior for mean of cov
+stanDF$m_cov <- 1
 ```
@@
 ```{r}
-comparison <- grid.arrange(saliva_pl, pos_samples_pl, nrow = 2, heights = c(6, 8), widths = 11)
-ggsave(plot = comparison, file = "../results/comparison.pdf", width = 11.5 / cm(1), height = 14 / cm(1))
+a  <- seq(-5, 5, .1)
+plot(a, dasym_laplace(a, stanDF$m_masks, stanDF$s_masks, stanDF$t_masks))
 ```
+
+### Model
+
+```{r}
+# fit model
+mod <- stan(file = "../models/comparison/air-detection-model.stan",
+            data = stanDF,
+            cores = 4, chains = 4, iter = 2000,
+            seed = 12345)
+
+summary(mod, pars = c("alpha", 
+                      # "tau_virus", paste0("beta_virus[", 1:stanDF$K_virus, "]"), 
+                      "beta_saliva", 
+                      "beta_cov", 
+                      "beta_masks", 
+                      "beta_air_cleaners", "beta_co2max"))$summary %>%
+  data.frame() %>%
+  add_rownames() %>%
+  as_data_frame() %>%
+  mutate_if(is.numeric, round, 2)
+```
+
+### Results
+
+```{r}
+# adjusted OR of mask wearing
+mod %>%
+  spread_draws(beta_cov) %>%
+  mutate(beta_cov = exp(beta_cov)) %>%
+  median_qi()
+
+pred_prob <- mod %>%
+  spread_draws(logit_cov, logit_nonCov) %>%
+  mutate(CoV = invlogit(logit_cov),
+         `non-Cov` = invlogit(logit_nonCov)) %>%
+  dplyr::select(-logit_cov, -logit_nonCov) %>%
+  melt(c(".chain", ".iteration", ".draw"))
+
+prob_pl <- ggplot(mapping = aes(x = variable, y = value, color = variable)) +
+  geom_errorbar(data = pred_prob %>% group_by(variable) %>% median_qi(value),
+                mapping = aes(ymin = .lower, ymax = .upper, color = variable),
+                width = .25) +
+  geom_boxplot(data = pred_prob,
+               position = position_dodge(width = .5), outlier.shape = NA, coef = 0, width = 0.25) +
+  stat_summary(data = pred_prob %>% dplyr::select(value, variable),
+               mapping = aes(x = variable, y = value, color = variable), geom = "point", fun = "median", 
+               position = position_dodge2(width = .5), size = 2, shape = 23, fill = "white") +
+  scale_y_continuous(expand = c(0,0), limits = c(0, NA), labels = function(x) x * 100) +
+  scale_color_manual(values = c(wes_palette("IsleofDogs1")[4], "darkred")) +
+  #scale_fill_manual(values = c(wes_palette("IsleofDogs1")[4], "darkred")) +
+  labs(y = "Prob. of airborne detection (%)", title = "c") +
+  theme_custom() +
+  theme(legend.position = "none",
+        axis.title.x = element_blank())
+
+prob_pl
+
+# CoV vs non-CoV
+pred_prob %>%
+  group_by(variable) %>%
+  median_qi() %>%
+  mutate_if(is.numeric, function(x) round(x * 100) )
+```
+
+```{r}
+design <- "12
+33"
+comparison <- saliva_pl + prob_pl + time_pl + plot_layout(design = design, widths = c(11, 5), heights = c(6, 9))
+ggsave(plot = comparison, file = "../results/comparison.pdf", width = 16 / cm(1), height = 15 / cm(1))
+```
+
+<!-- ## No. of positive samples -->
+
+<!-- ```{r} -->
+<!-- pos_samples <- data.frame( -->
+<!--   year_num = factor(c(rep(2022, 3), rep(2023, 3)), levels = c(2022, 2023)), -->
+<!--   type = factor(rep(c("Saliva", "Bioaerosols", "HEPA-filter"), times = 2), levels = c("Saliva", "Bioaerosols", "HEPA-filter")), -->
+<!--   n_pos = c(c(21, 10, 8), c(50, 2, 4)), -->
+<!--   n_tot = c(c(262, 130, 40), c(448, 105, 160)) -->
+<!-- ) %>% -->
+<!--   mutate(n_neg = n_tot - n_pos, -->
+<!--          year = ifelse(year_num == 2022, "Winter 2021/2022", "Winter 2022/2023"), -->
+<!--          year = factor(year, levels = c("Winter 2021/2022", "Winter 2022/2023"))) %>% -->
+<!--   dplyr::select(-n_tot,-n_neg) %>% -->
+<!--   reshape2::melt(c("year_num", "year", "type")) -->
+
+<!-- pos_samples_pl <- pos_samples %>% -->
+<!--   ggplot(aes(y = value, x = year_num, fill = year)) + -->
+<!--   geom_bar(stat = "identity") + -->
+<!--   geom_text(aes(label = value, color = year), size = 8 / cm(1), vjust = -1) + -->
+<!--   facet_wrap(~ type, nrow = 1) + -->
+<!--   scale_y_continuous(expand = expansion(add = c(0,5))) + -->
+<!--   scale_fill_manual(values = RColorBrewer::brewer.pal(9, "PuBu")[c(4,9)]) + -->
+<!--   scale_color_manual(values = RColorBrewer::brewer.pal(9, "PuBu")[c(4,9)]) + -->
+<!--   labs(y = "Frequency of positive samples", title = "b") + -->
+<!--   theme_custom() + -->
+<!--   theme(legend.position = "none", -->
+<!--         plot.title.position = "plot", -->
+<!--         strip.text = element_text(face = 2, vjust = 2, margin = margin(t = 10, b = 10)), -->
+<!--         panel.spacing.x = unit(1, "cm"), -->
+<!--         axis.title.x = element_blank(), -->
+<!--         panel.grid.major = element_blank(), -->
+<!--         panel.grid.minor = element_blank()) -->
+
+<!-- pos_samples_pl -->
+<!-- ``` -->
+
+<!-- ```{r} -->
+<!-- comparison <- grid.arrange(saliva_pl, pos_samples_pl, nrow = 2, heights = c(6, 8), widths = 11) -->
+<!-- ggsave(plot = comparison, file = "../results/comparison.pdf", width = 11.5 / cm(1), height = 14 / cm(1)) -->
+<!-- ``` -->
+
+<!-- ## Absences ## -->
+
+<!-- ```{r} -->
+<!-- mcid_2 <- readRDS("../data-clean/redcap-long.rds") %>% -->
+<!--   filter(respiratory_infection == "yes") %>% -->
+<!--   mutate(absent_days = date_back - date_absent) -->
+
+<!-- # 35 absences related to respiratory infection among 38 students -->
+
+<!-- summary(as.numeric(mcid_2$absent_days)) -->
+
+<!-- # on average 3.0 days out of school per absence (but weekends should be kept in mind) -->
+
+<!-- mcid_1 <- read_csv("../data-clean/mcid-1-redcap-cases.csv") %>% -->
+<!--   filter(!is_teacher, -->
+<!--          school_class == "School 1 (A)" | school_class == "School 2 (D)") %>% -->
+<!--   mutate(absent_days = date_end - date_start) -->
+
+<!-- # 40 absences related to respiratory infection among 44 students -->
+
+<!-- summary(as.numeric(mcid_1$absent_days)) -->
+
+<!-- #' on average 4.1 days out of school per absence (but weekends should be kept in mind) -->
+<!-- #' there is also an outlier with 30 days, but the median is still higher  -->
+
+<!-- rbind(mcid_1 %>% mutate(year = 2022) %>% dplyr::select(year, absent_days), -->
+<!--       mcid_2 %>% mutate(year = 2023) %>% dplyr::select(year, absent_days)) %>% -->
+<!--   mutate(absent_days = as.numeric(absent_days)) %>% -->
+<!--   ggplot(aes(fill = factor(year), x = year, y = absent_days)) + -->
+<!--   geom_boxplot() + -->
+<!--   scale_y_sqrt(breaks = c(0, 1:7, 10, 20, 30), limits = c(1,30)) -->
+<!-- ``` -->
 

### 2023-12-18 doc/comparison-mcid12/manuscript.tex

--- 2023-12-04 doc/comparison-mcid12/manuscript.tex
@@
 
-\title{\bfseries\LARGE\singlespacing{Brief report: SARS-CoV-2 more frequently detected in indoor air than other respiratory viruses}}
+\title{\bfseries\LARGE\singlespacing{Brief report:
+Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols}}
 
@@
 \begin{information}\normalfont
-\textbf{Word count:} Abstract 252, Manuscript 3,244 \\
+\textbf{Word count:} Abstract 150 (max 150), Manuscript 1,625 (max 1,500) \\
 \textbf{Display items:} Figure 1 \\
-\textbf{References:} 37 \\
-\textbf{Supporting information:} STROBE checklist, Supplementary Data with texts, tables, and figures
+\textbf{References:} 17 (max 12) \\
+%\textbf{Supporting information:} STROBE checklist, Supplementary Data with texts, tables, and figures
 % \par
@@
 \begin{abstract}\normalfont
-\noindent We compared air, air filters, and saliva samples from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave) and winter 2022/2023 (after the COVID-19 pandemic) in a Swiss school setting.
-Whereas during the pandemic we detected almost exclusively SARS-CoV-2 in human saliva (19~samples of SARS-CoV-2, 2~samples of non-SARS-CoV-2), the most common respiratory viruses after the pandemic were adenovirus, influenza~B, and rhinoviruses (3~samples of SARS-CoV-2, 47~samples of non-SARS-CoV-2).
-Despite the same study setting and environmental conditions, airborne detection of SARS-CoV-2 was relatively more frequent than the detection of non-SARS-CoV-2 viruses (10~samples of SARS-CoV-2 in 2021/2022 vs 2~samples of non-SARS-CoV-2 in 2022/2023).
-Our comparison suggests that respiratory viruses other than SARS-CoV-2 may be less detectable in the air and that close contact is required for transmission.
+\noindent We compared airborne and saliva samples from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave) and winter 2022/2023 (after the COVID-19 pandemic) in a Swiss school setting.
+While during the pandemic we detected almost exclusively SARS-CoV-2 in saliva (19~samples of SARS-CoV-2 vs 2~samples of non-SARS-CoV-2), after the pandemic the most common respiratory viruses were adenovirus, influenza, and rhinoviruses (3~samples of SARS-CoV-2 vs 47~samples of non-SARS-CoV-2).
+Despite the same study setting and methods used, airborne detection of SARS-CoV-2 was relatively more frequent than of non-SARS-CoV-2 viruses (10~samples of SARS-CoV-2 in 2021/2022 vs 2~samples of non-SARS-CoV-2 in 2022/2023).
+SARS-CoV-2 was also detected more frequently on the HEPA-filters of installed portable air cleaners (6~samples of SARS-CoV-2 vs 3~samples of non-SARS-CoV-2).
+Our comparison suggests that SARS-CoV-2 is easier to detect in the air than other endemic respiratory viruses, possibly facilitating long-range transmission.
 \medskip
@@
 
-\noindent\textbf{Keywords:} respiratory viruses, airborne transmission, molecular detection
+\noindent\textbf{Keywords:} respiratory viruses; SARS-CoV-2; influenza; airborne transmission; molecular detection
 % maximum of 3-5 keywords
@@
 
-%\section*{Introduction}
+\section*{Introduction}
+
+The transmission of respiratory viruses, such as SARS-CoV-2 and influenza, within schools and and other indoor environments is difficult to control\cite{Leung2020}.
+Respiratory viruses spread by multiple routes, including respiratory particles such as large droplets and small aerosols.
+Unlike larger droplets, which settle quickly, aerosols can remain suspended in the air for extended periods\cite{Wang2021}.
+Airborne transmission is probably significant, but the relative contribution to overall transmission is still unclear, particularly because detection of respiratory viruses in the air remains challenging\cite{Belser2023PLOSPath}.
+
+Airborne infectious pathogens are primarily found in smaller particles (<10$\mu$m) and the distribution is similar across a range of different pathogens\cite{Fennelly2020}.
+Thus, pathogen-carrying aerosols have the potential for long-range transmission, but the larger concentration of particles near the infectious person favours short-range transmission\cite{Tellier2009JTRSI,Wang2020}.
+The complexity of airborne transmission lies in the interplay between the physicochemical properties of pathogen-carrying aerosols and environmental conditions\cite{Wang2021}, making it challenging to devise effective public health strategies to mitigate the spread of respiratory infections in various indoor settings, including schools, offices, and hospitals.
+
+In this study, we compared saliva samples, bioaerosol samples, and samples from the HEPA filters of air cleaners that were collected as part of two studies that we conducted in a Swiss school setting in winter 2021/2022 (during the SARS-CoV-2 omicron wave, henceforth 2022)\cite{Banholzer2023PLoSMed} and winter 2022/2023 (after the COVID-19 pandemic, henceforth 2023)\cite{Banholzer2023submitted}.
+We show differences in the distribution of respiratory viruses and analyze the results from airborne molecular detection.
 
@@
 
-Data was collected in two classes of two secondary schools in winter 2021/2022 and two classrooms of one secondary school in winter 2022/2023 (age of students 14-17~years) in the canton of Solothurn, Switzerland, both over a seven-week study period from end of January to beginning of March.
-In winter 2021/2022, a general mask mandate was in place for the first 3~study weeks, followed by 2~weeks without intervention, and another 2~weeks of air cleaners installed in two classrooms.
-In winter 2022/2023, air cleaners were installed for half of the study period following a cross-over design.
-An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, temperature, and humidity in all classrooms.
+Data was collected in Swiss secondary schools (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from end of January to beginning of March.
+Two classes of two different secondary schools (school 1 and 2) participated in 2022 and two classrooms of one secondary school (\ie only school 2) participated in 2023.
+In 2022, a general mask mandate was in place for the first 3~study weeks, followed by 2~weeks without intervention, and another 2~weeks during which air cleaners were installed in the classrooms.
+In 2023, air cleaners were installed for half of the study period in the classrooms following a cross-over study design.
 
-School classes participated in repetitive weekly (winter 2021/2022) and bi-weekly (winter 2022/2023) testing for a panel of respiratory infections.
+All classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system, but were ventilated using passive window ventilation.
+An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, temperature, and humidity.
+In 2022, passive window ventilation occurred per recommendations of the national public health authorities.
+In school~1, ventilation was additionally assisted by a CO$_2$ guided opener of a small window at the top.
+In 2023, passive window ventilation occurred solely at the discretion of the teachers.
+
+Receptive testing for a panel of respiratory infections was done weekly in 2022 and bi-weekly in 2023.
 Airborne respiratory viruses were collected in each classroom with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France) and a BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft.
@@
 Saliva samples were analyzed directly without prior filtration.
-The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19 major respiratory viruses and viral subtypes, including SARS-CoV-2, influenza, respiratory syncytial, metapneumovirus, adenovirus, rhinovirus, and parainfluenza.
+The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19 major respiratory viruses and viral subtypes, including SARS-CoV-2, influenza A/B, respiratory syncytial, metapneumovirus, adenovirus, rhinovirus, and parainfluenza.
 
-Descriptive statistics were utilized to depict differences in the type, number, and proportion of respiratory viruses between winter 2021/2022 and winter 2022/2023.
+Descriptive statistics were used to present differences in the type, number, and proportion of respiratory viruses detected in saliva and airborne samples between 2022 and 2023.
 
@@
 
+In 2022, 44~students participated in weekly saliva testing.
+There were 21~positive saliva samples over the study, namely 19~SARS-CoV-2, 1~influenza~A, and 1~adenovirus.
+There were 10~positive bioaerosol samples, namely 9~SARS-CoV-2 and 1~adenovirus.
+There were 8~positive samples on the HEPA-filters of the air cleaners, namely 6~SARS-CoV-2, 1~influenza~A and 1~adenovirus.
+In 2023, 38~students from two classes of secondary school~2 participated in bi-weekly saliva testing.
+There were 50~positive saliva samples over the study, namely 15~influenza~B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, 1~parainfluenza virus.
+There were 2~positive bioaerosol samples, namely 1~rhinovirus and 1~adenovirus.
+There were 4~positive samples on the HEPA-filters of the air cleaners, namely 1~influenza~B, 1~rhinovirus, 1~adenovirus, and 1~SARS-CoV-2.
+
+In comparison, there was a clear shift in the distribution of respiratory viruses from predominantly SARS-CoV-2 in 2022 to predominantly non-SARS-CoV-2 viruses in 2023 (\Cref{fig:comparison}a).
+Molecular detection also varied between the two years, especially for airborne samples (\Cref{fig:comparison}b).
+For saliva, 21 out of 262~samples were positive in 2022 compared to 50 out of 448~samples in 2023.
+For bioaerosols, 10 out of 130~samples were positive in winter 2022 compared to 2 out of 105~samples in 2023 .
+For the HEPA-filters, 8 out of 40~swabs were positive in 2022 compared to 4 out of 160~swabs in 2023.
+
+\begin{figure}
+    \centering
+    \includegraphics{results/comparison.pdf}
+    \caption{Comparison of molecular detection of respiratory viruses between winter 2021/2022 (2022) and winter 2022/2023 (2023).
+\textbf{(a)}~Distribution of respiratory viruses found in saliva.
+IF: influenza~A/B, HRV: human rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: human metapneumovirus, PIV: parainfluenza virus.
+\textbf{(b)}~Number of positive samples in saliva, bioaerosols, and on the HEPA-filters of air cleaners.}
+    \label{fig:comparison}
+\end{figure}
+
+Daily mean CO$_2$ level was 1,064\,ppm (standard deviation [SD] 232\,pppm) in 2022 compared to 1,702\,ppm (SD 370\,pppm) in 2023, suggesting that the classrooms were less well ventilated after the COVID-19 pandemic.
+Daily mean relative humidity was 38\% (SD 6\%) in 2022 compared to 38\% (SD 5\%) in 2023.
+Daily mean temperature was 19°C (SD 2°C) in 2022 compared to 22°C (SD 1°C) in 2023.
+
 \section*{Discussion}
 
-Furthermore, the distribution of respiratory viruses in this study was very different from that in our previous study.
+We compared molecular data collected from two studies in Swiss secondary schools in winter 2021/2022 (2022) and 2022/2023 (2023).
+During the COVID-19 pandemic in 2022, we detected mainly SARS-CoV-2 in saliva and also in bioaerosols and on the HEPA-filters of air cleaners.
+After the pandemic in 2023, we detected a clear shift in the distribution of respiratory viruses towards mainly non-SARS-CoV-2 respiratory viruses such as influenza, adenovirus, and rhinovirus.
+These viruses were rarely detected in bioaerosols and on the HEPA-filters of air cleaners.
 
-% why could we not detect so many bioaerosols
+The differences between the two study periods (pandemic versus non-pandemic) suggest that SARS-CoV-2 is easier to detect in the air.
+One possible explanation is that SARS-CoV-2 can remain suspended and active in the air for extended periods of time, facilitating long-range transmission.
+In contrast, non-SARS-CoV-2 viruses, such as influenza and adenovirus, were rarely detected in the air, suggesting they settle or become inactive more quickly, and require prolonged close contact for transmission.
+However, these findings should be interpreted with caution as they originate from an analysis of secondary data collected in a real-world setting, not a rigorous experiment.
+Furthermore, although there may be variations in the airborne survival of respiratory viruses, it is important to note that prolonged close contact continues to play a significant role in transmission.
+This holds true even for viruses capable of long-range transmission, as exemplified by SARS-CoV-2\cite{Leung2020NatMed,Lind2023NatCommun}.
 
-We detected fewer respiratory viruses in bioaerosol samples compared to our previous study\cite{Banholzer2023PLoSMed} (1~sample of adenovirus and 1~sample of rhinovirus in this study vs 10~samples of SARS-CoV-2 in the previous study), especially in relation to the number of positive saliva samples (50~samples in this study vs 21~in the previous study).
+Technical reasons for the differences in airborne detection are unlikely because the bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
+Instead, airborne detection could have been influenced by environmental factors.
+Although temperature and relative humidity were similar in both study periods, the same levels may have different effects on the airborne survival of respiratory viruses\cite{Tellier2009JTRSI,Davis1971AM}.
+A change in ventilation conditions could also have had a profound effect, but CO$_2$ levels were even higher in winter 2023, suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
+We speculate that the difference in airborne molecular detection is due to the different characteristics of the respiratory viruses.
+Differences in the physicochemical properties of virus-laden aerosols (\eg particle size and viral load), which influence the generation and survival of airborne viral RNA\cite{Wang2021}, may facilitate the detection of SARS-CoV-2 in the air.
+This is because viral loads often do not exceed the detection limits of existing devices for bioaerosol sampling, which remains a challenge\cite{Belser2023PLOSPath}.
+For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.
+However, several studies , including ours, have shown that sporadically non-SARS-CoV-2 viruses such as influenza, rhinovirus, and adenovirus can also be detected in the air\cite{Bischoff2013JID,Pan2017mSphere,Myatt2004AJRCCM}.
+Finally, the ability to detect pathogens in bioaerosols may be related to the patient characteristics of the study participants.
+Several studies have suggested an association between the generation of pathogenic aerosols and infectiousness\cite{Leung2020NatMed,Bischoff2013JID}.
+Therefore, studies that include more infectious patients (intentionally or unintentionally) are likely to have a higher chance of detecting respiratory viruses in the air.
+
+In conclusion, our comparison of two years of respiratory viral transmission in a school setting suggests that SARS-CoV-2 is easier to detect in indoor air as compared to other endemic viruses such as influenza, adenovirus and rhinoviruses.
+Future studies should test this hypothesis experimentally and investigate the relative importance of different modes of transmission.
+
+% We detected fewer respiratory viruses in bioaerosol samples compared to our previous study\cite{Banholzer2023PLoSMed} (1~sample of adenovirus and 1~sample of rhinovirus in this study vs 10~samples of SARS-CoV-2 in the previous study), especially in relation to the number of positive saliva samples (50~samples in this study vs 21~in the previous study).
 Similarly, the number of viruses detected on the filters of the air cleaners was also lower in this study (4~positive samples in 160~swabs vs 8~positive samples in 80~swabs in the previous study).
-Importantly, there was a clear shift in the distribution of respiratory viruses in the current study compared to the previous study in 2022 during the SARS-CoV-2 Omicron variant.
+Importantly,  in the current study compared to the previous study in 2022 during the SARS-CoV-2 Omicron variant.
 Whereas in 2022 we detected almost exclusively SARS-CoV-2 in human saliva, the most common respiratory viruses in 2023 were adenovirus, influenza~B, and rhinoviruses (only three SARS-CoV-2-positive samples).
@@
 
-% other reasons
-
-Technical reasons for the differences in airborne detection are unlikely.
-The bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
-Instead, airborne detection could have been influenced by environmental factors.
-Although temperature (overall between 19 and 25$^{\circ}$C) and relative humidity (overall between 25 and 50\%) were similar in both studies, the same levels may have different effects on the airborne survival of respiratory viruses.
-For example, airborne survival of rhinovirus and adenovirus increases with relative humidity, whereas the opposite relationship has been documented for influenza and SARS-CoV-2\cite{Tellier2009JTRSI,Ahlawat2020AAQR,Biryukov2020mS,Karim1985CJM,Davis1971AM}.
-A change in ventilation conditions could also have had a profound effect, but CO$_2$ levels were higher compared to the previous study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
-Although lower ventilation was partially offset by a greater reduction in aerosol and particle concentrations from air cleaners, the removal of particles by air cleaners may be slower than by natural ventilation.
-
-% bioaerosol sampling in more depth
-
-We speculate that the difference in airborne molecular detection of respiratory viruses between 2022 (pandemic) and 2023 (non-pandemic) is due to the different characteristics of the respiratory viruses.
-Differences in the physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity), which influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}, may facilitate the detection of SARS-CoV-2 in the air.
-The reason is that viral loads do often not exceed the detection limits of existing devices for bioaerosol sampling, which remains a challenge\cite{Wang2021,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
-For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.
-However, several studies showed that influenza\cite{Bischoff2013JID,Pan2017mSphere}, rhinovirus\cite{Myatt2004AJRCCM}, and adenovirus\cite{Nguyen2017OFID,Pan2017mSphere} can also be detected in airborne samples.
-Finally, the ability to detect pathogens in bioaerosols may be related to patient characteristics of the study participants.
-Several studies suggested an association between the generation of pathogenic aerosols and infectiousness\cite{Leung2020NatMed,Bischoff2013JID,Escombe2008PLoSMed}.
-Therefore, studies that include more infectious patients (intentionally or unintentionally) are likely to have a higher chance of detecting respiratory viruses in the air.
-
-\section{Comparison with previous study}\label{sec:comparison-prev-study}
-
-% descriptive results
-
-In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners during the SARS-CoV-2 Omicron wave.
-\cref{fig:comparison} shows a comparison of the results.
-The distribution of respiratory infections is different (\cref{fig:comparison}a) because we almost exclusively detected SARS-CoV-2 in our previous study during the COVID-19 pandemic, as compared to a range of respiratory infections in this study during non-pandemic times.
-Although there were more positive saliva samples in this study, the number of respiratory infections detected in bioaerosols and on the HEPA-filters of the air cleaners was lower (\cref{fig:comparison}b).
-The reductions in particle concentrations were larger and with lower uncertainty in this study (\cref{fig:comparison}c).
-The risk of infection tended to be lower with air cleaners in this study where air cleaners were analyzed using a cross-over design, but not in our previous study where air cleaners were installed only at the end of the study when most students had already been infected with SARS-Cov-2(\cref{fig:comparison}d).
-
-% summary
-
-To summarize, we found similar reductions in particle concentrations and a tendency for the relative risk of infection to be lower with air cleaners.
-To some extent, the differences between the two studies may be explained by different adjustments.
-While we could not adjust for ventilation in our previous study, the estimates in this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see \Cref{sec:detailed-molecular}).
-Furthermore, the distribution of respiratory viruses in this study was very different from that in our previous study and molecular airborne detection was lower.
-
-\begin{figure}[!htpb]
-    \includegraphics[width=\linewidth]{../../results/comparison.pdf} 
-    \caption{\textbf{Comparison of the results between our previous study in 2022 during the COVID-19 pandemic (blue) and this study in 2023 during non-pandemic times (black).} \textbf{(a)}~Proportion of positive saliva samples by virus.
-Flu B: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.
-\textbf{(c)}~Estimated reduction in aerosol number and particle mass concentrations (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$) with air cleaners (posterior mean as squares and 95\%-CrI as lines).
-\textbf{(d)}~Adjusted relative risk (ARR) of infection with air cleaners (posterior mean as squares and 95\%-CrI as lines).}
-    \label{fig:comparison}
-\end{figure}
-
-% unlikely reasons for less airborne detection
-
-Technical reasons for the differences in airborne detection are unlikely.
-The bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
-Instead, airborne detection could have been influenced by environmental factors.
-Although temperature (overall between 19 and 25$^{\circ}$C) and relative humidity (overall between 25 and 50\%) were similar in both studies, the same levels may have different effects on the airborne survival of respiratory viruses.
-For example, airborne survival of rhinovirus and adenovirus increases with relative humidity, whereas the opposite relationship has been documented for influenza and SARS-CoV-2\cite{Tellier2009JTRSI,Ahlawat2020AAQR,Biryukov2020mS,Karim1985CJM,Davis1971AM}.
-A change in ventilation conditions could also have had a profound effect, but CO$_2$ levels were higher compared to the previous study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
-Although lower ventilation was partially offset by a greater reduction in aerosol and particle concentrations from air cleaners, the removal of particles by air cleaners may be slower than by natural ventilation.
-
-% likely reason for less airborne detection
-
-More likely, SARS-CoV-2 may be easier to detect in the air due to physicochemical properties of virus-laden aerosols, so that its viral loads more often exceed the detection limits of existing devices for bioaerosol sampling,  which remains a challenge\cite{Wang2021,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.
-For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.
-However, several studies showed that influenza\cite{Bischoff2013JID,Pan2017mSphere}, rhinovirus\cite{Myatt2004AJRCCM}, and adenovirus\cite{Nguyen2017OFID,Pan2017mSphere} can also be detected in airborne samples.
-
-\bibliography{../paper/references.bib}
+\bibliography{references.bib}
 

### 2023-12-18 analysis/study_comaprison.Rmd
minor changes to study comparison

@@
   group_by(year_num, class, week) %>%
-  summarize(co2.max = mean(co2.max, na.rm = T)) %>%
+  summarize(maxCO2_imputed = ifelse(is.na(co2.max), T, F), 
+            co2.max = mean(co2.max, na.rm = T)) %>%
   ungroup() %>%
@@
 pair_df <- left_join(pair_df, vent)
+
+pair_df %>%
+  group_by(year_num, class, week) %>%
+  slice(1) %>%
+  ungroup() %>%
+  filter(!maxCO2_imputed) %>%
+  group_by(year_num) %>%
+  summarize(mean(maxCO2), sd(maxCO2)) %>%
+  mutate_if(is.numeric, round, 0)
 
@@
 # add prior scales
-stanDF$s_saliva <- 1 / sd(stanDF$saliva)
-stanDF$s_cov <- 1 / sd(stanDF$cov)
-stanDF$s_masks <- .25 #/ sd(stanDF$masks)
-stanDF$s_air <- .25 #/ sd(stanDF$air_cleaners)
-stanDF$s_co2 <- .25
-stanDF$s_cov <- 1
+stanDF$s_saliva <- sd(stanDF$bioaerosol) / sd(stanDF$saliva)
+stanDF$s_cov <- sd(stanDF$bioaerosol) / sd(stanDF$cov)
+stanDF$s_masks <- .5 * sd(stanDF$bioaerosol) / sd(stanDF$masks)
+stanDF$s_air <- .5 * sd(stanDF$bioaerosol) / sd(stanDF$air_cleaners)
+stanDF$s_co2 <- sd(stanDF$bioaerosol)
 
@@
   mutate(beta_cov = exp(beta_cov)) %>%
-  median_qi()
+  median_qi() %>%
+  mutate_if(is.numeric, round, 1)
 

### 2023-12-18 doc/comparison-mcid12/manuscript.tex
Updates from Overleaf

@@
 \usepackage{import}
-\usepackage{../paper/main-lancet}
+\usepackage{main}
 
@@
 
-\title{\bfseries\LARGE\singlespacing{Brief report:
-Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols}}
+\title{\bfseries\LARGE\singlespacing{Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols}}
 
@@
 \begin{information}\normalfont
-\textbf{Word count:} Abstract 150 (max 150), Manuscript 1,625 (max 1,500) \\
-\textbf{Display items:} Figure 1 \\
-\textbf{References:} 17 (max 12) \\
+\textbf{Word count:} Abstract 219 (max 250), Manuscript 1,380 (max 1,200) \\
+\textbf{Display items:} Figure 1 (max 2) \\
+\textbf{References:} 15 (max 15) \\
 %\textbf{Supporting information:} STROBE checklist, Supplementary Data with texts, tables, and figures
@@
 \begin{abstract}\normalfont
-\noindent We compared airborne and saliva samples from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave) and winter 2022/2023 (after the COVID-19 pandemic) in a Swiss school setting.
-While during the pandemic we detected almost exclusively SARS-CoV-2 in saliva (19~samples of SARS-CoV-2 vs 2~samples of non-SARS-CoV-2), after the pandemic the most common respiratory viruses were adenovirus, influenza, and rhinoviruses (3~samples of SARS-CoV-2 vs 47~samples of non-SARS-CoV-2).
+\noindent \textbf{Objectives:} To compare molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols.
+
+\noindent \textbf{Methods:} We analyze saliva and bioaerosol samples from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave) and winter 2022/2023 (after the COVID-19 pandemic) in a Swiss school setting.
+We estimate the probability of weekly airborne detection of SARS-CoV-2 vs non-SARS-CoV-2 viruses using a Bayesian logistic regression model.
+
+\noindent \textbf{Results:} While during the pandemic SARS-CoV-2 was almost exclusively detected in saliva (19~samples of SARS-CoV-2 vs 2~samples of non-SARS-CoV-2), the most common respiratory viruses after the pandemic were adenovirus, influenza, and rhinoviruses (3~samples of SARS-CoV-2 vs 47~samples of non-SARS-CoV-2).
 Despite the same study setting and methods used, airborne detection of SARS-CoV-2 was relatively more frequent than of non-SARS-CoV-2 viruses (10~samples of SARS-CoV-2 in 2021/2022 vs 2~samples of non-SARS-CoV-2 in 2022/2023).
-SARS-CoV-2 was also detected more frequently on the HEPA-filters of installed portable air cleaners (6~samples of SARS-CoV-2 vs 3~samples of non-SARS-CoV-2).
+Based on our model, the posterior probability of weekly airborne detection was 31\% (95\%-credible interval [CrI] 13\%$-$54\%) for SARS-CoV-2 vs 11\% (95\%-CrI 4\%-23\%) for non-SARS-CoV-2.
+
+\noindent \textbf{Discussion:} There was a clear shift in the distribution of respiratory viruses from predominantely SARS-CoV-2 in 2022 to other respiratory viruses in 2023.
 Our comparison suggests that SARS-CoV-2 is easier to detect in the air than other endemic respiratory viruses, possibly facilitating long-range transmission.
+Future experimental studies are needed to confirm this finding.
 \medskip
@@
 
-Data was collected in Swiss secondary schools (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from end of January to beginning of March.
-Two classes of two different secondary schools (school 1 and 2) participated in 2022 and two classrooms of one secondary school (\ie only school 2) participated in 2023.
-In 2022, a general mask mandate was in place for the first 3~study weeks, followed by 2~weeks without intervention, and another 2~weeks during which air cleaners were installed in the classrooms.
-In 2023, air cleaners were installed for half of the study period in the classrooms following a cross-over study design.
-
+Data was collected in two Swiss secondary schools (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from end of January to beginning of March.
+Three classes participated in 2022 and two classes in 2023.
+Two classes shared the same classroom in 2022 due to half-class teaching, other classes had a separate classroom.
 All classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system, but were ventilated using passive window ventilation.
 An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, temperature, and humidity.
-In 2022, passive window ventilation occurred per recommendations of the national public health authorities.
-In school~1, ventilation was additionally assisted by a CO$_2$ guided opener of a small window at the top.
-In 2023, passive window ventilation occurred solely at the discretion of the teachers.
+A detailed comparison of the study settings can be found in \supp~Table~\zref{tab:comp_study}.
 
-Receptive testing for a panel of respiratory infections was done weekly in 2022 and bi-weekly in 2023.
+Repetitive testing for a panel of respiratory infections was done weekly in 2022 and bi-weekly in 2023.
 Airborne respiratory viruses were collected in each classroom with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France) and a BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft.
@@
 The removable parts were regularly autoclaved.
-Finally, swabs from the air cleaners' HEPA filters were collected at the end of the study in 2022 and after each intervention phase in 2023.
+Finally, swabs from the air cleaners' HEPA filters were collected.
 The HEPA filters were removed and divided into 20~fields.
@@
 Saliva and airborne samples were transported to the Institute of Infectious Diseases and stored at $-$80°C until further processing\cite{Huber2021}.
-
 Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL.
@@
 
-Descriptive statistics were used to present differences in the type, number, and proportion of respiratory viruses detected in saliva and airborne samples between 2022 and 2023.
+Descriptive statistics were used to present differences in the type and number of respiratory viruses detected in saliva and airborne samples between 2022 and 2023.
+A Bayesian logistic regression model was used to estimate the probability of detecting any SARS-CoV-2 vs non-SARS-CoV-2 in the air during a study week, adjusting for differences in the study settings and the interventions that were in place during the studies (see \supp~Text~\zref{sec:model} for a detailed model description).
 
@@
 
-In 2022, 44~students participated in weekly saliva testing.
-There were 21~positive saliva samples over the study, namely 19~SARS-CoV-2, 1~influenza~A, and 1~adenovirus.
+In 2022, 58~students participated in weekly saliva testing.
+There were 21~positive saliva samples over the study, namely 19~SARS-CoV-2, 1~influenza~A, and 1~adenovirus (\Cref{fig:comparison}a, left).
 There were 10~positive bioaerosol samples, namely 9~SARS-CoV-2 and 1~adenovirus.
 There were 8~positive samples on the HEPA-filters of the air cleaners, namely 6~SARS-CoV-2, 1~influenza~A and 1~adenovirus.
-In 2023, 38~students from two classes of secondary school~2 participated in bi-weekly saliva testing.
-There were 50~positive saliva samples over the study, namely 15~influenza~B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, 1~parainfluenza virus.
+In 2023, 38~students participated in bi-weekly saliva testing.
+There were 50~positive saliva samples over the study, namely 15~influenza~B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, 1~parainfluenza virus (\Cref{fig:comparison}a, right).
 There were 2~positive bioaerosol samples, namely 1~rhinovirus and 1~adenovirus.
@@
 
-In comparison, there was a clear shift in the distribution of respiratory viruses from predominantly SARS-CoV-2 in 2022 to predominantly non-SARS-CoV-2 viruses in 2023 (\Cref{fig:comparison}a).
-Molecular detection also varied between the two years, especially for airborne samples (\Cref{fig:comparison}b).
-For saliva, 21 out of 262~samples were positive in 2022 compared to 50 out of 448~samples in 2023.
-For bioaerosols, 10 out of 130~samples were positive in winter 2022 compared to 2 out of 105~samples in 2023 .
-For the HEPA-filters, 8 out of 40~swabs were positive in 2022 compared to 4 out of 160~swabs in 2023.
+We found 6 positive air-saliva samples in the same respective week (4~SARS-CoV-2 and 2~non-SARS-CoV-2 (\Cref{fig:comparison}b), suggesting they were paired samples.
+Based on our Bayesian logistic regression model, SARS-CoV-2 was more likely detected in bioaerosols than non-SARS-CoV-2 (adjusted odds ratio 3.6, 95\%-CrI 1.5$-$8.9).
+The probability of molecular airborne detection was 31\% (95\%-credible interval [CrI] 13\%$-$54\%) for SARS-CoV-2 vs 11\% (95\%-CrI 4\%-23\%) for non-SARS-CoV-2 (\Cref{fig:comparison}c).
+Our estimates are adjusted for differences in the daily maximum CO$_2$ level, which was 1,134\,ppm (standard deviation [SD] 277\,pppm) in 2022 compared to 2,224\,ppm (SD 321\,pppm) in 2023.
+Humidity and temperature were similar, with daily mean relative humidity of 38\% (SD 6\%) in 2022 vs 38\% (SD 5\%) in 2023 and daily mean temperature of 19°C (SD 2°C) in 2022 vs 22°C (SD 1°C) in 2023.
 
@@
 IF: influenza~A/B, HRV: human rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: human metapneumovirus, PIV: parainfluenza virus.
-\textbf{(b)}~Number of positive samples in saliva, bioaerosols, and on the HEPA-filters of air cleaners.}
+\textbf{(b)}~Positive samples in saliva and bioaerosols per study week.
+\textbf{(c)}~Probability of detecting any SARS-CoV-2 and non-SARS-CoV-2 in bioaerosols during a study week (posterior mean as dots, interquartile range as box, 95\%-CrI as error bars).}
     \label{fig:comparison}
@@
 
-Daily mean CO$_2$ level was 1,064\,ppm (standard deviation [SD] 232\,pppm) in 2022 compared to 1,702\,ppm (SD 370\,pppm) in 2023, suggesting that the classrooms were less well ventilated after the COVID-19 pandemic.
-Daily mean relative humidity was 38\% (SD 6\%) in 2022 compared to 38\% (SD 5\%) in 2023.
-Daily mean temperature was 19°C (SD 2°C) in 2022 compared to 22°C (SD 1°C) in 2023.
-
 \section*{Discussion}
 
+% In comparison,
+
 We compared molecular data collected from two studies in Swiss secondary schools in winter 2021/2022 (2022) and 2022/2023 (2023).
-During the COVID-19 pandemic in 2022, we detected mainly SARS-CoV-2 in saliva and also in bioaerosols and on the HEPA-filters of air cleaners.
-After the pandemic in 2023, we detected a clear shift in the distribution of respiratory viruses towards mainly non-SARS-CoV-2 respiratory viruses such as influenza, adenovirus, and rhinovirus.
-These viruses were rarely detected in bioaerosols and on the HEPA-filters of air cleaners.
+We detected predominantly SARS-CoV-2 in 2022 and predominantly non-SARS-CoV-2 viruses, such as influenza and adenovirus, in 2023.
+Molecular airborne detection of SARS-CoV-2 was more likely than of non-SARS-CoV-2.
 
@@
 One possible explanation is that SARS-CoV-2 can remain suspended and active in the air for extended periods of time, facilitating long-range transmission.
-In contrast, non-SARS-CoV-2 viruses, such as influenza and adenovirus, were rarely detected in the air, suggesting they settle or become inactive more quickly, and require prolonged close contact for transmission.
-However, these findings should be interpreted with caution as they originate from an analysis of secondary data collected in a real-world setting, not a rigorous experiment.
-Furthermore, although there may be variations in the airborne survival of respiratory viruses, it is important to note that prolonged close contact continues to play a significant role in transmission.
-This holds true even for viruses capable of long-range transmission, as exemplified by SARS-CoV-2\cite{Leung2020NatMed,Lind2023NatCommun}.
+In contrast, non-SARS-CoV-2 viruses were rarely detected in the air, suggesting they settle or become inactive more quickly.
+Therefore, prolonged close contact may be relatively more important for the transmission of respiratory viruses other than SARS-CoV-2, although previous studies suggest it also facilitates the transmission of SARS-CoV-2\cite{Leung2020NatMed,Lind2023NatCommun}.
 
 Technical reasons for the differences in airborne detection are unlikely because the bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
-Instead, airborne detection could have been influenced by environmental factors.
-Although temperature and relative humidity were similar in both study periods, the same levels may have different effects on the airborne survival of respiratory viruses\cite{Tellier2009JTRSI,Davis1971AM}.
-A change in ventilation conditions could also have had a profound effect, but CO$_2$ levels were even higher in winter 2023, suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic.
-We speculate that the difference in airborne molecular detection is due to the different characteristics of the respiratory viruses.
-Differences in the physicochemical properties of virus-laden aerosols (\eg particle size and viral load), which influence the generation and survival of airborne viral RNA\cite{Wang2021}, may facilitate the detection of SARS-CoV-2 in the air.
-This is because viral loads often do not exceed the detection limits of existing devices for bioaerosol sampling, which remains a challenge\cite{Belser2023PLOSPath}.
+Temperature and relative humidity were also similar in both study periods, but the same levels may have virus-specific effects on airborne survival\cite{Tellier2009JTRSI,Davis1971AM}.
+Ventilation changed profoundly, but CO$_2$ levels were even higher in 2023, possibly enhancing airborne survival.
+Additional differences in the study setting (\supp~Table~\zref
+{tab:comp_study}) that could have influenced airborne detection were adjusted for in the Bayesian model (\supp~Text~\zref{sec:model}).
+Therefore, we speculate that the difference in airborne detection is due to the different characteristics of the respiratory viruses which can influence the generation and survival of airborne viral RNA\cite{Wang2021}.
+Low viral loads do often not exceed the detection limits of existing devices for bioaerosol sampling.
 For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.
-However, several studies , including ours, have shown that sporadically non-SARS-CoV-2 viruses such as influenza, rhinovirus, and adenovirus can also be detected in the air\cite{Bischoff2013JID,Pan2017mSphere,Myatt2004AJRCCM}.
-Finally, the ability to detect pathogens in bioaerosols may be related to the patient characteristics of the study participants.
-Several studies have suggested an association between the generation of pathogenic aerosols and infectiousness\cite{Leung2020NatMed,Bischoff2013JID}.
-Therefore, studies that include more infectious patients (intentionally or unintentionally) are likely to have a higher chance of detecting respiratory viruses in the air.
+Finally, the ability to detect pathogens in bioaerosols may be related to the variation in infectiousness between study participants, which is associated with the generation of pathogenic aerosols\cite{Leung2020NatMed,Bischoff2013JID}.
 
-In conclusion, our comparison of two years of respiratory viral transmission in a school setting suggests that SARS-CoV-2 is easier to detect in indoor air as compared to other endemic viruses such as influenza, adenovirus and rhinoviruses.
-Future studies should test this hypothesis experimentally and investigate the relative importance of different modes of transmission.
-
-% We detected fewer respiratory viruses in bioaerosol samples compared to our previous study\cite{Banholzer2023PLoSMed} (1~sample of adenovirus and 1~sample of rhinovirus in this study vs 10~samples of SARS-CoV-2 in the previous study), especially in relation to the number of positive saliva samples (50~samples in this study vs 21~in the previous study).
-Similarly, the number of viruses detected on the filters of the air cleaners was also lower in this study (4~positive samples in 160~swabs vs 8~positive samples in 80~swabs in the previous study).
-Importantly,  in the current study compared to the previous study in 2022 during the SARS-CoV-2 Omicron variant.
-Whereas in 2022 we detected almost exclusively SARS-CoV-2 in human saliva, the most common respiratory viruses in 2023 were adenovirus, influenza~B, and rhinoviruses (only three SARS-CoV-2-positive samples).
-The shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.
+In conclusion, we observed a clear shift in the distribution of respiratory viruses from predominantly SARS-CoV-2 in 2022 to predominantly non-SARS-CoV-2 viruses in 2023.
+Molecular airborne detection of SARS-CoV-2 was more likely than of other endemic respiratory viruses, but future experimental studies are needed to confirm this finding.
 

### 2023-12-18 doc/comparison-mcid12/supplements.tex

--- 2023-12-04 doc/paper/supplements.tex
@@
 \title{\LARGE\singlespacing{\textbf{Supplmenetary Material} \\ \medskip
-Air cleaners and respiratory infections in schools: \\ A multi-measurement observational study}}
+Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols}}
 
@@
 % author list
-\author[1$\ddag$,2]{Nicolas Banholzer}
+\author[1,2]{Nicolas Banholzer}
 \author[2,3]{Philipp Jent}
@@
 \affil[8]{Institute of Educational Science, University of Bern, Bern, Switzerland}
-
-\affil[*]{Corresponding author: lukas.fenner@ispm.unibe.ch }
-
-\affil[$\ddag$]{These authors contributed equally to this work.}
 
@@
 
-\tableofcontents
-
-\listoffigures
-
-\listoftables
-
-%\listoffigures
-%\listoftables
-
-\newpage
-
-\section{Experimental determination of the CADR}\label{sec:cadr}
-
-The clean air delivery rate (CADR) is typically used to quantify the cleaning efficiency of air cleaners\cite{Nelson1993IndoorEnv}.
-It is expressed in m$^3$/s and describes the volumetric flow rate of particle-free, clean air that an air cleaner delivers into the room.
-The CADR is equal to the product of the filtration efficiency and the volumetric flow rate passing through the device.
-
-We followed the approach by K\"{u}pper et al.
-to determine the CADR\cite{Kuepper20199AAQR}.
-The approach is based on the release of fine aerosols (particle diameter $D$<1$\mu$m) in a closed, unventilated room by measuring the decay rate $k$ of the particle concentration $N(t)$ caused by the air cleaner, \ie
-\begin{align}
-    N(t) = N_0 \cdot e^{-k\cdot t}
-\end{align}
-Since the surfaces in the room (\eg walls and furniture) are also sinks for aerosol particles, measurements were carried out with ($k_\mathrm{cleaner}$) and without air cleaners($k_\mathrm{background}$).
-Based on that, we can compute the CADR as
-\begin{align}
-    \mathrm{CADR} = V \cdot k_\mathrm{cleaner} - V \cdot k_\mathrm{background},
-\end{align}
-where $V = 233 m^3$ is the volume of the classroom.
-
-We ensured that the particle concentrations in the experiments were not too high, thereby preventing bias due to coagulation (\ie the formation of a few large particles from many small particles).
-At the beginning of each experiment, fine aerosol particles were released for a few minutes by ultrasonic nebulization of an aqueous NaCl solution of 5\,g/l.
-The generated dehydrated salt particles have a broad size distribution in the range of 0.02$-$1.0$\mu$m.
-Decay rates were measured using 11~fine dust sensors (SPS30, Sensirion AG), which were placed at head height of the students.
-These sensors measured the number concentration of individual particles by counting the scattered light pulses they produce when passing through a laser beam.
-Because of their scattering properties, these sensors measure the number of particles larger than $\sim$0.3\,$\mu$m.
-
-Figure~\ref{fig:experiment} shows the particle number concentration with and without without the air cleaner.
-The exponential decrease is clearly visible as a linear curve in the logarithmic plot.
-Based on these experiments, we were able to determine the net CADR values of the air cleaner as 420\,m$^3$/h with an uncertainty of $\pm$10\% due to systematic differences in the setup and measurement errors of the instruments.
-Our estimated CADR is lower than the CADR reported by the manufacturer of 600m$^3$/h, most likely because the reported CADR was determined with comparably large particles (pollen in the super micrometer size range).
-Thus, we explain the lower filtration efficiency determined here with the fact that CADR is strongly dependent on the particle size, \ie large particles are easier to filter than fine aerosols.
-
-\begin{figure}
+\begin{table}[!htpb]
+    \caption{Differences in the study settings of 2022 and 2023.}
+    \label{tab:comp_study}
     \centering
-    \includegraphics[width=\linewidth]{doc/paper/experiment-particle-concentration-decay.png}
-    \caption[Measured concentration curves with and without air cleaner]{\textbf{Measured concentration curves with and without air cleaner.} The red dot shows the average concentration measured by the 11~distributed sensors of particles with $0.3\mu m < D < 1 \mu m$.
-The standard deviation is shown in gray as an error bar.
-The increase in concentration at the beginning of these experiments is due to the release of the NaCl particles during approximately 15-20\,min.
-The inhomogeneity of the concentration during the nebulization period is seen in larger error bars.
-Afterwards, the aerosol source is turned off, the concentration homogenizes quickly, and the concentration drop follows an exponential decay, which is used for the CADR calculation.}
-    \label{fig:experiment}
-\end{figure}
-
-\clearpage
-
-\section{Data collection}\label{sec:data-collection}
-
-\Cref{tab:data} in \supp provides an overview of the types of data that were collected in each classroom.
-
-\begin{table}[!htpb]
-    \footnotesize
-    \centering
-    \caption[Overview of collected data]{\textbf{Overview of collected data}. Type of data collected, method/device, and frequency in the rooms of classes A and B.}
-    \begin{tabular}{p{3.5cm}p{6cm} p{1cm} p{1cm} p{3cm}}
+    \renewcommand{\arraystretch}{1.5}
+    \begin{tabular}{lp{6cm}p{6cm}}
+        \toprule
+        & \textbf{2022 (during pandemic)} & \textbf{2023 (after pandemic)} \\
         \midrule
-    Type of data & Method / device & Class A & Class B & Frequency \\
-    \midrule
-    \textbf{Molecular data} \\
-    \midrule
-    Bioaerosol sampling & Bioaerosol sampling devices: BioSpot-VIVAS (Aerosol Devices Inc., Ft.
-Collins, Colorado, United States) and Coriolis Micro Air (Bertin Instruments Montigny-le/Bretonneux, France) & x & (x) \newline only Coriolis & daily \\
-    Saliva samples & Saliva samples & x & x & twice per week \\
-    Swabs from air cleaners & Swab samples from Xiaomi HEPA filters (Xiaomi Mi Air Pro 70m2, Shenzhen, China) & x & x & after each intervention phase (both classes) and before the vacation (only class B) \\
-    \midrule
-    \textbf{Environmental data} \\
-    \midrule
-    CO$_2$, aerosol/particle concentrations, humidity, temperature & Air quality device: AQ Guard (Palas GmbH, Karlsruhe, Germany) & x & x & daily by minute \\
-    Coughs & Sounds collected via microphones and coughs detected with a Machine Learning model\cite{Bertschinger2023CBMS} & x & x & daily by seconds \\
-    \midrule
-    \textbf{Epidemiological data} \\
-    \midrule
-    Absences & Survey & x & x & daily \\
-    Student characteristics & Survey & x & x & at study start \\
+        Participants & 2~classes from School~1 sharing one classroom (half-class teaching) and 1~class from School~2 & two classes from School~2 \\
+        Duration & 7 weeks per class (January 16 to March 11, 2023) & 7 weeks per class (January 24 to March 26, 2023) \\
+        Saliva testing & weekly (Wednesday) & bi-weekly (Tuesday and Thursday) \\
+        Bioaerosol sampling & daily with BioSpot-VIVAS and Coriolis in both classrooms & daily with BioSpot-VIVAS and Coriolis in one classroom and only Coriolis in the other classroom \\
+        HEPA-filters sampling & once at the end of the study per filter and classroom & two times per filter and classroom \\
+        Interventions & compulsory face mask wearing as mandated by the public health authorities at that time from the start of the study for a total of 6 study weeks ; and installation of portable air cleaners towards the end of the study in the study classrooms for a total of 4 study weeks & installation of portable air cleaners in the study classrooms using a cross-over design for a total of 7 weeks \\
+        Ventilation conditions & passive window ventilation as per recommendations of the national public health authorities; in School~1 ventilation was additionally assisted by a CO$_2$ guided opener of a small window at the top & passive window ventilation solely at the discretion of the teachers \\
         \bottomrule
     \end{tabular}
-    \label{tab:data}
 \end{table}
@@
 
-\section{Epidemiological line list data}\label{sec:case-data}
+\section{Model to estimate differences in airborne detection}\label{sec:model}
 
-Reports about absences were entered electronically into a REDcap database\cite{Harris2009,Harris2019} \Cref{tab:epi-data-line-list}.
-shows the line list data for respiratory cases based on epidemiological data.
-For each class and case, it shows the date when the student was absent, s/he reported symptoms, and returned to school.
-We also collected data on laboratory tests but none of the students reported a test result during the study.
-A case of respiratory infection was defined as an absence where the student reported a sickness with at least one of the following symptoms: fever, coughing, tiredness, loss of test or smell, sore throat, headache, aches and pains, diarrhea, difficulty breathing or shortness of breath, stomach.
-Students were asked to report the first date they experienced symptoms, so that a student absent on Monday could report Saturday or Sunday as the day when symptoms began to be felt.
-We will refer respiratory cases by their dates of symptom onset, which usually corresponded to the absence date, unless the student attended the school while experiencing symptoms.
-Informed consent for one student could not be obtained and the single absence of this student was considered unrelated to a respiratory infection.
+Our outcome is a binary variable $Y_{tcv}$ which is $1$ if any respiratory virus $v$ was detected in the air of classroom $c$ during study week $t$.
+We model this outcome with a Bayesian logistic regression model
+\begin{align}
+    \text{Bernoulli-Logit}(y|\mu) = \begin{cases}
+        \text{logit}^{-1}(\mu) & \text{if }y=1\text{, and} \\
+        1-\text{logit}^{-1}(\mu) & \text{if }y=0.
+    \end{cases}
+\end{align}
+The parameter $\mu$ is related to the type of respiratory virus and differences in the study setting (see \Cref{tab:comp_study}) as follows
+\begin{align}
+    \mu_{tcv} = \log(\text{offset}_c) + \alpha + \beta_1\,\cdot\text{saliva}_{tcv} + \beta_2\,\cdot\text{CoV}_{tcv} + \beta_3\,\cdot\text{mask}_{tcv} + \beta_4\,\cdot\text{aircl}_{tcv} + 
+\end{align}
+where
+\begin{itemize}
+    \item \textbf{offset}: Since one classroom in 2022 was shared by two classes, we set the offset of this classroom to 2 (double odds for airborne detection in this classroom).
+Conversely, since one classroom in 2023 only had one instead of two sampling devices in place, we set the offset of this classroom to 0.5 (half odds for airborne detection in this classroom).
+    \item  \textbf{saliva}: Binary variable indicating if there was a positive saliva sample in the same week. Airborne detection is more likely if the same virus was found in student saliva.
+    \item \textbf{CoV}: Binary variable indicating if virus $v$ in classroom $c$ in week $t$ was SARS-CoV-2.
+Parameter $\beta_2$ models the log odds of detecting any SARS-CoV-2 vs non-SARS-CoV-2 in bioaerosols.
+This is our parameter of interest.
+    \item \textbf{masks}: Binary variable indicating if face mask wearing was mandated. Masks possibly reduce the probability of airborne detection.
+    \item \textbf{aircl}: Binary variable indicating if portable air cleaners were installed in the classroom. Air cleaners possibly reduce the probability of airborne detection.
+    \item \textbf{CO}$_2$: Weekly average of the daily maximum CO$_2$ levels in each classroom.
+Higher CO$_2$ levels are a proxy of worse ventilation (outdoor air exchange) and possibly increase the probability of airborne detection.
+\end{itemize}
 
-{\footnotesize\begin{longtable}{l l l l}
-    \caption[Line list of respiratory cases over the study period]{Line list of respiratory cases over the study period.}\label{tab:epi-data-line-list} \\
-    \toprule
-    Class & Date of absence & Date of symptom onset & Date of return \\
-    \midrule
-    \input ../../results/epi-data/line-list-data
-    \bottomrule
-\end{longtable}}
+We model the intercept with a diffuse prior
+\begin{align}
+    \alpha \sim \text{Student}(\nu=5, \mu = 0, \sigma = 2.5) ~.
+\end{align}
+Bioaerosols were more frequently found in weeks with a positive saliva sample. Therefore, we model $\beta_1$ with an informative prior
+\begin{align}
+    \beta_1 \sim \text{Normal}(\mu = 5/3, \sigma = s_y / s_{\text{saliva}}),
+\end{align}
+where 5 is the number of weeks with paired samples and 3 is the number of weeks with a positive bioaerosol but without a positive saliva sample.
+The prior's standard deviation is computed as the ratio of the standard deviation of the outcome $s_y$ and the variable $s_{\text{saliva}}$.
+\\
+Our parameter of interest is modeled with a non-informative prior
+\begin{align}
+    \beta_2 \sim \text{Normal}(\mu = 0, \sigma = s_y / s_\text{CoV})~.
+\end{align}
+Masks and air cleaners possibly reduce the odds molecular airborne detection. To incorporate this prior believe into $\beta_3$ and $\beta_4$, we used an informative asymmetric Laplace prior 
+\begin{align}
+    \beta_3 &\sim \text{Asymmetric-Laplace}(\mu = -0.32, \sigma = 0.5\cdot s_y / s_{\text{masks}}, \tau = 0.75) \\
+    \beta_4 &\sim \text{Asymmetric-Laplace}(\mu = -0.27, \sigma = 0.5\cdot s_y / s_{\text{aircl}}, \tau = 0.75),
+\end{align}
+which has a 90\% prior probability for a negative intervention effect and is skewed towards smaller effects. \\
+Similarly, we incorporate our prior belief that higher CO$_2$ levels possibly increase the odds of airborne detection with the converse prior for $\beta_5$ as
+\begin{align}
+    \beta_3 &\sim \text{Asymmetric-Laplace}(\mu = 0.21, \sigma = 0.5\cdot s_y / s_{\text{masks}}, \tau = 0.25)~.
+\end{align}
+To summarize, we estimate the odd ratio for airborne detection of SARS-CoV-2, adjusting for differences in the setup of the classrooms (offset), positive saliva sample in the same week (saliva), the effects of interventions (masks and aircl), and differences in ventilation conditions (CO$_2$).
+We use informative priors because the sample size is small and insufficient to identify the effects of the confounders.
 
@@
 
-\section{Molecular line list data}\label{sec:mol-data}
-
-\Cref{tab:mol-data-line-list} shows the line list data for the laboratory test results from human saliva samples.
-For each class and test, it shows the date when the test was taken, the test result and (if positive) the virus that was detected.
-Recall that all students in class were sampled twice per week every Tuesday and Thursday.
-
-{\footnotesize\begin{longtable}{l l l l l}
-    \caption[Line list of molecular test results over the study period]{Line list of respiratory cases over the study period.
-IFB: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}\label{tab:mol-data-line-list} \\
-    \toprule
-    Class & Date of test & Weekday & Test result & Virus \\
-    \midrule
-    \input ../../results/mol-data/line-list-data
-    \bottomrule
-\end{longtable}}
-
-\clearpage
-
-\section{Modeling changes in particle concentrations}\label{sec:env-regression-model}
-
-We only analyze particle concentration data of the time periods during which students were in the classroom.
-Particle concentrations $Y$ on day $t$ in class $j$ are summarized daily with the mean across these time periods.
-The change in aerosol number and particle matter mass concentrations is then estimated using Bayesian log-linear regression models
-\begin{align*}
-    \log Y | \beta_0, \beta_1, \dots, \beta_6 &\sim \text{Normal}(\mu,\sigma) \\
-    \mu &= \beta_0 + \beta_1 \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Weekday} + \beta_4 \cdot \text{Students} \\
-    &\hphantom{= } + \beta_5 \cdot \text{AirChangeRate} + \beta_6 \cdot \text{Cases} \\
-    \beta_0 &\sim \text{Student-t}_5(\mu_y, 2.5s_{y}) \\
-    \beta_1, \dots, \beta_6 &\sim \text{Student-t}_5\left(0, 2.5\frac{s_{y}}{s_{x}}\right) \\
-    \sigma &\sim \text{Exponential}\left(\frac{1}{s_{y}}\right)~.
-\end{align*}
-The effect of air cleaners is adjusted for class- and weekday-specific effects, the number of students in school, the air change rate, and the cumulative number of cases related to respiratory infections.
-A log transform is applied to all continuous input variables.
-
-\clearpage
-
-\section{Modeling relative risk of infection}\label{sec:transmission-model}
-
-\subsection{Overall approach}
-
-The overall aim is to estimate the effects of air cleaners on the daily number of respiratory infections.
-The latter is unobserved/latent and inferred from the daily number of respiratory cases (absences related to respiratory infections by date of symptom onset), considering the delay from infection to symptom onset (\ie the incubation period).
-The effect of air cleaners is estimated with a Bayesian approach, which requires the specification of prior distributions for all model parameters.
-
-\subsection{Notation}
-
-\begin{tabular}{ll} 
-$j$  & class \\
-$t$ & days since start of study period \\
-$C_{jt}$  & number of  new cases among students in class $j$ at day $t$ (observed) \\
-$I_{jt}$  & number of new infections in class $j$ at day $t$ (unobserved)  \\
-$F_{jt}$  & number of new infections in class $j$ before day $t$ (unobserved)  \\
-$N_{jt}$  & cumulative number of infections in class $j$ until day $t$ (unobserved)  \\
-\end{tabular}
-
-\subsection{Relating the number of respiratory cases to the number of infections}
-
-The number of respiratory cases $C$ in class $j$ at time $t$ is modeled with a Negative Binomial distribution   
-\begin{align*}
-    C_{jt} \sim \text{Negative-Binomial}(\mu_{jt},\phi),
-\end{align*}
-where $\mu_{jt}$ is the expected number of new cases and $\phi$ is the parameter modeling over-dispersion.
-The expected number of new cases is the weighted sum of the number of new infections $I_{jt}$ in the  previous days
-\begin{align*}
-    \mu_{jt} = \sum_{s<t}I_{js} \cdot p_{\text{IN}}(t-s),
-\end{align*}
-where $p_{\text{IN}}(t-s)$ denotes the probability distribution of the incubation period.
-
-\subsection{Relating the number of new infections to the presence of interventions}
-
-The  number of new infections is related to the presence of air cleaners using a log-link
-\begin{align*}
-    \log I_{jt} &= \log F_{jt} - \log N_{jt} + \beta_0 + \beta_1 \cdot \text{AirCleaner}_{jt} + \beta_2 \cdot \text{Class}_t + \beta_3 \cdot \text{Students}_{jt} \\
-    &\hphantom{= }+ \beta_4 \cdot \text{AirChangeRate}_{jt} + \beta_5 \cdot \text{CoV}_t + \beta_6 \cdot \text{ILI}_t,
-\end{align*}
-where $F_{jt} = \sum_{t-7}^{t-1} I_{js}$ is the number of infections in the previous seven days (first model offset; a proxy for the number of contagious students), $N_{jt} = \sum_{s<t} I_{js}$ is the cumulative number of infections (second model offset; the inverse is a proxy for the number of susceptible students), $\beta_0$ is the rate of new infections without air cleaners (model intercept), and $\beta_1$ is the effect of air cleaners.
-The effect of air cleaners is adjusted for class-specific effects, the number of students in school, the air change rate, the proportion of positive tests for SARS-CoV-2 in the community, and the number of consultations for influenza-like illnesses in the community.
-
-Note that we deviated from the statistical analysis plan by using as model offset $\log F_{jt} - \log N_{jt}$ instead of just $\log N_{jt}$.
-This change improved the fit of our model.
-The reason is that the ratio $\log (F_{jt} / N_{jt})$ incorporates both ways by which new infections can naturally decrease over time, \ie a decrease in contagious students ($F_{jt}$) or a decrease in susceptible students ($1 / N_{jt}$).
-
-\subsection{Specifying the distribution of the incubation period}
-
-The virus of each respiratory infection could not be identified from the epidemiological data because the students never obtained a laboratory test result.
-As a consequence, different incubation periods need to be considered in $p_{\text{IN}}$, reflecting a combination of the virus-specific incubation periods.
-The combination is determined based on the weekly proportion of positive saliva samples for each virus found in the molecular analysis.
-Formally, let $p_{\text{IN}}^{v}$ be the distribution for the incubation period of respiratory virus $v$ and let $pp_{vw}$ be the proportion of positive saliva samples in study week $w$, then each week the combined incubation period is computed as the weighted sum of the virus-specific incubation periods
-\begin{align}
-    p_{\text{IN}}^w(s) = \sum_{v} p_{\text{IN}}^{v}(s) \cdot pp_{vw} \quad \forall s.
-\end{align}
-
-The prior distributions for the virus-specific incubation periods are based on estimates published in the literature\cite{McAloon2020,Lessler2009LancetID}.
-The distributions are shown in \Cref{fig:prior-pin}.
-Since we could not obtain prior estimates for the incubation period of metapneumovirus (MPV) from the literature, we instead formed a distribution by using the equally weighted average of the parameters from the other distributions.
-
-We estimate the virus-specific distributions from our data as part of fitting the overall model.
-Furthermore, note that $p_\text{IN}$ is discretized via $p_\text{IN}(s) = \int_0^{0.5} p_\text{IN}(\tau) \;\text{d}\tau$ for $s = 0$ and $p_\text{IN}(s) = \int_{s-0.5}^{s+0.5} p_\text{IN}(\tau) \;\text{d}\tau$ for $s > 0$, where $p_\text{IN}(\tau) \sim \textrm{Lognormal}(\mu, \sigma)$ is the density of the lognormal distribution.
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{../../results/epi-data/incubation-periods.pdf}
-    \caption[Choices of priors for the incubation periods]{Choices of priors for the distribution of the virus-specific incubation periods.
-IFB: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
-    \label{fig:prior-pin}
-\end{figure}
-
-\subsection{Adjusting for under-reporting of cases on weekends}
-
-Despite recording cases by date of symptom onset, we recorded a higher proportion of cases on Mondays than on weekends, suggesting recall bias and under-reporting of cases on weekends.
-To consider weekday effects in the reporting of cases, we re-weight the expected number of cases each week as follows.
-Let $k \in $(1: Saturday, 2: Sunday, 3: Monday, $\dots$, 7: Friday) denote the weekday with the week starting on Saturday.
-The re-weighted expected number of cases $\Tilde{\mu}$ (class and day indexes omitted) are computed as
-\begin{align*}
-    \Tilde{\mu}_{k} &= \mu_k \cdot \vartheta_k \cdot \left(\frac{\sum_k \mu_k}{\sum_k \nu_k \cdot \mu_k}\right)\\
-    \sum_k \vartheta_k &= 1,
-\end{align*}
-where $\nu_k$ is the weight for weekday $k$. These weights are modeled with a Dirichlet prior
-\begin{align*}
-    \vartheta &\sim \text{Dirichlet}(c) \\
-    c_k &= \sum_j \sum_t C_{jt} \mathbb{I}_{\text{Weekday}(t)=k} \quad \forall k
-\end{align*}
-where $c_k$ is the total number of cases reported for weekday $k$ and $\mathbb{I}$ is a binary indicator.
-
-\subsection{Modeling school-free days}
-
-Infections may have occurred during the week of vacation that falls into the study period.
-The expected number of infections and cases are computed during vacation, but vacation days are not modeled (\ie not incorporated into the model likelihood).
-In addition, we assume lower transmission of respiratory infections on days without school (weekends and vacations).
-We incorporate our prior belief into the model intercept $\beta_0$
-\begin{align*}
-    \beta_0 &= \alpha + \omega \cdot \text{NoSchool},
-\end{align*}
-where $\alpha$ is the rate of new infections on school days and $\alpha + \omega$ is the rate on days without school.
-We model $\omega$ with an informative prior for a 10\% decrease in new infections on school-free days
-\begin{align*}
-    \omega &\sim \text{Normal}(\log 1.1, 0.05)~.
-\end{align*}
-
-\subsection{Seeding infections before study start}
-
-Cases in the first week of the study could indicate infections before the study commenced.
-We will therefore seed our model $2 \cdot \text{m}\,$days before the study start, where $\text{m}$ is the average incubation period of the virus with the largest incubation period (\ie adenovirus).
-The number of infections before the study start will be modeled with an exponential prior
-\begin{align*}
-    I_{jt} \sim \text{Exponential}(1) \quad t = -2\cdot\text{m}+1, \dots, 0~. 
-\end{align*}
-
-Note that we deviated from the statistical analysis plan by changing the parameter of the Exponential distribution from $\lambda = 2 \cdot m$ to $\lambda = 1$.
-After inspecting the model fit, we realized that the model could not adequately fit the number of new cases at the start of study, because the expected number of new infections in the seeding period $\mu = 1 / \lambda = 1 / (2 \cdot m)$ were too low.
-After increasing the expected number of new infections in the seeding period to $\mu = 1$ the model could adequately capture new respiratory cases observed at the study start.
-
-\subsection{Priors for modeling parameters}
-
-We use weakly informative priors for all modeling parameters following recommendations on the choice of priors\cite{Gelman2008StatMed,Gelman2008StatAnnals,Gelman2020RegOther,Stan2020Priors,Gabry2023Priors}.
-The continuous adjustment variables are standardized to have zero mean and a standard deviation of 0.5.
-\begin{align*}
-    \alpha &\sim \text{Student-t}_5(0, 10) \\
-    \beta_2, \dots, \beta_6 &\sim \text{Student-t}_5(0, 2.5) \\
-    \frac{1}{\sqrt{\phi}} &\sim \text{Half-Normal}(0,1),
-\end{align*}
-where Student-t$_5$ is a Student-t distribution with 5 degrees of freedom.
-
-\clearpage
-
-\section{Modeling changes in cough frequency}\label{sec:aud-regression-model}
-
-We analyze the daily number of coughs $Y$ with a Negative Binomial regression model
-\begin{align*}
-    Y | \beta_0, \beta_1, \dots, \beta_6 &\sim \text{Negative-Binomial}(\mu,\phi) \\
-    \log \mu &= \log T + \beta_0 + \beta_1 \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Weekday} + \beta_4 \cdot \text{Students} + \beta_5 \cdot \text{Ventilation} + \beta_6 \cdot \text{Cases} \\
-    \beta_0 &\sim \text{Student-t}_5(0, 2.5) \\
-    \beta_1, \dots, \beta_6 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) \\
-    \frac{1}{\sqrt{\phi}} &\sim \text{Half-Normal}(0,1),
-\end{align*}
-where $T$ is the daily duration that students were in the classroom and $\phi$ is the parameter modeling over-dispersion in count data.
-
-In addition, we analyze the association between the daily number of coughs $Y$ and the number of positive saliva test resutls for respiratory viruses with a Negative Binomial hierarchical regression model
-\begin{align*}
-    Y | \beta_0, \theta_{[v]}, \theta, \tau &\sim \text{Negative-Binomial}(\mu,\phi) \\
-    \log \mu &= \log T + \beta_0 + \sum_v \theta_{[v]} \cdot \text{Saliva}_v \\
-    \beta_0 &\sim \text{Student-t}_5(0, 2.5) \\
-    \theta_{[\text{IFB}]}, \dots, \theta_{[\text{PIV}]} &\sim \text{Normal}\left(\theta, \tau\right) \\
-    \theta &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{v}}\right) \\
-    \tau &\sim \text{Half-Normal}\left(0, 1\right) \\
-    \frac{1}{\sqrt{\phi}} &\sim \text{Half-Normal}(0,1),
-\end{align*}
-where $\theta_{[v]}$ is the partially pooled estimate for respiratory virus $v \in$ (IFB, HRV, AdV, CoV, MPV, PIV), $\tau$ estimates variation between viruses, and $s_v$ is the average standard deviation across the count variables of the respiratory viruses.
-
-\section{Modeling positivity rate of human saliva samples}\label{sec:multinomial-model}
-
-Molecular analysis determined which saliva samples were positive for a respiratory virus.
-Let $p = 0, 1, \dots, P$ denote the virus, where $0$ refers to negative samples and $P$ is the number of different viruses detected over the study.
-The number of positive bioaerosol and saliva samples $y_p$ in samples of sizes $n$ is analyzed with a Multinomial logistic regression model
-\begin{align*}
-    y_0, y_1, \cdots, y_P | n, \beta_0, \beta_1, \beta_2, \beta_3 &\sim \text{Multinomial-Logit}(\theta_0, \theta_1, \dots \theta_P) \\
-    \theta &= \text{softmax}(\mu) \\
-    \mu &= \begin{cases}
-                0, & \text{if } p=0, \\
-                \beta_{0p} + \beta_{1} \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Susceptibles}_p, & \text{if } p>0.
-            \end{cases} \\
-    \beta_{0p} &\sim \text{Normal}(0, 2.5) \quad \forall p \\
-    \beta_1, \beta_2, \beta_3 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) 
-\end{align*}
-where $\text{softmax}(\mu) = \exp(\mu)/\sum\exp(\mu)$ and $s_{x}$ is the empirical standard deviation of each input variable.
-The negative test is set as the reference category.
-Overall variation in positive samples by virus will be modeled with $\beta_{0p}$ and the effect of air cleaners will be modeled with $\beta_1$.
-The effect of air cleaners is adjusted for class-specific effects and the decreasing number of susceptibles over time (for each virus computed as the number of students minus the total number of positive samples).
-
-Note that we deviated from the statistical analysis plan by estimating $\beta_{0p}$ and $\beta_1$ without a hierarchical prior.
-First, there was considerable variation in the overall positivity rate of each virus, so that we decided to use unpooled intercepts $\beta_{0p}$ instead of partially pooled intercepts $\beta_{0[p]}$ for each virus.
-Second, there was insufficient variation in the data to inform partially pooled estimates $\beta_{1[p]}$ for the effects of air cleaners.
-Therefore, we decided to only estimate the average effect across viruses, \ie the completely pooled estimate $\beta_1$.
-
-\clearpage
-
-\section{Detailed results for changes in particle concentrations}\label{sec:detailed-palas}
-
-There was a strong difference in particle concentrations between study conditions (\Cref{fig:palas-results}a).
-When adjusting for air change rates and multiple other factors, the aerosol number concentration decreased by 76\% (95\%-CrI 63\% to 86\%) with air cleaners (\Cref{fig:palas-results}b and \Cref{tab:palas-est-results}).
-The decrease in the concentration of larger particles (PM$_{10}$) was greater than the decrease in the concentration of smaller particles (PM$_1$ to PM$_{4}$).
-
-\begin{figure}[!htpb]
-\centering
-    \includegraphics[width=\linewidth]{../../results/env-data/particles-boxplot.pdf}
-    \includegraphics[width=\linewidth]{../../results/env-data/estimation-results-figure.pdf}
-    \caption[Analysis of particle concentrations and comparison between study conditions]{\textbf{Analysis of particle concentrations and comparison between study conditions}.
-\textbf{(a)}~Boxplot of the daily average values for aerosol number concentration (CN in 1/cm$^3$) and particle mass concentration (PM for particles of sizes <1 to <10~$\mu$m, respectively in $\mu$gm$^{-3}$).
-\textbf{(b)}~Estimated reduction in aerosol number and particle mass concentrations with air cleaners (posterior mean as dot and 50\%-, 80\%- and 95\%-CrI as lines, respectively).
-}
-    \label{fig:palas-results}
-\end{figure}
-
-\begin{table}[!htpb]
-    \caption[Estimated reduction in aerosol and particle concentrations with interventions]{Estimated reduction in aerosol number (CN) and particle mass (PM) concentrations with interventions (posterior mean and upper and lower estimate from the 95\%-CrI).}
-    \label{tab:palas-est-results}
-    \centering
-    \footnotesize
-    \begin{tabular}{l r r r}
-    \toprule
-    Variable & Mean & Lower & Upper \\
-    \midrule
-    \input ../../results/env-data/estimation-results-table.tex
-    \bottomrule
-    \end{tabular}
-\end{table}
-
-In the main paper, we compared particle concentrations between study conditions.
-\Cref{fig:env-descriptives-other-vars} compares additional environmental variables, \ie CO$_2$, air change rate, relative humidity, and temperature.
-The air change rate was estimated from indoor CO$_2$ levels using a transient mass balance model\cite{Batterman2017IJERPH}.
-There are no considerable differences between in any of the variables between study conditions.
-Furthermore, numerical estimation results for the reduction in particle concentrations with air cleaners are shown in \Cref{tab:palas-est-results}.
-
-\begin{figure}[!htb]
-\centering
-    \includegraphics[width=\linewidth]{../../results/env-data/otherVars-boxplot.pdf}
-    \caption[Boxplot of environmental variables by study condition]{Boxplot for the daily average values of each environmental variable by study condition.}
-    \label{fig:env-descriptives-other-vars}
-\end{figure}
-
-\clearpage
-
-\section{Detailed results for risk of infection model}\label{sec:detailed-redcap}
-
-Table~\ref{tab:estimation-results} presents the posterior mean, credible intervals and model diagnostics for all model parameters.
-See the main paper for a discussion of the effects of interventions.
-Here we briefly discuss some of the additional model parameters.
-
-The effective sample size (ESS) and the Gelman-Rubin convergence diagnostic ($\hat{R}$) indicate good estimation power. It further suggests that the Markov chains converged.
-
-The estimate for $\beta_2$ is positive, suggesting a higher risk of infection in class $B$, which is in line with the higher number of cases in this class.
-The estimate for $\beta_3$ is positive, suggesting a higher risk of infection as more students were in class.
-The estimate for $\beta_4$ is negative, suggesting a lower risk of infection with higher air change rates, in line with the intuition that the risk of infection is lower when the classroom is better ventilated.
-The estimate for $\beta_5$ and $\beta_6$ are both positive although only the former is distinguishable from zero, suggesting that higher transmission in the community was also associated with a higher risk of infection in the school.
-
-The overdispersion parameter ($\phi$) cannot be precisely estimated, but the credible interval suggests rather small overdispersion ($\phi \rightarrow \infty$).
-The school-free effect $\omega$, the weekday weights $\vartheta$, and the parameters for the distribution of the incubation period $\mu_{p_{\text{IN}}}$ and $\sigma_{p_{\text{IN}}}$ are all close to their priors, suggest that these parameters are barely informed by date but more by our prior assumptions.
-
-\begin{table}[!htpb]
-    \caption[Estimation results from infection risk model]{Estimation results from infection risk model for the number of new respiratory cases.}
-    \label{tab:estimation-results}
-    \footnotesize
-    \centering
-    \begin{tabular}{lrrrrr}
-    \toprule
-    Parameter & Mean & Lower 95\%-CrI & Upper 95\%-CrI & $\hat{R}$ & ESS \\
-    \midrule
-    \input ../../results/epi-data/estimation-results
-    \bottomrule
-    \multicolumn{6}{p{13cm}}{\scriptsize
-        ESS is the effective sample size, \ie the number of independent MCMC samples with estimation power equivalent to the total number of autocorrelated samples\cite{Stan2022}, and $\hat{R}$ is the Gelman-Rubin convergence diagnostic\cite{Gelman1992}.
-Low ESS or $\hat{R}$ or $\hat{R}>1.10$ indicate bad convergence of the model\cite{Gelman2013}.}
-    \end{tabular}
-\end{table}
-
-Our model accounts for the delay from infection to case confirmation and allows transmission to change only at the dates of interventions.
-It thus reflects overall trends in transmission during study conditions rather than day-to-day variation.
-To evaluate how well our model fits these trends, \Cref{fig:coverage} compares the model-estimated expected number of cases  with the observed number cases only on a weekly basis and across classes.
-Overall the estimates are in relatively good agreement.
-The 95\%-CrIs of the model-estimates always include the observed cases.
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{../../results/epi-data/model-fit.pdf}
-    \caption[Comparison of model-estimates with observed case data]{Model-estimated number of new respiratory cases (posterior mean as red line and 50\%-, 80\%, and 95\%-CrIs as shaded areas) and the observed number of new respiratory cases (black line) across classes by study week.}
-    \label{fig:coverage}
-\end{figure}
-
-% Our estimation results imply a small, potential benefit of air cleaners.
-Subject to considerable uncertainty, \Cref{fig:avoided-infections} shows the estimated number number of infections that would have occurred if air cleaners had been installed throughout the study (scenario~1) or not been installed (scenario~2).
-
-% \begin{figure}[!htpb]
-%     \includegraphics{../../results/epi-data/avoided-infections.pdf}
-%     \caption[Estimated number of infections with and without air cleaners]{Estimated mean number of infections (posterior mean as dot and 50\%, 80\%, and 95\%-CrI as shaded areas) if, in both classes, air cleaners had been installed throughout the study or not been installed.}
-%     \label{fig:avoided-infections}
-% \end{figure}
-
-\clearpage
-
-\section{Detailed results for cough frequency}\label{sec:detailed-results-coughing}
-
-Differences in the frequency of coughs between classes suggest an association with class-specific virus patterns.
-\Cref{fig:coughing-association} shows the relative risk of cough for each respiratory virus.
-Cough appeared more frequent for IFB and less frequent for AdV.
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{results/cough-data/coughs-virus-association.pdf}
-    \caption[Association between cough frequency and virus-specific positivity rate]{Estimated association (relative risk; posterior mean as dot, 50\%, 80\%, and 95\%-CrI as lines) between cough frequency and number of positive saliva test results by respiratory virus.
-IFB: Influenza B, HRV: rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: metapneumovirus, PIV: parainfluenza virus.}
-    \label{fig:coughing-association}
-\end{figure}
-
-\clearpage
-
-\section{Detailed results for positivity rate in saliva samples}
-\label{sec:detailed-molecular}
-
-There can be unknown delay between when students were infected and when they tested positive.
-Therefore, we performed a sensitivity analysis where we lagged the molecular test results (number of positive saliva samples) by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday).
-The results are shown in \Cref{fig:mol-estimation-results-sensitivity}.
-The estimated effect of air cleaners remains indistinguishable from zero regardless of the considered lag.
-\begin{figure}[!htb]
-\centering
-    \includegraphics[width=10cm]{../../results/mol-data/model-results.pdf}
-    \caption[Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples]{Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples.
-Shown is the adjusted relative risk (posterior mean as dot and 50\%-, 80\%, and 95\%-CrIs as lines) for air cleaners by lag of the molecular test results.}
-    \label{fig:mol-estimation-results-sensitivity}
-\end{figure}
-
-\clearpage
-
-\bibliography{references.bib}
-
 \end{document}

### 2024-01-03 analysis/study_comaprison.Rmd
cosmetic figure changes

@@
   filter(n > 0) %>%
-  mutate(year = ifelse(year_num == 2022, "Winter 2021/2022", "Winter 2022/2023"),
-         year = factor(year, levels = c("Winter 2021/2022", "Winter 2022/2023")),
+  mutate(year = ifelse(year_num == 2022, "2021/22", "2022/23"),
+         year = factor(year, levels = c("2021/22", "2022/23")),
          virus = factor(virus, levels = c("IF", "HRV", "AdV", "CoV", "MPV", "PIV"))) %>%
@@
   ungroup() %>%
-  mutate(labelPosition = ifelse(year == "Winter 2022/2023" & virus == "MPV", labelPosition - .015, labelPosition),
-         labelPosition = ifelse(year == "Winter 2021/2022" & virus == "IF", ymin, labelPosition),
-         labelPosition = ifelse(year == "Winter 2021/2022" & virus == "AdV", ymax, labelPosition))
+  mutate(labelPosition = ifelse(year == "2022/23" & virus == "MPV", labelPosition - .015, labelPosition),
+         labelPosition = ifelse(year == "2021/22" & virus == "IF", ymin, labelPosition),
+         labelPosition = ifelse(year == "2021/22" & virus == "AdV", ymax, labelPosition))
 
@@
   ggplot() +
-  facet_wrap(~ year_num, nrow = 1) +
+  facet_wrap(~ year, nrow = 1) +
   geom_rect(mapping = aes(ymax = ymax, ymin = ymin, xmax = 4, xmin = 3, fill = virus)) +
@@
   geom_text(mapping = aes(x = 5.25, y=labelPosition, label=label, color=virus), size = 8/cm(1)) +
-  geom_text(data = data.frame(x = 1, y = 1, year_num = c(2022, 2023), label = c(2022, 2023)), mapping = aes(x = x, label = label, y = y), 
+  geom_text(data = data.frame(x = 1, y = 1, year = c("2021/22", "2022/23"), label = c("2021/22", "2022/23")), mapping = aes(x = x, label = label, y = y), 
             size = 8/cm(1), fontface = 2) +
@@
          type_num = ifelse(type == "Bioaerosol", 1, 2),
-         classroom = factor(paste("Room", class), levels = c("Room A", "Room B")))
+         classroom = ifelse(class == "A", "Classroom 1", "Classroom 2"),
+         classroom = factor(classroom, levels = c("Classroom 1", "Classroom 2")))
 

### 2024-01-03 doc/comparison-mcid12/manuscript.tex
Updates from Overleaf

@@
 
-\title{\bfseries\LARGE\singlespacing{Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols}}
+\title{\bfseries\LARGE\singlespacing{Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva, air, and filter samples from schools during and after the pandemic}}
 
@@
 \author[1,2]{Nicolas Banholzer}
+\author[2,4]{Pascal Bittel}
 \author[2,3]{Philipp Jent}
-\author[2,4]{Pascal Bittel}
+\author[4]{Lavinia Furrer}
 \author[1]{Kathrin Zürcher}
-\author[4]{Lavinia Furrer}
-\author[1]{Simon Bertschinger}
-\author[5]{Ernest Weingartner}
-\author[2,4]{Alban Ramette}
+%\author[1]{Simon Bertschinger}
+%\author[5]{Ernest Weingartner}
+%\author[2,4]{Alban Ramette}
 \author[1,6,7]{Matthias Egger}
@@
 \affil[4]{Institute for Infectious Diseases, University of Bern, Bern, Switzerland}
-\affil[5]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
-\affil[6]{Population Health Sciences, University of Bristol, Bristol, UK}
-\affil[7]{Centre for Infectious Disease Epidemiology and Research, University of Cape Town, Cape Town, South Africa}
-\affil[8]{Institute of Educational Science, University of Bern, Bern, Switzerland}
+%\affil[5]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
+\affil[5]{Population Health Sciences, University of Bristol, Bristol, UK}
+\affil[6]{Centre for Infectious Disease Epidemiology and Research, University of Cape Town, Cape Town, South Africa}
+\affil[7]{Institute of Educational Science, University of Bern, Bern, Switzerland}
 
@@
 \begin{information}\normalfont
-\textbf{Word count:} Abstract 219 (max 250), Manuscript 1,380 (max 1,200) \\
+\textbf{Word count:} Abstract 248 (max 250), Manuscript 1,187 (max 1,200) \\
 \textbf{Display items:} Figure 1 (max 2) \\
-\textbf{References:} 15 (max 15) \\
+\textbf{References:} 14 (max 15) \\
 %\textbf{Supporting information:} STROBE checklist, Supplementary Data with texts, tables, and figures
@@
 \begin{abstract}\normalfont
-\noindent \textbf{Objectives:} To compare molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols.
+\noindent \textbf{Objectives:} To compare the prevalence of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols between two winters and model the probability of virus detection in classroom air for different viruses.
 
-\noindent \textbf{Methods:} We analyze saliva and bioaerosol samples from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave) and winter 2022/2023 (after the COVID-19 pandemic) in a Swiss school setting.
-We estimate the probability of weekly airborne detection of SARS-CoV-2 vs non-SARS-CoV-2 viruses using a Bayesian logistic regression model.
+\noindent \textbf{Methods:} We analyze saliva, air, and air cleaner filter samples from studies conducted in two  Swiss secondary schools (age 14-17 years) over seven weeks during the winters of 2021/22 and 2022/23.
+Two bioaerosol sampling devices and HEPA filters from air cleaners were used to collect airborne virus particles in five classrooms.
+Daily bioaerosol samples were pooled for each sampling device before PCR analysis of a panel of 19 respiratory viruses and viral subtypes.
+The probability of detection of airborne viruses was modelled using a Bayesian logistic regression model.
 
-\noindent \textbf{Results:} While during the pandemic SARS-CoV-2 was almost exclusively detected in saliva (19~samples of SARS-CoV-2 vs 2~samples of non-SARS-CoV-2), the most common respiratory viruses after the pandemic were adenovirus, influenza, and rhinoviruses (3~samples of SARS-CoV-2 vs 47~samples of non-SARS-CoV-2).
-Despite the same study setting and methods used, airborne detection of SARS-CoV-2 was relatively more frequent than of non-SARS-CoV-2 viruses (10~samples of SARS-CoV-2 in 2021/2022 vs 2~samples of non-SARS-CoV-2 in 2022/2023).
-Based on our model, the posterior probability of weekly airborne detection was 31\% (95\%-credible interval [CrI] 13\%$-$54\%) for SARS-CoV-2 vs 11\% (95\%-CrI 4\%-23\%) for non-SARS-CoV-2.
+\noindent \textbf{Results:} Three classes (58 students) participated in 2021/22, and two classes (38 students) in 2022/23 .
+During winter 2021/22, SARS-CoV-2 dominated in saliva (19 of 21 positive samples) and bioaerosols (9 of 10).
+One year later, there were 50 positive saliva samples, mostly influenza B, rhinovirus, and adenovirus, and two positive bioaerosol samples, one rhinovirus and one adenovirus.
+The probability of weekly airborne detection was 34\% (95\%-credible interval [CrI] 22\%$-$47\%) for SARS-CoV-2 and 10\% (95\%-CrI 5\%-16\%) for other respiratory viruses.
 
-\noindent \textbf{Discussion:} There was a clear shift in the distribution of respiratory viruses from predominantely SARS-CoV-2 in 2022 to other respiratory viruses in 2023.
-Our comparison suggests that SARS-CoV-2 is easier to detect in the air than other endemic respiratory viruses, possibly facilitating long-range transmission.
-Future experimental studies are needed to confirm this finding.
+\noindent \textbf{Conclusions:} There was a distinct shift in the distribution of respiratory viruses from SARS-CoV-2 during the Omicron wave to other respiratory viruses one year later.
+SARS-CoV-2 is more likely to be detected in the air than other endemic respiratory viruses, possibly reflecting viral characteristics that facilitate airborne long-range transmission.
 \medskip
@@
 
-The transmission of respiratory viruses, such as SARS-CoV-2 and influenza, within schools and and other indoor environments is difficult to control\cite{Leung2020}.
+The transmission of respiratory viruses, such as SARS-CoV-2 and influenza, in schools and other indoor environments is difficult to control\cite{Leung2020NatMed}.
+During the COVID-19 pandemic, non-pharmaceutical interventions and physical distancing reduced the spread of SARS-CoV-2 and other seasonal respiratory viruses, but a resurgence of respiratory cases followed the relaxation of these measures\cite{Poole2020LancetRespMed,Sauteur2022EuroSurv,Kandeel2023BMC}.
+Following epidemic peaks, a shift in the circulation of respiratory viruses occurs\cite{Pierangeli2012CMI}, which can be identified by frequent collection of non-invasive saliva samples\cite{Pasomsub2021CMI}.
+
 Respiratory viruses spread by multiple routes, including respiratory particles such as large droplets and small aerosols.
 Unlike larger droplets, which settle quickly, aerosols can remain suspended in the air for extended periods\cite{Wang2021}.
-Airborne transmission is probably significant, but the relative contribution to overall transmission is still unclear, particularly because detection of respiratory viruses in the air remains challenging\cite{Belser2023PLOSPath}.
+Airborne infectious pathogens are primarily found in smaller particles (<10$\mu$m) and the distribution is similar across varius pathogens\cite{Fennelly2020}.
+Thus, pathogen-carrying aerosols have the potential for long-range transmission, but the larger concentration of particles near the infectious person favours short-range transmission\cite{Wang2020}.
 
-Airborne infectious pathogens are primarily found in smaller particles (<10$\mu$m) and the distribution is similar across a range of different pathogens\cite{Fennelly2020}.
-Thus, pathogen-carrying aerosols have the potential for long-range transmission, but the larger concentration of particles near the infectious person favours short-range transmission\cite{Tellier2009JTRSI,Wang2020}.
-The complexity of airborne transmission lies in the interplay between the physicochemical properties of pathogen-carrying aerosols and environmental conditions\cite{Wang2021}, making it challenging to devise effective public health strategies to mitigate the spread of respiratory infections in various indoor settings, including schools, offices, and hospitals.
+%The complexity of airborne transmission lies in the interplay between the physicochemical properties of pathogen-carrying aerosols and environmental conditions\cite{Wang2021}, making it challenging to devise effective public health strategies to mitigate the spread of respiratory infections in various indoor settings, including schools, offices, and hospitals.
 
-In this study, we compared saliva samples, bioaerosol samples, and samples from the HEPA filters of air cleaners that were collected as part of two studies that we conducted in a Swiss school setting in winter 2021/2022 (during the SARS-CoV-2 omicron wave, henceforth 2022)\cite{Banholzer2023PLoSMed} and winter 2022/2023 (after the COVID-19 pandemic, henceforth 2023)\cite{Banholzer2023submitted}.
-We show differences in the distribution of respiratory viruses and analyze the results from airborne molecular detection.
+We compared saliva samples, bioaerosol samples, and samples from the HEPA-filters of air cleaners that were collected as part of two studies conducted in a Swiss school setting in winter 2021/22 (during the SARS-CoV-2 omicron wave)\cite{Banholzer2023PLoSMed} and winter 2022/23\cite{Banholzer2023medRxiv}.
 
@@
 
-Data was collected in two Swiss secondary schools (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from end of January to beginning of March.
-Three classes participated in 2022 and two classes in 2023.
-Two classes shared the same classroom in 2022 due to half-class teaching, other classes had a separate classroom.
-All classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system, but were ventilated using passive window ventilation.
+Data were collected in two secondary schools (students age 14-17~years) in the canton of Solothurn, Switzerland, during a seven-week study period from the end of January to the beginning of March.
+Three classes (two classrooms) participated in 2021/22 and two classes (two classrooms) in 2022/23.
 An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, temperature, and humidity.
@@
 
-Repetitive testing for a panel of respiratory infections was done weekly in 2022 and bi-weekly in 2023.
-Airborne respiratory viruses were collected in each classroom with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France) and a BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft.
+Testing for a panel of respiratory infections was performed weekly in 2021/22 and bi-weekly in 2022/23.
+Airborne respiratory viruses were collected in each classroom with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France) and the BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft.
 Collins, CO, USA)\cite{Lednicky2016AST}.
-The removable parts were regularly autoclaved.
-Finally, swabs from the air cleaners' HEPA filters were collected.
-The HEPA filters were removed and divided into 20~fields.
-One sterile Phosphate-Buffered Saline-moistened swab per field was then taken for a total of 20~swabs per filter.
-Saliva and airborne samples were transported to the Institute of Infectious Diseases and stored at $-$80°C until further processing\cite{Huber2021}.
-Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL.
+The HEPA filters from the portable air cleaner (Xiaomi Mi Air Pro 70m2, Shenzhen, China ) were removed and divided into 20~fields.
+One sterile Phosphate-Buffered Saline-moistened swab per field was then collected for a total of 20~swabs per filter.
+Saliva and airborne samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Huber2021}.
+Before real-time (RT)-PCR analysis, daily bioaerosol samples were pooled for each sampling device and filtered using Amicon Ultra-15 Centrifugal filters\cite{Banholzer2023PLoSMed}.
 Saliva samples were analyzed directly without prior filtration.
@@
 
-Descriptive statistics were used to present differences in the type and number of respiratory viruses detected in saliva and airborne samples between 2022 and 2023.
-A Bayesian logistic regression model was used to estimate the probability of detecting any SARS-CoV-2 vs non-SARS-CoV-2 in the air during a study week, adjusting for differences in the study settings and the interventions that were in place during the studies (see \supp~Text~\zref{sec:model} for a detailed model description).
+We used descriptive statistics to present differences in the type and number of respiratory viruses detected in saliva and airborne samples between 2021/22 and 2022/23.
+A Bayesian logistic regression model was used to estimate the probability of detecting any SARS-CoV-2 versus non-SARS-CoV-2 in the air during a study week, adjusting for differences in the study settings and the interventions implemented during the studies (see \supp~Text~\zref{sec:model} for a detailed model description).
 
@@
 
-In 2022, 58~students participated in weekly saliva testing.
-There were 21~positive saliva samples over the study, namely 19~SARS-CoV-2, 1~influenza~A, and 1~adenovirus (\Cref{fig:comparison}a, left).
-There were 10~positive bioaerosol samples, namely 9~SARS-CoV-2 and 1~adenovirus.
-There were 8~positive samples on the HEPA-filters of the air cleaners, namely 6~SARS-CoV-2, 1~influenza~A and 1~adenovirus.
-In 2023, 38~students participated in bi-weekly saliva testing.
-There were 50~positive saliva samples over the study, namely 15~influenza~B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, 1~parainfluenza virus (\Cref{fig:comparison}a, right).
-There were 2~positive bioaerosol samples, namely 1~rhinovirus and 1~adenovirus.
-There were 4~positive samples on the HEPA-filters of the air cleaners, namely 1~influenza~B, 1~rhinovirus, 1~adenovirus, and 1~SARS-CoV-2.
+In 2021/22, 58~students participated in weekly saliva testing.
+There were 21~positive saliva samples during the study, 19~SARS-CoV-2, one~influenza~A, and one~adenovirus (\Cref{fig:comparison}a, left).
+There were 10~positive bioaerosol samples, nine~SARS-CoV-2 and one~adenovirus.
+There were eight~positive samples on the HEPA-filters, six~SARS-CoV-2, one~influenza~A and one~adenovirus.
+In 2022/23, 38~students participated in bi-weekly saliva testing.
+There were 50~positive saliva samples, mostly influenza~B, rhinovirus, and adenovirus (\Cref{fig:comparison}a, right).
+There were two~positive bioaerosol samples, one~rhinovirus and one~adenovirus.
+There were four~positive samples on the HEPA-filters of the air cleaners, one~influenza~B, one~rhinovirus, one~adenovirus, and one~SARS-CoV-2.
+We found six positive air-saliva samples in the same week (4~SARS-CoV-2 and two~non-SARS-CoV-2 (\Cref{fig:comparison}b), suggesting they were paired samples.
 
-We found 6 positive air-saliva samples in the same respective week (4~SARS-CoV-2 and 2~non-SARS-CoV-2 (\Cref{fig:comparison}b), suggesting they were paired samples.
-Based on our Bayesian logistic regression model, SARS-CoV-2 was more likely detected in bioaerosols than non-SARS-CoV-2 (adjusted odds ratio 3.6, 95\%-CrI 1.5$-$8.9).
-The probability of molecular airborne detection was 31\% (95\%-credible interval [CrI] 13\%$-$54\%) for SARS-CoV-2 vs 11\% (95\%-CrI 4\%-23\%) for non-SARS-CoV-2 (\Cref{fig:comparison}c).
-Our estimates are adjusted for differences in the daily maximum CO$_2$ level, which was 1,134\,ppm (standard deviation [SD] 277\,pppm) in 2022 compared to 2,224\,ppm (SD 321\,pppm) in 2023.
-Humidity and temperature were similar, with daily mean relative humidity of 38\% (SD 6\%) in 2022 vs 38\% (SD 5\%) in 2023 and daily mean temperature of 19°C (SD 2°C) in 2022 vs 22°C (SD 1°C) in 2023.
+SARS-CoV-2 was more likely detected in bioaerosols than non-SARS-CoV-2 (adjusted odds ratio 4.8, 95\%-CrI 2.6$-$9.0).
+The probability of airborne molecular detection was 34\% (95\%-credible interval [CrI] 22\%$-$47\%) for SARS-CoV-2 versus 10\% (95\%-CrI 5\%-16\%) for non-SARS-CoV-2 (\Cref{fig:comparison}c).
+We adjusted estimates for differences in maximum daily CO$_2$, which increased from 1,134\,ppm (standard deviation [SD] 277\,pppm) in 2021/22 to 2,224\,ppm (SD 321\,pppm) in 2022/23.
+Relative humidity and temperature were similar at around 38\% and 20°C, respectively.
 
@@
     \includegraphics{results/comparison.pdf}
-    \caption{Comparison of molecular detection of respiratory viruses between winter 2021/2022 (2022) and winter 2022/2023 (2023).
+    \caption{Comparison of molecular detection of respiratory viruses between winter 2021/22 and winter 2022/23.
 \textbf{(a)}~Distribution of respiratory viruses found in saliva.
@@
 
-We compared molecular data collected from two studies in Swiss secondary schools in winter 2021/2022 (2022) and 2022/2023 (2023).
-We detected predominantly SARS-CoV-2 in 2022 and predominantly non-SARS-CoV-2 viruses, such as influenza and adenovirus, in 2023.
-Molecular airborne detection of SARS-CoV-2 was more likely than of non-SARS-CoV-2.
+We compared the molecular detection of respiratory viruses in saliva, air, and filter samples collected in two studies in Swiss secondary schools in the winter seasons of 2021/22 and 2022/23.
+In winter 2021/22, we predominantly identified SARS-CoV-2 in saliva, air, and air filter samples.
+Conversely, in 2022/23, we primarily detected non-SARS-CoV-2 viruses, such as influenza and adenoviruses, in saliva samples, but these were rarely found in air or filter samples.
 
-The differences between the two study periods (pandemic versus non-pandemic) suggest that SARS-CoV-2 is easier to detect in the air.
-One possible explanation is that SARS-CoV-2 can remain suspended and active in the air for extended periods of time, facilitating long-range transmission.
-In contrast, non-SARS-CoV-2 viruses were rarely detected in the air, suggesting they settle or become inactive more quickly.
-Therefore, prolonged close contact may be relatively more important for the transmission of respiratory viruses other than SARS-CoV-2, although previous studies suggest it also facilitates the transmission of SARS-CoV-2\cite{Leung2020NatMed,Lind2023NatCommun}.
+Overall, the likelihood of molecular airborne detection was substantially higher for SARS-CoV-2 compared to non-SARS-CoV-2 viruses, even when we adjusted for covariates and differences between the studies.
+A plausible explanation is that SARS-CoV-2 can remain airborne for extended durations, thus facilitating long-range transmission.
+In contrast, the infrequent detection of non-SARS-CoV-2 viruses in the air indicates quicker sedimentation or inactivation.
+Therefore, prolonged close contact may be relatively more important for transmission of respiratory viruses other than SARS-CoV-2, although close contact also facilitates transmission of SARS-CoV-2\cite{Leung2020NatMed,Lind2023NatCommun}.
 
-Technical reasons for the differences in airborne detection are unlikely because the bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed.
-Temperature and relative humidity were also similar in both study periods, but the same levels may have virus-specific effects on airborne survival\cite{Tellier2009JTRSI,Davis1971AM}.
-Ventilation changed profoundly, but CO$_2$ levels were even higher in 2023, possibly enhancing airborne survival.
-Additional differences in the study setting (\supp~Table~\zref
-{tab:comp_study}) that could have influenced airborne detection were adjusted for in the Bayesian model (\supp~Text~\zref{sec:model}).
-Therefore, we speculate that the difference in airborne detection is due to the different characteristics of the respiratory viruses which can influence the generation and survival of airborne viral RNA\cite{Wang2021}.
-Low viral loads do often not exceed the detection limits of existing devices for bioaerosol sampling.
-For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}.
-Finally, the ability to detect pathogens in bioaerosols may be related to the variation in infectiousness between study participants, which is associated with the generation of pathogenic aerosols\cite{Leung2020NatMed,Bischoff2013JID}.
+Technical factors are unlikely to account for the differences in airborne detection.
+The two studies employed identical bioaerosol samplers and laboratory methods, and no technical problems occurred.
+Temperature and relative humidity were also similar.
+Ventilation changed, with higher CO$_2$ levels in 2022/23 potentially enhancing airborne survival, but this and other differences were controlled for in the statistical analysis.
+Therefore, it is plausible that the difference in airborne detection may be due to differences in virus characteristics, particularly between SARS-CoV-2 and non-SARS-CoV-2 viruses, which may influence the generation and survival of airborne viral RNA\cite{Wang2021}.
+Non-SARS-CoV-2 respiratory viruses may produce fewer infectious bioaerosols, which can fall below the detection limit of current sampling devices\cite{Belser2023PLOSPath}.
 
-In conclusion, we observed a clear shift in the distribution of respiratory viruses from predominantly SARS-CoV-2 in 2022 to predominantly non-SARS-CoV-2 viruses in 2023.
-Molecular airborne detection of SARS-CoV-2 was more likely than of other endemic respiratory viruses, but future experimental studies are needed to confirm this finding.
+In conclusion, we observed a distinct shift in the distribution of respiratory viruses from SARS-CoV-2 in the winter of 2021/22 to non-SARS-CoV-2 viruses in 2022/23, reflecting the transition from epidemic to endemic transmission of SARS-CoV-2.
+Molecular detection of airborne SARS-CoV-2 was more frequent than other endemic respiratory viruses.
+Future studies should investigate the seasonality of SARS-CoV-2 and non-SARS-CoV-2 respiratory viruses and the contribution of prolonged close contact and contaminated air to the overall transmission of respiratory infections in congregate indoor settings.
 

### 2024-01-03 doc/comparison-mcid12/supplements.tex
Updates from Overleaf

@@
 
-\title{\LARGE\singlespacing{\textbf{Supplmenetary Material} \\ \medskip
-Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols}}
+\title{\LARGE\singlespacing{\textbf{Supplementary Material} \\ \medskip
+Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva, air, and filter samples from schools during and after the pandemic}}
 
@@
 % author list
+% author list
 \author[1,2]{Nicolas Banholzer}
+\author[2,4]{Pascal Bittel}
 \author[2,3]{Philipp Jent}
-\author[2,4]{Pascal Bittel}
+\author[4]{Lavinia Furrer}
 \author[1]{Kathrin Zürcher}
-\author[4]{Lavinia Furrer}
-\author[1]{Simon Bertschinger}
-\author[5]{Ernest Weingartner}
-\author[2,4]{Alban Ramette}
+%\author[1]{Simon Bertschinger}
+%\author[5]{Ernest Weingartner}
+%\author[2,4]{Alban Ramette}
 \author[1,6,7]{Matthias Egger}
@@
         \midrule
-        Participants & 2~classes from School~1 sharing one classroom (half-class teaching) and 1~class from School~2 & two classes from School~2 \\
-        Duration & 7 weeks per class (January 16 to March 11, 2023) & 7 weeks per class (January 24 to March 26, 2023) \\
-        Saliva testing & weekly (Wednesday) & bi-weekly (Tuesday and Thursday) \\
+        Participants & 2~classes from School~1 sharing one classroom (half-class teaching) and 1~class from School~2 with a separate classroom & two classes from School~2 with separate classrooms \\
+        Duration & 7 weeks per class (January 16 to March 11, 2023), excluding vacation & 7 weeks per class (January 24 to March 26, 2023), excluding vacation \\
+        Saliva testing & weekly (Wednesdays) & bi-weekly (Tuesdays and Thursdays) \\
         Bioaerosol sampling & daily with BioSpot-VIVAS and Coriolis in both classrooms & daily with BioSpot-VIVAS and Coriolis in one classroom and only Coriolis in the other classroom \\
-        HEPA-filters sampling & once at the end of the study per filter and classroom & two times per filter and classroom \\
-        Interventions & compulsory face mask wearing as mandated by the public health authorities at that time from the start of the study for a total of 6 study weeks ; and installation of portable air cleaners towards the end of the study in the study classrooms for a total of 4 study weeks & installation of portable air cleaners in the study classrooms using a cross-over design for a total of 7 weeks \\
-        Ventilation conditions & passive window ventilation as per recommendations of the national public health authorities; in School~1 ventilation was additionally assisted by a CO$_2$ guided opener of a small window at the top & passive window ventilation solely at the discretion of the teachers \\
+        HEPA-filters samples & collected once at the end of the study & collected twice after the phases where the air cleaner intervention was in place \\
+        Interventions & compulsory face mask wearing as mandated by the public health authorities at that time from the start of the study for a total of 6 out of 14 study weeks; and installation of portable air cleaners towards the end of the study in the study classrooms for a total of 4 out of 14 study weeks & installation of portable air cleaners in the study classrooms using a cross-over design for a total of 7 out of 14 study weeks\\
+        Ventilation conditions & all classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system; passive window ventilation as per recommendations of the national public health authorities; in School~1 ventilation was additionally assisted by a CO$_2$ guided opener of a small window at the top  & all classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system; passive window ventilation solely at the discretion of the teachers \\
         \bottomrule
@@
 
-Our outcome is a binary variable $Y_{tcv}$ which is $1$ if any respiratory virus $v$ was detected in the air of classroom $c$ during study week $t$.
+Our outcome is the binary variable $Y_{tcv}$ which is $1$ if any respiratory virus $v$ was detected in the air of classroom $c$ during study week $t$.
 We model this outcome with a Bayesian logistic regression model
@@
 \end{align}
-The parameter $\mu$ is related to the type of respiratory virus and differences in the study setting (see \Cref{tab:comp_study}) as follows
+The parameter $\mu$ is related to the type of respiratory virus detected (SARS-CoV-2 vs non-SARS-CoV-2) and takes into account differences in the study setting (see \Cref{tab:comp_study}) as follows
 \begin{align}
-    \mu_{tcv} = \log(\text{offset}_c) + \alpha + \beta_1\,\cdot\text{saliva}_{tcv} + \beta_2\,\cdot\text{CoV}_{tcv} + \beta_3\,\cdot\text{mask}_{tcv} + \beta_4\,\cdot\text{aircl}_{tcv} + 
+    \mu_{tcv} = \log(\text{offset}_c) + \alpha + \beta_1\,\cdot\text{saliva}_{tcv} + \beta_2\,\cdot\text{CoV}_{tcv} + \beta_3\,\cdot\text{mask}_{tcv} + \beta_4\,\cdot\text{aircl}_{tcv} + \beta_5\,\cdot\text{CO}_{2,tcv}.
 \end{align}
-where
-\begin{itemize}
-    \item \textbf{offset}: Since one classroom in 2022 was shared by two classes, we set the offset of this classroom to 2 (double odds for airborne detection in this classroom).
-Conversely, since one classroom in 2023 only had one instead of two sampling devices in place, we set the offset of this classroom to 0.5 (half odds for airborne detection in this classroom).
-    \item  \textbf{saliva}: Binary variable indicating if there was a positive saliva sample in the same week. Airborne detection is more likely if the same virus was found in student saliva.
-    \item \textbf{CoV}: Binary variable indicating if virus $v$ in classroom $c$ in week $t$ was SARS-CoV-2.
-Parameter $\beta_2$ models the log odds of detecting any SARS-CoV-2 vs non-SARS-CoV-2 in bioaerosols.
-This is our parameter of interest.
-    \item \textbf{masks}: Binary variable indicating if face mask wearing was mandated. Masks possibly reduce the probability of airborne detection.
-    \item \textbf{aircl}: Binary variable indicating if portable air cleaners were installed in the classroom. Air cleaners possibly reduce the probability of airborne detection.
-    \item \textbf{CO}$_2$: Weekly average of the daily maximum CO$_2$ levels in each classroom.
-Higher CO$_2$ levels are a proxy of worse ventilation (outdoor air exchange) and possibly increase the probability of airborne detection.
-\end{itemize}
+Since one classroom in 2022 was shared by two classes, we set the \textbf{offset} of this classroom to 2 (double odds for airborne detection in this classroom).
+Conversely, since one classroom in 2023 only had one instead of two sampling devices in place, we set the \textbf{offset} of this classroom to 0.5 (half odds for airborne detection in this classroom).
+Other classrooms have an offset of 1.
+The binary variable \textbf{saliva} indicates if there was a positive saliva sample in the same week, considering with $\beta_1$ that airborne detection was more frequent if the same virus was already detected in saliva.
+The binary variable \textbf{CoV} indicates if the detected virus was SARS-CoV-2, thus our parameter of interest $\beta_2$ is the estimated log odds for detecting any SARS-CoV-2 versus non-SARS-CoV-2 in bioaerosols.
+The binary variables \textbf{masks} and \textbf{aircl} indicate if compulsory face mask wearing was mandated or the portable air cleaners were installed in the classroom, respectively, considering with $\beta_3$ and $\beta_4$ that these interventions possibly reduce the probability of airborne detection.
+The continuous variable \textbf{CO}$_2$ is the weekly average of the maximum daily CO$_2$ levels per classroom, considering with $\beta_5$ that higher CO$_2$ levels (a proxy for ventilation through greater outdoor air exchange) possibly increase the probability of airborne detection.
 
-We model the intercept with a diffuse prior
+We model the effects of the aforementioned variables with a mix of  non-informative and informative priors. The intercept is modeled with a non-informative prior
 \begin{align}
@@
 \end{align}
-Bioaerosols were more frequently found in weeks with a positive saliva sample. Therefore, we model $\beta_1$ with an informative prior
+Since bioaerosols were more frequently detected in weeks with a positive saliva sample, we model $\beta_1$ with an informative prior
 \begin{align}
@@
 where 5 is the number of weeks with paired samples and 3 is the number of weeks with a positive bioaerosol but without a positive saliva sample.
-The prior's standard deviation is computed as the ratio of the standard deviation of the outcome $s_y$ and the variable $s_{\text{saliva}}$.
+The prior's hyperparameter for the standard deviation is computed as the ratio of the standard deviation of the outcome $s_y$ and the variable $s_{\text{saliva}}$.
 \\
-Our parameter of interest is modeled with a non-informative prior
+Our parameter of interest $\beta_2$ is modeled with a non-informative prior
 \begin{align}
@@
 \end{align}
-Masks and air cleaners possibly reduce the odds molecular airborne detection. To incorporate this prior believe into $\beta_3$ and $\beta_4$, we used an informative asymmetric Laplace prior 
+Since masks and air cleaners possibly reduce the odds of molecular airborne detection, we used an informative asymmetric Laplace prior for $\beta_3$ and $\beta_4$ as follows
 \begin{align}
@@
 \end{align}
-which has a 90\% prior probability for a negative intervention effect and is skewed towards smaller effects. \\
+which both place 90\% prior probability on negative intervention effects and are skewed towards smaller effect sizes. \\
 Similarly, we incorporate our prior belief that higher CO$_2$ levels possibly increase the odds of airborne detection with the converse prior for $\beta_5$ as
@@
 \end{align}
+
 To summarize, we estimate the odd ratio for airborne detection of SARS-CoV-2, adjusting for differences in the setup of the classrooms (offset), positive saliva sample in the same week (saliva), the effects of interventions (masks and aircl), and differences in ventilation conditions (CO$_2$).
-We use informative priors because the sample size is small and insufficient to identify the effects of the confounders.
+We use informative priors for the adjustments because the sample size and between-years comparison otherwise make it difficult to identify their effects.
 

### 2024-01-03 doc/paper/supplements.tex
Updates from Overleaf

@@
 
-\title{\LARGE\singlespacing{\textbf{Supplmenetary Material} \\ \medskip
+\title{\LARGE\singlespacing{\textbf{Supplementary Material} \\ \medskip
 Air cleaners and respiratory infections in schools: \\ A multi-measurement observational study}}
@@
 \affil[*]{Corresponding author: lukas.fenner@ispm.unibe.ch }
-
-\affil[$\ddag$]{These authors contributed equally to this work.}
 

### 2024-02-16 analysis/aud_analysis.Rmd
revisions

@@
           eps_file = "../results/cough-data/coughs-frequency-by-condition.eps", 
-          w = 7.25, h = 6)
+          w = 5.25, h = 6)
 

### 2024-02-16 analysis/env_analysis.Rmd
revisions

@@
           eps_file = "../results/env-data/aerosol-number-boxplot.eps", 
-          w = 7.25, h = 6)
+          w = 5.25, h = 6)
 

### 2024-02-16 analysis/epi_analysis.Rmd
revisions

@@
           eps_file = "../results/epi-data/cases_by_condition.eps",
-          w = 7.25, h = 6)
+          w = 5.25, h = 6)
 

### 2024-02-16 analysis/study_comaprison.Rmd
revisions

@@
              air_2023 %>% mutate(year_num = 2023))
+
+# viral loads
+vl_2022 <- readxl::read_xlsx("../data-clean/comparison/mcid-1-molecular-results.xlsx", "Viral Loads")
+vl_2023 <- readxl::read_xlsx("../data-clean/comparison/mcid-2-molecular-results.xlsx", "Viral Loads")
+
+vl <- rbind(vl_2022 %>% mutate(year_num = 2022),
+            vl_2023 %>% mutate(year_num = 2023))
 
@@
   slice(rep(1:n(), times = n)) %>%
-  mutate(year = ifelse(year_num == 2022, "Winter 2021/2022", "Winter 2022/2023"),
-         year = factor(year, levels = c("Winter 2021/2022", "Winter 2022/2023")),
+  mutate(year = ifelse(year_num == 2022, "2021/22", "2022/23"),
+         year = factor(year, levels = c("2021/22", "2022/23")),
          virus = factor(virus, levels = c("IF", "HRV", "AdV", "CoV", "MPV", "PIV")),
@@
 
-backgrd_y <- saliva_air_week %>%
+backgrd_y <- saliva %>%
+  group_by(school, class, week, year_num) %>%
+  slice(1) %>%
+  ungroup() %>%
+  mutate(
+    classroom = ifelse(school == 1, "Classroom 1", ifelse(class == "A", "Classroom 1", "Classroom 2")),
+    year = ifelse(year_num == 2022, "2021/22", "2022/23")
+  ) %>%
+  dplyr::select(year, classroom, week, intervention) %>%
   group_by(year, classroom, week) %>%
-  summarize(type_num = 1) %>%
+  slice(1) %>%
   ungroup() %>%
-  complete(year, classroom, week)
+  mutate(
+    intervention = fct_recode(intervention, !!! c("Mask wearing" = "masks", "Air cleaners" = "air cleaners", "No intervention" = "none")),
+    intervention = factor(intervention, levels = c("No intervention", "Mask wearing", "Air cleaners")),
+    type_num = 1)
 
@@
   facet_grid(classroom ~ year) +
-  geom_rect(data = backgrd_y,
-            mapping = aes(xmin = week - 0.5, xmax = week + 0.5, ymin = -Inf, ymax = Inf),
-            fill = ifelse(backgrd_y$week %% 2 == 0, "lightgray", "white"),
-            alpha = 0.5) +
+  geom_vline(data = backgrd_y, mapping = aes(xintercept = week + .5), alpha = .1) +
   geom_hline(data = data.frame(y = 1.5), mapping = aes(yintercept = y), linetype = "dashed") +
@@
              position = position_dodge(width = .75)) +
+  geom_rect(data = backgrd_y,
+            mapping = aes(xmin = week - 0.5, xmax = week + 0.5, ymin = -Inf, ymax = Inf, fill = intervention),
+            alpha = 0.2) +
   scale_color_manual(values = wes_palette("IsleofDogs1")) +
+  scale_fill_manual(values = c("grey", wes_palette("Zissou1")[1], wes_palette("GrandBudapest1")[2])) +
   scale_x_continuous(breaks = 1:7, limits = c(0.5, 7.5), expand = c(0,0)) +
@@
   annotate("text", y = 0.2, x = 0.65, label = "Bioaerosol", hjust = 0, size = 8 / cm(1)) +
-  labs(x = "Study week", title = "b") +
+  labs(x = "Study week", title = "b", fill = "Intervention") +
   theme_bw2() +
@@
         strip.text = element_text(face = 2, size = 8),
+        panel.background = element_rect(fill = "white"),
+        plot.background = element_rect(fill = "white"),
+        legend.position = "bottom",
+        legend.key.width = unit(0.4, "cm"),
+        legend.key.height = unit(0.5, "cm"),
         text = element_text(size = 8),
@@
 
+# posterior probability
+post_prob_p <- mod %>%
+  spread_draws(beta_cov) %>%
+  ungroup() %>%
+  summarize(p = round(sum(beta_cov > 1) / n() * 100)) %>%
+  dplyr::select(p) %>% unlist()
+
+# predicted probabilities
 pred_prob <- mod %>%
@@
 
-prob_pl <- ggplot(mapping = aes(x = variable, y = value, color = variable)) +
+prob_pl <- ggplot(pred_prob, aes(x = variable, y = value)) +
   geom_errorbar(data = pred_prob %>% group_by(variable) %>% median_qi(value),
-                mapping = aes(ymin = .lower, ymax = .upper, color = variable),
+                mapping = aes(x = variable, ymin = .lower, ymax = .upper, color = variable),
                 width = .25) +
   geom_boxplot(data = pred_prob,
+               mapping = aes(color = variable),
                position = position_dodge(width = .5), outlier.shape = NA, coef = 0, width = 0.25) +
@@
                position = position_dodge2(width = .5), size = 2, shape = 23, fill = "white") +
+  ggsignif::geom_signif(
+    annotations = paste0("Pr.=", post_prob_p, "%"), 
+    y_position = .55, 
+    xmin = 1, xmax = 2, 
+    textsize = 8 / cm(1), 
+    vjust = -.5, 
+    tip_length = .075) +
   scale_y_continuous(expand = c(0,0), limits = c(0, NA), labels = function(x) x * 100) +
@@
 
+## Viral Loads
+
+```{r}
+vl_saliva <- vl %>%
+  filter(type == "saliva") %>%
+  mutate(virus = ifelse(virus == "IFA" | virus == "IFB", "IF", virus), 
+         virus_type = ifelse(virus == "CoV", "CoV", "non-CoV"),
+         virus_type = factor(virus_type, levels = c("CoV", "non-CoV")),
+         year = ifelse(year_num == 2022, "2021/22", "2022/23"),
+         year = factor(year, levels = c("2021/22", "2022/23")),
+         virus = factor(virus, levels = c("IF", "HRV", "AdV", "CoV", "MPV", "PIV")))
+
+vl_test <- t.test(
+  vl_saliva$ct[vl_saliva$virus_type == "CoV"],
+  vl_saliva$ct[vl_saliva$virus_type == "non-CoV"])
+
+vl_saliva_pl <- vl_saliva %>%
+  ggplot(aes(x = virus_type, y = ct)) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_jitter(aes(shape = year, color = virus), width = .25, height = 0) +
+  ggsignif::geom_signif(
+    annotations = paste0("p=", round(vl_test$p.value, 2)), 
+    y_position = 44, 
+    xmin = 1, xmax = 2, 
+    textsize = 8 / cm(1), 
+    vjust = -.5, 
+    tip_length = .075) +
+  scale_color_manual(values = wes_palette("IsleofDogs1")) +
+  scale_shape_manual(values = c(1, 17)) +
+  scale_y_continuous(limits = c(NA, 46)) +
+  labs(y = "Ct value") +
+  theme_bw2() +
+  theme(axis.title.x = element_blank(),
+        legend.position = c(0.1, 0.87),
+        legend.background = element_rect(color = "black"),
+        plot.title.position = "plot",
+        legend.title = element_blank()) +
+  guides(color = "none")
+
+save_plot(vl_saliva_pl,
+          "../results/comparison-viral-loads.png",
+          w = 12, h = 8)
+```
+
+## Summary plot
+
 ```{r}
@@
 33"
-comparison <- saliva_pl + prob_pl + time_pl + plot_layout(design = design, widths = c(11, 5), heights = c(6, 9))
-ggsave(plot = comparison, file = "../results/comparison.pdf", width = 16 / cm(1), height = 15 / cm(1))
+comparison <- saliva_pl + prob_pl + time_pl + plot_layout(design = design, widths = c(11, 5), heights = c(6, 11))
+ggsave(plot = comparison, file = "../results/comparison.pdf", width = 16 / cm(1), height = 17 / cm(1))
+ggsave(plot = comparison, file = "../results/comparison.png", width = 16 / cm(1), height = 17 / cm(1))
 ```
-
-<!-- ## No. of positive samples -->
-
-<!-- ```{r} -->
-<!-- pos_samples <- data.frame( -->
-<!--   year_num = factor(c(rep(2022, 3), rep(2023, 3)), levels = c(2022, 2023)), -->
-<!--   type = factor(rep(c("Saliva", "Bioaerosols", "HEPA-filter"), times = 2), levels = c("Saliva", "Bioaerosols", "HEPA-filter")), -->
-<!--   n_pos = c(c(21, 10, 8), c(50, 2, 4)), -->
-<!--   n_tot = c(c(262, 130, 40), c(448, 105, 160)) -->
-<!-- ) %>% -->
-<!--   mutate(n_neg = n_tot - n_pos, -->
-<!--          year = ifelse(year_num == 2022, "Winter 2021/2022", "Winter 2022/2023"), -->
-<!--          year = factor(year, levels = c("Winter 2021/2022", "Winter 2022/2023"))) %>% -->
-<!--   dplyr::select(-n_tot,-n_neg) %>% -->
-<!--   reshape2::melt(c("year_num", "year", "type")) -->
-
-<!-- pos_samples_pl <- pos_samples %>% -->
-<!--   ggplot(aes(y = value, x = year_num, fill = year)) + -->
-<!--   geom_bar(stat = "identity") + -->
-<!--   geom_text(aes(label = value, color = year), size = 8 / cm(1), vjust = -1) + -->
-<!--   facet_wrap(~ type, nrow = 1) + -->
-<!--   scale_y_continuous(expand = expansion(add = c(0,5))) + -->
-<!--   scale_fill_manual(values = RColorBrewer::brewer.pal(9, "PuBu")[c(4,9)]) + -->
-<!--   scale_color_manual(values = RColorBrewer::brewer.pal(9, "PuBu")[c(4,9)]) + -->
-<!--   labs(y = "Frequency of positive samples", title = "b") + -->
-<!--   theme_custom() + -->
-<!--   theme(legend.position = "none", -->
-<!--         plot.title.position = "plot", -->
-<!--         strip.text = element_text(face = 2, vjust = 2, margin = margin(t = 10, b = 10)), -->
-<!--         panel.spacing.x = unit(1, "cm"), -->
-<!--         axis.title.x = element_blank(), -->
-<!--         panel.grid.major = element_blank(), -->
-<!--         panel.grid.minor = element_blank()) -->
-
-<!-- pos_samples_pl -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- comparison <- grid.arrange(saliva_pl, pos_samples_pl, nrow = 2, heights = c(6, 8), widths = 11) -->
-<!-- ggsave(plot = comparison, file = "../results/comparison.pdf", width = 11.5 / cm(1), height = 14 / cm(1)) -->
-<!-- ``` -->
-
-<!-- ## Absences ## -->
-
-<!-- ```{r} -->
-<!-- mcid_2 <- readRDS("../data-clean/redcap-long.rds") %>% -->
-<!--   filter(respiratory_infection == "yes") %>% -->
-<!--   mutate(absent_days = date_back - date_absent) -->
-
-<!-- # 35 absences related to respiratory infection among 38 students -->
-
-<!-- summary(as.numeric(mcid_2$absent_days)) -->
-
-<!-- # on average 3.0 days out of school per absence (but weekends should be kept in mind) -->
-
-<!-- mcid_1 <- read_csv("../data-clean/mcid-1-redcap-cases.csv") %>% -->
-<!--   filter(!is_teacher, -->
-<!--          school_class == "School 1 (A)" | school_class == "School 2 (D)") %>% -->
-<!--   mutate(absent_days = date_end - date_start) -->
-
-<!-- # 40 absences related to respiratory infection among 44 students -->
-
-<!-- summary(as.numeric(mcid_1$absent_days)) -->
-
-<!-- #' on average 4.1 days out of school per absence (but weekends should be kept in mind) -->
-<!-- #' there is also an outlier with 30 days, but the median is still higher  -->
-
-<!-- rbind(mcid_1 %>% mutate(year = 2022) %>% dplyr::select(year, absent_days), -->
-<!--       mcid_2 %>% mutate(year = 2023) %>% dplyr::select(year, absent_days)) %>% -->
-<!--   mutate(absent_days = as.numeric(absent_days)) %>% -->
-<!--   ggplot(aes(fill = factor(year), x = year, y = absent_days)) + -->
-<!--   geom_boxplot() + -->
-<!--   scale_y_sqrt(breaks = c(0, 1:7, 10, 20, 30), limits = c(1,30)) -->
-<!-- ``` -->
-
-<!-- ## Particle reduction -->
-
-<!-- ```{r} -->
-<!-- vars <- c("CO2", "rH", "Temp", "AER", "Cn", "PM1", "PM25", "PM4", "PM10") -->
-<!-- vars_labels_no_units <- c("CO[2]", "Relative~humidity", "Temperature", "Air~exchange~rate", -->
-<!--                           "CN", "PM[1]", "PM[2.5]", "PM[4]", "PM[10]") -->
-<!-- names(vars_labels_no_units) <- vars -->
-
-<!-- particle <- data.frame( -->
-<!--   year = factor(c(rep(2022, 15), rep(2023, 15)), levels = c(2022, 2023)), -->
-<!--   variable = factor(c(rep(c("Cn", "PM1", "PM25", "PM4", "PM10"), each = 3, times = 2)), levels = c("Cn", "PM1", "PM25", "PM4", "PM10")), -->
-<!--   interval = rep(c("Mean", "Lower", "Upper"), times = 5), -->
-<!--   estimate = c(c(39.24, 3.88, 68.80,  -->
-<!--                  41.05, 6.62, 67.98, -->
-<!--                  45.12, 19.66, 63.16, -->
-<!--                  46.49, 27.26, 61.72, -->
-<!--                  46.60, 29.99, 59.49), -->
-<!--                c(76.83, 63.20, 86.28, -->
-<!--                  76.79, 64.84, 85.46, -->
-<!--                  72.86, 64.24, 79.66, -->
-<!--                  70.77, 63.46, 76.97, -->
-<!--                  99.99, 99.94, 100.00)) -->
-<!-- ) %>% -->
-<!--   dcast(year + variable ~ interval) %>% -->
-<!--   ggplot(aes(y = year, color = year)) + -->
-<!--   facet_wrap(~ variable, nrow = 1, dir = "h", labeller = as_labeller(vars_labels_no_units, label_parsed)) + -->
-<!--   geom_errorbar(mapping = aes(xmin = Lower, xmax = Upper), lineend = "square", size = 1, width = .05) + -->
-<!--   geom_point(mapping = aes(x = Mean), shape = 23, size = 2, fill = "white") + -->
-<!--   scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) + -->
-<!--   scale_x_continuous(expand = expansion(mult = c(0, 0.1)), breaks = c(0, 50, 100), limits = c(0, 100)) + -->
-<!--   labs(y = "Study year", x = "Estimated reduction in particle concentration (%)", title = "c") +  -->
-<!--   theme_custom() + -->
-<!--   theme(legend.position = "none", -->
-<!--         plot.title.position = "plot", -->
-<!--         panel.spacing = unit(.5, "lines")) -->
-
-<!-- particle -->
-<!-- ``` -->
-
-<!-- ## Risk ratio -->
-
-<!-- ```{r} -->
-<!-- rr <- data.frame( -->
-<!--   year = factor(c(rep(2022, 3), rep(2023, 3)), levels = c(2022, 2023)), -->
-<!--   interval = rep(c("Mean", "Lower", "Upper"), times = 2), -->
-<!--   estimate = c(c(1.00, 0.22, 2.067), c(0.72, 0.44, 1.15)) -->
-<!-- ) %>% -->
-<!--   dcast(year ~ interval) %>% -->
-<!--   ggplot(aes(y = year, color = year)) + -->
-<!--   geom_vline(aes(xintercept = 1), linetype = "dashed") + -->
-<!--   geom_errorbar(mapping = aes(xmin = Lower, xmax = Upper), lineend = "square", size = 1, width = .05) + -->
-<!--   geom_point(mapping = aes(x = Mean), shape = 23, size = 2, fill = "white") + -->
-<!--   labs(y = "Study year", x = "ARR of infection with air cleaners (%)", title = "d") +  -->
-<!--   scale_color_manual(values = wes_palette("Darjeeling2")[c(2, 5)]) + -->
-<!--   theme_custom() + -->
-<!--   theme(legend.position = "none", -->
-<!--         plot.title.position = "plot") -->
-
-<!-- rr   -->
-<!-- ``` -->
-
-<!-- ## Combined  -->
-
-<!-- ```{r} -->
-<!-- comparison <- grid.arrange(saliva, pos_samples, particle, rr, ncol = 2, nrow = 2, heights = c(6, 4), widths = c(10,6)) -->
-<!-- ggsave(plot = comparison, file = "../results/comparison.pdf", width = 16 / cm(1), height = 10 / cm(1)) -->
-<!-- ``` -->

### 2024-02-16 doc/comparison-mcid12/supplements.tex
Updates from Overleaf

@@
 \title{\LARGE\singlespacing{\textbf{Supplementary Material} \\ \medskip
-Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva, air, and filter samples from schools during and after the pandemic}}
+Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and classroom air: a two winters tale}}
 
@@
         \toprule
-        & \textbf{2022 (during pandemic)} & \textbf{2023 (after pandemic)} \\
+        & \textbf{Winter 2021/22} & \textbf{Winter 2022/23} \\
         \midrule
-        Participants & 2~classes from School~1 sharing one classroom (half-class teaching) and 1~class from School~2 with a separate classroom & two classes from School~2 with separate classrooms \\
+        Participants & 2~classes from School~1 sharing one classroom (half-class teaching) and 1~class from School~2 with a separate classroom; School~1 and~2 are in different cities, which have grown together to form a community (\ie the schools are located nearby) & two classes from School~2 with separate (two) classrooms; none of the two classes already participated in winter 2021/22 \\
         Duration & 7 weeks per class (January 16 to March 11, 2023), excluding vacation & 7 weeks per class (January 24 to March 26, 2023), excluding vacation \\
@@
         HEPA-filters samples & collected once at the end of the study & collected twice after the phases where the air cleaner intervention was in place \\
-        Interventions & compulsory face mask wearing as mandated by the public health authorities at that time from the start of the study for a total of 6 out of 14 study weeks; and installation of portable air cleaners towards the end of the study in the study classrooms for a total of 4 out of 14 study weeks & installation of portable air cleaners in the study classrooms using a cross-over design for a total of 7 out of 14 study weeks\\
+        Interventions & compulsory face mask wearing as mandated by the public health authorities at that time from the start of the study, for a total of 6 out of 14 study weeks; and installation of portable air cleaners towards the end of the study in the study classrooms for a total of 4 out of 14 study weeks; in total, 10 study weeks with interventions & installation of portable air cleaners in the study classrooms using a cross-over design, for a total of 7 out of 14 study weeks; in total 7 study weeks with interventions\\
         Ventilation conditions & all classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system; passive window ventilation as per recommendations of the national public health authorities; in School~1 ventilation was additionally assisted by a CO$_2$ guided opener of a small window at the top  & all classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system; passive window ventilation solely at the discretion of the teachers \\
@@
 
-\section{Model to estimate differences in airborne detection}\label{sec:model}
+\begin{figure}
+    \centering
+    \includegraphics{}
+    \caption{Caption}
+    \label{fig:enter-label}
+\end{figure}
+
+\clearpage
+
+\section*{Model to estimate differences in airborne detection}\label{sec:model}
 
@@
 \begin{align}
-    \mu_{tcv} = \log(\text{offset}_c) + \alpha + \beta_1\,\cdot\text{saliva}_{tcv} + \beta_2\,\cdot\text{CoV}_{tcv} + \beta_3\,\cdot\text{mask}_{tcv} + \beta_4\,\cdot\text{aircl}_{tcv} + \beta_5\,\cdot\text{CO}_{2,tcv}.
+    \mu_{tcv} = \log(\text{offset}_c) + \alpha + \beta_1\,\cdot\text{saliva}_{tcv} + \beta_2\,\cdot\text{CoV}_{tcv} + \beta_3\,\cdot\text{mask}_{tc} + \beta_4\,\cdot\text{aircl}_{tc} + \beta_5\,\cdot\text{CO}_{2,tc}.
 \end{align}
@@
 Other classrooms have an offset of 1.
-The binary variable \textbf{saliva} indicates if there was a positive saliva sample in the same week, considering with $\beta_1$ that airborne detection was more frequent if the same virus was already detected in saliva.
-The binary variable \textbf{CoV} indicates if the detected virus was SARS-CoV-2, thus our parameter of interest $\beta_2$ is the estimated log odds for detecting any SARS-CoV-2 versus non-SARS-CoV-2 in bioaerosols.
-The binary variables \textbf{masks} and \textbf{aircl} indicate if compulsory face mask wearing was mandated or the portable air cleaners were installed in the classroom, respectively, considering with $\beta_3$ and $\beta_4$ that these interventions possibly reduce the probability of airborne detection.
-The continuous variable \textbf{CO}$_2$ is the weekly average of the maximum daily CO$_2$ levels per classroom, considering with $\beta_5$ that higher CO$_2$ levels (a proxy for ventilation through greater outdoor air exchange) possibly increase the probability of airborne detection.
+The binary variable \textbf{saliva} indicates if there was a positive saliva sample in the same week, with $\beta_1$ estimating whether airborne detection was more frequent if the same virus was already detected in saliva.
+This variable implicitly adjusts for community transmission, which would impact molecular detection both in saliva and airborne samples.
+The binary variable \textbf{CoV} indicates if the detected virus was SARS-CoV-2, thus our parameter of interest $\beta_2$ is estimating whether airborne detection of any SARS-CoV-2 is more likely than non-SARS-CoV-2 viruses.
+The binary variables \textbf{masks} and \textbf{aircl} indicate if compulsory face mask wearing was mandated or portable air cleaners were installed in the classroom, respectively, with $\beta_3$ and $\beta_4$ estimating the change in the probability of airborne detection due to the intervention.
+The continuous variable \textbf{CO}$_2$ is the weekly average of the maximum daily CO$_2$ levels per classroom, with $\beta_5$ estimating the impact of CO$_2$ levels (a proxy for classroom ventilation) on the probability of airborne detection.
 
-We model the effects of the aforementioned variables with a mix of  non-informative and informative priors. The intercept is modeled with a non-informative prior
+Considering the relatively small sample size and prior knowledge, we model the effects of the aforementioned variables with a mix of  non-informative and informative priors.
+The intercept is modeled with a non-informative prior
 \begin{align}
@@
 where 5 is the number of weeks with paired samples and 3 is the number of weeks with a positive bioaerosol but without a positive saliva sample.
-The prior's hyperparameter for the standard deviation is computed as the ratio of the standard deviation of the outcome $s_y$ and the variable $s_{\text{saliva}}$.
+The prior's hyperparameter for the standard deviation is computed as the ratio of the standard deviation of the outcome $s_y$ and predictor $s_{\text{saliva}}$.
 \\
@@
 \end{align}
-Since masks and air cleaners possibly reduce the odds of molecular airborne detection, we used an informative asymmetric Laplace prior for $\beta_3$ and $\beta_4$ as follows
+Since masks and air cleaners possibly reduce the odds of airborne detection, we used informative asymmetric Laplace priors for $\beta_3$ and $\beta_4$ as follows
 \begin{align}
@@
 \end{align}
-which both place 90\% prior probability on negative intervention effects and are skewed towards smaller effect sizes. \\
-Similarly, we incorporate our prior belief that higher CO$_2$ levels possibly increase the odds of airborne detection with the converse prior for $\beta_5$ as
+which both place 90\% prior probability on negative intervention effects (decreasing the odds of airborne detection) and are skewed towards smaller effect sizes. \\
+Similarly, we incorporate our prior belief that higher CO$_2$ levels possibly increase the odds of airborne detection with an informative asymmetric Laplace prior for $\beta_5$ as
 \begin{align}
-    \beta_3 &\sim \text{Asymmetric-Laplace}(\mu = 0.21, \sigma = 0.5\cdot s_y / s_{\text{masks}}, \tau = 0.25)~.
+    \beta_3 &\sim \text{Asymmetric-Laplace}(\mu = 0.21, \sigma = 0.5\cdot s_y / s_{\text{masks}}, \tau = 0.25)~,
 \end{align}
+which places 90\% prior probability on positive effects (increasing the odds of airborne detection).
 
-To summarize, we estimate the odd ratio for airborne detection of SARS-CoV-2, adjusting for differences in the setup of the classrooms (offset), positive saliva sample in the same week (saliva), the effects of interventions (masks and aircl), and differences in ventilation conditions (CO$_2$).
-We use informative priors for the adjustments because the sample size and between-years comparison otherwise make it difficult to identify their effects.
+To summarize, we estimate the probability of airborne detection of SARS-CoV-2, adjusting for differences in the setup of the classrooms (offset), positive saliva sample in the same week (saliva), the effects of interventions (masks and aircl), and differences in ventilation conditions (CO$_2$).
+We use informative priors for the parameters of the adjustments to incorporate prior knowledge and because the sample size and between-years comparison would otherwise make it difficult to identify these parameters.
 

### 2024-02-16 doc/paper/supplements.tex
Updates from Overleaf

@@
 
-\section{Modeling positivity rate of human saliva samples}\label{sec:multinomial-model}
+% \section{Modeling positivity rate of human saliva samples}\label{sec:multinomial-model}
 
-Molecular analysis determined which saliva samples were positive for a respiratory virus.
+% Molecular analysis determined which saliva samples were positive for a respiratory virus.
 Let $p = 0, 1, \dots, P$ denote the virus, where $0$ refers to negative samples and $P$ is the number of different viruses detected over the study.
 The number of positive bioaerosol and saliva samples $y_p$ in samples of sizes $n$ is analyzed with a Multinomial logistic regression model
-\begin{align*}
-    y_0, y_1, \cdots, y_P | n, \beta_0, \beta_1, \beta_2, \beta_3 &\sim \text{Multinomial-Logit}(\theta_0, \theta_1, \dots \theta_P) \\
-    \theta &= \text{softmax}(\mu) \\
-    \mu &= \begin{cases}
-                0, & \text{if } p=0, \\
-                \beta_{0p} + \beta_{1} \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Susceptibles}_p, & \text{if } p>0.
-            \end{cases} \\
-    \beta_{0p} &\sim \text{Normal}(0, 2.5) \quad \forall p \\
-    \beta_1, \beta_2, \beta_3 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) 
-\end{align*}
-where $\text{softmax}(\mu) = \exp(\mu)/\sum\exp(\mu)$ and $s_{x}$ is the empirical standard deviation of each input variable.
+% \begin{align*}
+%     y_0, y_1, \cdots, y_P | n, \beta_0, \beta_1, \beta_2, \beta_3 &\sim \text{Multinomial-Logit}(\theta_0, \theta_1, \dots \theta_P) \\
+%     \theta &= \text{softmax}(\mu) \\
+%     \mu &= \begin{cases}
+%                 0, & \text{if } p=0, \\
+%                 \beta_{0p} + \beta_{1} \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Susceptibles}_p, & \text{if } p>0.
+%             \end{cases} \\
+%     \beta_{0p} &\sim \text{Normal}(0, 2.5) \quad \forall p \\
+%     \beta_1, \beta_2, \beta_3 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) 
+% \end{align*}
+% where $\text{softmax}(\mu) = \exp(\mu)/\sum\exp(\mu)$ and $s_{x}$ is the empirical standard deviation of each input variable.
 The negative test is set as the reference category.
@@
 
-Note that we deviated from the statistical analysis plan by estimating $\beta_{0p}$ and $\beta_1$ without a hierarchical prior.
+% Note that we deviated from the statistical analysis plan by estimating $\beta_{0p}$ and $\beta_1$ without a hierarchical prior.
 First, there was considerable variation in the overall positivity rate of each virus, so that we decided to use unpooled intercepts $\beta_{0p}$ instead of partially pooled intercepts $\beta_{0[p]}$ for each virus.
@@
 
-\section{Detailed results for positivity rate in saliva samples}
-\label{sec:detailed-molecular}
+% \section{Detailed results for positivity rate in saliva samples}
+% \label{sec:detailed-molecular}
 
-There can be unknown delay between when students were infected and when they tested positive.
+% There can be unknown delay between when students were infected and when they tested positive.
 Therefore, we performed a sensitivity analysis where we lagged the molecular test results (number of positive saliva samples) by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday).
@@
 The estimated effect of air cleaners remains indistinguishable from zero regardless of the considered lag.
-\begin{figure}[!htb]
+% \begin{figure}[!htb]
+% \centering
+%     \includegraphics[width=10cm]{../../results/mol-data/model-results.pdf}
+%     \caption[Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples]{Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples.
+Shown is the adjusted relative risk (posterior mean as dot and 50\%-, 80\%, and 95\%-CrIs as lines) for air cleaners by lag of the molecular test results.}
+%     \label{fig:mol-estimation-results-sensitivity}
+% \end{figure}
+
+\clearpage
+
+\section{Possible transmission networks}\label{sec:transmission-networks}
+
+The distribution of positive saliva samples varied between classes.
+For example, all but one sample was positive for adenvovirus in class~A during the first three study weeks, while the vast majority of positive influenza~B samples were found in class~B.
+To illustrate possible transmission chains within classes, we linked positive saliva samples of the same virus that were less than one week apart.
+Based on this spatiotemporal analysis, we identified 10~possible transmission chains (\Cref{fig:molecular-descriptives}).
+The longest potential transmission chains occurred in January, referring to a cluster of adenovirus infections in class~A and influenza~B infections in class~B.
+Molecular genotyping to verify the proposed transmission network was unsuccessful because we could not amplify and sequence any of the gene targets.
+We also analyzed 105~bioaerosol samples and detected in two of them viral RNA (1~rhinovirus in class~A and 1~adenovirus in class~B).
+Similarly, we detected 1~influenza~B, 1~rhinovirus, 1~adenovirus, and 1~SARS-CoV-2 in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).
+
+\begin{figure}[!htpb]
     \centering
-    \includegraphics[width=10cm]{../../results/mol-data/model-results.pdf}
-    \caption[Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples]{Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples.
-Shown is the adjusted relative risk (posterior mean as dot and 50\%-, 80\%, and 95\%-CrIs as lines) for air cleaners by lag of the molecular test results.}
-    \label{fig:mol-estimation-results-sensitivity}
+    \includegraphics{../../results/mol-data/saliva-distribution.pdf}\hspace{.5cm}
+    \includegraphics{../../results/mol-data/network-plot.pdf}
+    \caption{\textbf{Molecular detection of respiratory viruses and transmission network based on spatiotemporal analysis of students' saliva samples}.
+\textbf{(a)}~Number of positive saliva samples by virus.
+\textbf{(b)}~Daily number of positive saliva samples (colored circles) and possible transmission chains within classes (directed arrows).
+Positive samples are linked if they belong to the same virus and are less than 1~week apart.
+Positive samples from the air and filters as blank squares aligned.
+IFB: influenza~B, HRV: human rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: human metapneumovirus, PIV: parainfluenza virus.}
+    \label{fig:molecular-descriptives}
 \end{figure}

### 2024-02-16 doc/comparison-mcid12/manuscript.tex

@@
 
-\title{\bfseries\LARGE\singlespacing{Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva, air, and filter samples from schools during and after the pandemic}}
+\title{\bfseries\LARGE\singlespacing{Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and classroom air: a two winters tale}}
 
@@
 \begin{information}\normalfont
-\textbf{Word count:} Abstract 248 (max 250), Manuscript 1,187 (max 1,200) \\
+\textbf{Word count:} Abstract 248 (max 250), Manuscript 1,230 (max 1,200) \\
 \textbf{Display items:} Figure 1 (max 2) \\
-\textbf{References:} 14 (max 15) \\
+\textbf{References:} 15 (max 15) \\
 %\textbf{Supporting information:} STROBE checklist, Supplementary Data with texts, tables, and figures
@@
 Daily bioaerosol samples were pooled for each sampling device before PCR analysis of a panel of 19 respiratory viruses and viral subtypes.
-The probability of detection of airborne viruses was modelled using a Bayesian logistic regression model.
+The probability of detection of airborne viruses was modelled using an adjusted Bayesian logistic regression model.
 
-\noindent \textbf{Results:} Three classes (58 students) participated in 2021/22, and two classes (38 students) in 2022/23 .
+\noindent \textbf{Results:} Three classes (58 students) participated in 2021/22, and two classes (38 students) in 2022/23.
 During winter 2021/22, SARS-CoV-2 dominated in saliva (19 of 21 positive samples) and bioaerosols (9 of 10).
@@
 The transmission of respiratory viruses, such as SARS-CoV-2 and influenza, in schools and other indoor environments is difficult to control\cite{Leung2020NatMed}.
-During the COVID-19 pandemic, non-pharmaceutical interventions and physical distancing reduced the spread of SARS-CoV-2 and other seasonal respiratory viruses, but a resurgence of respiratory cases followed the relaxation of these measures\cite{Poole2020LancetRespMed,Sauteur2022EuroSurv,Kandeel2023BMC}.
+During the COVID-19 pandemic, non-pharmaceutical interventions and physical distancing reduced the spread of SARS-CoV-2 and other seasonal respiratory viruses, but a resurgence of respiratory infections followed the relaxation of these measures\cite{Poole2020LancetRespMed,Sauteur2022EuroSurv,Kandeel2023BMC}.
 Following epidemic peaks, a shift in the circulation of respiratory viruses occurs\cite{Pierangeli2012CMI}, which can be identified by frequent collection of non-invasive saliva samples\cite{Pasomsub2021CMI}.
 
-Respiratory viruses spread by multiple routes, including respiratory particles such as large droplets and small aerosols.
+Respiratory viruses spread via multiple routes, including respiratory particles such as large droplets and small aerosols.
 Unlike larger droplets, which settle quickly, aerosols can remain suspended in the air for extended periods\cite{Wang2021}.
-Airborne infectious pathogens are primarily found in smaller particles (<10$\mu$m) and the distribution is similar across varius pathogens\cite{Fennelly2020}.
-Thus, pathogen-carrying aerosols have the potential for long-range transmission, but the larger concentration of particles near the infectious person favours short-range transmission\cite{Wang2020}.
+Airborne infectious pathogens are primarily found in smaller particles and the distribution is similar across various pathogens\cite{Fennelly2020}.
+Thus, pathogen-carrying aerosols have the potential for long-range transmission, but the larger concentration of particles near the infectious person favors short-range transmission\cite{Wang2020}.
 
@@
 Collins, CO, USA)\cite{Lednicky2016AST}.
-The HEPA filters from the portable air cleaner (Xiaomi Mi Air Pro 70m2, Shenzhen, China ) were removed and divided into 20~fields.
-One sterile Phosphate-Buffered Saline-moistened swab per field was then collected for a total of 20~swabs per filter.
+The HEPA filters from the portable air cleaner (Xiaomi Mi Air Pro 70m2, Shenzhen, China) were removed and divided into 20~fields.
+For each field, one swab moistened with sterile Phosphate-Buffered Saline was collected, amounting to a total of 20~swabs per filter.
 Saliva and airborne samples were transported to the laboratory and stored at $-$80°C until further processing\cite{Huber2021}.
-Before real-time (RT)-PCR analysis, daily bioaerosol samples were pooled for each sampling device and filtered using Amicon Ultra-15 Centrifugal filters\cite{Banholzer2023PLoSMed}.
-Saliva samples were analyzed directly without prior filtration.
-The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19 major respiratory viruses and viral subtypes, including SARS-CoV-2, influenza A/B, respiratory syncytial, metapneumovirus, adenovirus, rhinovirus, and parainfluenza.
+Before real-time (RT)-PCR analysis, daily bioaerosol samples were pooled for each sampling device and enriched using Amicon Ultra-15 Centrifugal filters as described previously\cite{Banholzer2023PLoSMed}.
+Saliva samples were analyzed directly without prior filtration/enrichment.
+The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19 major respiratory viruses and viral subtypes, including SARS-CoV-2, influenza A/B virus, respiratory syncytial virus, metapneumovirus, adenovirus, rhinovirus, and parainfluenza virus.
 
 We used descriptive statistics to present differences in the type and number of respiratory viruses detected in saliva and airborne samples between 2021/22 and 2022/23.
-A Bayesian logistic regression model was used to estimate the probability of detecting any SARS-CoV-2 versus non-SARS-CoV-2 in the air during a study week, adjusting for differences in the study settings and the interventions implemented during the studies (see \supp~Text~\zref{sec:model} for a detailed model description).
+A Bayesian logistic regression model was used to estimate the probability of detecting any SARS-CoV-2 versus non-SARS-CoV-2 viruses in the air during a study week, adjusting for differences in the study settings and the interventions implemented during the studies (see \supp~Text~\zref{sec:model} for a detailed model description).
 
@@
 In 2021/22, 58~students participated in weekly saliva testing.
-There were 21~positive saliva samples during the study, 19~SARS-CoV-2, one~influenza~A, and one~adenovirus (\Cref{fig:comparison}a, left).
+There were 21~positive saliva samples during the study, 19~SARS-CoV-2, one~influenza~A virus, and one~adenovirus (\Cref{fig:comparison}a, left).
 There were 10~positive bioaerosol samples, nine~SARS-CoV-2 and one~adenovirus.
-There were eight~positive samples on the HEPA-filters, six~SARS-CoV-2, one~influenza~A and one~adenovirus.
+There were eight~positive samples on the HEPA-filters, six~SARS-CoV-2, one~influenza~A virus and one~adenovirus.
 In 2022/23, 38~students participated in bi-weekly saliva testing.
-There were 50~positive saliva samples, mostly influenza~B, rhinovirus, and adenovirus (\Cref{fig:comparison}a, right).
+There were 50~positive saliva samples, mostly influenza~B virus, rhinovirus, and adenovirus (\Cref{fig:comparison}a, right).
 There were two~positive bioaerosol samples, one~rhinovirus and one~adenovirus.
-There were four~positive samples on the HEPA-filters of the air cleaners, one~influenza~B, one~rhinovirus, one~adenovirus, and one~SARS-CoV-2.
-We found six positive air-saliva samples in the same week (4~SARS-CoV-2 and two~non-SARS-CoV-2 (\Cref{fig:comparison}b), suggesting they were paired samples.
+There were four~positive samples on the HEPA-filters of the air cleaners, one~influenza~B virus, one~rhinovirus, one~adenovirus, and one~SARS-CoV-2.
+We found six positive air-saliva samples in the same week (4~SARS-CoV-2 and two~non-SARS-CoV-2 viruses (\Cref{fig:comparison}b), suggesting they were paired samples.
 
-SARS-CoV-2 was more likely detected in bioaerosols than non-SARS-CoV-2 (adjusted odds ratio 4.8, 95\%-CrI 2.6$-$9.0).
-The probability of airborne molecular detection was 34\% (95\%-credible interval [CrI] 22\%$-$47\%) for SARS-CoV-2 versus 10\% (95\%-CrI 5\%-16\%) for non-SARS-CoV-2 (\Cref{fig:comparison}c).
-We adjusted estimates for differences in maximum daily CO$_2$, which increased from 1,134\,ppm (standard deviation [SD] 277\,pppm) in 2021/22 to 2,224\,ppm (SD 321\,pppm) in 2022/23.
+SARS-CoV-2 was more likely detected in bioaerosols than other respiratory viruses (adjusted odds ratio 4.8, 95\%-CrI 2.6$-$9.0).
+The probability of airborne molecular detection was 34\% (95\%-credible interval [CrI] 22\%$-$47\%) for SARS-CoV-2 versus 10\% (95\%-CrI 5\%-16\%) for non-SARS-CoV-2 viruses (\Cref{fig:comparison}c).
+We adjusted estimates for differences in maximum daily CO$_2$, which increased from 1,134\,ppm (standard deviation [SD] 277\,ppm) in 2021/22 to 2,224\,ppm (SD 321\,ppm) in 2022/23.
 Relative humidity and temperature were similar at around 38\% and 20°C, respectively.
@@
 \textbf{(a)}~Distribution of respiratory viruses found in saliva.
-IF: influenza~A/B, HRV: human rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: human metapneumovirus, PIV: parainfluenza virus.
+IF: influenza~A/B virus, HRV: human rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: human metapneumovirus, PIV: parainfluenza virus.
 \textbf{(b)}~Positive samples in saliva and bioaerosols per study week.
-\textbf{(c)}~Probability of detecting any SARS-CoV-2 and non-SARS-CoV-2 in bioaerosols during a study week (posterior mean as dots, interquartile range as box, 95\%-CrI as error bars).}
+\textbf{(c)}~Probability of detecting any SARS-CoV-2 and non-SARS-CoV-2 viruses in bioaerosols during a study week (posterior mean as dots, interquartile range as box, 95\%-CrI as error bars).}
     \label{fig:comparison}
@@
 In winter 2021/22, we predominantly identified SARS-CoV-2 in saliva, air, and air filter samples.
-Conversely, in 2022/23, we primarily detected non-SARS-CoV-2 viruses, such as influenza and adenoviruses, in saliva samples, but these were rarely found in air or filter samples.
+Conversely, in 2022/23, we primarily detected non-SARS-CoV-2 viruses, such as influenza viruses and adenoviruses, in saliva samples, but these were rarely found in air or filter samples.
 
 Overall, the likelihood of molecular airborne detection was substantially higher for SARS-CoV-2 compared to non-SARS-CoV-2 viruses, even when we adjusted for covariates and differences between the studies.
-A plausible explanation is that SARS-CoV-2 can remain airborne for extended durations, thus facilitating long-range transmission.
-In contrast, the infrequent detection of non-SARS-CoV-2 viruses in the air indicates quicker sedimentation or inactivation.
+Besides differences in virus circulation in the population during the study periods, a plausible explanation is that SARS-CoV-2 can remain airborne for extended durations, thus facilitating long-range transmission, matching the observation of superspreading events during the pandemic.
+This is in contrast to other respiratory viruses, where airborne detection was found to be infrequent in our studies.
 Therefore, prolonged close contact may be relatively more important for transmission of respiratory viruses other than SARS-CoV-2, although close contact also facilitates transmission of SARS-CoV-2\cite{Leung2020NatMed,Lind2023NatCommun}.
@@
 Therefore, it is plausible that the difference in airborne detection may be due to differences in virus characteristics, particularly between SARS-CoV-2 and non-SARS-CoV-2 viruses, which may influence the generation and survival of airborne viral RNA\cite{Wang2021}.
-Non-SARS-CoV-2 respiratory viruses may produce fewer infectious bioaerosols, which can fall below the detection limit of current sampling devices\cite{Belser2023PLOSPath}.
+Non-SARS-CoV-2 respiratory virus infections may results in smaller amounts of exhaled bioaerosols, falling below the detection limit of current sampling devices\cite{Belser2023PLOSPath}.
 
@@
 Molecular detection of airborne SARS-CoV-2 was more frequent than other endemic respiratory viruses.
-Future studies should investigate the seasonality of SARS-CoV-2 and non-SARS-CoV-2 respiratory viruses and the contribution of prolonged close contact and contaminated air to the overall transmission of respiratory infections in congregate indoor settings.
+Future studies should investigate the seasonality of SARS-CoV-2 and non-SARS-CoV-2 respiratory viruses and the contribution of close contact versues airborne long-range transmission to overall transmission of respiratory infections in congregate indoor settings.
 

### 2024-02-16 doc/comparison-mcid12/supplements.tex

@@
 
-\begin{figure}
+\begin{figure}[!htpb]
     \centering
-    \includegraphics{}
-    \caption{Caption}
-    \label{fig:enter-label}
+    \includegraphics{results/comparison-viral-loads.png}
+    \caption{Comparison of Ct values for positive saliva samples between SARS-CoV-2 (CoV) and non-SARS-CoV-2 (non-CoV) viruses.
+A statistical comparison was performed pooling the Ct values across the two study periods (winter 2021/22 and 2022/23), with the p-value (p) based on a numerical two-sided two sample t-test.}
+    \label{fig:viral-loads}
 \end{figure}

### 2024-02-16 analysis/study_comaprison.Rmd
revisions

@@
   theme(axis.title.x = element_blank(),
-        legend.position = c(0.1, 0.87),
-        legend.background = element_rect(color = "black"),
+        legend.position = "top",
         plot.title.position = "plot",
         legend.title = element_blank()) +
-  guides(color = "none")
+  guides(color = guide_legend(nrow = 2))
 
@@
           "../results/comparison-viral-loads.png",
-          w = 12, h = 8)
+          w = 10, h = 12)
 ```

### 2025-03-26 analysis/study_comaprison.Rmd
commit remaining files

@@
 ggsave(plot = comparison, file = "../results/comparison.png", width = 16 / cm(1), height = 17 / cm(1))
+ggsave(plot = comparison, file = "../results/comparison.tiff", width = 16 / cm(1), height = 17 / cm(1))
 ```

### 2025-03-26 doc/comparison-mcid12/supplements.tex

@@
 \begin{table}[!htpb]
-    \caption{Differences in the study settings of 2022 and 2023.}
+    \caption{Comparison of the study settings in winter 2021/22 and 2022/23.}
     \label{tab:comp_study}
@@
         \midrule
-        Participants & 2~classes from School~1 sharing one classroom (half-class teaching) and 1~class from School~2 with a separate classroom; School~1 and~2 are in different cities, which have grown together to form a community (\ie the schools are located nearby) & two classes from School~2 with separate (two) classrooms; none of the two classes already participated in winter 2021/22 \\
-        Duration & 7 weeks per class (January 16 to March 11, 2023), excluding vacation & 7 weeks per class (January 24 to March 26, 2023), excluding vacation \\
+        Participants & two classes from School~1 sharing one classroom (half-class teaching) and one class from School~2 with a separate classroom; one teacher per lesson per classroom & two classes from School~2 with two separate classrooms (none of the two classes had already participated in winter 2021/23); one teacher per lesson per classroom \\
+        Setting & School~1 and~2 are in different towns, which have grown together over the years and are situated in a urban/peri-urban environment & School 2 is the same as in the previous winter \\
+        Duration & seven weeks per class (January 16 to March 11, 2023), excluding vacation & seven weeks per class (January 24 to March 26, 2023), excluding vacation \\
         Saliva testing & weekly (Wednesdays) & bi-weekly (Tuesdays and Thursdays) \\
@@
         HEPA-filters samples & collected once at the end of the study & collected twice after the phases where the air cleaner intervention was in place \\
-        Interventions & compulsory face mask wearing as mandated by the public health authorities at that time from the start of the study, for a total of 6 out of 14 study weeks; and installation of portable air cleaners towards the end of the study in the study classrooms for a total of 4 out of 14 study weeks; in total, 10 study weeks with interventions & installation of portable air cleaners in the study classrooms using a cross-over design, for a total of 7 out of 14 study weeks; in total 7 study weeks with interventions\\
+        Interventions & compulsory face mask wearing as mandated by the public health authorities at that time from the start of the study, for a total of six out of 14 study weeks; and installation of portable air cleaners towards the end of the study in the study classrooms for a total of four out of 14 study weeks; in total, ten study weeks with interventions & installation of portable air cleaners in the study classrooms using a cross-over design, for a total of seven out of 14 study weeks; in total seven study weeks with interventions\\
         Ventilation conditions & all classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system; passive window ventilation as per recommendations of the national public health authorities; in School~1 ventilation was additionally assisted by a CO$_2$ guided opener of a small window at the top  & all classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system; passive window ventilation solely at the discretion of the teachers \\
@@
 The binary variable \textbf{saliva} indicates if there was a positive saliva sample in the same week, with $\beta_1$ estimating whether airborne detection was more frequent if the same virus was already detected in saliva.
-This variable implicitly adjusts for community transmission, which would impact molecular detection both in saliva and airborne samples.
 The binary variable \textbf{CoV} indicates if the detected virus was SARS-CoV-2, thus our parameter of interest $\beta_2$ is estimating whether airborne detection of any SARS-CoV-2 is more likely than non-SARS-CoV-2 viruses.
